---
title: "Example used for the overview figures"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "overview_figure_example" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "5b90232d-7e46-48b6-b05c-dbaebffebb89")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/overview_"),
  echo = FALSE)
```

This script generates some example data used for the overview figures.

```{r packages}
library("conflicted")
library(dacomp)
library(ggplot2)
library(patchwork)
library(zCompositions)
```

```{r helper-functions}
clr_transform <- function(x) {
  log_x <- log(x)
  sweep(log_x, 1, rowMeans(log_x), "-")
}

shannon_index <- function(x) {
  p <- x / sum(x)
  p <- p[p > 0]
  -sum(p * log(p))
}

pooled_coordinate_variance <- function(y1, y2) {
  n1 <- nrow(y1)
  n2 <- nrow(y2)
  
  ((n1 - 1) * apply(y1, 2, stats::var) +
    (n2 - 1) * apply(y2, 2, stats::var)) / (n1 + n2 - 2)
}

comp_equiv_components <- function(y, group) {
  group <- droplevels(factor(group))
  
  if (nlevels(group) != 2) {
    stop("Compositional equivalence requires exactly two groups.")
  }
  
  group_levels <- levels(group)
  y1 <- y[group == group_levels[[1]], , drop = FALSE]
  y2 <- y[group == group_levels[[2]], , drop = FALSE]
  
  n1 <- nrow(y1)
  n2 <- nrow(y2)
  mean1 <- colMeans(y1)
  mean2 <- colMeans(y2)
  mean_diff <- mean1 - mean2
  pooled_var <- pooled_coordinate_variance(y1, y2)
  
  valid_taxa <- is.finite(pooled_var) & pooled_var > .Machine$double.eps
  scale_factor <- n1 * n2 / (n1 + n2)
  
  standardized_diff <- rep(NA_real_, length(mean_diff))
  names(standardized_diff) <- names(mean_diff)
  standardized_diff[valid_taxa] <- mean_diff[valid_taxa] / sqrt(pooled_var[valid_taxa])
  
  coordinate_stat <- scale_factor * standardized_diff^2
  names(coordinate_stat) <- names(mean_diff)
  statistic <- max(coordinate_stat[valid_taxa], na.rm = TRUE)
  
  list(
    statistic = statistic,
    n1 = n1,
    n2 = n2,
    group1 = group_levels[[1]],
    group2 = group_levels[[2]],
    mean1 = mean1,
    mean2 = mean2,
    mean_diff = mean_diff,
    pooled_var = pooled_var,
    standardized_diff = standardized_diff,
    coordinate_stat = coordinate_stat,
    valid_taxa = valid_taxa
  )
}
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Original data set

```{r}
set.seed(1)

# Rows shown in the schematic
row_names <- c(
  "1", "...", "k", "m", "...", "n1",
  "n1+1", "n1+2", "...", "l", "...", "n"
)

# First three displayed genera represent columns 1, i, and p in the figure
visible_counts <- matrix(
  c(
    127,  3,  8,
      0, 15,  4,
      0,  0,  1,
    242,  1,  0,
      5, 24,  0,
      0,  1,  0,
      0, 37, 14,
     14,  2,  3
  ),
  ncol = 3,
  byrow = TRUE
)

# Insert rows that are displayed as "..."
counts_first3 <- matrix(NA_integer_, nrow = length(row_names), ncol = 3)
counts_first3[c(1, 3, 4, 6, 7, 8, 10, 12), ] <- visible_counts

# Fill the "..." rows with arbitrary counts
counts_first3[is.na(counts_first3)] <- sample(
  0:20,
  sum(is.na(counts_first3)),
  replace = TRUE,
  prob = c(0.35, rep(0.65 / 20, 20))
)

# Add further genera to the right with heterogeneous abundances and zeros
n_extra <- 17

# Genus-specific baseline abundances: some genera rare, some abundant
genus_mean <- rgamma(n_extra, shape = 0.7, rate = 0.08)

# Sample-specific scaling: some samples have higher counts overall
sample_factor <- rgamma(length(row_names), shape = 2, rate = 2)

extra_counts <- sapply(seq_len(n_extra), function(j) {
  mu_j <- sample_factor * genus_mean[j]
  
  # Overdispersed counts
  x <- rnbinom(length(row_names), mu = mu_j, size = 1.2)
  
  # Add extra zeros, especially for rare genera
  zero_prob <- plogis(1.5 - 0.12 * genus_mean[j])
  x[rbinom(length(row_names), size = 1, prob = zero_prob) == 1] <- 0
  
  x
})

extra_counts <- as.matrix(extra_counts)

# Keep the two additional schematic rows sparse across the full data set.
sparse_row_idx <- match(c("m", "n1+2"), row_names)
extra_counts[sparse_row_idx, ] <- 0
extra_counts[sparse_row_idx, 1] <- 1

# Combined count matrix
counts <- cbind(counts_first3, extra_counts)

rownames(counts) <- row_names
colnames(counts) <- c(
  "Genus_1",
  "Genus_i",
  "Genus_p",
  paste0("Genus_extra_", seq_len(n_extra))
)

sample_group <- factor(
  ifelse(seq_along(row_names) <= match("n1", row_names), "Group 1", "Group 2"),
  levels = c("Group 1", "Group 2")
)

counts
```

**Genera shown in the thesis:**

```{r}
counts[, 1:3]
```

## Alpha diversity

```{r}
# Shannon index computed from the unfiltered, untransformed count matrix.
alpha_diversity <- data.frame(
  sample = rownames(counts),
  group = sample_group,
  shannon = apply(counts, 1, shannon_index),
  row.names = NULL
)

alpha_diversity_table <- alpha_diversity
alpha_diversity_table$shannon <- round(alpha_diversity_table$shannon, 2)
alpha_diversity_table
```

```{r alpha_diversity_boxplot, fig.width=4.5, fig.height=5.5, dpi=250}
group_colors <- c(
  "Group 1" = "#2456D9",
  "Group 2" = "#3F7F2A"
)

group_fills <- c(
  "Group 1" = "#E8F0FF",
  "Group 2" = "#EAF5E7"
)

ggplot(alpha_diversity, aes(x = group, y = shannon, fill = group, color = group)) +
  geom_boxplot(
    width = 0.55,
    linewidth = 0.8,
    alpha = 0.75,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    size = 5,
    alpha = 0.85
  ) +
  scale_color_manual(values = group_colors) +
  scale_fill_manual(values = group_fills) +
  labs(
    x = NULL,
    y = "Within-sample diversity"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.y = element_text(face = "bold", size = 16),
    axis.text.x = element_text(face = "bold", size = 15),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )
```


## Total sum scaling

```{r}
# Total sum scaling / relative abundances
lib_size <- rowSums(counts)
rel_abund <- sweep(counts, 1, lib_size, "/")

# Show the corresponding relative abundances
round(rel_abund, 3)
```

**Genera shown in the thesis:**

```{r}
# Or only first 3 relative abundances
round(rel_abund[, 1:3], 3)
```

## Multiplicative replacement

```{r, warning=FALSE}
# Multiplicative replacement as used in chapter 7:
# zero replacement on relative abundances with the global minimum non-zero
# relative abundance as detection limit and frac = 0.65.
zero_replacement_fraction <- 0.65
detection_limit <- min(rel_abund[rel_abund > 0], na.rm = TRUE)

detection_limit_mat <- matrix(
  detection_limit,
  nrow = nrow(rel_abund),
  ncol = ncol(rel_abund),
  dimnames = dimnames(rel_abund)
)

rel_abund_repl <- zCompositions::multRepl(
  rel_abund,
  label = 0,
  dl = detection_limit_mat,
  frac = zero_replacement_fraction,
  z.warning = 0.9,
  z.delete = FALSE
)

rel_abund_repl <- as.matrix(rel_abund_repl)
rownames(rel_abund_repl) <- rownames(rel_abund)
colnames(rel_abund_repl) <- colnames(rel_abund)

print("Row sums: ")
rowSums(rel_abund_repl)

round(rel_abund_repl, 3)
```

**Genera shown in the thesis:**

```{r}
round(rel_abund_repl[, 1:3], 3)
```

## CLR transformation

```{r}
# Centered log-ratio transformation after multiplicative replacement
counts_clr <- clr_transform(rel_abund_repl)

round(counts_clr, 3)
```

**Genera shown in the thesis:**

```{r}
round(counts_clr[, 1:3], 3)
```

## Beta diversity

```{r}
# Between-sample distances after multiplicative replacement and CLR transformation.
# This corresponds to Euclidean distance on the CLR scale.
sample_info <- data.frame(
  sample_id = paste0("S", seq_len(nrow(counts_clr))),
  sample = rownames(counts_clr),
  group = sample_group,
  row.names = NULL
)

counts_clr_beta <- counts_clr
rownames(counts_clr_beta) <- sample_info$sample_id

beta_distance <- dist(counts_clr_beta, method = "euclidean")
beta_distance_matrix <- as.matrix(beta_distance)

beta_distance_table <- beta_distance_matrix
rownames(beta_distance_table) <- sample_info$sample
colnames(beta_distance_table) <- sample_info$sample

round(beta_distance_table, 2)
```

```{r beta_diversity_plots, fig.width=9, fig.height=4.5, dpi=250}
# Fake display data with distances in a range typical for larger microbiome
# studies. These values are used only for the overview figure.
beta_display_ids <- paste0("D", seq_len(5))
beta_display_labels <- c("1", "...", "n1", "...", "n")

beta_distance_display_matrix <- matrix(
  c(
    0.00, 0.34, 0.47, 0.73, 0.68,
    0.34, 0.00, 0.41, 0.86, 0.77,
    0.47, 0.41, 0.00, 0.69, 0.74,
    0.73, 0.86, 0.69, 0.00, 0.36,
    0.68, 0.77, 0.74, 0.36, 0.00
  ),
  nrow = length(beta_display_ids),
  byrow = TRUE,
  dimnames = list(beta_display_ids, beta_display_ids)
)

beta_distance_plot_df <- as.data.frame(as.table(beta_distance_display_matrix))
names(beta_distance_plot_df) <- c("sample_1", "sample_2", "distance")
beta_distance_plot_df$sample_1 <- factor(
  beta_distance_plot_df$sample_1,
  levels = beta_display_ids
)
beta_distance_plot_df$sample_2 <- factor(
  beta_distance_plot_df$sample_2,
  levels = rev(beta_display_ids)
)
beta_distance_plot_df$distance_label <- ifelse(
  beta_distance_plot_df$distance == 0,
  "0",
  sprintf("%.2f", beta_distance_plot_df$distance)
)

beta_display_label_lookup <- setNames(beta_display_labels, beta_display_ids)

# Fake ordination coordinates to mimic a larger data set.
set.seed(123)
n_fake_group <- 26
pcoa_df <- data.frame(
  Axis1 = c(
    rnorm(n_fake_group, mean = -1.15, sd = 0.42),
    rnorm(n_fake_group, mean = 1.25, sd = 0.48)
  ),
  Axis2 = c(
    rnorm(n_fake_group, mean = 0.10, sd = 0.78),
    rnorm(n_fake_group, mean = -0.05, sd = 0.72)
  ),
  group = factor(
    rep(c("Group 1", "Group 2"), each = n_fake_group),
    levels = c("Group 1", "Group 2")
  )
)

distance_heatmap <- ggplot(
  beta_distance_plot_df,
  aes(x = sample_1, y = sample_2, fill = distance)
) +
  geom_tile(color = "grey55", linewidth = 0.25) +
  geom_text(aes(label = distance_label), size = 5) +
  scale_x_discrete(labels = beta_display_label_lookup) +
  scale_y_discrete(labels = beta_display_label_lookup) +
  scale_fill_gradient(low = "white", high = "tomato3", guide = "none") +
  coord_equal() +
  labs(
    x = NULL,
    y = NULL,
    fill = "Distance",
    title = "Between-sample diversity"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "none"
  )

ordination_plot <- ggplot(pcoa_df, aes(x = Axis1, y = Axis2, color = group)) +
  geom_point(size = 3.1, alpha = 0.9) +
  scale_color_manual(values = group_colors) +
  labs(
    x = "Axis 1",
    y = "Axis 2",
    color = NULL,
    title = "Ordination of sample distances"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

distance_heatmap + ordination_plot +
  plot_layout(widths = c(1.15, 1), guides = "collect") &
  theme(legend.position = "bottom")
```

## Regression

```{r}
# Fake coefficient matrix for the overview figure. Rows represent model terms,
# columns represent taxa. Values are illustrative feature-level effects.
regression_terms <- c("Intercept", "Group", "Covariate 1", "...", "Covariate q")
regression_taxa <- c("Genus 1", "Genus 2", "...", "Genus p")

regression_coef_matrix <- matrix(
  c(
     0.05, -0.42,  0.00, -0.18,
    -0.68, -0.31,  0.08,  0.55,
     0.74,  0.03,  0.10,  0.69,
     0.00,  0.12,  0.08,  0.06,
    -0.36, -0.10, -0.22,  0.27
  ),
  nrow = length(regression_terms),
  byrow = TRUE,
  dimnames = list(regression_terms, regression_taxa)
)

round(regression_coef_matrix, 2)
```

```{r regression_coefficient_heatmap, fig.width=4.6, fig.height=4.2, dpi=250}
regression_coef_df <- as.data.frame(as.table(regression_coef_matrix))
names(regression_coef_df) <- c("term", "genus", "coefficient")
regression_coef_df$term <- factor(
  regression_coef_df$term,
  levels = rev(regression_terms)
)
regression_coef_df$genus <- factor(
  regression_coef_df$genus,
  levels = regression_taxa
)

ggplot(regression_coef_df, aes(x = genus, y = term, fill = coefficient)) +
  geom_tile(color = "grey55", linewidth = 0.25) +
  scale_fill_gradient2(
    low = "#2456D9",
    mid = "white",
    high = "#D7191C",
    midpoint = 0,
    limits = c(-0.8, 0.8),
    breaks = c(-0.8, 0, 0.8),
    labels = c("-", "0", "+"),
    name = NULL
  ) +
  coord_equal() +
  labs(
    x = NULL,
    y = NULL,
    title = "Coefficient matrix"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  )
```

## Compositional equivalence

```{r}
# Real compositional equivalence components for the example data.
# The statistic is the maximum coordinate-wise standardized squared difference
# between group CLR means.
comp_equiv <- comp_equiv_components(
  y = counts_clr,
  group = sample_group
)

clr_group_means <- rbind(
  comp_equiv$mean1,
  comp_equiv$mean2
)
rownames(clr_group_means) <- c(comp_equiv$group1, comp_equiv$group2)

round(clr_group_means, 3)
```

**Genera shown in the thesis:**

```{r}
round(clr_group_means[, 1:3], 3)
```

```{r}
comp_equiv_contrasts <- data.frame(
  genus = names(comp_equiv$coordinate_stat),
  mean_group1 = as.numeric(comp_equiv$mean1),
  mean_group2 = as.numeric(comp_equiv$mean2),
  mean_diff_clr = as.numeric(comp_equiv$mean_diff),
  standardized_diff = as.numeric(comp_equiv$standardized_diff),
  m_i = as.numeric(comp_equiv$coordinate_stat),
  valid_genus = as.logical(comp_equiv$valid_taxa),
  row.names = NULL
)

comp_equiv_contrasts <- comp_equiv_contrasts[order(
  comp_equiv_contrasts$m_i,
  decreasing = TRUE,
  na.last = TRUE
), ]

comp_equiv_statistic <- max(comp_equiv_contrasts$m_i, na.rm = TRUE)
comp_equiv_statistic

comp_equiv_contrasts
```

```{r compositional_equivalence_contrasts, fig.width=5, fig.height=3, dpi=250}
comp_equiv_plot_df <- comp_equiv_contrasts[comp_equiv_contrasts$valid_genus, ]
comp_equiv_plot_df <- comp_equiv_plot_df[order(comp_equiv_plot_df$m_i), ]
comp_equiv_plot_df$genus <- factor(
  comp_equiv_plot_df$genus,
  levels = comp_equiv_plot_df$genus
)

comp_equiv_plot_df <- comp_equiv_plot_df[1:10, ]

comp_equiv_plot_df$is_maximum <- comp_equiv_plot_df$m_i == max(
  comp_equiv_plot_df$m_i,
  na.rm = TRUE
)

ggplot(comp_equiv_plot_df, aes(x = m_i, y = genus)) +
  geom_segment(aes(x = 0, xend = m_i, yend = genus), color = "grey72", 
               linewidth = 1) +
  geom_point(aes(color = is_maximum), size = 3) +
  scale_color_manual(
    values = c("FALSE" = "grey35", "TRUE" = "#D55E00"),
    guide = "none"
  ) +
  labs(
    x = expression("Coordinate-wise contrast " * m[i]),
    y = NULL,
    #title = expression("Compositional equivalence contrasts: " * M == max[i](m[i]))
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(face = "bold")
  )
```


## Differential abundance with dacomp

```{r differential_abundance_dacomp, warning=FALSE, message=FALSE}
# dacomp uses the raw count matrix. Here we run a small illustrative dacomp
# analysis and print the dacomp-rarefied count matrix returned by dacomp.test().
set.seed(42)

dacomp_counts <- counts
rownames(dacomp_counts) <- paste0("S", seq_len(nrow(dacomp_counts)))

dacomp_group <- sample_group

dacomp_refs <- dacomp::dacomp.select_references(
  X = dacomp_counts,
  minimal_TA = 1,
  verbose = FALSE
)

dacomp_reference_taxa <- dacomp_refs$selected_references

if (length(dacomp_reference_taxa) == 0) {
  dacomp_reference_taxa <- head(
    order(colSums(dacomp_counts), decreasing = TRUE),
    min(5, ncol(dacomp_counts))
  )
}

dacomp_res <- dacomp::dacomp.test(
  X = dacomp_counts,
  y = dacomp_group,
  ind_reference_taxa = dacomp_reference_taxa,
  test = dacomp::DACOMP.TEST.NAME.WILCOXON,
  nr_perm = 1000,
  q = 0.1,
  verbose = FALSE,
  return_rarefied_values = TRUE,
  Test_All = TRUE
)

dacomp_transformed_matrix <- dacomp_res$rarefied_counts
dimnames(dacomp_transformed_matrix) <- dimnames(dacomp_counts)

dacomp_transformed_table <- dacomp_transformed_matrix
rownames(dacomp_transformed_table) <- rownames(counts)

dacomp_transformed_table
```

**Genera shown in the thesis:**

```{r}
dacomp_transformed_table[, c("Genus_1", "Genus_i", "Genus_p")]
```

```{r differential_abundance_boxplots, fig.width=8, fig.height=4, dpi=250}
dacomp_plot_genera <- c("Genus_1", "Genus_i", "Genus_p")
dacomp_plot_labels <- c(
  "Genus_1" = "Genus 1",
  "Genus_i" = "Genus i",
  "Genus_p" = "Genus p"
)

dacomp_boxplot_df <- as.data.frame(as.table(
  dacomp_transformed_matrix[, dacomp_plot_genera, drop = FALSE]
))
names(dacomp_boxplot_df) <- c("sample_id", "genus", "dacomp_value")
dacomp_boxplot_df$group <- rep(dacomp_group, times = length(dacomp_plot_genera))
dacomp_boxplot_df$genus <- factor(
  dacomp_boxplot_df$genus,
  levels = dacomp_plot_genera,
  labels = dacomp_plot_labels[dacomp_plot_genera]
)

ggplot(
  dacomp_boxplot_df,
  aes(x = group, y = dacomp_value, fill = group, color = group)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.75,
    linewidth = 0.7,
    width = 0.58
  ) +
  geom_jitter(width = 0.08, height = 0, size = 2.4, alpha = 0.85) +
  facet_wrap(~ genus, scales = "free_y", nrow = 1) +
  scale_color_manual(values = group_colors) +
  scale_fill_manual(values = group_fills) +
  labs(
    x = NULL,
    y = "dacomp-rarefied count"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```



## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
