---
title: "Example used for the overview figures"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "overview_figure_example" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "5b90232d-7e46-48b6-b05c-dbaebffebb89")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/overview_"),
  echo = FALSE)
```

This script generates some example data used for the overview figures.

```{r packages}
library("conflicted")
library(ggplot2)
library(patchwork)
library(zCompositions)
```

```{r helper-functions}
clr_transform <- function(x) {
  log_x <- log(x)
  sweep(log_x, 1, rowMeans(log_x), "-")
}

shannon_index <- function(x) {
  p <- x / sum(x)
  p <- p[p > 0]
  -sum(p * log(p))
}
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Original data set

```{r}
set.seed(1)

# Rows shown in the schematic
row_names <- c(
  "1", "...", "k", "m", "...", "n1",
  "n1+1", "n1+2", "...", "l", "...", "n"
)

# First three displayed genera represent columns 1, i, and p in the figure
visible_counts <- matrix(
  c(
    127,  3,  8,
      0, 15,  4,
      0,  0,  1,
    242,  1,  0,
      5, 24,  0,
      0,  1,  0,
      0, 37, 14,
     14,  2,  3
  ),
  ncol = 3,
  byrow = TRUE
)

# Insert rows that are displayed as "..."
counts_first3 <- matrix(NA_integer_, nrow = length(row_names), ncol = 3)
counts_first3[c(1, 3, 4, 6, 7, 8, 10, 12), ] <- visible_counts

# Fill the "..." rows with arbitrary counts
counts_first3[is.na(counts_first3)] <- sample(
  0:20,
  sum(is.na(counts_first3)),
  replace = TRUE,
  prob = c(0.35, rep(0.65 / 20, 20))
)

# Add further genera to the right with heterogeneous abundances and zeros
n_extra <- 17

# Genus-specific baseline abundances: some genera rare, some abundant
genus_mean <- rgamma(n_extra, shape = 0.7, rate = 0.08)

# Sample-specific scaling: some samples have higher counts overall
sample_factor <- rgamma(length(row_names), shape = 2, rate = 2)

extra_counts <- sapply(seq_len(n_extra), function(j) {
  mu_j <- sample_factor * genus_mean[j]
  
  # Overdispersed counts
  x <- rnbinom(length(row_names), mu = mu_j, size = 1.2)
  
  # Add extra zeros, especially for rare genera
  zero_prob <- plogis(1.5 - 0.12 * genus_mean[j])
  x[rbinom(length(row_names), size = 1, prob = zero_prob) == 1] <- 0
  
  x
})

extra_counts <- as.matrix(extra_counts)

# Keep the two additional schematic rows sparse across the full data set.
sparse_row_idx <- match(c("m", "n1+2"), row_names)
extra_counts[sparse_row_idx, ] <- 0
extra_counts[sparse_row_idx, 1] <- 1

# Combined count matrix
counts <- cbind(counts_first3, extra_counts)

rownames(counts) <- row_names
colnames(counts) <- c(
  "Genus_1",
  "Genus_i",
  "Genus_p",
  paste0("Genus_extra_", seq_len(n_extra))
)

sample_group <- factor(
  ifelse(seq_along(row_names) <= match("n1", row_names), "Group 1", "Group 2"),
  levels = c("Group 1", "Group 2")
)

counts
```

**Genera shown in the thesis:**

```{r}
counts[, 1:3]
```

## Alpha diversity

```{r}
# Shannon index computed from the unfiltered, untransformed count matrix.
alpha_diversity <- data.frame(
  sample = rownames(counts),
  group = sample_group,
  shannon = apply(counts, 1, shannon_index),
  row.names = NULL
)

alpha_diversity_table <- alpha_diversity
alpha_diversity_table$shannon <- round(alpha_diversity_table$shannon, 2)
alpha_diversity_table
```

```{r alpha_diversity_boxplot, fig.width=4.5, fig.height=5.5, dpi=250}
group_colors <- c(
  "Group 1" = "#2456D9",
  "Group 2" = "#3F7F2A"
)

group_fills <- c(
  "Group 1" = "#E8F0FF",
  "Group 2" = "#EAF5E7"
)

ggplot(alpha_diversity, aes(x = group, y = shannon, fill = group, color = group)) +
  geom_boxplot(
    width = 0.55,
    linewidth = 0.8,
    alpha = 0.75,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    size = 5,
    alpha = 0.85
  ) +
  scale_color_manual(values = group_colors) +
  scale_fill_manual(values = group_fills) +
  labs(
    x = NULL,
    y = "Within-sample diversity"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.y = element_text(face = "bold", size = 16),
    axis.text.x = element_text(face = "bold", size = 15),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )
```


## Total sum scaling

```{r}
# Total sum scaling / relative abundances
lib_size <- rowSums(counts)
rel_abund <- sweep(counts, 1, lib_size, "/")

# Show the corresponding relative abundances
round(rel_abund, 3)
```

**Genera shown in the thesis:**

```{r}
# Or only first 3 relative abundances
round(rel_abund[, 1:3], 3)
```

## Multiplicative replacement

```{r, warning=FALSE}
# Multiplicative replacement as used in chapter 7:
# zero replacement on relative abundances with the global minimum non-zero
# relative abundance as detection limit and frac = 0.65.
zero_replacement_fraction <- 0.65
detection_limit <- min(rel_abund[rel_abund > 0], na.rm = TRUE)

detection_limit_mat <- matrix(
  detection_limit,
  nrow = nrow(rel_abund),
  ncol = ncol(rel_abund),
  dimnames = dimnames(rel_abund)
)

rel_abund_repl <- zCompositions::multRepl(
  rel_abund,
  label = 0,
  dl = detection_limit_mat,
  frac = zero_replacement_fraction,
  z.warning = 0.9,
  z.delete = FALSE
)

rel_abund_repl <- as.matrix(rel_abund_repl)
rownames(rel_abund_repl) <- rownames(rel_abund)
colnames(rel_abund_repl) <- colnames(rel_abund)

print("Row sums: ")
rowSums(rel_abund_repl)

round(rel_abund_repl, 3)
```

**Genera shown in the thesis:**

```{r}
round(rel_abund_repl[, 1:3], 3)
```

## CLR transformation

```{r}
# Centered log-ratio transformation after multiplicative replacement
counts_clr <- clr_transform(rel_abund_repl)

round(counts_clr, 3)
```

**Genera shown in the thesis:**

```{r}
round(counts_clr[, 1:3], 3)
```

## Beta diversity

```{r}
# Between-sample distances after multiplicative replacement and CLR transformation.
# This corresponds to Euclidean distance on the CLR scale.
sample_info <- data.frame(
  sample_id = paste0("S", seq_len(nrow(counts_clr))),
  sample = rownames(counts_clr),
  group = sample_group,
  row.names = NULL
)

counts_clr_beta <- counts_clr
rownames(counts_clr_beta) <- sample_info$sample_id

beta_distance <- dist(counts_clr_beta, method = "euclidean")
beta_distance_matrix <- as.matrix(beta_distance)

beta_distance_table <- beta_distance_matrix
rownames(beta_distance_table) <- sample_info$sample
colnames(beta_distance_table) <- sample_info$sample

round(beta_distance_table, 2)
```

```{r beta_diversity_plots, fig.width=9, fig.height=4.5, dpi=250}
beta_display_idx <- c(1, 2, 8, 11, 12)
beta_display_ids <- sample_info$sample_id[beta_display_idx]
beta_display_labels <- c("1", "...", "n1+2", "...", "n")

beta_distance_display_matrix <- beta_distance_matrix[
  beta_display_ids,
  beta_display_ids,
  drop = FALSE
]

beta_distance_plot_df <- as.data.frame(as.table(beta_distance_display_matrix))
names(beta_distance_plot_df) <- c("sample_1", "sample_2", "distance")
beta_distance_plot_df$sample_1 <- factor(
  beta_distance_plot_df$sample_1,
  levels = beta_display_ids
)
beta_distance_plot_df$sample_2 <- factor(
  beta_distance_plot_df$sample_2,
  levels = rev(beta_display_ids)
)
beta_distance_plot_df$distance_label <- ifelse(
  beta_distance_plot_df$distance == 0,
  "0",
  sprintf("%.2f", beta_distance_plot_df$distance)
)

beta_display_label_lookup <- setNames(beta_display_labels, beta_display_ids)

pcoa_object <- cmdscale(beta_distance, k = 2, eig = TRUE)
pcoa_df <- data.frame(
  sample_id = rownames(pcoa_object$points),
  Axis1 = pcoa_object$points[, 1],
  Axis2 = pcoa_object$points[, 2],
  row.names = NULL
)
pcoa_df <- merge(pcoa_df, sample_info, by = "sample_id", sort = FALSE)

distance_heatmap <- ggplot(
  beta_distance_plot_df,
  aes(x = sample_1, y = sample_2, fill = distance)
) +
  geom_tile(color = "grey55", linewidth = 0.25) +
  geom_text(aes(label = distance_label), size = 4) +
  scale_x_discrete(labels = beta_display_label_lookup) +
  scale_y_discrete(labels = beta_display_label_lookup) +
  scale_fill_gradient(low = "white", high = "tomato3", guide = "none") +
  coord_equal() +
  labs(
    x = NULL,
    y = NULL,
    fill = "Distance",
    title = "Between-sample diversity"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "none"
  )

ordination_plot <- ggplot(pcoa_df, aes(x = Axis1, y = Axis2, color = group)) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = group_colors) +
  labs(
    x = "Axis 1",
    y = "Axis 2",
    color = NULL,
    title = "Ordination of sample distances"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

distance_heatmap + ordination_plot +
  plot_layout(widths = c(1.15, 1), guides = "collect") &
  theme(legend.position = "bottom")
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
