---
title: "Alpha diversity analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "02_alpha_diversity" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "41b33e9b-a351-49eb-bdc8-cdeddd77796c")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params, "/alpha_"),
  echo = FALSE)

conflicts_prefer(dplyr::filter)
```

## Load packages

```{r packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(ggtext)
library(ggpubr)
library(patchwork)

# Diversity packages
library(breakaway)
library(DivNet)
library(progress)
library(permApprox)

# Parallel computing
library(future)
library(future.apply)
library(progressr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data

```{r load_data}
phylo_asv   <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_asv.rds"))
phylo_genus <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_genus.rds"))
sampdata <- sample_data(phylo_asv) %>% as("data.frame")
otutab_asv   <- as(otu_table(phylo_asv), "matrix")
otutab_genus <- as(otu_table(phylo_genus), "matrix")
```

### ASV level

```{r}
phylo_asv
```

### Genus level

```{r}
phylo_genus
```


## Functions

### Function for significance plot

```{r, echo=TRUE}
modify_div_plot <- function(strip_chart, xlab = "", breaks = NULL, labels = NULL,
                            binwidth = 0.5, dotsize = 1, ylim = c(0, 75),
                            ylab = "", ybreaks = NULL,
                            textsize = 18, title = "") {
  strip_chart <- strip_chart +
    stat_summary(fun.data = median_hilow, fun.args=0.50, show.legend=FALSE,
                 geom="crossbar", alpha=0.25, width=0.7) +
    geom_dotplot(binaxis = "y", binwidth = binwidth, 
                 stackdir = "center", width = 0.5, alpha = 0.7,
                 show.legend = FALSE, dotsize = dotsize) +
    labs(x=NULL,
         y=ylab) +
    scale_fill_manual(values = cols) +
    scale_x_discrete(name = xlab,
                     breaks = breaks, 
                     labels = labels) +
    theme_classic() +
    theme(axis.text.x = element_markdown(),
          axis.text.y = element_markdown(),
          legend.text = element_markdown(),
          text = element_text(size = textsize)) +
    ggtitle(title)
  
  if (!is.null(ybreaks)) {
    strip_chart <- strip_chart + 
      scale_y_continuous(breaks=ybreaks, limits = ylim)
  } else {
    strip_chart <- strip_chart + 
      scale_y_continuous(limits = ylim)
  }
  strip_chart
}

```

### Function for significance line

```{r, echo=TRUE}
add_sig_line <- function(strip_chart, pval, diversity,
                         space_line = 0.05, space_lab = 0.05, nfeat = nfeat, 
                         linewidth = 0.8, size = 5) {
  
  if (nfeat == 2) {
    pmat_x <- c(1)
    pmat_y <- c(1)
    line_left <- c(1)
    line_right <- c(2)
    lab_x <- c(1.5)
  } else if (nfeat == 3) {
    pmat_x <- c(1, 2, 2)
    pmat_y <- c(1, 2, 1)
    line_left <- c(1, 2, 1)
    line_right <- c(2, 3, 3)
    lab_x <- c(1.5, 2.5, 2)
  } else {
    pmat_x <- c(1, 2, 3, 2, 3, 3)
    pmat_y <- c(1, 2, 3, 1, 2, 1)
    line_left <- c(1, 2, 3, 1, 2, 1)
    line_right <- c(2, 3, 4, 3, 4, 4)
    lab_x <- c(1.5, 2.5, 3.5, 2, 3, 2.5)
  }
  
  pos_line <- max(diversity) - space_line + 0.1 * space_line
  pos_lab <- pos_line + space_lab
  
  for (i in seq_along(pmat_x)) {
    if (is.null(dim(pval))) {
      label <- pval
    } else {
      label <- pval[pmat_x[i], pmat_y[i]]
    }
    
    if (label <= 0.05) {
      pos_line <- pos_line + space_line
      pos_lab <- pos_lab + space_line
      
      strip_chart <- strip_chart +
        geom_line(data = tibble(x = c(line_left[i], line_right[i]), 
                                y = c(pos_line, pos_line)), 
                  aes(x=x, y=y),
                  inherit.aes = FALSE, 
                  linewidth = linewidth) +
        #geom_text(x = lab_x[i], y = pos_lab, label = signif(label, digits = 3))
        annotate("text", lab_x[i], y = pos_lab, 
                 label = signif(label, digits = 3), size = size)
    } 
  }
  strip_chart
}
```

### Permutation test functions

```{r, echo=TRUE}
make_perm_chunks <- function(n_perm, n_chunks = NULL) {
  if (is.null(n_chunks)) {
    n_chunks <- min(parallelly::availableCores(), n_perm)
  }
  split(seq_len(n_perm), cut(seq_len(n_perm), breaks = n_chunks, labels = FALSE))
}

perm_kruskal <- function(x, g, n_perm = 9999, seed = NULL,
                         n_chunks = NULL, progress = TRUE) {
  g    <- droplevels(as.factor(g))
  keep <- complete.cases(x, g)
  x    <- x[keep]
  g    <- droplevels(g[keep])

  # Asymptotic test called once (for p_asymp only)
  obs     <- kruskal.test(x ~ g)
  p_asymp <- obs$p.value

  # Pre-compute fixed quantities — ranks never change under permutation
  r    <- rank(x)
  n    <- length(x)
  gi   <- as.integer(g)            # integer group labels, faster than factor
  K    <- nlevels(g)
  nj   <- tabulate(gi, nbins = K)  # group sizes
  coef <- 12 / (n * (n + 1L))
  adj  <- 3 * (n + 1L)

  # Observed H with the same formula used in permutations (ensures fair comparison)
  Rj_obs <- c(rowsum(r, gi))
  obs_H  <- coef * sum(Rj_obs^2 / nj) - adj

  chunks <- make_perm_chunks(n_perm, n_chunks)

  if (progress) {
    prog <- progressr::progressor(steps = length(chunks))
  }

  perm_list <- future.apply::future_lapply(
    chunks,
    function(idx) {
      out <- numeric(length(idx))
      for (b in seq_along(idx)) {
        Rj     <- c(rowsum(r, sample(gi)))
        out[b] <- coef * sum(Rj^2 / nj) - adj
      }
      if (progress) prog()
      out
    },
    future.seed = if (is.null(seed)) TRUE else seed
  )
  
  perm_H <- unlist(perm_list, use.names = FALSE)
  p_perm <- (sum(perm_H >= obs_H) + 1) / (n_perm + 1)

  pa <- permApprox::perm_approx(
    obs_stats     = obs_H,
    perm_stats    = matrix(perm_H, ncol = 1),
    alternative   = "greater",
    adjust_method = "none",
    verbose       = FALSE,
    gpd_ctrl      = permApprox::make_gpd_ctrl(
      sample_size     = min(nj),
      # With large n_perm the discreteness screen incorrectly flags the KW H
      # distribution (tied ranks) and falls back to empirical; disable it.
      discrete_screen = FALSE,
      exceed0_min     = 250L,
      # For large permutation numbers, the number of starting exceedances is 
      # reduced below 25% to reduce runtime.
      exceed0         = min(0.25, (5000 / n_perm))
    )
  )

  list(
    H        = obs_H,
    H_perm   = perm_H,
    p_asymp  = p_asymp,
    p_perm   = p_perm,
    p_approx = pa$p_unadjusted[1]
  )
}

perm_pairwise <- function(x, g, n_perm = 9999, p_adjust = "BH",
                          seed = NULL, n_chunks = NULL, progress = TRUE) {
  g <- droplevels(as.factor(g))
  keep <- complete.cases(x, g)
  x <- x[keep]
  g <- droplevels(g[keep])
  
  lvls  <- levels(g)
  pairs <- combn(lvls, 2, simplify = FALSE)
  
  p_vals        <- numeric(length(pairs))
  p_vals_approx <- numeric(length(pairs))
  
  for (i in seq_along(pairs)) {
    pair <- pairs[[i]]
    idx  <- g %in% pair

    xi <- x[idx]
    gi <- droplevels(g[idx])

    n1 <- sum(gi == levels(gi)[1])
    n2 <- sum(gi == levels(gi)[2])
    EW <- n1 * n2 / 2

    # Pre-compute fixed quantities for fast W calculation
    ri     <- rank(xi)                  # ranks fixed across all permutations
    gi_int <- as.integer(gi)            # 1 = first level, 2 = second level
    # W = (sum of ranks in group 1) - n1*(n1+1)/2  [Mann-Whitney U convention]
    W_obs <- sum(ri[gi_int == 1L]) - n1 * (n1 + 1) / 2
    obs_T <- abs(W_obs - EW)

    chunks <- make_perm_chunks(n_perm, n_chunks)

    if (progress) {
      message("Running pairwise permutations for: ", pair[1], " vs ", pair[2])
      prog <- progressr::progressor(steps = length(chunks))
    }

    pair_seed <- if (is.null(seed)) TRUE else seed + i

    perm_list <- future.apply::future_lapply(
      chunks,
      function(idx_perm) {
        out      <- numeric(length(idx_perm))
        rank_adj <- n1 * (n1 + 1) / 2
        for (b in seq_along(idx_perm)) {
          gi_perm <- sample(gi_int)
          W_perm  <- sum(ri[gi_perm == 1L]) - rank_adj
          out[b]  <- abs(W_perm - EW)
        }
        if (progress) prog()
        out
      },
      future.seed = pair_seed
    )
    
    perm_T <- unlist(perm_list, use.names = FALSE)
    
    p_vals[i] <- (sum(perm_T >= obs_T) + 1) / (n_perm + 1)
    
    pa <- permApprox::perm_approx(
      obs_stats     = obs_T,
      perm_stats    = matrix(perm_T, ncol = 1),
      alternative   = "greater",
      adjust_method = "none",
      verbose       = FALSE,
      gpd_ctrl      = permApprox::make_gpd_ctrl(
        sample_size     = min(n1, n2),
        discrete_screen = FALSE,
        exceed0_min     = 250L,
        exceed0         = min(0.25, (5000 / n_perm))
      )
    )
    
    p_vals_approx[i] <- pa$p_unadjusted[1]
  }
  
  p_adj        <- p.adjust(p_vals, method = p_adjust)
  p_adj_approx <- p.adjust(p_vals_approx, method = p_adjust)
  
  make_mat <- function(vals) {
    mat <- matrix(
      NA_real_,
      nrow = length(lvls) - 1,
      ncol = length(lvls) - 1,
      dimnames = list(lvls[-1], lvls[-length(lvls)])
    )
    
    for (i in seq_along(pairs)) {
      r <- which(lvls == pairs[[i]][2]) - 1
      c <- which(lvls == pairs[[i]][1])
      if (r >= 1 && c <= ncol(mat)) mat[r, c] <- vals[i]
    }
    
    mat
  }
  
  list(
    p_matrix         = make_mat(p_adj),
    p_matrix_approx  = make_mat(p_adj_approx),
    p_raw            = p_vals,
    p_raw_approx     = p_vals_approx,
    pairs            = pairs
  )
}
```


### Plot functions

```{r}
plot_richness <- function(div_dataframe, feature_name, 
                          xlab = "", breaks = NULL, labels = NULL,
                          binwidth = 0.4, dotsize = 1.2, 
                          ylab = "Richness (breakaway)", 
                          ybreaks = NULL, title = "",
                          ylim = c(0, 75), textsize = 18,
                          space_line = 2, space_lab = 5,
                          ptest = NULL,
                          p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=richness, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(richness ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$richness,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$richness,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}

plot_shannon <- function(div_dataframe, feature_name, 
                         xlab = "", breaks = NULL, labels = NULL,
                         binwidth = 0.04, dotsize = 1, ylim = c(0, 4), 
                         space_line = 0.3, space_lab = 0.15,
                         ylab = "Shannon index (DivNet)", 
                         ybreaks = NULL, title = "", textsize = 18,
                         ptest = NULL,
                         p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=shannon, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(shannon ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$shannon,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$shannon,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}

plot_gini <- function(div_dataframe, feature_name, 
                      xlab = "", breaks = NULL, labels = NULL,
                      binwidth = 0.01, dotsize = 1, ylim = c(0, 1.1), 
                      space_line = 0.08, space_lab = 0.04,
                      ylab = "Gini-Simpson index (DivNet)", 
                      ybreaks = NULL, title = "", textsize = 18,
                      ptest = NULL,
                      p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=gini, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(gini ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$gini,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$gini,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```


## Colors

```{r}
cols = c("#009E73", "#56B4E9", "#E69F00", "#D55E00", 
         "#CC79A7", "#F0E442", "#0072B2", "gray")
```

## Compute richness (on ASV level)

### Common richness

Computing alpha diversity the common way (not model-based) using the `estimate_richness` function from `phyloseq` package.

```{r}
alpha_div_common_asv <- estimate_richness(phylo_asv)
head(alpha_div_common_asv)
```

### Breakaway richness

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_asv)
```

```{r}
breakaway_rich_asv <- 
  data.frame(SampleID = names(frequencytablelist),
             richness = unlist(lapply(frequencytablelist, 
                                      function(x) breakaway(x)$estimate))) 
```

### Comparison: classical vs. breakaway

Observed (classical) richness simply counts the number of non-zero ASVs in a sample,
so it is directly inflated by sequencing depth. Breakaway estimates the true number of
taxa by modelling unobserved low-abundance species, and should therefore be more robust
to variation in library size. Here we compare both metrics against library size
(total reads per sample) to assess how strongly each is confounded by sequencing depth.

```{r richness_libsize_data_asv}
lib_size_asv <- data.frame(
  SampleID     = names(sample_sums(phylo_asv)),
  library_size = as.numeric(sample_sums(phylo_asv))
)

obs_rich_asv <- data.frame(
  SampleID = rownames(alpha_div_common_asv),
  observed  = alpha_div_common_asv$Observed
)

richness_comp_asv <- lib_size_asv %>%
  inner_join(obs_rich_asv, by = "SampleID") %>%
  inner_join(rename(breakaway_rich_asv, breakaway = richness), by = "SampleID")

head(richness_comp_asv)
```

**Spearman correlation between library size and each richness metric:**

```{r richness_libsize_cor_asv}
cor_obs_asv <- cor.test(richness_comp_asv$library_size, richness_comp_asv$observed,
                        method = "spearman", exact = FALSE)
cor_bw_asv  <- cor.test(richness_comp_asv$library_size, richness_comp_asv$breakaway,
                        method = "spearman", exact = FALSE)

tibble(
  Metric   = c("Observed richness", "Richness (breakaway)"),
  rho      = round(c(cor_obs_asv$estimate, cor_bw_asv$estimate), 3),
  p_value  = signif(c(cor_obs_asv$p.value, cor_bw_asv$p.value), 3)
)
```

```{r richness_vs_libsize_asv, fig.width=10, fig.height=5}
richness_comp_asv %>%
  pivot_longer(c(observed, breakaway),
               names_to  = "method",
               values_to = "richness") %>%
  mutate(method = factor(method,
                         levels = c("observed", "breakaway"),
                         labels = c("Observed richness", "Richness (breakaway)"))) %>%
  ggplot(aes(x = library_size, y = richness)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  ggpubr::stat_cor(method = "spearman",
                   label.x.npc = "left", label.y.npc = "top", size = 4.5) +
  facet_wrap(~method, scales = "free_y") +
  labs(x = "Library size (reads)", y = "Richness") +
  theme_classic(base_size = 14)
```

```{r observed_vs_breakaway_asv, fig.width=7, fig.height=5}
ggplot(richness_comp_asv, aes(x = observed, y = breakaway, color = library_size)) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, color = "grey40") +
  geom_point(alpha = 1, size = 3) +
  ggpubr::stat_cor(method = "spearman",
                   label.x.npc = "left", label.y.npc = "top", size = 4.5,
                   color = "black") +
  scale_color_viridis_c(name = "Library size", labels = scales::label_comma()) +
  labs(x = "Observed richness", y = "Richness (breakaway)") +
  theme_classic(base_size = 14)
```

```{r freq_spectrum_asv, fig.width=7, fig.height=4}
# For each read-count class 0–10, count how many genera hit that count per sample
freq_spectrum <- map_dfr(1:10, function(k) {
  counts_per_sample <- colSums(otutab_asv == k)
  tibble(
    count     = k,
    mean_taxa = mean(counts_per_sample),
    sd_taxa   = sd(counts_per_sample)
  )
})

ggplot(freq_spectrum, aes(x = factor(count), y = mean_taxa)) +
  geom_errorbar(aes(ymin = pmax(0, mean_taxa - sd_taxa),
                    ymax = mean_taxa + sd_taxa),
                width = 0.25, linewidth = 0.5, color = "grey40") +
  geom_point(size = 3) +
  scale_x_discrete(labels = 1:10) +
  scale_y_continuous(
    trans        = "log1p",
    breaks       = c(0, 1, 2, 5, 10, 50, 100, 500, 1000, 2000),
    minor_breaks = NULL
  ) +
  labs(x = "Read count per sample",
       y = "Mean number of ASVs per sample") +
  theme_bw(base_size = 13) +
  theme(plot.caption = element_text(hjust = 0, size = 10))
```

## Compute richness (on genus level)

### Common richness

Computing richness the common way (not model-based) using the `estimate_richness` function from `phyloseq` package.

```{r}
alpha_div_common <- estimate_richness(phylo_genus)
head(alpha_div_common)
```

### Breakaway richness

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_genus)
```

```{r}
breakaway_rich <- 
  data.frame(SampleID = names(frequencytablelist),
             richness = unlist(lapply(frequencytablelist, 
                                      function(x) breakaway(x)$estimate))) 

alpha_diversity_rich <- breakaway_rich
head(alpha_diversity_rich)
```

### Comparison: classical vs. breakaway 

```{r richness_libsize_data}
lib_size <- data.frame(
  SampleID     = names(sample_sums(phylo_genus)),
  library_size = as.numeric(sample_sums(phylo_genus))
)

obs_rich <- data.frame(
  SampleID = rownames(alpha_div_common),
  observed  = alpha_div_common$Observed
)

richness_comp <- lib_size %>%
  inner_join(obs_rich, by = "SampleID") %>%
  inner_join(rename(breakaway_rich, breakaway = richness), by = "SampleID")

head(richness_comp)
```

```{r seq_depth}
richness_comp %>%
  ggplot(aes(x = library_size)) +
  geom_histogram(bins = 40, fill = "gray", colour = "white") +
  geom_vline(xintercept = median(sample_sums(phylo_asv)), linetype = "dashed", 
             colour = "black", linewidth = 0.9) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x     = "Read depth (counts)",
    y     = "Number of samples",
    title = "Per-sample sequencing depth",
    subtitle = paste0("Median: ", scales::comma(median(sample_sums(phylo_asv))),
                      "  |  Range: ", scales::comma(min(sample_sums(phylo_asv))),
                      " – ", scales::comma(max(sample_sums(phylo_asv))))
  ) +
  theme(
    axis.title        = element_text(size = 13),
    axis.text         = element_text(size = 11),
    plot.title        = element_text(size = 14)
  ) +
  theme_bw()
```

**Spearman correlation between library size and each richness metric:**

```{r richness_libsize_cor}
cor_obs <- cor.test(richness_comp$library_size, richness_comp$observed,
                    method = "spearman", exact = FALSE)
cor_bw  <- cor.test(richness_comp$library_size, richness_comp$breakaway,
                    method = "spearman", exact = FALSE)

tibble(
  Metric   = c("Observed richness", "Richness (breakaway)"),
  rho      = round(c(cor_obs$estimate, cor_bw$estimate), 3),
  p_value  = signif(c(cor_obs$p.value, cor_bw$p.value), 3)
)
```

```{r richness_vs_libsize, fig.width=10, fig.height=5}
richness_comp %>%
  pivot_longer(c(observed, breakaway),
               names_to  = "method",
               values_to = "richness") %>%
  mutate(method = factor(method,
                         levels = c("observed", "breakaway"),
                         labels = c("Observed richness", "Richness (breakaway)"))) %>%
  ggplot(aes(x = library_size, y = richness)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  ggpubr::stat_cor(method = "spearman",
                   label.x.npc = "left", label.y.npc = "top", size = 4.5) +
  facet_wrap(~method, scales = "free_y") +
  labs(x = "Library size (reads)", y = "Richness") +
  theme_bw(base_size = 14)
```

```{r observed_vs_breakaway, fig.width=5, fig.height=4}
plot_obs_est_richness <- ggplot(richness_comp, aes(x = observed, y = breakaway, color = library_size)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60",
              linewidth = 0.8) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_viridis_c(name = "Library size", labels = scales::label_comma()) +
  labs(x = "Observed richness (classical)", y = "Estimated richness (breakaway)") +
  theme_bw(base_size = 14)
plot_obs_est_richness
```

Breakaway estimates closely track observed richness. The frequency spectrum below
explains why: unseen-species estimators such as breakaway extrapolate unobserved
diversity from the ratio of rare taxa (singletons, doubletons, …). At genus level,
read counts are aggregated across ASVs, so genera with very few reads are rare.

```{r freq_spectrum_genus, fig.width=7, fig.height=4}
# For each read-count class 0–10, count how many genera hit that count per sample
freq_spectrum <- map_dfr(0:10, function(k) {
  counts_per_sample <- colSums(otutab_genus == k)
  tibble(
    count     = k,
    mean_taxa = mean(counts_per_sample),
    sd_taxa   = sd(counts_per_sample)
  )
})

x_labels <- c("0\n(absent)", "1\n(singleton)", "2\n(doubleton)", "3\n(tripleton)",
               as.character(4:10))

ggplot(freq_spectrum, aes(x = factor(count), y = mean_taxa)) +
  geom_errorbar(aes(ymin = pmax(0, mean_taxa - sd_taxa),
                    ymax = mean_taxa + sd_taxa),
                width = 0.25, linewidth = 0.5, color = "grey40") +
  geom_point(size = 3) +
  scale_x_discrete(labels = 0:10) +
  scale_y_continuous(
    trans        = "log1p",
    breaks       = c(0, 1, 2, 5, 10, 50, 100, 200),
    minor_breaks = NULL
  ) +
  labs(x = "Read count per sample",
       y = "Mean number of genera per sample") +
  theme_bw(base_size = 13) +
  theme(plot.caption = element_text(hjust = 0, size = 10))
```

Legend: Dots = mean across 592 samples; bars = \u00b11 SD.

~211 genera are absent in the average sample (count = 0, top-left point). No sample
contains any singleton genus (count = 1, pinned at zero): ASV-level rareness
disappears when reads are aggregated to genus. The f~2~–f~10~ band contains only
~0.6–0.9 genera on average per sample — an extremely sparse frequency tail.
Without singleton signal, estimators that rely on f~1~/f~2~ (Chao1, breakaway)
have no basis to project unseen genera, which is why breakaway estimates
closely track observed richness throughout this analysis.

## DivNet diversity measures

Find base taxon (with minimum number of non-zero counts)

```{r}
nonzero_elem <- apply(otutab_genus, 1, function(x) sum(x == 0))
base_taxon <- which.min(nonzero_elem)
nonzero_elem[base_taxon]
```

```{r}
if (!file.exists(path_target("divnet_genus.rds"))) {
  divnet_start <- proc.time()
  divnet_genus <- phylo_genus %>%
    DivNet::divnet(ncores = 1, 
                   base = names(base_taxon),
                   tuning = "fast")
  divnet_runtime <- proc.time() - divnet_start
  message("divnet runtime: ", round(divnet_runtime["elapsed"], 1), " seconds")
  
  saveRDS(divnet_genus, file = path_target("divnet_genus.rds"))
  saveRDS(divnet_runtime, file = path_target("divnet_genus_runtime.rds"))
} else {
  divnet_genus <- readRDS(path_target("divnet_genus.rds"))
  divnet_runtime <- readRDS(path_target("divnet_genus_runtime.rds"))
}

print("Runtime:")
divnet_runtime
```

```{r}
# Use the no-formula genus-level model for per-sample estimates (used in strip charts)
alpha_diversity_div <- 
  data.frame(SampleID = names(divnet_genus$shannon),
             shannon = unlist(lapply(divnet_genus$shannon, function(x) x$estimate)),
             simpson = unlist(lapply(divnet_genus$simpson, function(x) x$estimate)))

alpha_diversity_div <- alpha_diversity_div %>%
  mutate(inv_simpson = 1 / simpson,
         gini = 1 - simpson)
```

```{r}
alpha_diversity <- inner_join(alpha_diversity_rich, 
                              alpha_diversity_div, 
                              by = "SampleID")
```


```{r}
head(alpha_diversity)
```


## Permutation tests

Results are cached separately per variable (one RDS file each). Delete a variable's
file and rerun this chunk to recompute only that variable. Each test uses $10^6$
permutations. Results are reused by the plotting functions below and summarised in
the comparison table.

```{r perm_tests, warning=FALSE}
cores <- 10

future::plan(
  future::multisession,
  workers = cores
)

progressr::handlers("txtprogressbar")
options(progressr.enable = TRUE)

# Helper: build metadata_alpha for a given variable
prep_meta <- function(col, extra_filter = NULL) {
  d <- sampdata[, c("SampleID", col)]
  colnames(d) <- c("SampleID", "feature")
  rownames(d) <- NULL
  if (col == "Sex") {
    d$feature <- fct_recode(as.factor(d$feature), female = "f", male = "m")
  } else if (col == "cesarean") {
    d$feature <- fct_recode(as.factor(d$feature), no = "0", yes = "1")
  } else {
    d$feature <- as.factor(d$feature)
  }
  d <- d %>% tidyr::drop_na(feature)
  if (!is.null(extra_filter)) d <- d %>% filter(feature != extra_filter)
  d$feature <- droplevels(d$feature)
  inner_join(d, alpha_diversity, by = "SampleID")
}

vars_def <- list(
  Country        = list(col = "Country",         filter = NULL),
  Sex            = list(col = "Sex",              filter = NULL),
  Cesarean       = list(col = "cesarean",         filter = NULL),
  Breastfeeding  = list(col = "breast_dur_cat1",  filter = NULL),
  Exclusive_BF   = list(col = "breast_excl_cat1", filter = "1"),
  Prenatal_smoke = list(col = "pregsmoke",        filter = NULL),
  Siblings       = list(col = "sibs_numb_cat",    filter = NULL)
)

measures_def <- c("richness", "shannon", "gini")

perm_results <- list()

progressr::with_progress({
  for (vname in names(vars_def)) {
    rds_path <- path_target(paste0("perm_results_", vname, ".rds"))
    
    if (!file.exists(rds_path)) {
      message("Running permutation tests for: ", vname)
      meta    <- prep_meta(vars_def[[vname]]$col, vars_def[[vname]]$filter)
      nfeat_v <- nlevels(meta$feature)
      var_res <- list()
      set.seed(123456)
      for (m in measures_def) {
        key <- paste0(vname, "_", m)
        x   <- meta[[m]]
        g   <- meta$feature
        res <- perm_kruskal(x, g, n_perm = 1e6)
        res$p_pairwise <- NULL
        if (nfeat_v > 2 && res$p_perm <= 0.05) {
          res$p_pairwise <- perm_pairwise(x, g, n_perm = 1e6)  # returns list(p_matrix, p_matrix_approx)
        }
        var_res[[key]] <- res
      }
      saveRDS(var_res, rds_path)
    } else {
      var_res <- readRDS(rds_path)
    }
    
    perm_results <- c(perm_results, var_res)
  }
})
```

### Comparison of asymptotic vs. permutation p-values

```{r perm_table}
var_labels <- c(
  Country        = "Country",
  Sex            = "Sex",
  Cesarean       = "C-section",
  Breastfeeding  = "Breastfeeding duration",
  Exclusive_BF   = "Exclusive breastfeeding",
  Prenatal_smoke = "Prenatal smoking",
  Siblings       = "Number of siblings"
)
measure_labels <- c(
  richness_genus = "Richness (breakaway)",
  shannon        = "Shannon (DivNet)",
  gini           = "Gini-Simpson (DivNet)"
)

fmt_p <- function(p) formatC(p, format = "e", digits = 2)

perm_table <- purrr::imap_dfr(perm_results, function(res, key) {
  measure_part <- stringr::str_extract(key, "richness_genus|shannon|gini")
  var_part     <- sub(paste0("_", measure_part, "$"), "", key)
  tibble::tibble(
    Variable  = var_labels[var_part],
    Measure   = measure_labels[measure_part],
    H         = round(res$H, 2),
    p_asymp   = fmt_p(res$p_asymp),
    p_perm    = fmt_p(res$p_perm),
    p_approx  = fmt_p(res$p_approx)
  )
})

knitr::kable(perm_table,
             col.names = c("Variable", "Measure", "H statistic",
                           "p (KW asymptotic)", "p (permutation)", "p (GPD-refined)"),
             align = "llrrrr")
```

## Plots 

### Study center

```{r}
feature_name <- "Country"
feat_key     <- "Country"
```



```{r}
metadata <- sampdata[, c("SampleID", "Country")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>%
  mutate(feature = factor(feature)) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

```{r Country_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.5, dotsize = 1.2, 
              ylim = c(0, 90), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r Country_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r Country_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r Country_2m, fig.width=14, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```


### Sex

```{r}
feature_name <- "Sex"
feat_key     <- "Sex"
```



```{r}
metadata <- sampdata[, c("SampleID", "Sex")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, female = "f", male = "m")) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

```{r sex_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```


```{r sex_shannon_2m, fig.width=5, fig.height=5}
# Shannon 
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r sex_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r sex_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

### C-section

```{r}
feature_name <- "C-section"
feat_key     <- "Cesarean"
```



```{r}
metadata <- sampdata[, c("SampleID", "cesarean")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  drop_na(feature) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

```{r cesarean_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r cesarean_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r cesarean_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r cesarean_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg, ncol = 3, nrow = 1)
```


### Breast feeding

```{r}
feature_name <- "Breast feeding"
feat_key     <- "Breastfeeding"
```



```{r}
metadata <- sampdata[, c("SampleID", "breast_dur_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```


```{r breastfeed_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "\u2265 2"),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r breastfeed_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "\u2265 2"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r breastfeed_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "\u2265 2"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```


```{r breastfeed_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

### Exclusive breast feeding

```{r}
feature_name <- "Exclusive breast feeding"
feat_key     <- "Exclusive_BF"
```



```{r}
metadata <- sampdata[, c("SampleID", "breast_excl_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature) %>%
  filter(feature != "1")

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```


```{r breast_excl_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(droplevels(metadata_alpha)$feature),
              labels = c("0", "\u2265 2"),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r breast_excl_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(droplevels(metadata_alpha)$feature),
             labels = c("0", "\u2265 2"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r breast_excl_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(droplevels(metadata_alpha)$feature),
          labels = c("0", "\u2265 2"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r breast_excl_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

### Smoke during pregnancy

```{r}
feature_name <- "Smoke during Pregnancy"
feat_key     <- "Prenatal_smoke"
```



```{r}
metadata <- sampdata[, c("SampleID", "pregsmoke")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature) %>%
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

```{r pregsmoke_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r pregsmoke_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r pregsmoke_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r pregsmoke_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

### Number of siblings

```{r}
feature_name <- "Number of siblings"
feat_key     <- "Siblings"
```



```{r}
metadata <- sampdata[, c("SampleID", "sibs_numb_cat")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

```{r sibsnumb_richness_2m, fig.width=5, fig.height=5}
# Richness
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "&gt;1"),
              ptest = perm_results[[paste0(feat_key, "_richness")]])
```



```{r sibsnumb_shannon_2m, fig.width=5, fig.height=5}
# Shannon
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "&gt;1"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```



```{r sibsnumb_simpson_2m, fig.width=5, fig.height=5}
# Gini-Simpson index
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "&gt;1"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```



```{r sibsnumb_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```


## Model-based analysis

The permutation tests above treat each sample's diversity estimate as an exact
observation. A model-based alternative, introduced by Willis & Martin (2020), uses
`breakaway::betta()` — a mixed-effects regression model that explicitly accounts
for the per-sample estimation variance ($\delta_i^2$) from breakaway/DivNet while
simultaneously estimating residual biological heterogeneity ($\sigma^2$). We
present two versions:

1. **Intercept-only**: estimates the grand mean of each diversity metric with
   proper variance weighting, using the full sample set. This is the baseline
   approach from Willis & Martin (2020) and can serve as a foundation for
   permutation tests that respect estimation uncertainty.
2. **With covariates**: adds cohort-level predictors to jointly estimate
   associations with diversity, restricted to samples with complete covariate data.

---

### Intercept-only model

The intercept-only `betta` model fits

$$Y_i = \mu + \varepsilon_i + \delta_i,$$

where $Y_i$ is the diversity estimate for sample $i$, $\mu$ is the global mean,
$\varepsilon_i \sim \mathcal{N}(0,\,\sigma^2)$ captures residual biological
variation (estimated from the data), and $\delta_i$ is the estimation error with
known variance supplied by breakaway/DivNet.

The estimated $\sigma^2$ (labelled `global` in the betta output) captures
between-sample variation after accounting for estimation noise. A large $\sigma^2$
relative to the typical $\delta_i^2$ means biological heterogeneity dominates —
which is a prerequisite for permutation tests to have power beyond a naïve
Kruskal–Wallis test on raw estimates. The BLUP-like fitted values from this model
could replace raw estimates in permutation tests, downweighting samples with high
estimation uncertainty and providing a fairer test statistic when $\delta_i^2$
varies substantially across samples.

#### Richness

```{r betta_int_richness}
# Cache full-sample breakaway runs (estimate + SE needed for betta)
if (!file.exists(path_target("breakaway_rich_full.rds"))) {
  rich_full <- lapply(frequencytablelist, breakaway::breakaway)
  saveRDS(rich_full, file = path_target("breakaway_rich_full.rds"))
} else {
  rich_full <- readRDS(path_target("breakaway_rich_full.rds"))
}

rich_est_int <- sapply(rich_full, function(x) x$estimate)
rich_ses_int <- pmax(sapply(rich_full, function(x) x$error), 0.5)

X_int_rich <- matrix(1, nrow = length(rich_est_int), ncol = 1,
                     dimnames = list(NULL, "(Intercept)"))

betta_rich_int <- breakaway::betta(rich_est_int, rich_ses_int, X_int_rich,
                                   p.digits = 30)
betta_rich_int$table
```

```{r}
# Predicted values: intercept-only model predicts the grand mean for every sample
rich_predicted_int <- as.numeric(X_int_rich %*% betta_rich_int$table[, 1])

richness_plot_data_int <- data.frame(
  SampleID           = names(rich_est_int),
  richness_predicted = rich_predicted_int
) |>
  left_join(select(richness_comp, SampleID, library_size), by = "SampleID")
```

```{r observed_vs_predicted_richness_int_libsize, fig.width=7, fig.height=4}
plot_obs_est_richness_int <-
  ggplot(richness_plot_data_int |> inner_join(obs_rich, by = "SampleID"),
         aes(x = observed, y = richness_predicted, color = library_size)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_viridis_c(name = "Library size", labels = scales::label_comma()) +
  labs(x = "Observed richness (classical)",
       y = "Predicted richness (intercept-only)") +
  theme_bw(base_size = 14)
plot_obs_est_richness_int
```

Since the intercept-only model has no covariates, all samples receive the same
predicted value (the grand mean), so the points form a horizontal band. The
vertical spread reflects true biological variation plus estimation noise that
the model cannot explain without predictors.

#### Shannon

```{r betta_int_shannon}
betta_est_int_s <- divnet_genus$shannon %>% summary() %>% dplyr::pull(estimate)
betta_ses_int_s <- sqrt(divnet_genus$`shannon-variance`)

X_int_s <- matrix(1, nrow = length(betta_est_int_s), ncol = 1,
                  dimnames = list(NULL, "(Intercept)"))

betta_shannon_int <- breakaway::betta(betta_est_int_s, betta_ses_int_s, X_int_s,
                                      p.digits = 30)
betta_shannon_int$table
```

#### Gini-Simpson

```{r betta_int_gini}
gini_est_int <- 1 - (divnet_genus$simpson %>% summary() %>% dplyr::pull(estimate))
gini_ses_int <- sqrt(divnet_genus$`simpson-variance`)

X_int_g <- matrix(1, nrow = length(gini_est_int), ncol = 1,
                  dimnames = list(NULL, "(Intercept)"))

betta_gini_int <- breakaway::betta(gini_est_int, gini_ses_int, X_int_g,
                                   p.digits = 30)
betta_gini_int$table
```


---

### With covariates

```{r}
# Drop samples with NAs in any formula covariate
# Also exclude samples with 1-month exclusive breastfeeding (sparse category)
cov_vars <- c("Country", "Sex", "cesarean", "breast_dur_cat1",
              "breast_excl_cat1", "pregsmoke", "sibs_numb_cat")
sdata <- data.frame(sample_data(phylo_genus))
complete_samples <- rownames(sdata)[complete.cases(sdata[, cov_vars])]
complete_samples <- complete_samples[sdata[complete_samples, "breast_excl_cat1"] != "1"]
phylo_genus_cov <- prune_samples(complete_samples, phylo_genus)
```

The strip charts above show unadjusted, per-sample DivNet estimates. Here we use
the covariate model (`divnet_genus_cov`, family level) together with
`breakaway::betta` to jointly estimate associations between all covariates and
Shannon diversity, properly accounting for the uncertainty in the DivNet estimates.
Note that this model uses only the `r nsamples(phylo_genus_cov)` samples with
complete covariate data.

```{r}
if (!file.exists(path_target("divnet_genus_cov.rds"))) {
  set.seed(123456)
  divnet_start <- proc.time()
  divnet_genus_cov <- phylo_genus_cov %>%
    DivNet::divnet(formula = ~ Country + Sex + cesarean + breast_dur_cat1 + 
                     breast_excl_cat1 + pregsmoke + sibs_numb_cat,
                   ncores = 1, 
                   tuning = "fast")
  
  divnet_runtime <- proc.time() - divnet_start
  
  saveRDS(divnet_genus_cov, file = path_target("divnet_genus_cov.rds"))
  saveRDS(divnet_runtime, file = path_target("divnet_genus_cov_runtime.rds"))
} else {
  divnet_genus_cov <- readRDS(path_target("divnet_genus_cov.rds"))
  divnet_runtime <- readRDS(path_target("divnet_genus_cov_runtime.rds"))
}

print("Runtime:")
divnet_runtime
```

**Computed measures:**

```{r}
names(divnet_genus_cov)
```

#### Richness

```{r}
# Rerun breakaway only on the model samples (n = 486) to obtain aligned SEs.
# One sample has a near-zero SE (~5e-95); this is floored at 0.5
# (conservative relative to the estimate scale of ~6–54) to avoid
# a singular covariance matrix in betta().
model_samples  <- rownames(divnet_genus_cov$X)
flist_cov      <- frequencytablelist[model_samples]
rich_cov       <- lapply(flist_cov, breakaway::breakaway)
rich_estimates <- sapply(rich_cov, function(x) x$estimate)
rich_ses       <- pmax(sapply(rich_cov, function(x) x$error), 0.5)

betta_rich <- breakaway::betta(rich_estimates, rich_ses, divnet_genus_cov$X,
                                  p.digits = 30)
betta_rich$table
```

Extract predicted richness values (conditional on covariates):

```{r}
# Predicted values from the model
rich_predicted <- divnet_genus_cov$X %*% betta_rich$table[, 1]

richness_with_model <- data.frame(
  SampleID = model_samples,
  richness_observed = rich_estimates,
  richness_predicted = as.numeric(rich_predicted),
  richness_se = rich_ses
)

head(richness_with_model)

sdata <- filter(sdata, SampleID %in% richness_with_model$SampleID)

richness_plot_data <- richness_with_model %>%
  left_join(select(sdata, SampleID, breast_dur_cat1), by = "SampleID") %>%
  left_join(select(richness_comp, SampleID, library_size), by = "SampleID")
```


```{r observed_vs_predicted_richness_libsize, fig.width=7, fig.height=4}
plot_obs_est_richness_cov <- 
  ggplot(richness_plot_data |> inner_join(obs_rich, by = "SampleID"), 
         aes(x = observed, y = richness_predicted,
             color = library_size)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60",
              linewidth = 0.6) +
  #coord_cartesian(xlim = c(0, 55), ylim = c(15, 45)) +
  scale_color_viridis_c(name = "Library size", labels = scales::label_comma()) +
  labs(x = "Observed richness (classical)",
       y = "Predicted richness (model-based)") +
  theme_bw(base_size = 14)
plot_obs_est_richness_cov
```

```{r observed_vs_predicted_richness_libsize_combined, fig.width=8, fig.height=4}
(plot_obs_est_richness | plot_obs_est_richness_cov) +
  plot_layout(guides = 'collect')
```


Observed vs. predicted richness (with covariates):

```{r observed_vs_predicted_richness, fig.width=7, fig.height=4}
ggplot(richness_plot_data, aes(x = richness_observed, y = richness_predicted, 
                                color = breast_dur_cat1)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60", 
              linewidth = 0.6) +
  scale_color_manual(name = "breast_dur_cat1", 
                     values = cols[1:length(unique(richness_plot_data$breast_dur_cat1))]) +
  labs(x = "Observed richness (breakaway)", 
       y = "Predicted richness (model-based)") +
  theme_classic(base_size = 14)
```

#### Shannon

```{r}
betta_estimates <- divnet_genus_cov$shannon %>% summary() %>% dplyr::pull(estimate)
betta_ses       <- sqrt(divnet_genus_cov$`shannon-variance`)

betta_shannon <- breakaway::betta(betta_estimates, betta_ses, divnet_genus_cov$X,
                                  p.digits = 30)
betta_shannon$table
```

Global test of whether any covariate is associated with Shannon diversity:

```{r}
betta_shannon$global
```

`betta_shannon$global` is a likelihood ratio test of the null hypothesis that all covariates
jointly have no association with Shannon diversity (i.e., all βs except the intercept equal
zero). The two returned values are the chi-squared test statistic and its p-value,
respectively. A significant result indicates that the model as a whole fits better than an
intercept-only model, meaning at least one covariate is meaningfully associated with Shannon
diversity. It does not identify which covariates are responsible — that is addressed by the
individual p-values in `betta_shannon$table` above.

#### Gini-Simpson

```{r}
gini_estimates <- 1 - (divnet_genus_cov$simpson %>% summary() %>% dplyr::pull(estimate))
gini_ses       <- sqrt(divnet_genus_cov$`simpson-variance`)

betta_gini <- breakaway::betta(gini_estimates, gini_ses, divnet_genus_cov$X,
                                  p.digits = 30)
betta_gini$table
```



#### Combined coefficient plot

```{r betta_forest_combined, fig.width=10, fig.height=6}
term_labels <- c(
  "CountryGermany"      = "Country: Germany",
  "CountrySwitzerland"  = "Country: Switzerland",
  "Sexm"                = "Sex: male",
  "cesarean"            = "Caesarean delivery",
  "breast_dur_cat11"    = "Breastfeeding duration: 1 month",
  "breast_dur_cat1>=2"  = "Breastfeeding duration: \u22652 months",
  "breast_excl_cat11"   = "Exclusive BF: 1 month",
  "breast_excl_cat1>=2" = "Exclusive BF: \u22652 months",
  "pregsmoke"           = "Prenatal smoking",
  "sibs_numb_cat1"      = "Siblings: 1",
  "sibs_numb_cat>1"     = "Siblings: >1"
)

tidy_betta <- function(betta_res, measure_label) {
  as.data.frame(betta_res$table) %>%
    tibble::rownames_to_column("term") %>%
    filter(term != "(Intercept)") %>%
    mutate(
      lower   = Estimates - 1.96 * `Standard Errors`,
      upper   = Estimates + 1.96 * `Standard Errors`,
      sig     = `p-values` < 0.05,
      measure = measure_label,
      term    = dplyr::recode(term, !!!term_labels)
    )
}

bind_rows(
  tidy_betta(betta_shannon, "Shannon (DivNet)"),
  tidy_betta(betta_gini,   "Gini-Simpson (DivNet)"),
  tidy_betta(betta_rich,   "Richness (breakaway)")
) %>%
  mutate(
    measure = factor(measure, levels = c("Richness (breakaway)",
                                         "Shannon (DivNet)",
                                         "Gini-Simpson (DivNet)")),
    term = factor(term, levels = rev(unique(term)))
  ) %>%
  ggplot(aes(x = Estimates, y = term, color = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = "#1B7837"),
                     labels = c("FALSE" = "no", "TRUE"  = "yes")) +
  facet_wrap(~measure, scales = "free_x", nrow = 1) +
  labs(x = "Coefficient estimate (95% CI)", y = NULL, color = "p < 0.05") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

##### Investigating the C-section effect: confounding by breastfeeding duration

The positive association of caesarean delivery with Shannon diversity is counterintuitive
biologically. A likely explanation is confounding by breastfeeding: C-section delivery is
associated with shorter breastfeeding duration, and shorter breastfeeding is associated with
higher Shannon diversity (the dominant effect in the model). The following two plots
investigate this.

```{r cesarean_breastfeed_confounding, fig.width=9, fig.height=4}
plot_dat <- alpha_diversity_div %>%
  left_join(
    sdata %>%
      select(SampleID, cesarean, breast_dur_cat1),
    by = "SampleID"
  ) %>%
  filter(!is.na(cesarean), !is.na(breast_dur_cat1)) %>%
  mutate(
    cesarean = factor(cesarean, levels = c(1, 0), labels = c("C-section", "Vaginal")),
    breast_dur_cat1 = factor(breast_dur_cat1,
                             levels = c("0", "1", ">=2"),
                             labels = c("< 1 month", "1 month", "≥ 2 months"))
  )

# Panel 1: breastfeeding duration distribution by delivery mode
p1 <- plot_dat %>%
  count(cesarean, breast_dur_cat1) %>%
  group_by(cesarean) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = cesarean, y = pct, fill = breast_dur_cat1)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Delivery mode", y = "Proportion", fill = "Breastfeeding\nduration") +
  theme_bw()

# Panel 2: Shannon diversity by delivery mode, stratified by breastfeeding duration
p2 <- ggplot(plot_dat, aes(x = cesarean, y = shannon)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_wrap(~breast_dur_cat1) +
  labs(x = "Delivery mode", y = "Shannon index (DivNet)") +
  theme_bw()

cowplot::plot_grid(p1, p2, rel_widths = c(1, 1.8))
```

The left panel shows that C-section infants are less likely to be breastfed for ≥ 2 months
compared to vaginally born infants. Since prolonged breastfeeding is the strongest single
predictor of *lower* Shannon diversity in this cohort, C-section infants appear more diverse
in the unadjusted comparison. The right panel confirms this: within each breastfeeding
duration stratum, the delivery mode difference is substantially attenuated or absent.

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list_files_target}
projthis::proj_dir_info(path_target())
```
