---
title: "Alpha diversity analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "02_alpha_diversity" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "41b33e9b-a351-49eb-bdc8-cdeddd77796c")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params, "/alpha_"),
  echo = FALSE)
```

## Load packages

```{r packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(ggtext)
library(ggpubr)

# Diversity packages
library(breakaway)
library(DivNet)
library(progress)
library(permApprox)
library(permApprox)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data

```{r load-data}
phylo_asv   <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_asv.rds"))
phylo_genus <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_genus.rds"))
sampdata <- sample_data(phylo_asv) %>% as("data.frame")
otutab_asv   <- as(otu_table(phylo_asv), "matrix")
otutab_genus <- as(otu_table(phylo_genus), "matrix")
```

### ASV level

```{r}
phylo_asv
```

### Genus level

```{r}
phylo_genus
```


## Colors

```{r}
cols = c("#009E73", "#56B4E9", "#E69F00", "#D55E00", 
         "#CC79A7", "#F0E442", "#0072B2", "gray")
```

## Functions

### Function for significance plot

```{r}
modify_div_plot <- function(strip_chart, xlab = "", breaks = NULL, labels = NULL,
                            binwidth = 0.5, dotsize = 1, ylim = c(0, 75),
                            ylab = "", ybreaks = NULL,
                            textsize = 18, title = "") {
  strip_chart <- strip_chart +
    stat_summary(fun.data = median_hilow, fun.args=0.50, show.legend=FALSE,
                 geom="crossbar", alpha=0.25, width=0.7) +
    geom_dotplot(binaxis = "y", binwidth = binwidth, 
                 stackdir = "center", width = 0.5, alpha = 0.7,
                 show.legend = FALSE, dotsize = dotsize) +
    labs(x=NULL,
         y=ylab) +
    scale_fill_manual(values = cols) +
    scale_x_discrete(name = xlab,
                     breaks = breaks, 
                     labels = labels) +
    theme_classic() +
    theme(axis.text.x = element_markdown(),
          axis.text.y = element_markdown(),
          legend.text = element_markdown(),
          text = element_text(size = textsize)) +
    ggtitle(title)
  
  if (!is.null(ybreaks)) {
    strip_chart <- strip_chart + 
      scale_y_continuous(breaks=ybreaks, limits = ylim)
  } else {
    strip_chart <- strip_chart + 
      scale_y_continuous(limits = ylim)
  }
  strip_chart
}

```

### Function for significance line

```{r}
add_sig_line <- function(strip_chart, pval, diversity,
                         space_line = 0.05, space_lab = 0.05, nfeat = nfeat, 
                         linewidth = 0.8, size = 5) {
  
  if (nfeat == 2) {
    pmat_x <- c(1)
    pmat_y <- c(1)
    line_left <- c(1)
    line_right <- c(2)
    lab_x <- c(1.5)
  } else if (nfeat == 3) {
    pmat_x <- c(1, 2, 2)
    pmat_y <- c(1, 2, 1)
    line_left <- c(1, 2, 1)
    line_right <- c(2, 3, 3)
    lab_x <- c(1.5, 2.5, 2)
  } else {
    pmat_x <- c(1, 2, 3, 2, 3, 3)
    pmat_y <- c(1, 2, 3, 1, 2, 1)
    line_left <- c(1, 2, 3, 1, 2, 1)
    line_right <- c(2, 3, 4, 3, 4, 4)
    lab_x <- c(1.5, 2.5, 3.5, 2, 3, 2.5)
  }
  
  pos_line <- max(diversity) - space_line + 0.1 * space_line
  pos_lab <- pos_line + space_lab
  
  for (i in seq_along(pmat_x)) {
    if (is.null(dim(pval))) {
      label <- pval
    } else {
      label <- pval[pmat_x[i], pmat_y[i]]
    }
    
    if (label <= 0.05) {
      pos_line <- pos_line + space_line
      pos_lab <- pos_lab + space_line
      
      strip_chart <- strip_chart +
        geom_line(data = tibble(x = c(line_left[i], line_right[i]), 
                                y = c(pos_line, pos_line)), 
                  aes(x=x, y=y),
                  inherit.aes = FALSE, 
                  linewidth = linewidth) +
        #geom_text(x = lab_x[i], y = pos_lab, label = signif(label, digits = 3))
        annotate("text", lab_x[i], y = pos_lab, 
                 label = signif(label, digits = 3), size = size)
    } 
  }
  strip_chart
}
```

### Permutation test functions

```{r}
# Kruskal-Wallis test with permutation p-value and GPD-refined p-value.
# Uses the KW H statistic on permuted group labels.
perm_kruskal <- function(x, g, n_perm = 9999, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  g      <- droplevels(as.factor(g))
  obs    <- kruskal.test(x ~ g)
  obs_H  <- obs$statistic
  
  pb <- progress::progress_bar$new(
    format = "  Permutations [:bar] :current/:total (:percent)",
    total = n_perm, clear = FALSE, width = 60
  )
  perm_H <- sapply(1:n_perm, function(i) {
    pb$tick()
    kruskal.test(x ~ sample(g))$statistic
  })
  
  p_perm <- (sum(perm_H >= obs_H) + 1) / (n_perm + 1)
  
  pa <- permApprox::perm_approx(
    obs_stats     = unname(obs_H),
    perm_stats    = matrix(perm_H, ncol = 1),
    alternative   = "greater",
    adjust_method = "none",
    verbose       = FALSE,
    gpd_ctrl      = permApprox::make_gpd_ctrl(sample_size = min(table(g)))
  )
  
  list(H        = unname(obs_H),
       p_asymp  = obs$p.value,
       p_perm   = p_perm,
       p_approx = pa$p_unadjusted[1])
}

# Pairwise permutation Wilcoxon tests with BH correction.
# For each pair, permutes group labels and uses the W statistic (two-sided).
# Returns both empirical and GPD-refined BH-adjusted p-value matrices.
perm_pairwise <- function(x, g, n_perm = 9999, p_adjust = "BH", seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  g     <- droplevels(as.factor(g))
  pairs <- combn(levels(g), 2, simplify = FALSE)
  
  n_pairs <- length(pairs)
  pb <- progress::progress_bar$new(
    format = "  Pairwise tests [:bar] :current/:total (:percent)",
    total = n_pairs, clear = FALSE, width = 60
  )

  p_vals        <- numeric(n_pairs)
  p_vals_approx <- numeric(n_pairs)
  
  for (i in seq_along(pairs)) {
    pb$tick()
    pair   <- pairs[[i]]
    idx    <- g %in% pair
    xi     <- x[idx]
    gi     <- droplevels(g[idx])
    obs_W  <- wilcox.test(xi ~ gi, exact = FALSE)$statistic
    n1     <- sum(gi == levels(gi)[1])
    n2     <- sum(gi == levels(gi)[2])
    EW     <- n1 * n2 / 2
    perm_W <- replicate(n_perm, wilcox.test(xi ~ sample(gi), exact = FALSE)$statistic)
    
    p_vals[i] <- (sum(abs(perm_W - EW) >= abs(obs_W - EW)) + 1) / (n_perm + 1)
    
    pa <- permApprox::perm_approx(
      obs_stats     = obs_W,
      perm_stats    = matrix(perm_W, ncol = 1),
      alternative   = "two_sided",
      null_center   = EW,
      adjust_method = "none",
      verbose       = FALSE,
      gpd_ctrl      = permApprox::make_gpd_ctrl(sample_size = min(n1, n2))
    )
    p_vals_approx[i] <- pa$p_unadjusted[1]
  }

  # BH correction across pairs
  p_adj        <- p.adjust(p_vals,        method = p_adjust)
  p_adj_approx <- p.adjust(p_vals_approx, method = p_adjust)

  # Build upper-triangular-style matrices matching pairwise.wilcox.test() format
  lvls <- levels(g)
  make_mat <- function(vals) {
    mat <- matrix(NA,
                  nrow     = length(lvls) - 1,
                  ncol     = length(lvls) - 1,
                  dimnames = list(lvls[-1], lvls[-length(lvls)]))
    for (i in seq_along(pairs)) {
      r <- which(lvls == pairs[[i]][2]) - 1
      c <- which(lvls == pairs[[i]][1])
      if (r >= 1 && c <= ncol(mat)) mat[r, c] <- vals[i]
    }
    mat
  }
  
  list(p_matrix        = make_mat(p_adj),
       p_matrix_approx = make_mat(p_adj_approx))
}
```

### Plot function for richness

```{r}
plot_richness <- function(div_dataframe, feature_name, 
                          xlab = "", breaks = NULL, labels = NULL,
                          binwidth = 0.4, dotsize = 1.2, 
                          ylab = "Breakaway richness", 
                          ybreaks = NULL, title = "",
                          ylim = c(0, 75), textsize = 18,
                          space_line = 2, space_lab = 5,
                          ptest = NULL,
                          p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=richness_genus, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(richness_genus ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$richness_genus,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$richness_genus,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```

### Plot function for Shannon

```{r}
plot_shannon <- function(div_dataframe, feature_name, 
                         xlab = "", breaks = NULL, labels = NULL,
                         binwidth = 0.04, dotsize = 1, ylim = c(0, 4), 
                         space_line = 0.3, space_lab = 0.15,
                         ylab = "Shannon index (DivNet)", 
                         ybreaks = NULL, title = "", textsize = 18,
                         ptest = NULL,
                         p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=shannon, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(shannon ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$shannon,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$shannon,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```

### Plot function for Gini-Simpson

```{r}
plot_gini <- function(div_dataframe, feature_name, 
                      xlab = "", breaks = NULL, labels = NULL,
                      binwidth = 0.01, dotsize = 1, ylim = c(0, 1.1), 
                      space_line = 0.08, space_lab = 0.04,
                      ylab = "Gini-Simpson index (DivNet)", 
                      ybreaks = NULL, title = "", textsize = 18,
                      ptest = NULL,
                      p_type = c("p_approx", "p_perm")) {
  p_type <- match.arg(p_type)
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=gini, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  if (is.null(ptest)) {
    pval_kt <- pval <- kruskal.test(gini ~ feature,
                                    data = div_dataframe)$p.value
    if (pval_kt <= 0.05 && nfeat > 2) {
      pval <- pairwise.wilcox.test(div_dataframe$gini,
                                   g = div_dataframe$feature,
                                   p.adjust.method = "BH")$p.value
    }
  } else {
    pval_kt <- ptest[[p_type]]
    pval    <- if (nfeat > 2 && pval_kt <= 0.05) {
      if (p_type == "p_approx") ptest$p_pairwise$p_matrix_approx
      else                      ptest$p_pairwise$p_matrix
    } else pval_kt
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$gini,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```


## Compute alpha diversity

### Alpha diversity on ASV level

Computing alpha diversity the common way

```{r}
alpha_div_common <- estimate_richness(phylo_asv)
head(alpha_div_common)
```

### Breakaway richness on ASV level

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_asv)
```

```{r}
rich_asv <- unlist(lapply(frequencytablelist, function(x) breakaway(x)$estimate))

alpha_diversity_rich <- data.frame(SampleID = names(rich_asv),
                                   age = "2 months",
                                   richness_otu = rich_asv)
head(alpha_diversity_rich)
```

### Alpha diversity on genus level

Computing alpha diversity the common way

```{r}
alpha_div_common_genus <- estimate_richness(phylo_genus)
head(alpha_div_common_genus)
```

### Breakaway  richness on genus level

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_genus)
```

```{r}
rich_genus <- 
  data.frame(SampleID = names(frequencytablelist),
             richness_genus = unlist(lapply(frequencytablelist, 
                                            function(x) breakaway(x)$estimate))) 
```


```{r}
alpha_diversity_rich <- inner_join(alpha_diversity_rich, 
                                      rich_genus, by = "SampleID")
head(alpha_diversity_rich)
```

## DivNet diversity measures

Due to the long runtime only on genus level

Find base taxon (with minimum number of non-zero counts)

```{r}
nonzero_elem <- apply(otutab_genus, 1, function(x) sum(x == 0))
base_taxon <- which.min(nonzero_elem)
nonzero_elem[base_taxon]
```

```{r}
if (!file.exists(path_target("divnet_genus.rds"))) {
  divnet_start <- proc.time()
  divnet_genus <- phylo_genus %>%
    DivNet::divnet(ncores = 1, 
                   base = names(base_taxon),
                   tuning = "fast")
  divnet_runtime <- proc.time() - divnet_start
  message("divnet runtime: ", round(divnet_runtime["elapsed"], 1), " seconds")
  
  saveRDS(divnet_genus, file = path_target("divnet_genus.rds"))
  saveRDS(divnet_runtime, file = path_target("divnet_genus_runtime.rds"))
} else {
  divnet_genus <- readRDS(path_target("divnet_genus.rds"))
  divnet_runtime <- readRDS(path_target("divnet_genus_runtime.rds"))
}
```

```{r}
# Use the no-formula genus-level model for per-sample estimates (used in strip charts)
alpha_diversity_div <- 
  data.frame(SampleID = names(divnet_genus$shannon),
             shannon = unlist(lapply(divnet_genus$shannon, function(x) x$estimate)),
             simpson = unlist(lapply(divnet_genus$simpson, function(x) x$estimate)))

alpha_diversity_div <- alpha_diversity_div %>%
  mutate(inv_simpson = 1 / simpson,
         gini = 1 - simpson)
```

```{r}
alpha_diversity <- inner_join(alpha_diversity_rich, 
                              alpha_diversity_div, 
                              by = "SampleID")
```


```{r}
head(alpha_diversity)
```

## Permutation tests

Results are cached separately per variable (one RDS file each). Delete a variable's
file and rerun this chunk to recompute only that variable. Each test uses 9 999
permutations. Results are reused by the plotting functions below and summarised in
the comparison table.

```{r perm-tests}
# Helper: build metadata_alpha for a given variable
prep_meta <- function(col, extra_filter = NULL) {
  d <- sampdata[, c("SampleID", col)]
  colnames(d) <- c("SampleID", "feature")
  rownames(d) <- NULL
  if (col == "Sex") {
    d$feature <- fct_recode(as.factor(d$feature), female = "f", male = "m")
  } else if (col == "cesarean") {
    d$feature <- fct_recode(as.factor(d$feature), no = "0", yes = "1")
  } else {
    d$feature <- as.factor(d$feature)
  }
  d <- d |> tidyr::drop_na(feature)
  if (!is.null(extra_filter)) d <- d |> filter(feature != extra_filter)
  d$feature <- droplevels(d$feature)
  inner_join(d, alpha_diversity, by = "SampleID")
}

vars_def <- list(
  Country        = list(col = "Country",         filter = NULL),
  Sex            = list(col = "Sex",              filter = NULL),
  Cesarean       = list(col = "cesarean",         filter = NULL),
  Breastfeeding  = list(col = "breast_dur_cat1",  filter = NULL),
  Exclusive_BF   = list(col = "breast_excl_cat1", filter = "1"),
  Prenatal_smoke = list(col = "pregsmoke",        filter = NULL),
  Siblings       = list(col = "sibs_numb_cat",    filter = NULL)
)
measures_def <- c("richness_genus", "shannon", "gini")

perm_results <- list()

for (vname in names(vars_def)) {
  rds_path <- path_target(paste0("perm_results_", vname, ".rds"))

  if (!file.exists(rds_path)) {
    message("Running permutation tests for: ", vname)
    meta    <- prep_meta(vars_def[[vname]]$col, vars_def[[vname]]$filter)
    nfeat_v <- nlevels(meta$feature)
    var_res <- list()
    set.seed(123456)
    for (m in measures_def) {
      key <- paste0(vname, "_", m)
      x   <- meta[[m]]
      g   <- meta$feature
      res <- perm_kruskal(x, g, n_perm = 9999)
      res$p_pairwise <- NULL
      if (nfeat_v > 2 && res$p_perm <= 0.05) {
        res$p_pairwise <- perm_pairwise(x, g, n_perm = 9999)  # returns list(p_matrix, p_matrix_approx)
      }
      var_res[[key]] <- res
    }
    saveRDS(var_res, rds_path)
  } else {
    var_res <- readRDS(rds_path)
  }

  perm_results <- c(perm_results, var_res)
}
```

### Comparison of asymptotic vs. permutation p-values

```{r perm-table}
var_labels <- c(
  Country        = "Country",
  Sex            = "Sex",
  Cesarean       = "C-section",
  Breastfeeding  = "Breastfeeding duration",
  Exclusive_BF   = "Exclusive breastfeeding",
  Prenatal_smoke = "Prenatal smoking",
  Siblings       = "Number of siblings"
)
measure_labels <- c(
  richness_genus = "Breakaway richness",
  shannon        = "Shannon (DivNet)",
  gini           = "Gini-Simpson (DivNet)"
)

fmt_p <- function(p) formatC(p, format = "e", digits = 2)

perm_table <- purrr::imap_dfr(perm_results, function(res, key) {
  measure_part <- stringr::str_extract(key, "richness_genus|shannon|gini")
  var_part     <- sub(paste0("_", measure_part, "$"), "", key)
  tibble::tibble(
    Variable  = var_labels[var_part],
    Measure   = measure_labels[measure_part],
    H         = round(res$H, 2),
    p_asymp   = fmt_p(res$p_asymp),
    p_perm    = fmt_p(res$p_perm),
    p_approx  = fmt_p(res$p_approx)
  )
})

knitr::kable(perm_table,
             col.names = c("Variable", "Measure", "H statistic",
                           "p (KW asymptotic)", "p (permutation)", "p (GPD-refined)"),
             align = "llrrrr")
```

## Study center

```{r}
feature_name <- "Country"
feat_key     <- "Country"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "Country")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>%
  mutate(feature = factor(feature)) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

# metadata_alpha[which.max(metadata_alpha$richness_genus), "richness_genus"] <-
#   max(metadata_alpha$richness_genus) / 2

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r Country_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.5, dotsize = 1.2, 
              ylim = c(0, 90), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r Country_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r Country_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all

```{r Country_2m, fig.width=14, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```


## Sex

```{r}
feature_name <- "Sex"
feat_key     <- "Sex"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "Sex")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, female = "f", male = "m")) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r sex_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r sex_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r sex_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all

```{r sex_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## C-section

```{r}
feature_name <- "C-section"
feat_key     <- "Cesarean"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "cesarean")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  drop_na(feature) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r cesarean_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r cesarean_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r cesarean_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all

```{r cesarean_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg, ncol = 3, nrow = 1)
```


## Breast feeding

```{r}
feature_name <- "Breast feeding"
feat_key     <- "Breastfeeding"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "breast_dur_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r breastfeed_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "\u2265 2"),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r breastfeed_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "\u2265 2"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r breastfeed_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "\u2265 2"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all
```{r breastfeed_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Exclusive breast feeding

```{r}
feature_name <- "Exclusive breast feeding"
feat_key     <- "Exclusive_BF"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "breast_excl_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature) %>%
  filter(feature != "1")

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r breast_excl_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "\u2265 2"),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r breast_excl_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "\u2265 2"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r breast_excl_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "\u2265 2"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```


#### Plot all
```{r breast_excl_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Smoke during pregnancy

```{r}
feature_name <- "Smoke during Pregnancy"
feat_key     <- "Prenatal_smoke"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "pregsmoke")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature) %>%
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r pregsmoke_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r pregsmoke_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r pregsmoke_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all

```{r pregsmoke_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Number of siblings

```{r}
feature_name <- "Number of siblings"
feat_key     <- "Siblings"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "sibs_numb_cat")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r sibsnumb_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "&gt;1"),
              ptest = perm_results[[paste0(feat_key, "_richness_genus")]])
```

#### Shannon

```{r sibsnumb_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "&gt;1"),
             ptest = perm_results[[paste0(feat_key, "_shannon")]]) 
```

#### Gini-Simpson index

```{r sibsnumb_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "&gt;1"),
          ptest = perm_results[[paste0(feat_key, "_gini")]])
```

#### Plot all

```{r sibsnumb_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```


## Model-based analysis: regression on Shannon diversity (betta)

The strip charts above show unadjusted, per-sample DivNet estimates. Here we use
the covariate model (`divnet_genus_cov`, family level) together with
`breakaway::betta` to jointly estimate associations between all covariates and
Shannon diversity, properly accounting for the uncertainty in the DivNet estimates.
Note that this model uses only the `r nsamples(phylo_genus_cov)` samples with
complete covariate data.

```{r}
# Drop samples with NAs in any formula covariate
# Also exclude samples with 1-month exclusive breastfeeding (sparse category)
cov_vars <- c("Country", "Sex", "cesarean", "breast_dur_cat1",
              "breast_excl_cat1", "pregsmoke", "sibs_numb_cat")
sdata <- data.frame(sample_data(phylo_genus))
complete_samples <- rownames(sdata)[complete.cases(sdata[, cov_vars])]
complete_samples <- complete_samples[sdata[complete_samples, "breast_excl_cat1"] != "1"]
phylo_genus_cov <- prune_samples(complete_samples, phylo_genus)

if (!file.exists(path_target("divnet_genus_cov.rds"))) {
  set.seed(123456)
  divnet_start <- proc.time()
  divnet_genus_cov <- phylo_genus_cov %>%
    DivNet::divnet(formula = ~ Country + Sex + cesarean + breast_dur_cat1 + 
                     breast_excl_cat1 + pregsmoke + sibs_numb_cat,
                   ncores = 1, 
                   tuning = "fast")
  
  divnet_runtime <- proc.time() - divnet_start
  
  saveRDS(divnet_genus_cov, file = path_target("divnet_genus_cov.rds"))
  saveRDS(divnet_runtime, file = path_target("divnet_genus_cov_runtime.rds"))
} else {
  divnet_genus_cov <- readRDS(path_target("divnet_genus_cov.rds"))
  divnet_runtime <- readRDS(path_target("divnet_genus_cov_runtime.rds"))
}
```

```{r}
names(divnet_genus_cov)
```

```{r}
betta_estimates <- divnet_genus_cov$shannon |> summary() |> dplyr::pull(estimate)
betta_ses       <- sqrt(divnet_genus_cov$`shannon-variance`)

betta_shannon <- breakaway::betta(betta_estimates, betta_ses, divnet_genus_cov$X)
betta_shannon$table
```

```{r}
gini_estimates <- 1 - (divnet_genus_cov$simpson |> summary() |> dplyr::pull(estimate))
gini_ses       <- sqrt(divnet_genus_cov$`simpson-variance`)

betta_gini <- breakaway::betta(gini_estimates, gini_ses, divnet_genus_cov$X)
betta_gini$table
```

```{r}
# Rerun breakaway only on the model samples (n = 486) to obtain aligned SEs.
# One sample has a near-zero SE (~5e-95); this is floored at 0.5
# (conservative relative to the estimate scale of ~6–54) to avoid
# a singular covariance matrix in betta().
model_samples  <- rownames(divnet_genus_cov$X)
flist_cov      <- frequencytablelist[model_samples]
rich_cov       <- lapply(flist_cov, breakaway::breakaway)
rich_estimates <- sapply(rich_cov, function(x) x$estimate)
rich_ses       <- pmax(sapply(rich_cov, function(x) x$error), 0.5)

betta_rich <- breakaway::betta(rich_estimates, rich_ses, divnet_genus_cov$X)
betta_rich$table
```

**Interpretation** (reference group: Austria, female, vaginal birth, breastfeeding duration < 1 month,
not exclusively breastfed, no prenatal smoking, no siblings):

- **Country:** Swiss infants show lower genus-level Shannon diversity than Austrian infants
  (β = −0.113, p < 0.001); German infants do not differ significantly (β = +0.002, p = 0.800).
- **Sex:** Male infants show marginally lower Shannon diversity than females (β = −0.008,
  p = 0.059), a difference that does not reach conventional significance at the genus level.
- **Caesarean delivery:** Associated with slightly higher Shannon diversity (β = +0.019, p = 0.016).
  This is somewhat counterintuitive — caesarean birth typically disrupts vertical transmission
  of vaginal/gut microbiota — and may reflect confounding with feeding mode or other factors.
- **Breastfeeding duration:** The dominant effect in the model. Compared to very short/no
  breastfeeding (< 1 month), one month of breastfeeding is associated with modestly higher
  diversity (β = +0.039, p = 0.001), while ≥ 2 months is associated with substantially lower
  diversity (β = −0.359, p < 0.001). This non-monotone pattern likely reflects the well-documented
  HMO-driven enrichment of *Bifidobacterium* in longer-breastfed infants, which reduces
  overall evenness (Shannon diversity) at the genus level.
- **Exclusive breastfeeding:** Duration ≥ 2 months is associated with slightly lower
  diversity (β = −0.058, p < 0.001), consistent with the breastfeeding duration effect above.
- **Prenatal smoking:** Associated with higher infant Shannon diversity (β = +0.073, p < 0.001).
  The direction is unexpected and warrants cautious interpretation; residual confounding
  cannot be excluded.
- **Number of siblings:** Having more than one sibling is associated with slightly lower Shannon
  diversity (> 1 sibling: β = −0.022, p < 0.001); having exactly one sibling shows no significant
  effect (β = −0.004, p = 0.391). The negative direction is counterintuitive given the expected
  increase in microbial exposure from older siblings, and may interact with breastfeeding or
  other covariates.

Global test of whether any covariate is associated with Shannon diversity:

```{r}
betta_shannon$global
```

`betta_shannon$global` is a likelihood ratio test of the null hypothesis that all covariates
jointly have no association with Shannon diversity (i.e., all βs except the intercept equal
zero). The two returned values are the chi-squared test statistic and its p-value,
respectively. A significant result indicates that the model as a whole fits better than an
intercept-only model, meaning at least one covariate is meaningfully associated with Shannon
diversity. It does not identify which covariates are responsible — that is addressed by the
individual p-values in `betta_shannon$table` above.


### Combined coefficient plot

```{r betta_forest_combined, fig.width=10, fig.height=6}
term_labels <- c(
  "CountryGermany"      = "Country: Germany",
  "CountrySwitzerland"  = "Country: Switzerland",
  "Sexm"                = "Sex: male",
  "cesarean"            = "Caesarean delivery",
  "breast_dur_cat11"    = "Breastfeeding duration: 1 month",
  "breast_dur_cat1>=2"  = "Breastfeeding duration: \u22652 months",
  "breast_excl_cat11"   = "Exclusive BF: 1 month",
  "breast_excl_cat1>=2" = "Exclusive BF: \u22652 months",
  "pregsmoke"           = "Prenatal smoking",
  "sibs_numb_cat1"      = "Siblings: 1",
  "sibs_numb_cat>1"     = "Siblings: >1"
)

tidy_betta <- function(betta_res, measure_label) {
  as.data.frame(betta_res$table) |>
    tibble::rownames_to_column("term") |>
    filter(term != "(Intercept)") |>
    mutate(
      lower   = Estimates - 1.96 * `Standard Errors`,
      upper   = Estimates + 1.96 * `Standard Errors`,
      sig     = `p-values` < 0.05,
      measure = measure_label,
      term    = dplyr::recode(term, !!!term_labels)
    )
}

bind_rows(
  tidy_betta(betta_shannon, "Shannon (DivNet)"),
  tidy_betta(betta_gini,   "Gini-Simpson (DivNet)"),
  tidy_betta(betta_rich,   "Breakaway richness")
) |>
  mutate(
    measure = factor(measure, levels = c("Breakaway richness",
                                         "Shannon (DivNet)",
                                         "Gini-Simpson (DivNet)")),
    term = factor(term, levels = rev(unique(term)))
  ) |>
  ggplot(aes(x = Estimates, y = term, color = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = "#1B7837"),
                     labels = c("FALSE" = "no", "TRUE"  = "yes")) +
  facet_wrap(~measure, scales = "free_x", nrow = 1) +
  labs(x = "Coefficient estimate (95% CI)", y = NULL, color = "p < 0.05") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

#### Investigating the C-section effect: confounding by breastfeeding duration

The positive association of caesarean delivery with Shannon diversity is counterintuitive
biologically. A likely explanation is confounding by breastfeeding: C-section delivery is
associated with shorter breastfeeding duration, and shorter breastfeeding is associated with
higher Shannon diversity (the dominant effect in the model). The following two plots
investigate this.

```{r cesarean_breastfeed_confounding, fig.width=9, fig.height=4}
plot_dat <- alpha_diversity_div |>
  left_join(
    sdata |>
      select(SampleID, cesarean, breast_dur_cat1),
    by = "SampleID"
  ) |>
  filter(!is.na(cesarean), !is.na(breast_dur_cat1)) |>
  mutate(
    cesarean = factor(cesarean, levels = c(1, 0), labels = c("C-section", "Vaginal")),
    breast_dur_cat1 = factor(breast_dur_cat1,
                             levels = c("0", "1", ">=2"),
                             labels = c("< 1 month", "1 month", "≥ 2 months"))
  )

# Panel 1: breastfeeding duration distribution by delivery mode
p1 <- plot_dat |>
  count(cesarean, breast_dur_cat1) |>
  group_by(cesarean) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = cesarean, y = pct, fill = breast_dur_cat1)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Delivery mode", y = "Proportion", fill = "Breastfeeding\nduration") +
  theme_bw()

# Panel 2: Shannon diversity by delivery mode, stratified by breastfeeding duration
p2 <- ggplot(plot_dat, aes(x = cesarean, y = shannon)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_wrap(~breast_dur_cat1) +
  labs(x = "Delivery mode", y = "Shannon index (DivNet)") +
  theme_bw()

cowplot::plot_grid(p1, p2, rel_widths = c(1, 1.8))
```

The left panel shows that C-section infants are less likely to be breastfed for ≥ 2 months
compared to vaginally born infants. Since prolonged breastfeeding is the strongest single
predictor of *lower* Shannon diversity in this cohort, C-section infants appear more diverse
in the unadjusted comparison. The right panel confirms this: within each breastfeeding
duration stratum, the delivery mode difference is substantially attenuated or absent.

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
