---
title: "Alpha diversity analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "02_alpha_diversity" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "41b33e9b-a351-49eb-bdc8-cdeddd77796c")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params, "/alpha_"),
  echo = FALSE)
```

## Load packages

```{r packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(ggtext)
library(ggpubr)

# Diversity packages
library(breakaway)
library(DivNet)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data

```{r load-data}
phylo_asv   <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_asv.rds"))
phylo_genus <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_genus.rds"))
sampdata <- sample_data(phylo_asv) %>% as("data.frame")
otutab_asv   <- as(otu_table(phylo_asv), "matrix")
otutab_genus <- as(otu_table(phylo_genus), "matrix")
```

### ASV level

```{r}
phylo_asv
```

### Genus level

```{r}
phylo_genus
```


## Colors

```{r}
cols = c("#009E73", "#56B4E9", "#E69F00", "#D55E00", 
         "#CC79A7", "#F0E442", "#0072B2", "gray")
```

## Functions

### Function for significance plot

```{r}
modify_div_plot <- function(strip_chart, xlab = "", breaks = NULL, labels = NULL,
                            binwidth = 0.5, dotsize = 1, ylim = c(0, 75),
                            ylab = "", ybreaks = NULL,
                            textsize = 18, title = "") {
  strip_chart <- strip_chart +
    stat_summary(fun.data = median_hilow, fun.args=0.50, show.legend=FALSE,
                 geom="crossbar", alpha=0.25, width=0.7) +
    geom_dotplot(binaxis = "y", binwidth = binwidth, 
                 stackdir = "center", width = 0.5, alpha = 0.7,
                 show.legend = FALSE, dotsize = dotsize) +
    labs(x=NULL,
         y=ylab) +
    scale_fill_manual(values = cols) +
    scale_x_discrete(name = xlab,
                     breaks = breaks, 
                     labels = labels) +
    theme_classic() +
    theme(axis.text.x = element_markdown(),
          axis.text.y = element_markdown(),
          legend.text = element_markdown(),
          text = element_text(size = textsize)) +
    ggtitle(title)
  
  if (!is.null(ybreaks)) {
    strip_chart <- strip_chart + 
      scale_y_continuous(breaks=ybreaks, limits = ylim)
  } else {
    strip_chart <- strip_chart + 
      scale_y_continuous(limits = ylim)
  }
  strip_chart
}

```

### Function for significance line

```{r}
add_sig_line <- function(strip_chart, pval, diversity,
                         space_line = 0.05, space_lab = 0.05, nfeat = nfeat, 
                         linewidth = 0.8, size = 5) {
  
  if (nfeat == 2) {
    pmat_x <- c(1)
    pmat_y <- c(1)
    line_left <- c(1)
    line_right <- c(2)
    lab_x <- c(1.5)
  } else if (nfeat == 3) {
    pmat_x <- c(1, 2, 2)
    pmat_y <- c(1, 2, 1)
    line_left <- c(1, 2, 1)
    line_right <- c(2, 3, 3)
    lab_x <- c(1.5, 2.5, 2)
  } else {
    pmat_x <- c(1, 2, 3, 2, 3, 3)
    pmat_y <- c(1, 2, 3, 1, 2, 1)
    line_left <- c(1, 2, 3, 1, 2, 1)
    line_right <- c(2, 3, 4, 3, 4, 4)
    lab_x <- c(1.5, 2.5, 3.5, 2, 3, 2.5)
  }
  
  pos_line <- max(diversity) - space_line + 0.1 * space_line
  pos_lab <- pos_line + space_lab
  
  for (i in seq_along(pmat_x)) {
    if (is.null(dim(pval))) {
      label <- pval
    } else {
      label <- pval[pmat_x[i], pmat_y[i]]
    }
    
    if (label <= 0.05) {
      pos_line <- pos_line + space_line
      pos_lab <- pos_lab + space_line
      
      strip_chart <- strip_chart +
        geom_line(data = tibble(x = c(line_left[i], line_right[i]), 
                                y = c(pos_line, pos_line)), 
                  aes(x=x, y=y),
                  inherit.aes = FALSE, 
                  linewidth = linewidth) +
        #geom_text(x = lab_x[i], y = pos_lab, label = signif(label, digits = 3))
        annotate("text", lab_x[i], y = pos_lab, 
                 label = signif(label, digits = 3), size = size)
    } 
  }
  strip_chart
}
```

### Plot function for richness

```{r}
plot_richness <- function(div_dataframe, feature_name, 
                          xlab = "", breaks = NULL, labels = NULL,
                          binwidth = 0.4, dotsize = 1.2, 
                          ylab = "Breakaway richness", 
                          ybreaks = NULL, title = "",
                          ylim = c(0, 75), textsize = 18,
                          space_line = 2, space_lab = 5) {
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=richness_genus, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  pval_kt <- pval <- kruskal.test(richness_genus ~ feature, 
                                  data = div_dataframe)$p.value
  pval_kt
  
  if (pval_kt <= 0.05 && nfeat > 2) {
    pval <- pairwise.wilcox.test(div_dataframe$richness_genus,
                                 g = div_dataframe$feature,
                                 p.adjust.method = "BH")$p.value
    pval
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$richness_genus,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```

### Plot function for Shannon

```{r}
plot_shannon <- function(div_dataframe, feature_name, 
                         xlab = "", breaks = NULL, labels = NULL,
                         binwidth = 0.04, dotsize = 1, ylim = c(0, 4), 
                         space_line = 0.3, space_lab = 0.15,
                         ylab = "Shannon index (DivNet)", 
                         ybreaks = NULL, title = "", textsize = 18) {
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=shannon, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  pval_kt <- pval <- kruskal.test(shannon ~ feature, 
                                  data = div_dataframe)$p.value
  pval_kt
  
  if (pval_kt <= 0.05 && nfeat > 2) {
    pval <- pairwise.wilcox.test(div_dataframe$shannon,
                                 g = div_dataframe$feature,
                                 p.adjust.method = "BH")$p.value
    pval
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$shannon,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```

### Plot function for Gini-Simpson

```{r}
plot_gini <- function(div_dataframe, feature_name, 
                      xlab = "", breaks = NULL, labels = NULL,
                      binwidth = 0.01, dotsize = 1, ylim = c(0, 1.1), 
                      space_line = 0.08, space_lab = 0.04,
                      ylab = "Gini-Simpson index (DivNet)", 
                      ybreaks = NULL, title = "", textsize = 18) {
  
  # Remove levels with zero observations
  div_dataframe <- droplevels(div_dataframe[div_dataframe$feature %in% 
                                              names(which(table(div_dataframe$feature) > 0)), ])
  nfeat <- nlevels(div_dataframe$feature)
  
  strip_chart <- div_dataframe %>%
    ggplot(aes(x=feature, y=gini, fill=feature)) 
  
  strip_chart <- modify_div_plot(strip_chart, 
                                 xlab = xlab, breaks = breaks, labels = labels,
                                 binwidth = binwidth, dotsize = dotsize, 
                                 ylim = ylim, ylab = ylab, ybreaks = ybreaks,
                                 textsize = textsize, title = title) 
  
  # Significance test
  pval_kt <- pval <- kruskal.test(gini ~ feature, 
                                  data = div_dataframe)$p.value
  pval_kt
  
  if (pval_kt <= 0.05 && nfeat > 2) {
    pval <- pairwise.wilcox.test(div_dataframe$gini,
                                 g = div_dataframe$feature,
                                 p.adjust.method = "BH")$p.value
    pval
  }
  
  if (pval_kt <= 0.05) {
    strip_chart <- add_sig_line(strip_chart, pval = pval, nfeat = nfeat,
                                diversity = div_dataframe$gini,
                                space_line = space_line, 
                                space_lab = space_lab)
  }
  strip_chart
}
```


## Compute alpha diversity

### Alpha diversity on ASV level

Computing alpha diversity the common way

```{r}
alpha_div_common <- estimate_richness(phylo_asv)
head(alpha_div_common)
```

### Breakaway richness on ASV level

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_asv)
```

```{r}
rich_asv <- unlist(lapply(frequencytablelist, function(x) breakaway(x)$estimate))

alpha_diversity_rich <- data.frame(SampleID = names(rich_asv),
                                   age = "2 months",
                                   richness_otu = rich_asv)
head(alpha_diversity_rich)
```

### Alpha diversity on genus level

Computing alpha diversity the common way

```{r}
alpha_div_common_genus <- estimate_richness(phylo_genus)
head(alpha_div_common_genus)
```

### Breakaway  richness on genus level

```{r}
frequencytablelist <- build_frequency_count_tables(otutab_genus)
```

```{r}
rich_genus <- 
  data.frame(SampleID = names(frequencytablelist),
             richness_genus = unlist(lapply(frequencytablelist, 
                                            function(x) breakaway(x)$estimate))) 
```


```{r}
alpha_diversity_rich <- inner_join(alpha_diversity_rich, 
                                      rich_genus, by = "SampleID")
head(alpha_diversity_rich)
```

## DivNet diversity measures

Due to the long runtime only on genus level

Find base taxon (with minimum number of non-zero counts)

```{r}
nonzero_elem <- apply(otutab_genus, 1, function(x) sum(x == 0))
base_taxon <- which.min(nonzero_elem)
nonzero_elem[base_taxon]
```

```{r}
if (!file.exists(path_target("divnet_genus.rds"))) {
  divnet_result <- phylo_genus %>%
    DivNet::divnet(ncores = 8, 
                   base = names(base_taxon),
                   tuning = "fast")
  
  saveRDS(divnet_result, file = path_target("divnet_genus.rds"))
} else {
  divnet_result <- readRDS(path_target("divnet_genus.rds"))
}
```

```{r}
phylo_family <- tax_glom(phylo_genus, taxrank="Family")

# Drop samples with NAs in any formula covariate
cov_vars <- c("Country", "Sex", "cesarean", "breast_dur_cat1",
              "breast_excl_cat1", "pregsmoke", "sibs_numb_cat")
sdata <- data.frame(sample_data(phylo_family))
complete_samples <- rownames(sdata)[complete.cases(sdata[, cov_vars])]
phylo_family_cov <- prune_samples(complete_samples, phylo_family)

if (!file.exists(path_target("divnet_family_cov.rds"))) {
  divnet_result <- phylo_family_cov %>%
    DivNet::divnet(formula = ~ Country + Sex + cesarean + breast_dur_cat1 + 
                     breast_excl_cat1 + pregsmoke + sibs_numb_cat,
                   ncores = 8, 
                   tuning = "fast")
  
  saveRDS(divnet_result, file = path_target("divnet_family_cov.rds"))
} else {
  divnet_result <- readRDS(path_target("divnet_family_cov.rds"))
}
```

```{r}
names(divnet_result)
```

```{r}
alpha_diversity_div <- 
  data.frame(SampleID = names(divnet_result$shannon),
             shannon = unlist(lapply(divnet_result$shannon, function(x) x$estimate)),
             simpson = unlist(lapply(divnet_result$simpson, function(x) x$estimate)))

alpha_diversity_div <- alpha_diversity_div %>%
  mutate(inv_simpson = 1 / simpson,
         gini = 1 - simpson)
```

### Regression on Shannon diversity (betta)

Using `breakaway::betta` to regress DivNet Shannon estimates on the covariates.
The design matrix is taken directly from the fitted DivNet model (`divnet_result$X`),
which encodes the same formula used above.

```{r}
betta_estimates <- divnet_result$shannon %>% summary() %$% estimate
betta_ses       <- sqrt(divnet_result$`shannon-variance`)

betta_result <- breakaway::betta(betta_estimates, betta_ses, divnet_result$X)
betta_result$table
```

**Interpretation** (reference group: Austria, female, vaginal birth, breastfeeding duration < 1 month,
not exclusively breastfed, no prenatal smoking, no siblings):

- **Country:** Swiss infants show substantially lower Shannon diversity than Austrian infants
  (β = −0.156, p < 0.001); German infants do not differ significantly (p = 0.226).
- **Sex:** Male infants have slightly but significantly lower Shannon diversity than females
  (β = −0.024, p < 0.001).
- **Caesarean delivery:** Associated with higher Shannon diversity (β = +0.050, p < 0.001).
  This is somewhat counterintuitive — caesarean birth typically disrupts vertical transmission
  of vaginal/gut microbiota — and may reflect confounding with feeding mode or other factors.
- **Breastfeeding duration:** The largest effect in the model. Compared to very short/no
  breastfeeding (< 1 month), one month of breastfeeding is associated with higher diversity
  (β = +0.117, p < 0.001), while ≥ 2 months is associated with substantially lower diversity
  (β = −0.382, p < 0.001). This non-monotone pattern likely reflects the well-documented
  HMO-driven enrichment of *Bifidobacterium* in longer-breastfed infants, which reduces
  overall evenness (Shannon diversity).
- **Exclusive breastfeeding:** The 1-month exclusive breastfeeding category is not
  significant (p = 0.708; n = 2, very sparse). Duration ≥ 2 months is associated with
  slightly lower diversity (β = −0.046, p < 0.001), consistent with the duration effect above.
- **Prenatal smoking:** Associated with higher infant Shannon diversity (β = +0.116, p < 0.001).
  The direction is unexpected and warrants cautious interpretation; residual confounding
  cannot be excluded.
- **Number of siblings:** Having siblings is associated with slightly lower Shannon diversity
  (1 sibling: β = −0.013, p = 0.011; > 1 sibling: β = −0.033, p < 0.001). This direction
  is counterintuitive given the expected increase in microbial exposure from older siblings,
  and may interact with breastfeeding duration or other covariates in the model.

Global test of whether any covariate is associated with Shannon diversity:

```{r}
betta_result$global
```

```{r}
alpha_diversity <- inner_join(alpha_diversity_rich, 
                              alpha_diversity_div, 
                              by = "SampleID")
```


```{r}
head(alpha_diversity)
```

## Study center

```{r}
feature_name <- "Country"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "Country")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>%
  mutate(feature = factor(feature)) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

# metadata_alpha[which.max(metadata_alpha$richness_genus), "richness_genus"] <-
#   max(metadata_alpha$richness_genus) / 2

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r Country_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.5, dotsize = 1.2, 
              ylim = c(0, 90), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature))
```

#### Shannon

```{r Country_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature)) 
```

#### Gini-Simpson index

```{r Country_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature))
```

#### Plot all

```{r Country_2m, fig.width=14, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

```{r}
table(sampdata$Country, sampdata$Farmer)
```


## Sex

```{r}
feature_name <- "Sex"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "Sex")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, female = "f", male = "m")) %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r sex_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature))
```

#### Shannon

```{r sex_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature)) 
```

#### Gini-Simpson index

```{r sex_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature))
```

#### Plot all

```{r sex_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## C-section

```{r}
feature_name <- "C-section"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "cesarean")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  drop_na(feature) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r cesarean_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature))
```

#### Shannon

```{r cesarean_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature)) 
```

#### Gini-Simpson index

```{r cesarean_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature))
```

#### Plot all

```{r cesarean_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg, ncol = 3, nrow = 1)
```


## Breast feeding

```{r}
feature_name <- "Breast feeding"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "breast_dur_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r breastfeed_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "\u2265 2"))
```

#### Shannon

```{r breastfeed_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "\u2265 2")) 
```

#### Gini-Simpson index

```{r breastfeed_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "\u2265 2"))
```

#### Plot all
```{r breastfeed_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Exclusive breast feeding

```{r}
feature_name <- "Exclusive breast feeding"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "breast_excl_cat1")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r breast_excl_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.3, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "\u2265 2"))
```

#### Shannon

```{r breast_excl_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "\u2265 2")) 
```

#### Gini-Simpson index

```{r breast_excl_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "\u2265 2"))
```


#### Plot all
```{r breast_excl_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Smoke during pregnancy

```{r}
feature_name <- "Smoke during Pregnancy"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "pregsmoke")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature) %>%
  mutate(feature = as.factor(feature)) %>%
  mutate(feature = fct_recode(feature, no = "0", yes = "1")) %>%
  mutate(feature = factor(feature,
                          levels=c("yes", 
                                   "no")))

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r pregsmoke_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature))
```

#### Shannon

```{r pregsmoke_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature)) 
```

#### Gini-Simpson index

```{r pregsmoke_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature))
```

#### Plot all

```{r pregsmoke_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## BMI mother

```{r}
feature_name <- "BMI mother"
```


### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "BMI_mum_fact")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>%
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r bmi_mother_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = levels(metadata$feature))
```

#### Shannon

```{r bmi_mother_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = levels(metadata$feature)) 
```

#### Gini-Simpson index

```{r bmi_mother_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = levels(metadata$feature))
```

#### Plot all

```{r bmi_mother_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```

## Number of siblings

```{r}
feature_name <- "Number of siblings"
```

### 2 months

```{r}
metadata <- sampdata[, c("SampleID", "sibs_numb_cat")]

colnames(metadata) <- c("SampleID", "feature")
rownames(metadata) <- NULL

metadata <- metadata %>% 
  drop_na(feature)

metadata_alpha <- inner_join(metadata, alpha_diversity,
                             by = "SampleID")

nfeat <- nlevels(metadata_alpha$feature)
table(metadata$feature)
```

#### Richness

```{r sibsnumb_richness_2m, fig.width=5, fig.height=5}
pr <- plot_richness(metadata_alpha, feature_name, binwidth = 0.4, dotsize = 1.2, 
              ylim = c(0, 80), space_line = 6, space_lab = 3,
              xlab = feature_name, 
              breaks = levels(metadata$feature),
              labels = c("0", "1", "&gt;1"))
```

#### Shannon

```{r sibsnumb_shannon_2m, fig.width=5, fig.height=5}
ps <- plot_shannon(metadata_alpha, feature_name, binwidth = 0.04, dotsize = 1, 
             ylim = c(0, 4), space_line = 0.3, space_lab = 0.15,
             xlab = feature_name, 
             breaks = levels(metadata$feature),
             labels = c("0", "1", "&gt;1")) 
```

#### Gini-Simpson index

```{r sibsnumb_simpson_2m, fig.width=5, fig.height=5}
pg <- plot_gini(metadata_alpha, feature_name, binwidth = 0.01, dotsize = 1, 
          ylim = c(0, 1.3), space_line = 0.09, space_lab = 0.04,
          xlab = feature_name, 
          breaks = levels(metadata$feature),
          labels = c("0", "1", "&gt;1"))
```

#### Plot all

```{r sibsnumb_2m, fig.width=12, fig.height=5}
ggarrange(pr, ps, pg,
          ncol = 3, nrow = 1)
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
