---
title: "Mini example: zero replacement before CLR transformation"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "CLR_mini_example" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "7aac30f8-96f0-4cb8-b30e-a3c181fc95c0")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params, "/clr_exmpl_"),
  echo = FALSE)
```

```{r}
library("conflicted")
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

# Aim

This file reproduces the small toy example used to compare multiplicative replacement and additive offset replacement before applying the centered log-ratio (CLR) transformation.

The example is not intended as a simulation study. Its purpose is to show that the two replacement strategies can lead to different transformed values and therefore affect downstream quantities such as Aitchison distances, marginal transformed values, group contrasts, and pairwise log-ratios used in association or proportionality analyses.

# Packages

```{r packages}
library(dplyr)
library(tidyr)
library(tibble)
library(knitr)
library(ggplot2)
library(tidyverse)
library(ggh4x)


conflicts_prefer(dplyr::select)
```

# Helper functions

```{r helper-functions}
clr <- function(x) {
  log(x) - mean(log(x))
}

rel_norm <- function(x) {
  x / sum(x)
}

multiplicative_replacement <- function(x, delta = 1) {
  if (any(x == 0)) {
    as.matrix(multRepl(x, label = 0, frac = 1, dl = rep(delta, length(x))))
  } else {
    x
  }
}

additive_offset_replacement <- function(x, delta = 1) {
  x + delta
}

make_table <- function(mat, group) {
  as_tibble(mat, rownames = "sample") |>
    mutate(group = group, .after = sample) |>
    mutate(sum = rowSums(mat))
}

apply_by_row <- function(mat, fun, ...) {
  out <- t(apply(mat, 1, fun, ...))
  colnames(out) <- colnames(mat)
  rownames(out) <- rownames(mat)
  out
}
```

# Original toy data

```{r original-data}
X <- matrix(
  c(
    0, 4, 96,
    1, 4, 95,
    0, 8, 92,
    1, 8, 91
  ),
  nrow = 4,
  byrow = TRUE
)

rownames(X) <- paste0("S", 1:4)
colnames(X) <- paste0("Taxon ", 1:3)

group <- c("A", "A", "B", "B")

delta <- 1

original_table <- make_table(X, group)

kable(original_table, digits = 3, caption = "Original toy count data.")
```

# Replacement strategies

The multiplicative replacement uses a small value \(\delta\) for zeros and rescales the positive components so that the sample sum is preserved. The additive offset replacement adds \(\delta\) to all components.

```{r replacements}
X_mult <- apply_by_row(X, multiplicative_replacement, delta = delta)
X_add <- apply_by_row(X, additive_offset_replacement, delta = delta)

mult_table <- make_table(X_mult, group)
add_table <- make_table(X_add, group)

kable(mult_table, digits = 3, caption = "Toy data after multiplicative replacement with delta = 1.")
kable(add_table, digits = 3, caption = "Toy data after additive offset replacement with delta = 1.")
```

# CLR-transformed values

```{r clr-values}
CLR_mult <- apply_by_row(X_mult, clr)
CLR_add <- apply_by_row(X_add, clr)

clr_mult_table <- make_table(CLR_mult, group)
clr_add_table <- make_table(CLR_add, group)

# The row sums of CLR values are zero up to numerical precision.
clr_mult_table <- clr_mult_table |> mutate(sum = round(sum, 12))
clr_add_table <- clr_add_table |> mutate(sum = round(sum, 12))

kable(clr_mult_table, digits = 3, caption = "CLR values after multiplicative replacement.")
kable(clr_add_table, digits = 3, caption = "CLR values after additive offset replacement.")
```



# Visualization of all transformed values

```{r original_transformed_values, fig.height=4.5, fig.width=8, dpi = 200}
# Compact plot: original counts and CLR-transformed values ----------------
# Fixed horizontal positions for samples within each group
pos_sample <- position_dodge(width = 0.35)

# Original counts
count_values <- as_tibble(X, rownames = "sample") |>
  mutate(group = group, .after = sample) |>
  pivot_longer(
    cols = starts_with("Taxon "),
    names_to = "taxon",
    values_to = "value"
  ) |>
  mutate(
    representation = "Original counts"
  )

# CLR-transformed values
clr_values <- bind_rows(
  as_tibble(CLR_mult, rownames = "sample") |>
    mutate(
      group = group,
      representation = "MultRepl + CLR",
      .after = sample
    ),
  as_tibble(CLR_add, rownames = "sample") |>
    mutate(
      group = group,
      representation = "Offset + CLR",
      .after = sample
    )
) |>
  pivot_longer(
    cols = starts_with("Taxon "),
    names_to = "taxon",
    values_to = "value"
  )

# Combine into one plotting data set
plot_values <- bind_rows(count_values, clr_values) |>
  mutate(
    representation = factor(
      representation,
      levels = c(
        "Original counts",
        "MultRepl + CLR",
        "Offset + CLR"
      )
    ),
    sample = factor(sample, levels = rownames(X)),
    taxon = factor(taxon, levels = colnames(X))
  )

# Plot
p_counts_and_clr_compact <- ggplot(
  plot_values,
  aes(x = group, y = value, group = sample)
) +
  geom_hline(
    data = plot_values |> filter(representation != "Original counts"),
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +
  geom_point(
    aes(shape = sample),
    size = 2,
    alpha = 0.9,
    position = pos_sample
  ) +
  ggh4x::facet_grid2(
    representation ~ taxon,
    scales = "free_y",
    independent = "y",
    switch = "y"
  ) +
  labs(
    x = "Group",
    y = NULL,
    shape = "Sample"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.spacing.x = unit(1.2, "lines"),
    panel.spacing.y = unit(1.2, "lines"),
    axis.title.x = element_text(size = 12),
    strip.text.x = element_text(size = 11),
    strip.text.y.left = element_text(size = 11, angle = 90),
    strip.placement = "outside",
    legend.position = "right"
  )

p_counts_and_clr_compact
```


# Distance-based analysis

We compare samples S1 and S2. These samples are nearly identical after multiplicative replacement but differ more clearly after additive offset replacement.

```{r distance-example}
dist_mult_S1_S2 <- sqrt(sum((CLR_mult["S1", ] - CLR_mult["S2", ])^2))
dist_add_S1_S2 <- sqrt(sum((CLR_add["S1", ] - CLR_add["S2", ])^2))

distance_results <- tibble(
  comparison = "S1 vs S2",
  `Multiplicative replacement + CLR` = dist_mult_S1_S2,
  `Additive offset + CLR` = dist_add_S1_S2
)

kable(distance_results, digits = 3, caption = "Euclidean distances between CLR-transformed samples S1 and S2.")
```

# Differential distribution example

For differential distribution analyses of transformed taxa, the transformed values themselves are the object of comparison. Here we inspect the CLR-transformed values of Taxon 2.

```{r differential-distribution}
taxon <- "Taxon 2"

taxon2_values <- tibble(
  sample = rownames(X),
  group = group,
  original_count = X[, taxon],
  clr_mult = CLR_mult[, taxon],
  clr_add = CLR_add[, taxon]
)

kable(taxon2_values, digits = 3, caption = "CLR-transformed values of Taxon 2 under the two replacement strategies.")
```

```{r differential_distribution_plot}
taxon2_values_long <- taxon2_values |>
  select(sample, group, clr_mult, clr_add) |>
  pivot_longer(
    cols = c(clr_mult, clr_add),
    names_to = "method",
    values_to = "clr_value"
  ) |>
  mutate(
    method = recode(
      method,
      clr_mult = "MultRepl + CLR",
      clr_add = "Offset + CLR"
    )
  )

ggplot(taxon2_values_long, aes(x = group, y = clr_value)) +
  geom_point(size = 2, position = position_jitter(width = 0.04, height = 0)) +
  facet_wrap(~ method) +
  labs(
    x = "Group",
    y = "CLR-transformed value of Taxon 2",
    title = "Transformed values entering differential distribution analyses"
  ) +
  theme_minimal(base_size = 11)
```

# Differential abundance-style group contrast

A simple feature-level group comparison on the CLR scale can be represented by the difference between the mean transformed values in group B and group A. This is not a realistic differential abundance test, because the toy data contain only two samples per group. It illustrates only that the transformed group difference depends on the zero-handling strategy.

```{r da-contrast}
da_contrast <- taxon2_values |>
  summarise(
    delta_mult = mean(clr_mult[group == "B"]) - mean(clr_mult[group == "A"]),
    delta_add = mean(clr_add[group == "B"]) - mean(clr_add[group == "A"])
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "method",
    values_to = "group_difference_B_minus_A"
  ) |>
  mutate(
    method = recode(
      method,
      delta_mult = "MultRepl + CLR",
      delta_add = "Offset + CLR"
    )
  )

kable(da_contrast, digits = 3, caption = "Difference in mean CLR-transformed values of Taxon 2 between groups B and A.")
```

# Association and proportionality example

For association and proportionality analyses, pairwise log-ratios and their variation across samples are central. We compare Taxon 2 and Taxon 3, which are positive in all four samples. Multiplicative replacement preserves their ratios, whereas additive offset replacement changes them.

```{r association-example}
logratio_original <- log(X[, "Taxon 2"] / X[, "Taxon 3"])
logratio_mult <- log(X_mult[, "Taxon 2"] / X_mult[, "Taxon 3"])
logratio_add <- log(X_add[, "Taxon 2"] / X_add[, "Taxon 3"])

# Equivalent coordinate differences after CLR transformation
nclr_diff_mult <- CLR_mult[, "Taxon 2"] - CLR_mult[, "Taxon 3"]
clr_diff_add <- CLR_add[, "Taxon 2"] - CLR_add[, "Taxon 3"]

logratio_table <- tibble(
  sample = rownames(X),
  group = group,
  original_logratio_T2_T3 = logratio_original,
  mult_logratio_T2_T3 = logratio_mult,
  add_logratio_T2_T3 = logratio_add,
  mult_clr_difference_T2_T3 = nclr_diff_mult,
  add_clr_difference_T2_T3 = clr_diff_add
)

kable(logratio_table, digits = 3, caption = "Sample-wise log-ratios between Taxon 2 and Taxon 3.")
```

```{r association-summary}
association_summary <- tibble(
  method = c("Original", "MultRepl + CLR", "Offset + CLR"),
  variance_logratio_T2_T3 = c(
    var(logratio_original),
    var(logratio_mult),
    var(logratio_add)
  )
)

kable(association_summary, digits = 4, caption = "Variance of the log-ratio between Taxon 2 and Taxon 3.")
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
