---
title: "CLR example"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "CLR_example" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "e2a8506c-47b1-4f13-9fd3-b5bee3577bc5")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/clr_example_"),
  echo = FALSE)
```


```{r packages}
library("conflicted")
library(tidyverse)
library(patchwork)
conflicts_prefer(dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Zero-aware CLR example for thesis chapter 2

```{r}
set.seed(42)

# Helper functions -------------------------------------------------------

closure <- function(x) {
  x / sum(x)
}

clr <- function(x) {
  log(x) - mean(log(x))
}

# Multiplicative replacement on proportions.
# Zeros receive delta, non-zero parts are scaled proportionally.
mult_repl <- function(x, delta_factor = 0.65) {
  x <- closure(x)
  z <- x == 0
  n_zero <- sum(z)
  
  if (n_zero == 0) return(x)
  if (n_zero == length(x)) stop("Composition contains only zeros.")
  
  min_pos <- min(x[!z])
  delta <- delta_factor * min_pos
  
  # Ensure replacement does not exceed the unit sum.
  delta <- min(delta, 1 / (2 * n_zero))
  
  x_new <- x
  x_new[z] <- delta
  x_new[!z] <- x[!z] * (1 - n_zero * delta)
  
  closure(x_new)
}

clr_mult_repl <- function(x, delta_factor = 0.65) {
  clr(mult_repl(x, delta_factor = delta_factor))
}

# Offset-based CLR / emCLR-style illustration.
# A small positive offset is added to all components before CLR.
offset_clr <- function(x, delta_factor = 0.65) {
  x <- closure(x)
  
  if (all(x == 0)) stop("Composition contains only zeros.")
  
  min_pos <- min(x[x > 0])
  delta <- delta_factor * min_pos
  
  clr(x + delta)
}

# Modified CLR illustration.
# Zeros remain zero; the geometric mean is computed from positive parts only.
# Positive transformed values are shifted to be strictly positive.
mclr <- function(x, shift = 1e-6) {
  x <- closure(x)
  pos <- x > 0
  
  if (!any(pos)) stop("Composition contains only zeros.")
  
  g_pos <- exp(mean(log(x[pos])))
  out <- rep(0, length(x))
  out[pos] <- log(x[pos] / g_pos)
  
  # Positive entries are shifted so that non-zero components are > 0.
  out[pos] <- out[pos] - min(out[pos]) + shift
  
  out
}

# Simulation -------------------------------------------------------------

set.seed(123)

n <- 50
p <- 12

taxa <- paste0("Taxon_", seq_len(p))
samples <- paste0("Sample_", seq_len(n))

# Latent mean structure:
# first taxa are common, later taxa become increasingly rare
mu <- seq(3, -3, length.out = p)

# Moderate correlation structure
Sigma <- matrix(0.25, nrow = p, ncol = p)
diag(Sigma) <- 1

# Simulate latent logistic-normal compositions
Z <- MASS::mvrnorm(n = n, mu = mu, Sigma = Sigma)
P <- t(apply(exp(Z), 1, closure))
colnames(P) <- taxa
rownames(P) <- samples

# Heterogeneous and partly shallow library sizes to induce sampling zeros
lib_size <- sample(
  c(seq(300, 1000, by = 100), seq(1500, 5000, by = 500)),
  size = n,
  replace = TRUE,
  prob = c(rep(2, length(seq(300, 1000, by = 100))),
           rep(1, length(seq(1500, 5000, by = 500))))
)

# Multinomial sequencing counts
X_mat <- map2_dfr(
  .x = seq_len(n),
  .y = lib_size,
  .f = function(k, m) {
    tibble(
      sample = samples[k],
      taxon = taxa,
      count = as.numeric(rmultinom(1, size = m, prob = P[k, ]))
    )
  }
) |>
  pivot_wider(names_from = taxon, values_from = count) |>
  column_to_rownames("sample") |>
  as.matrix()

# Add controlled additional sparsity among rare taxa.
# This is artificial, but useful for the diagnostic figure.
rare_taxa <- paste0("Taxon_", 7:12)
zero_probs <- seq(0.10, 0.65, length.out = length(rare_taxa))

for (j in seq_along(rare_taxa)) {
  taxon_j <- rare_taxa[j]
  zero_id <- runif(n) < zero_probs[j]
  X_mat[zero_id, taxon_j] <- 0
}

# Remove samples that accidentally became all zero
X_mat <- X_mat[rowSums(X_mat) > 0, ]

# Recompute relative abundances and transformations ----------------------

rel_mat <- t(apply(X_mat, 1, closure))

clr_mult_mat <- t(apply(X_mat, 1, clr_mult_repl))
clr_offset_mat <- t(apply(X_mat, 1, offset_clr))
mclr_mat <- t(apply(X_mat, 1, mclr))
colnames(mclr_mat) <- colnames(clr_mult_mat)
```


```{r}
# Collect transformed data in long format -------------------------------

make_long <- function(mat, method) {
  as_tibble(mat, rownames = "sample") |>
    pivot_longer(-sample, names_to = "taxon", values_to = "value") |>
    mutate(method = method)
}

plot_dat <- bind_rows(
  make_long(rel_mat, "Relative abundance"),
  make_long(clr_mult_mat, "Multiplicative replacement + CLR"),
  make_long(clr_offset_mat, "Offset-based CLR"),
  make_long(mclr_mat, "mCLR")
)

plot_dat$method <- factor(
  plot_dat$method,
  levels = c(
    "Relative abundance",
    "Multiplicative replacement + CLR",
    "Offset-based CLR",
    "mCLR"
  )
)
```


```{r heatmap, fig.width=10, fig.height=6}
# Plot 1: Four heatmaps in one row ---------------------------------------

library(RColorBrewer)
library(scales)
library(patchwork)

# Choose direction of the palette.
# With rev(), low/negative values are blue and high/positive values are red.
spectral_cols <- rev(brewer.pal(11, "RdBu"))

# Relative abundance heatmap --------------------------------------------

plot_dat_rel <- make_long(rel_mat, "Relative abundance")

p_heat_rel <- plot_dat_rel |>
  mutate(sample = factor(sample, levels = rownames(X_mat))) |>
  ggplot(aes(x = taxon, y = sample, fill = value)) +
  geom_tile(color = "grey90", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = spectral_cols[6:11],
    name = "Relative abundance"
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Relative abundance"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )

# Transformed data -------------------------------------------------------

plot_dat_transformed <- bind_rows(
  make_long(clr_mult_mat, "Multiplicative replacement + CLR"),
  make_long(clr_offset_mat, "Offset-based CLR"),
  make_long(mclr_mat, "mCLR")
)

# Robust symmetric color limits around zero.
# This improves contrast without changing the data.
color_lim <- quantile(abs(plot_dat_transformed$value), probs = 0.98, na.rm = TRUE)

make_heat_transformed <- function(mat, title) {
  make_long(mat, title) |>
    mutate(sample = factor(sample, levels = rownames(X_mat))) |>
    ggplot(aes(x = taxon, y = sample, fill = value)) +
    geom_tile(color = "grey90", linewidth = 0.2) +
    scale_fill_gradientn(
      colours = spectral_cols,
      values = rescale(c(-color_lim, 0, color_lim)),
      limits = c(-color_lim, color_lim),
      oob = squish,
      name = "Transformed value"
    ) +
    labs(
      x = NULL,
      y = NULL,
      title = title
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
}

p_heat_mult <- make_heat_transformed(
  clr_mult_mat,
  "Multiplicative replacement + CLR"
)

p_heat_offset <- make_heat_transformed(
  clr_offset_mat,
  "Offset-based CLR"
)

p_heat_mclr <- make_heat_transformed(
  mclr_mat,
  "mCLR"
)

# Combine all four plots in one row -------------------------------------

p_heat <- p_heat_rel + p_heat_mult + p_heat_offset + p_heat_mclr +
  plot_layout(nrow = 1, widths = c(1, 1, 1, 1), guides = "collect") &
  theme(
    legend.position = "bottom"
  )

p_heat <- p_heat +
  plot_annotation(
    title = "Effect of zero-aware transformations on the data matrix",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)
    )
  )

p_heat
```

```{r}
# Plot 2: Ratio preservation among originally non-zero components --------

# Choose two taxa for which we inspect the original non-zero ratio.
taxon_a <- "Taxon_2"
taxon_b <- "Taxon_3"

ratio_dat <- tibble(
  sample = rownames(X_mat),
  count_a = X_mat[, taxon_a],
  count_b = X_mat[, taxon_b]
) |>
  filter(count_a > 0, count_b > 0) |>
  mutate(
    original_log_ratio = log(count_a / count_b),
    mult_clr_diff = clr_mult_mat[sample, taxon_a] - clr_mult_mat[sample, taxon_b],
    offset_clr_diff = clr_offset_mat[sample, taxon_a] - clr_offset_mat[sample, taxon_b],
    mclr_diff = mclr_mat[sample, taxon_a] - mclr_mat[sample, taxon_b]
  ) |>
  pivot_longer(
    cols = c(mult_clr_diff, offset_clr_diff, mclr_diff),
    names_to = "method",
    values_to = "transformed_difference"
  ) |>
  mutate(
    method = recode(
      method,
      mult_clr_diff = "Multiplicative replacement + CLR",
      offset_clr_diff = "Offset-based CLR",
      mclr_diff = "mCLR"
    )
  )

p_ratio <- ratio_dat |>
  ggplot(aes(x = original_log_ratio, y = transformed_difference)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~ method, nrow = 1) +
  coord_equal() +
  labs(
    x = paste0("Original log-ratio log(", taxon_a, "/", taxon_b, ")"),
    y = "Difference between transformed coordinates",
    title = "Preservation of ratios among originally non-zero taxa"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

p_ratio
```

```{r}
# Plot 3: Effect on pairwise sample distances ----------------------------

dist_mult <- dist(clr_mult_mat)
dist_offset <- dist(clr_offset_mat)
dist_mclr <- dist(mclr_mat)

dist_dat <- tibble(
  mult_repl_clr = as.vector(dist_mult),
  offset_clr = as.vector(dist_offset),
  mclr = as.vector(dist_mclr)
)

p_dist_offset <- ggplot(dist_dat, aes(x = mult_repl_clr, y = offset_clr)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(alpha = 0.6) +
  labs(
    x = "Distance after multiplicative replacement + CLR",
    y = "Distance after offset-based CLR",
    title = "Sample distances: offset-based CLR"
  ) +
  theme_minimal(base_size = 11)

p_dist_mclr <- ggplot(dist_dat, aes(x = mult_repl_clr, y = mclr)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(alpha = 0.6) +
  labs(
    x = "Distance after multiplicative replacement + CLR",
    y = "Distance after mCLR",
    title = "Sample distances: mCLR"
  ) +
  theme_minimal(base_size = 11)

p_dist <- p_dist_offset + p_dist_mclr

p_dist
```

```{r}
# Distribution plots across samples --------------------------------------

library(tidyverse)
library(RColorBrewer)
library(patchwork)

# Select taxa with different zero frequencies ----------------------------

zero_summary <- tibble(
  taxon = colnames(X_mat),
  zero_prop = colMeans(X_mat == 0),
  mean_rel_abundance = colMeans(rel_mat)
) |>
  arrange(desc(zero_prop), desc(mean_rel_abundance))

zero_summary

# Automatically choose:
# - two taxa with many zeros
# - two taxa with few or no zeros
taxa_sparse <- zero_summary |>
  arrange(desc(zero_prop), desc(mean_rel_abundance)) |>
  slice_head(n = 2) |>
  pull(taxon)

taxa_common <- zero_summary |>
  arrange(zero_prop, desc(mean_rel_abundance)) |>
  slice_head(n = 2) |>
  pull(taxon)

taxa_selected <- sort(unique(c(taxa_sparse, taxa_common)))

taxa_selected
```

```{r}
# Build long data set ----------------------------------------------------

dist_dat <- bind_rows(
  make_long(rel_mat, "Relative abundance"),
  make_long(clr_mult_mat, "Multiplicative replacement + CLR"),
  make_long(clr_offset_mat, "Offset-based CLR"),
  make_long(mclr_mat, "mCLR")
) |>
  filter(taxon %in% taxa_selected) |>
  mutate(
    method = factor(
      method,
      levels = c(
        "Relative abundance",
        "Multiplicative replacement + CLR",
        "Offset-based CLR",
        "mCLR"
      )
    ),
    taxon = factor(taxon, levels = taxa_selected)
  )

# Add zero frequency labels for facet strips -----------------------------

taxon_labels <- zero_summary |>
  filter(taxon %in% taxa_selected) |>
  mutate(
    label = paste0(taxon, "\nzeros: ", round(100 * zero_prop), "%")
  ) |>
  select(taxon, label) |>
  deframe()

# Shared plotting settings -----------------------------------------------

spectral_cols <- rev(brewer.pal(11, "Spectral"))

make_distribution_plot <- function(data, method_name, x_label) {
  data |>
    filter(method == method_name) |>
    ggplot(aes(x = value, fill = taxon)) +
    geom_density(alpha = 0.55, linewidth = 0.4, adjust = 1.1) +
    facet_wrap(
      ~ taxon,
      scales = "free",
      labeller = labeller(taxon = taxon_labels)
    ) +
    scale_fill_manual(
      values = spectral_cols[seq_len(length(taxa_selected))],
      guide = "none"
    ) +
    labs(
      x = x_label,
      y = "Density",
      title = method_name
    ) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

p_dist_rel <- make_distribution_plot(
  dist_dat,
  "Relative abundance",
  "Relative abundance"
)

p_dist_mult <- make_distribution_plot(
  dist_dat,
  "Multiplicative replacement + CLR",
  "Transformed value"
)

p_dist_offset <- make_distribution_plot(
  dist_dat,
  "Offset-based CLR",
  "Transformed value"
)

p_dist_mclr <- make_distribution_plot(
  dist_dat,
  "mCLR",
  "Transformed value"
)
```


```{r distributions, fig.width=10, fig.height=5}
# Combine four distribution plots in one row -----------------------------

p_distribution <- p_dist_rel + p_dist_mult + p_dist_offset + p_dist_mclr +
  plot_layout(nrow = 1)

p_distribution <- p_distribution +
  plot_annotation(
    title = "Across-sample distributions of selected taxa after zero-aware transformations",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)
    )
  )

p_distribution
```

```{r violin, fig.width=10, fig.height=5}
# Alternative: violin + points -------------------------------------------

make_violin_plot <- function(data, method_name, y_label) {
  data |>
    filter(method == method_name) |>
    ggplot(aes(x = taxon, y = value, fill = taxon)) +
    geom_violin(
      alpha = 0.65,
      linewidth = 0.3,
      trim = FALSE
    ) +
    # geom_boxplot(
    #   width = 0.15,
    #   outlier.shape = NA,
    #   alpha = 0.8,
    #   linewidth = 0.3
    # ) +
    geom_jitter(
      width = 0.08,
      size = 0.9,
      alpha = 0.7
    ) +
    scale_x_discrete(labels = taxon_labels) +
    scale_fill_manual(
      values = spectral_cols[seq_len(length(taxa_selected))],
      guide = "none"
    ) +
    labs(
      x = NULL,
      y = y_label,
      title = method_name
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

p_vio_rel <- make_violin_plot(
  dist_dat,
  "Relative abundance",
  "Relative abundance"
)

p_vio_mult <- make_violin_plot(
  dist_dat,
  "Multiplicative replacement + CLR",
  "Transformed value"
)

p_vio_offset <- make_violin_plot(
  dist_dat,
  "Offset-based CLR",
  "Transformed value"
)

p_vio_mclr <- make_violin_plot(
  dist_dat,
  "mCLR",
  "Transformed value"
)

p_distribution_violin <- p_vio_rel + p_vio_mult + p_vio_offset + p_vio_mclr +
  plot_layout(nrow = 1) +
  plot_annotation(
    title = "Across-sample distributions of selected taxa after zero-aware transformations",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)
    )
  )

p_distribution_violin
```

```{r}
# Build long data set ----------------------------------------------------

hist_dat <- bind_rows(
  make_long(rel_mat, "Relative abundance"),
  make_long(clr_mult_mat, "Multiplicative replacement + CLR"),
  make_long(clr_offset_mat, "Offset-based CLR"),
  make_long(mclr_mat, "mCLR")
) |>
  filter(taxon %in% taxa_selected) |>
  mutate(
    method = factor(
      method,
      levels = c(
        "Relative abundance",
        "Multiplicative replacement + CLR",
        "Offset-based CLR",
        "mCLR"
      )
    ),
    taxon = factor(taxon, levels = taxa_selected)
  )

# Taxon labels with zero proportions ------------------------------------

taxon_labels <- zero_summary |>
  filter(taxon %in% taxa_selected) |>
  mutate(
    label = paste0(taxon, "\nzeros: ", round(100 * zero_prop), "%")
  ) |>
  select(taxon, label) |>
  deframe()

spectral_cols <- rev(brewer.pal(11, "Spectral"))

# Function for one transformation ---------------------------------------

make_histogram_plot <- function(data, method_name, x_label) {
  data_method <- data |>
    filter(method == method_name)
  
  ggplot(data_method, aes(x = value, fill = taxon)) +
    geom_histogram(
      bins = 12,
      color = "white",
      linewidth = 0.25,
      alpha = 0.85
    ) +
    facet_wrap(
      ~ taxon,
      scales = "free_x",
      labeller = labeller(taxon = taxon_labels)
    ) +
    scale_fill_manual(
      values = spectral_cols[seq_len(length(taxa_selected))],
      guide = "none"
    ) +
    labs(
      x = x_label,
      y = "Number of samples",
      title = method_name
    ) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

# One histogram panel per transformation --------------------------------

p_hist_rel <- make_histogram_plot(
  hist_dat,
  "Relative abundance",
  "Relative abundance"
)

p_hist_mult <- make_histogram_plot(
  hist_dat,
  "Multiplicative replacement + CLR",
  "Transformed value"
)

p_hist_offset <- make_histogram_plot(
  hist_dat,
  "Offset-based CLR",
  "Transformed value"
)

p_hist_mclr <- make_histogram_plot(
  hist_dat,
  "mCLR",
  "Transformed value"
)

# Combine in one row -----------------------------------------------------

p_histograms <- p_hist_rel + p_hist_mult + p_hist_offset + p_hist_mclr +
  plot_layout(nrow = 1) +
  plot_annotation(
    title = "Across-sample distributions of selected taxa after zero-aware transformations",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)
    )
  )

p_histograms
```

```{r histograms, fig.width=10, fig.height=5}
p_histograms_grid <- hist_dat |>
  ggplot(aes(x = value, fill = taxon)) +
  geom_histogram(
    bins = 12,
    color = "white",
    linewidth = 0.25,
    alpha = 0.85
  ) +
  facet_grid(
    method ~ taxon,
    scales = "free_x",
    labeller = labeller(taxon = taxon_labels)
  ) +
  scale_fill_manual(
    values = spectral_cols[seq_len(length(taxa_selected))],
    guide = "none"
  ) +
  labs(
    x = "Value",
    y = "Number of samples",
    title = "Across-sample distributions of selected taxa after zero-aware transformations"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14)
  )

p_histograms_grid
```
## Diagnostic 1: Log-ratio preservation

Here, we plot on they axis for each pair of non-zero counts:

$$
\Delta_{\text{log-ratio}}
=
\underbrace{
\left(T_i(x_k)-T_j(x_k)\right)
}_{\text{log-ratio after transformation}}
-
\underbrace{
\log\left(\frac{x_{ki}}{x_{kj}}\right)
}_{\text{original log-ratio}}.
$$

```{r}
compute_logratio_error <- function(X, transformed_mat, method_name) {
  taxa <- colnames(X)
  
  purrr::map_dfr(seq_len(nrow(X)), function(k) {
    x <- X[k, ]
    pos_taxa <- taxa[x > 0]
    
    if (length(pos_taxa) < 2) return(NULL)
    
    pairs <- combn(pos_taxa, 2, simplify = FALSE)
    
    purrr::map_dfr(pairs, function(pair) {
      i <- pair[1]
      j <- pair[2]
      
      original_lr <- log(x[i] / x[j])
      transformed_lr <- transformed_mat[k, i] - transformed_mat[k, j]
      
      tibble(
        sample = rownames(X)[k],
        taxon_i = i,
        taxon_j = j,
        method = method_name,
        original_log_ratio = original_lr,
        transformed_log_ratio = transformed_lr,
        log_ratio_error = transformed_lr - original_lr
      )
    })
  })
}

logratio_error_dat <- bind_rows(
  compute_logratio_error(X_mat, clr_mult_mat, "Multiplicative replacement + CLR"),
  compute_logratio_error(X_mat, clr_offset_mat, "Offset-based CLR"),
  compute_logratio_error(X_mat, mclr_mat, "mCLR")
) |>
  mutate(
    method = factor(
      method,
      levels = c(
        "Multiplicative replacement + CLR",
        "Offset-based CLR",
        "mCLR"
      )
    )
  )

logratio_error_dat <- logratio_error_dat |>
  mutate(
    method_short = recode(
      as.character(method),
      "Multiplicative replacement + CLR" = "multRepl + CLR",
      "Offset-based CLR" = "Offset + CLR",
      "mCLR" = "mCLR"
    ),
    method_short = factor(
      method_short,
      levels = c("multRepl + CLR", "Offset + CLR", "mCLR")
    )
  )

p_logratio_error <- ggplot(logratio_error_dat, aes(x = method_short, y = log_ratio_error)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(width = 0.45, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.12, alpha = 0.25, size = 0.8) +
  labs(
    x = NULL,
    y = expression(Delta["log-ratio"]),
    title = "Preservation of log-ratios among originally non-zero taxa",
    subtitle = "Each point represents one sample-specific taxon pair with positive counts for both taxa"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )

p_logratio_error
```

This diagnostic is particularly relevant for association and proportionality analyses, where pairwise log-ratio information and its variation across samples are central to the inferential target. For CLR-based association measures and proportionality analyses, multiplicative replacement followed by CLR is therefore used as the primary strategy in this thesis. The mCLR transformation also preserves positive-positive log-ratios, but it retains zeros as exact zeros and induces a different sample geometry; it is therefore used only when it is part of a method-specific model or transformation.

## Diagnostic 2: Distance distortion and zero burden

```{r}
pairwise_zero_burden <- function(X) {
  sample_pairs <- combn(rownames(X), 2, simplify = FALSE)
  p <- ncol(X)
  
  purrr::map_dfr(sample_pairs, function(pair) {
    s1 <- pair[1]
    s2 <- pair[2]
    
    tibble(
      sample_1 = s1,
      sample_2 = s2,
      zero_burden = (sum(X[s1, ] == 0) + sum(X[s2, ] == 0)) / (2 * p)
    )
  })
}

dist_to_tibble <- function(D, name) {
  D_mat <- as.matrix(D)
  sample_pairs <- combn(rownames(D_mat), 2, simplify = FALSE)
  
  purrr::map_dfr(sample_pairs, function(pair) {
    tibble(
      sample_1 = pair[1],
      sample_2 = pair[2],
      distance = D_mat[pair[1], pair[2]],
      method = name
    )
  })
}

dist_dat_long <- bind_rows(
  dist_to_tibble(dist(clr_mult_mat), "Multiplicative replacement + CLR"),
  dist_to_tibble(dist(clr_offset_mat), "Offset-based CLR"),
  dist_to_tibble(dist(mclr_mat), "mCLR")
)

dist_ref <- dist_dat_long |>
  filter(method == "Multiplicative replacement + CLR") |>
  select(sample_1, sample_2, distance_ref = distance)

zero_burden_dat <- pairwise_zero_burden(X_mat)

dist_distortion_dat <- dist_dat_long |>
  filter(method != "Multiplicative replacement + CLR") |>
  left_join(dist_ref, by = c("sample_1", "sample_2")) |>
  left_join(zero_burden_dat, by = c("sample_1", "sample_2")) |>
  mutate(
    distance_difference = distance / distance_ref,
    method = factor(
      method,
      levels = c("Offset-based CLR", "mCLR")
    )
  )

p_distance_distortion <- ggplot(
  dist_distortion_dat,
  aes(x = zero_burden, y = distance_difference)
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.45, size = 1.2) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ method, nrow = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Proportion of zero entries in each sample pair",
    y = "Difference in pairwise distance relative to multReplCLR",
    title = "Distance distortion depends on sparsity"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

p_distance_distortion
```

For offset-based CLR, the distance differences are small and mostly slightly negative. This means that offset-based CLR gives a sample geometry that is broadly similar to multiplicative replacement + CLR, but tends to compress distances a little. The effect becomes somewhat stronger for highly sparse sample pairs. That makes sense because adding a positive offset to all components reduces the extremeness of zeros, but also changes the ratios among originally non-zero taxa.

For mCLR, the distance differences are much larger and mostly positive. This means that mCLR often makes sample pairs substantially farther apart than multiplicative replacement + CLR. The effect is clearly stronger when sample pairs contain more zeros. This is also expected: mCLR leaves zeros at exactly zero and transforms only the positive components relative to their positive-part geometric mean, so it induces a different geometry from standard CLR-based Aitchison geometry.

This plot supports the argument that zero handling is not neutral for distance-based analyses. If the downstream method uses Euclidean distances between transformed compositions, as in Aitchison distance, ordination, PERMANOVA, or CLR-based multivariate regression, then the zero-aware transformation directly changes the geometry on which the analysis is based.

## Diagnostic 3: Marginal distributions of one common and one sparse taxon

```{r}
taxon_common <- zero_summary |>
  arrange(zero_prop, desc(mean_rel_abundance)) |>
  slice(1) |>
  pull(taxon)

taxon_sparse <- zero_summary |>
  arrange(desc(zero_prop), desc(mean_rel_abundance)) |>
  slice(1) |>
  pull(taxon)

taxa_selected_simple <- c(taxon_common, taxon_sparse)

hist_dat_simple <- bind_rows(
  make_long(rel_mat, "Relative abundance"),
  make_long(clr_mult_mat, "Multiplicative replacement + CLR"),
  make_long(clr_offset_mat, "Offset-based CLR"),
  make_long(mclr_mat, "mCLR")
) |>
  filter(taxon %in% taxa_selected_simple) |>
  mutate(
    method = factor(
      method,
      levels = c(
        "Relative abundance",
        "Multiplicative replacement + CLR",
        "Offset-based CLR",
        "mCLR"
      )
    ),
    taxon = factor(taxon, levels = taxa_selected_simple)
  )

taxon_labels_simple <- zero_summary |>
  filter(taxon %in% taxa_selected_simple) |>
  mutate(label = paste0(taxon, "\nzeros: ", round(100 * zero_prop), "%")) |>
  select(taxon, label) |>
  deframe()

p_marginal_hist <- hist_dat_simple |>
  ggplot(aes(x = value)) +
  geom_histogram(
    bins = 12,
    color = "white",
    fill = "grey45",
    linewidth = 0.25
  ) +
  facet_grid(
    method ~ taxon,
    scales = "free_x",
    labeller = labeller(taxon = taxon_labels_simple)
  ) +
  labs(
    x = "Value",
    y = "Number of samples",
    title = "Marginal distributions of a common and a sparse taxon"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_marginal_hist
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
