---
title: "Generate exponential data for permutation p-value simulations (Mann–Whitney U)"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "01_data_gen" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "a2428bff-9ba8-452e-89b9-af01268d0f2f")
```

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkgs}
library(conflicted)
library(dplyr)
library(tidyr)
library(tibble)
library(furrr)
library(progressr)

conflicts_prefer(dplyr::filter, dplyr::select)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

# Parameter settings

```{r, echo=TRUE}
# Which part to generate this run: "accuracy" or "error"
part <- "accuracy"

# ---------------- Accuracy ----------------
acc_n_fix      <- 100                    # per-group size
acc_n_vec      <- c(25, 50, 75, 100, 250, 500, 750, 1000)
acc_d_fix      <- 1.0                    # delta on the *mean* scale (mean2 = 1 + d)
acc_d_vec      <- c(0.5, 1.0, 1.5, 2.0)
acc_n_perm_fix <- 1000
acc_n_perm_vec <- c(500, 1000, 5000, 10000)
acc_n_test     <- 1
acc_n_rep      <- 1000

# ---------------- Error rates ----------------
err_n_fix      <- 100
err_n_vec      <- c(25, 50, 75, 100, 250, 500, 750, 1000)
err_d_fix      <- 1.0
err_d_vec      <- c(0.5, 1.0, 1.5, 2.0)
err_n_perm_fix <- 1000
err_n_perm_vec <- c(500, 1000, 5000, 10000)
err_pi1_fix    <- 0.05
err_pi1_vec    <- c(0.00, 0.01, 0.05, 0.10)  # non-null fractions
err_n_test     <- 1000
err_n_rep      <- 100

cores <- 8
```

This script generates **exponential** data for two groups:

* Group 1: Exp(rate = 1) ⇒ mean = 1
* Group 2: Exp(rate = 1/(1 + δ)) ⇒ mean = 1 + δ

We use the **Mann–Whitney U** statistic (via ranks; U = W − n(n+1)/2) and store:

* `1:n_perm` = permutation U-statistics,
* `uobs` = observed U-statistic,
* `p_wilcox` = asymptotic Wilcoxon p-value (two-sided; `exact = FALSE, correct = FALSE`).

# Helpers

```{r helpers, echo=TRUE}
# One two-sample Wilcoxon/Mann–Whitney (U) test on exponentials:
# returns observed U, vector of permuted U's (length n_perm), and classical p
get_one_test_U <- function(n, delta, n_perm = 1000) {
  # Data: Exp(rate1 = 1), Exp(rate2 = 1/(1+delta))  ⇒ means 1 vs 1+delta
  x1 <- rexp(n, rate = 1)
  x2 <- rexp(n, rate = 1 / (1 + delta))

  # Pooled ranks
  x  <- c(x1, x2)
  nn <- length(x)
  r  <- rank(x, ties.method = "average")

  # Wilcoxon rank-sum for group 1, then shift to U
  W_obs <- sum(r[seq_len(n)])
  shift <- n * (n + 1) / 2
  U_obs <- W_obs - shift

  # Permutation: sample group-1 indices; compute W_perm and U_perm
  # (matrix sampling is faster than a loop for moderate sizes)
  idx_mat <- replicate(n_perm, sample.int(nn, n, replace = FALSE))
  W_perm  <- colSums(matrix(r[idx_mat], nrow = n, ncol = n_perm))
  U_perm  <- W_perm - shift

  # Classical Wilcoxon p-value (asymptotic, no cont. correction)
  p_wcx <- stats::wilcox.test(x1, x2,
                              alternative = "two.sided",
                              exact = FALSE, correct = FALSE)$p.value

  list(U_obs = U_obs, U_perm = U_perm, p_wilcox = p_wcx)
}

# Build matrix [n_test x (n_perm+2)] with columns 1:n_perm, "uobs", "p_wilcox"
# If delta_vec has length n_test, it can mix H0/H1; otherwise a scalar delta for all tests.
make_panel_matrix_U <- function(n, delta_vec, n_perm = 1000) {
  if (length(delta_vec) == 1L) delta_vec <- rep(delta_vec, length.out = 1L)
  n_test <- length(delta_vec)

  M <- matrix(NA_real_, nrow = n_test, ncol = n_perm + 2)
  colnames(M) <- c(as.character(seq_len(n_perm)), "uobs", "p_wilcox")

  for (t in seq_len(n_test)) {
    rs <- get_one_test_U(n = n, delta = delta_vec[t], n_perm = n_perm)
    M[t, 1:n_perm]     <- rs$U_perm
    M[t, "uobs"]       <- rs$U_obs
    M[t, "p_wilcox"]   <- rs$p_wilcox
  }
  M
}

# Reproducible seed streams per (setting, rep) using L'Ecuyer-CMRG
make_seed_stream <- function(n_streams, master_seed) {
  set.seed(master_seed, kind = "L'Ecuyer-CMRG")
  s <- .Random.seed
  streams <- vector("list", n_streams)
  for (i in seq_len(n_streams)) {
    s <- parallel::nextRNGStream(s)
    streams[[i]] <- s
  }
  streams
}
```

## Accuracy

We vary **n** at **d = 1**, vary **d** at **n = 100**, and vary **n_perm** at (n=100, d=1).

```{r}
acc_grid_vary_n <- tidyr::crossing(
  n      = acc_n_vec,
  d      = acc_d_fix,
  n_perm = acc_n_perm_fix
)
acc_grid_vary_d <- tidyr::crossing(
  n      = acc_n_fix,
  d      = acc_d_vec,
  n_perm = acc_n_perm_fix
)
acc_grid_vary_nperm <- tibble(
  n      = acc_n_fix,
  d      = acc_d_fix,
  n_perm = acc_n_perm_vec
)

acc_grid <- bind_rows(acc_grid_vary_n, acc_grid_vary_d, acc_grid_vary_nperm) %>%
  distinct() %>%
  mutate(n_test = acc_n_test,
         n_rep  = acc_n_rep)
```

```{r accuracy, eval=FALSE}
out_path <- path_target("sim_accuracy_data")
dir.create(out_path, showWarnings = FALSE, recursive = TRUE)

handlers("txtprogressbar")
plan(multisession, workers = cores)

with_progress({
  p <- progressor(steps = nrow(acc_grid))

  invisible(furrr::future_walk(
    split(acc_grid, seq_len(nrow(acc_grid))),
    function(row) {
      row <- tibble::as_tibble(row)
      n      <- row$n[[1]]
      d      <- row$d[[1]]
      n_perm <- row$n_perm[[1]]
      n_rep  <- row$n_rep[[1]]

      out_file <- file.path(
        out_path,
        sprintf("sim_acc_n%d_d%.1f_nperm%d.rds", n, d, n_perm)
      )
      if (file.exists(out_file)) { p(); return(invisible()) }

      streams  <- make_seed_stream(n_streams = n_rep, master_seed = 987654)
      res_list <- vector("list", n_rep)

      # Accuracy: 1 test per rep; if d>0 it's H1 (shift in means), if d==0 it's H0
      for (r in seq_len(n_rep)) {
        .Random.seed <<- streams[[r]]
        mat <- make_panel_matrix_U(n = n, delta_vec = d, n_perm = n_perm)
        res_list[[r]] <- mat
      }

      saveRDS(res_list, file = out_file)
      p()
    },
    .options = furrr::furrr_options(seed = TRUE)
  ))
})

plan(sequential)
```

## Error rates

Here, we vary **n** at **d = 1**; vary **d** at **n = 100**; plus focused **(n=100, d=1)** with multiple **n_perm** and **pi1**.

```{r}
err_grid_vary_n <- tidyr::crossing(
  n      = err_n_vec,
  d      = err_d_fix,
  n_perm = err_n_perm_fix,
  pi1    = err_pi1_fix
)

err_grid_vary_d <- tidyr::crossing(
  n      = err_n_fix,
  d      = err_d_vec,
  n_perm = err_n_perm_fix,
  pi1    = err_pi1_fix
)

err_grid_vary_nperm <- tibble(
  n      = err_n_fix,
  d      = err_d_fix,
  n_perm = err_n_perm_vec,
  pi1    = err_pi1_fix
)

err_grid_vary_pi1 <- tibble(
  n      = err_n_fix,
  d      = err_d_fix,
  n_perm = err_n_perm_fix,
  pi1    = err_pi1_vec
)

err_grid <- bind_rows(err_grid_vary_n, err_grid_vary_d, 
                      err_grid_vary_nperm, err_grid_vary_pi1) %>%
  distinct() %>%
  mutate(n_test = err_n_test,
         n_rep  = err_n_rep)
```

```{r error, eval=FALSE}
out_path <- path_target("sim_error_data")
dir.create(out_path, showWarnings = FALSE, recursive = TRUE)

handlers("txtprogressbar")
plan(multisession, workers = cores)

with_progress({
  p <- progressor(steps = nrow(err_grid))

  invisible(furrr::future_walk(
    split(err_grid, seq_len(nrow(err_grid))),
    function(row) {
      row <- tibble::as_tibble(row)
      n      <- row$n[[1]]
      d      <- row$d[[1]]
      n_perm <- row$n_perm[[1]]
      pi1    <- row$pi1[[1]]
      n_test <- row$n_test[[1]]
      n_rep  <- row$n_rep[[1]]

      out_file <- file.path(
        out_path,
        sprintf("sim_err_n%d_d%.1f_nperm%d_pi1%.2f.rds", n, d, n_perm, pi1)
      )
      if (file.exists(out_file)) { p(); return(invisible()) }

      streams  <- make_seed_stream(n_streams = n_rep, master_seed = 987654)
      res_list <- vector("list", n_rep)

      n_alt    <- round(n_test * pi1)
      # Non-nulls have delta = d (mean2 = 1 + d); nulls have delta = 0
      delta_vec <- c(rep(d, n_alt), rep(0, n_test - n_alt))

      for (r in seq_len(n_rep)) {
        .Random.seed <<- streams[[r]]
        mat <- make_panel_matrix_U(n = n, delta_vec = delta_vec, n_perm = n_perm)
        res_list[[r]] <- mat
      }

      saveRDS(res_list, file = out_file)
      p()
    },
    .options = furrr::furrr_options(seed = TRUE)
  ))
})

plan(sequential)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
