---
title: "Wilcoxon (Mann–Whitney U) on exponential data - Test the SLLS epsilon rule"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "01_test_eps_rule" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "efcad4fd-ecce-4206-b899-b1726dff38ea")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/expl_wilcox_eps_rule_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

In this script we evaluate our **standardized lifted log-saturation (SLLS)** ε-rule with the **Wilcoxon rank-sum / Mann–Whitney test**, using the **(U) statistic** throughout.

For equal group sizes (n), (U) counts pairwise “wins” of group 1 over group 2 and takes values in ([0, n^2]) with (E_0[U]=n^2/2). It is related to the Wilcoxon rank-sum (W) by a constant shift:

$$
U = W - \frac{n(n+1)}{2}.
$$

We fit a GPD to the **right tail** of the permutation distribution of (U) and compare unconstrained vs. constrained (SLLS) fits.

Unlike the two-sample $t$-test setting, the Wilcoxon (U) statistic is **nonparametric** and has a different distributional behavior under the null, which provides a useful test case for evaluating the robustness of our approach. 

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)

library(ggplot2)
library(patchwork)
library(pheatmap)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(future)
library(future.apply)
library(kableExtra)
library(furrr)
library(progressr)
library(patchwork)
library(digest) # For seed stream

# For permApprox:
library(fdrtool)
library(fitdistrplus)
library(goftest)
library(limma)

conflicts_prefer(dplyr::select, dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r load_permApprox}
# Load permApprox functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(permApprox_functions, source))
```

## Data Simulation

```{r}
cores <- 8

n_tests <- 1000
n_per_group_vec <- c(50, 100, 250, 500, 750, 1000)
B <- 1000

# Effect sizes for Exp means: mean1 = 1, mean2 = 1 + delta
effect_blocks <- c(0, 0.5, 1, 1.5, 2)
delta_vec <- rep(effect_blocks, each = n_tests / length(effect_blocks))

```

We simulate two groups from **exponential distributions**:

* Group 1: mean (=1) (rate (=1))
* Group 2: mean (=1+\delta) (rate (=1/(1+\delta)))
* Sample sizes per group: `r n_per_group_vec`
* Number of tests: `r n_tests`
* Effect sizes: `r paste(effect_blocks, collapse = " ")`
* Tests per effect size: `r paste0("d=", effect_blocks, ": ",table(delta_vec), collapse = ", ")`

We apply the **Wilcoxon rank-sum test** (Mann–Whitney $U$ test) to compare the two exponential samples.

* Statistic: **Mann–Whitney (U)** (converted from rank-sum (W))
* P-values: classical Wilcoxon (`exact = FALSE, correct = FALSE`) for reference; permutation/GPD for tail modeling

## Simulate data

```{r}
### Helpers

# Stable 32-bit integer from arbitrary keys
seed_from <- function(...) {
  hx <- digest(paste(..., collapse="|"), algo = "xxhash32")
  s  <- as.integer(strtoi(hx, base = 16L))
  if (is.na(s) || s == 0L) 1L else s
}

# One test -> (U_obs, U_perm[B], classical p, empirical p)
.sim_one_U <- function(n_per_group, B, delta) {
  # Data (Exp means 1 vs 1+delta)
  x1 <- rexp(n_per_group, rate = 1)
  x2 <- rexp(n_per_group, rate = 1 / (1 + as.numeric(delta)))

  # Pooled ranks and W (sum of ranks of sample 1)
  x_comb <- c(x1, x2)
  n_comb <- length(x_comb)
  r_all  <- rank(x_comb, ties.method = "average")
  W_obs  <- sum(r_all[seq_len(n_per_group)])

  # Convert to U: U = W - n(n+1)/2
  shift <- n_per_group * (n_per_group + 1) / 2
  U_obs <- W_obs - shift

  # Permutations: sample “group 1” indices; sum ranks -> W_perm -> U_perm
  perm_idx_mat <- replicate(B, sample.int(n_comb, n_per_group))
  W_perm <- colSums(matrix(r_all[perm_idx_mat], nrow = n_per_group, ncol = B))
  U_perm <- W_perm - shift

  # Classical Wilcoxon p (asymptotic, no continuity corr)
  p_cl <- wilcox.test(x1, x2, alternative = "two.sided",
                      exact = FALSE, correct = FALSE)$p.value

  # Two-sided empirical p from U
  greater <- sum(U_perm >= U_obs)
  lesser  <- sum(U_perm <= U_obs)
  p_emp   <- (2 * min(greater, lesser) + 1) / (B + 1)

  list(U_obs = U_obs, U_perm = U_perm, p_cl = p_cl, p_emp = p_emp)
}


simulate_once_wilcoxU <- function(n_per_group, B, n_tests, delta_vec, path_target) {
  fn_obs  <- path_target(paste0("obs_U_stats_n", n_per_group, "_B", B, ".rds"))
  fn_perm <- path_target(paste0("perm_U_stats_n", n_per_group, "_B", B, ".rds"))
  fn_pcl  <- path_target(paste0("pval_wilcox_n", n_per_group, "_B", B, ".rds"))
  fn_pemp <- path_target(paste0("pval_empwilcox_n", n_per_group, "_B", B, ".rds"))

  if (all(file.exists(fn_obs, fn_perm, fn_pcl, fn_pemp))) return(invisible(NULL))

  message("Simulating (Wilcoxon/Mann–Whitney U, Exponential) for n=", n_per_group, ", B=", B, " ...")

  # Parallel plan + deterministic substreams per g
  master_seed <- seed_from("wilcox-U-exp", n_per_group, B, n_tests)
  oplan <- future::plan(multisession, workers = max(1, future::availableCores() - 1))
  on.exit(future::plan(oplan), add = TRUE)

  obs_stats  <- numeric(n_tests)
  pval_wcx   <- numeric(n_tests)
  pval_emp   <- numeric(n_tests)
  perm_stats <- matrix(NA_real_, nrow = B, ncol = n_tests)

  with_progress({
    p <- progressr::progressor(along = seq_len(n_tests))

    res_list <- future_lapply(
      seq_len(n_tests),
      function(g, p) {
        p()
        delta_g <- as.numeric(as.character(delta_vec[g]))
        .sim_one_U(n_per_group, B, delta_g)
      },
      p = p,
      future.seed = master_seed
    )

    for (g in seq_len(n_tests)) {
      obs_stats[g]    <- res_list[[g]]$U_obs
      pval_wcx[g]     <- res_list[[g]]$p_cl
      pval_emp[g]     <- res_list[[g]]$p_emp
      perm_stats[, g] <- res_list[[g]]$U_perm
    }
  })

  saveRDS(obs_stats,  fn_obs)
  saveRDS(perm_stats, fn_perm)
  saveRDS(pval_wcx,   fn_pcl)
  saveRDS(pval_emp,   fn_pemp)
  invisible(NULL)
}

```


```{r simulate_data}
# Run all (n, B) scenarios
scenarios <- tidyr::expand_grid(n_per_group = n_per_group_vec, B = B)
scenarios %>% purrr::pwalk(~ simulate_once_wilcoxU(..1, ..2, n_tests, delta_vec, path_target))

# factor for later grouping
delta_vec <- factor(delta_vec, levels = effect_blocks)
```

## Histograms

```{r histograms}
n_per_group <- 100
B <- 1000

perm_stats <- readRDS(path_target(paste0("perm_U_stats_n", n_per_group, "_B", B, ".rds")))

# Choose up to the first six columns
cols <- 1:min(6, ncol(perm_stats))

# Prepare long format for faceting
df <- as.data.frame(perm_stats[, cols, drop = FALSE])
colnames(df) <- paste0("Test ", cols)

df_long <- df %>%
  mutate(.row = row_number()) %>%
  pivot_longer(-.row, names_to = "stat", values_to = "value")

ggplot(df_long, aes(x = value)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ stat, scales = "free_x") +
  labs(
    title = "Permutation statistics: Histograms of first six tests (n = 100)",
    x = expression(U[perm]), y = "Count"
  ) +
  theme_minimal()

```


## Plot function


```{r}
# ---------- helper: step = 20→10→5 up to 100, then 50 beyond ----------
.compute_log_breaks <- function(ymin, ymax, min_ticks = 5, breaks_incl = NULL) {
  
  ymin <- max(ymin, .Machine$double.xmin)
  ymax <- min(max(ymin, ymax), 1)
  
  e_min <- floor(log10(ymin))
  e_max <- min(0, ceiling(log10(ymax)))  # p-values ≤ 1
  
  # choose step candidates depending on how deep the tail goes
  if (abs(e_min) <= 100) {
    try_steps <- c(20, 10, 5)
  } else {
    try_steps <- 50
  }
  
  brks <- NULL
  
  for (s in try_steps) {
    start_exp <- -s * ceiling(abs(e_min) / s)
    exps <- seq(from = start_exp, to = -s, by = s)
    exps <- exps[exps >= e_min & exps <= e_max]
    cand <- 10^exps
    cand <- cand[cand >= ymin & cand <= ymax]
    
    if (!is.null(breaks_incl)) {
      add <- breaks_incl[breaks_incl > 0 & breaks_incl <= 1]
      cand <- sort(unique(c(cand, add)))
      cand <- cand[cand >= ymin & cand <= ymax]
    }
    
    if (length(cand) >= min_ticks) { brks <- cand; break }
    brks <- cand  # fallback if all fail
  }
  
  labs <- paste0("1e", 
                 formatC(log10(brks), format = "f",
                         digits = if (any(abs(log10(brks) - 
                                              round(log10(brks))) > 1e-9)) 1 else 0))
  list(breaks = brks, labels = labs)
}

plot_pvals <- function(df, sampsize, ylim = NULL, log_scale = TRUE, 
                       title = NULL, eps_rule = NULL,
                       min_ticks = 5, breaks_incl = 1e-3) {

  df <- dplyr::mutate(df, effect_size = as.character(effect_size))

  p <- ggplot(df, aes(x = obs_stats)) +
    geom_point(aes(y = p_wilcox, shape = "Wilcoxon"),
               color = "black", size = 2, stroke = 0.5, alpha = 0.5) +
    geom_point(aes(y = p_constr, shape = "approximated", color = effect_size),
               size = 1.6, alpha = 0.6) +
    scale_shape_manual(
      name = "P-value type:",
      values = c("approximated" = 16, "Wilcoxon" = 1)
    ) +
    scale_color_manual(
      name = "Effect size:",
      values = c("0"="gold","0.5"="#ffb347","1"="#ff8c00","1.5"="#e55b00","2"="red")
    ) +
    labs(
      title = ifelse(!is.null(title), title,
                     paste0("n = ", sampsize, "; epsilon rule: ", eps_rule)),
      x = expression(U[obs]),
      y = expression(p)
    ) +
    theme_minimal() +
    theme(
      axis.text    = element_text(size = 14),
      axis.title   = element_text(size = 15),
      plot.title   = element_text(size = 15),
      strip.text   = element_text(size = 15),
      legend.title = element_text(size = 16),
      legend.text  = element_text(size = 16)
    )  +
    guides(
      color = guide_legend(override.aes = list(size = 2.5)),
      shape = guide_legend(override.aes = list(size = 2.5))
    )

  # range
  if (is.null(ylim)) {
    y_min <- suppressWarnings(min(c(df$p_wilcox, df$p_constr), na.rm = TRUE))
    y_max <- suppressWarnings(max(c(df$p_wilcox, df$p_constr), na.rm = TRUE))
  } else {
    y_min <- ylim[1]; y_max <- ylim[2]
    p <- p + coord_cartesian(ylim = ylim)
  }
  
  if (log_scale) {
    br <- .compute_log_breaks(ymin = y_min, ymax = y_max,
                              min_ticks = min_ticks,
                              breaks_incl = breaks_incl)
    p <- p + scale_y_log10(breaks = br$breaks, labels = br$labels, minor_breaks = NULL)
  }
  
  p
}

```

## Function to run permApprox

This function will be used for all p-values approximations with constrained GPD fit.

```{r run_permApprox_function, echo=TRUE}
run_permApprox <- function(rule_name, eps_fct, constraint = "support_at_obs",
                           cores = 8) {
  
  path_out <- path_target()
  
  # ---- pre-check: which outputs already exist?
  out_files <- paste0(path_out, "/permapprox_constr_", rule_name,
                      "_n", n_per_group_vec, "_B", B, ".rds")
  missing_idx <- which(!file.exists(out_files))
  missing_n   <- n_per_group_vec[missing_idx]
  
  # ---- run only for missing n's
  if (length(missing_n) > 0) {
    handlers("txtprogressbar")
    plan(multisession, workers = cores)
    
    with_progress({
      p <- progressor(along = n_per_group_vec)
      
      future_map(
        n_per_group_vec,
        function(n_per_group) {
          # file name for this n
          fname_constr <- paste0(path_out, "/permapprox_constr_", rule_name, "_n",
                                 n_per_group, "_B", B, ".rds")
          
          # skip if already computed
          if (file.exists(fname_constr)) {
            p(message = sprintf("n=%s: skip (exists)", n_per_group))
            return(invisible(NULL))
          }
          
          # Load simulated data for this n
          obs_stats  <- readRDS(paste0(path_out, "/obs_U_stats_n",  n_per_group, 
                                       "_B", B, ".rds"))
          perm_stats <- readRDS(paste0(path_out, "/perm_U_stats_n", n_per_group, 
                                       "_B", B, ".rds"))
          
          # GPD control
          gpd_ctrl_constr <- make_gpd_ctrl(
            fit_method   = "LME",
            tol = 1e-7,
            include_obs  = FALSE,
            constraint   = constraint,
            thresh_method= "ftr",
            exceed0      = 250,
            exceed_min   = 10,
            thresh_step  = 10,
            gof_test     = "ad",
            gof_alpha    = 0.05,
            cores        = 1,
            verbose      = TRUE
          )
          
          # compute and save
          papprox_constr <- compute_p_values(
            obs_stats     = obs_stats,
            perm_stats    = perm_stats,
            eps_fct       = eps_fct,
            sampsize      = n_per_group,
            alternative   = "two_sided",
            null_center   = "mean",   # mean of U-perms is n^2/2
            method        = "gpd",
            gpd_ctrl      = gpd_ctrl_constr,
            adjust_method = "BH"
          )
          
          saveRDS(papprox_constr, file = fname_constr)
          p(message = sprintf("n=%s: done", n_per_group))
          invisible(NULL)
        },
        .options = furrr_options(seed = TRUE)  # reproducible RNG per worker
      )
    })
    
    plan(sequential)
  } # else: nothing to compute; fall through to loading
  

  # Combine results for all n_per_group values
  results_all <- purrr::map_dfr(n_per_group_vec, function(n_per_group) {
    obs_stats    <- readRDS(path_target(paste0("obs_U_stats_n",  n_per_group, 
                                               "_B", B, ".rds")))
    pval_wilcox  <- readRDS(path_target(paste0("pval_wilcox_n",  n_per_group, 
                                               "_B", B, ".rds")))
    papprox_constr <- readRDS(path_target(paste0("permapprox_constr_", rule_name, 
                                                 "_n", n_per_group, "_B", B, ".rds")))

    tibble(
      idx         = seq_along(obs_stats),
      n_per_group = n_per_group,
      effect_size = delta_vec,
      obs_stats   = obs_stats,         # on U-scale
      p_wilcox    = pval_wilcox,
      gpd_fit     = papprox_constr$gpd_fit$fitted,
      p_constr    = papprox_constr$p_unadjusted,
      epsilon     = papprox_constr$gpd_fit$epsilon
    )
  })

  results_all
}

```

## Unconstrained GPD fit

For comparison, we again start with the unconstrained fit.

```{r run_permApprox_unconstr}
path_out <- path_target()

# ---- pre-check: which outputs already exist?
out_files <- paste0(path_out, "/permapprox_unconstr_n", 
                    n_per_group_vec, "_B", B, ".rds")
missing_idx <- which(!file.exists(out_files))
missing_n   <- n_per_group_vec[missing_idx]

# ---- run only for missing n's
if (length(missing_n) > 0) {
  
  handlers("txtprogressbar")
  plan(multisession, workers = 6)
  
  with_progress({
    p <- progressor(along = n_per_group_vec)
    
    future_map(
      n_per_group_vec,
      function(n_per_group) {
        # file name for this n
        fname_unconstr <- paste0(path_out, "/permapprox_unconstr_n", 
                                 n_per_group, "_B", B, ".rds")
        
        # skip if already computed
        if (file.exists(fname_unconstr)) {
          p(message = sprintf("n=%s: skip (exists)", n_per_group))
          return(invisible(NULL))
        }
        
        # Load simulated data for this n
        obs_stats  <- readRDS(paste0(path_out, "/obs_U_stats_n",  
                                     n_per_group, "_B", B, ".rds"))
        perm_stats <- readRDS(paste0(path_out, "/perm_U_stats_n", 
                                     n_per_group, "_B", B, ".rds"))
        
        # GPD control: keep cores=1 to avoid nested parallelism
        gpd_ctrl_unconstr <- make_gpd_ctrl(
          fit_method   = "LME",
          tol          = 1e-7,
          include_obs  = FALSE,
          constraint   = "unconstrained",
          thresh_method= "ftr",
          exceed0      = 250,
          exceed_min   = 10,
          thresh_step  = 10,
          gof_test     = "ad",
          gof_alpha    = 0.05,
          cores        = 1,
          verbose      = TRUE
        )
        
        # Compute and save
        papprox_unconstr <- compute_p_values(
          obs_stats     = obs_stats,
          perm_stats    = perm_stats,
          alternative   = "two_sided",
          null_center   = "mean",   # on U-scale
          method        = "gpd",
          gpd_ctrl      = gpd_ctrl_unconstr,
          adjust_method = "BH"
        )
        
        saveRDS(papprox_unconstr, file = fname_unconstr)
        p(message = sprintf("n=%s: done", n_per_group))
        invisible(NULL)
      },
      .options = furrr_options(seed = TRUE)  # reproducible RNG per worker
    )
  })
  
  plan(sequential)
}
```

```{r}
results_all <- purrr::map_dfr(n_per_group_vec, function(n_per_group) {
  obs_stats     <- readRDS(path_target(paste0("obs_U_stats_n",  n_per_group, "_B", B, ".rds")))
  pval_wilcox   <- readRDS(path_target(paste0("pval_wilcox_n",  n_per_group, "_B", B, ".rds")))
  papprox_unconstr <- readRDS(path_target(paste0("permapprox_unconstr_n", n_per_group, "_B", B, ".rds")))

  tibble(
    idx         = seq_along(obs_stats),
    n_per_group = n_per_group,
    effect_size = delta_vec,
    obs_stats   = obs_stats,                 # U-scale
    p_wilcox    = pval_wilcox,
    gpd_fit     = papprox_unconstr$gpd_fit$fitted,
    p_unconstr  = papprox_unconstr$p_unadjusted
  )
})

```

```{r pvals_unconstr, fig.width = 10, fig.height = 6, dpi = 200}
for (n_val in sort(unique(results_all$n_per_group))) {
  df_n <- results_all %>% filter(n_per_group == n_val)

  min_nonzero <- min(c(df_n$p_wilcox[df_n$p_wilcox > 0],
                       df_n$p_unconstr[df_n$p_unconstr > 0]),
                     na.rm = TRUE)
  zero_repl <- min_nonzero / 100

  df_n <- df_n %>%
    mutate(
      p_wilcox_plot    = ifelse(p_wilcox   == 0, zero_repl, p_wilcox),
      p_unconstr_plot  = ifelse(p_unconstr == 0, zero_repl, p_unconstr)
    )

  p <- ggplot(df_n, aes(x = obs_stats)) +
    geom_point(aes(y = p_wilcox_plot, shape = "Wilcoxon"),
               color = "black", size = 2, stroke = 0.5, alpha = 0.5) +
    geom_point(aes(y = p_unconstr_plot, shape = "Unconstrained GPD", color = effect_size),
               size = 1.6, alpha = 0.6) +
    scale_shape_manual(name = "p-value type",
                       values = c("Unconstrained GPD" = 16, "Wilcoxon" = 1)) +
    scale_color_manual(name = "Effect size",
                       values = c("0"="gold",
                                  "0.5"="#ffb347",
                                  "1"="#ff8c00",
                                  "1.5"="#e55b00",
                                  "2"="red")) +
    scale_y_log10(
      breaks = c(zero_repl, 1e-40, 1e-20, 1e-2),
      labels = c("0", "1e-40", "1e-20", "1e-2"),
      limits = c(zero_repl, 1),
      minor_breaks = NULL
    ) +
    labs(
      title = paste("Unconstrained permApprox vs. Mann-Whitney U (n =", n_val, ")"),
      x = expression(U[obs]),
      y = expression(p)
    ) +
    theme_minimal() +
    theme(
      axis.text  = element_text(size = 14),
      axis.title = element_text(size = 16),
      plot.title = element_text(size = 16),
      strip.text = element_text(size = 15),
      legend.title = element_text(size = 14),
      legend.text  = element_text(size = 12)
    )

  print(p)
}
```

## Non-standardized Lifted Log-Saturation (LLS) rule

```{r}
eps_n_power <- function(obs_stats = NULL, sampsize = NULL, C = 1000, p = 2, ...) {
  abs(obs_stats) * (C / sampsize^p)
}

# Build T_cap from choice: "perm" or "tmax"
# - cap_base = "tmax": uses observed maximum across tests
# - cap_base = "perm": uses high quantile of |perm_stats| inflated by g(n)
.build_Tcap <- function(cap_base, t_obs, perm_stats, n, 
                        q_ref = 0.99, alpha = 0.5) {
  if (identical(cap_base, "tmax")) {
    # Observed cap (same for all tests)
    T_cap <- rep(max(abs(t_obs), na.rm = TRUE), length(t_obs))

  } else if (identical(cap_base, "perm")) {
    # Permutation cap: high quantile of |perm|, inflated by g(n)
    if (is.null(dim(perm_stats))) perm_stats <- matrix(perm_stats, ncol = 1)
    Tabs <- abs(perm_stats)
    Tq <- apply(Tabs, 2, stats::quantile, probs = q_ref, na.rm = TRUE, names = FALSE)
    
    # Inflation g(n) = max(1, alpha * sqrt(n))
    infl <- pmax(1, alpha * sqrt(n))
    T_cap <- Tq * infl

    # Broadcast if needed
    if (length(T_cap) == 1L) T_cap <- rep(T_cap, length(t_obs))
    if (length(T_cap) != length(t_obs)) T_cap <- rep(T_cap[1L], length(t_obs))

  } else {
    stop("cap_base must be 'perm' or 'tmax'.")
  }

  # Guardrails: enforce positive finite caps
  bad <- !is.finite(T_cap) | T_cap <= 0
  if (any(bad)) {
    fallback <- max(T_cap[!bad], 1)
    T_cap[bad] <- fallback
  }
  T_cap
}

# Simplified LLS using unified T_cap
eps_LLS <- function(obs_stats,
                    perm_stats,
                    sampsize,
                    cap_base = "tmax",
                    k_factor = 1000,
                    target_factor = 0.25,
                    rho_lift = 3,
                    floor_eps = 1e-6,
                    q_ref = 0.99,   # quantile of |T_perm|
                    alpha = 0.5,    # inflation constant
                    ...) {
  cap_base <- if (is.character(cap_base)) match.arg(cap_base) else cap_base
  Tobs <- abs(obs_stats)
  n <- sampsize
  
  # Build cap T_cap according to cap_base
  T_cap <- .build_Tcap(cap_base, Tobs, perm_stats, n,
                       q_ref = q_ref, alpha = alpha)
  
  # Curvature (k) and plateau (E_target) scale with 1/n
  k <- k_factor * (500 / n)
  E_target <- target_factor * (500 / n)
  
  # Degenerate guard: if cap fails, return conservative constant
  if (!any(is.finite(T_cap)) || all(T_cap <= 0)) {
    return(rep(max(E_target + rho_lift, floor_eps), length(Tobs)))
  }
  
  # Scale-free index s and smooth Wendland C^2 lift
  s <- pmin(Tobs / T_cap, 1)                # clamp to [0,1]
  psi_wendland_c2 <- (1 - s)^4 * (1 + 4*s)  # lift vanishes at s=1
  
  # Simplified LLS core (no A,B), numerically stable via log1p
  eps_core <- E_target * log1p(k * s) / log1p(k) + rho_lift * psi_wendland_c2
  
  # Floor to avoid vanishing constraints
  pmax(eps_core, floor_eps)
}
```

```{r rule_LLS}
n_val <- 100
  obs_stats <- readRDS(path_target(paste0("obs_U_stats_n", n_val, "_B", B, ".rds")))
  perm_stats <- readRDS(path_target(paste0("perm_U_stats_n", n_val, "_B", B, ".rds")))
  
  p <- tibble(
    t_obs = abs(obs_stats),
    eps_n2 = eps_n_power(obs_stats, sampsize = n_val),
    eps_LLS= eps_LLS(obs_stats, perm_stats = perm_stats, sampsize = n_val)
  ) %>%
    pivot_longer(-t_obs, names_to = "rule", values_to = "eps") %>%
    ggplot(aes(t_obs, eps, color = rule)) +
    geom_line(size = 1.2, alpha = 0.7) +
    labs(
      title = paste0("Epsilon scaling comparison (n = ", n_val, ")"),
      x = expression(W[obs]),
      y = expression(epsilon)
    ) +
    theme_minimal(base_size = 14)
  print(p)
```


```{r run_permApprox_constr_LLS}
results_all <- run_permApprox(rule_name = "LLS", eps_fct = eps_LLS)
```

```{r pvals_constr_LLS}
for (n_val in n_per_group_vec) {
  df_n <- results_all %>% filter(n_per_group == n_val)#, gpd_fit == TRUE)
  p <- plot_pvals(df = df_n, sampsize = n_val, eps_rule = "LLS")
  print(p)
}
```

As expected, the LLS rule does not work for the Wilcoxon test statistics as their range is extremely large.

## Standardized Lifted Log-Saturation (SLLS) rule

```{r}
# Build T_cap from choice: "perm" or "tmax"
# - cap_base = "tmax": uses observed maximum across tests
# - cap_base = "perm": uses high quantile of |perm_stats| inflated by g(n)
.build_Tcap <- function(cap_base, t_obs, perm_stats, n, 
                        q_ref = 0.99, alpha = 0.5) {
  if (identical(cap_base, "tmax")) {
    # Observed cap (same for all tests)
    T_cap <- rep(max(abs(t_obs), na.rm = TRUE), length(t_obs))

  } else if (identical(cap_base, "perm")) {
    # Permutation cap: high quantile of |perm|, inflated by g(n)
    if (is.null(dim(perm_stats))) perm_stats <- matrix(perm_stats, ncol = 1)
    Tabs <- abs(perm_stats)
    Tq <- apply(Tabs, 2, stats::quantile, probs = q_ref, na.rm = TRUE, names = FALSE)
    
    # Inflation g(n) = max(1, alpha * sqrt(n))
    infl <- pmax(1, alpha * sqrt(n))
    T_cap <- Tq * infl

    # Broadcast if needed
    if (length(T_cap) == 1L) T_cap <- rep(T_cap, length(t_obs))
    if (length(T_cap) != length(t_obs)) T_cap <- rep(T_cap[1L], length(t_obs))

  } else {
    stop("cap_base must be 'perm' or 'tmax'.")
  }

  # Guardrails: enforce positive finite caps
  bad <- !is.finite(T_cap) | T_cap <= 0
  if (any(bad)) {
    fallback <- max(T_cap[!bad], 1)
    T_cap[bad] <- fallback
  }
  T_cap
}

# Standardized Lifted Log-Saturation (SLLS) using unified Z-cap
# - Standardize to Z using permutation mean/sd per test
# - Build the cap on the Z-scale via .build_Tcap (cap_base = "perm" or "tmax")
# - Apply simplified LLS on Z
# - Map epsilon back to T-scale via sigma_j
eps_SLLS <- function(obs_stats,
                     perm_stats,
                     sampsize,
                     cap_base      = "tmax",  # Z-cap choice
                     alpha         = 0.5, # g(n) = max(1, alpha*sqrt(n))
                     q_ref         = 0.99, # quantile of |Z_perm|
                     rho_lift      = 3,
                     k_factor      = 1000,
                     target_factor = 0.25,
                     floor_epsZ    = 1e-6,
                     ...) {
  cap_base <- if (is.character(cap_base)) match.arg(cap_base) else cap_base
  n <- sampsize
  stopifnot(n > 0)
  
  # Shape inputs --------------------------------------------------------------
  # Allow vector perm_stats for single test
  if (is.null(dim(perm_stats))) perm_stats <- matrix(perm_stats, ncol = 1)
  Tobs <- as.numeric(obs_stats)
  
  # Z-standardization per test -----------------------------------------------
  mu  <- colMeans(perm_stats, na.rm = TRUE)
  sig <- apply(perm_stats, 2, sd, na.rm = TRUE)

  # Guard against non-finite or zero SDs: replace with a small positive fallback
  valid_sig <- sig[is.finite(sig) & sig > 0]
  sig_fallback <- if (length(valid_sig)) max(min(valid_sig), 1e-12) else 1e-12
  sig_ok <- sig
  sig_ok[!is.finite(sig_ok) | sig_ok <= 0] <- sig_fallback

  Zobs  <- abs((Tobs - mu) / sig_ok)
  Zperm <- sweep(sweep(perm_stats, 2, mu, "-"), 2, sig_ok, "/")  # same shape as perm_stats

  # Build Z-cap (unified) -----------------------------------------------------
  # Uses: "tmax" => max(Zobs); "perm" => Q_q(|Zperm|) * g(n) with g(n)=max(1, alpha*sqrt(n))
  T_cap_Z <- .build_Tcap(cap_base = cap_base,
                         t_obs    = Zobs,
                         perm_stats = Zperm,
                         n        = n,
                         q_ref    = q_ref,
                         alpha    = alpha)

  # LLS core on Z -------------------------------------------------------------
  k        <- k_factor      * (500 / n)
  E_target <- target_factor * (500 / n)

  if (!any(is.finite(T_cap_Z)) || all(T_cap_Z <= 0)) {
    # Conservative constant if the cap failed
    epsZ <- rep(max(E_target + rho_lift, floor_epsZ), length(Zobs))
  } else {
    s   <- pmin(Zobs / T_cap_Z, 1)        # scale-free index in [0,1]
    psi <- (1 - s)^4 * (1 + 4*s)          # Wendland C^2 lift
    epsZ_core <- E_target * log1p(k * s) / log1p(k) + rho_lift * psi
    epsZ <- pmax(epsZ_core, floor_epsZ)
  }

  # Map back to T-scale -------------------------------------------------------
  epsT <- epsZ * sig_ok
  epsT
}
```

```{r rule_SLLS}
n_val <- 100
  obs_stats <- readRDS(path_target(paste0("obs_U_stats_n", n_val, "_B", B, ".rds")))
  perm_stats <- readRDS(path_target(paste0("perm_U_stats_n", n_val, "_B", B, ".rds")))
  
  p <- tibble(
    t_obs = abs(obs_stats),
    eps_n2 = eps_n_power(obs_stats, sampsize = n_val),
    eps_SLLS= eps_SLLS(obs_stats, perm_stats = perm_stats, sampsize = n_val)
  ) %>%
    pivot_longer(-t_obs, names_to = "rule", values_to = "eps") %>%
    ggplot(aes(t_obs, eps, color = rule)) +
    geom_line(size = 1.2, alpha = 0.7) +
    labs(
      title = paste0("Epsilon scaling comparison (n = ", n_val, ")"),
      x = expression(W[obs]),
      y = expression(epsilon)
    ) +
    theme_minimal(base_size = 14)
  print(p)
```

```{r run_permApprox_constr_SLLS}
results_all <- run_permApprox(rule_name = "SLLS", eps_fct = eps_SLLS)
```

```{r pvals_constr_SLLS}
for (n_val in n_per_group_vec) {
  df_n <- results_all %>% filter(n_per_group == n_val)#, gpd_fit == TRUE)
  p <- plot_pvals(df = df_n, sampsize = n_val, eps_rule = "SLLS")
  print(p)
}
```

## Constrained vs. unconstrained fit

```{r}
# Load & combine results: wilcox, unconstrained GPD, constrained GPD
rule_name <- "SLLS"   # <- change if you used a different rule_name

results_three <- purrr::map_dfr(n_per_group_vec, function(n_per_group) {

  # base inputs
  obs_stats   <- readRDS(path_target(paste0("obs_U_stats_n", 
                                            n_per_group, "_B", B, ".rds")))
  pval_wilcox  <- readRDS(path_target(paste0("pval_wilcox_n", 
                                             n_per_group, "_B", B, ".rds")))

  # GPD results
  papprox_unconstr <- readRDS(path_target(paste0("permapprox_unconstr_n", 
                                                 n_per_group, "_B", B, ".rds")))
  papprox_constr   <- readRDS(path_target(paste0("permapprox_constr_", 
                                                 rule_name, "_n", n_per_group, "_B", B, ".rds")))

  tibble::tibble(
    idx           = seq_along(obs_stats),
    n_per_group   = n_per_group,
    effect_size   = factor(delta_vec, levels = c(0, 0.5, 1, 1.5, 2)),
    obs_stats     = abs(obs_stats),
    p_wilcox       = pval_wilcox,
    p_unconstr    = papprox_unconstr$p_unadjusted,
    p_constr      = papprox_constr$p_unadjusted,
    gpd_unconstr_fit = papprox_unconstr$gpd_fit$fitted,
    gpd_constr_fit   = papprox_constr$gpd_fit$fitted
  )
})
```


```{r}
# Plot function: three p-value types in one panel
plot_pvals_three <- function(df, sampsize, log_scale = TRUE) {
  df_plot <- df

  # find smallest nonzero across all three types for stable zero replacement on log scale
  min_nonzero <- min(
    c(df_plot$p_wilcox[df_plot$p_wilcox > 0],
      df_plot$p_unconstr[df_plot$p_unconstr > 0],
      df_plot$p_constr[df_plot$p_constr > 0]),
    na.rm = TRUE
  )
  min_nonzero <- max(min_nonzero, 1e-320)
  zero_repl   <- min_nonzero / 100
  
  df_long <- df_plot %>%
    dplyr::mutate(
      p_wilcox_plot     = ifelse(p_wilcox     == 0, zero_repl, p_wilcox),
      p_unconstr_plot  = ifelse(p_unconstr  == 0, zero_repl, p_unconstr),
      p_constr_plot    = ifelse(p_constr    == 0, zero_repl, p_constr)
    ) %>%
    tidyr::pivot_longer(
      cols = c(p_wilcox_plot, p_unconstr_plot, p_constr_plot),
      names_to = "ptype", values_to = "pval"
    ) %>%
    dplyr::mutate(
      ptype = dplyr::recode(ptype,
                            p_wilcox_plot    = "Mann-Whitney U",
                            p_unconstr_plot = "Unconstrained",
                            p_constr_plot   = "Constrained"
      ),
      # set the legend order
      ptype = factor(ptype, levels = c("Mann-Whitney U", "Constrained", 
                                       "Unconstrained"))
    ) %>%
    # this controls the draw order in a single geom_point layer
    dplyr::arrange(ptype)
  
  p <- ggplot2::ggplot(df_long, ggplot2::aes(x = obs_stats, y = pval, 
                                             color = ptype, size = ptype)) +
    ggplot2::geom_point(alpha = 0.6) +
    ggplot2::scale_color_manual(
      name   = "P-value type:",
      values = c("Mann-Whitney U" = "grey40", 
                 "Unconstrained" = "#1f77b4",
                 "Constrained" = "darkorange")
    ) + 
    scale_size_manual(
      values = c("Mann-Whitney U" = 1.6,
                 "Constrained" = 1.6,
                 "Unconstrained" = 1.2)  # smaller dots here
    ) +
    ggplot2::labs(
      title = paste0("n = ", sampsize),
      x = expression(U[obs]),
      y = expression(p)
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      axis.text    = ggplot2::element_text(size = 14),
      axis.title   = ggplot2::element_text(size = 16),
      plot.title   = ggplot2::element_text(size = 16),
      legend.title = ggplot2::element_text(size = 14),
      legend.text  = ggplot2::element_text(size = 12)
    ) +
    guides(size = "none")

  if (log_scale) {
    p <- p +
      ggplot2::scale_y_log10(
        breaks = c(zero_repl, 1e-300, 1e-200, 1e-100, 1e-80, 1e-60, 1e-40, 
                   1e-20, 1e-2),
        labels = c("0", "1e-300", "1e-200", "1e-100", "1e-80", "1e-60", "1e-40", 
                   "1e-20", "1e-2"),
        limits = c(zero_repl, 1),
        minor_breaks = NULL
      )
  }

  return(p)
}
```


```{r pvals_unconstrained_vs_constrained, fig.width=10, fig.height=6, dpi=200}
# Make the plots for each sample size
for (n_val in n_per_group_vec) {
  df_n <- dplyr::filter(results_three, n_per_group == n_val)
  cat("n =", n_val,
      "| zeros (wilcox, unconst, const):",
      sum(df_n$p_wilcox == 0, na.rm = TRUE), ",",
      sum(df_n$p_unconstr == 0, na.rm = TRUE), ",",
      sum(df_n$p_constr == 0, na.rm = TRUE), "\n")
  print(plot_pvals_three(df_n, sampsize = n_val))
}

```

## Combined plot

```{r}
results_lls  <- run_permApprox(rule_name = "LLS", eps_fct = eps_factor)
results_slls <- run_permApprox(rule_name = "SLLS", eps_fct = eps_factor)
```

```{r eps_rules_combined, fig.width=10, fig.height=4, dpi=200}
sampsize <- 100

p_lls <- results_lls  %>% dplyr::filter(n_per_group == sampsize) %>%
  plot_pvals(sampsize = sampsize, title = "Lifted log-saturation",
             ylim = c(1e-45, 1))

p_slls <- results_slls %>% dplyr::filter(n_per_group == sampsize) %>%
  plot_pvals(sampsize = sampsize, 
             title = "Standardized lifted log-saturation",
             ylim = c(1e-45, 1)) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )


# All four plots
(p_lls | p_slls) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
