---
title: "Wilcoxon rank-sum on exponential data - Test epsilon rule for single tests"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "02_single_tests" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "f6f7de73-abcc-4ebe-b039-ea102462293e")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/expl_wilcox_single_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

In this script we investigate the performance of our **final $\varepsilon$-rule (the SLLS rule)** when applied to data generated from exponential distributions and analyzed with the **Wilcoxon rank-sum test**.

In contrast to the previous script, which depicted a multiple-testing setting, we are examining a single-test setting here. That means, each test is treated independently: the $p$-value approximation is computed per test, and the choice of $\varepsilon$ depends only on the test’s own permutation distribution rather than on the maximum statistic across multiple tests. 

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(ggplot2)
library(patchwork)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(future)
library(future.apply)
library(kableExtra)
library(furrr)
library(fdrtool)
library(fitdistrplus)
library(goftest)
library(limma)
library(progressr)

conflicts_prefer(dplyr::select, dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r load_permApprox}
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))

# Load permApprox functions
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$",
                                   full.names = TRUE)
invisible(sapply(permApprox_functions, source))

# Load epsilon functions
eps_rule_path <- file.path(proj_path, 
                           "functions/permApprox_eps_fct/slls_eps_rule.R")
source(eps_rule_path)
```

## Data Simulation

```{r}
n_tests <- 1000
n_per_group_vec <- c(50, 100, 250, 500, 750, 1000)
B <- 1000

effect_blocks <- c(0, 0.5, 1, 1.5, 2)
delta_vec <- rep(effect_blocks, each = n_tests / length(effect_blocks)) %>% 
  factor(levels = effect_blocks)
```

We reuse the data from the former simulation study, where we simulated two groups from **exponential distributions** with increasing effect sizes:

* Group 1: mean (=1) (rate (=1))
* Group 2: mean (=1+\delta) (rate (=1/(1+\delta)))
* Sample sizes per group: `r n_per_group_vec`
* Number of tests: `r n_tests`
* Effect sizes: `r paste(effect_blocks, collapse = " ")`
* Tests per effect size: `r paste0("d=", effect_blocks, ": ",table(delta_vec), collapse = ", ")`

We apply the **Wilcoxon rank-sum test** (Mann–Whitney $U$ test) to compare the two exponential samples.

* Statistic: **Mann–Whitney (U)** (converted from rank-sum (W))
* P-values: classical Wilcoxon (`exact = FALSE, correct = FALSE`) for reference; permutation/GPD for tail modeling

## Plot function

```{r}
plot_pvals <- function(df, sampsize, ylim = NULL, log_scale = TRUE) {
  p <- ggplot(df, aes(x = obs_stats)) +
    # Mann–Whitney U in black
    geom_point(
      aes(y = p_wilcox, shape = "Mann–Whitney U"),
      color = "black", size = 2, stroke = 0.5, alpha = 0.5
    ) +
    # Constrained GPD p-values colored by effect size
    geom_point(
      aes(y = p_constr, shape = "Constrained GPD", color = effect_size),
      size = 1.6, alpha = 0.6
    ) +
    scale_shape_manual(
      name = "P-value type:",
      values = c("Constrained GPD" = 16, "Mann–Whitney U" = 1)
    ) +
    scale_color_manual(
      name = "Effect size:",
      values = c("0"="gold","0.5"="#ffb347","1"="#ff8c00","1.5"="#e55b00","2"="red")
    ) +
    labs(
      title = 
        paste0("Mann–Whitney U vs. constrained GPD fit (single tests, n = ", 
               sampsize, ")"),
      x = expression(U[obs]),
      y = expression(p)
    ) +
    theme_minimal() +
    theme(
      axis.text  = element_text(size = 14),
      axis.title = element_text(size = 16),
      plot.title = element_text(size = 16),
      strip.text = element_text(size = 15),
      legend.title = element_text(size = 14),
      legend.text  = element_text(size = 12)
    )
  if (log_scale) p <- p + scale_y_log10(labels = scales::scientific)
  if (!is.null(ylim)) p <- p + coord_cartesian(ylim = ylim)
  p
}

```

## Function to run permApprox

```{r run_permApprox_function}
run_permApprox <- function(rule_name, eps_fct, constraint = "support_at_obs",
                           cores = 8) {

  for (n_per_group in n_per_group_vec) {

    # Load simulated data (Mann–Whitney U, from the multi-test simulator with B in name)
    obs_stats   <- readRDS(path_source("01_test_eps_rule",
                                       paste0("obs_U_stats_n",  n_per_group, 
                                              "_B", B, ".rds")))
    perm_stats  <- readRDS(path_source("01_test_eps_rule",
                                       paste0("perm_U_stats_n", n_per_group, 
                                              "_B", B, ".rds")))
    pval_wilcox <- readRDS(path_source("01_test_eps_rule",
                                       paste0("pval_wilcox_n",  n_per_group, 
                                              "_B", B, ".rds")))

    fname_constr <- path_target(paste0("single_test_constr_", rule_name, "_n",
                                       n_per_group, "_B", B, ".rds"))
    if (file.exists(fname_constr)) next

    cat("n:", n_per_group, "\n")

    gpd_ctrl_constr <- make_gpd_ctrl(
      fit_method   = "LME",
      tol          = 1e-7,
      include_obs  = FALSE,
      constraint   = constraint,
      thresh_method= "ftr",
      exceed0      = 250,
      exceed_min   = 10,
      thresh_step  = 10,
      gof_test     = "ad",
      gof_alpha    = 0.05,
      cores        = 1,
      verbose      = TRUE
    )

    # Parallelize across tests (each test is independent)
    oplan <- future::plan(multisession, workers = cores)
    on.exit(future::plan(oplan), add = TRUE)

    with_progress({
      p <- progressr::progressor(along = seq_len(n_tests))

      papprox_list <- future_lapply(
        seq_len(n_tests),
        function(m, p) {
          p()
          compute_p_values(
            obs_stats     = obs_stats[m],          # single number (U_obs)
            perm_stats    = perm_stats[, m],       # vector (U_perm)
            eps_fct       = eps_fct,               # SLLS
            sampsize      = n_per_group,
            alternative   = "two_sided",
            null_center   = "mean",                # mean(U_perm) ~ n^2/2
            method        = "gpd",
            gpd_ctrl      = gpd_ctrl_constr,
            adjust_method = "BH"
          )
        },
        p = p,
        future.seed = TRUE   # not used by compute_p_values(), but silences RNG warnings
      )
    })

    saveRDS(papprox_list, file = fname_constr)
  }

  # ---- Combine results for all n_per_group values ----
  results_all <- purrr::map_dfr(n_per_group_vec, function(n_per_group) {

    obs_stats <- readRDS(path_source("01_test_eps_rule",
                                     paste0("obs_U_stats_n",  n_per_group, 
                                            "_B", B, ".rds")))
    pval_wilcox <- readRDS(path_source("01_test_eps_rule",
                                       paste0("pval_wilcox_n",  n_per_group, 
                                              "_B", B, ".rds")))
    papprox_list <- readRDS(path_target(paste0("single_test_constr_", rule_name, 
                                               "_n", n_per_group, "_B", B, ".rds")))

    stopifnot(length(papprox_list) == length(obs_stats),
              length(pval_wilcox)  == length(obs_stats))

    purrr::map_dfr(seq_along(papprox_list), function(m) {
      fit <- papprox_list[[m]]
      tibble::tibble(
        idx         = m,
        n_per_group = n_per_group,
        effect_size = delta_vec[m],
        obs_stats   = obs_stats[m],                     # U-scale (no abs)
        p_wilcox    = pval_wilcox[m],                   # “Mann–Whitney U (asymptotic)”
        gpd_fit     = ifelse(is.null(fit$gpd_fit), NA, fit$gpd_fit$fitted),
        p_constr    = fit$p_unadjusted,
        epsilon     = ifelse(is.null(fit$gpd_fit), NA, fit$gpd_fit$epsilon)
      )
    })
  })

  results_all
}
```

## Run permApprox 

```{r run_permApprox_constr_SLLS}
results_all <- run_permApprox(rule_name = "SLLS", eps_fct = eps_slls_perm)
```

## Constrained GPD vs Mann-Whitney U p-values

```{r pvalues_wilcox_vs_constrained_SLSL, fig.width=10, fig.height=6, dpi=200}
for (n_val in n_per_group_vec) {
  df_n <- results_all %>% filter(n_per_group == n_val)
  p <- plot_pvals(df = df_n, sampsize = n_val)
  print(p)
}
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
