---
title: "Mann–Whitney U - Exceedances & Permutations (SLLS vs. unconstrained)"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "03_exceedances"
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "395a0ba5-185a-4db3-a7c0-d661b80ab60f")
```

In this script we extend the exceedance study to **iterate over the number of permutations** `B` *and* the starting exceedances `exceed0`.
We compare **unconstrained** vs. **constrained (SLLS rule)** GPD fits against the **Mann–Whitney U** reference.

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/expl_wilcox_exceed_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
# Load packages
library(conflicted)

library(ggplot2)
library(patchwork)
library(pheatmap)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(future)
library(future.apply)
library(kableExtra)
library(furrr)
library(patchwork)
library(digest)     # seed_from

# For permApprox:
library(fdrtool)
library(fitdistrplus)
library(goftest)
library(limma)
library(progressr)

conflicts_prefer(dplyr::select, dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r, message=FALSE, warning=FALSE}
# Load permApprox functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(permApprox_functions, source))
```

## Design

```{r design}
cores <- 8

n_tests            <- 1000
n_per_group_vec    <- c(100, 500)      # per-group n
B_vec              <- c(1000, 10000)   # permutations
exceed0_vec        <- c(100, 250, 0.25, 0.10) # absolute OR proportion of B

# Effect sizes by blocks
effect_blocks <- c(0, 0.5, 1, 1.5, 2)
delta_vec     <- rep(effect_blocks, each = n_tests / length(effect_blocks)) %>%
  factor(levels = effect_blocks)

# Scenario grid
scenarios <- tidyr::expand_grid(
  n_per_group = n_per_group_vec,
  B           = B_vec,
  exceed0     = exceed0_vec
)

scenarios
```

> Notes:
>
> * If `exceed0 < 1`, we treat it as a **proportion of B** (e.g., `0.25 → ceiling(0.25*B)`); otherwise it’s used as a fixed integer.
> * We never trim the tail to fewer exceedances than requested if the observed statistic already sits in the tail.

## Simulation (once per $n, B$)

```{r}
### Helpers

# Stable 32-bit integer from arbitrary keys
seed_from <- function(...) {
  hx <- digest(paste(..., collapse="|"), algo = "xxhash32")
  s  <- as.integer(strtoi(hx, base = 16L))
  if (is.na(s) || s == 0L) 1L else s
}

# One test -> (U_obs, U_perm[B], classical p asymptotic)
.sim_one_U <- function(n_per_group, B, delta) {
  # Exponential data
  x1 <- rexp(n_per_group, rate = 1)                       # mean 1
  x2 <- rexp(n_per_group, rate = 1 / (1 + as.numeric(delta)))  # mean 1+delta

  # Pooled ranks and W (sum ranks of group 1), then convert to U
  x_comb <- c(x1, x2)
  n_comb <- length(x_comb)
  r_all  <- rank(x_comb, ties.method = "average")

  W_obs  <- sum(r_all[seq_len(n_per_group)])
  shift  <- n_per_group * (n_per_group + 1) / 2
  U_obs  <- W_obs - shift

  # Permutations: indices of group 1 (columns), then W -> U
  perm_idx_mat <- replicate(B, sample.int(n_comb, n_per_group))
  W_perm <- colSums(matrix(r_all[perm_idx_mat], nrow = n_per_group, ncol = B))
  U_perm <- W_perm - shift

  # Asymptotic Wilcoxon p (no continuity corr)
  p_cl <- wilcox.test(x1, x2, alternative = "two.sided",
                      exact = FALSE, correct = FALSE)$p.value

  list(U_obs = U_obs, U_perm = U_perm, p_cl = p_cl)
}

simulate_once_U <- function(n_per_group, B, n_tests, delta_vec, path_target) {
  fn_obs <- path_target(paste0("obs_U_stats_n",  n_per_group, "_B", B, ".rds"))
  fn_perm<- path_target(paste0("perm_U_stats_n", n_per_group, "_B", B, ".rds"))
  fn_pwc <- path_target(paste0("pval_wilcox_n", n_per_group, "_B", B, ".rds"))
  
  if (all(file.exists(fn_obs, fn_perm, fn_pwc))) return(invisible(NULL))

  message("Simulating (Mann–Whitney U, Exponential) for n=", n_per_group, ", B=", B, " ...")

  # scenario-specific master seed (subset-invariant)
  master_seed <- seed_from("wilcoxU-exp", n_per_group, B, n_tests)

  oplan <- future::plan(multisession, workers = cores)
  on.exit(future::plan(oplan), add = TRUE)

  obs_stats  <- numeric(n_tests)
  pval_wcx   <- numeric(n_tests)
  perm_stats <- matrix(NA_real_, nrow = B, ncol = n_tests)

  with_progress({
    p <- progressr::progressor(along = seq_len(n_tests))

    res_list <- future_lapply(
      seq_len(n_tests),
      function(g, p) {
        p()
        delta_g <- as.numeric(as.character(delta_vec[g]))
        .sim_one_U(n_per_group, B, delta_g)
      },
      p = p,
      future.seed = master_seed
    )

    for (g in seq_len(n_tests)) {
      obs_stats[g]    <- res_list[[g]]$U_obs
      pval_wcx[g]     <- res_list[[g]]$p_cl
      perm_stats[, g] <- res_list[[g]]$U_perm
    }
  })

  saveRDS(obs_stats,  fn_obs)
  saveRDS(perm_stats, fn_perm)
  saveRDS(pval_wcx,   fn_pwc)
  invisible(NULL)
}
```

```{r simulate, eval=TRUE}
# Create full grid, then simulate each (n, B) scenario once
scenarios_data <- expand_grid(n_per_group = n_per_group_vec, B = B_vec) 

scenarios_data %>% 
  pwalk(~ simulate_once_U(..1, ..2, n_tests, delta_vec, path_target))
```

## Helpers for filenames & exceed0 handling

```{r helpers}
# Tag exceed0 in filenames robustly (keeps proportion info)
tag_exceed0 <- function(exceed0) {
  if (exceed0 < 1) {
    paste0("p", gsub("\\.", "", sprintf("%.4f", exceed0)))
  } else {
    paste0(exceed0)
  }
}
```

## permApprox runs

We compute both **unconstrained** and **constrained (SLLS)** $p$-values for every scenario.
All results are saved per-scenario and then **combined safely** afterward.

```{r run-permapprox}
run_permapprox_unconstr <- function(n_per_group, B, exceed0) {
  ex_tag <- tag_exceed0(exceed0)
  fn_out <- path_target(paste0("permapprox_unconstr_n", n_per_group, 
                               "_B", B, "_ex", ex_tag,".rds"))
  if (file.exists(fn_out)) return(invisible(NULL))

  message("Unconstrained: n=", n_per_group, ", B=", B, ", exceed0=", exceed0)
  obs_stats  <- readRDS(path_target(paste0("obs_U_stats_n",  n_per_group, "_B", B, ".rds")))
  perm_stats <- readRDS(path_target(paste0("perm_U_stats_n", n_per_group, "_B", B, ".rds")))

  gpd_ctrl <- make_gpd_ctrl(
    fit_method   = "LME",
    include_obs  = FALSE,
    constraint   = "unconstrained",
    thresh_method= "ftr",
    exceed0      = exceed0,
    exceed_min   = 10,
    thresh_step  = 10,
    gof_test     = "ad",
    gof_alpha    = 0.05,
    cores        = 1,
    verbose      = TRUE
  )

  res <- compute_p_values(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,
    alternative   = "two_sided",
    null_center   = "mean",   # on U-scale
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl,
    adjust_method = "BH"
  )

  saveRDS(res, fn_out)
  invisible(NULL)
}

run_permapprox_constr <- function(n_per_group, B, exceed0,
                                  rule_name = "slls_std",
                                  eps_fct   = eps_SLLS) {
  ex_tag <- tag_exceed0(exceed0)
  fn_out <- path_target(paste0("permapprox_constr_n", n_per_group, 
                               "_B", B, "_ex", ex_tag,".rds"))
  if (file.exists(fn_out)) return(invisible(NULL))

  message("Constrained(SLLS): n=", n_per_group, ", B=", B, ", exceed0=", exceed0)
  obs_stats  <- readRDS(path_target(paste0("obs_U_stats_n",  n_per_group, "_B", B, ".rds")))
  perm_stats <- readRDS(path_target(paste0("perm_U_stats_n", n_per_group, "_B", B, ".rds")))

  gpd_ctrl <- make_gpd_ctrl(
    fit_method   = "LME",
    include_obs  = FALSE,
    constraint   = "support_at_obs",
    thresh_method= "ftr",
    exceed0      = exceed0,
    exceed_min   = 10,
    thresh_step  = 10,
    gof_test     = "ad",
    gof_alpha    = 0.05,
    cores        = 1,
    verbose      = TRUE
  )

  res <- compute_p_values(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,
    eps_fct       = eps_fct,
    sampsize      = n_per_group,
    alternative   = "two_sided",
    null_center   = "mean",
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl,
    adjust_method = "BH"
  )

  saveRDS(res, fn_out)
  invisible(NULL)
}

# Run for all scenarios
walk2(seq_len(nrow(scenarios)), 1, function(i, .) {
  sc <- scenarios[i,]
  run_permapprox_unconstr(sc$n_per_group, sc$B, sc$exceed0)
  run_permapprox_constr(sc$n_per_group, sc$B, sc$exceed0)
})
```

## Safe combination of results

```{r combine}
combine_results <- function(scenarios, delta_vec) {
  purrr::pmap_dfr(list(scenarios$n_per_group, scenarios$B, scenarios$exceed0),
                  function(n_per_group, B, exceed0) {
    ex_tag     <- tag_exceed0(exceed0)
    fn_obs     <- path_target(paste0("obs_U_stats_n",  n_per_group, "_B", B, ".rds"))
    fn_pwc     <- path_target(paste0("pval_wilcox_n", n_per_group, "_B", B, ".rds"))
    fn_un      <- path_target(paste0("permapprox_unconstr_n", n_per_group, 
                                     "_B", B, "_ex", ex_tag,".rds"))
    fn_con     <- path_target(paste0("permapprox_constr_n", n_per_group, 
                                     "_B", B, "_ex", ex_tag,".rds"))
    
    obs_stats   <- readRDS(fn_obs)
    pval_wcx    <- readRDS(fn_pwc)
    res_un      <- readRDS(fn_un)
    res_con     <- readRDS(fn_con)
    
    tibble(
      idx             = seq_along(obs_stats),
      n_per_group     = n_per_group,
      B               = B,
      exceed0         = exceed0,
      effect_size     = delta_vec,
      obs_stats       = obs_stats,              # U-scale
      p_U_asympt      = pval_wcx,               # Mann–Whitney U
      p_unconstr      = res_un$p_unadjusted,
      p_constr        = res_con$p_unadjusted,
      method_constr   = res_con$method_used,
      method_unconstr = res_un$method_used,
      gpd_un_fit      = res_un$gpd_fit$fitted,
      gpd_con_fit     = res_con$gpd_fit$fitted,
      epsilon         = res_con$gpd_fit$epsilon
    )
  })
}

results_all <- combine_results(scenarios, delta_vec)
glimpse(results_all)
```

## Plotting

```{r plotting-helpers}
zero_safe_long <- function(df, factor = 100, cols = c("p_U_asympt","p_unconstr","p_constr")) {
  # smallest nonzero across requested columns
  min_nonzero <- min(unlist(lapply(cols, function(cl)
    df[[cl]][df[[cl]] > 0])), na.rm = TRUE)
  min_nonzero <- max(min_nonzero, 1e-320)
  zero_repl   <- min_nonzero / factor

  df2 <- df
  for (cl in cols) df2[[paste0(cl, "_plot")]] <- ifelse(df2[[cl]] == 0, zero_repl, df2[[cl]])

  list(df = df2, zero_repl = zero_repl)
}

plot_pvals_three <- function(df, title = NULL, log_scale = TRUE,
                             facet_by = c("n_per_group","B","exceed0"),
                             facet_ncol = 3,
                             show_types = c("Mann-Whitney U","constrained","unconstrained"),
                             exceed0_order = c(100, 250, 0.1, 0.25),
                             exceed0_digits = 2,
                             zero_factor = 100,
                             y_breaks = c(1e-300, 1e-200, 1e-100, 1e-60, 1e-40, 1e-20, 1e-2)) {

  label_exceed0 <- function(x, digits = 2) {
    ifelse(x < 1,
           paste0(formatC(x, format = "f", digits = digits), "\u00D7B"),
           as.character(x))
  }

  z <- zero_safe_long(df, factor = zero_factor)
  df2 <- z$df
  zero_repl <- z$zero_repl

  df_long <- df2 %>%
    dplyr::transmute(
      obs_stats, effect_size,
      n_per_group, B, exceed0,
      `Mann-Whitney U` = p_U_asympt_plot,
      constrained      = p_constr_plot,
      unconstrained    = p_unconstr_plot
    ) %>%
    tidyr::pivot_longer(
      cols = c(`Mann-Whitney U`, constrained, unconstrained),
      names_to = "ptype", values_to = "pval"
    ) %>%
    dplyr::filter(ptype %in% show_types) %>%
    dplyr::mutate(
      ptype = factor(ptype, levels = c("Mann-Whitney U","constrained","unconstrained")),
      effect_size = factor(effect_size, levels = c(0,0.5,1,1.5,2)),
      exceed0_lab = label_exceed0(exceed0, digits = exceed0_digits),
      exceed0_lab = factor(exceed0_lab,
                           levels = label_exceed0(exceed0_order, digits = exceed0_digits))
    ) %>%
    arrange(ptype)   # draw U first, then constrained, then unconstrained

  p <- ggplot2::ggplot(df_long, ggplot2::aes(x = obs_stats, y = pval, color = ptype, size = ptype)) +
    ggplot2::geom_point() +
    ggplot2::scale_color_manual(
      name   = "P-value type:",
      values = c("Mann-Whitney U" = "black", 
                 "constrained"    = alpha("darkorange2", alpha = 0.6), 
                 "unconstrained"  = "#1f77b4")
    ) +
    ggplot2::scale_size_manual(values = c("Mann-Whitney U"=1.6,
                                          "constrained"    =1.6,
                                          "unconstrained"  =1),
                               guide = "none") +
    ggplot2::labs(
      title = title %||% "P-values vs. U_obs",
      x = expression(U[obs]),
      y = expression(p)
    ) +
    ggplot2::theme_minimal(base_size = 14) +
    ggplot2::theme(
      axis.text       = element_text(size = 14),
      axis.title      = element_text(size = 15),
      strip.text      = element_text(size = 15),
      legend.title    = element_text(size = 14),
      legend.text     = element_text(size = 14),
      legend.position = "bottom")

  sci02_with_zero <- function(zero) {
    function(x) {
      lab <- sprintf("%.0e", x)
      lab[x < zero*1.1] <- "0"
      lab
    }
  }

  if (log_scale) {
    p <- p + ggplot2::scale_y_log10(
      breaks = c(zero_repl, y_breaks),
      labels = sci02_with_zero(zero_repl),
      limits = c(zero_repl, 1),
      minor_breaks = NULL
    )
  }

  facet_by <- ifelse(facet_by == "exceed0", "exceed0_lab", facet_by)
  facet_by <- intersect(facet_by, c("n_per_group","B","exceed0_lab"))

  lab_eq <- labeller(
    B = function(x) paste0("B = ", x),
    exceed0_lab = function(x) paste0("exc0 = ", x),
    n_per_group = function(x) paste0("n = ", x)
  )

  if (length(facet_by) > 0) {
    p <- p + ggplot2::facet_wrap(as.formula(paste("~", paste(facet_by, collapse = " + "))),
                                 ncol = facet_ncol, scales = "free_x",
                                 labeller = lab_eq)
  }

  p
}
```

### n = 100

#### n=100, B=1000, varying exceed0

```{r pvals_Uobs_n100_B1000_vary_exc0, fig.width=10, fig.height=4, dpi = 200}
df_n100_B1k <- results_all %>% filter(n_per_group == 100, exceed0 %in% c(100, 250),
                                      B == 1000, gpd_con_fit)
plot_pvals_three(df_n100_B1k,
                 title = "n=100, B=1000, varying exceed0",
                 facet_by = "exceed0", exceed0_order = c(100, 250),
                 zero_factor = 1e20,
                 y_breaks = c(1e-60, 1e-40, 1e-20, 1e-2))
```

#### n=100, B=10000, varying exceed0

```{r pvals_Uobs_n100_B10000_vary_exc0, fig.width=10, fig.height=6, dpi = 200}
df_n100_B10k <- results_all %>% filter(n_per_group == 100, B == 10000, gpd_con_fit)
plot_pvals_three(df_n100_B10k,
                 title = "n=100, B=10000, varying exceed0",
                 facet_by = "exceed0", facet_ncol = 2,
                 zero_factor = 1e20,
                 exceed0_order = c(100, 250, 0.1, 0.25))
```

#### n=100, varying B and exceed0

```{r pvals_Uobs_n100_vary_B_exc0, fig.width=10, fig.height=8, dpi = 200}
df_n100_B1k10k <- rbind(df_n100_B1k, df_n100_B10k)
plot_pvals_three(df_n100_B1k10k,
                 title = "n=100, varying B and exceed0",
                 facet_by = c("B","exceed0"), facet_ncol = 2,
                 zero_factor = 1e20,
                 exceed0_order = c(100, 250, 0.1, 0.25))
```

### n = 500

#### n=500, B=1000, varying exceed0

```{r pvals_Uobs_n500_B1000_vary_exc0, fig.width=10, fig.height=4, dpi = 200}
df_n500_B1k <- results_all %>% filter(n_per_group == 500, exceed0 %in% c(100, 250),
                                      B == 1000, gpd_con_fit)
plot_pvals_three(df_n500_B1k,
                 title = "n=500, B=1000, varying exceed0",
                 facet_by = "exceed0", exceed0_order = c(100, 250),
                 y_breaks = c(1e-300, 1e-200, 1e-150, 1e-100, 1e-75, 1e-50, 1e-25, 1e-2),
                 zero_factor = 1e30)
```

#### n=500, B=10000, varying exceed0

```{r pvals_Uobs_n500_B10000_vary_exc0, fig.width=10, fig.height=6, dpi = 200}
df_n500_B10k <- results_all %>% filter(n_per_group == 500, B == 10000, gpd_con_fit)
plot_pvals_three(df_n500_B10k,
                 title = "n=500, B=10000, varying exceed0",
                 facet_by = "exceed0", facet_ncol = 2,
                 exceed0_order = c(100, 250, 0.1, 0.25),
                 y_breaks = c(1e-300, 1e-200, 1e-150, 1e-100, 1e-50, 1e-2),
                 zero_factor = 1e20)
```

#### n=500, varying B and exceed0

```{r pvals_Uobs_n500_vary_B_exc0, fig.width=10, fig.height=8, dpi = 200}
df_n500_B1k10k <- rbind(df_n500_B1k, df_n500_B10k)
plot_pvals_three(df_n500_B1k10k,
                 title = "n=500, varying B and exceed0",
                 facet_by = c("B","exceed0"), facet_ncol = 2,
                 exceed0_order = c(100, 250, 0.1, 0.25),
                 y_breaks = c(1e-300, 1e-200, 1e-150, 1e-100, 1e-50, 1e-2),
                 zero_factor = 1e5)
```

## Zero counts

```{r zero-counts}
zero_tbl <- results_all %>%
  summarize(
    zeros_U_asympt  = sum(p_U_asympt  == 0, na.rm = TRUE),
    zeros_unconstr  = sum(p_unconstr  == 0, na.rm = TRUE),
    zeros_constr    = sum(p_constr    == 0, na.rm = TRUE),
    .by = c(n_per_group, B, exceed0)
  ) %>%
  arrange(n_per_group, B, exceed0)

zero_tbl %>% knitr::kable()
```

## How often are the approximated p-values smaller?

```{r}
# Approximated vs. asymptotic U: how often are approximations smaller?
# (i.e., more optimistic than U-asymptotic)
cmp_tbl <- results_all %>%
  mutate(
    unconst_lt_U = p_unconstr < p_U_asympt,
    const_lt_U   = p_constr   < p_U_asympt
  ) %>%
  summarize(
    n_tests          = dplyr::n(),
    unconst_smaller  = sum(unconst_lt_U, na.rm = TRUE),
    const_smaller    = sum(const_lt_U,   na.rm = TRUE),
    unconst_share    = unconst_smaller / n_tests,
    const_share      = const_smaller   / n_tests,
    .by = c(n_per_group, B, exceed0)
  ) %>%
  arrange(n_per_group, B, exceed0) %>%
  mutate(
    unconst_share = scales::percent(unconst_share, accuracy = 0.1),
    const_share   = scales::percent(const_share,   accuracy = 0.1)
  )

# Pretty table
cmp_tbl %>%
  dplyr::rename(
    n      = n_per_group,
    B_perm = B,
    exc0   = exceed0,
    `# smaller (unconstr)` = unconst_smaller,
    `# smaller (constr)`   = const_smaller,
    `% smaller (unconstr)`= unconst_share,
    `% smaller (constr)`   = const_share
  ) %>%
  knitr::kable(caption = "Counts and shares of approximated p-values smaller than Mann-Whitney U p-values.")

```

## Files written

```{r list-files}
projthis::proj_dir_info(path_target())
```
