---
title: "Generate GPD data for simulation studies"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "01_data_gen" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "bd977ab5-754a-40a8-a775-cdac50375411")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(fig.width = 10, 
                      fig.height = 5, 
                      dpi = 200,
                      fig.path = paste0('figures/', params, "/"),
                      echo = FALSE)
```

This script generates random samples from the Generalized Pareto Distribution (GPD).

## Load packages

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)
library(kableExtra)
library(eva)
library(furrr)
library(progressr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Data generation for unconstrained study

We draw random samples from the GPD distribution. The following parameter settings are considered:

```{r}
shapes <- c(-0.6, -0.4, -0.2, 0, 0.2, 0.4, 0.6)
scale <- 1
sample_sizes <- c(25, 50, 75, 100, 250, 500, 750, 1000)
num_reps <- 500

cat("Shape values: ", shapes,
    "\nScale: ", scale,
    "\nSample sizes: ", sample_sizes,
    "\nRepetitions: ", num_reps)
```

```{r}
# Create directories for saving data
data_dir <- file.path(path_target())
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r, eval=FALSE}
# Generate seeds
set.seed(123456)
seeds <- sample(10000000:99999999, num_reps*length(shapes))
stopifnot(!anyDuplicated(seeds))

# Simulate and save GPD data
for(shape_idx in seq_along(shapes)){
  
  # Current seed
  seeds_tmp <- seeds[(1:num_reps)+(num_reps*(shape_idx-1))]
  
  # Current shape
  shape <- shapes[shape_idx]
  
  for (n in sample_sizes) {
    result_file <- file.path(data_dir, paste0("shape-", shape, "_n-", n, ".RDS"))
    
    if (!file.exists(result_file)) {
          sim_data <- lapply(
      seq_len(num_reps), function(r) {
        set.seed(seeds_tmp[r])
        eva::rgpd(n = n, scale = scale, shape = shape)
      })
    
    # Save each combination as a separate file
    saveRDS(sim_data, file = result_file)
    }
  }
}
```

## Data generation for boundary misspecification study

```{r}
true_boundaries <- seq(6, 2, by = -0.5)  # from edge case to strong misspecification
shapes <- (-scale / true_boundaries)
scale <- 1
boundaries <- -scale / shapes
sample_sizes <- c(1000)
num_reps <- 500

cat("\nBoundaries: ", boundaries,
    "\nShape values: ", shapes,
    "\nScale: ", scale,
    "\nSample sizes: ", sample_sizes,
    "\nRepetitions: ", num_reps)
```

```{r}
# Create directories for saving data
data_dir <- file.path(path_target())
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r, eval=FALSE}
# Generate seeds
set.seed(123456)
seeds <- sample(10000000:99999999, num_reps*length(shapes))
stopifnot(!anyDuplicated(seeds))

# Simulate and save GPD data
for(bound_idx in seq_along(true_boundaries)){
  
  # Current seed
  seeds_tmp <- seeds[(1:num_reps)+(num_reps*(shape_idx-1))]
  
  # Boundary and shape
  bound <- true_boundaries[bound_idx]
  shape <- -scale / bound
  
  for (n in sample_sizes) {
    result_file <- file.path(data_dir, paste0("bound-", bound, "_n-", n, ".RDS"))
    
    if (!file.exists(result_file)) {
      sim_data <- lapply(
        seq_len(num_reps), function(r) {
          set.seed(seeds_tmp[r])
          eva::rgpd(n = n, scale = scale, shape = shape)
        })
      
      # Save each combination as a separate file
      saveRDS(
        sim_data,
        file = result_file
      )
    }
  }
}
```

## Data generation for asymptotic behavior study

```{r}
shapes <- -0.2
scale <- 1
sample_sizes <- c(50, 100, 200, 500, 1000, 5000, 10000, 50000)
num_reps <- 500

cat("Shape: ", shapes,
    "\nScale: ", scale,
    "\nSample sizes: ", sample_sizes,
    "\nRepetitions: ", num_reps)
```

```{r}
# Create directories for saving data
data_dir <- file.path(path_target())
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r, eval=FALSE}
# Generate seeds
set.seed(123456)
seeds <- sample(10000000:99999999, num_reps*length(shapes))
stopifnot(!anyDuplicated(seeds))

# Simulate and save GPD data
for(shape_idx in seq_along(shapes)){
  
  # Current seed
  seeds_tmp <- seeds[(1:num_reps)+(num_reps*(shape_idx-1))]
  
  # Current shape
  shape <- shapes[shape_idx]
  
  for (n in sample_sizes) {
    result_file <- file.path(data_dir, paste0("shape-", shape, "_n-", n, ".RDS"))
    
    if (!file.exists(result_file)) {
      sim_data <- lapply(
        seq_len(num_reps), function(r) {
          set.seed(seeds_tmp[r])
          eva::rgpd(n = n, scale = scale, shape = shape)
        })
      
      # Save each combination as a separate file
      saveRDS(
        sim_data,
        file = result_file
      )
    }
  }
}
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
