---
title: "Simulation study: Asymptotic behavior of constrained GPD estimation under boundary misspecification"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "03_asymptotic" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "cce9dcb1-e519-4e92-89e6-f629fd4cb7bc")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 5, 
  fig.height = 3, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/sim_GPD_asymp_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

This simulation study investigates the behavior of constrained GPD parameter estimation with a fixed shape parameter and a fixed support constraint $s_c$. As the sample size increases, the maximum observed value approaches the true boundary, reducing the degree of misspecification induced by the constraint. We analyze how this convergence affects estimation accuracy across different GPD fitting methods.


```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)
library(kableExtra)
library(eva)
library(furrr)
library(progressr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r, message=FALSE, warning=FALSE}
# Load permApprox functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(permApprox_functions, source))
```

## Define parameters

```{r}
constraint <- 5
scale <- 1
shape <- -0.2
true_bound <- -scale / shape
sample_sizes <- c(50, 100, 200, 500, 1000, 5000, 10000, 50000)
num_reps <- 500 # Number of repetitions
method <- "LME"
cores <- 8
set.seed(123456)

cat("Shape:", shape,
    "\nScale: ", scale,
    "\nBoundary: ", true_bound,
    "\nSupport constraint: ", constraint,
    "\nMethod: ", method,
    "\nSample sizes: ", sample_sizes,
    "\nRepetitions: ", num_reps)
```

## Setting

```{r}
estimation_settings <- tidyr::crossing(
  n = sample_sizes
) %>%
  mutate(
    shape = shape,
    scale = scale,
    bound = true_bound,
    constraint = constraint
  )

estimation_settings %>% head()
cat("Number of settings: ", nrow(estimation_settings))
```


## Perform estimations

```{r}
# Data directory
data_dir <- file.path(path_source("01_data_gen"))

# Directory setup for saving results
results_dir <- file.path(path_target())
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r, eval=FALSE}
cores <- length(sample_sizes)

# Set up progress handler
handlers("txtprogressbar")

# Set number of cores
if (parallel::detectCores() < cores) {
  cores <- parallel::detectCores() - 2
}

# Set up parallel plan
plan(multisession, workers = cores)

# Define estimation function
estimate_one_setting <- function(i) {
  
  # Extract setting
  setting <- estimation_settings[i, ]
  n <- setting$n
  data_shape <- setting$shape
  data_scale <- setting$scale
  data_bound <- setting$bound
  constraint <- setting$constraint
  
  # Load data
  data_file <- file.path(data_dir, 
                         paste0("shape-", shape, "_n-", n, ".RDS"))
  gpd_samples <- readRDS(data_file)
  
  # Construct output file path
  result_file <- file.path(results_dir, 
                           paste0("shape-", shape,
                                  "_n-", n,
                                  ".RDS"))
  
  if (!file.exists(result_file)) {
    # Estimate parameters
    fit_results <- purrr::imap_dfr(gpd_samples, ~ {
      data <- .x
      
      data_max <- max(data)
      
      # Unconstrained fit
      fit_uc <- fit_gpd(
        data = data, 
        thresh = 0,
        fit_method = method,
        tol = 1e-7, 
        epsilon = 0,
        constraint = "unconstrained",
        support_boundary = NULL
      )
      
      if (fit_uc$shape < 0) {
        bound_uc <- -fit_uc$scale / fit_uc$shape
        dgpd_uc <- POT::dgpd(constraint, 
                             scale = fit_uc$scale, 
                             shape = fit_uc$shape)
      } else {
        bound_uc <- dgpd_uc <- NA
      }
      
      constr_active <- bound_uc < constraint
      
      # Constrained fit
      fit_c <- fit_gpd(
        data = data, 
        thresh = 0,
        fit_method = method,
        tol = 1e-7, 
        epsilon = 0,
        constraint = "support_at_obs",
        support_boundary = constraint
      )
      
      if (fit_c$shape < 0) {
        bound_c <- -fit_c$scale / fit_c$shape
        dgpd_c <- POT::dgpd(constraint, 
                            scale = fit_c$scale, 
                            shape = fit_c$shape)
      } else {
        bound_c <- dgpd_c <- NA
      }
      
      tibble(constraint = constraint,
             data_bound = true_bound,
             data_shape = shape,
             data_scale = scale,
             n = n,
             method = method,
             data_max = data_max,
             constr_active = constr_active,
             shape_uc = fit_uc$shape,
             scale_uc = fit_uc$scale,
             bound_uc = bound_uc,
             dgpd_uc = dgpd_uc,
             shape_c = fit_c$shape,
             scale_c = fit_c$scale,
             bound_c = bound_c,
             dgpd_c = dgpd_c,
             repetition = .y)
    })
    
    # Save results
    saveRDS(fit_results, file = result_file)
  }
}

# Run estimation in parallel with progress
with_progress({
  pb <- progressor(steps = nrow(estimation_settings))
  
  gpdfitlist <- future_map(seq_len(nrow(estimation_settings)), function(i) {
    pb()
    estimate_one_setting(i)
  })
})

# Optionally stop the plan (not necessary in most cases)
plan(sequential)
```

## Combine results

```{r}
# List and load all result files
all_result_files <- list.files(results_dir, full.names = TRUE)

gpdfitlist <- purrr::map_dfr(all_result_files, ~ {
  res <- readRDS(.x)
  res %>% mutate(across(where(is.list), ~ unlist(base::unname(.))))
}) %>%
  dplyr::filter(n != 200)
```

## Results table

```{r}
gpdfitlist[1:20, ] %>%
  kable
```


## Plots

```{r}
# Compute errors

error_df <- gpdfitlist %>%
  pivot_longer(
    cols = c(shape_uc, shape_c, scale_uc, scale_c, bound_uc, bound_c),
    names_to = c("param", "fit_type"),
    names_pattern = "(.*)_(uc|c)",
    values_to = "estimate"
  ) %>%
  mutate(
    fit_type = recode(fit_type,
                      "uc" = "Unconstrained",
                      "c"  = "Constrained"),
    true_value = case_when(
      param == "shape" ~ data_shape,
      param == "scale" ~ data_scale,
      param == "bound" ~ data_bound
    ),
    error = estimate - true_value,
    abs_error = abs(error)
  )

```


### Shape error

Difference between the estimated shape $\hat{\xi}$ and the true shape used for data generation $\xi_0$.

```{r error_shape_boxplots}
error_df %>%
  dplyr::filter(param == "shape") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(xi) - xi[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Scale error

Difference between the estimated scale $\hat{\sigma}$ and the true scale used for data generation $\sigma_0$.

```{r error_scale_boxplots}
error_df %>%
  dplyr::filter(param == "scale") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(sigma) - sigma[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Boundary error

Difference between the estimated support boundary $\hat{s} = \frac{-\hat{\sigma}}{\hat{\xi}}$ and the boundary that results from the true parameters used for data generation $s_0=\frac{-\sigma_0}{\xi_0}$.

```{r error_bound_boxplots}
error_df %>%
  dplyr::filter(param == "bound") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(s) - s[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Boundary error (log scale)

Same as before but on log scale.

```{r error_bound_log_boxplots}
error_df %>%
  dplyr::filter(param == "bound") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(s) - s[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_y_log10()
```


### Difference between boundary and constraint

Difference between the estimated boundary $\hat{s} = \frac{-\hat{\sigma}}{\hat{\xi}}$ and the support constraint (here **5**).

```{r diff_bound_constraint_boxplots}
delta_df <- gpdfitlist %>%
  select(method, data_bound, constraint, bound_uc, bound_c, constr_active, n) %>%
  pivot_longer(cols = c(bound_uc, bound_c),
               names_to = "fit_type",
               values_to = "estimated_bound") %>%
  mutate(
    fit_type = recode(fit_type,
                      "bound_uc" = "Unconstrained",
                      "bound_c"  = "Constrained"),
    delta = estimated_bound - constraint
  ) %>% 
  dplyr::filter(data_bound <= 5)

p <- ggplot(delta_df, aes(x = factor(n), y = delta, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    title = expression("Difference between estimated boundary and constraint"),
    x = "n",
    y = expression(hat(s) - s[c]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50")

print(p)
```



## Plots (constraint active)

In the following plots the results are filtered so that only replicates are used where the constraint is active. 

The constraint is active if the unconstraint fit would lead to a boundary below the constraint. 

```{r}
# Compute errors

error_df <- gpdfitlist %>%
  dplyr::filter(constr_active == TRUE) %>%
  pivot_longer(
    cols = c(shape_uc, shape_c, scale_uc, scale_c, bound_uc, bound_c),
    names_to = c("param", "fit_type"),
    names_pattern = "(.*)_(uc|c)",
    values_to = "estimate"
  ) %>%
  mutate(
    fit_type = recode(fit_type,
                      "uc" = "Unconstrained",
                      "c"  = "Constrained"),
    true_value = case_when(
      param == "shape" ~ data_shape,
      param == "scale" ~ data_scale,
      param == "bound" ~ data_bound
    ),
    error = estimate - true_value,
    abs_error = abs(error)
  )

```


### Shape error

Difference between the estimated shape $\hat{\xi}$ and the true shape used for data generation $\xi_0$.

```{r error_shape_active_boxplots}
error_df %>%
  dplyr::filter(param == "shape") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(xi) - xi[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Scale error

Difference between the estimated scale $\hat{\sigma}$ and the true scale used for data generation $\sigma_0$.

```{r error_scale_active_boxplots}
error_df %>%
  dplyr::filter(param == "scale") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(sigma) - sigma[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Boundary error

Difference between the estimated boundary $\hat{s} = \frac{-\hat{\sigma}}{\hat{\xi}}$ and the boundary that results from the true parameters used for data generation $s_0 = \frac{-\sigma_0}{\xi_0}$.

```{r error_bound_active_boxplots}
error_df %>%
  dplyr::filter(param == "bound") %>%
  ggplot(aes(x = factor(n), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = "n",
    y = expression(hat(s) - s[0]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Difference between boundary and constraint

Difference between the estimated boundary $\hat{s} = \frac{-\hat{\sigma}}{\hat{\xi}}$ and the support constraint (here **5**).

```{r diff_bound_constraint_active_boxplots}
delta_df <- gpdfitlist %>%
  select(method, data_bound, constraint, bound_uc, bound_c, constr_active, n) %>%
  pivot_longer(cols = c(bound_uc, bound_c),
               names_to = "fit_type",
               values_to = "estimated_bound") %>%
  mutate(
    fit_type = recode(fit_type,
                      "bound_uc" = "Unconstrained",
                      "bound_c"  = "Constrained"),
    delta = estimated_bound - constraint
  ) %>% 
  dplyr::filter(data_bound <= 5)

p <- ggplot(delta_df, aes(x = factor(n), y = delta, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    #title = expression("Difference between estimated boundary and constraint"),
    x = "n",
    y = expression(hat(s) - s[c]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50")

print(p)
```

```{r diff_bound_constraint_active_boxplots_ylim}
p + coord_cartesian(ylim = c(-3, 10))
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
