---
title: "Simulation study: Robustness of GPD estimation under boundary misspecification"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "04_misspecification"
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "1e46eb1d-6a01-485d-aa23-9c71660a8549")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 7, 
  fig.height = 5, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/sim_GPD_misspec_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

We evaluate how GPD parameter estimation methods behave under varying degrees of misspecification with respect to a boundary constraint. We simulate data where the true upper boundary $s_0$ of the GPD is below a fixed constraint point $s_c$ and compare constrained and unconstrained fits.

```{r, eval=FALSE}
devtools::install_github("stefpeschel/permApprox")
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)
library(kableExtra)
library(eva)
library(furrr)
library(progressr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r, message=FALSE, warning=FALSE}
# Load permApprox functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(permApprox_functions, source))
```

## Example cases

We firstly look at three example cases: 

* Well specified model: true boundary > constraint (constraint inactive)
* Edge case: true boundary = constraint
* Misspecified model: true boundary < constraint (constraint always active)

```{r}
# Plot function
plot_example <- function(true_bound, scale = 1, n = 1000, c = 5, method = "MLE1D", rep = 1) {
  shape <- -scale / true_bound
  
  cat("Shape: ", shape)
  cat("\nConstraint: ", c)
  cat("\nTrue boundary: ", true_bound)
  
  tol <- case_when(
    method == "LME" ~ 1e-7,
    method == "MLE1D" ~ 1e-10,
    method == "MLE2D" ~ 1e-10,
    method == "NLS2" ~ 1e-9,
    method == "WNLLSM" ~1e-9,
    TRUE ~ NA_real_
  )
  
  # Load data
  data_file <- file.path(path_source("01_data_gen"), paste0("bound-", true_bound, "_n-", n, ".RDS"))
  gpd_samples <- readRDS(data_file)
  data <- gpd_samples[[rep]]
  
  cat("\nData maximumum: ", max(data))
  
  # Unconstrained GPD fit
  fit_uc <- fit_gpd(
    data = data, 
    thresh = 0,
    fit_method = method,
    tol = tol, 
    epsilon = 0,
    constraint = "unconstrained",
    support_boundary = NULL
  )
  
  bound_uc <- -fit_uc$scale / fit_uc$shape
  
  # Constrained GPD fit
  fit_c <- fit_gpd(
    data = data, 
    thresh = 0,
    fit_method = method,
    tol = tol, 
    epsilon = 0,
    constraint = "support_at_obs",
    support_boundary = c
  )
  
  bound_c <- -fit_c$scale / fit_c$shape
  
  # Create x values to evaluate densities on
  x_vals <- seq(min(data), max(data), length.out = 1000)
  
  # Evaluate densities
  gpd_density_uc <- eva::dgpd(x_vals, scale = fit_uc$scale, shape = fit_uc$shape)
  gpd_density_c  <- eva::dgpd(x_vals, scale = fit_c$scale, shape = fit_c$shape)
  
  # Combine into a data frame
  df_plot <- data.frame(
    x = x_vals,
    dens_uc = gpd_density_uc,
    dens_c  = gpd_density_c
  )
  
  # Combine densities into long format for plotting with ggplot
  df_plot_long <- df_plot %>%
    pivot_longer(cols = c("dens_uc", "dens_c"),
                 names_to = "fit_type",
                 values_to = "density") %>%
    mutate(fit_type = recode(fit_type,
                             "dens_uc" = "Unconstrained fit",
                             "dens_c" = "Constrained fit"))
  
  # Create a data frame for boundary lines
  bound_df <- data.frame(
    xintercept = c(true_bound, bound_uc, bound_c),
    fit_type = factor(c("True boundary", "Unconstrained fit", "Constrained fit"),
                      levels = c("True boundary", "Unconstrained fit", "Constrained fit"))
  )
  
  # Plot
  p <- ggplot() +
    geom_histogram(aes(x = data, y = after_stat(density)),
                   bins = 30, fill = "lightblue", color = "black", alpha = 0.6) +
    geom_line(data = df_plot_long, aes(x = x, y = density, color = fit_type), linewidth = 0.6) +
    geom_vline(data = bound_df, aes(xintercept = xintercept, color = fit_type, linetype = fit_type),
               linewidth = 0.7, alpha = 0.5) +
    scale_color_manual(values = c("True boundary" = "darkgray", 
                                  "Unconstrained fit" = "blue", 
                                  "Constrained fit" = "red")) +
    scale_linetype_manual(values = c("True boundary" = "solid", 
                                     "Unconstrained fit" = "dotted", 
                                     "Constrained fit" = "dashed")) +
    labs(
      x = "x",
      y = "Density",
      color = "Fit type",
      linetype = "Support boundary",
      title = paste0("Fitted GPD with bound = ", true_bound,
                    ", shape = ", round(shape, 3),
                    ", and scale = ", round(scale, 3))
    ) +
    theme_minimal()
  
  cat("\nUnconstrained boundary:", bound_uc)
  cat("\nConstrained boundary:", bound_c)
  
  p
}

```

### Well specified case (boundary > cutoff)

```{r}
print(plot_example(true_bound = 6))
```


### Edge case (boundary = cutoff)

```{r}
print(plot_example(true_bound = 5, rep = 2))
```

### Misspecification (boundary < cutoff)

```{r}
print(plot_example(true_bound = 4))
```


## Define parameters

```{r}
constraint <- 5
scale <- 1
true_boundaries <- seq(6, 2, by = -0.5)  # from edge case to strong misspecification
shapes <- -scale / true_boundaries
sample_sizes <- 1000
num_reps <- 500 # Number of repetitions
cores <- 8

cat("True boundaries:", true_boundaries,
    "\nShape values:", shapes,
    "\nScale: ", scale,
    "\nConstraint: ", constraint,
    "\nSample sizes: ", sample_sizes,
    "\nRepetitions: ", num_reps)
```

The GPD parameters are estimated for all considered parameter combinations using the following estimation methods:

* LME (Likelihood moment estimator) by Zhang (2007)
* WNLLSM (Weighted Nonlinear Least Squares Likelihood Moments Estimation) by Zhao et al. (2019)
* MLE-1D (one-dimensional MLE) by Castillo & Serra (2015)
* MLE-2D (Maximum Likelihood Estimation)
* NLS2 (Quantile Estimation) by Song & Song (2012)
* ZSE (Zhang Stephens Estimator; Bayesian methods) by Zhang & Stephens (2009)

```{r}
# Methods for parameter fitting
fit_methods <- c("MOM", "PWM", "LME", "MLE1D", "MLE2D", "NLS2", "WNLLSM", "ZSE")
fit_methods <- factor(fit_methods, levels = fit_methods)
```

## Settings

```{r}
# Create combinations of shape, sample size, and method
estimation_settings <- tidyr::crossing(bound = true_boundaries,
                                       n = sample_sizes, 
                                       method = fit_methods)

estimation_settings %>% head()
cat("Number of settings: ", nrow(estimation_settings))
```


## Perform estimations

Constrained and unconstrained fitting for all settings ...

```{r}
# Data directory
data_dir <- file.path(path_source("01_data_gen"))

# Directory setup for saving results
results_dir <- file.path(path_target())
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r, eval=FALSE}
# Set up progress handler
handlers("txtprogressbar")

# Set number of cores
if (parallel::detectCores() < cores) {
  cores <- parallel::detectCores() - 2
}

# Set up parallel plan
plan(multisession, workers = cores)

# Define estimation function
estimate_one_setting <- function(i) {
  
  # Extract setting
  setting <- estimation_settings[i, ]
  true_bound <- setting$bound
  n <- setting$n
  method <- as.character(setting$method)
  shape <- -scale / true_bound
  
  tol <- case_when(
    method == "LME" ~ 1e-7,
    method == "MLE1D" ~ 1e-10,
    method == "MLE2D" ~ 1e-10,
    method == "NLS2" ~ 1e-9,
    method == "WNLLSM" ~ 1e-9,
    TRUE ~ NA_real_
  )
  
  # Load data
  data_file <- file.path(data_dir, 
                         paste0("bound-", true_bound, "_n-", n, ".RDS"))
  gpd_samples <- readRDS(data_file)
  
  # Construct output file path
  result_file <- file.path(results_dir, 
                           paste0("bound-", true_bound,
                                  "_n-", n,
                                  "_method-", method, 
                                  ".RDS"))
  
  if (!file.exists(result_file)) {
    # Estimate parameters
    fit_results <- purrr::imap_dfr(gpd_samples, ~ {
      data <- .x
      
      data_max <- max(data)
      constr_active <- data_max < constraint
      
      # Unconstrained fit
      fit_uc <- fit_gpd(
        data = data, 
        thresh = 0,
        fit_method = method,
        tol = tol, 
        epsilon = 0,
        constraint = "unconstrained",
        support_boundary = NULL
      )
      
      if (fit_uc$shape < 0) {
        bound_uc <- -fit_uc$scale / fit_uc$shape
        dgpd_uc <- POT::dgpd(constraint, 
                             scale = fit_uc$scale, 
                             shape = fit_uc$shape)
      } else {
        bound_uc <- dgpd_uc <- NA
      }
      
      # Constrained fit
      fit_c <- fit_gpd(
        data = data, 
        thresh = 0,
        fit_method = method,
        tol = tol, 
        epsilon = 0,
        constraint = "support_at_obs",
        support_boundary = constraint
      )
      
      if (fit_c$shape < 0) {
        bound_c <- -fit_c$scale / fit_c$shape
        dgpd_c <- POT::dgpd(constraint, 
                            scale = fit_c$scale, 
                            shape = fit_c$shape)
      } else {
        bound_c <- dgpd_c <- NA
      }
      
      tibble(constraint = constraint,
             data_bound = true_bound,
             data_shape = shape,
             data_scale = scale,
             n = n,
             method = method,
             data_max = data_max,
             constr_active = constr_active,
             shape_uc = fit_uc$shape,
             scale_uc = fit_uc$scale,
             bound_uc = bound_uc,
             dgpd_uc = dgpd_uc,
             shape_c = fit_c$shape,
             scale_c = fit_c$scale,
             bound_c = bound_c,
             dgpd_c = dgpd_c,
             repetition = .y)
    })
    
    # Save results
    saveRDS(fit_results, file = result_file)
  }
}

# Run estimation in parallel with progress
with_progress({
  pb <- progressor(steps = nrow(estimation_settings))
  
  gpdfitlist <- future_map(seq_len(nrow(estimation_settings)), function(i) {
    pb()
    estimate_one_setting(i)
  })
})

# Optionally stop the plan (not necessary in most cases)
plan(sequential)
```

```{r}
## Combine results

# List and load all result files
all_result_files <- list.files(results_dir, full.names = TRUE)

gpdfitlist <- purrr::map_dfr(all_result_files, ~ {
  res <- readRDS(.x)
  res %>% mutate(
    across(where(is.list), ~ unlist(unname(.))),
    constraint = NULL)
})

# Determine if constraint is active
gpdfitlist <- gpdfitlist %>%
  mutate(constr_active = bound_uc < constraint) %>%
  dplyr::filter(data_bound <= 5)

# Remove two outliers
#sum(gpdfitlist$bound_c > 500)
#sum(gpdfitlist$bound_uc > 500)

gpdfitlist <- gpdfitlist %>%
  dplyr::filter(bound_c < 500, bound_uc < 500,
                !method %in% c("MOM", "PWM"))
```

## Results table

```{r}
gpdfitlist[1:20, ] %>%
  kable
```


## Plots

```{r}
# Plot parameters
axis_text_size <- 12
axis_title_size <- 16
legend_text_size <- 12
legend_title_size <- 16
title_size <- 16
strip_text <- 14
```


We plot the error in shape, scale, boundary, and the difference between the estimated boundary and the constraint value. The true boundary $b_0$ is always on the x axis.

We expect the error to decrease as the boundary approaches the constraint value (5 in this case).

```{r}
# Compute errors

error_df <- gpdfitlist %>%
  pivot_longer(
    cols = c(shape_uc, shape_c, scale_uc, scale_c, bound_uc, bound_c),
    names_to = c("param", "fit_type"),
    names_pattern = "(.*)_(uc|c)",
    values_to = "estimate"
  ) %>%
  mutate(
    fit_type = recode(fit_type,
                      "uc" = "Unconstrained",
                      "c"  = "Constrained"),
    true_value = case_when(
      param == "shape" ~ data_shape,
      param == "scale" ~ data_scale,
      param == "bound" ~ data_bound
    ),
    error = estimate - true_value,
    abs_error = abs(error)
  )

```


### Shape error

Difference between the estimated shape $\hat{\xi}$ and the true shape used for data generation $\xi_0$.

```{r error_shape_boxplots}
error_df %>%
  dplyr::filter(param == "shape") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(xi) - xi[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

```

### Scale error

Difference between the estimated scale $\hat{\sigma}$ and the true scale used for data generation $\sigma_0$.

```{r error_scale_boxplots}
error_df %>%
  dplyr::filter(param == "scale") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(sigma) - sigma[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

```

### Boundary error

Difference between the estimated boundary $\frac{-\hat{\sigma}}{\hat{\xi}}$ and the boundary that results from the true parameters used for data generation $\frac{-\sigma_0}{\xi_0}$.

```{r error_bound_boxplots}
p_bound_box <- error_df %>%
  dplyr::filter(param == "bound") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(s) - s[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")
p_bound_box
```

```{r error_bound_jitter_constr}
p_bound_jitter <- error_df %>%
  dplyr::filter(param == "bound", fit_type == "Constrained",
                !is.na(constr_active)) %>%
  ggplot(aes(x = factor(data_bound), y = error)) +
  geom_jitter(aes(color = constr_active),
              width = 0.2, alpha = 0.7, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50") +
  facet_wrap(~ method) +
  scale_color_manual(
    values = c("FALSE" = "#1F78B4", "TRUE" = "#FF7F00"),
    labels = c("FALSE" = "No", "TRUE" = "Yes")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(s) - s[0]),
    color = "Constraint active",
    shape = "Fit type",
    title = "Constrained fit"
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 2))
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

p_bound_jitter
```


### Boundary error (set y-axis limits)

```{r error_bound_boxplots_ylim}
p_bound_box + coord_cartesian(ylim = c(-1, 10))

```

```{r error_bound_jitter_constr_ylim}
p_bound_jitter + coord_cartesian(ylim = c(0, 10))

```

### Difference between boundary and constraint

Difference between the estimated boundary $\frac{-\hat{\sigma}}{\hat{\xi}}$ and the constraint (here **5**).

```{r diff_bound_constraint_boxplots}
delta_df <- gpdfitlist %>%
  select(method, data_bound, constraint, bound_uc, bound_c, constr_active) %>%
  pivot_longer(cols = c(bound_uc, bound_c),
               names_to = "fit_type",
               values_to = "estimated_bound") %>%
  mutate(
    fit_type = recode(fit_type,
                      "bound_uc" = "Unconstrained",
                      "bound_c"  = "Constrained"),
    delta = estimated_bound - constraint
  ) %>% 
  dplyr::filter(data_bound <= 5)

p <- ggplot(delta_df, aes(x = factor(data_bound), y = delta, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  facet_wrap(~ method) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    #title = expression("Difference between estimated boundary and constraint value"),
    x = expression(s[0] * " = " * -sigma[0]/xi[0]),
    y = expression(hat(s) - s[c]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50")

print(p)
```

```{r diff_bound_constraint_boxplots_ylim}
p + coord_cartesian(ylim = c(-4, 6))
```


```{r diff_bound_constraint_jitter_constr}
p <- delta_df %>%
  dplyr::filter(fit_type == "Constrained",
                !is.na(constr_active)) %>%
  ggplot(aes(x = factor(data_bound), y = delta)) +
  geom_jitter(aes(color = constr_active),
              width = 0.2, alpha = 0.7, size = 0.7) +
  facet_wrap(~ method) +
  scale_color_manual(
    values = c("FALSE" = "#1F78B4", "TRUE" = "#FF7F00"),
    labels = c("FALSE" = "No", "TRUE" = "Yes")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(s) - s[c]),
    color = "Constraint active",
    shape = "Fit type",
    title = "Constrained fit"
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 2))
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")
p
```
```{r diff_bound_constraint_jitter_constr_ylim}
p + coord_cartesian(ylim = c(0, 6))
```

## Plots (constraint active)

Same plots as before, but only replicates with active constraint are considered.

```{r}
# Compute errors

error_df <- gpdfitlist %>%
  dplyr::filter(constr_active == TRUE) %>%
  pivot_longer(
    cols = c(shape_uc, shape_c, scale_uc, scale_c, bound_uc, bound_c),
    names_to = c("param", "fit_type"),
    names_pattern = "(.*)_(uc|c)",
    values_to = "estimate"
  ) %>%
  mutate(
    fit_type = recode(fit_type,
                      "uc" = "Unconstrained",
                      "c"  = "Constrained"),
    true_value = case_when(
      param == "shape" ~ data_shape,
      param == "scale" ~ data_scale,
      param == "bound" ~ data_bound
    ),
    error = estimate - true_value,
    abs_error = abs(error)
  )

```

### Shape error

Difference between the estimated shape $\hat{\xi}$ and the true shape used for data generation $\xi_0$.

```{r error_shape_boxplots_active}
error_df %>%
  dplyr::filter(param == "shape") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(xi) - xi[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

```


### Scale error

Difference between the estimated scale $\hat{\sigma}$ and the true scale used for data generation $\sigma_0$.

```{r error_scale_boxplots_active}
error_df %>%
  dplyr::filter(param == "scale") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(sigma) - sigma[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

```

### Boundary error

Difference between the estimated boundary $\frac{-\hat{\sigma}}{\hat{\xi}}$ and the boundary that results from the true parameters used for data generation $\frac{-\sigma_0}{\xi_0}$.

```{r error_bound_boxplots_active}
p_bound_box <- error_df %>%
  dplyr::filter(param == "bound") %>%
  ggplot(aes(x = factor(data_bound), y = error, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    x = expression(s[0] *"=" * -sigma[0] / xi[0]),
    y = expression(hat(s) - s[0]),
    fill = "Fit type"
  ) +
  facet_wrap(~ method) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")
p_bound_box
```

### Difference between boundary and constraint

Difference between the estimated boundary $\frac{-\hat{\sigma}}{\hat{\xi}}$ and the constraint (here **5**).

```{r diff_bound_constraint_boxplots_active}
delta_df <- gpdfitlist %>%
  dplyr::filter(constr_active == TRUE) %>%
  select(method, data_bound, constraint, bound_uc, bound_c, constr_active) %>%
  pivot_longer(cols = c(bound_uc, bound_c),
               names_to = "fit_type",
               values_to = "estimated_bound") %>%
  mutate(
    fit_type = recode(fit_type,
                      "bound_uc" = "Unconstrained",
                      "bound_c"  = "Constrained"),
    delta = estimated_bound - constraint
  ) %>% 
  dplyr::filter(data_bound <= 5)

p <- ggplot(delta_df, aes(x = factor(data_bound), y = delta, fill = fit_type)) +
  geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  facet_wrap(~ method) +
  scale_fill_manual(
    values = c("Unconstrained" = "#1F78B4", "Constrained" = "#FF7F00")
  ) +
  labs(
    #title = expression("Difference between estimated boundary and constraint"),
    x = expression(s[0] * " = " * -sigma[0]/xi[0]),
    y = expression(hat(s) - s[c]),
    fill = "Fit type"
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50")

print(p)
```

## Constraint validation

```{r}
gpdfitlist <- gpdfitlist %>%
  mutate(
    valid_uc = ifelse(shape_uc < 0, bound_uc > constraint, NA),
    valid_c  = ifelse(shape_c  < 0, bound_c  > constraint, NA)
  )
```

### Constraint satisfatcion rate

```{r}
cat("Number of replicates where constraint is not satisfied: ", sum(!gpdfitlist$valid_c))
```

```{r}
gpdfitlist %>%
  pivot_longer(cols = c(valid_uc, valid_c),
               names_to = "fit_type",
               values_to = "constraint_held") %>%
  mutate(fit_type = recode(fit_type,
                           valid_uc = "Unconstrained",
                           valid_c = "Constrained")) %>%
  group_by(method, data_bound, fit_type) %>%
  summarise(valid_constr_rate = mean(constraint_held, na.rm = TRUE), .groups = "drop") %>%
  dplyr::filter(fit_type == "Constrained") %>%
  select(method, data_bound, valid_constr_rate)

```

## GPD density at constraint point

```{r density_cutoff_boxplots}
dgpd_df <- gpdfitlist %>%
  dplyr::filter(!is.na(dgpd_c), data_bound <= 5) %>%
  select(method, data_bound, dgpd_c, constr_active)


ggplot(dgpd_df, aes(x = factor(data_bound), y = dgpd_c, fill = constr_active)) +
  geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  facet_wrap(~ method) +
  scale_y_log10() +
  scale_fill_manual(
    values = c("FALSE" = "#1F78B4", "TRUE" = "#FF7F00"),
    labels = c("FALSE" = "No", "TRUE" = "Yes")
  ) +
  labs(
    #title = expression("Estimated density at cutoff under constrained fit"),
    x = expression(s[0] * " = " * -sigma[0]/xi[0]),
    y = expression(f~(s[0])),
    fill = "Constraint active"
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title_size),
        legend.text = element_text(size = legend_text_size),
        legend.title = element_text(size = legend_title_size),
        plot.title = element_text(size = title_size),
        strip.text = element_text(size = strip_text)
  ) +
  theme(legend.position = "bottom")

```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
