---
title: "Simulation study: Robustness of GPD estimation under varying evaluation points"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
params:
  name: "05_eval_point" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "39c54387-5000-48bc-a7db-a7ed41e6e144")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 5, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/sim_GPD_evalpoint_"),
  message = FALSE, 
  warning = FALSE,
  echo = FALSE)
```

This simulation study examines how GPD parameter estimation behaves under a boundary constraint that requires the estimated support to include a predefined evaluation point above the true boundary. We assess the impact of varying sample sizes and distances (ε) between the evaluation point and the boundary across multiple estimation methods.


```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)
library(kableExtra)
library(eva)
library(furrr)
library(progressr)
conflicts_prefer(dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r, message=FALSE, warning=FALSE}
# Load permApprox functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
permApprox_path <- file.path(proj_path, "functions/permApprox_eps_fct/R")
permApprox_functions <- list.files(permApprox_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(permApprox_functions, source))
```

## Parameter grid

```{r}
# Fixed parameters
shape <- -0.2
scale <- 1
true_bound <- -scale / shape  # = 5
num_reps <- 500

# Varying parameters
sample_sizes <- c(25, 50, 75, 100, 250, 500, 750, 1000)
epsilons <- c(-1, 0, 1, 2, 3)

cat("Shape:", shape,
    "\nScale: ", scale,
    "\nTrue_bound: ", true_bound,
    "\nSample sizes: ", sample_sizes,
    "\nEpsilons: ", epsilons,
    "\nRepetitions: ", num_reps)
```

The GPD parameters are estimated for all considered parameter combinations using the following estimation methods:

* LME (Likelihood moment estimator) by Zhang (2007)
* WNLLSM (Weighted Nonlinear Least Squares Likelihood Moments Estimation) by Zhao et al. (2019)
* MLE-1D (one-dimensional MLE) by Castillo & Serra (2015)
* MLE-2D (Maximum Likelihood Estimation)
* NLS2 (Quantile Estimation) by Song & Song (2012)
* ZSE (Zhang Stephens Estimator; Bayesian methods) by Zhang & Stephens (2009)

```{r}
# Methods for parameter fitting
fit_methods <- c("LME", "MLE1D", "MLE2D", "NLS2", "WNLLSM", "ZSE")
fit_methods <- factor(fit_methods, levels = fit_methods)
```


```{r}
estimation_settings <- tidyr::crossing(
  n = sample_sizes,
  eps = epsilons,
  method = fit_methods
) %>%
  mutate(
    eval_point = true_bound + eps,
    shape = shape,
    scale = scale,
    bound = true_bound
  )

estimation_settings %>% head()
cat("Number of settings: ", nrow(estimation_settings))
```


## Perform estimations

Constrained and unconstrained fitting for all settings ...

```{r}
# Data directory
data_dir <- file.path(path_source("01_data_gen"))

# Directory setup for saving results
results_dir <- file.path(path_target())
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
# Define estimation function
estimate_one_setting <- function(i) {
  
  # Extract setting
  setting <- estimation_settings[i, ]
  shape <- setting$shape
  bound <- setting$bound
  n <- setting$n
  epsilon <- setting$eps
  eval_point <- setting$eval_point
  method <- as.character(setting$method)
  
  tol <- case_when(
    method == "LME" ~ 1e-7,
    method == "MLE1D" ~ 1e-10,
    method == "MLE2D" ~ 1e-10,
    method == "NLS2" ~ 1e-9,
    method == "WNLLSM" ~1e-9,
    TRUE ~ NA_real_
  )
  
  # Construct output file path
  result_file <- file.path(results_dir, 
                           paste0("shape-", shape,
                                  "_n-", n,
                                  "_eps-", epsilon,
                                  "_", method, 
                                  ".RDS"))
  
  if (!file.exists(result_file)) {
    # Load data
    data_file <- file.path(data_dir, 
                           paste0("shape-", shape, "_n-", n, ".RDS"))
    gpd_samples <- readRDS(data_file)
    
    fit_results <- purrr::imap_dfr(gpd_samples, ~ {
      data <- .x
      data_max <- max(data)
      
      # Unconstrained fit
      fit_uc <- fit_gpd(
        data = data,
        thresh = 0,
        fit_method = method,
        tol = tol,
        epsilon = 0,
        constraint = "unconstrained",
        support_boundary = NULL
      )

      if (fit_uc$shape < 0) {
        bound_uc <- -fit_uc$scale / fit_uc$shape
        dgpd_uc <- POT::dgpd(eval_point, scale = fit_uc$scale, shape = fit_uc$shape)
      } else {
        bound_uc <- dgpd_uc <- NA
      }
      
      constr_active <- bound_uc < eval_point

      # Constrained fit
      fit_c <- fit_gpd(
        data = data,
        thresh = 0,
        fit_method = method,
        tol = tol,
        epsilon = 0,
        constraint = "support_at_obs",
        support_boundary = eval_point
      )

      if (fit_c$shape < 0) {
        bound_c <- -fit_c$scale / fit_c$shape
        dgpd_c <- POT::dgpd(eval_point, scale = fit_c$scale, shape = fit_c$shape)
      } else {
        bound_c <- dgpd_c <- NA
      }
      
      tibble(
        data_shape = shape,
        data_scale = scale,
        data_bound = true_bound,
        eval_point = eval_point,
        eps = epsilon,
        n = n,
        method = method,
        data_max = data_max,
        constr_active = constr_active,
        shape = fit_c$shape,
        scale = fit_c$scale,
        bound = bound_c,
        dgpd = dgpd_c,
        repetition = .y
      )
    })
    
    # Save results
    saveRDS(fit_results, file = result_file)
  }
}
```

```{r, eval=FALSE}
# Set up progress handler
handlers("txtprogressbar")

# Set number of cores
cores <- 1

# Set up parallel plan
plan(multisession, workers = cores)

# Run estimation in parallel with progress
with_progress({
  pb <- progressor(steps = nrow(estimation_settings))
  
  gpdfitlist <- future_map(seq_len(nrow(estimation_settings)), function(i) {
    estimate_one_setting(i)
    pb()
  })
})

# Optionally stop the plan (not necessary in most cases)
plan(sequential)
```

```{r}
## Combine results

# List and load all result files
all_result_files <- list.files(results_dir, full.names = TRUE)

gpdfitlist <- purrr::map_dfr(all_result_files, ~ {
  res <- readRDS(.x)
  res %>% mutate(across(where(is.list), ~ unlist(unname(.))))
})

# # Remove two outliers
# sum(gpdfitlist$bound_c > 500)
# sum(gpdfitlist$bound_uc > 500)
# 
# gpdfitlist <- gpdfitlist %>%
#   dplyr::filter(bound_c < 500, bound_uc < 500)
```

## Results table

```{r}
gpdfitlist[1:20, ] %>%
  kable
```

## Heatmap (SD of dgpd)

```{r}
# ---- choose replacement rule: "global" or "per_setting" ----------------------
replace_mode <- "global"   # or "global"

# base data (constraint active only)
dgpd_base <- gpdfitlist %>%
  dplyr::filter(eps %in% c(1, 2, 3),
                constr_active == TRUE,
                !is.na(dgpd))

if (replace_mode == "global") {
  # smallest strictly positive across ALL reps/settings
  min_pos <- min(dgpd_base$dgpd[dgpd_base$dgpd > 0], na.rm = TRUE)
  dgpd_log_ready <- dgpd_base %>%
    dplyr::mutate(dgpd_log = ifelse(dgpd > 0, log(dgpd), log(min_pos)))
} else { # "per_setting" = per (eps, n, method)
  dgpd_log_ready <- dgpd_base %>%
    dplyr::group_by(eps, n, method) %>%
    dplyr::mutate(
      min_pos_grp = suppressWarnings(min(dgpd[dgpd > 0], na.rm = TRUE)),
      # if a group has no positive values at all -> set to NA
      repl_val    = dplyr::if_else(is.finite(min_pos_grp), min_pos_grp, NA_real_),
      dgpd_log    = dplyr::if_else(dgpd > 0, log(dgpd), log(repl_val))
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-min_pos_grp, -repl_val)
}

# ---- summarise SD of log-dgpd per (eps, n, method) ---------------------------
dgpd_sd <- dgpd_log_ready %>%
  dplyr::group_by(eps, n, method) %>%
  dplyr::summarise(
    n_active = dplyr::n(),
    sd_dgpd  = sd(dgpd_log, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  # keep tiles with at least 2 reps; others NA (blank)
  dplyr::mutate(
    sd_dgpd = dplyr::if_else(n_active >= 2, sd_dgpd, NA_real_),
    n       = factor(n),
    method  = as.character(method)
  )

```

```{r}
# Heatmap (row-wise scaling, same look & feel)
plot_dgpd_sd_heat <- function(df, 
                              digits = 3, 
                              scale_mode = c("row","global"),
                              scale_low = 0.4,
                              scale_mid = 0.7,
                              scale_high = 0.95) {
  
  scale_mode <- match.arg(scale_mode)
  
  # Make LME bold
  method_labels <- setNames(
    lapply(levels(factor(df$method)), function(m) {
      if (m == "LME") bquote(bold(.(m) * "*")) else m
    }),
    levels(factor(df$method))
  )
  
  # per-row "best" = smallest SD
  df <- df %>%
    dplyr::group_by(eps, n) %>%
    dplyr::mutate(
      row_min = min(sd_dgpd, na.rm = TRUE),
      is_best = !is.na(sd_dgpd) & (abs(sd_dgpd - row_min) <= .Machine$double.eps^0.5),
      lab     = ifelse(is.na(sd_dgpd), "", sprintf(paste0("%.", digits, "f"), sd_dgpd))
    ) %>%
    dplyr::ungroup()

  if (scale_mode == "row") {
    df <- df %>%
      dplyr::group_by(eps, n) %>%
      dplyr::mutate(
        row_max = max(sd_dgpd, na.rm = TRUE),
        scaled  = dplyr::if_else(row_max > row_min,
                                 (sd_dgpd - row_min)/(row_max - row_min), 0)
      ) %>%
      dplyr::ungroup()

    fill_mapping <- aes(fill = scaled)
    fill_scale   <- scale_fill_gradient(low = "yellow", high = "red",
                                        name = "SD")
    
  } else { # global
    vmin <- min(df$sd_dgpd, na.rm = TRUE)
    vmax <- max(df$sd_dgpd, na.rm = TRUE)
    q_low  <- quantile(df$sd_dgpd, probs = scale_low, na.rm = TRUE)
    q_mid  <- quantile(df$sd_dgpd, probs = scale_mid, na.rm = TRUE)
    q_high <- quantile(df$sd_dgpd, probs = scale_high, na.rm = TRUE)
    
    fill_mapping <- aes(fill = sd_dgpd)
    fill_scale   <- scale_fill_gradientn(
      colours = c("white", "yellow", "orange", "red", "red3"),
      values  = scales::rescale(c(vmin, q_low, q_mid, q_high, vmax), 
                                from = c(vmin, vmax)),
      limits = c(vmin, vmax),
      name = "SD",
      guide   = guide_colourbar(
        frame.colour   = "black",
        frame.linewidth = 0.3,
        ticks.colour   = "black"
      ))
  }

  ggplot(df, aes(x = method, y = n)) +
    geom_tile(fill_mapping, color = "grey30", linewidth = 0.3) +
    geom_text(aes(label = lab), size = 3) +
    fill_scale +
    facet_wrap(~ eps, labeller = label_bquote(epsilon == .(eps))) +
    labs(
      #title   = "SD of f(x_eval) (constraint active)",
      #subtitle = bquote(sqrt(Var(f[GPD](x[eval])))),
      x = "Method", y = "n"
    ) +
    theme_minimal() +
    scale_x_discrete(labels = method_labels) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text  = element_text(size = 12, face = "bold")
    )
}
```

```{r heat_sd_dgpd, fig.width=12, fig.height=4, dpi = 200}
plot_dgpd_sd_heat(dgpd_sd, digits = 3, scale_mode = "global")
```

## Plots

### Colors

```{r}
# Okabe Ito palette (colorblind-friendly)
okabe_ito_cols <- c("black",   "#E69F00", "#56B4E9", 
                    "#009E73", "#F0E442", "#0072B2",
                    "#D55E00", "#CC79A7", "gray60")

my_cols <- c(
  "MOM"    = "#6C6C6C",
  "PWM"    = "#6C6C6C",
  "LME"    = "#009E73", 
  "MLE1D"  = "#56B4E9", 
  "MLE2D"  = "#0072B2",
  "NLS2"   = "#E69F00",
  "WNLLSM" = "#D55E00",
  "ZSE"    = "#CC79A7"
)
```

```{r}
# Compute errors
error_df <- gpdfitlist %>%
  pivot_longer(cols = c(shape, scale, bound), names_to = "param", 
               values_to = "estimate") %>%
  mutate(
    true_value = case_when(
      param == "shape" ~ data_shape,
      param == "scale" ~ data_scale,
      param == "bound" ~ data_bound
    ),
    error = estimate - true_value,
    abs_error = abs(error)
  )

```


```{r}
# Plot function
plot_error_by_eps <- function(df, param_name, eps_val, title_label, y_label,
                              axis_text_size = 14,
                              axis_title_size = 16,
                              legend_text_size = 12,
                              legend_title_size = 16,
                              title_size = 16,
                              col_palette) {
  eps_lab <- paste0("ε = ", eps_val)
  
  df %>%
    filter(param == param_name, eps == eps_val) %>%
    ggplot(aes(x = factor(n), y = error, fill = method)) +
    geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_fill_manual(values = col_palette) +
    labs(
      title = paste(title_label, "|", eps_lab),
      x = "n",
      y = y_label,
      fill = "Method"
    ) +
    theme_minimal() +
    theme(axis.text = element_text(size = axis_text_size),
          axis.title = element_text(size = axis_title_size),
          legend.text = element_text(size = legend_text_size),
          legend.title = element_text(size = legend_title_size),
          plot.title = element_text(size = title_size)
    )
}
```


### Shape error

```{r error_shape_by_n}
for (eps in epsilons) {
  p <- error_df %>% plot_error_by_eps(
    param_name = "shape",
    eps_val = eps,
    col_palette = my_cols,
    title_label = "Estimation error in shape",
    y_label = expression(hat(xi) - xi[0])
  ) +  coord_cartesian(ylim = c(-0.1, 1))
  
  print (p)
}

```

### Scale error

```{r error_scale_by_n}
for (eps in epsilons) {
  p <- error_df %>% plot_error_by_eps(
    param_name = "scale",
    eps_val = eps,
    col_palette = my_cols,
    title_label = "Estimation error in scale",
    y_label = expression(hat(sigma) - sigma[0])
  ) #+  coord_cartesian(ylim = c(-0.1, 1))
  
  print (p)
}

```

### Boundary error

```{r error_bound_by_n}
for (eps in epsilons) {
  p <- error_df %>% plot_error_by_eps(
    param_name = "bound",
    eps_val = eps,
    col_palette = my_cols,
    title_label = "Estimation error in boundary",
    y_label = expression(hat(s) - s[0])
  ) +  coord_cartesian(ylim = c(0, 10))
  
  print (p)
}

```

### Difference boundary to eval_point

```{r}
# Plot function
plot_delta_bound_by_eps <- function(df, eps_val, title = NULL,
                              axis_text_size = 14,
                              axis_title_size = 16,
                              legend_text_size = 12,
                              legend_title_size = 16,
                              title_size = 16,
                              col_palette) {
  eps_lab <- paste0("ε = ", eps_val)
  
  df <- df %>%
    filter(eps == eps_val) %>%
    mutate(delta = bound - eval_point)
  
  df %>%
    ggplot(aes(x = factor(n), y = delta, fill = method)) +
    geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
    #geom_hline(yintercept = 0, linetype = "dashed") +
    scale_fill_manual(values = col_palette) +
    labs(
      title = paste0("Difference between boundary and eval point ",
                     title, " | ", eps_lab),
      x = "n",
      y = expression(hat(s) - x[eval]),
      fill = "Method"
    ) +
    theme_minimal() +
    theme(axis.text = element_text(size = axis_text_size),
          axis.title = element_text(size = axis_title_size),
          legend.text = element_text(size = legend_text_size),
          legend.title = element_text(size = legend_title_size),
          plot.title = element_text(size = title_size)
    )
}
```

```{r diff_bound_evalpoint_by_n}
for (eps in epsilons) {
  p <- gpdfitlist %>% 
    filter(!is.na(bound), bound < 500) %>%
    plot_delta_bound_by_eps(
      eps_val = eps, col_palette = my_cols,
    ) +  coord_cartesian(ylim = c(0, 10))
  
  print (p)
}
```

### Difference boundary to eval_point (log scale)

```{r diff_bound_evalpoint_log_by_n}
for (eps in epsilons) {
  p <- gpdfitlist %>% 
    filter(!is.na(bound), bound < 500) %>%
    plot_delta_bound_by_eps(
      eps_val = eps, col_palette = my_cols,
      title = "(log scale)"
    ) + scale_y_log10()
  
  print (p)
}
```

### Difference boundary to eval_point (constraint active, log scale)

```{r diff_bound_evalpoint_active_by_n}
for (eps in epsilons) {
  p <- gpdfitlist %>%
    filter(constr_active == TRUE) %>% 
    plot_delta_bound_by_eps(
      eps_val = eps, col_palette = my_cols,
      title = "(constraint active, log scale)"
    ) + scale_y_log10()
  
  print (p)
}
```

### GPD density at eval_point

```{r}
# Plot function
plot_dgpd_by_eps <- function(df, eps_val, title = NULL,
                             axis_text_size = 14,
                             axis_title_size = 16,
                             legend_text_size = 12,
                             legend_title_size = 16,
                             title_size = 16,
                             col_palette) {
  eps_lab <- paste0("ε = ", eps_val)
  
  df <- df %>%
    filter(eps == eps_val)
  
  df %>%
    ggplot(aes(x = factor(n), y = dgpd, fill = method)) +
    geom_boxplot(outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
    #geom_hline(yintercept = 0, linetype = "dashed") +
    scale_fill_manual(
      values = col_palette,
      labels = setNames(
        lapply(names(col_palette), function(m) {
          if (m == "LME") bquote(bold("LME" * "*")) else m
        }),
        names(col_palette)
      )
    ) +
    labs(
      title = paste0("GPD density at evaluation point ",
                     title, " | ", eps_lab),
      x = "n",
      y = expression(f[widehat(GPD)](x[eval])),
      fill = "Method"
    ) +
    theme_minimal() +
    theme(axis.text = element_text(size = axis_text_size),
          axis.title = element_text(size = axis_title_size),
          legend.text = element_text(size = legend_text_size),
          legend.title = element_text(size = legend_title_size),
          plot.title = element_text(size = title_size)
    ) +
    scale_y_log10()
}
```

```{r dgpd_log_by_n}
for (eps in epsilons) {
  p <- gpdfitlist %>%
    plot_dgpd_by_eps(
      eps_val = eps, col_palette = my_cols,
      title = ""
    )
  
  print (p)
}
```

### GPD density at eval_point (constraint active)

```{r dgpd_active_by_n, fig.width = 10, fig.height = 5, dpi = 200}
for (eps in epsilons) {
  p <- gpdfitlist %>%
    filter(constr_active == TRUE) %>% 
    plot_dgpd_by_eps(
      eps_val = eps, col_palette = my_cols,
      title = "(constraint active)"
    )
  
  print (p)
}
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
