---
title: "Example figures for the methods chapter"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "example_figures" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "a8d40ab2-6177-4d45-bb75-de680b946e54")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/netcomi_"),
  echo = FALSE)
```

```{r packages}
library("conflicted")
library("dplyr")
library("ggplot2")
library("igraph")
library("NetCoMi")
library("patchwork")
library("tidyr")


conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::select)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Introduction

This script collects small, self-contained figures used to illustrate concepts in
the NetCoMi chapter. The examples are meant to be visually simple rather than
data-driven: each figure isolates one methodological step so that it can be used
as explanatory material in the thesis text or during a presentation.

## Association transformations

The following plot compares signed and unsigned transformations from sparse
associations to dissimilarities and the corresponding similarities. The signed
transformation treats negative associations as large dissimilarities and small
similarities, whereas the unsigned transformation uses only association strength.

```{r asso_diss_transformation, fig.width=8, fig.height=3, dpi=200}
cols <- c("darkorange2", "#1f77b4")

association_grid <- tibble(
  association = seq(-1, 1, length.out = 501)
)

association_transforms <- association_grid |>
  mutate(
    Signed = sqrt(0.5 * (1 - association)),
    Unsigned = sqrt(1 - association^2)
  ) |>
  pivot_longer(
    cols = -association,
    names_to = "Transformation",
    values_to = "Dissimilarity"
  ) |>
  mutate(
    Similarity = 1 - Dissimilarity
  )

transformation_colors <- c(
  "Signed" = cols[1],
  "Unsigned" = cols[2]
)

base_transform_plot <- ggplot(
  association_transforms,
  aes(x = association, color = Transformation)
) +
  geom_vline(
    xintercept = 0,
    linewidth = 1,
    linetype = "dashed",
    color = "gray40"
  ) +
  scale_color_manual(values = transformation_colors) +
  scale_x_continuous(
    breaks = seq(-1, 1, by = 0.5),
    limits = c(-1, 1)
  ) +
  labs(
    x = "Association",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

dissimilarity_plot <- base_transform_plot +
  geom_line(aes(y = Dissimilarity), linewidth = 1.1) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.25),
    limits = c(-0.05, 1.05),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    y = "Dissimilarity"
  )

similarity_plot <- base_transform_plot +
  geom_line(aes(y = Similarity), linewidth = 1.1) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.25),
    limits = c(-0.05, 1.05),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    y = "Similarity"
  )

dissimilarity_plot + similarity_plot +
  plot_layout(guides = "collect") &
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 14),
    legend.position = "bottom")
```

## Centrality measures

The following example uses a modular core-periphery network to illustrate how
different centrality measures highlight different structural roles. The network
contains two subnetworks with different internal densities, peripheral leaf-like
nodes, and one low-degree bridge node connecting the subnetworks. This structure
is kept separate from the larger graphlet examples below so that the highlighted
centrality roles remain visible.

```{r}
set.seed(42)

connect_components <- function(graph) {
  component_info <- igraph::components(graph)
  if (component_info$no == 1) {
    return(graph)
  }

  representative_nodes <- vapply(
    seq_len(component_info$no),
    function(component) which(component_info$membership == component)[1],
    numeric(1)
  )
  connection_edges <- c(rbind(
    head(representative_nodes, -1),
    tail(representative_nodes, -1)
  ))

  igraph::add_edges(graph, connection_edges)
}

sample_centrality_ecological_network <- function(
    subnetwork_sizes = c(15, 15),
    n_leaves_per_subnetwork = 6,
    p_within = c(0.12, 0.28),
    n_bridge_targets = 3) {
  n_core <- sum(subnetwork_sizes)
  n_bridge_nodes <- 1
  n_leaves <- length(subnetwork_sizes) * n_leaves_per_subnetwork
  n_total <- n_core + n_bridge_nodes + n_leaves

  subnetworks <- lapply(
    seq_along(subnetwork_sizes),
    function(subnetwork_index) {
      repeat {
        sampled_subnetwork <- igraph::sample_gnp(
          n = subnetwork_sizes[subnetwork_index],
          p = p_within[subnetwork_index],
          directed = FALSE,
          loops = FALSE
        )
        if (igraph::components(sampled_subnetwork)$no == 1) {
          return(sampled_subnetwork)
        }
      }
    }
  )
  graph <- Reduce(igraph::disjoint_union, subnetworks)
  graph <- igraph::add_vertices(graph, n_total - n_core)

  subnetwork_starts <- cumsum(c(1, head(subnetwork_sizes, -1)))
  subnetwork_ends <- cumsum(subnetwork_sizes)
  subnetwork_hubs <- vapply(
    seq_along(subnetwork_sizes),
    function(subnetwork) {
      module_nodes <- subnetwork_starts[subnetwork]:subnetwork_ends[subnetwork]
      module_nodes[which.max(igraph::degree(graph, v = module_nodes))]
    },
    numeric(1)
  )
  subnetwork_connectors <- lapply(
    seq_along(subnetwork_sizes),
    function(subnetwork) {
      module_nodes <- subnetwork_starts[subnetwork]:subnetwork_ends[subnetwork]
      connector_candidates <- setdiff(module_nodes, subnetwork_hubs[subnetwork])
      connector_candidates[
        order(igraph::degree(graph, v = connector_candidates))
      ][seq_len(n_bridge_targets)]
    }
  )

  bridge_node <- n_core + 1
  leaf_nodes <- (n_core + n_bridge_nodes + 1):n_total
  ecological_edges <- c()

  ecological_edges <- c(
    ecological_edges,
    as.vector(rbind(subnetwork_connectors[[1]], bridge_node)),
    as.vector(rbind(bridge_node, subnetwork_connectors[[2]]))
  )

  leaf_groups <- split(
    leaf_nodes,
    rep(seq_along(subnetwork_sizes), each = n_leaves_per_subnetwork)
  )
  for (subnetwork in seq_along(leaf_groups)) {
    hub_index <- as.numeric(names(leaf_groups)[subnetwork])
    for (leaf_node in leaf_groups[[subnetwork]]) {
      ecological_edges <- c(
        ecological_edges,
        leaf_node,
        subnetwork_hubs[hub_index]
      )
    }
  }

  graph <- igraph::add_edges(graph, ecological_edges)
  igraph::simplify(graph, remove.multiple = TRUE, remove.loops = TRUE)
}

centrality_network <- sample_centrality_ecological_network()
centrality_layout <- igraph::layout_with_fr(centrality_network)

centrality_values <- list(
  "Degree" = igraph::degree(centrality_network),
  "Betweenness" = igraph::betweenness(
    centrality_network,
    directed = FALSE,
    normalized = TRUE
  ),
  "Closeness" = igraph::closeness(
    centrality_network,
    normalized = TRUE
  ),
  "Eigenvector" = igraph::eigen_centrality(
    centrality_network,
    directed = FALSE
  )$vector
)

scale_to_unit <- function(x) {
  if (diff(range(x)) == 0) {
    return(rep(0, length(x)))
  }
  (x - min(x)) / diff(range(x))
}

centrality_palette <- colorRampPalette(
  c("gray80", "#E69F00", "#B2182B")
)(101)

plot_centrality_network <- function(graph, layout, values, main,
                                    vertex_size = 8, title_size = 1) {
  scaled_values <- scale_to_unit(values)
  node_colors <- centrality_palette[
    pmax(1, round(scaled_values * 100) + 1)
  ]

  plot(
    graph,
    layout = layout,
    vertex.size = vertex_size,
    vertex.label = NA,
    vertex.color = node_colors,
    vertex.frame.color = "white",
    edge.color = adjustcolor("gray20", alpha.f = 0.35),
    edge.width = 0.6,
    main = NA
  )
  title(main = main, cex.main = title_size, line = 0.4)
}
```


```{r centrality_measures_network, fig.width=10, fig.height=3, dpi=250}
old_par_centrality <- par(no.readonly = TRUE)
par(
  mfrow = c(1, 4),
  mar = c(0.5, 0, 2.4, 0),
  oma = c(0, 0, 0, 0)
)

for (centrality_name in names(centrality_values)) {
  plot_centrality_network(
    centrality_network,
    centrality_layout,
    centrality_values[[centrality_name]],
    centrality_name,
    vertex_size = 10,
    title_size = 1.8
  )
}

par(old_par_centrality)
```

Degree highlights taxa with many direct neighbors, here mainly module hubs to
which several peripheral nodes are attached. Betweenness emphasizes the
low-degree bridge node that lies on shortest paths between the two subnetworks,
even though this node is not highly connected itself. Closeness
favors nodes that are globally well positioned with respect to the whole
network, often near the bridge between subnetworks. Eigenvector centrality
highlights nodes embedded in central and well-connected neighborhoods, rather
than nodes that are merely bridges, and is therefore higher in the denser
subnetwork.

## Standard networks and graphlet correlation matrices

The following examples compare three standard network models with the same
number of nodes as the filtered genus-level data set used in the application
chapter: an Erdos-Renyi graph, a scale-free Barabasi-Albert graph, and a
stochastic block model. The Erdos-Renyi graph is generated with the same edge
count as the stochastic block model. For each graph, the corresponding graphlet
correlation matrix is computed from the binary adjacency matrix.

```{r network_gcms_functions, fig.width=8, fig.height=13, dpi=250}
set.seed(42)

n_genera <- 117
connect_components <- function(graph) {
  component_info <- igraph::components(graph)
  if (component_info$no == 1) {
    return(graph)
  }

  representative_nodes <- vapply(
    seq_len(component_info$no),
    function(component) which(component_info$membership == component)[1],
    numeric(1)
  )
  connection_edges <- c(rbind(
    head(representative_nodes, -1),
    tail(representative_nodes, -1)
  ))

  igraph::add_edges(graph, connection_edges)
}

sample_ecological_core_periphery <- function(
    core_sizes = c(16, 16, 16, 16),
    n_connectors = 11,
    n_peripheral = 42,
    p_within = 0.35,
    p_between = 0.005) {
  n_core <- sum(core_sizes)
  n_total <- n_core + n_connectors + n_peripheral
  module_probabilities <- matrix(
    p_between,
    nrow = length(core_sizes),
    ncol = length(core_sizes)
  )
  diag(module_probabilities) <- p_within

  graph <- igraph::sample_sbm(
    n = n_core,
    pref.matrix = module_probabilities,
    block.sizes = core_sizes,
    directed = FALSE,
    loops = FALSE
  )
  graph <- igraph::add_vertices(graph, n_total - n_core)

  module_starts <- cumsum(c(1, head(core_sizes, -1)))
  module_ends <- cumsum(core_sizes)
  module_hubs <- vapply(
    seq_along(core_sizes),
    function(module) {
      module_nodes <- module_starts[module]:module_ends[module]
      module_nodes[which.max(igraph::degree(graph, v = module_nodes))]
    },
    numeric(1)
  )

  connector_nodes <- (n_core + 1):(n_core + n_connectors)
  peripheral_nodes <- (n_core + n_connectors + 1):n_total
  ecological_edges <- c()

  for (connector_node in connector_nodes) {
    connected_modules <- sample(seq_along(core_sizes), 2)
    for (module in connected_modules) {
      ecological_edges <- c(
        ecological_edges,
        connector_node,
        sample(module_starts[module]:module_ends[module], 2)
      )
    }
  }

  for (peripheral_index in seq_along(peripheral_nodes)) {
    peripheral_node <- peripheral_nodes[peripheral_index]
    if (peripheral_index <= 0.75 * length(peripheral_nodes)) {
      target_node <- sample(module_hubs, 1)
    } else {
      target_node <- sample(connector_nodes, 1)
    }

    ecological_edges <- c(ecological_edges, peripheral_node, target_node)
    if (runif(1) < 0.12) {
      ecological_edges <- c(
        ecological_edges,
        peripheral_node,
        sample(module_hubs, 1)
      )
    }
  }

  connect_components(igraph::add_edges(graph, ecological_edges))
}

add_peripheral_leaves_to_hubs <- function(graph, n_leaves = 25, hub_pool = 8) {
  n_core <- igraph::vcount(graph)
  graph <- igraph::add_vertices(graph, n_leaves)
  hub_nodes <- order(igraph::degree(graph, v = seq_len(n_core)),
                     decreasing = TRUE)[seq_len(hub_pool)]
  leaf_nodes <- (n_core + 1):(n_core + n_leaves)
  leaf_targets <- sample(
    hub_nodes,
    size = n_leaves,
    replace = TRUE,
    prob = igraph::degree(graph, v = hub_nodes)
  )
  leaf_edges <- as.vector(rbind(leaf_nodes, leaf_targets))

  igraph::add_edges(graph, leaf_edges)
}

g_ba <- igraph::sample_pa(
  n = n_genera,
  m = 2,
  directed = FALSE
)
n_ba_peripheral <- 25
g_ba_peripheral <- igraph::sample_pa(
  n = n_genera - n_ba_peripheral,
  m = 2,
  directed = FALSE
)
g_ba_peripheral <- add_peripheral_leaves_to_hubs(
  g_ba_peripheral,
  n_leaves = n_ba_peripheral
)
module_sizes <- c(22, 20, 19, 18, 20, 18)
module_probabilities <- matrix(
  0.003,
  nrow = length(module_sizes),
  ncol = length(module_sizes)
)
diag(module_probabilities) <- c(0.24, 0.22, 0.23, 0.21, 0.22, 0.20)
g_sbm <- igraph::sample_sbm(
  n = n_genera,
  pref.matrix = module_probabilities,
  block.sizes = module_sizes,
  directed = FALSE,
  loops = FALSE
)
module_starts <- cumsum(c(1, head(module_sizes, -1)))
module_edges <- cbind(
  module_starts[-length(module_starts)],
  module_starts[-1]
)
g_sbm <- igraph::add_edges(g_sbm, t(module_edges))
g_sbm <- connect_components(g_sbm)
g_ecological <- sample_ecological_core_periphery()
g_er <- igraph::sample_gnm(
  n = n_genera,
  m = igraph::ecount(g_sbm),
  directed = FALSE,
  loops = FALSE
)

standard_networks <- list(
  "Erdos-Renyi" = g_er,
  "Barabasi-Albert" = g_ba,
  "Stochastic block model" = g_sbm
)
stopifnot(
  all(vapply(standard_networks, igraph::vcount, numeric(1)) == n_genera),
  igraph::vcount(g_ecological) == n_genera,
  igraph::vcount(g_ba_peripheral) == n_genera
)

gcm_orbits <- c(0, 2, 5, 7, 8, 10, 11, 6, 9, 4, 1)

calc_standard_gcm <- function(graph) {
  adja <- igraph::as_adjacency_matrix(graph, sparse = FALSE)
  NetCoMi::calcGCM(as.matrix(adja), orbits = gcm_orbits)[["gcm"]]
}

standard_gcms <- lapply(standard_networks, calc_standard_gcm)
network_layouts <- lapply(
  standard_networks,
  igraph::layout_with_fr
)
ba_peripheral_gcm <- calc_standard_gcm(g_ba_peripheral)
ba_peripheral_layout <- igraph::layout_with_fr(g_ba_peripheral)
ecological_gcm <- calc_standard_gcm(g_ecological)
ecological_layout <- igraph::layout_with_fr(g_ecological)

plot_standard_network <- function(graph, layout, main, title_size = 1) {
  plot(
    graph,
    layout = layout,
    vertex.size = 4,
    vertex.label = NA,
    vertex.color = "gray40",
    vertex.frame.color = NA,
    edge.color = adjustcolor("gray20", alpha.f = 0.45),
    edge.width = 0.5,
    main = NA
  )
  title(main = main, cex.main = title_size, line = 0.4)
}

plot_gcm_heatmap <- function(gcm, main, title_size = 1,
                             orbit_size = 0.9, legend_size = 0.9) {
  gcm <- gcm[paste0("O", gcm_orbits), paste0("O", gcm_orbits)]
  orbit_labels <- sub("^O", "", rownames(gcm))
  heat_cols <- c(
    colorRampPalette(c("#2166AC", "white"))(101),
    colorRampPalette(c("white", "#E69F00", "#B2182B"))(101)[-1]
  )
  old_xpd <- par("xpd")
  on.exit(par(xpd = old_xpd), add = TRUE)

  image(
    x = seq(0.5, ncol(gcm) + 0.5, by = 1),
    y = seq(0.5, nrow(gcm) + 0.5, by = 1),
    z = t(gcm[nrow(gcm):1, ]),
    col = heat_cols,
    zlim = c(-1, 1),
    xlim = c(0.5, ncol(gcm) + 2.2),
    axes = FALSE,
    xlab = "",
    ylab = "",
    main = ""
  )
  axis(
    side = 3,
    at = seq_along(orbit_labels),
    labels = orbit_labels,
    tick = FALSE,
    line = -0.8,
    cex.axis = orbit_size,
    gap.axis = -1
  )
  axis(
    side = 2,
    at = seq_along(orbit_labels),
    labels = rev(orbit_labels),
    tick = FALSE,
    las = 1,
    line = -0.5,
    cex.axis = orbit_size,
    gap.axis = -1
  )
  title(main = main, line = 2.1, cex.main = title_size)
  rect(0.5, 0.5, ncol(gcm) + 0.5, nrow(gcm) + 0.5)

  legend_x <- c(ncol(gcm) + 0.9, ncol(gcm) + 1.15)
  legend_y <- seq(0.5, nrow(gcm) + 0.5, length.out = length(heat_cols) + 1)
  image(
    x = legend_x,
    y = legend_y,
    z = matrix(seq(-1, 1, length.out = length(heat_cols)), nrow = 1),
    col = heat_cols,
    zlim = c(-1, 1),
    add = TRUE
  )
  text(
    x = ncol(gcm) + 1.7,
    y = seq(1, nrow(gcm), length.out = 5),
    labels = seq(-1, 1, by = 0.5),
    cex = legend_size
  )
}
```

### Erdos-Renyi network

The Erdos-Renyi network is used as a random baseline. Edges are placed without
an explicit preference for hubs or modules, so any local structure arises from
chance rather than from an imposed ecological organization.

```{r network_gcms_ER, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Erdos-Renyi"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM shows broadly positive correlations among most graphlet orbits. This is
expected for a random graph of this size and density: nodes with higher degree
also tend to participate more often in many other small graphlet positions,
whereas clear opposing orbit roles are largely absent.

### Barabasi-Albert network

The Barabasi-Albert network represents a scale-free hub structure. New nodes
preferentially attach to already well-connected nodes, producing a few hubs and
many nodes with low to moderate degree.

```{r network_gcms_BA, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Barabasi-Albert"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM has more visible block structure than the random baseline. Orbits
associated with hub participation are highly correlated with each other. Since
the network is generated with two edges for each newly added node, it contains
few or no terminal nodes; the terminal-orbit group is therefore less pronounced
than in ecological networks with many endpoints.

### Barabasi-Albert network with peripheral leaves

This exploratory variant keeps a scale-free Barabasi-Albert core but adds
degree-one peripheral taxa attached preferentially to hubs. It therefore
combines hub-dominated structure with endpoints, a feature that is often more
plausible for ecological or microbiome networks than a pure Barabasi-Albert
model with `m = 2`.

```{r network_gcms_BA_peripheral, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Barabasi-Albert with leaves"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  g_ba_peripheral,
  ba_peripheral_layout,
  paste(network_name, "network")
)
plot_gcm_heatmap(
  ba_peripheral_gcm,
  paste(network_name, "GCM")
)
```

Compared with the pure Barabasi-Albert graph, this network should strengthen
the contrast between hub-related orbits and terminal-node roles. Hubs remain
important for orbits associated with degree and internal graphlet positions,
while the added leaves increase counts for terminal orbits and can make their
relationship to the hub-dominated part of the network easier to see.

```{r network_gcms_BA_peripheral_thesis, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Barabasi-Albert with leaves"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.5, 1.5, 1.5),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  g_ba_peripheral,
  ba_peripheral_layout,
  main = NA
)
plot_gcm_heatmap(
  ba_peripheral_gcm,
  main = NA
)
```

### Stochastic block model

The stochastic block model represents a modular ecological network. Edges are
more likely within modules than between modules, mimicking groups of taxa that
share similar niches or co-occurrence patterns.

```{r network_gcms_SB, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Stochastic block model"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM reflects the modular structure through correlated groups of orbits.
Dense within-module neighborhoods increase the dependence among graphlet roles
that occur in locally clustered regions, while between-module links contribute
less strongly. The result is more structured than the Erdos-Renyi baseline but
less hub-dominated than the Barabasi-Albert network.

### Core-periphery ecological network

The core-periphery ecological network combines dense module cores with connector
taxa and many peripheral taxa. This creates a contrast between taxa embedded in
locally dense communities and taxa attached as leaves or sparse connectors.

```{r network_gcms_Eco, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Core-periphery ecological"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  g_ecological,
  ecological_layout,
  paste(network_name, "network")
)
plot_gcm_heatmap(
  ecological_gcm,
  paste(network_name, "GCM")
)
```

The GCM shows the clearest contrast between graphlet roles. Orbits 0, 2, 5, and
7 describe degree and internal or branching positions in open structures, while
orbits 8, 10, and 11 describe participation in cycle- or triangle-related
structures. In this example, these roles are concentrated in the module cores
and connector taxa. Their negative correlations with orbits 6 and 9 indicate
that nodes frequently participating in core, branching, or cyclic graphlets are
not usually the same nodes that occupy terminal graphlet positions. Conversely,
nodes with high counts for terminal roles tend to be peripheral taxa attached to
the network boundary and therefore have low counts for the denser core-related
orbits.

### Combined overview

The combined figure places the three standard network models and their GCMs
next to each other to make the structural differences easier to compare. This is
the version intended for the thesis appendix.

```{r network_gcms, fig.width=8, fig.height=9.75, dpi=250}
par(
  mfrow = c(3, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

for (network_name in names(standard_networks)) {
  plot_standard_network(
    standard_networks[[network_name]],
    network_layouts[[network_name]],
    paste(network_name, "network"),
    title_size = 1.5
  )
  plot_gcm_heatmap(
    standard_gcms[[network_name]],
    paste(network_name, "GCM"),
    title_size = 1.5,
    orbit_size = 1.15,
    legend_size = 1.15
  )
}
```

Across the three standard models, the GCMs become more structured as the
networks move away from a homogeneous random baseline. The random network mainly shows degree-driven positive orbit correlations, the Barabasi-Albert network highlights hub-related dependencies, and the stochastic block model emphasizes module-related dependencies.

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
