---
title: "Example figures for the methods chapter"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "example_figures" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "a8d40ab2-6177-4d45-bb75-de680b946e54")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/netcomi_"),
  echo = FALSE)
```

```{r packages}
library("conflicted")
library("dplyr")
library("ggplot2")
library("igraph")
library("NetCoMi")
library("patchwork")
library("tidyr")


conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::select)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Introduction

This script collects small, self-contained figures used to illustrate concepts in
the NetCoMi chapter. The examples are meant to be visually simple rather than
data-driven: each figure isolates one methodological step so that it can be used
as explanatory material in the thesis text or during a presentation.

## Association transformations

The following plot compares signed and unsigned transformations from sparse
associations to dissimilarities and the corresponding similarities. The signed
transformation treats negative associations as large dissimilarities and small
similarities, whereas the unsigned transformation uses only association strength.

```{r asso_diss_transformation, fig.width=8, fig.height=3, dpi=200}
cols <- c("darkorange2", "#1f77b4")

association_grid <- tibble(
  association = seq(-1, 1, length.out = 501)
)

association_transforms <- association_grid |>
  mutate(
    Signed = sqrt(0.5 * (1 - association)),
    Unsigned = sqrt(1 - association^2)
  ) |>
  pivot_longer(
    cols = -association,
    names_to = "Transformation",
    values_to = "Dissimilarity"
  ) |>
  mutate(
    Similarity = 1 - Dissimilarity
  )

transformation_colors <- c(
  "Signed" = cols[1],
  "Unsigned" = cols[2]
)

base_transform_plot <- ggplot(
  association_transforms,
  aes(x = association, color = Transformation)
) +
  geom_vline(
    xintercept = 0,
    linewidth = 0.3,
    linetype = "dotted"
  ) +
  scale_color_manual(values = transformation_colors) +
  scale_x_continuous(
    breaks = seq(-1, 1, by = 0.5),
    limits = c(-1, 1)
  ) +
  labs(
    x = "Association",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

dissimilarity_plot <- base_transform_plot +
  geom_line(aes(y = Dissimilarity), linewidth = 1.1) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.25),
    limits = c(-0.05, 1.05),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    y = "Dissimilarity"
  )

similarity_plot <- base_transform_plot +
  geom_line(aes(y = Similarity), linewidth = 1.1) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.25),
    limits = c(-0.05, 1.05),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    y = "Similarity"
  )

dissimilarity_plot + similarity_plot +
  plot_layout(guides = "collect") &
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 14),
    legend.position = "bottom")
```

## Standard networks and graphlet correlation matrices

The following examples show four standard networks with the same number of
nodes as the filtered genus-level data set used in the application chapter. An
Erdos-Renyi graph is compared with a scale-free Barabasi-Albert graph and a
stochastic block model, which represents a modular ecological network with
denser within-module than between-module connectivity. The last network adds
connector and peripheral taxa to this modular structure to create a stronger
contrast between core/module and leaf-like graphlet roles. The Erdos-Renyi graph
is generated with the same edge count as the stochastic block model. For each
graph, the corresponding graphlet correlation matrix is computed from the binary
adjacency matrix.

```{r standard_network_gcms_functions, fig.width=8, fig.height=13, dpi=250}
set.seed(20260619)

n_genera <- 117
connect_components <- function(graph) {
  component_info <- igraph::components(graph)
  if (component_info$no == 1) {
    return(graph)
  }

  representative_nodes <- vapply(
    seq_len(component_info$no),
    function(component) which(component_info$membership == component)[1],
    numeric(1)
  )
  connection_edges <- c(rbind(
    head(representative_nodes, -1),
    tail(representative_nodes, -1)
  ))

  igraph::add_edges(graph, connection_edges)
}

sample_ecological_core_periphery <- function(
    core_sizes = c(16, 16, 16, 16),
    n_connectors = 11,
    n_peripheral = 42,
    p_within = 0.35,
    p_between = 0.005) {
  n_core <- sum(core_sizes)
  n_total <- n_core + n_connectors + n_peripheral
  module_probabilities <- matrix(
    p_between,
    nrow = length(core_sizes),
    ncol = length(core_sizes)
  )
  diag(module_probabilities) <- p_within

  graph <- igraph::sample_sbm(
    n = n_core,
    pref.matrix = module_probabilities,
    block.sizes = core_sizes,
    directed = FALSE,
    loops = FALSE
  )
  graph <- igraph::add_vertices(graph, n_total - n_core)

  module_starts <- cumsum(c(1, head(core_sizes, -1)))
  module_ends <- cumsum(core_sizes)
  module_hubs <- vapply(
    seq_along(core_sizes),
    function(module) {
      module_nodes <- module_starts[module]:module_ends[module]
      module_nodes[which.max(igraph::degree(graph, v = module_nodes))]
    },
    numeric(1)
  )

  connector_nodes <- (n_core + 1):(n_core + n_connectors)
  peripheral_nodes <- (n_core + n_connectors + 1):n_total
  ecological_edges <- c()

  for (connector_node in connector_nodes) {
    connected_modules <- sample(seq_along(core_sizes), 2)
    for (module in connected_modules) {
      ecological_edges <- c(
        ecological_edges,
        connector_node,
        sample(module_starts[module]:module_ends[module], 2)
      )
    }
  }

  for (peripheral_index in seq_along(peripheral_nodes)) {
    peripheral_node <- peripheral_nodes[peripheral_index]
    if (peripheral_index <= 0.75 * length(peripheral_nodes)) {
      target_node <- sample(module_hubs, 1)
    } else {
      target_node <- sample(connector_nodes, 1)
    }

    ecological_edges <- c(ecological_edges, peripheral_node, target_node)
    if (runif(1) < 0.12) {
      ecological_edges <- c(
        ecological_edges,
        peripheral_node,
        sample(module_hubs, 1)
      )
    }
  }

  connect_components(igraph::add_edges(graph, ecological_edges))
}

g_ba <- igraph::sample_pa(
  n = n_genera,
  m = 2,
  directed = FALSE
)
module_sizes <- c(22, 20, 19, 18, 20, 18)
module_probabilities <- matrix(
  0.003,
  nrow = length(module_sizes),
  ncol = length(module_sizes)
)
diag(module_probabilities) <- c(0.24, 0.22, 0.23, 0.21, 0.22, 0.20)
g_sbm <- igraph::sample_sbm(
  n = n_genera,
  pref.matrix = module_probabilities,
  block.sizes = module_sizes,
  directed = FALSE,
  loops = FALSE
)
module_starts <- cumsum(c(1, head(module_sizes, -1)))
module_edges <- cbind(
  module_starts[-length(module_starts)],
  module_starts[-1]
)
g_sbm <- igraph::add_edges(g_sbm, t(module_edges))
g_sbm <- connect_components(g_sbm)
g_ecological <- sample_ecological_core_periphery()
g_er <- igraph::sample_gnm(
  n = n_genera,
  m = igraph::ecount(g_sbm),
  directed = FALSE,
  loops = FALSE
)

standard_networks <- list(
  "Erdos-Renyi" = g_er,
  "Barabasi-Albert" = g_ba,
  "Stochastic block model" = g_sbm,
  "Core-periphery ecological" = g_ecological
)

gcm_orbits <- c(0, 2, 5, 7, 8, 10, 11, 6, 9, 4, 1)

calc_standard_gcm <- function(graph) {
  adja <- igraph::as_adjacency_matrix(graph, sparse = FALSE)
  NetCoMi::calcGCM(as.matrix(adja), orbits = gcm_orbits)[["gcm"]]
}

standard_gcms <- lapply(standard_networks, calc_standard_gcm)
network_layouts <- lapply(
  standard_networks,
  igraph::layout_with_fr
)

plot_standard_network <- function(graph, layout, main) {
  plot(
    graph,
    layout = layout,
    vertex.size = 3.5,
    vertex.label = NA,
    vertex.color = "#009E73",
    vertex.frame.color = NA,
    edge.color = adjustcolor("gray20", alpha.f = 0.45),
    edge.width = 0.5,
    main = main
  )
}

plot_gcm_heatmap <- function(gcm, main) {
  gcm <- gcm[paste0("O", gcm_orbits), paste0("O", gcm_orbits)]
  orbit_labels <- sub("^O", "", rownames(gcm))
  heat_cols <- c(
    colorRampPalette(c("#2166AC", "white"))(101),
    colorRampPalette(c("white", "#E69F00", "#B2182B"))(101)[-1]
  )
  old_xpd <- par("xpd")
  on.exit(par(xpd = old_xpd), add = TRUE)

  image(
    x = seq(0.5, ncol(gcm) + 0.5, by = 1),
    y = seq(0.5, nrow(gcm) + 0.5, by = 1),
    z = t(gcm[nrow(gcm):1, ]),
    col = heat_cols,
    zlim = c(-1, 1),
    xlim = c(0.5, ncol(gcm) + 2.2),
    axes = FALSE,
    xlab = "",
    ylab = "",
    main = ""
  )
  axis(
    side = 3,
    at = seq_along(orbit_labels),
    labels = orbit_labels,
    tick = FALSE,
    line = -0.8,
    cex.axis = 0.9
  )
  axis(
    side = 2,
    at = seq_along(orbit_labels),
    labels = rev(orbit_labels),
    tick = FALSE,
    las = 1,
    line = -0.5,
    cex.axis = 0.9
  )
  title(main = main, line = 2.1)
  rect(0.5, 0.5, ncol(gcm) + 0.5, nrow(gcm) + 0.5)

  legend_x <- c(ncol(gcm) + 0.9, ncol(gcm) + 1.15)
  legend_y <- seq(0.5, nrow(gcm) + 0.5, length.out = length(heat_cols) + 1)
  image(
    x = legend_x,
    y = legend_y,
    z = matrix(seq(-1, 1, length.out = length(heat_cols)), nrow = 1),
    col = heat_cols,
    zlim = c(-1, 1),
    add = TRUE
  )
  text(
    x = ncol(gcm) + 1.8,
    y = seq(1, nrow(gcm), length.out = 5),
    labels = seq(-1, 1, by = 0.5),
    cex = 0.9
  )
}
```

```{r}
old_par <- par(no.readonly = TRUE)
```

### Erdos-Renyi network

The Erdos-Renyi network is used as a random baseline. Edges are placed without
an explicit preference for hubs or modules, so any local structure arises from
chance rather than from an imposed ecological organization.

```{r standard_network_gcms_ER, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Erdos-Renyi"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM shows broadly positive correlations among most graphlet orbits. This is
expected for a random graph of this size and density: nodes with higher degree
also tend to participate more often in many other small graphlet positions,
whereas clear opposing orbit roles are largely absent.

### Barabasi-Albert network

The Barabasi-Albert network represents a scale-free hub structure. New nodes
preferentially attach to already well-connected nodes, producing a few hubs and
many nodes with low to moderate degree.

```{r standard_network_gcms_BA, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Barabasi-Albert"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM has more visible block structure than the random baseline. Orbits
associated with hub participation are highly correlated with each other, while
orbits describing peripheral or degree-one roles form a second group. This
matches the interpretation of scale-free networks as being dominated by
hub-related graphlet dependencies.

### Stochastic block model

The stochastic block model represents a modular ecological network. Edges are
more likely within modules than between modules, mimicking groups of taxa that
share similar niches or co-occurrence patterns.

```{r standard_network_gcms_SB, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Stochastic block model"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM reflects the modular structure through correlated groups of orbits.
Dense within-module neighborhoods increase the dependence among graphlet roles
that occur in locally clustered regions, while between-module links contribute
less strongly. The result is more structured than the Erdos-Renyi baseline but
less hub-dominated than the Barabasi-Albert network.

### Core-periphery ecological network

The core-periphery ecological network combines dense module cores with connector
taxa and many peripheral taxa. This creates a contrast between taxa embedded in
locally dense communities and taxa attached as leaves or sparse connectors.

```{r standard_network_gcms_Eco, fig.width=8, fig.height=3.5, dpi=250}
network_name <- "Core-periphery ecological"

par(
  mfrow = c(1, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

plot_standard_network(
  standard_networks[[network_name]],
  network_layouts[[network_name]],
  paste(network_name, "network")
)
plot_gcm_heatmap(
  standard_gcms[[network_name]],
  paste(network_name, "GCM")
)
```

The GCM shows the clearest contrast between graphlet roles. Orbits 0, 2, 5, and
7 describe degree and internal or branching positions in open structures, while
orbits 8, 10, and 11 describe participation in cycle- or triangle-related
structures. In this example, these roles are concentrated in the module cores
and connector taxa. Their negative correlations with orbits 6 and 9 indicate
that nodes frequently participating in core, branching, or cyclic graphlets are
not usually the same nodes that occupy terminal graphlet positions. Conversely,
nodes with high counts for terminal roles tend to be peripheral taxa attached to
the network boundary and therefore have low counts for the denser core-related
orbits.

### Combined overview

The combined figure places the four network types and their GCMs next to each
other to make the structural differences easier to compare.

```{r standard_network_gcms, fig.width=8, fig.height=13, dpi=250}
par(
  mfrow = c(4, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

for (network_name in names(standard_networks)) {
  plot_standard_network(
    standard_networks[[network_name]],
    network_layouts[[network_name]],
    paste(network_name, "network")
  )
  plot_gcm_heatmap(
    standard_gcms[[network_name]],
    paste(network_name, "GCM")
  )
}
```

Across the four examples, the GCMs become more structured as the networks move
away from a homogeneous random baseline. The random network mainly shows
degree-driven positive orbit correlations, the Barabasi-Albert network
highlights hub-related dependencies, the stochastic block model emphasizes
module-related dependencies, and the core-periphery network introduces the
strongest separation between dense-core and peripheral graphlet roles.

The following plot is used for the thesis:

```{r standard_network_gcms_two_panels, fig.width=8, fig.height=6, dpi=250}
par(
  mfrow = c(2, 2),
  mar = c(0.5, 1.6, 3.8, 2.8),
  oma = c(0, 0, 0, 0)
)

for (network_name in c("Barabasi-Albert", "Core-periphery ecological")) {
  plot_standard_network(
    standard_networks[[network_name]],
    network_layouts[[network_name]],
    paste(network_name, "network")
  )
  plot_gcm_heatmap(
    standard_gcms[[network_name]],
    paste(network_name, "GCM")
  )
}

par(old_par)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
