---
title: "Build reduced data set"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
always_allow_html: true
params:
  name: "00_build_dataset" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "8ad93c89-8d43-4945-9813-3d10fac8034c")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/app_dacomp_"),
  echo = FALSE)

refit_permApprox <- FALSE
```

In this script, we load the original PASTURE data (not published) and build the reduced data set, which is the basis for all analyses conducted in this project. 

```{r packages}
library(conflicted)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(kableExtra)
library(pheatmap)
library(ggrepel)
library(patchwork)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data & set parameters

```{r load_data}
phylo_genus <- readRDS(path_source("PASTURE_data", paste0("bact_2m_genus.rds")))
phylo_asv <- readRDS(path_source("PASTURE_data", paste0("bact_2m_asv.rds")))

phylo_genus
phylo_asv
```

## Select variables of interest

```{r}
keep_countries <- c("Germany", "Switzerland", "Austria")
keep_vars      <- c("SampleID", "idx", "Country", "Sex", "cesarean",
                    "breast_dur_cat1", "breast_excl_cat1", "pregsmoke",
                    "sibs_numb_cat")
rename_vars    <- c("SampleID", "idx", "Country", "Sex", "C_section",
                    "BF_duration", "EBF_duration", "Smoking", "Siblings")

# Filter samples (rows) by country and variables (columns)
phylo_genus <- prune_samples(
  sample_data(phylo_genus)$Country %in% keep_countries,
  phylo_genus
)
sample_data(phylo_genus) <- sample_data(phylo_genus) %>%
  as("data.frame") %>%
  select(all_of(keep_vars)) %>%
  setNames(rename_vars)

phylo_asv <- prune_samples(
  sample_data(phylo_asv)$Country %in% keep_countries,
  phylo_asv
)
sample_data(phylo_asv) <- sample_data(phylo_asv) %>%
  as("data.frame") %>%
  select(all_of(keep_vars)) %>%
  setNames(rename_vars)

# Extract sampdata if needed downstream
sampdata <- sample_data(phylo_genus) %>% as("data.frame")
```

### ASV level

```{r}
phylo_asv
```


### Genus level

```{r}
phylo_genus
```

## Filter

Remove taxa absent from all samples, then remove samples with zero total counts.

```{r filter}
# Remove taxa with zero counts across all samples
phylo_asv   <- prune_taxa(taxa_sums(phylo_asv)   > 0, phylo_asv)
phylo_genus <- prune_taxa(taxa_sums(phylo_genus) > 0, phylo_genus)

# Remove samples with zero total counts
phylo_asv   <- prune_samples(sample_sums(phylo_asv)   > 0, phylo_asv)
phylo_genus <- prune_samples(sample_sums(phylo_genus) > 0, phylo_genus)
```

### ASV level

```{r}
phylo_asv
```

### Genus level

```{r}
phylo_genus
```

```{r}
# Save data
saveRDS(phylo_asv, path_target("bact_phylo_2m_asv.rds"))
saveRDS(phylo_genus, path_target("bact_phylo_2m_genus.rds"))
```

## Sample metadata with display labels

```{r}
# sampdata with display-friendly names (hyphen in C-section, underscores elsewhere)
sampdata_labels <- sampdata %>%
  rename(`C-section` = C_section,
         `BF duration` = BF_duration,
         `EBF duration` = EBF_duration) %>%
  mutate(
    Country     = factor(Country, levels = keep_countries),
    Sex         = factor(Sex, levels = c("f", "m"), labels = c("Female", "Male")),
    `C-section` = factor(`C-section`, levels = c(0, 1), labels = c("No", "Yes")),
    Smoking     = factor(Smoking, levels = c(0, 1), labels = c("No", "Yes"))
  )

saveRDS(sampdata_labels, path_target("sampdata.rds"))
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
