---
title: "Taxonomic aggregation and filtering"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "02_filtering" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "7426994c-47c1-4984-bc1e-aeb070dbe32b")
```


```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params, "/filter_"),
  echo = FALSE)
```

## Load packages

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(kableExtra)
library(ggtext)
library(patchwork)

conflicts_prefer(dplyr::filter)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data

```{r load_data}
phylo_asv   <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_asv.rds"))
phylo_genus <- readRDS(path_source("00_build_dataset", "bact_phylo_2m_genus.rds"))
sampdata    <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
otutab_asv  <- as(otu_table(phylo_asv), "matrix")
```


**Summary of overall sample counts:**

```{r}
summary(colSums(otutab_asv))
```


## Helper functions

```{r}
otu_matrix_samples_by_taxa <- function(phylo_obj) {
  mat <- otu_table(phylo_obj) |> 
    as("matrix")
  
  if (taxa_are_rows(phylo_obj)) {
    mat <- t(mat)
  }
  
  mat
}

zero_prop <- function(phylo_obj) {
  mat <- otu_matrix_samples_by_taxa(phylo_obj)
  mean(mat == 0)
}

taxon_prevalence <- function(phylo_obj) {
  mat <- otu_matrix_samples_by_taxa(phylo_obj)
  colMeans(mat > 0)
}

safe_tax_glom <- function(phylo_obj, taxrank) {
  if (!taxrank %in% rank_names(phylo_obj)) {
    return(NULL)
  }
  
  out <- phyloseq::tax_glom(
    phylo_obj,
    taxrank = taxrank,
    NArm = FALSE
  )
  
  out <- prune_taxa(taxa_sums(out) > 0, out)
  
  out
}

format_prevalence_threshold <- function(threshold) {
  if (threshold == 0) {
    "No prevalence filtering"
  } else {
    paste0("$\\geq ", threshold * 100, "\\%$ of samples")
  }
}
```


## Number of taxa across taxonomic levels

```{r}
tax_levels <- c("Phylum", "Class", "Order", "Family", "Genus")

taxonomic_overview_levels <- map_dfr(
  tax_levels,
  function(level) {
    phylo_level <- safe_tax_glom(phylo_asv, taxrank = level)
    
    if (is.null(phylo_level)) {
      tibble(
        taxonomic_level = level,
        n_taxa = NA_integer_,
        zero_entries = NA_real_
      )
    } else {
      tibble(
        taxonomic_level = level,
        n_taxa = ntaxa(phylo_level),
        zero_entries = 100 * zero_prop(phylo_level)
      )
    }
  }
)

taxonomic_overview <- bind_rows(
  taxonomic_overview_levels,
  tibble(
    taxonomic_level = "ASV",
    n_taxa = ntaxa(phylo_asv),
    zero_entries = 100 * zero_prop(phylo_asv)
  )
) |>
  mutate(
    zero_entries = round(zero_entries, 1)
  )

taxonomic_overview
```

```{r}
write_csv(
  taxonomic_overview,
  path_target("taxonomic_overview.csv")
)

taxonomic_overview_tex <- knitr::kable(
  taxonomic_overview,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Taxonomic level", "Number of taxa", "Zero entries (\\%)"),
  caption = "\\textbf{Number of taxa across taxonomic levels.} Number of taxa obtained at different taxonomic levels after restricting the data to the reduced analysis subset and removing taxa with zero total counts."
)

writeLines(
  taxonomic_overview_tex,
  path_target("taxonomic_overview.tex")
)
```

```{r}
knitr::kable(
  taxonomic_overview,
  col.names = c("Taxonomic level", "Number of taxa", "Zero entries (%)")
)
```

## Phylum-level relative abundance by covariate

```{r phylum-data-prep}
# Melt phylum-level phyloseq, keep top 5 phyla, lump rest as "Other"
phylum_melt <- phylo_asv %>%
  tax_glom("Phylum") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt() %>%
  mutate(Phylum = str_remove(Phylum, "^p__"))

top5_phyla <- phylum_melt %>%
  group_by(Phylum) %>%
  summarise(m = mean(Abundance), .groups = "drop") %>%
  slice_max(m, n = 5) %>%
  pull(Phylum)

# Merge with nicely-labelled metadata
phylum_meta <- phylum_melt %>%
  mutate(Phylum = if_else(Phylum %in% top5_phyla, Phylum, "Other")) %>%
  group_by(Sample, Phylum) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  left_join(
    sampdata %>%
      rownames_to_column("Sample"),
    by = "Sample"
  )
```

```{r phylum-plot-function}
phylum_colors <- c(
  Actinobacteriota  = "#44BFA3",
  Bacteroidota      = "#F28E2B",
  Firmicutes        = "#7B9FC7",
  Proteobacteria    = "#F1959B",
  Verrucomicrobiota = "#8DB86B",
  Other             = "#F2E750"
)

# Stack order: Actinobacteriota on top, matching reference style
phylum_stack_levels <- c("Firmicutes", "Proteobacteria", "Bacteroidota",
                         "Verrucomicrobiota", "Other", "Actinobacteriota")

phylum_labels <- c(
  Actinobacteriota  = "*Actinobacteriota*",
  Bacteroidota      = "*Bacteroidota*",
  Firmicutes        = "*Firmicutes*",
  Proteobacteria    = "*Proteobacteria*",
  Verrucomicrobiota = "*Verrucomicrobiota*",
  Other             = "Other"
)

plot_phylum_by_group <- function(data, var, x_label) {
  data %>%
    dplyr::filter(!is.na(.data[[var]])) %>%
    group_by(.data[[var]], Phylum) %>%
    summarise(mean_abund = mean(Abundance), .groups = "drop") %>%
    mutate(Phylum = factor(Phylum, levels = phylum_stack_levels)) %>%
    ggplot(aes(x = .data[[var]], y = mean_abund, fill = Phylum)) +
    geom_col(width = 0.6) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    scale_fill_manual(values = phylum_colors, labels = phylum_labels) +
    labs(x = x_label, y = "Mean Relative Abundance (%)", fill = "Phylum",
         title = x_label) +
    theme(legend.text = element_markdown(),
          legend.position = "bottom")
}
```

```{r phylum-by-covariate, fig.width=12, fig.height=12}
vars <- list(
  list(var = "Country",      label = "Country"),
  list(var = "Sex",          label = "Sex"),
  list(var = "C-section",    label = "C-section"),
  list(var = "BF duration",  label = "BF duration"),
  list(var = "EBF duration", label = "EBF duration"),
  list(var = "Smoking",      label = "Smoking"),
  list(var = "Siblings",     label = "Siblings")
)

plots <- lapply(vars, function(v) plot_phylum_by_group(phylum_meta, v$var, v$label))

# Remove y-axis from all but the leftmost column (every 3rd plot starting at index 1)
plots <- lapply(seq_along(plots), function(i) {
  if (i %% 3 == 1) plots[[i]]
  else plots[[i]] + theme(axis.title.y = element_blank(),
                          axis.text.y  = element_blank(),
                          axis.ticks.y = element_blank())
})

wrap_plots(plots, ncol = 3, guides = "collect") +
  plot_annotation(
    title = "Phylum-level composition by covariate",
    theme = theme(plot.title = element_text(size = 16))
  ) &
  theme(
    legend.position   = "right",
    legend.title      = element_text(size = 14),
    legend.text       = element_markdown(size = 14),
    axis.title        = element_text(size = 13),
    axis.text         = element_text(size = 12),
    plot.title        = element_text(size = 14)
  )
```

## ASV prevalence vs. total abundance

```{r prevalence_text_asv, fig.width=7, fig.height=4}
data.frame(
  prevalence   = rowSums(otutab_asv > 0) / ncol(otutab_asv),
  total_counts = rowSums(otutab_asv)
) %>%
  ggplot(aes(x = prevalence, y = total_counts)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x        = "Prevalence (fraction of samples)",
    y        = "Total counts",
    title    = "ASV prevalence vs. total abundance",
    subtitle = paste0(
      sum(rowSums(otutab_asv > 0) != 0), " ASVs in total; ",
      sum(rowSums(otutab_asv > 0) / ncol(otutab_asv) < 0.1),
      " ASVs present in <10% of samples"
    )
  ) +
  theme(
    axis.title        = element_text(size = 13),
    axis.text         = element_text(size = 11),
    plot.title        = element_text(size = 14)
  ) +
  theme_bw()
```

```{r prevalence_asv, fig.width=7, fig.height=2.5}
data.frame(
  prevalence   = rowSums(otutab_asv > 0) / ncol(otutab_asv),
  total_counts = rowSums(otutab_asv)
) %>%
  ggplot(aes(x = prevalence, y = total_counts)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x        = "Prevalence (fraction of samples)",
    y        = "Total counts"
  ) +
  theme_bw() +
  theme(
    axis.title        = element_text(size = 13),
    axis.text         = element_text(size = 12),
    plot.title        = element_text(size = 14)
  )
```

## Most abundant genera

```{r}
genus_mat <- otu_matrix_samples_by_taxa(phylo_genus)

genus_rel_mat <- genus_mat / rowSums(genus_mat)

genus_tax <- tax_table(phylo_genus) |> 
  as("matrix") |> 
  as.data.frame() |> 
  rownames_to_column("taxon_id")

genus_labels <- genus_tax |>
  transmute(
    taxon_id,
    genus = if_else(
      is.na(Genus) | Genus == "",
      taxon_id,
      Genus
    )
  )

top_genera <- tibble(
  taxon_id = colnames(genus_mat),
  mean_rel_abundance = colMeans(genus_rel_mat, na.rm = TRUE) * 100,
  prevalence = colMeans(genus_mat > 0) * 100
) |>
  left_join(genus_labels, by = "taxon_id") |>
  arrange(desc(mean_rel_abundance)) |>
  slice_head(n = 10) |>
  transmute(
    genus,
    mean_rel_abundance = round(mean_rel_abundance, 1),
    prevalence = round(prevalence, 1)
  )

top_genera
```

```{r}
write_csv(
  top_genera,
  path_target("top_genera.csv")
)

top_genera_tex <- top_genera |>
  mutate(
    genus = paste0("\\emph{", genus, "}")
  ) |>
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    col.names = c("Genus", "Mean rel. abundance (\\%)", "Prevalence (\\%)"),
    caption = "\\textbf{Most abundant genera.} Genera in the reduced PASTURE data set ranked by mean relative abundance across samples. For each genus, the mean relative abundance and prevalence are reported."
  )

writeLines(
  top_genera_tex,
  path_target("top_genera.tex")
)
```

```{r}
top_genera |>
  knitr::kable(
    col.names = c("Genus", "Mean rel. abundance (%)", "Prevalence (%)")
  )
```

## Prevalence filtering at genus level

For all analyses except alpha diversity, genera observed in less than 1% of samples are removed. The table below summarizes the effect of different prevalence thresholds on the number of retained genera and the proportion of zero entries.

```{r}
prevalence_thresholds <- c(0, 0.01, 0.05, 0.10, 0.20)

genus_prev <- taxon_prevalence(phylo_genus)

prevalence_filtering <- map_dfr(
  prevalence_thresholds,
  function(threshold) {
    keep_taxa <- names(genus_prev)[genus_prev >= threshold]
    
    phylo_filtered <- prune_taxa(
      keep_taxa,
      phylo_genus
    )
    
    tibble(
      prevalence_threshold = threshold,
      prevalence_threshold_label = format_prevalence_threshold(threshold),
      n_genera = ntaxa(phylo_filtered),
      zero_entries = 100 * zero_prop(phylo_filtered)
    )
  }
) |>
  mutate(
    zero_entries = round(zero_entries, 1)
  )

prevalence_filtering
```

```{r}
write_csv(
  prevalence_filtering,
  path_target("prevalence_filtering.csv")
)

prevalence_filtering_tex <- prevalence_filtering |>
  select(
    prevalence_threshold_label,
    n_genera,
    zero_entries
  ) |>
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    col.names = c("Prevalence threshold", "Number of genera", "Zero entries (\\%)"),
    caption = "\\textbf{Effect of prevalence filtering on the genus-level count table.} Number of retained genera and proportion of zero entries for different prevalence thresholds. Prevalence is defined as the proportion of samples in which a genus has positive counts."
  )

writeLines(
  prevalence_filtering_tex,
  path_target("prevalence_filtering.tex")
)
```

```{r}
prevalence_filtering |>
  select(
    prevalence_threshold_label,
    n_genera,
    zero_entries
  ) |>
  knitr::kable(
    col.names = c("Prevalence threshold", "Number of genera", "Zero entries (%)")
  )
```

```{r}
genus_prev_threshold <- 0.01

phylo_genus_prev10 <- prune_taxa(
  names(genus_prev)[genus_prev >= genus_prev_threshold],
  phylo_genus
)

saveRDS(
  phylo_genus,
  path_target("bact_phylo_2m_genus_unfiltered.rds")
)

saveRDS(
  phylo_genus_prev10,
  path_target("bact_phylo_2m_genus_prev01.rds")
)
```

```{r zero_fraction_plot, fig.width=7, fig.height=2.5, dpi = 200}
count_mat <- get_sample_by_taxa(phylo_genus_prev10)

tibble(
  genus = colnames(count_mat),
  zero_fraction = colMeans(count_mat == 0)
) |>
  ggplot(aes(x = zero_fraction)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  scale_x_continuous(
    labels = scales::percent_format(),
    breaks = seq(0, 1, by = 0.1)
  ) +
  labs(
    x = "Zero fraction",
    y = "Number of genera"
  ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 13)
  ) 
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```


