---
title: "Regression models for compositional responses"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "05_regression" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "9b9d6ec0-81d9-43a9-9e04-c99b38578ef2")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/regression_"),
  echo = FALSE)
```
```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(permApprox)
library(kableExtra)
library(furrr)
library(future)
library(progressr)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
n_perm <- 999
permutation_seed <- 42
cores <- 10

covariates_regression <- c(
  "Country", "Sex", "C-section", "BF duration",
  "EBF duration", "Smoking", "Siblings"
)
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

Counts are first transformed to relative abundances. Zeros are then replaced with
the multiplicative replacement implemented in `zCompositions::multRepl()`, using
the global minimum non-zero relative abundance as detection limit and
`frac = 0.65`. The CLR transformation is applied after replacement.

```{r helper_functions}
get_sample_by_taxa <- function(phylo) {
  mat <- as(otu_table(phylo), "matrix")
  
  if (taxa_are_rows(phylo)) {
    mat <- t(mat)
  }
  
  mat <- as.matrix(mat)
  storage.mode(mat) <- "numeric"
  
  mat
}

counts_to_rel_abund <- function(count_mat) {
  lib_size <- rowSums(count_mat)
  
  if (any(lib_size == 0)) {
    stop("At least one sample has library size zero.")
  }
  
  sweep(count_mat, 1, lib_size, "/")
}

compute_multrepl_clr <- function(rel_abund_mat, frac = 0.65) {
  positive_values <- rel_abund_mat[rel_abund_mat > 0]
  
  if (length(positive_values) == 0) {
    stop("No positive relative abundances found.")
  }
  
  detection_limit <- min(positive_values, na.rm = TRUE)
  detection_limit_mat <- matrix(
    detection_limit,
    nrow = nrow(rel_abund_mat),
    ncol = ncol(rel_abund_mat),
    dimnames = dimnames(rel_abund_mat)
  )
  
  rel_abund_repl <- zCompositions::multRepl(
    rel_abund_mat,
    label = 0,
    dl = detection_limit_mat,
    frac = frac,
    z.warning = 0.9,
    z.delete = FALSE
  )
  
  rel_abund_repl <- as.matrix(rel_abund_repl)
  rownames(rel_abund_repl) <- rownames(rel_abund_mat)
  colnames(rel_abund_repl) <- colnames(rel_abund_mat)
  
  log_repl <- log(rel_abund_repl)
  z <- sweep(log_repl, 1, rowMeans(log_repl), "-")
  
  list(
    z = z,
    rel_abund_repl = rel_abund_repl,
    detection_limit = detection_limit,
    replacement_value = frac * detection_limit,
    frac = frac
  )
}

safe_reformulate <- function(vars) {
  if (length(vars) == 0) {
    return(stats::as.formula("~ 1"))
  }
  
  reformulate(sprintf("`%s`", vars))
}

make_regression_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country        = factor(Country),
      Sex            = factor(Sex),
      `C-section`    = factor(`C-section`),
      `BF duration`  = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", "\u22652 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", "\u22652 months")
      ),
      Smoking        = factor(Smoking),
      Siblings       = factor(Siblings)
    )
}

multivariate_rss <- function(y, design_mat) {
  fit <- lm.fit(x = design_mat, y = y)
  sum(fit$residuals^2)
}

regression_stat <- function(y, metadata, full_formula, reduced_formula) {
  full_design <- model.matrix(full_formula, data = metadata)
  red_design  <- model.matrix(reduced_formula, data = metadata)
  
  rss_red  <- multivariate_rss(y = y, design_mat = red_design)
  rss_full <- multivariate_rss(y = y, design_mat = full_design)
  
  rss_red - rss_full
}

empirical_pvalue <- function(obs_stat, perm_stats) {
  (1 + sum(perm_stats >= obs_stat, na.rm = TRUE)) / (1 + length(perm_stats))
}

run_regression_permutation_test <- function(
    y,
    metadata,
    covariate,
    all_covariates,
    n_perm = 999
) {
  full_formula <- safe_reformulate(all_covariates)
  reduced_covariates <- setdiff(all_covariates, covariate)
  reduced_formula <- safe_reformulate(reduced_covariates)
  
  obs_stat <- regression_stat(
    y = y,
    metadata = metadata,
    full_formula = full_formula,
    reduced_formula = reduced_formula
  )
  
  perm_stats <- replicate(
    n_perm,
    {
      metadata_perm <- metadata
      metadata_perm[[covariate]] <- sample(
        metadata_perm[[covariate]],
        size = nrow(metadata_perm),
        replace = FALSE
      )
      
      regression_stat(
        y = y,
        metadata = metadata_perm,
        full_formula = full_formula,
        reduced_formula = reduced_formula
      )
    }
  )
  
  tibble(
    variable = covariate,
    statistic_obs = obs_stat,
    p_empirical = empirical_pvalue(obs_stat, perm_stats),
    n_perm = n_perm,
    n_exceed = sum(perm_stats >= obs_stat, na.rm = TRUE),
    perm_stats = list(as.numeric(perm_stats))
  )
}

sig_stars <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ ""
  )
}

fmt_p_latex <- function(p_num) {
  p_str <- formatC(p_num, format = "e", digits = 2)
  stars <- sig_stars(p_num)
  
  ifelse(stars != "", paste0(p_str, "$^{", stars, "}$"), p_str)
}

fmt_p_table <- function(p_num) {
  formatC(p_num, format = "e", digits = 2)
}

format_coefficient_label <- function(term) {
  term <- stringr::str_remove_all(term, "`")
  
  dplyr::case_when(
    stringr::str_starts(term, "Country") ~ paste0(
      "Country: ",
      stringr::str_remove(term, "^Country")
    ),
    stringr::str_starts(term, "Sex") ~ paste0(
      "Sex: ",
      stringr::str_to_lower(stringr::str_remove(term, "^Sex"))
    ),
    stringr::str_starts(term, "C-section") ~ paste0(
      "C-section: ",
      stringr::str_remove(term, "^C-section")
    ),
    stringr::str_starts(term, "BF duration") ~ paste0(
      "BF duration: ",
      stringr::str_remove(term, "^BF duration")
    ),
    stringr::str_starts(term, "EBF duration") ~ paste0(
      "EBF duration: ",
      stringr::str_remove(term, "^EBF duration")
    ),
    stringr::str_starts(term, "Smoking") ~ paste0(
      "Smoking: ",
      stringr::str_remove(term, "^Smoking")
    ),
    stringr::str_starts(term, "Siblings") ~ paste0(
      "Siblings: ",
      stringr::str_remove(term, "^Siblings")
    ),
    TRUE ~ term
  )
}
```

## Prepare CLR response matrix

```{r prepare_clr_response}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_reg <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_regression_metadata()

counts_raw <- counts_raw[sampdata_reg$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_reg$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

regression_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

regression_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("regression_multrepl_clr_object.rds")
)

write_csv(
  regression_preprocessing_summary,
  file = path_target("regression_preprocessing_summary.csv")
)
```

## Regression analysis setup

Samples with `EBF duration == "1 month"` are removed before fitting the
regression models, because this sparse group contains only three samples and may
confound the regression-based global tests.

All covariates are tested on the same complete-case analysis set. That is, after
removing the sparse EBF group, samples with missing values in any of the model
covariates are excluded once, and the resulting sample set is used for all
full-versus-reduced model comparisons. This ensures that differences in the test
statistics reflect the covariate being omitted rather than changes in the sample
composition. The number of taxa is fixed by the filtered genus-level phyloseq
object used to construct the CLR response matrix.

```{r regression_setup}
model_data <- sampdata_reg %>%
  filter(`EBF duration` != "1 month" | is.na(`EBF duration`)) %>%
  select(SampleID, all_of(covariates_regression)) %>%
  filter(if_all(all_of(covariates_regression), ~ !is.na(.x))) %>%
  droplevels()

model_sample_ids <- model_data$SampleID
response_clr <- counts_clr[model_sample_ids, , drop = FALSE]

stopifnot(identical(rownames(response_clr), model_data$SampleID))

covariates_regression <- covariates_regression[
  vapply(
    covariates_regression,
    function(var) length(unique(model_data[[var]])) >= 2,
    logical(1)
  )
]

model_data <- model_data %>%
  select(SampleID, all_of(covariates_regression)) %>%
  droplevels()

regression_model_summary <- tibble(
  n_samples = nrow(model_data),
  n_taxa = ncol(response_clr),
  excluded_ebf_one_month = sum(sampdata_reg$`EBF duration` == "1 month", na.rm = TRUE),
  n_covariates = length(covariates_regression)
)

covariate_level_summary <- tibble(
  variable = covariates_regression,
  n_levels = vapply(
    covariates_regression,
    function(var) nlevels(model_data[[var]]),
    integer(1)
  ),
  levels = vapply(
    covariates_regression,
    function(var) paste(levels(model_data[[var]]), collapse = "; "),
    character(1)
  )
)

regression_model_summary
covariate_level_summary
```

Thus, the multivariate regression model is fitted to
`r regression_model_summary$n_samples` samples and
`r regression_model_summary$n_taxa` genus-level taxa for every covariate test.

```{r save_regression_setup}
write_csv(
  regression_model_summary,
  file = path_target("regression_model_summary.csv")
)

write_csv(
  covariate_level_summary,
  file = path_target("regression_covariate_level_summary.csv")
)
```

## CLR-based multivariate regression

The fitted model uses the CLR-transformed microbial profile as multivariate
response and compares the full model with reduced models in which one covariate
is omitted. For each covariate, the test statistic is the reduction in
multivariate residual sum of squares.

```{r full_model_coefficients}
full_formula <- safe_reformulate(covariates_regression)
full_design <- model.matrix(full_formula, data = model_data)
full_fit <- lm.fit(x = full_design, y = response_clr)

coefficient_matrix <- as_tibble(
  full_fit$coefficients,
  rownames = "coefficient"
)

saveRDS(
  list(
    formula = full_formula,
    design = full_design,
    coefficients = full_fit$coefficients,
    taxa = colnames(response_clr),
    sample_ids = model_data$SampleID
  ),
  file = path_target("regression_full_model_fit.rds")
)

coefficient_matrix[seq_len(min(10, nrow(coefficient_matrix))), seq_len(min(6, ncol(coefficient_matrix)))]
```

### Coefficient heatmap

The full multivariate regression yields one coefficient for each model term and
CLR coordinate. The heatmap below visualizes these coefficients as a compact
overview of the fitted multivariate effect pattern. Rows correspond to the model
matrix terms; for categorical variables, these are contrasts relative to the
reference level. Columns correspond to genus-level taxa. Red values indicate a
positive coefficient on the CLR scale, whereas blue values indicate a negative
coefficient. The heatmap is descriptive and is not used for taxon-level
inference in this community-level analysis.

```{r coefficient_heatmap, fig.width=14, fig.height=6}
coef_mat <- full_fit$coefficients
coef_mat <- coef_mat[rownames(coef_mat) != "(Intercept)", , drop = FALSE]

tax_labels <- colnames(coef_mat)
tax_tab <- tryCatch(
  as(tax_table(phylo_genus), "matrix"),
  error = function(e) NULL
)

if (!is.null(tax_tab) && "Genus" %in% colnames(tax_tab)) {
  common_taxa <- intersect(colnames(coef_mat), rownames(tax_tab))
  genus_labels <- as.character(tax_tab[common_taxa, "Genus"])
  genus_labels <- stringr::str_remove(genus_labels, "^g__")
  
  missing_label <- is.na(genus_labels) |
    genus_labels == "" |
    genus_labels == "NA"
  
  genus_labels[missing_label] <- common_taxa[missing_label]
  tax_labels[match(common_taxa, colnames(coef_mat))] <- genus_labels
}

tax_labels <- make.unique(tax_labels)

coef_plot_mat <- coef_mat
colnames(coef_plot_mat) <- tax_labels
rownames(coef_plot_mat) <- format_coefficient_label(rownames(coef_plot_mat))

row_order <- if (nrow(coef_plot_mat) > 1) {
  hclust(dist(coef_plot_mat))$order
} else {
  seq_len(nrow(coef_plot_mat))
}

col_order <- if (ncol(coef_plot_mat) > 1) {
  hclust(dist(t(coef_plot_mat)))$order
} else {
  seq_len(ncol(coef_plot_mat))
}

coef_heatmap_df <- as_tibble(
  coef_plot_mat[row_order, col_order, drop = FALSE],
  rownames = "coefficient"
) %>%
  pivot_longer(
    cols = -coefficient,
    names_to = "taxon",
    values_to = "coefficient_value"
  ) %>%
  mutate(
    coefficient = factor(
      coefficient,
      levels = rev(rownames(coef_plot_mat)[row_order])
    ),
    taxon = factor(
      taxon,
      levels = colnames(coef_plot_mat)[col_order]
    )
  )

coef_limit <- max(abs(coef_heatmap_df$coefficient_value), na.rm = TRUE)

ggplot(coef_heatmap_df, aes(x = taxon, y = coefficient, fill = coefficient_value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-coef_limit, coef_limit),
    name = "Coefficient"
  ) +
  labs(
    x = "",
    y = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, 
                               hjust = 1, size = 10),
    axis.text.y = element_text(size = 14),
    #axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 14),
    legend.position = "right"
  )

coef_heatmap_df %>%
  mutate(
    coefficient = as.character(coefficient),
    taxon = as.character(taxon)
  ) %>%
  write_csv(path_target("regression_coefficient_heatmap_data.csv"))
```

```{r coefficient_heatmap_top20, fig.width=10, fig.height=6}
top_taxa_df <- coef_heatmap_df %>%
  group_by(taxon) %>%
  summarise(
    max_abs_coefficient = max(abs(coefficient_value), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  slice_max(
    order_by = max_abs_coefficient,
    n = 20,
    with_ties = FALSE
  )

top_taxa <- top_taxa_df %>%
  pull(taxon)

top20_coef_mat <- coef_plot_mat[, as.character(top_taxa), drop = FALSE]

top20_row_order <- if (nrow(top20_coef_mat) > 1) {
  hclust(dist(top20_coef_mat))$order
} else {
  seq_len(nrow(top20_coef_mat))
}

# Keep taxa ordered by decreasing maximum absolute coefficient.
top20_col_order <- seq_along(top_taxa)

coef_heatmap_top20_df <- as_tibble(
  top20_coef_mat[top20_row_order, top20_col_order, drop = FALSE],
  rownames = "coefficient"
) %>%
  pivot_longer(
    cols = -coefficient,
    names_to = "taxon",
    values_to = "coefficient_value"
  ) %>%
  mutate(
    coefficient = factor(
      coefficient,
      levels = rev(rownames(top20_coef_mat)[top20_row_order])
    ),
    taxon = factor(
      taxon,
      levels = colnames(top20_coef_mat)[top20_col_order]
    )
  )

top20_coef_limit <- max(abs(coef_heatmap_top20_df$coefficient_value), na.rm = TRUE)

ggplot(coef_heatmap_top20_df, aes(x = taxon, y = coefficient, fill = coefficient_value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-top20_coef_limit, top20_coef_limit),
    name = "Coefficient"
  ) +
  labs(
    x = "",
    y = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, 
                               hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    #axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 14),
    legend.position = "right"
  )

coef_heatmap_top20_df %>%
  mutate(
    coefficient = as.character(coefficient),
    taxon = as.character(taxon)
  ) %>%
  write_csv(path_target("regression_coefficient_heatmap_top20_data.csv"))
```

## Permutation-based global covariate tests

```{r parallel_setup}
future::plan(
  future::multisession,
  workers = cores
)

progressr::handlers("txtprogressbar")
options(progressr.enable = TRUE)
```

```{r regression_permutation_tests}
set.seed(permutation_seed)
path_regression <- path_target("regression_results_multrepl_clr.rds")

if (!file.exists(path_regression)) {
  regression_results <- progressr::with_progress({
    p <- progressr::progressor(steps = length(covariates_regression))
    
    furrr::future_map_dfr(
      covariates_regression,
      function(covariate) {
        out <- run_regression_permutation_test(
          y = response_clr,
          metadata = model_data,
          covariate = covariate,
          all_covariates = covariates_regression,
          n_perm = n_perm
        )
        
        p(message = sprintf("Regression test: %s", covariate))
        
        out
      },
      .options = furrr::furrr_options(seed = permutation_seed)
    )
  })
  
  saveRDS(regression_results, file = path_regression)
} else {
  regression_results <- readRDS(path_regression)
}

regression_results <- regression_results %>%
  mutate(
    variable = factor(variable, levels = covariates_regression),
    n_samples = nrow(model_data),
    n_taxa = ncol(response_clr),
    .after = variable
  )

regression_results %>%
  select(variable, n_samples, n_taxa, statistic_obs, p_empirical, n_exceed, n_perm) %>%
  mutate(
    p_empirical = fmt_p_table(p_empirical)
  )
```

```{r regression_results_csv}
regression_results %>%
  select(variable, n_samples, n_taxa, statistic_obs, p_empirical, n_exceed, n_perm) %>%
  write_csv(path_target("regression_results_multrepl_clr.csv"))
```

### Permutation distributions

The following plot shows a subset of the empirical null distributions obtained
from the permutation statistics. Six covariate-specific distributions are shown
to provide a compact visual check of the permutation results.

```{r regression_perm_dist, fig.width=8, fig.height=4}
plot_variables <- regression_results %>%
  arrange(variable) %>%
  slice_head(n = 6) %>%
  pull(variable)

perm_dist_df <- regression_results %>%
  #filter(variable %in% plot_variables) %>%
  select(variable, perm_stats) %>%
  mutate(
    perm_id = purrr::map(perm_stats, seq_along)
  ) %>%
  tidyr::unnest(c(perm_stats, perm_id)) %>%
  rename(statistic_perm = perm_stats)

obs_df <- regression_results %>%
  #filter(variable %in% plot_variables) %>%
  select(variable, statistic_obs, p_empirical)

ggplot(perm_dist_df, aes(x = statistic_perm)) +
  geom_histogram(bins = 35, fill = "grey80", color = "white") +
  geom_vline(
    data = obs_df,
    aes(xintercept = statistic_obs),
    color = "#B2182B",
    linewidth = 0.8
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    x = expression(t[j]^"*"),
    y = "Number of permutations"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

### p-value refinement with permApprox

```{r regression_permapprox}
path_regression_permapprox <- path_target("regression_permapprox_results_multrepl_clr.rds")

regression_obs_stats <- regression_results$statistic_obs
regression_perm_stats <- do.call(cbind, regression_results$perm_stats)

gpd_ctrl <- permApprox::make_gpd_ctrl(
  constraint = "support_at_obs",
  sample_size = min(regression_results$n_samples)
)

if (!file.exists(path_regression_permapprox)) {
  regression_permapprox <- permApprox::perm_approx(
    obs_stats = regression_obs_stats,
    perm_stats = regression_perm_stats,
    method = "gpd",
    alternative = "greater",
    null_center = "median",
    approx_thresh = 0.1,
    adjust_method = "none",
    gpd_ctrl = gpd_ctrl,
    cores = cores,
    verbose = TRUE
  )
  
  saveRDS(regression_permapprox, file = path_regression_permapprox)
} else {
  regression_permapprox <- readRDS(path_regression_permapprox)
}

regression_permapprox
```

```{r regression_permapprox_table}
regression_results_all <- regression_results %>%
  select(variable, n_samples, n_taxa, statistic_obs, n_exceed, n_perm, p_empirical) %>%
  mutate(
    p_permapprox = regression_permapprox$p_unadjusted,
    method_used = regression_permapprox$method_used
  )

regression_results_all %>%
  arrange(p_permapprox) %>%
  mutate(
    p_empirical = fmt_p_table(p_empirical),
    p_permapprox = fmt_p_table(p_permapprox)
  )
```

```{r regression_latex}
regression_tex_df <- regression_results_all %>%
  arrange(variable) %>%
  transmute(
    Variable = as.character(variable),
    RSS = round(statistic_obs, 2),
    n_exceed = n_exceed,
    p_emp = fmt_p_latex(p_empirical),
    p_papp = fmt_p_latex(p_permapprox)
  )

tex_regression <- kbl(
  regression_tex_df,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Variable", "$t_j$", "\\# exceed",
                "$p$ (emp.)", "$p$ (GPD-refined)"),
  align = "lrrrr",
  caption = paste(
    "CLR-based multivariate regression results for cohort variables.",
    "$t_j$ is the reduction in multivariate residual sum of squares when the",
    "covariate is added to the model.",
    "\\# exceed: number of permuted statistics $\\geq$ observed.",
    "$p$ (emp.): empirical permutation $p$-value ($999$ permutations);",
    "$p$ (GPD-refined): $p$-value refined with \\texttt{permApprox}."
  ),
  label = "application:regression"
) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(
    general = "$^{***}\\:p<0.001$; $^{**}\\:p<0.01$; $^{*}\\:p<0.05$; $^{.}\\:p<0.1$.",
    general_title = "",
    escape = FALSE,
    threeparttable = TRUE
  )

tex_regression
save_kable(tex_regression, file = path_target("regression_table.tex"))
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
