---
title: "Compositional equivalence"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "06_comp_equivalence" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "312648b4-8c26-4e76-aff7-ab6062b7ba2e")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/comp_equi_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(permApprox)
library(kableExtra)
library(furrr)
library(future)
library(progressr)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
n_perm <- 999
permutation_seed <- 42
alpha_perm_approx <- 0.05
cores <- 10

contrast_order <- c(
  "C-section",
  "BF duration",
  "EBF duration",
  "Sex",
  "Smoking",
  "Siblings"
)

diagnostic_contrasts <- c("C-section", "BF duration", "EBF duration")
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

Counts are transformed to relative abundances, zeros are replaced by
multiplicative replacement, and the CLR transformation is applied after
replacement. This follows the same preprocessing strategy as the beta-diversity
and regression analyses. The shared preprocessing helpers are defined in
`functions.R`.

```{r helper_functions}
make_comp_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

pooled_coordinate_variance <- function(y1, y2) {
  n1 <- nrow(y1)
  n2 <- nrow(y2)
  
  ((n1 - 1) * apply(y1, 2, stats::var) +
    (n2 - 1) * apply(y2, 2, stats::var)) / (n1 + n2 - 2)
}

comp_equiv_components <- function(y, group) {
  group <- droplevels(factor(group))
  
  if (nlevels(group) != 2) {
    stop("Compositional equivalence requires exactly two groups.")
  }
  
  group_levels <- levels(group)
  y1 <- y[group == group_levels[[1]], , drop = FALSE]
  y2 <- y[group == group_levels[[2]], , drop = FALSE]
  
  n1 <- nrow(y1)
  n2 <- nrow(y2)
  
  if (n1 < 2 || n2 < 2) {
    stop("Both groups must contain at least two samples.")
  }
  
  mean1 <- colMeans(y1)
  mean2 <- colMeans(y2)
  mean_diff <- mean1 - mean2
  pooled_var <- pooled_coordinate_variance(y1, y2)
  
  valid_taxa <- is.finite(pooled_var) & pooled_var > .Machine$double.eps
  scale_factor <- n1 * n2 / (n1 + n2)
  
  standardized_diff <- rep(NA_real_, length(mean_diff))
  standardized_diff[valid_taxa] <- mean_diff[valid_taxa] / sqrt(pooled_var[valid_taxa])
  
  coordinate_stat <- scale_factor * standardized_diff^2
  statistic <- max(coordinate_stat[valid_taxa], na.rm = TRUE)
  
  list(
    statistic = statistic,
    n1 = n1,
    n2 = n2,
    group1 = group_levels[[1]],
    group2 = group_levels[[2]],
    n_taxa = ncol(y),
    n_taxa_tested = sum(valid_taxa),
    n_taxa_zero_variance = sum(!valid_taxa),
    mean_diff = mean_diff,
    pooled_var = pooled_var,
    standardized_diff = standardized_diff,
    coordinate_stat = coordinate_stat,
    valid_taxa = valid_taxa
  )
}

run_comp_equiv_permutation_test <- function(
    y,
    metadata,
    variable,
    group_levels,
    contrast_label,
    n_perm = 999
) {
  test_data <- metadata %>%
    select(SampleID, group = all_of(variable)) %>%
    filter(!is.na(group), group %in% group_levels) %>%
    mutate(group = factor(group, levels = group_levels)) %>%
    droplevels()
  
  y_sub <- y[test_data$SampleID, , drop = FALSE]
  
  stopifnot(identical(rownames(y_sub), test_data$SampleID))
  
  obs <- comp_equiv_components(y = y_sub, group = test_data$group)
  
  perm_stats <- replicate(
    n_perm,
    {
      group_perm <- sample(test_data$group, size = nrow(test_data), replace = FALSE)
      comp_equiv_components(y = y_sub, group = group_perm)$statistic
    }
  )
  
  tibble(
    variable = variable,
    contrast = contrast_label,
    group1 = obs$group1,
    group2 = obs$group2,
    n1 = obs$n1,
    n2 = obs$n2,
    n_samples = obs$n1 + obs$n2,
    n_taxa = obs$n_taxa,
    n_taxa_tested = obs$n_taxa_tested,
    n_taxa_zero_variance = obs$n_taxa_zero_variance,
    statistic_obs = obs$statistic,
    p_empirical = empirical_pvalue(obs$statistic, perm_stats),
    n_exceed = sum(perm_stats >= obs$statistic, na.rm = TRUE),
    n_perm = n_perm,
    perm_stats = list(as.numeric(perm_stats)),
    components = list(obs)
  )
}

taxon_labels_from_phyloseq <- function(phylo) {
  tax_tab <- tryCatch(
    as(tax_table(phylo), "matrix"),
    error = function(e) NULL
  )
  
  tax_ids <- taxa_names(phylo)
  tax_labels <- tax_ids
  
  if (!is.null(tax_tab) && "Genus" %in% colnames(tax_tab)) {
    genus <- as.character(tax_tab[tax_ids, "Genus"])
    genus <- stringr::str_remove(genus, "^g__")
    missing_label <- is.na(genus) | genus == "" | genus == "NA"
    genus[missing_label] <- tax_ids[missing_label]
    tax_labels <- make.unique(genus)
  }
  
  tibble(taxon_id = tax_ids, taxon = tax_labels)
}

extract_taxon_contrasts <- function(comp_results, taxon_labels) {
  purrr::map_dfr(
    seq_len(nrow(comp_results)),
    function(i) {
      comp <- comp_results$components[[i]]
      
      tibble(
        variable = as.character(comp_results$variable[[i]]),
        contrast = as.character(comp_results$contrast[[i]]),
        group1 = comp_results$group1[[i]],
        group2 = comp_results$group2[[i]],
        taxon_id = names(comp$mean_diff),
        mean_diff_clr = as.numeric(comp$mean_diff),
        standardized_diff = as.numeric(comp$standardized_diff),
        coordinate_stat = as.numeric(comp$coordinate_stat),
        valid_taxon = as.logical(comp$valid_taxa)
      )
    }
  ) %>%
    left_join(taxon_labels, by = "taxon_id")
}

reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  stats::reorder(paste(x, within, sep = sep), by, FUN = fun, ...)
}

scale_y_reordered <- function(..., sep = "___") {
  scale_y_discrete(labels = function(x) gsub(paste0(sep, ".*$"), "", x), ...)
}

sig_stars <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ ""
  )
}

fmt_p_latex <- function(p_num) {
  p_str <- formatC(p_num, format = "e", digits = 2)
  stars <- sig_stars(p_num)
  
  ifelse(stars != "", paste0(p_str, "$^{", stars, "}$"), p_str)
}

fmt_p_table <- function(p_num) {
  formatC(p_num, format = "e", digits = 2)
}

center_max_statistic <- function(statistic, n_taxa) {
  statistic - 2 * log(n_taxa) + log(log(n_taxa))
}

gumbel_cdf_cao <- function(x) {
  exp(-exp(-x / 2) / sqrt(pi))
}

gumbel_density_cao <- function(x) {
  exp(-x / 2) * gumbel_cdf_cao(x) / (2 * sqrt(pi))
}

gumbel_upper_tail_cao <- function(x) {
  1 - gumbel_cdf_cao(x)
}
```

## Prepare CLR matrix

```{r prepare_clr_matrix}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_comp <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_comp_metadata()

counts_raw <- counts_raw[sampdata_comp$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_comp$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

comp_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

comp_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("comp_equivalence_multrepl_clr_object.rds")
)

write_csv(
  comp_preprocessing_summary,
  file = path_target("comp_equivalence_preprocessing_summary.csv")
)
```

## Contrast setup

Compositional equivalence is a two-group procedure. Country is omitted, and the
number of siblings is collapsed to a two-level variable contrasting children
without siblings against children with at least one sibling. The sparse
`EBF duration == "1 month"` category is set to missing only for the EBF-specific
contrast setup; those samples remain in the CLR matrix and are available for all
other contrasts.

```{r contrast_setup}
sampdata_test <- sampdata_comp %>%
  mutate(
    `EBF duration` = replace(
      `EBF duration`,
      `EBF duration` == "1 month",
      NA
    )
  )

contrast_spec <- tibble::tribble(
  ~variable,       ~contrast,       ~group1,      ~group2,
  "C-section",    "C-section",     "Cesarean",  "Vaginal",
  "BF duration",  "BF duration",   "0 months",  ">=2 months",
  "EBF duration", "EBF duration",  "0 months",  ">=2 months",
  "Sex",          "Sex",           "Female",    "Male",
  "Smoking",      "Smoking",       "No",        "Yes",
  "Siblings",     "Siblings",      "0",         ">=1"
) %>%
  mutate(
    variable = factor(variable, levels = contrast_order),
    contrast = factor(contrast, levels = contrast_order)
  )

contrast_level_summary <- purrr::map_dfr(
  seq_len(nrow(contrast_spec)),
  function(i) {
    spec <- contrast_spec[i, ]
    vals <- sampdata_test[[as.character(spec$variable)]]
    
    tibble(
      variable = as.character(spec$variable),
      group1 = spec$group1,
      group2 = spec$group2,
      n_group1 = sum(vals == spec$group1, na.rm = TRUE),
      n_group2 = sum(vals == spec$group2, na.rm = TRUE),
      n_missing_or_other = sum(is.na(vals) | !vals %in% c(spec$group1, spec$group2))
    )
  }
)

contrast_level_summary
```

```{r save_contrast_setup}
write_csv(
  contrast_level_summary,
  file = path_target("comp_equivalence_contrast_level_summary.csv")
)
```

## Compositional equivalence tests

For each contrast, the statistic is the largest standardized squared difference
between group CLR means. Group labels are permuted while the CLR-transformed
microbial profiles are kept fixed.

```{r parallel_setup}
future::plan(
  future::multisession,
  workers = cores
)

progressr::handlers("txtprogressbar")
options(progressr.enable = TRUE)
```

```{r comp_equivalence_tests}
set.seed(permutation_seed)
path_comp_results <- path_target("comp_equivalence_results_multrepl_clr.rds")

if (!file.exists(path_comp_results)) {
  comp_results <- progressr::with_progress({
    p <- progressr::progressor(steps = nrow(contrast_spec))
    
    furrr::future_map_dfr(
      seq_len(nrow(contrast_spec)),
      function(i) {
        spec <- contrast_spec[i, ]
        
        out <- run_comp_equiv_permutation_test(
          y = counts_clr,
          metadata = sampdata_test,
          variable = as.character(spec$variable),
          group_levels = c(spec$group1, spec$group2),
          contrast_label = as.character(spec$contrast),
          n_perm = n_perm
        )
        
        p(message = sprintf("Compositional equivalence: %s", spec$contrast))
        
        out
      },
      .options = furrr::furrr_options(seed = permutation_seed)
    )
  })
  
  saveRDS(comp_results, file = path_comp_results)
} else {
  comp_results <- readRDS(path_comp_results)
}

comp_results <- comp_results %>%
  mutate(
    variable = factor(variable, levels = contrast_order),
    contrast = factor(contrast, levels = contrast_order)
  )

comp_results %>%
  select(
    contrast, group1, group2, n1, n2, n_taxa_tested,
    statistic_obs, p_empirical, n_exceed, n_perm
  ) %>%
  mutate(p_empirical = fmt_p_table(p_empirical))
```

## p-value refinement with permApprox

```{r comp_equivalence_permapprox}
path_comp_permapprox <- path_target("comp_equivalence_permapprox_results_multrepl_clr.rds")

comp_obs_stats <- comp_results$statistic_obs
comp_perm_stats <- do.call(cbind, comp_results$perm_stats)

gpd_ctrl <- permApprox::make_gpd_ctrl(
  constraint = "support_at_obs",
  sample_size = min(comp_results$n_samples)
)

if (!file.exists(path_comp_permapprox)) {
  comp_permapprox <- permApprox::perm_approx(
    obs_stats = comp_obs_stats,
    perm_stats = comp_perm_stats,
    method = "gpd",
    alternative = "greater",
    null_center = "median",
    approx_thresh = 0.1,
    adjust_method = "none",
    gpd_ctrl = gpd_ctrl,
    cores = cores,
    verbose = TRUE
  )
  
  saveRDS(comp_permapprox, file = path_comp_permapprox)
} else {
  comp_permapprox <- readRDS(path_comp_permapprox)
}

comp_permapprox
```

```{r comp_equivalence_results_table}
comp_results_all <- comp_results %>%
  select(
    contrast, group1, group2, n1, n2, n_samples, n_taxa, n_taxa_tested,
    n_taxa_zero_variance, statistic_obs, n_exceed, n_perm, p_empirical
  ) %>%
  mutate(
    statistic_centered = center_max_statistic(statistic_obs, n_taxa_tested),
    p_asymptotic_gumbel = gumbel_upper_tail_cao(statistic_centered),
    p_permapprox = comp_permapprox$p_unadjusted,
    method_used = comp_permapprox$method_used
  )

comp_results_all %>%
  arrange(contrast) %>%
  mutate(
    p_empirical = fmt_p_table(p_empirical),
    p_asymptotic_gumbel = fmt_p_table(p_asymptotic_gumbel),
    p_permapprox = fmt_p_table(p_permapprox)
  )
```

```{r comp_equivalence_results_csv}
comp_results_all %>%
  mutate(
    contrast = as.character(contrast)
  ) %>%
  write_csv(path_target("comp_equivalence_results_multrepl_clr.csv"))
```

```{r comp_equivalence_latex}
comp_tex_df <- comp_results_all %>%
  arrange(contrast) %>%
  transmute(
    Contrast = as.character(contrast),
    M = round(statistic_obs, 2),
    Mc = round(statistic_centered, 2),
    n_exceed = n_exceed,
    p_emp = fmt_p_latex(p_empirical),
    p_papp = fmt_p_latex(p_permapprox),
    p_asy = fmt_p_latex(p_asymptotic_gumbel)
  )

tex_comp_equivalence <- kbl(
  comp_tex_df,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Contrast", "$M_n$", "$M_n^c$", 
                "\\# exceed", "$p$ (emp.)",
                "$p$ (GPD-refined)", "$p$ (asym.)"),
  align = "lrrrlll",
  caption = paste(
    "Compositional equivalence results for selected two-group contrasts.",
    "$M_n$ is the maximum standardized squared CLR mean difference.",
    "$M_n^c = M_n - 2\\log(p) + \\log\\log(p)$ is the centered statistic",
    "used for the asymptotic Gumbel approximation.",
    "\\# exceed: number of permuted statistics $\\geq$ observed.",
    "$p$ (emp.): empirical permutation $p$-value ($999$ permutations);",
    "$p$ (GPD-refined): $p$-value refined with \\texttt{permApprox};",
    "$p$ (asym.): asymptotic Gumbel $p$-value."
  ),
  label = "application:comp-equivalence"
) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(
    general = "$^{***}\\:p<0.001$; $^{**}\\:p<0.01$; $^{*}\\:p<0.05$; $^{.}\\:p<0.1$.",
    general_title = "",
    escape = FALSE,
    threeparttable = TRUE
  )

tex_comp_equivalence
save_kable(tex_comp_equivalence, file = path_target("comp_equivalence_table.tex"))
```

## Asymptotic Gumbel comparison

The maximum statistic can also be centered as
$M_n^c = M_n - 2\log(p) + \log\log(p)$ and compared with the limiting Gumbel
distribution used by Cao et al. The permutation statistics are
centered in the same way, using the number of CLR coordinates tested in each
contrast. This comparison is diagnostic: the permutation-based p-values remain
the primary inferential results in this analysis.

```{r comp_equivalence_gumbel_comparison_data}
perm_gumbel_df <- comp_results %>%
  select(contrast, n_taxa_tested, perm_stats) %>%
  mutate(perm_id = purrr::map(perm_stats, seq_along)) %>%
  tidyr::unnest(c(perm_stats, perm_id)) %>%
  rename(statistic_perm = perm_stats) %>%
  mutate(
    contrast = factor(contrast, levels = contrast_order),
    statistic_centered_perm = center_max_statistic(
      statistic_perm,
      n_taxa_tested
    )
  )

obs_gumbel_df <- comp_results_all %>%
  mutate(
    contrast = factor(contrast, levels = contrast_order),
    label = paste0(
      "p-val. (perm.) = ", formatC(p_empirical, format = "e", digits = 2),
      "\n",
      "p-val. (asy.) = ",
      formatC(p_asymptotic_gumbel, format = "e", digits = 2),
      "\n",
      "p-val. (GPD) = ", formatC(p_permapprox, format = "e", digits = 2)
    )
  )

gumbel_curve_df <- purrr::map_dfr(
  contrast_order,
  function(current_contrast) {
    perm_vals <- perm_gumbel_df %>%
      filter(contrast == current_contrast) %>%
      pull(statistic_centered_perm)
    
    obs_val <- obs_gumbel_df %>%
      filter(contrast == current_contrast) %>%
      pull(statistic_centered)
    
    x_grid <- seq(
      min(c(perm_vals, obs_val), na.rm = TRUE) - 1,
      max(c(perm_vals, obs_val), na.rm = TRUE) + 1,
      length.out = 400
    )
    
    tibble(
      contrast = current_contrast,
      statistic_centered = x_grid,
      density = gumbel_density_cao(x_grid)
    )
  }
) %>%
  mutate(contrast = factor(contrast, levels = contrast_order))
```

```{r comp_equivalence_gumbel_comparison_plot, fig.width=9, fig.height=6, dpi = 200}
perm_gumbel_df %>%
  filter(contrast %in% contrast_order) %>%
  ggplot(aes(x = statistic_centered_perm)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 45,
    fill = "grey70",
    color = "grey50",
    alpha = 0.85
  ) +
  geom_line(
    data = gumbel_curve_df,
    aes(x = statistic_centered, y = density),
    color = "gray20",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_vline(
    data = obs_gumbel_df %>% filter(contrast %in% contrast_order),
    aes(xintercept = statistic_centered),
    color = "tomato2",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_text(
    data = obs_gumbel_df %>% filter(contrast %in% contrast_order),
    aes(x = Inf, y = Inf, label = label),
    hjust = 1.05,
    vjust = 1.2,
    size = 3.5,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ contrast, scales = "free", ncol = 3) +
  labs(
    x = expression(M[n] - 2 * log(p) + log(log(p))),
    y = "Density"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 14)
  )
```

## Permutation distributions

The following diagnostic plot shows selected empirical null distributions. These
plots are used as a visual check of the permutation tail before relying on the
GPD-based p-value refinement.

```{r comp_equivalence_perm_dist_selected, fig.width=8, fig.height=4.5}
perm_dist_df <- comp_results %>%
  select(contrast, perm_stats) %>%
  mutate(perm_id = purrr::map(perm_stats, seq_along)) %>%
  tidyr::unnest(c(perm_stats, perm_id)) %>%
  rename(statistic_perm = perm_stats) %>%
  mutate(contrast = factor(contrast, levels = contrast_order))

obs_df <- comp_results_all %>%
  mutate(
    contrast = factor(contrast, levels = contrast_order),
    label = paste0(
      "p_emp = ", formatC(p_empirical, format = "e", digits = 2),
      ifelse(p_empirical < alpha_perm_approx, "*", ""), "\n",
      "p_pApp = ", formatC(p_permapprox, format = "e", digits = 2),
      ifelse(p_permapprox < alpha_perm_approx, "*", "")
    )
  )

perm_dist_df %>%
  filter(contrast %in% contrast_order) %>%
  ggplot(aes(x = statistic_perm)) +
  geom_histogram(bins = 40, fill = "grey75", color = "white") +
  geom_vline(
    data = obs_df %>% filter(contrast %in% contrast_order),
    aes(xintercept = statistic_obs),
    color = "#B2182B",
    linewidth = 0.8
  ) +
  geom_text(
    data = obs_df %>% filter(contrast %in% contrast_order),
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05,
    vjust = 1.3,
    size = 3.5,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ contrast, scales = "free", ncol = 3) +
  labs(
    x = expression(M[n]^"*"),
    y = "Number of permutations"
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Taxon-wise CLR contrasts

The test is global and does not provide taxon-level inference. The following
plots are descriptive: they show the taxa with the largest coordinate-wise
contributions to the maximum statistic for each contrast.

```{r taxon_contrast_data}
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)

taxon_contrasts <- extract_taxon_contrasts(
  comp_results = comp_results,
  taxon_labels = taxon_labels
)

taxon_contrasts %>%
  write_csv(path_target("comp_equivalence_taxon_contrasts.csv"))

top10_taxon_contrasts <- taxon_contrasts %>%
  filter(valid_taxon) %>%
  group_by(contrast) %>%
  slice_max(order_by = coordinate_stat, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    contrast = factor(contrast, levels = contrast_order),
    taxon_plot = reorder_within(taxon, coordinate_stat, contrast)
  )

top10_taxon_contrasts %>%
  select(-taxon_plot) %>%
  write_csv(path_target("comp_equivalence_top10_taxon_contrasts.csv"))
```

```{r taxon_contrast_top10_all, fig.width=12, fig.height=14}
ggplot(
  top10_taxon_contrasts,
  aes(x = coordinate_stat, y = taxon_plot)
) +
  geom_segment(
    aes(x = 0, xend = coordinate_stat, yend = taxon_plot),
    color = "grey70"
  ) +
  geom_point(aes(color = standardized_diff), size = 2) +
  scale_color_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Std. CLR\ndifference"
  ) +
  scale_y_reordered() +
  facet_wrap(~ contrast, scales = "free", ncol = 2) +
  labs(
    x = "Coordinate-wise contribution to maximum statistic",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

As we will use `EBF duration` as grouping variable in later analyses, we generate an isolated plot for this variable here, which will be shown in the thesis chapter.

```{r taxon_contrast_top10_selected, fig.width=7, fig.height=3}
top10_taxon_contrasts %>%
  #filter(contrast %in% diagnostic_contrasts) %>%
  filter(contrast == "EBF duration") %>%
  ggplot(aes(
    x = coordinate_stat,
    y = taxon_plot
  )) +
  geom_segment(
    aes(x = 0, xend = coordinate_stat, yend = taxon_plot),
    color = "grey70"
  ) +
  geom_point(aes(color = standardized_diff), size = 2) +
  scale_color_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Std. CLR\ndifference"
  ) +
  scale_y_reordered() +
  #facet_wrap(~ contrast, scales = "free", ncol = 1) +
  labs(
    x = "Coordinate-wise contribution to maximum statistic",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 11),
    legend.text = element_text(size = 11)
  )
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
