---
title: "Exploratory taxon-level analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
params:
  name: "07_exploration" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "0fb365aa-5986-41d1-9910-df766aa9b502")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8,
  fig.height = 5,
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/expl_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(pheatmap)
library(ggrepel)
library(kableExtra)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data:
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
n_top_taxa <- 20
n_selected_taxa <- 6
n_prevalence_taxa <- 10
n_clr_variance_taxa <- 10
n_rank_label_taxa <- 5

ebf_palette <- c("non-EBF" = "#1F78B4", "EBF" = "#E66101")
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

The exploratory analyses use the filtered genus-level object that is also used
for differential abundance and differential distribution analyses. Relative
abundances are used for compositional summaries, while multiplicative zero
replacement followed by CLR transformation is used for transformed-abundance
visualizations.

```{r helper_functions}
make_exploration_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      EBF_group = case_when(
        `EBF duration` == "0 months" ~ "non-EBF",
        `EBF duration` == ">=2 months" ~ "EBF",
        TRUE ~ NA_character_
      ),
      EBF_group = factor(EBF_group, levels = c("non-EBF", "EBF")),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

taxon_labels_from_phyloseq <- function(phylo) {
  tax_tab <- tax_table(phylo) %>%
    as("matrix") %>%
    as_tibble(rownames = "taxon_id")

  label_col <- dplyr::case_when(
    "Genus" %in% colnames(tax_tab) ~ "Genus",
    "genus" %in% colnames(tax_tab) ~ "genus",
    TRUE ~ NA_character_
  )

  if (is.na(label_col)) {
    labels <- tax_tab$taxon_id
  } else {
    labels <- tax_tab[[label_col]]
    labels <- ifelse(is.na(labels) | labels == "", tax_tab$taxon_id, labels)
  }

  setNames(make.unique(labels), tax_tab$taxon_id)
}

tax_table_tibble <- function(phylo) {
  tax_table(phylo) %>%
    as("matrix") %>%
    as_tibble(rownames = "taxon_id")
}

mean_or_na <- function(x) {
  if (all(is.na(x))) {
    NA_real_
  } else {
    mean(x, na.rm = TRUE)
  }
}
```

## Prepare matrices

```{r prepare_matrices}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_expl <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_exploration_metadata()

counts_raw <- counts_raw[sampdata_expl$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_expl$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)
taxon_table <- tax_table_tibble(phylo_genus)

exploration_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac,
  n_non_ebf = sum(sampdata_expl$EBF_group == "non-EBF", na.rm = TRUE),
  n_ebf = sum(sampdata_expl$EBF_group == "EBF", na.rm = TRUE),
  n_excluded_from_ebf_comparison = sum(is.na(sampdata_expl$EBF_group))
)

exploration_summary
```

```{r save_preprocessed_objects}
saveRDS(
  list(
    sample_data = sampdata_expl,
    counts_raw = counts_raw,
    counts_rel = counts_rel,
    counts_clr = counts_clr,
    clr_object = clr_object
  ),
  file = path_target("exploration_preprocessed_objects.rds")
)

write_csv(
  exploration_summary,
  file = path_target("exploration_summary.csv")
)
```

## Taxon summaries

```{r taxon_summary}
taxon_summary <- tibble(
  taxon_id = colnames(counts_raw),
  taxon = base::unname(taxon_labels[colnames(counts_raw)]),
  total_count = colSums(counts_raw),
  prevalence = colMeans(counts_raw > 0),
  mean_relative_abundance = colMeans(counts_rel),
  median_relative_abundance = apply(counts_rel, 2, median),
  max_relative_abundance = apply(counts_rel, 2, max),
  mean_clr = colMeans(counts_clr),
  median_clr = apply(counts_clr, 2, median),
  clr_variance = apply(counts_clr, 2, var)
) %>%
  left_join(taxon_table, by = "taxon_id") %>%
  arrange(desc(mean_relative_abundance), desc(prevalence))

top_taxa_ids <- taxon_summary %>%
  slice_head(n = min(n_top_taxa, nrow(taxon_summary))) %>%
  pull(taxon_id)

selected_taxa_ids <- taxon_summary %>%
  slice_head(n = min(n_selected_taxa, nrow(taxon_summary))) %>%
  pull(taxon_id)

prevalence_taxa_ids <- taxon_summary %>%
  slice_head(n = min(n_prevalence_taxa, nrow(taxon_summary))) %>%
  pull(taxon_id)

taxon_summary
```

```{r save_taxon_summary}
write_csv(
  taxon_summary,
  file = path_target("taxon_level_summary.csv")
)

top_taxa_table <- taxon_summary %>%
  slice_head(n = n_top_taxa) %>%
  transmute(
    Taxon = taxon,
    Prevalence = scales::percent(prevalence, accuracy = 0.1),
    `Mean relative abundance` = scales::percent(mean_relative_abundance, 
                                                accuracy = 0.01),
    `Median relative abundance` = scales::percent(median_relative_abundance, 
                                                  accuracy = 0.01),
    `Maximum relative abundance` = scales::percent(max_relative_abundance, 
                                                   accuracy = 0.01)
  )

top_taxa_kable <- kbl(
  top_taxa_table,
  format = "latex",
  booktabs = TRUE,
  caption = "Most abundant genera in the filtered genus-level PASTURE data.",
  label = "application:exploration-top-taxa"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 8)

top_taxa_kable

save_kable(
  top_taxa_kable,
  path_target("exploration_top_taxa_table.tex")
)
```

## Prevalence and abundance


```{r prevalence_histogram, fig.width=7, fig.height=4.2, dpi=250}
taxon_summary %>%
  ggplot(aes(x = prevalence)) +
  geom_histogram(binwidth = 0.05, fill = "#1F78B4", color = "white", alpha = 0.85) +
  scale_x_continuous(labels = scales::percent_format(), breaks = seq(0, 1, by = 0.25)) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(
    title = "Taxon prevalence",
    x = "Proportion of samples with positive count",
    y = "Number of genera"
  ) +
  theme_minimal()
```

```{r abundance_histogram, fig.width=7, fig.height=4.2, dpi=250}
taxon_summary %>%
  ggplot(aes(x = mean_relative_abundance)) +
  geom_histogram(bins = 40, fill = "#1F78B4", color = "white", alpha = 0.85) +
  scale_x_log10(labels = scales::percent_format()) +
  labs(
    title = "Mean relative abundance",
    x = "Mean relative abundance (log scale)",
    y = "Number of genera"
  ) +
  theme_minimal()
```

```{r prevalence_abundance, fig.width=8, fig.height=4.1, dpi=200}
set.seed(42)
taxa_to_label <- taxon_summary %>%
  slice_head(n = 10)

taxon_summary %>%
  ggplot(aes(x = prevalence, y = mean_relative_abundance)) +
  geom_point(alpha = 0.65, color = "grey35") +
  geom_point(
    data = taxa_to_label,
    color = "#E66101",
    size = 2
  ) +
  ggrepel::geom_text_repel(
    data = taxa_to_label,
    aes(label = stringr::str_trunc(taxon, width = 30)),
    size = 4,
    nudge_x = 0.04,
    min.segment.length = 0,
    segment.color = "grey45",
    segment.size = 0.3,
    segment.alpha = 0.9,
    box.padding = 0.8,
    point.padding = 0.2,
    max.overlaps = Inf,
    seed = 42
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_y_log10(
    labels = function(x) {
      paste0(format(signif(x * 100, 3), trim = TRUE, scientific = FALSE, 
                    drop0trailing = TRUE), "%")
    }
  ) +
  coord_cartesian(clip = "off", xlim = c(0, 1.35)) +
  labs(
    #title = "Prevalence-abundance relationship",
    x = "Prevalence",
    y = "Mean relative abundance"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5),
    panel.grid.minor = element_blank()
  )
```

```{r rank_abundance_curve, fig.width=7, fig.height=4.2, dpi=250}
rank_abundance_df <- taxon_summary %>%
  arrange(desc(mean_relative_abundance)) %>%
  mutate(
    abundance_rank = row_number(),
    highlight_taxon = abundance_rank <= 10,
    label_taxon = abundance_rank <= n_rank_label_taxa
  )

ggplot(
  rank_abundance_df,
  aes(x = abundance_rank, y = mean_relative_abundance)
) +
  geom_line(color = "grey55", linewidth = 0.5) +
  geom_point(color = "grey35", alpha = 0.75, size = 1.8) +
  geom_point(
    data = filter(rank_abundance_df, highlight_taxon),
    color = "#E66101",
    size = 2.2
  ) +
  ggrepel::geom_text_repel(
    data = filter(rank_abundance_df, label_taxon),
    aes(label = stringr::str_trunc(taxon, width = 28)),
    size = 3.2,
    min.segment.length = 0,
    segment.color = "grey45",
    segment.size = 0.3,
    box.padding = 0.5,
    point.padding = 0.2,
    max.overlaps = Inf,
    seed = 42
  ) +
  scale_y_log10(
    labels = function(x) {
      paste0(format(signif(x * 100, 3), trim = TRUE, scientific = FALSE,
                    drop0trailing = TRUE), "%")
    }
  ) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 8)) +
  labs(
    title = "Rank-abundance curve",
    x = "Genus rank by mean relative abundance",
    y = "Mean relative abundance"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```

**Most variable genera after CLR transformation**

```{r clr_variance_by_taxon, fig.width=7, fig.height=3.5, dpi=200}
taxon_summary %>%
  arrange(desc(clr_variance)) %>%
  slice_head(n = min(n_clr_variance_taxa, nrow(taxon_summary))) %>%
  mutate(
    taxon = factor(taxon, levels = rev(taxon))
  ) %>%
  ggplot(aes(x = clr_variance, y = taxon)) +
  geom_segment(aes(x = 0, xend = clr_variance, yend = taxon),
               color = "grey70", linewidth = 0.5) +
  geom_point(color = "#1F78B4", size = 2.2) +
  labs(
    #title = "Most variable genera after CLR transformation",
    x = "Variance of CLR coordinates",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Dominant taxonomic composition

```{r stacked_barplot_data}
top_bar_taxa <- taxon_summary %>%
  slice_head(n = 8) %>%
  pull(taxon_id)

composition_taxon_levels <- c(base::unname(taxon_labels[top_bar_taxa]), "Other")

composition_colors <- c("#009E73", "#56B4E9", "#E69F00", "#D55E00", 
                        "#F0E442", "#CC79A7", "#91D1C2", "#0072B2", 
                        "Other" = "gray")

names(composition_colors) <- composition_taxon_levels

composition_df <- as_tibble(counts_rel[, top_bar_taxa, drop = FALSE], rownames = "SampleID") %>%
  mutate(Other = 1 - rowSums(across(all_of(top_bar_taxa)))) %>%
  pivot_longer(
    cols = -SampleID,
    names_to = "taxon_id",
    values_to = "relative_abundance"
  ) %>%
  mutate(
    taxon = case_when(
      taxon_id == "Other" ~ "Other",
      TRUE ~ base::unname(taxon_labels[taxon_id])
    ),
    taxon = factor(taxon, levels = composition_taxon_levels)
  ) %>%
  left_join(
    sampdata_expl %>%
      select(SampleID, EBF_group) %>%
      mutate(EBF_group = forcats::fct_na_value_to_level(EBF_group, level = "not compared")),
    by = "SampleID"
  )

composition_group_df <- composition_df %>%
  group_by(EBF_group, taxon) %>%
  summarise(
    mean_relative_abundance = mean(relative_abundance),
    .groups = "drop"
  )

composition_group_df
```

```{r relative_abundance_barplot, fig.width=9, fig.height=4, dpi=200}
ggplot(
  composition_group_df,
  aes(x = EBF_group, y = mean_relative_abundance, fill = taxon)
) +
  geom_col(width = 0.68, color = "white", linewidth = 0.25) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = composition_colors) +
  labs(
    #title = "Mean genus-level composition by EBF group",
    x = NULL,
    y = "Mean relative abundance",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "right"
  )
```

## CLR coordinatess

```{r clr_heatmap_data}
heatmap_taxa <- taxon_summary %>%
  slice_head(n = min(n_top_taxa, nrow(taxon_summary))) %>%
  pull(taxon_id)

heatmap_taxa_labels <- taxon_labels[heatmap_taxa]
heatmap_taxa_labels[is.na(heatmap_taxa_labels) | heatmap_taxa_labels == ""] <-
  heatmap_taxa[is.na(heatmap_taxa_labels) | heatmap_taxa_labels == ""]
heatmap_taxa_labels <- make.unique(base::unname(heatmap_taxa_labels))
names(heatmap_taxa_labels) <- heatmap_taxa

heatmap_mat <- counts_clr[, heatmap_taxa, drop = FALSE]
colnames(heatmap_mat) <- heatmap_taxa_labels

heatmap_sample_data <- sampdata_expl %>%
  filter(!is.na(EBF_group))

heatmap_mat_ebf <- heatmap_mat[heatmap_sample_data$SampleID, , drop = FALSE]

annotation_col <- data.frame(EBF_group = heatmap_sample_data$EBF_group)
rownames(annotation_col) <- rownames(heatmap_mat_ebf)
annotation_colors <- list(EBF_group = ebf_palette)
```

```{r clr_heatmap_top20_unclustered, fig.width=10, fig.height=5, dpi=250}
pheatmap(
  t(heatmap_mat_ebf),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "CLR coordinates (top 20 genera)"
)
```

```{r clr_heatmap_top20_clustered, fig.width=10, fig.height=5, dpi=250}
pheatmap(
  t(heatmap_mat_ebf),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  main = "CLR coordinates (top 20 genera)"
)
```

## Selected taxon distributions

```{r selected_taxon_data}
selected_taxon_labels <- setNames(
  stringr::str_trunc(taxon_labels[selected_taxa_ids], width = 30),
  selected_taxa_ids
)

selected_clr_df <- as_tibble(
  counts_clr[heatmap_sample_data$SampleID, selected_taxa_ids, drop = FALSE],
  rownames = "SampleID"
) %>%
  left_join(
    heatmap_sample_data %>% select(SampleID, EBF_group),
    by = "SampleID"
  ) %>%
  pivot_longer(
    cols = all_of(selected_taxa_ids),
    names_to = "taxon_id",
    values_to = "clr_abundance"
  ) %>%
  mutate(
    taxon = factor(
      selected_taxon_labels[taxon_id],
      levels = selected_taxon_labels[selected_taxa_ids]
    )
  )

selected_transform_df <- as_tibble(
  counts_rel[heatmap_sample_data$SampleID, selected_taxa_ids, drop = FALSE],
  rownames = "SampleID"
) %>%
  pivot_longer(
    cols = all_of(selected_taxa_ids),
    names_to = "taxon_id",
    values_to = "relative_abundance"
  ) %>%
  left_join(
    selected_clr_df %>% select(SampleID, taxon_id, clr_abundance, taxon),
    by = c("SampleID", "taxon_id")
  ) %>%
  mutate(
    originally_zero = relative_abundance == 0,
    zero_status = factor(
      if_else(originally_zero, "Originally zero", "Observed"),
      levels = c("Observed", "Originally zero")
    )
  )

selected_transform_density_df <- selected_transform_df %>%
  transmute(
    SampleID,
    taxon,
    zero_status,
    scale = "Relative abundance",
    value = relative_abundance
  ) %>%
  bind_rows(
    selected_transform_df %>%
      transmute(
        SampleID,
        taxon,
        zero_status,
        scale = "CLR coordinates",
        value = clr_abundance
      )
  ) %>%
  mutate(
    scale = factor(
      scale,
      levels = c("Relative abundance", "CLR coordinates")
    )
  )

selected_zero_df <- as_tibble(
  counts_raw[heatmap_sample_data$SampleID, prevalence_taxa_ids, 
             drop = FALSE] > 0,
  rownames = "SampleID"
) %>%
  left_join(
    heatmap_sample_data %>% select(SampleID, EBF_group),
    by = "SampleID"
  ) %>%
  pivot_longer(
    cols = all_of(prevalence_taxa_ids),
    names_to = "taxon_id",
    values_to = "observed"
  ) %>%
  group_by(EBF_group, taxon_id) %>%
  summarise(prevalence = mean(observed), .groups = "drop") %>%
  mutate(
    taxon = factor(
      stringr::str_trunc(taxon_labels[taxon_id], width = 30),
      levels = rev(stringr::str_trunc(taxon_labels[prevalence_taxa_ids], 
                                      width = 30))
    )
  )

selected_clr_df
```

**Distributional change from relative to CLR coordinates**

```{r rel_to_clr_histogram_comparison, fig.width=9, fig.height=6, dpi=250}
ggplot(selected_transform_density_df, aes(x = value)) +
  geom_histogram(
    data = filter(selected_transform_density_df, zero_status == "Observed"),
    bins = 40,
    fill = "grey75",
    color = "grey35",
    alpha = 0.8,
    linewidth = 0.2
  ) +
  geom_histogram(
    data = filter(selected_transform_density_df, 
                  zero_status == "Originally zero"),
    bins = 40,
    fill = "#E66101",
    color = NA,
    alpha = 0.85,
    linewidth = 0
  ) +
  facet_grid(taxon ~ scale, scales = "free", switch = "y") +
  labs(
    #title = "Distributional change from relative to CLR coordinates",
    x = "Abundance value",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 13),
    strip.text.y = element_text(size = 14),
    strip.text.x = element_text(size = 16, face = "bold"),
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

```{r rel_to_clr_scatter, fig.width=10, fig.height=6, dpi=250}
ggplot(
  selected_transform_df,
  aes(x = relative_abundance, y = clr_abundance, color = zero_status)
) +
  geom_point(alpha = 0.65, size = 0.9) +
  scale_x_continuous(
    labels = function(x) {
      paste0(format(signif(x * 100, 3), trim = TRUE, scientific = FALSE,
                    drop0trailing = TRUE), "%")
    }
  ) +
  scale_color_manual(
    values = c("Observed" = "grey35", "Originally zero" = "#E66101")
  ) +
  scale_x_log10() +
  facet_wrap(~ taxon, scales = "free_y") +
  labs(
    title = "Relative abundance and CLR coordinates",
    x = "Relative abundance",
    y = "CLR coordinates",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

```{r selected_taxon_violin, fig.width=10, fig.height=5, dpi=250}
ggplot(selected_clr_df, aes(x = EBF_group, y = clr_abundance, fill = EBF_group)) +
  geom_violin(alpha = 0.55, trim = FALSE, color = NA) +
  geom_boxplot(width = 0.16, outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = ebf_palette, guide = "none") +
  facet_wrap(~ taxon, scales = "free_y") +
  labs(
    #title = "CLR coordinatess of selected dominant genera",
    x = NULL,
    y = "CLR coordinates"
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 16),
    panel.grid.minor = element_blank()
  )
```

```{r selected_taxon_violin_jitter, fig.width=10, fig.height=5, dpi=250}
ggplot(selected_clr_df, aes(x = EBF_group, y = clr_abundance, fill = EBF_group)) +
  geom_violin(alpha = 0.55, trim = FALSE, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.05, alpha = 0.3, size = 0.4) +
  scale_fill_manual(values = ebf_palette, guide = "none") +
  facet_wrap(~ taxon, scales = "free_y") +
  labs(
    #title = "CLR coordinatess of selected dominant genera",
    x = NULL,
    y = "CLR coordinates"
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 16),
    panel.grid.minor = element_blank()
  )
```

```{r selected_taxon_density, fig.width=12, fig.height=4, dpi=200}
ggplot(selected_clr_df, aes(x = clr_abundance, fill = EBF_group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = ebf_palette) +
  facet_wrap(~ taxon, scales = "free") +
  labs(
    #title = "CLR coordinates densities",
    x = "CLR coordinates",
    y = "Density",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 16),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r selected_taxon_prevalence_by_group, fig.width=8, fig.height=4, dpi=250}
ggplot(selected_zero_df, aes(x = prevalence, y = taxon, color = EBF_group)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.45)) +
  scale_color_manual(values = ebf_palette) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    #title = "Prevalence of selected dominant genera by EBF group",
    x = "Prevalence",
    y = NULL,
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 13),
    legend.text = element_text(size = 13),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r save_selected_taxon_data}
write_csv(
  selected_zero_df,
  file = path_target("selected_taxa_prevalence_by_group.csv")
)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
