---
title: "Differential abundance analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "08_diff_abundance" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "8400abc1-adab-467f-99ae-0f1d4d512bda")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/DA_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(dacomp)
library(permApprox)
library(kableExtra)
library(ggrepel)
library(furrr)
library(future)
library(progressr)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
permutation_seed <- 42
cores <- 10

q_DSFDR <- 0.1
dacomp_perm_grid <- c(1e3, 1e4, 1e5, 1e6, 1e7)
refit_dacomp_refs <- FALSE
refit_dacomp <- FALSE
write_dacomp_stats_files <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

Counts are transformed to relative abundances, zeros are replaced by
multiplicative replacement, and the CLR transformation is applied after
replacement. The dacomp test itself uses raw counts, as recommended by dacomp,
but the relative-abundance and CLR objects are stored to keep the preprocessing
consistent with the other analyses in this chapter. The shared preprocessing
helpers are defined in `functions.R`.

```{r helper_functions}
make_da_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

fmt_nperm <- function(n) {
  format(n, scientific = TRUE)
}

make_dacomp_result_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_res_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

make_dacomp_runtime_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_runtime_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

make_dacomp_stats_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_stats_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

run_dacomp <- function(X_counts, y_groups, res_refs, n_perm, q = 0.1) {
  dacomp.test(
    X = X_counts,
    y = y_groups,
    ind_reference_taxa = res_refs,
    test = DACOMP.TEST.NAME.WILCOXON,
    nr_perm = n_perm,
    q = q,
    verbose = FALSE,
    return_rarefied_values = TRUE,
    Test_All = TRUE
  )
}

extract_dacomp_stats <- function(res_dacomp) {
  stats_mat <- res_dacomp$stats_matrix
  
  if (is.null(stats_mat)) {
    stop("The dacomp result does not contain `stats_matrix`.")
  }
  
  obs_stats <- stats_mat[1, ]
  perm_stats <- stats_mat[-1, , drop = FALSE]
  colnames(perm_stats) <- names(obs_stats)
  
  list(
    obs_stats = obs_stats,
    perm_stats = perm_stats
  )
}
```

## Prepare relative abundance and CLR matrices

```{r prepare_clr_matrix}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_da <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_da_metadata()

counts_raw <- counts_raw[sampdata_da$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_da$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

da_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

da_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("diff_abundance_multrepl_clr_object.rds")
)

write_csv(
  da_preprocessing_summary,
  file = path_target("diff_abundance_preprocessing_summary.csv")
)
```

## Differential abundance setup

The differential abundance analysis compares children with no exclusive
breastfeeding (`EBF duration == "0 months"`) to children exclusively breastfed
for at least two months (`EBF duration == ">=2 months"`). Samples in the sparse
`1 month` category and samples with missing EBF duration are excluded from this
two-group comparison.

```{r dacomp_setup}
dacomp_sample_data <- sampdata_da %>%
  mutate(
    EBF_group = case_when(
      `EBF duration` == "0 months" ~ "non-EBF",
      `EBF duration` == ">=2 months" ~ "EBF",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EBF_group)) %>%
  mutate(EBF_group = factor(EBF_group, levels = c("non-EBF", "EBF"))) %>%
  select(SampleID, `EBF duration`, EBF_group)

X_counts <- counts_raw[dacomp_sample_data$SampleID, , drop = FALSE]
X_rel <- counts_rel[dacomp_sample_data$SampleID, , drop = FALSE]
X_clr <- counts_clr[dacomp_sample_data$SampleID, , drop = FALSE]
y_groups <- dacomp_sample_data$EBF_group

stopifnot(identical(rownames(X_counts), dacomp_sample_data$SampleID))
stopifnot(identical(rownames(X_clr), dacomp_sample_data$SampleID))

dacomp_group_summary <- dacomp_sample_data %>%
  count(EBF_group, name = "n_samples")

dacomp_analysis_summary <- tibble(
  n_samples = nrow(X_counts),
  n_taxa = ncol(X_counts),
  n_non_ebf = sum(y_groups == "non-EBF"),
  n_ebf = sum(y_groups == "EBF"),
  excluded_ebf_one_month = sum(sampdata_da$`EBF duration` == "1 month", na.rm = TRUE),
  excluded_missing_ebf = sum(is.na(sampdata_da$`EBF duration`))
)

dacomp_group_summary
dacomp_analysis_summary
```

```{r save_dacomp_setup}
write_csv(
  dacomp_group_summary,
  file = path_target("diff_abundance_group_summary.csv")
)

write_csv(
  dacomp_analysis_summary,
  file = path_target("diff_abundance_analysis_summary.csv")
)

saveRDS(
  list(
    sample_data = dacomp_sample_data,
    X_counts = X_counts,
    X_rel = X_rel,
    X_clr = X_clr,
    y_groups = y_groups
  ),
  file = path_target("diff_abundance_analysis_objects.rds")
)
```

## dacomp reference selection

```{r dacomp_select_references}
path_dacomp_refs <- path_target("dacomp_reference_selection.rds")

if (!file.exists(path_dacomp_refs) || isTRUE(refit_dacomp_refs)) {
  set.seed(permutation_seed)
  
  res_refs <- dacomp.select_references(
    X = X_counts,
    minimal_TA = 50,
    verbose = TRUE
  )
  
  saveRDS(res_refs, file = path_dacomp_refs)
} else {
  res_refs <- readRDS(path_dacomp_refs)
}

dacomp_reference_summary <- tibble(
  n_taxa = ncol(X_counts),
  n_reference_taxa = length(res_refs$selected_references),
  n_test_taxa = ncol(X_counts) - length(res_refs$selected_references),
  minimal_TA = 50
)

dacomp_reference_summary
```

```{r save_dacomp_reference_summary}
write_csv(
  dacomp_reference_summary,
  file = path_target("dacomp_reference_summary.csv")
)
```

### Reference score diagnostic

```{r dacomp_reference_score_data}
dacomp_reference_score_df <- tibble(
  taxon = colnames(X_counts),
  prevalence = colMeans(X_counts > 0),
  score = as.numeric(res_refs$scores),
  reference_status = if_else(
    seq_along(colnames(X_counts)) %in% res_refs$selected_references,
    "reference",
    "test"
  )
) %>%
  mutate(
    reference_status = factor(reference_status, levels = c("test", "reference"))
  )

write_csv(
  dacomp_reference_score_df,
  file = path_target("dacomp_reference_score_data.csv")
)

dacomp_reference_score_df
```

```{r dacomp_reference_score_plot, fig.width=8, fig.height=4, dpi=200}
dacomp_reference_labels <- dacomp_reference_score_df %>%
  filter(reference_status == "reference", prevalence >= 0.25)

ggplot(dacomp_reference_score_df, aes(x = prevalence, y = score)) +
  geom_point(aes(color = reference_status), size = 2, alpha = 0.9) +
  ggrepel::geom_text_repel(
    data = dacomp_reference_labels,
    aes(label = taxon),
    color = "black",
    size = 4.5,
    box.padding = 0.6,
    point.padding = 0.3,
    min.segment.length = grid::unit(0, "lines"),
    segment.alpha = 0.5,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  scale_color_manual(
    values = c("test" = "grey70", "reference" = "forestgreen"),
    name = NULL
  ) +
  labs(
    #title = "Prevalence vs dacomp score (genus level)",
    x = "Prevalence (proportion of samples with count > 0)",
    y = "dacomp median SD score"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## dacomp permutation tests

The following chunk runs the long dacomp permutation analyses. Each permutation
count is cached in a separate result file and runtime file. If the result file
already exists, the corresponding run is skipped unless `refit_dacomp` is set to
`TRUE`.

```{r run_dacomp_tests}
dacomp_runtime_tbl <- purrr::map_dfr(
  dacomp_perm_grid,
  function(n_perm) {
    set.seed(permutation_seed)
    
    fname_res <- make_dacomp_result_path(n_perm, seed = permutation_seed)
    fname_rt <- make_dacomp_runtime_path(n_perm, seed = permutation_seed)
    
    if (!file.exists(fname_res) || isTRUE(refit_dacomp)) {
      cat(
        "Running dacomp:",
        "n_perm =", fmt_nperm(n_perm),
        "| seed =", permutation_seed,
        "\n"
      )
      
      runtime <- system.time({
        res_dacomp <- run_dacomp(
          X_counts = X_counts,
          y_groups = y_groups,
          res_refs = res_refs,
          n_perm = n_perm,
          q = q_DSFDR
        )
      })
      
      saveRDS(res_dacomp, file = fname_res)
      
      runtime_row <- tibble(
        n_perm = n_perm,
        elapsed_sec = unname(runtime["elapsed"]),
        seed = permutation_seed,
        result_file = basename(fname_res)
      )
      
      saveRDS(runtime_row, file = fname_rt)
    } else {
      cat(
        "Skipping cached dacomp run:",
        "n_perm =", fmt_nperm(n_perm),
        "| seed =", permutation_seed,
        "\n"
      )
      
      runtime_row <- if (file.exists(fname_rt)) {
        readRDS(fname_rt)
      } else {
        tibble(
          n_perm = n_perm,
          elapsed_sec = NA_real_,
          seed = permutation_seed,
          result_file = basename(fname_res)
        )
      }
    }
    
    runtime_row
  }
) %>%
  arrange(n_perm)

dacomp_runtime_tbl
```

```{r save_dacomp_runtime}
write_csv(
  dacomp_runtime_tbl,
  file = path_target("dacomp_runtime.csv")
)
```

## Runtime preparation

The dacomp runtime rows are initialized here from the cached dacomp runs. The
final runtime comparison table and figure are created after the permApprox
runtime has been added.

```{r save_dacomp_stats_for_permapprox}
dacomp_stats_index_tbl <- purrr::map_dfr(
  dacomp_runtime_tbl$n_perm,
  function(n_perm) {
    fname_res <- make_dacomp_result_path(n_perm, seed = permutation_seed)
    fname_stats <- make_dacomp_stats_path(n_perm, seed = permutation_seed)
    
    if (
      isTRUE(write_dacomp_stats_files) &&
        file.exists(fname_res) &&
        !file.exists(fname_stats)
    ) {
      res_dacomp <- readRDS(fname_res)
      dacomp_stats <- extract_dacomp_stats(res_dacomp)
      saveRDS(dacomp_stats, file = fname_stats)
    }
    
    tibble(
      n_perm = n_perm,
      seed = permutation_seed,
      result_file = basename(fname_res),
      stats_file = basename(fname_stats),
      result_exists = file.exists(fname_res),
      stats_exists = file.exists(fname_stats)
    )
  }
)

dacomp_stats_index_tbl
```

```{r save_dacomp_stats_index}
write_csv(
  dacomp_stats_index_tbl,
  file = path_target("dacomp_stats_index.csv")
)
```

```{r runtime_dacomp_rows, include=FALSE}
runtime_total_tbl <- dacomp_runtime_tbl %>%
  transmute(
    line = "dacomp only",
    method = "dacomp",
    setting = "dacomp",
    n_perm = n_perm,
    elapsed_s_total = elapsed_sec,
    seed = seed,
    result_file = result_file
  ) %>%
  mutate(
    elapsed_s = round(elapsed_s_total, 2),
    elapsed_min = round(elapsed_s_total / 60, 2),
    elapsed_h = round(elapsed_s_total / 3600, 3)
  ) %>%
  arrange(line, n_perm)
```

## p-value refinement with permApprox

The dacomp result with 1000 permutations is used as the empirical permutation
sample for the GPD tail approximation. We fit both the unconstrained
`permApprox(U)` variant and the constrained `permApprox(C)` variant. The main
comparison and volcano plot use the constrained variant. Since dacomp's
permutation statistics are non-negative test statistics and large values are
more extreme, the GPD tail approximation is fitted to the right tail.

```{r permapprox_helpers}
refit_permapprox <- FALSE
permapprox_n_perm <- 1e3
permapprox_constraint <- "unconstrained"
permapprox_constrained_constraint <- "support_at_obs"
permapprox_exceed0 <- 250
permapprox_alternative <- "greater"
alpha_diff_abundance <- 0.01

make_permapprox_result_path <- function(
    n_perm = permapprox_n_perm,
    constraint = permapprox_constraint,
    exceed0 = permapprox_exceed0,
    alternative = permapprox_alternative,
    seed = permutation_seed
) {
  path_target(
    paste0(
      "permapprox_nperm=", fmt_nperm(n_perm),
      "_constraint=", constraint,
      "_exceed0=", exceed0,
      "_alternative=", alternative,
      "_seed=", seed,
      ".rds"
    )
  )
}

make_permapprox_ctrl <- function(
    sample_size,
    constraint = permapprox_constraint,
    exceed0 = permapprox_exceed0
) {
  permApprox::make_gpd_ctrl(
    fit_method = "lme",
    include_obs = FALSE,
    constraint = constraint,
    sample_size = sample_size,
    thresh_method = "fix",
    exceed0 = exceed0,
    exceed0_min = exceed0,
    exceed_min = 10,
    thresh_step = 10
  )
}

expand_permapprox_result <- function(res, valid, n_total) {
  n_valid <- sum(valid)
  
  for (nm in names(res)) {
    x <- res[[nm]]
    
    if (is.atomic(x) && length(x) == n_valid) {
      out <- rep(NA, n_total)
      out[valid] <- x
      res[[nm]] <- out
    } else if (is.data.frame(x) && nrow(x) == n_valid) {
      out <- x[rep(1, n_total), , drop = FALSE]
      out[,] <- NA
      out[valid, ] <- x
      rownames(out) <- NULL
      res[[nm]] <- out
    } else if (is.matrix(x) && nrow(x) == n_valid) {
      out <- matrix(
        NA,
        nrow = n_total,
        ncol = ncol(x),
        dimnames = list(NULL, colnames(x))
      )
      out[valid, ] <- x
      res[[nm]] <- out
    }
  }
  
  res$valid_test <- valid
  res
}

run_permapprox_from_dacomp <- function(
    res_dacomp,
    sample_size,
    cores = 1,
    constraint = permapprox_constraint,
    exceed0 = permapprox_exceed0,
    alternative = permapprox_alternative
) {
  dacomp_stats <- extract_dacomp_stats(res_dacomp)
  obs_stats <- as.numeric(dacomp_stats$obs_stats)
  names(obs_stats) <- names(dacomp_stats$obs_stats)
  perm_stats <- dacomp_stats$perm_stats
  
  valid <- is.finite(obs_stats) & colSums(is.finite(perm_stats)) > 0
  
  if (!any(valid)) {
    stop("No valid dacomp statistics available for permApprox.")
  }
  
  gpd_ctrl <- make_permapprox_ctrl(
    sample_size = sample_size,
    constraint = constraint,
    exceed0 = exceed0
  )
  
  res <- permApprox::perm_approx(
    obs_stats = obs_stats[valid],
    perm_stats = perm_stats[, valid, drop = FALSE],
    method = "gpd",
    alternative = alternative,
    approx_thresh = 0.1,
    adjust_method = "BH",
    gpd_ctrl = gpd_ctrl,
    cores = cores,
    verbose = TRUE
  )
  
  expand_permapprox_result(
    res,
    valid = valid,
    n_total = length(obs_stats)
  )
}

dacomp_to_tbl <- function(res_dacomp, n_perm, taxa_ids, res_refs) {
  tibble(
    taxon_id = taxa_ids,
    statistic = as.numeric(res_dacomp$stats_matrix[1, ]),
    p_raw = as.numeric(res_dacomp$p.values.test),
    p_adj = as.numeric(res_dacomp$p.values.test.adjusted),
    is_reference = seq_along(taxa_ids) %in% res_refs$selected_references,
    method = "dacomp",
    n_perm = n_perm,
    constraint = NA_character_,
    exceed0 = NA_real_
  ) %>%
    mutate(
      p_adj = if_else(is.infinite(p_adj), 1, p_adj)
    )
}

permapprox_to_tbl <- function(
    papp,
    n_perm,
    taxa_ids,
    res_refs,
    constraint = permapprox_constraint,
    exceed0 = permapprox_exceed0
) {
  tibble(
    taxon_id = taxa_ids,
    statistic = NA_real_,
    p_raw = as.numeric(papp$p_unadjusted),
    p_adj = as.numeric(papp$p_values),
    is_reference = seq_along(taxa_ids) %in% res_refs$selected_references,
    method = "permApprox",
    n_perm = n_perm,
    constraint = constraint,
    exceed0 = exceed0,
    method_used = papp$method_used
  )
}
```

```{r run_permapprox}
path_permapprox <- make_permapprox_result_path()
path_dacomp_1e3 <- make_dacomp_result_path(
  permapprox_n_perm,
  seed = permutation_seed
)

if (!file.exists(path_dacomp_1e3)) {
  stop("Missing dacomp result required for permApprox: ", path_dacomp_1e3)
}

if (!file.exists(path_permapprox) || isTRUE(refit_permapprox)) {
  res_dacomp_1e3 <- readRDS(path_dacomp_1e3)
  
  permapprox_runtime <- system.time({
    da_permapprox <- run_permapprox_from_dacomp(
      res_dacomp = res_dacomp_1e3,
      sample_size = min(dacomp_group_summary$n_samples),
      cores = cores,
      constraint = permapprox_constraint,
      exceed0 = permapprox_exceed0,
      alternative = permapprox_alternative
    )
  })
  
  saveRDS(
    list(
      papprox = da_permapprox,
      elapsed_sec = unname(permapprox_runtime["elapsed"]),
      n_perm = permapprox_n_perm,
      constraint = permapprox_constraint,
      exceed0 = permapprox_exceed0,
      alternative = permapprox_alternative,
      seed = permutation_seed
    ),
    file = path_permapprox
  )
  
  rm(res_dacomp_1e3)
} else {
  message("Skipping cached permApprox fit: ", basename(path_permapprox))
}

permapprox_result <- readRDS(path_permapprox)
summary(permapprox_result$papprox)
```

```{r run_permapprox_constrained}
path_permapprox_constrained <- make_permapprox_result_path(
  constraint = permapprox_constrained_constraint
)

if (!file.exists(path_permapprox_constrained) || isTRUE(refit_permapprox)) {
  res_dacomp_1e3 <- readRDS(path_dacomp_1e3)
  
  permapprox_constrained_runtime <- system.time({
    da_permapprox_constrained <- run_permapprox_from_dacomp(
      res_dacomp = res_dacomp_1e3,
      sample_size = min(dacomp_group_summary$n_samples),
      cores = cores,
      constraint = permapprox_constrained_constraint,
      exceed0 = permapprox_exceed0,
      alternative = permapprox_alternative
    )
  })
  
  saveRDS(
    list(
      papprox = da_permapprox_constrained,
      elapsed_sec = unname(permapprox_constrained_runtime["elapsed"]),
      n_perm = permapprox_n_perm,
      constraint = permapprox_constrained_constraint,
      exceed0 = permapprox_exceed0,
      alternative = permapprox_alternative,
      seed = permutation_seed
    ),
    file = path_permapprox_constrained
  )
  
  rm(res_dacomp_1e3)
} else {
  message(
    "Skipping cached constrained permApprox fit: ",
    basename(path_permapprox_constrained)
  )
}

permapprox_constrained_result <- readRDS(path_permapprox_constrained)
summary(permapprox_constrained_result$papprox)
```

```{r update_runtime_with_permapprox}
dacomp_1e3_runtime <- dacomp_runtime_tbl %>%
  filter(n_perm == permapprox_result$n_perm) %>%
  slice(1)

permapprox_runtime_row <- tibble(
  line = "dacomp + permApprox(U)",
  method = "dacomp + permApprox(U)",
  setting = paste0(
    permapprox_result$constraint,
    ", exceed0=", permapprox_result$exceed0
  ),
  n_perm = permapprox_result$n_perm,
  elapsed_s_total = dacomp_1e3_runtime$elapsed_sec + permapprox_result$elapsed_sec,
  seed = permapprox_result$seed,
  result_file = basename(path_permapprox)
)

permapprox_constrained_runtime_row <- tibble(
  line = "dacomp + permApprox(C)",
  method = "dacomp + permApprox(C)",
  setting = paste0(
    permapprox_constrained_result$constraint,
    ", exceed0=", permapprox_constrained_result$exceed0
  ),
  n_perm = permapprox_constrained_result$n_perm,
  elapsed_s_total = dacomp_1e3_runtime$elapsed_sec +
    permapprox_constrained_result$elapsed_sec,
  seed = permapprox_constrained_result$seed,
  result_file = basename(path_permapprox_constrained)
)

runtime_total_tbl <- bind_rows(
  runtime_total_tbl,
  permapprox_runtime_row,
  permapprox_constrained_runtime_row
) %>%
  mutate(
    elapsed_s = round(elapsed_s_total, 2),
    elapsed_min = round(elapsed_s_total / 60, 2),
    elapsed_h = round(elapsed_s_total / 3600, 3)
  ) %>%
  arrange(method, n_perm)

runtime_total_tbl
```

```{r save_runtime_with_permapprox}
saveRDS(
  runtime_total_tbl,
  file = path_target("runtime_total_tbl.rds")
)

write_csv(
  runtime_total_tbl,
  file = path_target("runtime_total_tbl.csv")
)
```

## Runtime comparison

The runtime comparison includes the empirical dacomp runs and the combined
runtimes for the 1000-permutation dacomp run plus the unconstrained and
constrained permApprox tail approximations.

```{r runtime_comparison_display}
runtime_total_tbl %>%
  select(line, n_perm, elapsed_s, elapsed_min, elapsed_h) %>%
  knitr::kable()
```

```{r runtime_plot_data}
runtime_plot_tbl <- runtime_total_tbl %>%
  filter(!is.na(elapsed_s_total), !is.na(n_perm)) %>%
  mutate(
    line = factor(
      line,
      levels = c(
        "dacomp only",
        "dacomp + permApprox(U)",
        "dacomp + permApprox(C)"
      )
    )
  )

runtime_plot_tbl
```

```{r runtime_plot, fig.width=7.5, fig.height=3.6, dpi=300}
x_grid <- sort(unique(runtime_plot_tbl$n_perm))
minute_grid <- c(1, 2, 5, 10, 30, 60, 120, 300)
second_grid <- c(1, 3, 10, minute_grid * 60)

ggplot(runtime_plot_tbl, aes(
  x = n_perm,
  y = elapsed_s_total,
  colour = line,
  group = line
)) +
  geom_hline(yintercept = second_grid, color = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = x_grid, color = "grey90", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.2, stroke = 0.9) +
  scale_x_log10(
    breaks = x_grid,
    labels = scales::label_scientific()
  ) +
  scale_y_log10(
    name = "Runtime (seconds)",
    breaks = second_grid,
    labels = scales::label_number(),
    sec.axis = sec_axis(
      trans = ~ . / 60,
      name = "Runtime (minutes)",
      breaks = minute_grid,
      labels = scales::label_number()
    )
  ) +
  scale_colour_manual(
    values = c(
      "dacomp only" = "black",
      "dacomp + permApprox(U)" = "darkorange2",
      "dacomp + permApprox(C)" = "#1f77b4"
    ),
    drop = FALSE
  ) +
  labs(x = "Number of permutations", colour = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = c(1, 0.1),
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(
      fill = "white",
      color = "grey80",
      linewidth = 0.3
    ),
    legend.key = element_rect(fill = "white")
  )
```

## Differential abundance results

```{r taxon_labels}
taxon_labels_from_phyloseq <- function(phylo) {
  tax_tab <- tryCatch(
    as(tax_table(phylo), "matrix"),
    error = function(e) NULL
  )
  
  tax_ids <- taxa_names(phylo)
  tax_labels <- tax_ids
  
  if (!is.null(tax_tab) && "Genus" %in% colnames(tax_tab)) {
    genus <- as.character(tax_tab[tax_ids, "Genus"])
    genus <- stringr::str_remove(genus, "^g__")
    missing_label <- is.na(genus) | genus == "" | genus == "NA"
    genus[missing_label] <- tax_ids[missing_label]
    tax_labels <- make.unique(genus)
  }
  
  tibble(taxon_id = tax_ids, taxon = tax_labels)
}

taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)
taxa_ids <- colnames(X_counts)

format_diff_abundance_method_label <- function(method, n_perm, constraint) {
  case_when(
    method == "dacomp" ~ paste0("dacomp (", fmt_nperm(n_perm), ")"),
    method == "permApprox" & constraint == "unconstrained" ~
      paste0("dacomp + permApprox(U) (", fmt_nperm(n_perm), ")"),
    method == "permApprox" & constraint == "support_at_obs" ~
      paste0("dacomp + permApprox(C) (", fmt_nperm(n_perm), ")"),
    method == "permApprox" ~ paste0("dacomp + permApprox (", fmt_nperm(n_perm), ")"),
    TRUE ~ method
  )
}
```

```{r effect_sizes}
mean_rel_non_ebf <- colMeans(X_rel[y_groups == "non-EBF", , drop = FALSE])
mean_rel_ebf <- colMeans(X_rel[y_groups == "EBF", , drop = FALSE])
mean_clr_non_ebf <- colMeans(X_clr[y_groups == "non-EBF", , drop = FALSE])
mean_clr_ebf <- colMeans(X_clr[y_groups == "EBF", , drop = FALSE])

fc_pseudocount <- 1e-6

effect_size_tbl <- tibble(
  taxon_id = taxa_ids,
  mean_rel_non_ebf = as.numeric(mean_rel_non_ebf[taxa_ids]),
  mean_rel_ebf = as.numeric(mean_rel_ebf[taxa_ids]),
  mean_clr_non_ebf = as.numeric(mean_clr_non_ebf[taxa_ids]),
  mean_clr_ebf = as.numeric(mean_clr_ebf[taxa_ids]),
  log2_fc = log2((mean_rel_ebf + fc_pseudocount) /
                   (mean_rel_non_ebf + fc_pseudocount)),
  clr_difference = mean_clr_ebf - mean_clr_non_ebf,
  prevalence_non_ebf = colMeans(X_counts[y_groups == "non-EBF", , drop = FALSE] > 0),
  prevalence_ebf = colMeans(X_counts[y_groups == "EBF", , drop = FALSE] > 0)
) %>%
  left_join(taxon_labels, by = "taxon_id")

effect_size_tbl
```

```{r collect_dacomp_results}
selected_dacomp_perm_grid <- dacomp_perm_grid

dacomp_results_long <- purrr::map_dfr(
  selected_dacomp_perm_grid,
  function(n_perm) {
    fname_res <- make_dacomp_result_path(n_perm, seed = permutation_seed)
    
    if (!file.exists(fname_res)) {
      stop("Missing dacomp result file: ", fname_res)
    }
    
    res_dacomp <- readRDS(fname_res)
    out <- dacomp_to_tbl(
      res_dacomp = res_dacomp,
      n_perm = n_perm,
      taxa_ids = taxa_ids,
      res_refs = res_refs
    )
    
    rm(res_dacomp)
    gc()
    
    out
  }
)

permapprox_results_long <- permapprox_to_tbl(
  papp = permapprox_result$papprox,
  n_perm = permapprox_result$n_perm,
  taxa_ids = taxa_ids,
  res_refs = res_refs,
  constraint = permapprox_result$constraint,
  exceed0 = permapprox_result$exceed0
)

permapprox_constrained_results_long <- permapprox_to_tbl(
  papp = permapprox_constrained_result$papprox,
  n_perm = permapprox_constrained_result$n_perm,
  taxa_ids = taxa_ids,
  res_refs = res_refs,
  constraint = permapprox_constrained_result$constraint,
  exceed0 = permapprox_constrained_result$exceed0
)

diff_abundance_results_long <- bind_rows(
  dacomp_results_long,
  permapprox_results_long,
  permapprox_constrained_results_long
) %>%
  left_join(effect_size_tbl, by = "taxon_id") %>%
  mutate(
    method_label = format_diff_abundance_method_label(method, n_perm, constraint),
    panel_label = case_when(
      method == "dacomp" & n_perm == 1e3 ~ "dacomp\n1000 permutations",
      method == "dacomp" & n_perm == 1e7 ~ "dacomp\n1e+07 permutations",
      method == "permApprox" & constraint == "support_at_obs" & n_perm == 1e3 ~
        "dacomp + permApprox\n1000 permutations",
      TRUE ~ method_label
    ),
    panel_label = factor(
      panel_label,
      levels = c(
        "dacomp\n1000 permutations",
        "dacomp\n1e+07 permutations",
        "dacomp + permApprox\n1000 permutations"
      )
    )
  )

diff_abundance_results_long %>%
  count(method, n_perm, constraint, exceed0)
```

```{r save_diff_abundance_results}
saveRDS(
  diff_abundance_results_long,
  file = path_target("diff_abundance_results_long.rds")
)

write_csv(
  diff_abundance_results_long,
  file = path_target("diff_abundance_results_long.csv")
)

diff_abundance_results_wide <- diff_abundance_results_long %>%
  select(
    taxon_id, taxon, log2_fc, clr_difference, mean_rel_non_ebf, mean_rel_ebf,
    prevalence_non_ebf, prevalence_ebf, method_label, p_raw, p_adj
  ) %>%
  pivot_wider(
    names_from = method_label,
    values_from = c(p_raw, p_adj),
    names_glue = "{.value}_{method_label}"
  )

write_csv(
  diff_abundance_results_wide,
  file = path_target("diff_abundance_results_wide.csv")
)
```

## P-value tables for significant taxa

The following tables collect the taxa that are significant in at least one of
the selected dacomp or constrained permApprox analyses. Both unadjusted and
adjusted p-values are shown because the volcano plot is based on unadjusted
p-values, while the adjusted p-values give the multiple-testing corrected view
of the same taxa.

```{r pvalue_table_helpers}
pvalue_table_methods <- c(
  paste0("dacomp (", fmt_nperm(1e3), ")"),
  paste0("dacomp (", fmt_nperm(1e6), ")"),
  paste0("dacomp (", fmt_nperm(1e7), ")"),
  paste0("dacomp + permApprox(C) (", fmt_nperm(permapprox_n_perm), ")")
)

pvalue_table_col_names <- c(
  "Taxon",
  "dacomp statistic",
  "dacomp ($10^3$)",
  "dacomp ($10^6$)",
  "dacomp ($10^7$)",
  "\\texttt{permApprox} (C)"
)

fmt_p_table <- function(p_num, digits = 2) {
  if_else(
    is.na(p_num),
    NA_character_,
    formatC(p_num, format = "e", digits = digits)
  )
}

fmt_p_star <- function(p_num, alpha = alpha_diff_abundance, digits = 2) {
  p_fmt <- fmt_p_table(p_num, digits = digits)
  
  if_else(
    !is.na(p_num) & p_num < alpha,
    paste0(p_fmt, "*"),
    p_fmt
  )
}

make_significant_pvalue_table <- function(results_long, p_col, alpha) {
  statistic_tbl <- results_long %>%
    filter(method == "dacomp", n_perm == permapprox_n_perm) %>%
    select(taxon_id, statistic)
  
  p_wide <- results_long %>%
    filter(method_label %in% pvalue_table_methods) %>%
    select(taxon_id, taxon, method_label, p_value = all_of(p_col)) %>%
    pivot_wider(names_from = method_label, values_from = p_value) %>%
    left_join(statistic_tbl, by = "taxon_id") %>%
    select(taxon_id, taxon, statistic, all_of(pvalue_table_methods))
  
  p_wide %>%
    filter(if_any(all_of(pvalue_table_methods), ~ !is.na(.x) & .x < alpha)) %>%
    arrange(.data[[pvalue_table_methods[[1]]]]) %>%
    mutate(statistic = round(statistic, 2))
}
```

### Unadjusted p-values

```{r significant_raw_pvalues_table}
significant_raw_pvalues <- make_significant_pvalue_table(
  results_long = diff_abundance_results_long,
  p_col = "p_raw",
  alpha = alpha_diff_abundance
)

significant_raw_pvalues_print <- significant_raw_pvalues %>%
  mutate(
    across(
      all_of(pvalue_table_methods),
      ~ fmt_p_star(.x, alpha = alpha_diff_abundance)
    )
  )

significant_raw_pvalues_print %>%
  select(-taxon_id) %>%
  knitr::kable(
    caption = paste0(
      "Unadjusted p-values for taxa significant in at least one selected ",
      "differential abundance analysis. Asterisks mark p < ",
      alpha_diff_abundance,
      "."
    )
  )
```

```{r save_significant_raw_pvalues_table}
write_csv(
  significant_raw_pvalues,
  file = path_target("diff_abundance_significant_raw_pvalues.csv")
)
```

```{r significant_raw_pvalues_latex}
latex_raw_pvalues <- significant_raw_pvalues %>%
  select(taxon, statistic, all_of(pvalue_table_methods)) %>%
  mutate(
    statistic = round(statistic, 2),
    across(
      all_of(pvalue_table_methods),
      ~ fmt_p_star(.x, alpha = alpha_diff_abundance)
    )
  )

tex_raw_pvalues <- latex_raw_pvalues %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    col.names = pvalue_table_col_names,
    align = "lrrrrr",
    caption = paste0(
      "Unadjusted p-values for taxa significant in at least one selected ",
      "differential abundance analysis. Asterisks mark p-values below ",
      "$\\alpha = ", alpha_diff_abundance, "$."
    ),
    label = "application:diff-abundance-pvalues"
  ) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(
    general = paste0(
      "The permApprox column uses the constrained GPD tail approximation ",
      "based on the 1000-permutation dacomp run."
    ),
    general_title = "",
    escape = FALSE,
    threeparttable = TRUE
  )

tex_raw_pvalues
save_kable(
  tex_raw_pvalues,
  file = path_target("diff_abundance_raw_pvalues_table.tex")
)
```

### Adjusted p-values

```{r significant_adjusted_pvalues_table}
significant_adjusted_pvalues <- make_significant_pvalue_table(
  results_long = diff_abundance_results_long,
  p_col = "p_adj",
  alpha = alpha_diff_abundance
)

significant_adjusted_pvalues_print <- significant_adjusted_pvalues %>%
  mutate(
    across(
      all_of(pvalue_table_methods),
      ~ fmt_p_star(.x, alpha = alpha_diff_abundance)
    )
  )

significant_adjusted_pvalues_print %>%
  select(-taxon_id) %>%
  knitr::kable(
    caption = paste0(
      "Adjusted p-values for taxa significant in at least one selected ",
      "differential abundance analysis. Asterisks mark adjusted p < ",
      alpha_diff_abundance,
      "."
    )
  )
```

```{r save_significant_adjusted_pvalues_table}
write_csv(
  significant_adjusted_pvalues,
  file = path_target("diff_abundance_significant_adjusted_pvalues.csv")
)
```

```{r significant_adjusted_pvalues_latex}
latex_adjusted_pvalues <- significant_adjusted_pvalues %>%
  select(taxon, statistic, all_of(pvalue_table_methods)) %>%
  mutate(
    statistic = round(statistic, 2),
    across(
      all_of(pvalue_table_methods),
      ~ fmt_p_star(.x, alpha = alpha_diff_abundance)
    )
  )

tex_adjusted_pvalues <- latex_adjusted_pvalues %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    col.names = pvalue_table_col_names,
    align = "lrrrrr",
    caption = paste0(
      "Adjusted p-values for taxa significant in at least one selected ",
      "differential abundance analysis. Asterisks mark adjusted p-values ",
      "below $\\alpha = ", alpha_diff_abundance, "$."
    ),
    label = "application:diff-abundance-adjusted-pvalues"
  ) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  footnote(
    general = paste0(
      "The permApprox column uses the constrained GPD tail approximation ",
      "based on the 1000-permutation dacomp run."
    ),
    general_title = "",
    escape = FALSE,
    threeparttable = TRUE
  )

tex_adjusted_pvalues
save_kable(
  tex_adjusted_pvalues,
  file = path_target("diff_abundance_adjusted_pvalues_table.tex")
)
```

```{r significant_results_table}
diff_abundance_sig_table <- diff_abundance_results_long %>%
  filter(p_raw < alpha_diff_abundance) %>%
  arrange(p_raw, desc(abs(log2_fc))) %>%
  transmute(
    method = method_label,
    taxon,
    log2_fc = round(log2_fc, 2),
    mean_rel_non_ebf = signif(mean_rel_non_ebf, 3),
    mean_rel_ebf = signif(mean_rel_ebf, 3),
    p_raw = formatC(p_raw, format = "e", digits = 2),
    p_adj = formatC(p_adj, format = "e", digits = 2)
  )

diff_abundance_sig_table %>%
  knitr::kable(
    caption = paste0(
      "Taxa with raw p-values below ", alpha_diff_abundance,
      " in at least one selected differential abundance analysis."
    )
  )
```

```{r save_significant_results_table}
write_csv(
  diff_abundance_sig_table,
  file = path_target("diff_abundance_significant_results.csv")
)
```

## Volcano plot

The volcano plot compares the empirical dacomp p-values with 1000 and
`10^7` permutations to the constrained permApprox refinement based on the 1000
permutation run. The left panel highlights the empirical p-value bound for 1000
permutations, while the permApprox panel shows how the GPD tail approximation
can resolve smaller p-values from the same empirical permutation sample.

```{r volcano_plot_data}
volcano_y_cutoff <- 30
volcano_y_ellipsis <- 33
volcano_y_top <- 36
volcano_label_p_threshold <- 1e-5

volcano_plot_df <- diff_abundance_results_long %>%
  filter(!is.na(panel_label)) %>%
  mutate(
    p_plot_raw = p_raw,
    neglog10_p_raw = if_else(!is.na(p_plot_raw) & p_plot_raw > 0,
                             -log10(p_plot_raw),
                             NA_real_),
    significant = !is.na(p_plot_raw) & p_plot_raw < alpha_diff_abundance,
    direction = case_when(
      significant & log2_fc > 0 ~ "Higher in 'EBF'",
      significant & log2_fc < 0 ~ "Higher in 'non-EBF'",
      TRUE ~ "No difference"
    ),
    direction = factor(
      direction,
      levels = c("Higher in 'EBF'", "Higher in 'non-EBF'", "No difference")
    )
  )

has_large_p <- any(volcano_plot_df$neglog10_p_raw > volcano_y_cutoff, 
                   na.rm = TRUE)
has_zero_p <- any(volcano_plot_df$p_plot_raw == 0, na.rm = TRUE)
max_large_neglog10 <- max(volcano_plot_df$neglog10_p_raw, na.rm = TRUE)
volcano_y_top_label <- if (has_zero_p) {
  "Inf"
} else {
  as.character(max(100, ceiling(max_large_neglog10 / 10) * 10))
}

volcano_plot_df <- volcano_plot_df %>%
  mutate(
    neglog10_p_display = case_when(
      is.na(neglog10_p_raw) ~ NA_real_,
      p_plot_raw == 0 ~ volcano_y_top,
      neglog10_p_raw > volcano_y_cutoff ~ volcano_y_top,
      TRUE ~ neglog10_p_raw
    )
  )

if (has_large_p || has_zero_p) {
  volcano_y_breaks <- c(0, 10, 20, 30, volcano_y_ellipsis, volcano_y_top)
  volcano_y_labels <- c("0", "10", "20", "30", "...", volcano_y_top_label)
  volcano_y_limits <- c(0, volcano_y_top + 2)
} else {
  volcano_y_breaks <- pretty(c(0, max_large_neglog10))
  volcano_y_labels <- as.character(volcano_y_breaks)
  volcano_y_limits <- c(0, max(volcano_y_breaks, na.rm = TRUE))
}

volcano_label_taxa <- volcano_plot_df %>%
  group_by(taxon_id) %>%
  summarise(
    label_taxon = any(p_plot_raw < volcano_label_p_threshold, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(label_taxon) %>%
  pull(taxon_id)

volcano_label_df <- volcano_plot_df %>%
  filter(taxon_id %in% volcano_label_taxa, !is.na(neglog10_p_display))

volcano_plot_df
```

```{r save_volcano_plot_data}
write_csv(
  volcano_plot_df,
  file = path_target("diff_abundance_volcano_plot_data.csv")
)
```

```{r volcano_plot_raw, fig.width=10, fig.height=5, dpi=200}
set.seed(permutation_seed)

ggplot(volcano_plot_df, aes(x = log2_fc, y = neglog10_p_display)) +
  geom_point(aes(color = direction), alpha = 0.8, size = 2.6) +
  geom_hline(
    yintercept = c(3, 7),
    linetype = "dotted",
    color = "grey40",
    linewidth = 0.5
  ) +
  geom_hline(
    yintercept = -log10(alpha_diff_abundance),
    linetype = "dashed",
    color = "grey55"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey55"
  ) +
  ggrepel::geom_text_repel(
    data = volcano_label_df,
    aes(label = taxon),
    size = 4,
    box.padding = 0.8,
    point.padding = 0.25,
    min.segment.length = grid::unit(0, "lines"),
    segment.alpha = 0.45,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  facet_wrap(~ panel_label, nrow = 1) +
  scale_color_manual(
    values = c(
      "Higher in 'EBF'" = "darkorange2",
      "Higher in 'non-EBF'" = "#1f77b4",
      "No difference" = "grey70"
    ),
    drop = FALSE,
    name = NULL
  ) +
  scale_y_continuous(
    breaks = volcano_y_breaks,
    labels = volcano_y_labels,
    limits = volcano_y_limits
  ) +
  labs(
    x = "log2 fold-change",
    y = expression(-log[10](p))
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.spacing = grid::unit(1.5, "lines"),
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 16)
  )
```

## Permutation distributions

The volcano plot summarizes the p-values, but it is useful to inspect the
empirical permutation distributions behind the GPD approximation. The following
diagnostic plot shows all taxa labelled in the volcano plot. The red vertical
line marks the observed dacomp statistic used for the right-tail GPD fit.

```{r permutation_distribution_data}
res_dacomp_permapprox <- readRDS(path_dacomp_1e3)
dacomp_stats_permapprox <- extract_dacomp_stats(res_dacomp_permapprox)

dacomp_1e7_pvalues <- diff_abundance_results_long %>%
  filter(method == "dacomp", n_perm == 1e7) %>%
  select(taxon_id, p_dacomp_1e7 = p_raw)

dacomp_1e3_obs_stats <- tibble(
  taxon_id = taxa_ids,
  statistic_obs = as.numeric(dacomp_stats_permapprox$obs_stats)
)

diagnostic_taxa <- volcano_label_taxa

perm_dist_df <- purrr::map_dfr(
  diagnostic_taxa,
  function(taxon_id_i) {
    taxon_idx <- match(taxon_id_i, taxa_ids)
    
    tibble(
      taxon_id = taxon_id_i,
      perm_stat = as.numeric(dacomp_stats_permapprox$perm_stats[, taxon_idx])
    )
  }
) %>%
  left_join(taxon_labels, by = "taxon_id")

obs_dist_df <- diff_abundance_results_long %>%
  filter(
    method == "permApprox",
    constraint == "support_at_obs",
    taxon_id %in% diagnostic_taxa
  ) %>%
  left_join(dacomp_1e7_pvalues, by = "taxon_id") %>%
  left_join(dacomp_1e3_obs_stats, by = "taxon_id") %>%
  transmute(
    taxon_id,
    taxon,
    statistic_obs,
    p_dacomp_1e7,
    p_permapprox_constrained = p_raw,
    label = paste0(
      "T_obs = ", formatC(statistic_obs, format = "f", digits = 2), "\n",
      "p_emp = ", formatC(p_dacomp_1e7, format = "e", digits = 2), "\n",
      "p_pApp = ",
      formatC(p_permapprox_constrained, format = "e", digits = 2)
    )
  )

diagnostic_taxon_order <- obs_dist_df %>%
  arrange(p_permapprox_constrained) %>%
  pull(taxon)

perm_dist_df <- perm_dist_df %>%
  mutate(taxon = factor(taxon, levels = diagnostic_taxon_order))

obs_dist_df <- obs_dist_df %>%
  mutate(taxon = factor(taxon, levels = diagnostic_taxon_order))

write_csv(
  perm_dist_df,
  file = path_target("diff_abundance_perm_distribution_data.csv")
)

write_csv(
  obs_dist_df,
  file = path_target("diff_abundance_perm_distribution_obs.csv")
)

obs_dist_df
```

```{r permutation_distribution_plot, fig.width=10.5, fig.height=5, dpi=200}
ggplot(perm_dist_df, aes(x = perm_stat)) +
  geom_blank(
    data = obs_dist_df,
    aes(x = statistic_obs, y = 0),
    inherit.aes = FALSE
  ) +
  geom_histogram(binwidth = 1, fill = "grey75", color = "white") +
  geom_hline(yintercept = 0, color = "grey85", linewidth = 0.4) +
  geom_vline(
    data = obs_dist_df,
    aes(xintercept = statistic_obs),
    color = "tomato",
    linewidth = 0.9
  ) +
  geom_text(
    data = obs_dist_df,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05,
    vjust = 1.25,
    size = 4,
    inherit.aes = FALSE
  ) +
  coord_cartesian(xlim = c(0, 50)) +
  facet_wrap(~ taxon, scales = "free_x", ncol = 4) +
  labs(
    x = "Permuted dacomp statistic",
    y = "Count"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

## Rarefied count diagnostics

dacomp internally rarefies the count table before computing the Wilcoxon-type
statistics. The following boxplots show the corresponding dacomp-rarefied
counts directly for the same taxa labelled in the volcano plot. Zeros are shown
as individual points, while the boxplots are computed from the non-zero counts
only to make the positive count distributions easier to compare.

```{r rarefied_count_boxplot_data}
rarefied_counts <- res_dacomp_permapprox$rarefied_counts
dimnames(rarefied_counts) <- dimnames(X_counts)

rarefied_taxa <- intersect(volcano_label_taxa, colnames(rarefied_counts))

rarefied_taxon_order <- diff_abundance_results_long %>%
  filter(
    method == "permApprox",
    constraint == "support_at_obs",
    taxon_id %in% rarefied_taxa
  ) %>%
  arrange(p_raw) %>%
  pull(taxon_id)

rarefied_boxplot_df <- rarefied_counts[, rarefied_taxa, drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(
    tibble(SampleID = rownames(X_counts), group = y_groups),
    by = "SampleID"
  ) %>%
  pivot_longer(
    cols = all_of(rarefied_taxa),
    names_to = "taxon_id",
    values_to = "rarefied_count"
  ) %>%
  left_join(taxon_labels, by = "taxon_id") %>%
  mutate(
    group = factor(group, levels = c("non-EBF", "EBF")),
    taxon_plot = stringr::str_trunc(taxon, width = 20, side = "right"),
    taxon = factor(
      taxon,
      levels = taxon_labels$taxon[match(rarefied_taxon_order, taxon_labels$taxon_id)]
    ),
    taxon_plot = factor(
      taxon_plot,
      levels = stringr::str_trunc(levels(taxon), width = 20, side = "right")
    )
  )

write_csv(
  rarefied_boxplot_df,
  file = path_target("diff_abundance_rarefied_boxplot_data.csv")
)

rarefied_boxplot_df
```

```{r rarefied_count_boxplots, fig.width=11, fig.height=5, dpi=200}
zero_count_plot_value <- 0.5

rarefied_boxplot_plot_df <- rarefied_boxplot_df %>%
  mutate(
    rarefied_count_plot = if_else(
      rarefied_count == 0,
      zero_count_plot_value,
      rarefied_count
    )
  )

rarefied_boxplot_nonzero_df <- rarefied_boxplot_plot_df %>%
  filter(rarefied_count > 0)

ggplot(rarefied_boxplot_plot_df, aes(x = group, y = rarefied_count_plot)) +
  geom_jitter(aes(color = group), width = 0.15, alpha = 0.5, size = 1.2) +
  geom_boxplot(
    data = rarefied_boxplot_nonzero_df,
    aes(fill = group),
    outlier.shape = NA,
    alpha = 0.75
  ) +
  facet_wrap(~ taxon_plot, scales = "free_y", ncol = 4) +
  scale_y_log10(
    breaks = c(zero_count_plot_value, 1, 10, 100),
    labels = c("0", "1", "10", "100")
  ) +
  scale_fill_manual(
    values = c("non-EBF" = "#1f77b4", "EBF" = "darkorange2"),
    name = ""
  ) +
  scale_color_manual(
    values = c("non-EBF" = "#1f77b4", "EBF" = "darkorange2"),
    name = ""
  ) +
  labs(
    x = NULL,
    y = "dacomp-rarefied count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 14),
    legend.position = "bottom",
    axis.text.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r cleanup_dacomp_permapprox_object, include=FALSE}
rm(res_dacomp_permapprox)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
