---
title: "Differential abundance analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "08_diff_abundance" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "8400abc1-adab-467f-99ae-0f1d4d512bda")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/DA_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(dacomp)
library(kableExtra)
library(furrr)
library(future)
library(progressr)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
permutation_seed <- 42
cores <- 10

q_DSFDR <- 0.1
dacomp_perm_grid <- c(1e3, 1e4, 1e5, 1e6, 1e7)
refit_dacomp_refs <- FALSE
refit_dacomp <- FALSE
write_dacomp_stats_files <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

Counts are transformed to relative abundances, zeros are replaced by
multiplicative replacement, and the CLR transformation is applied after
replacement. The dacomp test itself uses raw counts, as recommended by dacomp,
but the relative-abundance and CLR objects are stored to keep the preprocessing
consistent with the other analyses in this chapter. The shared preprocessing
helpers are defined in `functions.R`.

```{r helper_functions}
make_da_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

fmt_nperm <- function(n) {
  format(n, scientific = TRUE)
}

make_dacomp_result_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_res_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

make_dacomp_runtime_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_runtime_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

make_dacomp_stats_path <- function(n_perm, seed = permutation_seed) {
  path_target(
    paste0("dacomp_stats_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
  )
}

run_dacomp <- function(X_counts, y_groups, res_refs, n_perm, q = 0.1) {
  dacomp.test(
    X = X_counts,
    y = y_groups,
    ind_reference_taxa = res_refs,
    test = DACOMP.TEST.NAME.WILCOXON,
    nr_perm = n_perm,
    q = q,
    verbose = FALSE,
    return_rarefied_values = TRUE,
    Test_All = TRUE
  )
}

extract_dacomp_stats <- function(res_dacomp) {
  stats_mat <- res_dacomp$stats_matrix
  
  if (is.null(stats_mat)) {
    stop("The dacomp result does not contain `stats_matrix`.")
  }
  
  obs_stats <- stats_mat[1, ]
  perm_stats <- stats_mat[-1, , drop = FALSE]
  colnames(perm_stats) <- names(obs_stats)
  
  list(
    obs_stats = obs_stats,
    perm_stats = perm_stats
  )
}
```

## Prepare relative abundance and CLR matrices

```{r prepare_clr_matrix}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_da <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_da_metadata()

counts_raw <- counts_raw[sampdata_da$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_da$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

da_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

da_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("diff_abundance_multrepl_clr_object.rds")
)

write_csv(
  da_preprocessing_summary,
  file = path_target("diff_abundance_preprocessing_summary.csv")
)
```

## Differential abundance setup

The differential abundance analysis compares children with no exclusive
breastfeeding (`EBF duration == "0 months"`) to children exclusively breastfed
for at least two months (`EBF duration == ">=2 months"`). Samples in the sparse
`1 month` category and samples with missing EBF duration are excluded from this
two-group comparison.

```{r dacomp_setup}
dacomp_sample_data <- sampdata_da %>%
  mutate(
    EBF_group = case_when(
      `EBF duration` == "0 months" ~ "non-EBF",
      `EBF duration` == ">=2 months" ~ "EBF",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EBF_group)) %>%
  mutate(EBF_group = factor(EBF_group, levels = c("non-EBF", "EBF"))) %>%
  select(SampleID, `EBF duration`, EBF_group)

X_counts <- counts_raw[dacomp_sample_data$SampleID, , drop = FALSE]
X_rel <- counts_rel[dacomp_sample_data$SampleID, , drop = FALSE]
X_clr <- counts_clr[dacomp_sample_data$SampleID, , drop = FALSE]
y_groups <- dacomp_sample_data$EBF_group

stopifnot(identical(rownames(X_counts), dacomp_sample_data$SampleID))
stopifnot(identical(rownames(X_clr), dacomp_sample_data$SampleID))

dacomp_group_summary <- dacomp_sample_data %>%
  count(EBF_group, name = "n_samples")

dacomp_analysis_summary <- tibble(
  n_samples = nrow(X_counts),
  n_taxa = ncol(X_counts),
  n_non_ebf = sum(y_groups == "non-EBF"),
  n_ebf = sum(y_groups == "EBF"),
  excluded_ebf_one_month = sum(sampdata_da$`EBF duration` == "1 month", na.rm = TRUE),
  excluded_missing_ebf = sum(is.na(sampdata_da$`EBF duration`))
)

dacomp_group_summary
dacomp_analysis_summary
```

```{r save_dacomp_setup}
write_csv(
  dacomp_group_summary,
  file = path_target("diff_abundance_group_summary.csv")
)

write_csv(
  dacomp_analysis_summary,
  file = path_target("diff_abundance_analysis_summary.csv")
)

saveRDS(
  list(
    sample_data = dacomp_sample_data,
    X_counts = X_counts,
    X_rel = X_rel,
    X_clr = X_clr,
    y_groups = y_groups
  ),
  file = path_target("diff_abundance_analysis_objects.rds")
)
```

## dacomp reference selection

```{r dacomp_select_references}
path_dacomp_refs <- path_target("dacomp_reference_selection.rds")

if (!file.exists(path_dacomp_refs) || isTRUE(refit_dacomp_refs)) {
  set.seed(permutation_seed)
  
  res_refs <- dacomp.select_references(
    X = X_counts,
    minimal_TA = 50,
    verbose = TRUE
  )
  
  saveRDS(res_refs, file = path_dacomp_refs)
} else {
  res_refs <- readRDS(path_dacomp_refs)
}

dacomp_reference_summary <- tibble(
  n_taxa = ncol(X_counts),
  n_reference_taxa = length(res_refs$selected_references),
  n_test_taxa = ncol(X_counts) - length(res_refs$selected_references),
  minimal_TA = 50
)

dacomp_reference_summary
```

```{r save_dacomp_reference_summary}
write_csv(
  dacomp_reference_summary,
  file = path_target("dacomp_reference_summary.csv")
)
```

## dacomp permutation tests

The following chunk runs the long dacomp permutation analyses. Each permutation
count is cached in a separate result file and runtime file. If the result file
already exists, the corresponding run is skipped unless `refit_dacomp` is set to
`TRUE`.

```{r run_dacomp_tests}
dacomp_runtime_tbl <- purrr::map_dfr(
  dacomp_perm_grid,
  function(n_perm) {
    set.seed(permutation_seed)
    
    fname_res <- make_dacomp_result_path(n_perm, seed = permutation_seed)
    fname_rt <- make_dacomp_runtime_path(n_perm, seed = permutation_seed)
    
    if (!file.exists(fname_res) || isTRUE(refit_dacomp)) {
      cat(
        "Running dacomp:",
        "n_perm =", fmt_nperm(n_perm),
        "| seed =", permutation_seed,
        "\n"
      )
      
      runtime <- system.time({
        res_dacomp <- run_dacomp(
          X_counts = X_counts,
          y_groups = y_groups,
          res_refs = res_refs,
          n_perm = n_perm,
          q = q_DSFDR
        )
      })
      
      saveRDS(res_dacomp, file = fname_res)
      
      runtime_row <- tibble(
        n_perm = n_perm,
        elapsed_sec = unname(runtime["elapsed"]),
        seed = permutation_seed,
        result_file = basename(fname_res)
      )
      
      saveRDS(runtime_row, file = fname_rt)
    } else {
      cat(
        "Skipping cached dacomp run:",
        "n_perm =", fmt_nperm(n_perm),
        "| seed =", permutation_seed,
        "\n"
      )
      
      runtime_row <- if (file.exists(fname_rt)) {
        readRDS(fname_rt)
      } else {
        tibble(
          n_perm = n_perm,
          elapsed_sec = NA_real_,
          seed = permutation_seed,
          result_file = basename(fname_res)
        )
      }
    }
    
    runtime_row
  }
) %>%
  arrange(n_perm)

dacomp_runtime_tbl
```

```{r save_dacomp_runtime}
write_csv(
  dacomp_runtime_tbl,
  file = path_target("dacomp_runtime.csv")
)
```

## Runtime comparison

This section prepares the runtime comparison table and figure. At present, it
contains the dacomp-only rows. The dacomp result files also retain the
permutation statistics needed to add permApprox runtimes later.

```{r save_dacomp_stats_for_permapprox}
dacomp_stats_index_tbl <- purrr::map_dfr(
  dacomp_runtime_tbl$n_perm,
  function(n_perm) {
    fname_res <- make_dacomp_result_path(n_perm, seed = permutation_seed)
    fname_stats <- make_dacomp_stats_path(n_perm, seed = permutation_seed)
    
    if (
      isTRUE(write_dacomp_stats_files) &&
        file.exists(fname_res) &&
        !file.exists(fname_stats)
    ) {
      res_dacomp <- readRDS(fname_res)
      dacomp_stats <- extract_dacomp_stats(res_dacomp)
      saveRDS(dacomp_stats, file = fname_stats)
    }
    
    tibble(
      n_perm = n_perm,
      seed = permutation_seed,
      result_file = basename(fname_res),
      stats_file = basename(fname_stats),
      result_exists = file.exists(fname_res),
      stats_exists = file.exists(fname_stats)
    )
  }
)

dacomp_stats_index_tbl
```

```{r save_dacomp_stats_index}
write_csv(
  dacomp_stats_index_tbl,
  file = path_target("dacomp_stats_index.csv")
)
```

```{r runtime_comparison_table}
runtime_total_tbl <- dacomp_runtime_tbl %>%
  transmute(
    line = "dacomp only",
    method = "dacomp",
    setting = "dacomp",
    n_perm = n_perm,
    elapsed_s_total = elapsed_sec,
    seed = seed,
    result_file = result_file
  ) %>%
  mutate(
    elapsed_s = round(elapsed_s_total, 2),
    elapsed_min = round(elapsed_s_total / 60, 2),
    elapsed_h = round(elapsed_s_total / 3600, 3)
  ) %>%
  arrange(line, n_perm)

runtime_total_tbl
```

```{r save_runtime_comparison_table}
saveRDS(
  runtime_total_tbl,
  file = path_target("runtime_total_tbl.rds")
)

write_csv(
  runtime_total_tbl,
  file = path_target("runtime_total_tbl.csv")
)
```

```{r runtime_comparison_display}
runtime_total_tbl %>%
  select(line, n_perm, elapsed_s, elapsed_min, elapsed_h) %>%
  knitr::kable()
```

```{r runtime_plot_data}
runtime_plot_tbl <- runtime_total_tbl %>%
  filter(!is.na(elapsed_s_total), !is.na(n_perm)) %>%
  mutate(
    line = factor(
      line,
      levels = c(
        "dacomp only",
        "dacomp + permApprox (SLLS, k0=250)",
        "dacomp + permApprox (SLLS, k0=0.25B)"
      )
    )
  )

runtime_plot_tbl
```

```{r runtime_plot, fig.width=7.5, fig.height=3.6, dpi=300}
x_grid <- sort(unique(runtime_plot_tbl$n_perm))
minute_grid <- c(1, 2, 5, 10, 30, 60, 120, 300)
second_grid <- c(1, 3, 10, minute_grid * 60)

ggplot(runtime_plot_tbl, aes(
  x = n_perm,
  y = elapsed_s_total,
  colour = line,
  group = line
)) +
  geom_hline(yintercept = second_grid, color = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = x_grid, color = "grey90", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.2, stroke = 0.9) +
  scale_x_log10(
    breaks = x_grid,
    labels = scales::label_scientific()
  ) +
  scale_y_log10(
    name = "Runtime (seconds)",
    breaks = second_grid,
    labels = scales::label_number(),
    sec.axis = sec_axis(
      trans = ~ . / 60,
      name = "Runtime (minutes)",
      breaks = minute_grid,
      labels = scales::label_number()
    )
  ) +
  scale_colour_manual(
    values = c(
      "dacomp only" = "black",
      "dacomp + permApprox (SLLS, k0=250)" = "darkorange2",
      "dacomp + permApprox (SLLS, k0=0.25B)" = "#1f77b4"
    ),
    drop = FALSE
  ) +
  labs(x = "Number of permutations", colour = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = c(1, 0.1),
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(
      fill = "white",
      color = "grey80",
      linewidth = 0.3
    ),
    legend.key = element_rect(fill = "white")
  )
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
