---
title: "Differential distribution analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "09_diff_distribution" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "8ecbd0ba-b360-460e-aba4-302f35d02cda")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/DD_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(arm)
library(eva)
library(BiocParallel)
library(SingleCellExperiment)
library(permApprox)
library(kableExtra)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
n_perm <- 999
permutation_seed <- 42
cores <- 10
alpha_diff_distribution <- 0.05

refit_waddr <- FALSE
refit_permapprox <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

Counts are transformed to relative abundances, zeros are replaced by
multiplicative replacement, and the CLR transformation is applied after
replacement. For the adapted two-stage differential distribution analysis, the
original count matrix defines whether a taxon is zero in a sample, while the
non-zero Wasserstein tests are computed on the corresponding CLR values.

```{r helper_functions}
make_dd_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

sci <- function(x, digits = 2) {
  formatC(x, format = "e", digits = digits)
}

safe_fisher_combine <- function(p1, p2) {
  p1[!is.na(p1) & p1 <= 0] <- .Machine$double.xmin
  p2[!is.na(p2) & p2 <= 0] <- .Machine$double.xmin
  
  purrr::map2_dbl(p1, p2, function(a, b) {
    x <- c(a, b)
    
    if (sum(is.na(x)) == 0) {
      stats::pchisq(-2 * sum(log(x)), df = 2 * length(x), lower.tail = FALSE)
    } else if (sum(is.na(x)) == 1) {
      x[!is.na(x)]
    } else {
      NA_real_
    }
  })
}

taxon_labels_from_phyloseq <- function(phylo) {
  tax_tab <- tax_table(phylo) %>%
    as("matrix") %>%
    as_tibble(rownames = "taxon_id")
  
  label_col <- dplyr::case_when(
    "Genus" %in% colnames(tax_tab) ~ "Genus",
    "genus" %in% colnames(tax_tab) ~ "genus",
    TRUE ~ NA_character_
  )
  
  if (is.na(label_col)) {
    labels <- tax_tab$taxon_id
  } else {
    labels <- tax_tab[[label_col]]
    labels <- ifelse(is.na(labels) | labels == "", tax_tab$taxon_id, labels)
  }
  
  setNames(make.unique(labels), tax_tab$taxon_id)
}

normalise_wass_result <- function(res, n_perm) {
  if (length(res$wass.values) == n_perm) {
    return(res)
  }
  
  res$wass.values <- rep(NA_real_, n_perm)
  res
}

run_adapted_waddr <- function(count_mat, clr_mat, groups, n_perm, seed) {
  taxon_ids <- colnames(clr_mat)
  group_levels <- levels(groups)
  
  if (length(group_levels) != 2) {
    stop("The adapted waddR analysis requires exactly two groups.")
  }
  
  counts_taxa_by_samples <- t(count_mat)
  p_zero <- testZeroes(counts_taxa_by_samples, as.character(groups))
  p_zero <- setNames(as.numeric(p_zero), taxon_ids)
  
  wass_res_list <- purrr::imap(
    taxon_ids,
    function(taxon_id, taxon_idx) {
      if (taxon_idx %% 100 == 0) {
        message("Running Wasserstein test for taxon ", taxon_idx, " / ", length(taxon_ids))
      }
      
      keep_nonzero <- count_mat[, taxon_id] > 0
      x1 <- clr_mat[groups == group_levels[1] & keep_nonzero, taxon_id]
      x2 <- clr_mat[groups == group_levels[2] & keep_nonzero, taxon_id]
      
      set.seed(seed + taxon_idx)
      res <- suppressWarnings(.wassersteinTestSp(x1, x2, permnum = n_perm))
      normalise_wass_result(res, n_perm = n_perm)
    }
  )
  
  names(wass_res_list) <- taxon_ids
  
  wass_stats <- do.call(
    rbind,
    lapply(wass_res_list, function(res) {
      as.numeric(res[names(res) != "wass.values"])
    })
  )
  
  colnames(wass_stats) <- names(wass_res_list[[1]])[
    names(wass_res_list[[1]]) != "wass.values"
  ]
  rownames(wass_stats) <- taxon_ids
  
  perm_stats <- do.call(
    cbind,
    lapply(wass_res_list, function(res) as.numeric(res$wass.values))
  )
  colnames(perm_stats) <- taxon_ids
  
  results <- as_tibble(wass_stats, rownames = "taxon_id") %>%
    mutate(
      p.zero = base::unname(p_zero[taxon_id]),
      p.combined = safe_fisher_combine(p.zero, pval),
      p.adj.nonzero = p.adjust(pval, method = "BH"),
      p.adj.zero = p.adjust(p.zero, method = "BH"),
      p.adj.combined = p.adjust(p.combined, method = "BH"),
      n_nonzero_group1 = colSums(count_mat[groups == group_levels[1], , drop = FALSE] > 0),
      n_nonzero_group2 = colSums(count_mat[groups == group_levels[2], , drop = FALSE] > 0),
      n_group1 = sum(groups == group_levels[1]),
      n_group2 = sum(groups == group_levels[2])
    )
  
  list(
    results = results,
    perm_stats = perm_stats,
    group_levels = group_levels
  )
}

make_permapprox_results <- function(obs_stats, perm_stats, sample_size, cores) {
  valid <- !is.na(obs_stats) & colSums(!is.na(perm_stats)) > 0
  
  if (!any(valid)) {
    stop("No valid Wasserstein statistics available for permApprox.")
  }
  
  gpd_ctrl_unconstr <- permApprox::make_gpd_ctrl(
    constraint = "unconstrained",
    sample_size = sample_size
  )
  
  gpd_ctrl_constr <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = sample_size
  )
  
  expand_permapprox_result <- function(res, valid, n_total) {
    n_valid <- sum(valid)
    
    for (nm in names(res)) {
      x <- res[[nm]]
      
      if (is.atomic(x) && length(x) == n_valid) {
        out <- rep(NA, n_total)
        out[valid] <- x
        res[[nm]] <- out
      } else if (is.data.frame(x) && nrow(x) == n_valid) {
        out <- x[rep(1, n_total), , drop = FALSE]
        out[,] <- NA
        out[valid, ] <- x
        rownames(out) <- NULL
        res[[nm]] <- out
      } else if (is.matrix(x) && nrow(x) == n_valid) {
        out <- matrix(
          NA,
          nrow = n_total,
          ncol = ncol(x),
          dimnames = list(NULL, colnames(x))
        )
        out[valid, ] <- x
        res[[nm]] <- out
      }
    }
    
    res$valid_test <- valid
    res
  }
  
  unconstrained <- permApprox::perm_approx(
      obs_stats = obs_stats[valid],
      perm_stats = perm_stats[, valid, drop = FALSE],
      method = "gpd",
      alternative = "greater",
      null_center = "mean",
      approx_thresh = 1,
      adjust_method = "none",
      gpd_ctrl = gpd_ctrl_unconstr,
      cores = cores,
      verbose = TRUE
  )
  
  constrained <- permApprox::perm_approx(
      obs_stats = obs_stats[valid],
      perm_stats = perm_stats[, valid, drop = FALSE],
      method = "gpd",
      alternative = "greater",
      null_center = "mean",
      approx_thresh = 1,
      adjust_method = "none",
      gpd_ctrl = gpd_ctrl_constr,
      cores = cores,
      verbose = TRUE
  )
  
  list(
    unconstrained = expand_permapprox_result(
      unconstrained,
      valid = valid,
      n_total = length(obs_stats)
    ),
    constrained = expand_permapprox_result(
      constrained,
      valid = valid,
      n_total = length(obs_stats)
    )
  )
}

extract_gpd_column <- function(x, column) {
  if (!is.null(x$gpd_fit) && column %in% colnames(x$gpd_fit)) {
    return(x$gpd_fit[[column]])
  }
  
  rep(NA_real_, length(x$p_unadjusted))
}
```

## Load modified waddR implementation

The adapted analysis uses the low-level Wasserstein permutation test from the
modified local `waddR` implementation. This version exports the permutation
statistics required by `permApprox`.

```{r source_waddr_functions, message=FALSE, warning=FALSE}
waddR_lib <- system.file("libs", .Platform$r_arch, "waddR.dll", package = "waddR")

if (waddR_lib == "") {
  waddR_lib <- system.file("libs", "waddR.dll", package = "waddR")
}

if (waddR_lib == "") {
  stop("Could not find the installed waddR DLL required by the local R functions.")
}

if (!"waddR" %in% names(getLoadedDLLs())) {
  dyn.load(waddR_lib)
}

thesis_path <- dirname(application_path)
waddR_path <- file.path(thesis_path, "ch06_permapprox", "functions", "waddR", "R")
waddR_functions <- list.files(waddR_path, pattern = "\\.R$", full.names = TRUE)

if (length(waddR_functions) == 0) {
  stop("No local waddR R files found at: ", waddR_path)
}

invisible(sapply(waddR_functions, source))
```

## Prepare relative abundance and CLR matrices

```{r prepare_clr_matrix}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_dd <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_dd_metadata()

counts_raw <- counts_raw[sampdata_dd$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_dd$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

dd_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

dd_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("diff_distribution_multrepl_clr_object.rds")
)

write_csv(
  dd_preprocessing_summary,
  file = path_target("diff_distribution_preprocessing_summary.csv")
)
```

## Differential distribution setup

The differential distribution analysis compares children with no exclusive
breastfeeding (`EBF duration == "0 months"`) to children exclusively breastfed
for at least two months (`EBF duration == ">=2 months"`). Samples in the sparse
`1 month` category and samples with missing EBF duration are excluded from this
two-group comparison.

```{r diff_distribution_setup}
dd_sample_data <- sampdata_dd %>%
  mutate(
    EBF_group = case_when(
      `EBF duration` == "0 months" ~ "non-EBF",
      `EBF duration` == ">=2 months" ~ "EBF",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EBF_group)) %>%
  mutate(EBF_group = factor(EBF_group, levels = c("non-EBF", "EBF"))) %>%
  select(SampleID, `EBF duration`, EBF_group)

X_counts <- counts_raw[dd_sample_data$SampleID, , drop = FALSE]
X_rel <- counts_rel[dd_sample_data$SampleID, , drop = FALSE]
X_clr <- counts_clr[dd_sample_data$SampleID, , drop = FALSE]
y_groups <- dd_sample_data$EBF_group
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)
dd_group_palette <- c("non-EBF" = "#1F78B4", "EBF" = "#E66101")

stopifnot(identical(rownames(X_counts), dd_sample_data$SampleID))
stopifnot(identical(rownames(X_clr), dd_sample_data$SampleID))

dd_group_summary <- dd_sample_data %>%
  count(EBF_group, name = "n_samples")

dd_analysis_summary <- tibble(
  n_samples = nrow(X_counts),
  n_taxa = ncol(X_counts),
  n_non_ebf = sum(y_groups == "non-EBF"),
  n_ebf = sum(y_groups == "EBF"),
  excluded_ebf_one_month = sum(sampdata_dd$`EBF duration` == "1 month", na.rm = TRUE),
  excluded_missing_ebf = sum(is.na(sampdata_dd$`EBF duration`)),
  n_perm = n_perm,
  seed = permutation_seed
)

dd_group_summary
dd_analysis_summary
```

```{r save_diff_distribution_setup}
write_csv(
  dd_group_summary,
  file = path_target("diff_distribution_group_summary.csv")
)

write_csv(
  dd_analysis_summary,
  file = path_target("diff_distribution_analysis_summary.csv")
)

saveRDS(
  list(
    sample_data = dd_sample_data,
    X_counts = X_counts,
    X_rel = X_rel,
    X_clr = X_clr,
    y_groups = y_groups
  ),
  file = path_target("diff_distribution_analysis_objects.rds")
)
```

## Adapted two-stage waddR analysis

The zero stage is computed from the original counts. The non-zero Wasserstein
stage uses CLR values from samples where the original count of the tested taxon
is positive.

```{r run_adapted_waddr}
path_waddr <- path_target(
  paste0("waddr_adapted_ts_nperm=", n_perm, "_seed=", permutation_seed, ".rds")
)

if (!file.exists(path_waddr) || isTRUE(refit_waddr)) {
  register(SerialParam())
  
  runtime_waddr <- system.time({
    waddr_adapted <- run_adapted_waddr(
      count_mat = X_counts,
      clr_mat = X_clr,
      groups = y_groups,
      n_perm = n_perm,
      seed = permutation_seed
    )
  })
  
  waddr_adapted$runtime <- runtime_waddr
  saveRDS(waddr_adapted, file = path_waddr)
} else {
  waddr_adapted <- readRDS(path_waddr)
}

waddr_results <- waddr_adapted$results
waddr_perm_stats <- waddr_adapted$perm_stats

waddr_results %>%
  arrange(pval) %>%
  head(n = 10)
```

```{r waddr_summary}
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)

waddr_summary <- waddr_results %>%
  summarise(
    n_taxa = n(),
    n_nonzero_testable = sum(!is.na(pval)),
    n_zero_p_waddr = sum(pval == 0, na.rm = TRUE),
    n_zero_stage_testable = sum(!is.na(p.zero)),
    n_zero_p_combined = sum(p.combined == 0, na.rm = TRUE),
    n_significant_nonzero_raw = sum(pval < alpha_diff_distribution, na.rm = TRUE),
    n_significant_nonzero_bh = sum(p.adj.nonzero < alpha_diff_distribution, na.rm = TRUE),
    n_significant_combined_raw = sum(p.combined < alpha_diff_distribution, na.rm = TRUE),
    n_significant_combined_bh = sum(p.adj.combined < alpha_diff_distribution, na.rm = TRUE)
  )

waddr_summary
```

```{r save_waddr_results}
write_csv(
  waddr_summary,
  file = path_target("waddr_adapted_summary.csv")
)

write_csv(
  waddr_results %>% mutate(taxon = base::unname(taxon_labels[taxon_id])),
  file = path_target("waddr_adapted_results.csv")
)
```

## p-value refinement with permApprox

```{r run_permapprox}
path_permapprox <- path_target(
  paste0("permapprox_waddr_adapted_nperm=", n_perm, "_seed=", permutation_seed, ".rds")
)

obs_stats <- waddr_results$`d.wass^2`
names(obs_stats) <- waddr_results$taxon_id

if (!file.exists(path_permapprox) || isTRUE(refit_permapprox)) {
  dd_permapprox <- make_permapprox_results(
    obs_stats = obs_stats,
    perm_stats = waddr_perm_stats,
    sample_size = min(dd_group_summary$n_samples),
    cores = cores
  )
  
  saveRDS(dd_permapprox, file = path_permapprox)
} else {
  dd_permapprox <- readRDS(path_permapprox)
}

summary(dd_permapprox$constrained)
```

```{r combine_results}
dd_results_all <- waddr_results %>%
  mutate(
    taxon = base::unname(taxon_labels[taxon_id]),
    p_permapprox_unconstrained = dd_permapprox$unconstrained$p_unadjusted,
    p_permapprox_constrained = dd_permapprox$constrained$p_unadjusted,
    p_combined_permapprox_unconstrained = safe_fisher_combine(
      p.zero,
      p_permapprox_unconstrained
    ),
    p_combined_permapprox_constrained = safe_fisher_combine(
      p.zero,
      p_permapprox_constrained
    ),
    p_adj_permapprox_unconstrained = p.adjust(
      p_permapprox_unconstrained,
      method = "BH"
    ),
    p_adj_permapprox_constrained = p.adjust(
      p_permapprox_constrained,
      method = "BH"
    ),
    p_adj_combined_permapprox_unconstrained = p.adjust(
      p_combined_permapprox_unconstrained,
      method = "BH"
    ),
    p_adj_combined_permapprox_constrained = p.adjust(
      p_combined_permapprox_constrained,
      method = "BH"
    ),
    method_used_unconstrained = dd_permapprox$unconstrained$method_used,
    method_used_constrained = dd_permapprox$constrained$method_used,
    shape_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox$unconstrained,
      "shape"
    ),
    shape_permapprox_constrained = extract_gpd_column(
      dd_permapprox$constrained,
      "shape"
    ),
    scale_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox$unconstrained,
      "scale"
    ),
    scale_permapprox_constrained = extract_gpd_column(
      dd_permapprox$constrained,
      "scale"
    )
  )

dd_permapprox_summary <- dd_results_all %>%
  summarise(
    n_taxa = n(),
    n_testable_waddr = sum(!is.na(pval)),
    n_testable_permapprox_unconstrained = sum(!is.na(p_permapprox_unconstrained)),
    n_testable_permapprox_constrained = sum(!is.na(p_permapprox_constrained)),
    n_zero_waddr = sum(pval == 0, na.rm = TRUE),
    n_zero_permapprox_unconstrained = sum(p_permapprox_unconstrained == 0, na.rm = TRUE),
    n_zero_permapprox_constrained = sum(p_permapprox_constrained == 0, na.rm = TRUE),
    n_sig_waddr_bh = sum(p.adj.nonzero < alpha_diff_distribution, na.rm = TRUE),
    n_sig_permapprox_unconstrained_bh = sum(
      p_adj_permapprox_unconstrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_permapprox_constrained_bh = sum(
      p_adj_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_combined_waddr_bh = sum(p.adj.combined < alpha_diff_distribution, na.rm = TRUE),
    n_sig_combined_permapprox_unconstrained_bh = sum(
      p_adj_combined_permapprox_unconstrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_combined_permapprox_constrained_bh = sum(
      p_adj_combined_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    )
  )

dd_permapprox_summary
```

```{r save_permapprox_results}
saveRDS(
  dd_results_all,
  file = path_target("diff_distribution_results_all.rds")
)

write_csv(
  dd_results_all,
  file = path_target("diff_distribution_results_all.csv")
)

write_csv(
  dd_permapprox_summary,
  file = path_target("diff_distribution_permapprox_summary.csv")
)
```

## P-values from waddR and permApprox

```{r pvalue_plot_data}
plot_methods <- tibble(
  method_key = c(
    "waddR",
    "permApprox_unconstrained",
    "permApprox_constrained"
  ),
  method_label = c(
    "waddR",
    "waddR + permApprox(U)",
    "waddR + permApprox(C)"
  ),
  p_col = c(
    "pval",
    "p_permapprox_unconstrained",
    "p_permapprox_constrained"
  )
)

pvalue_plot_df <- purrr::map_dfr(
  seq_len(nrow(plot_methods)),
  function(i) {
    p_col <- plot_methods$p_col[i]
    
    dd_results_all %>%
      transmute(
        taxon_id,
        taxon,
        method_key = plot_methods$method_key[i],
        method_label = plot_methods$method_label[i],
        d_wass2 = `d.wass^2`,
        pvalue = .data[[p_col]],
        p_waddr = pval,
        p_permapprox_unconstrained,
        p_permapprox_constrained
      )
  }
) %>%
  mutate(
    zero_class = case_when(
      is.na(p_waddr) ~ "na_waddR",
      p_waddr > 0 ~ "nonzero_waddR",
      p_waddr == 0 & p_permapprox_unconstrained == 0 ~ "zero_both_unconstrained",
      p_waddr == 0 ~ "zero_waddR_only",
      TRUE ~ NA_character_
    ),
    zero_class = factor(
      zero_class,
      levels = c(
        "nonzero_waddR",
        "zero_waddR_only",
        "zero_both_unconstrained",
        "na_waddR"
      )
    ),
    method_key = factor(method_key, levels = plot_methods$method_key),
    method_label = factor(method_label, levels = plot_methods$method_label)
  )

facet_labels <- pvalue_plot_df %>%
  group_by(method_label) %>%
  summarise(
    n_zero = sum(pvalue == 0, na.rm = TRUE),
    n_non_na = sum(!is.na(pvalue)),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(method_label, "\n(zeros: ", n_zero, " / ", n_non_na, ")")
  )

facet_labeller <- setNames(facet_labels$label, facet_labels$method_label)

pvalue_plot_df
```

```{r pvalues_waddr_permapprox, fig.width=10, fig.height=4.5, dpi=250}
zero_repl <- 1e-318
log_breaks <- c(1e-3, 1e-25, 1e-50, 1e-75, 1e-100, 1e-200, 1e-300, zero_repl)

pvalue_plot_df %>%
  mutate(
    pvalue_plot = case_when(
      is.na(pvalue) ~ NA_real_,
      pvalue == 0 ~ zero_repl,
      pvalue < zero_repl ~ NA_real_,
      TRUE ~ pvalue
    )
  ) %>%
  ggplot(aes(x = d_wass2, y = pvalue_plot)) +
  geom_point(
    data = function(x) filter(x, zero_class == "nonzero_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2.4,
    alpha = 0.85,
    stroke = 1
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_waddR_only"),
    aes(color = zero_class, shape = zero_class),
    size = 2.4,
    alpha = 0.9,
    stroke = 0.4
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_both_unconstrained"),
    aes(color = zero_class, shape = zero_class),
    size = 2,
    alpha = 0.9,
    stroke = 0.3
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "na_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2,
    alpha = 0.9,
    stroke = 0.6
  ) +
  geom_hline(yintercept = 1 / (n_perm + 1), linetype = "dashed", color = "black") +
  facet_wrap(
    ~ method_label,
    nrow = 1,
    labeller = as_labeller(facet_labeller)
  ) +
  scale_y_log10(
    breaks = log_breaks,
    labels = function(x) {
      out <- scales::label_scientific()(x)
      out[x <= zero_repl * 10] <- "0"
      out
    }
  ) +
  coord_cartesian(ylim = c(zero_repl / 2, 1)) +
  scale_color_manual(
    values = c(
      "nonzero_waddR" = "grey40",
      "zero_waddR_only" = "#1f77b4",
      "zero_both_unconstrained" = "darkorange2",
      "na_waddR" = "grey20"
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  scale_shape_manual(
    values = c(
      "nonzero_waddR" = 1,
      "zero_waddR_only" = 17,
      "zero_both_unconstrained" = 16,
      "na_waddR" = 7
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  labs(
    title = "P-values (waddR vs. permApprox)",
    subtitle = paste0("B = ", n_perm),
    x = expression(d[W]^2),
    y = expression(p),
    color = NULL,
    shape = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.spacing = unit(1.6, "lines"),
    strip.text = element_text(size = 11),
    legend.position = c(0.23, 0.32),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )
```

## Distribution diagnostics for selected taxa

These plots focus on taxa for which the empirical waddR p-value is zero. If no
such taxa occur, the same diagnostic plots are shown for the taxa with the
smallest constrained permApprox p-values.

```{r selected_taxa_distribution_data}
selected_taxa_df <- dd_results_all %>%
  filter(!is.na(`d.wass^2`), !is.na(p_permapprox_constrained)) %>%
  mutate(is_zero_waddr = !is.na(pval) & pval == 0) %>%
  arrange(desc(is_zero_waddr), p_permapprox_constrained, `d.wass^2`) %>%
  slice_head(n = 6) %>%
  mutate(
    taxon_label = taxon,
    taxon_label = if_else(
      is.na(taxon_label) | taxon_label == "",
      taxon_id,
      taxon_label
    ),
    panel_label = paste0(
      stringr::str_trunc(taxon_label, width = 22),
      "\n(d = ", signif(`d.wass^2`, 3),
      "; pApp(C) = ", sci(p_permapprox_constrained), ")"
    )
  )

selected_taxa_ids <- selected_taxa_df$taxon_id
selected_panel_labels <- setNames(selected_taxa_df$panel_label, selected_taxa_ids)
selected_plot_title_suffix <- if (any(selected_taxa_df$is_zero_waddr)) {
  "waddR p = 0"
} else {
  "smallest constrained permApprox p-values"
}

selected_clr_df <- as_tibble(X_clr[, selected_taxa_ids, drop = FALSE], rownames = "SampleID") %>%
  left_join(dd_sample_data %>% select(SampleID, EBF_group), by = "SampleID") %>%
  pivot_longer(
    cols = all_of(selected_taxa_ids),
    names_to = "taxon_id",
    values_to = "clr_abundance"
  ) %>%
  mutate(
    taxon_id = factor(taxon_id, levels = selected_taxa_ids)
  )

if (is.null(colnames(waddr_perm_stats))) {
  colnames(waddr_perm_stats) <- waddr_results$taxon_id
}

selected_perm_df <- waddr_perm_stats[, selected_taxa_ids, drop = FALSE] %>%
  as_tibble() %>%
  mutate(perm_id = row_number(), .before = 1) %>%
  pivot_longer(
    cols = all_of(selected_taxa_ids),
    names_to = "taxon_id",
    values_to = "perm_value"
  ) %>%
  left_join(
    selected_taxa_df %>%
      select(taxon_id, d_wass2 = `d.wass^2`),
    by = "taxon_id"
  ) %>%
  mutate(
    taxon_id = factor(taxon_id, levels = selected_taxa_ids)
  )

selected_taxa_df
```

### Densities

```{r selected_taxa_density, fig.width=10, fig.height=5, dpi=250}
ggplot(selected_clr_df, aes(x = clr_abundance, fill = EBF_group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = dd_group_palette) +
  facet_wrap(
    ~ taxon_id,
    scales = "free",
    labeller = as_labeller(selected_panel_labels)
  ) +
  labs(
    title = paste("CLR-transformed abundance densities (", selected_plot_title_suffix, ")", sep = ""),
    x = "CLR-transformed abundance",
    y = "Density",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines")
  )
```

### Histograms

```{r selected_taxa_hist, fig.width=10, fig.height=5, dpi=250}
ggplot(selected_clr_df, aes(x = clr_abundance, fill = EBF_group)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = dd_group_palette) +
  facet_wrap(
    ~ taxon_id,
    scales = "free",
    labeller = as_labeller(selected_panel_labels)
  ) +
  labs(
    title = paste("CLR-transformed abundance histograms (", selected_plot_title_suffix, ")", sep = ""),
    x = "CLR-transformed abundance",
    y = "Count",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines")
  )
```

### Permutation distributions

```{r selected_taxa_perm_dist, fig.width=10, fig.height=5, dpi=250}
ggplot(selected_perm_df, aes(x = perm_value)) +
  geom_histogram(bins = 30, fill = "grey70", color = "white") +
  geom_vline(aes(xintercept = d_wass2), color = "#D55E00", linewidth = 0.8) +
  facet_wrap(
    ~ taxon_id,
    scales = "free",
    labeller = as_labeller(selected_panel_labels)
  ) +
  labs(
    title = paste("Permutation distributions (", selected_plot_title_suffix, ")", sep = ""),
    x = expression(d[W]^2),
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines")
  )
```

## Top taxa

```{r table_top_taxa, results="asis"}
n_top <- 20

top_taxa_tbl <- dd_results_all %>%
  filter(!is.na(p_permapprox_constrained)) %>%
  arrange(p_permapprox_constrained) %>%
  slice_head(n = n_top) %>%
  transmute(
    Taxon = taxon,
    `$d_W^2$` = round(`d.wass^2`, 3),
    `$p_{0}$` = sci(p.zero),
    `$p_{+}$ (waddR)` = sci(pval),
    `$p_{+}$ (permApprox C)` = sci(p_permapprox_constrained),
    `$p_{comb}$ (permApprox C)` = sci(p_combined_permapprox_constrained),
    `$q_{comb}$ (permApprox C)` = sci(p_adj_combined_permapprox_constrained)
  )

top_taxa_table <- kbl(
  top_taxa_tbl,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste0(
    "Top ", n_top,
    " taxa in the adapted two-stage differential distribution analysis, ",
    "sorted by constrained \\texttt{permApprox} p-values for the non-zero Wasserstein stage."
  ),
  label = "application:diff-distribution-top-taxa"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 8)

top_taxa_table

save_kable(
  top_taxa_table,
  path_target("diff_distribution_top_taxa_table.tex")
)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
