---
title: "Differential distribution analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "09_diff_distribution" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "8ecbd0ba-b360-460e-aba4-302f35d02cda")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/DD_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(tidyverse)
library(zCompositions)
library(arm)
library(eva)
library(BiocParallel)
library(SingleCellExperiment)
library(permApprox)
library(kableExtra)
library(ggrepel)
library(patchwork)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Set global parameters

```{r}
zero_replacement_fraction <- 0.65
n_perm <- 999
permutation_seed <- 42
cores <- 10
alpha_diff_distribution <- 0.01

refit_waddr <- FALSE
refit_permapprox <- FALSE
refit_waddr_complete_clr <- FALSE
refit_permapprox_complete_clr <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

## Helper functions

Counts are transformed to relative abundances, zeros are replaced by
multiplicative replacement, and the CLR transformation is applied after
replacement. For the adapted two-stage differential distribution analysis, the
original count matrix defines whether a taxon is zero in a sample, while the
non-zero Wasserstein tests are computed on the corresponding CLR values.

```{r helper_functions}
make_dd_metadata <- function(sampdata) {
  sampdata %>%
    mutate(
      Country = factor(Country),
      Sex = factor(Sex),
      `C-section` = factor(
        as.character(`C-section`),
        levels = c("Yes", "No"),
        labels = c("Cesarean", "Vaginal")
      ),
      `BF duration` = factor(
        as.character(`BF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      `EBF duration` = factor(
        as.character(`EBF duration`),
        levels = c("0", "1", ">=2"),
        labels = c("0 months", "1 month", ">=2 months")
      ),
      Smoking = factor(Smoking),
      Siblings = factor(
        dplyr::case_when(
          is.na(Siblings) ~ NA_character_,
          as.character(Siblings) == "0" ~ "0",
          TRUE ~ ">=1"
        ),
        levels = c("0", ">=1")
      )
    )
}

sci <- function(x, digits = 2) {
  formatC(x, format = "e", digits = digits)
}

safe_fisher_combine <- function(p1, p2) {
  p1[!is.na(p1) & p1 <= 0] <- .Machine$double.xmin
  p2[!is.na(p2) & p2 <= 0] <- .Machine$double.xmin
  
  purrr::map2_dbl(p1, p2, function(a, b) {
    x <- c(a, b)
    
    if (sum(is.na(x)) == 0) {
      stats::pchisq(-2 * sum(log(x)), df = 2 * length(x), lower.tail = FALSE)
    } else if (sum(is.na(x)) == 1) {
      x[!is.na(x)]
    } else {
      NA_real_
    }
  })
}

taxon_labels_from_phyloseq <- function(phylo) {
  tax_tab <- tax_table(phylo) %>%
    as("matrix") %>%
    as_tibble(rownames = "taxon_id")
  
  label_col <- dplyr::case_when(
    "Genus" %in% colnames(tax_tab) ~ "Genus",
    "genus" %in% colnames(tax_tab) ~ "genus",
    TRUE ~ NA_character_
  )
  
  if (is.na(label_col)) {
    labels <- tax_tab$taxon_id
  } else {
    labels <- tax_tab[[label_col]]
    labels <- ifelse(is.na(labels) | labels == "", tax_tab$taxon_id, labels)
  }
  
  setNames(make.unique(labels), tax_tab$taxon_id)
}

normalise_wass_result <- function(res, n_perm) {
  if (length(res$wass.values) == n_perm) {
    return(res)
  }
  
  res$wass.values <- rep(NA_real_, n_perm)
  res
}

run_adapted_waddr <- function(count_mat, clr_mat, groups, n_perm, seed) {
  taxon_ids <- colnames(clr_mat)
  group_levels <- levels(groups)
  
  if (length(group_levels) != 2) {
    stop("The adapted waddR analysis requires exactly two groups.")
  }
  
  counts_taxa_by_samples <- t(count_mat)
  p_zero <- testZeroes(counts_taxa_by_samples, as.character(groups))
  p_zero <- setNames(as.numeric(p_zero), taxon_ids)
  
  wass_res_list <- purrr::imap(
    taxon_ids,
    function(taxon_id, taxon_idx) {
      if (taxon_idx %% 100 == 0) {
        message("Running Wasserstein test for taxon ", taxon_idx, " / ", 
                length(taxon_ids))
      }
      
      keep_nonzero <- count_mat[, taxon_id] > 0
      x1 <- clr_mat[groups == group_levels[1] & keep_nonzero, taxon_id]
      x2 <- clr_mat[groups == group_levels[2] & keep_nonzero, taxon_id]
      
      set.seed(seed + taxon_idx)
      res <- suppressWarnings(.wassersteinTestSp(x1, x2, permnum = n_perm))
      normalise_wass_result(res, n_perm = n_perm)
    }
  )
  
  names(wass_res_list) <- taxon_ids
  
  wass_stats <- do.call(
    rbind,
    lapply(wass_res_list, function(res) {
      as.numeric(res[names(res) != "wass.values"])
    })
  )
  
  colnames(wass_stats) <- names(wass_res_list[[1]])[
    names(wass_res_list[[1]]) != "wass.values"
  ]
  rownames(wass_stats) <- taxon_ids
  
  perm_stats <- do.call(
    cbind,
    lapply(wass_res_list, function(res) as.numeric(res$wass.values))
  )
  colnames(perm_stats) <- taxon_ids
  
  results <- as_tibble(wass_stats, rownames = "taxon_id") %>%
    mutate(
      p.zero = base::unname(p_zero[taxon_id]),
      p.combined = safe_fisher_combine(p.zero, pval),
      p.adj.nonzero = p.adjust(pval, method = "BH"),
      p.adj.zero = p.adjust(p.zero, method = "BH"),
      p.adj.combined = p.adjust(p.combined, method = "BH"),
      n_nonzero_group1 = colSums(count_mat[groups == group_levels[1], , drop = FALSE] > 0),
      n_nonzero_group2 = colSums(count_mat[groups == group_levels[2], , drop = FALSE] > 0),
      n_group1 = sum(groups == group_levels[1]),
      n_group2 = sum(groups == group_levels[2])
    )
  
  list(
    results = results,
    perm_stats = perm_stats,
    group_levels = group_levels
  )
}

run_complete_clr_waddr <- function(clr_mat, groups, n_perm, seed) {
  taxon_ids <- colnames(clr_mat)
  group_levels <- levels(groups)
  
  if (length(group_levels) != 2) {
    stop("The complete-CLR waddR analysis requires exactly two groups.")
  }
  
  wass_res_list <- purrr::imap(
    taxon_ids,
    function(taxon_id, taxon_idx) {
      if (taxon_idx %% 100 == 0) {
        message("Running complete-CLR Wasserstein test for taxon ", taxon_idx,
                " / ", length(taxon_ids))
      }
      
      x1 <- clr_mat[groups == group_levels[1], taxon_id]
      x2 <- clr_mat[groups == group_levels[2], taxon_id]
      
      set.seed(seed + taxon_idx)
      res <- suppressWarnings(.wassersteinTestSp(x1, x2, permnum = n_perm))
      normalise_wass_result(res, n_perm = n_perm)
    }
  )
  
  names(wass_res_list) <- taxon_ids
  
  wass_stats <- do.call(
    rbind,
    lapply(wass_res_list, function(res) {
      as.numeric(res[names(res) != "wass.values"])
    })
  )
  
  colnames(wass_stats) <- names(wass_res_list[[1]])[
    names(wass_res_list[[1]]) != "wass.values"
  ]
  rownames(wass_stats) <- taxon_ids
  
  perm_stats <- do.call(
    cbind,
    lapply(wass_res_list, function(res) as.numeric(res$wass.values))
  )
  colnames(perm_stats) <- taxon_ids
  
  results <- as_tibble(wass_stats, rownames = "taxon_id") %>%
    mutate(
      p.adj = p.adjust(pval, method = "BH"),
      n_group1 = sum(groups == group_levels[1]),
      n_group2 = sum(groups == group_levels[2])
    )
  
  list(
    results = results,
    perm_stats = perm_stats,
    group_levels = group_levels
  )
}

make_permapprox_results <- function(obs_stats, perm_stats, sample_size, cores) {
  valid <- is.finite(obs_stats) & colSums(is.finite(perm_stats)) > 0
  
  if (!any(valid)) {
    stop("No valid Wasserstein statistics available for permApprox.")
  }
  
  gpd_ctrl_unconstr <- permApprox::make_gpd_ctrl(
    constraint = "unconstrained",
    sample_size = sample_size
  )
  
  gpd_ctrl_constr <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = sample_size
  )
  
  expand_permapprox_result <- function(res, valid, n_total) {
    n_valid <- sum(valid)
    
    for (nm in names(res)) {
      x <- res[[nm]]
      
      if (is.atomic(x) && length(x) == n_valid) {
        out <- rep(NA, n_total)
        out[valid] <- x
        res[[nm]] <- out
      } else if (is.data.frame(x) && nrow(x) == n_valid) {
        out <- x[rep(1, n_total), , drop = FALSE]
        out[,] <- NA
        out[valid, ] <- x
        rownames(out) <- NULL
        res[[nm]] <- out
      } else if (is.matrix(x) && nrow(x) == n_valid) {
        out <- matrix(
          NA,
          nrow = n_total,
          ncol = ncol(x),
          dimnames = list(NULL, colnames(x))
        )
        out[valid, ] <- x
        res[[nm]] <- out
      }
    }
    
    res$valid_test <- valid
    res
  }
  
  unconstrained <- permApprox::perm_approx(
      obs_stats = obs_stats[valid],
      perm_stats = perm_stats[, valid, drop = FALSE],
      method = "gpd",
      alternative = "greater",
      null_center = "mean",
      approx_thresh = 1,
      adjust_method = "none",
      gpd_ctrl = gpd_ctrl_unconstr,
      cores = cores,
      verbose = TRUE
  )
  
  constrained <- permApprox::perm_approx(
      obs_stats = obs_stats[valid],
      perm_stats = perm_stats[, valid, drop = FALSE],
      method = "gpd",
      alternative = "greater",
      null_center = "mean",
      approx_thresh = 1,
      adjust_method = "none",
      gpd_ctrl = gpd_ctrl_constr,
      cores = cores,
      verbose = TRUE
  )
  
  list(
    unconstrained = expand_permapprox_result(
      unconstrained,
      valid = valid,
      n_total = length(obs_stats)
    ),
    constrained = expand_permapprox_result(
      constrained,
      valid = valid,
      n_total = length(obs_stats)
    )
  )
}

extract_gpd_column <- function(x, column) {
  if (!is.null(x$gpd_fit) && column %in% colnames(x$gpd_fit)) {
    return(x$gpd_fit[[column]])
  }
  
  rep(NA_real_, length(x$p_unadjusted))
}

make_distribution_diagnostic_data <- function(
    results,
    perm_stats,
    clr_mat,
    count_mat,
    sample_data,
    selected_taxa,
    include_original_zeros = TRUE
) {
  selected_taxa_ids <- selected_taxa$taxon_id
  
  value_df <- as_tibble(
    clr_mat[, selected_taxa_ids, drop = FALSE],
    rownames = "SampleID"
  ) %>%
    pivot_longer(
      cols = all_of(selected_taxa_ids),
      names_to = "taxon_id",
      values_to = "clr_abundance"
    ) %>%
    left_join(sample_data %>% select(SampleID, EBF_group), by = "SampleID")
  
  if (!include_original_zeros) {
    count_df <- as_tibble(
      count_mat[, selected_taxa_ids, drop = FALSE],
      rownames = "SampleID"
    ) %>%
      pivot_longer(
        cols = all_of(selected_taxa_ids),
        names_to = "taxon_id",
        values_to = "count"
      )
    
    value_df <- value_df %>%
      left_join(count_df, by = c("SampleID", "taxon_id")) %>%
      filter(count > 0)
  }
  
  value_df <- value_df %>%
    mutate(taxon_id = factor(taxon_id, levels = selected_taxa_ids))
  
  quantile_df <- value_df %>%
    filter(!is.na(clr_abundance), !is.na(EBF_group)) %>%
    group_by(taxon_id, EBF_group) %>%
    arrange(clr_abundance, .by_group = TRUE) %>%
    mutate(cum_prob = (row_number() - 0.5) / n()) %>%
    ungroup()
  
  if (is.null(colnames(perm_stats))) {
    colnames(perm_stats) <- results$taxon_id
  }
  
  perm_df <- perm_stats[, selected_taxa_ids, drop = FALSE] %>%
    as_tibble() %>%
    mutate(perm_id = row_number(), .before = 1) %>%
    pivot_longer(
      cols = all_of(selected_taxa_ids),
      names_to = "taxon_id",
      values_to = "perm_value"
    ) %>%
    left_join(
      selected_taxa %>%
        select(taxon_id, d_wass2 = `d.wass^2`),
      by = "taxon_id"
    ) %>%
    mutate(taxon_id = factor(taxon_id, levels = selected_taxa_ids))
  
  list(
    value_df = value_df,
    quantile_df = quantile_df,
    perm_df = perm_df
  )
}

plot_distribution_diagnostics <- function(
    diag_data,
    panel_labels,
    group_palette,
    ncol,
    axis_title_size = 16,
    axis_text_size = 14,
    strip_text_size = 14,
    legend_text_size = 16,
    legend_title_size = 16,
    row_spacing = 18
) {
  hist_plot <- ggplot(
    diag_data$value_df,
    aes(x = clr_abundance, fill = EBF_group)
  ) +
    geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
    scale_fill_manual(values = group_palette) +
    facet_wrap(
      ~ taxon_id,
      ncol = ncol,
      scales = "free",
      labeller = as_labeller(panel_labels)
    ) +
    labs(x = "CLR coordinate", y = "Count", fill = NULL) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = axis_title_size),
      axis.text = element_text(size = axis_text_size),
      strip.text = element_text(size = strip_text_size),
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.margin = margin(b = row_spacing)
    )
  
  quantile_plot <- ggplot(
    diag_data$quantile_df,
    aes(x = cum_prob, y = clr_abundance, color = EBF_group)
  ) +
    geom_line(linewidth = 0.7) +
    scale_color_manual(values = group_palette) +
    facet_wrap(
      ~ taxon_id,
      ncol = ncol,
      scales = "free_y",
      labeller = as_labeller(panel_labels)
    ) +
    labs(
      x = "Cumulative probability",
      y = "CLR coordinate",
      color = NULL
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = axis_title_size),
      axis.text = element_text(size = axis_text_size),
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      strip.text = element_blank(),
      plot.margin = margin(b = row_spacing)
    )
  
  obs_stat_df <- diag_data$perm_df %>%
    distinct(taxon_id, d_wass2)
  
  perm_plot <- ggplot(diag_data$perm_df, aes(x = perm_value)) +
    geom_histogram(bins = 30, fill = "grey70", color = "white") +
    geom_vline(
      data = obs_stat_df,
      aes(xintercept = d_wass2, color = "Observed test statistic"),
      linewidth = 0.8
    ) +
    facet_wrap(
      ~ taxon_id,
      ncol = ncol,
      scales = "free",
      labeller = as_labeller(panel_labels)
    ) +
    scale_color_manual(
      values = c("Observed test statistic" = "tomato"),
      breaks = "Observed test statistic",
      name = NULL
    ) +
    labs(x = expression(d[W]^2), y = "Permutation count") +
    theme_minimal() +
    theme(
      axis.title = element_text(size = axis_title_size),
      axis.text = element_text(size = axis_text_size),
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size),
      panel.grid.minor = element_blank(),
      strip.text = element_blank()
    )
  
  hist_plot / quantile_plot / perm_plot +
    plot_layout(guides = "collect", heights = c(1, 1, 1)) &
    theme(
      legend.position = "bottom",
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size)
    )
}
```

## Load modified waddR implementation

The adapted analysis uses the low-level Wasserstein permutation test from the
modified local `waddR` implementation. This version exports the permutation
statistics required by `permApprox`.

```{r source_waddr_functions, message=FALSE, warning=FALSE}
waddR_lib <- system.file("libs", .Platform$r_arch, "waddR.dll", package = "waddR")

if (waddR_lib == "") {
  waddR_lib <- system.file("libs", "waddR.dll", package = "waddR")
}

if (waddR_lib == "") {
  stop("Could not find the installed waddR DLL required by the local R functions.")
}

if (!"waddR" %in% names(getLoadedDLLs())) {
  dyn.load(waddR_lib)
}

thesis_path <- dirname(application_path)
waddR_path <- file.path(thesis_path, "ch06_permapprox", "functions", "waddR", "R")
waddR_functions <- list.files(waddR_path, pattern = "\\.R$", full.names = TRUE)

if (length(waddR_functions) == 0) {
  stop("No local waddR R files found at: ", waddR_path)
}

invisible(sapply(waddR_functions, source))
```

## Prepare relative abundance and CLR matrices

```{r prepare_clr_matrix}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_dd <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw))) %>%
  make_dd_metadata()

counts_raw <- counts_raw[sampdata_dd$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_dd$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

clr_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_clr <- clr_object$z

dd_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction = mean(counts_raw == 0),
  detection_limit = clr_object$detection_limit,
  replacement_value = clr_object$replacement_value,
  replacement_fraction = clr_object$frac
)

dd_preprocessing_summary
```

```{r save_clr_objects}
saveRDS(
  clr_object,
  file = path_target("diff_distribution_multrepl_clr_object.rds")
)

write_csv(
  dd_preprocessing_summary,
  file = path_target("diff_distribution_preprocessing_summary.csv")
)
```

## Differential distribution setup

The differential distribution analysis compares children with no exclusive
breastfeeding (`EBF duration == "0 months"`) to children exclusively breastfed
for at least two months (`EBF duration == ">=2 months"`). Samples in the sparse
`1 month` category and samples with missing EBF duration are excluded from this
two-group comparison.

```{r diff_distribution_setup}
dd_sample_data <- sampdata_dd %>%
  mutate(
    EBF_group = case_when(
      `EBF duration` == "0 months" ~ "non-EBF",
      `EBF duration` == ">=2 months" ~ "EBF",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EBF_group)) %>%
  mutate(EBF_group = factor(EBF_group, levels = c("non-EBF", "EBF"))) %>%
  select(SampleID, `EBF duration`, EBF_group)

X_counts <- counts_raw[dd_sample_data$SampleID, , drop = FALSE]
X_rel <- counts_rel[dd_sample_data$SampleID, , drop = FALSE]
X_clr <- counts_clr[dd_sample_data$SampleID, , drop = FALSE]
y_groups <- dd_sample_data$EBF_group
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)
dd_group_palette <- c("non-EBF" = "#1F78B4", "EBF" = "#E66101")

stopifnot(identical(rownames(X_counts), dd_sample_data$SampleID))
stopifnot(identical(rownames(X_clr), dd_sample_data$SampleID))

dd_group_summary <- dd_sample_data %>%
  count(EBF_group, name = "n_samples")

dd_analysis_summary <- tibble(
  n_samples = nrow(X_counts),
  n_taxa = ncol(X_counts),
  n_non_ebf = sum(y_groups == "non-EBF"),
  n_ebf = sum(y_groups == "EBF"),
  excluded_ebf_one_month = sum(sampdata_dd$`EBF duration` == "1 month", na.rm = TRUE),
  excluded_missing_ebf = sum(is.na(sampdata_dd$`EBF duration`)),
  n_perm = n_perm,
  seed = permutation_seed
)

dd_group_summary
dd_analysis_summary
```

```{r save_diff_distribution_setup}
write_csv(
  dd_group_summary,
  file = path_target("diff_distribution_group_summary.csv")
)

write_csv(
  dd_analysis_summary,
  file = path_target("diff_distribution_analysis_summary.csv")
)

saveRDS(
  list(
    sample_data = dd_sample_data,
    X_counts = X_counts,
    X_rel = X_rel,
    X_clr = X_clr,
    y_groups = y_groups
  ),
  file = path_target("diff_distribution_analysis_objects.rds")
)
```

## Adapted two-stage approach

### waddR analysis

The zero stage is computed from the original counts. The non-zero Wasserstein
stage uses CLR values from samples where the original count of the tested taxon
is positive.

```{r run_adapted_waddr}
path_waddr <- path_target(
  paste0("waddr_adapted_ts_nperm=", n_perm, "_seed=", permutation_seed, ".rds")
)

if (!file.exists(path_waddr) || isTRUE(refit_waddr)) {
  register(SerialParam())
  
  runtime_waddr <- system.time({
    waddr_adapted <- run_adapted_waddr(
      count_mat = X_counts,
      clr_mat = X_clr,
      groups = y_groups,
      n_perm = n_perm,
      seed = permutation_seed
    )
  })
  
  waddr_adapted$runtime <- runtime_waddr
  saveRDS(waddr_adapted, file = path_waddr)
} else {
  waddr_adapted <- readRDS(path_waddr)
}

waddr_results <- waddr_adapted$results
waddr_perm_stats <- waddr_adapted$perm_stats

waddr_results %>%
  arrange(pval) %>%
  head(n = 10)
```

```{r waddr_summary}
taxon_labels <- taxon_labels_from_phyloseq(phylo_genus)

waddr_summary <- waddr_results %>%
  summarise(
    n_taxa = n(),
    n_nonzero_testable = sum(!is.na(pval)),
    n_zero_p_waddr = sum(pval == 0, na.rm = TRUE),
    n_zero_stage_testable = sum(!is.na(p.zero)),
    n_zero_p_combined = sum(p.combined == 0, na.rm = TRUE),
    n_significant_nonzero_raw = sum(pval < alpha_diff_distribution, na.rm = TRUE),
    n_significant_nonzero_bh = sum(p.adj.nonzero < alpha_diff_distribution, na.rm = TRUE),
    n_significant_combined_raw = sum(p.combined < alpha_diff_distribution, na.rm = TRUE),
    n_significant_combined_bh = sum(p.adj.combined < alpha_diff_distribution, na.rm = TRUE)
  )

waddr_summary
```

```{r save_waddr_results}
write_csv(
  waddr_summary,
  file = path_target("waddr_adapted_summary.csv")
)

write_csv(
  waddr_results %>% mutate(taxon = base::unname(taxon_labels[taxon_id])),
  file = path_target("waddr_adapted_results.csv")
)
```

### p-value refinement with permApprox

```{r run_permapprox}
path_permapprox <- path_target(
  paste0("permapprox_waddr_adapted_nperm=", n_perm, "_seed=", permutation_seed, ".rds")
)

obs_stats <- waddr_results$`d.wass^2`
names(obs_stats) <- waddr_results$taxon_id

if (!file.exists(path_permapprox) || isTRUE(refit_permapprox)) {
  dd_permapprox <- make_permapprox_results(
    obs_stats = obs_stats,
    perm_stats = waddr_perm_stats,
    sample_size = min(dd_group_summary$n_samples),
    cores = cores
  )
  
  saveRDS(dd_permapprox, file = path_permapprox)
} else {
  dd_permapprox <- readRDS(path_permapprox)
}

summary(dd_permapprox$constrained)
```

```{r combine_results}
dd_results_all <- waddr_results %>%
  mutate(
    taxon = base::unname(taxon_labels[taxon_id]),
    p_permapprox_unconstrained = dd_permapprox$unconstrained$p_unadjusted,
    p_permapprox_constrained = dd_permapprox$constrained$p_unadjusted,
    p_combined_permapprox_unconstrained = safe_fisher_combine(
      p.zero,
      p_permapprox_unconstrained
    ),
    p_combined_permapprox_constrained = safe_fisher_combine(
      p.zero,
      p_permapprox_constrained
    ),
    p_adj_permapprox_unconstrained = p.adjust(
      p_permapprox_unconstrained,
      method = "BH"
    ),
    p_adj_permapprox_constrained = p.adjust(
      p_permapprox_constrained,
      method = "BH"
    ),
    p_adj_combined_permapprox_unconstrained = p.adjust(
      p_combined_permapprox_unconstrained,
      method = "BH"
    ),
    p_adj_combined_permapprox_constrained = p.adjust(
      p_combined_permapprox_constrained,
      method = "BH"
    ),
    method_used_unconstrained = dd_permapprox$unconstrained$method_used,
    method_used_constrained = dd_permapprox$constrained$method_used,
    shape_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox$unconstrained,
      "shape"
    ),
    shape_permapprox_constrained = extract_gpd_column(
      dd_permapprox$constrained,
      "shape"
    ),
    scale_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox$unconstrained,
      "scale"
    ),
    scale_permapprox_constrained = extract_gpd_column(
      dd_permapprox$constrained,
      "scale"
    )
  )

dd_permapprox_summary <- dd_results_all %>%
  summarise(
    n_taxa = n(),
    n_testable_waddr = sum(!is.na(pval)),
    n_testable_permapprox_unconstrained = sum(!is.na(p_permapprox_unconstrained)),
    n_testable_permapprox_constrained = sum(!is.na(p_permapprox_constrained)),
    n_zero_waddr = sum(pval == 0, na.rm = TRUE),
    n_zero_permapprox_unconstrained = sum(p_permapprox_unconstrained == 0, na.rm = TRUE),
    n_zero_permapprox_constrained = sum(p_permapprox_constrained == 0, na.rm = TRUE),
    n_sig_waddr_bh = sum(p.adj.nonzero < alpha_diff_distribution, na.rm = TRUE),
    n_sig_permapprox_unconstrained_bh = sum(
      p_adj_permapprox_unconstrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_permapprox_constrained_bh = sum(
      p_adj_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_combined_waddr_bh = sum(p.adj.combined < alpha_diff_distribution, na.rm = TRUE),
    n_sig_combined_permapprox_unconstrained_bh = sum(
      p_adj_combined_permapprox_unconstrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_combined_permapprox_constrained_bh = sum(
      p_adj_combined_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    )
  )

dd_permapprox_summary
```

```{r save_permapprox_results}
saveRDS(
  dd_results_all,
  file = path_target("diff_distribution_results_all.rds")
)

write_csv(
  dd_results_all,
  file = path_target("diff_distribution_results_all.csv")
)

write_csv(
  dd_permapprox_summary,
  file = path_target("diff_distribution_permapprox_summary.csv")
)
```

### P-values from separated-zero waddR and permApprox

This diagnostic follows the zero-p-value plot used in the permApprox single-cell
application. It separates taxa with non-zero empirical waddR p-values from taxa
for which waddR returns zero, and then shows whether unconstrained or
constrained permApprox resolves those zeros.

```{r pvalue_plot_data}
plot_methods <- tibble(
  method_key = c(
    "waddR",
    "permApprox_unconstrained",
    "permApprox_constrained"
  ),
  method_label = c(
    "waddR",
    "waddR + permApprox(U)",
    "waddR + permApprox(C)"
  ),
  p_col = c(
    "pval",
    "p_permapprox_unconstrained",
    "p_permapprox_constrained"
  )
)

pvalue_plot_df <- purrr::map_dfr(
  seq_len(nrow(plot_methods)),
  function(i) {
    p_col <- plot_methods$p_col[i]
    
    dd_results_all %>%
      transmute(
        taxon_id,
        taxon,
        method_key = plot_methods$method_key[i],
        method_label = plot_methods$method_label[i],
        d_wass2 = `d.wass^2`,
        pvalue = .data[[p_col]],
        p_waddr = pval,
        p_permapprox_unconstrained,
        p_permapprox_constrained
      )
  }
) %>%
  mutate(
    zero_class = case_when(
      is.na(p_waddr) ~ "na_waddR",
      p_waddr > 0 ~ "nonzero_waddR",
      p_waddr == 0 & p_permapprox_unconstrained == 0 ~ "zero_both_unconstrained",
      p_waddr == 0 ~ "zero_waddR_only",
      TRUE ~ NA_character_
    ),
    zero_class = factor(
      zero_class,
      levels = c(
        "nonzero_waddR",
        "zero_waddR_only",
        "zero_both_unconstrained",
        "na_waddR"
      )
    ),
    method_key = factor(method_key, levels = plot_methods$method_key),
    method_label = factor(method_label, levels = plot_methods$method_label)
  )

facet_labels <- pvalue_plot_df %>%
  group_by(method_label) %>%
  summarise(
    n_zero = sum(pvalue == 0, na.rm = TRUE),
    n_non_na = sum(!is.na(pvalue)),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(method_label, "\n(zeros: ", n_zero, " / ", n_non_na, ")")
  )

facet_labeller <- setNames(facet_labels$label, facet_labels$method_label)

pvalue_plot_df
```

```{r pvalues_waddr_permapprox, fig.width=10, fig.height=4.5, dpi=250}
zero_repl <- 1e-60
log_breaks <- c(1e-3, 1e-25, 1e-50, zero_repl)

pvalue_plot_df %>%
  mutate(
    pvalue_plot = case_when(
      is.na(pvalue) ~ NA_real_,
      pvalue == 0 ~ zero_repl,
      pvalue < zero_repl ~ NA_real_,
      TRUE ~ pvalue
    )
  ) %>%
  ggplot(aes(x = d_wass2, y = pvalue_plot)) +
  geom_point(
    data = function(x) filter(x, zero_class == "nonzero_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2.4,
    alpha = 0.85,
    stroke = 1
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_waddR_only"),
    aes(color = zero_class, shape = zero_class),
    size = 2.4,
    alpha = 0.9,
    stroke = 0.4
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_both_unconstrained"),
    aes(color = zero_class, shape = zero_class),
    size = 2,
    alpha = 0.9,
    stroke = 0.3
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "na_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2,
    alpha = 0.9,
    stroke = 0.6
  ) +
  geom_hline(yintercept = 1 / (n_perm + 1), linetype = "dashed", color = "black") +
  facet_wrap(
    ~ method_label,
    nrow = 1,
    labeller = as_labeller(facet_labeller)
  ) +
  scale_y_log10(
    breaks = log_breaks,
    labels = function(x) {
      out <- scales::label_scientific()(x)
      out[x <= zero_repl * 10] <- "0"
      out
    }
  ) +
  coord_cartesian(ylim = c(zero_repl / 2, 1)) +
  scale_color_manual(
    values = c(
      "nonzero_waddR" = "grey40",
      "zero_waddR_only" = "#1f77b4",
      "zero_both_unconstrained" = "darkorange2",
      "na_waddR" = "grey20"
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  scale_shape_manual(
    values = c(
      "nonzero_waddR" = 16,
      "zero_waddR_only" = 17,
      "zero_both_unconstrained" = 16,
      "na_waddR" = 7
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(d[W]^2),
    y = expression(p),
    color = NULL,
    shape = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.spacing = unit(1.6, "lines"),
    strip.text = element_text(size = 11),
    legend.position = c(0.23, 0.32),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )
```

### Two-stage p-value components

This panel plot separates the three p-values used in the adapted two-stage
analysis: the non-zero Wasserstein component, the zero component, and the
combined p-value. Colours indicate BH significance within each component.

```{r two_stage_pvalue_components_data}
alpha_diff_distribution_label <- paste0("BH < ", alpha_diff_distribution)
two_stage_component_label_n <- 6
volcano_label_width <- 22

two_stage_pvalue_components_df <- bind_rows(
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Non-zero distribution",
      d_wass2 = `d.wass^2`,
      pvalue = p_permapprox_constrained,
      qvalue = p_adj_permapprox_constrained
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Zero prevalence",
      d_wass2 = `d.wass^2`,
      pvalue = p.zero,
      qvalue = p.adj.zero
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Combined",
      d_wass2 = `d.wass^2`,
      pvalue = p_combined_permapprox_constrained,
      qvalue = p_adj_combined_permapprox_constrained
    )
) %>%
  filter(is.finite(d_wass2), is.finite(pvalue)) %>%
  mutate(
    component = factor(
      component,
      levels = c("Non-zero distribution", "Zero prevalence", "Combined")
    ),
    neglog10_p = -log10(pvalue),
    significance_class = if_else(
      qvalue < alpha_diff_distribution,
      alpha_diff_distribution_label,
      "Not significant"
    ),
    significance_class = factor(
      significance_class,
      levels = c(alpha_diff_distribution_label, "Not significant")
    )
  )

two_stage_component_labels <- two_stage_pvalue_components_df %>%
  group_by(component) %>%
  summarise(
    n_sig = sum(significance_class == alpha_diff_distribution_label, na.rm = TRUE),
    n_testable = n(),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(
      component,
      "\n(",
      alpha_diff_distribution_label,
      ": ",
      n_sig,
      " / ",
      n_testable,
      ")"
    )
  )

two_stage_component_labeller <- setNames(
  two_stage_component_labels$label,
  two_stage_component_labels$component
)

two_stage_component_label_df <- two_stage_pvalue_components_df %>%
  filter(significance_class == alpha_diff_distribution_label) %>%
  group_by(component) %>%
  arrange(qvalue, desc(neglog10_p), .by_group = TRUE) %>%
  slice_head(n = two_stage_component_label_n) %>%
  ungroup() %>%
  mutate(
    label_taxon = stringr::str_trunc(
      if_else(
        is.na(taxon) | taxon == "",
        taxon_id,
        taxon
      ),
      width = volcano_label_width
    )
  )

two_stage_pvalue_components_df
```

```{r volcano_two_stage_three_panels, fig.width=10, fig.height=4.5, dpi=250}
volcano_combined_two_stage <- ggplot(
  two_stage_pvalue_components_df,
  aes(x = d_wass2, y = neglog10_p)
) +
  geom_point(
    aes(color = significance_class),
    size = 2.3,
    alpha = 0.85
  ) +
  geom_text_repel(
    data = two_stage_component_label_df,
    aes(label = label_taxon),
    size = 4,
    min.segment.length = 0,
    box.padding = 0.5,
    max.overlaps = Inf,
    segment.color = "grey50",
    segment.alpha = 0.55,
    seed = permutation_seed
  ) +
  geom_hline(
    yintercept = -log10(alpha_diff_distribution),
    linetype = "dashed",
    color = "grey35"
  ) +
  facet_wrap(
    ~ component,
    nrow = 1,
    labeller = as_labeller(two_stage_component_labeller)
  ) +
  scale_color_manual(
    values = c(
      setNames("#D55E00", alpha_diff_distribution_label),
      "Not significant" = "grey60"
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(d[W]^2),
    y = expression(-log[10](p)),
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 14),
    legend.text = element_text(size = 16),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    legend.position = "bottom"
  )
volcano_combined_two_stage
```

```{r table_two_stage_significant_taxa, results="asis"}
two_stage_sig_taxa_tbl <- dd_results_all %>%
  filter(
    p_adj_permapprox_constrained < alpha_diff_distribution |
      p.adj.zero < alpha_diff_distribution |
      p_adj_combined_permapprox_constrained < alpha_diff_distribution
  ) %>%
  arrange(
    p_adj_combined_permapprox_constrained,
    p_adj_permapprox_constrained,
    p.adj.zero
  ) %>%
  transmute(
    Taxon = taxon,
    `$d_W^2$` = round(`d.wass^2`, 3),
    `$p_{+}$` = sci(p_permapprox_constrained),
    `$q_{+}$` = sci(p_adj_permapprox_constrained),
    `$p_{0}$` = sci(p.zero),
    `$q_{0}$` = sci(p.adj.zero),
    `$p_{comb}$` = sci(p_combined_permapprox_constrained),
    `$q_{comb}$` = sci(p_adj_combined_permapprox_constrained)
  )

two_stage_sig_taxa_table <- kbl(
  two_stage_sig_taxa_tbl,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste0(
    "Genera significant at BH < ",
    alpha_diff_distribution,
    " in at least one component of the adapted two-stage differential ",
    "distribution analysis. The non-zero and combined p-values use the ",
    "constrained \\texttt{permApprox} refinement."
  ),
  label = "application:diff-distribution-two-stage-significant-taxa"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 8)

two_stage_sig_taxa_table

save_kable(
  two_stage_sig_taxa_table,
  path_target("diff_distribution_two_stage_significant_taxa_table.tex")
)
```

### Volcano plot

The differential distribution volcano plot uses the observed squared
Wasserstein distance as the effect statistic and the constrained permApprox
combined p-value for the zero and non-zero stages as the significance measure.

```{r volcano_plot_data}
volcano_q_threshold <- alpha_diff_distribution
volcano_label_n <- 12

dd_volcano_plot_df <- dd_results_all %>%
  filter(is.finite(`d.wass^2`), is.finite(p_combined_permapprox_constrained)) %>%
  mutate(
    neglog10_p_combined = -log10(p_combined_permapprox_constrained),
    significance_class = case_when(
      p_adj_combined_permapprox_constrained < volcano_q_threshold ~
        "combined_bh_significant",
      TRUE ~ "not_significant"
    ),
    significance_class = factor(
      significance_class,
      levels = c(
        "combined_bh_significant",
        "not_significant"
      )
    ),
    label_taxon = stringr::str_trunc(
      if_else(
        is.na(taxon) | taxon == "",
        taxon_id,
        taxon
      ),
      width = volcano_label_width
    )
  )

volcano_label_df <- dd_volcano_plot_df %>%
  arrange(p_adj_combined_permapprox_constrained, desc(`d.wass^2`)) %>%
  slice_head(n = volcano_label_n)

dd_volcano_plot_df
```

```{r save_volcano_plot_data}
write_csv(
  dd_volcano_plot_df,
  file = path_target("diff_distribution_volcano_plot_data.csv")
)
```

```{r volcano_plot, fig.width=8, fig.height=5, dpi=250}
ggplot(
  dd_volcano_plot_df,
  aes(x = `d.wass^2`, y = neglog10_p_combined)
) +
  geom_point(
    aes(color = significance_class),
    size = 2.4,
    alpha = 0.85
  ) +
  geom_hline(
    yintercept = -log10(alpha_diff_distribution),
    linetype = "dashed",
    color = "grey35"
  ) +
  geom_text_repel(
    data = volcano_label_df,
    aes(label = label_taxon),
    size = 3,
    min.segment.length = 0,
    box.padding = 0.35,
    max.overlaps = Inf,
    seed = permutation_seed
  ) +
  scale_color_manual(
    values = c(
      "combined_bh_significant" = "#D55E00",
      "not_significant" = "grey60"
    ),
    labels = c(
      "combined_bh_significant" = paste0("Combined ", alpha_diff_distribution_label),
      "not_significant" = "Not significant"
    ),
    drop = FALSE
  ) +
  labs(
    title = "Differential distribution volcano plot",
    subtitle = "Adapted waddR with constrained permApprox p-value refinement",
    x = expression(d[W]^2),
    y = expression(-log[10](p[combined])),
    color = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

### Distribution diagnostics for selected genera

These plots focus on the non-zero Wasserstein stage. CLR-transformed
abundances are shown only for samples where the tested genus had a positive
original count, matching the distributions used in the adapted two-stage
waddR test.

```{r selected_taxa_distribution_data}
n_distribution_taxa <- 4

selected_taxa_df <- dd_results_all %>%
  filter(!is.na(`d.wass^2`), !is.na(p_permapprox_constrained)) %>%
  arrange(p_adj_permapprox_constrained, p_permapprox_constrained) %>%
  slice_head(n = n_distribution_taxa) %>%
  mutate(
    taxon_label = taxon,
    taxon_label = if_else(
      is.na(taxon_label) | taxon_label == "",
      taxon_id,
      taxon_label
    ),
    panel_label = paste0(
      stringr::str_trunc(taxon_label, width = 22),
      "\n(d = ", signif(d.wass, 3),
      "; p = ", sci(p_permapprox_constrained), ")"
    )
  )

selected_taxa_ids <- selected_taxa_df$taxon_id
selected_panel_labels <- setNames(selected_taxa_df$panel_label, selected_taxa_ids)

selected_distribution_data <- make_distribution_diagnostic_data(
  results = waddr_results,
  perm_stats = waddr_perm_stats,
  clr_mat = X_clr,
  count_mat = X_counts,
  sample_data = dd_sample_data,
  selected_taxa = selected_taxa_df,
  include_original_zeros = FALSE
)

selected_taxa_df
```

```{r selected_taxa_distribution_diagnostics, fig.width=10, fig.height=7, dpi=250}
plot_distribution_diagnostics(
  diag_data = selected_distribution_data,
  panel_labels = selected_panel_labels,
  group_palette = dd_group_palette,
  ncol = n_distribution_taxa,
  axis_title_size = 14,
  axis_text_size = 14,
  strip_text_size = 13,
  legend_text_size = 16,
  legend_title_size = 16,
  row_spacing = 20
)
```

### Top taxa

```{r table_top_taxa, results="asis"}
n_top <- 20

top_taxa_tbl <- dd_results_all %>%
  filter(!is.na(p_permapprox_constrained)) %>%
  arrange(p_permapprox_constrained) %>%
  slice_head(n = n_top) %>%
  transmute(
    Taxon = taxon,
    `$d_W^2$` = round(`d.wass^2`, 3),
    `$p_{0}$` = sci(p.zero),
    `$p_{+}$ (waddR)` = sci(pval),
    `$p_{+}$ (permApprox C)` = sci(p_permapprox_constrained),
    `$p_{comb}$ (permApprox C)` = sci(p_combined_permapprox_constrained),
    `$q_{comb}$ (permApprox C)` = sci(p_adj_combined_permapprox_constrained)
  )

top_taxa_table <- kbl(
  top_taxa_tbl,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste0(
    "Top ", n_top,
    " taxa in the adapted two-stage differential distribution analysis, ",
    "sorted by constrained \\texttt{permApprox} p-values for the non-zero Wasserstein stage."
  ),
  label = "application:diff-distribution-top-taxa"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 8)

top_taxa_table

save_kable(
  top_taxa_table,
  path_target("diff_distribution_top_taxa_table.tex")
)
```

## One-stage complete CLR approach

### Complete CLR waddR analysis

As a contrast to the separated-zero approach, this analysis runs the
Wasserstein test on the complete CLR-transformed distributions. Samples with an
original zero count are therefore included through their multiplicative
replacement CLR value rather than being handled by a separate zero stage.

```{r run_complete_clr_waddr}
path_waddr_complete <- path_target(
  paste0("waddr_complete_clr_nperm=", n_perm, "_seed=", permutation_seed, ".rds")
)

if (!file.exists(path_waddr_complete) || isTRUE(refit_waddr_complete_clr)) {
  register(SerialParam())
  
  runtime_waddr_complete <- system.time({
    waddr_complete_clr <- run_complete_clr_waddr(
      clr_mat = X_clr,
      groups = y_groups,
      n_perm = n_perm,
      seed = permutation_seed
    )
  })
  
  waddr_complete_clr$runtime <- runtime_waddr_complete
  saveRDS(waddr_complete_clr, file = path_waddr_complete)
} else {
  waddr_complete_clr <- readRDS(path_waddr_complete)
}

waddr_complete_results <- waddr_complete_clr$results
waddr_complete_perm_stats <- waddr_complete_clr$perm_stats

waddr_complete_results %>%
  arrange(pval) %>%
  head(n = 10)
```

```{r waddr_complete_summary}
waddr_complete_summary <- waddr_complete_results %>%
  summarise(
    n_taxa = n(),
    n_testable = sum(!is.na(pval)),
    n_zero_p_waddr = sum(pval == 0, na.rm = TRUE),
    n_significant_raw = sum(pval < alpha_diff_distribution, na.rm = TRUE),
    n_significant_bh = sum(p.adj < alpha_diff_distribution, na.rm = TRUE)
  )

waddr_complete_summary
```

```{r save_waddr_complete_results}
write_csv(
  waddr_complete_summary,
  file = path_target("waddr_complete_clr_summary.csv")
)

write_csv(
  waddr_complete_results %>% mutate(taxon = base::unname(taxon_labels[taxon_id])),
  file = path_target("waddr_complete_clr_results.csv")
)
```

### p-value refinement with permApprox for complete CLR waddR

```{r run_complete_clr_permapprox}
path_permapprox_complete <- path_target(
  paste0(
    "permapprox_waddr_complete_clr_nperm=",
    n_perm,
    "_seed=",
    permutation_seed,
    ".rds"
  )
)

obs_stats_complete <- waddr_complete_results$`d.wass^2`
names(obs_stats_complete) <- waddr_complete_results$taxon_id

if (!file.exists(path_permapprox_complete) || isTRUE(refit_permapprox_complete_clr)) {
  dd_permapprox_complete <- make_permapprox_results(
    obs_stats = obs_stats_complete,
    perm_stats = waddr_complete_perm_stats,
    sample_size = min(dd_group_summary$n_samples),
    cores = cores
  )
  
  saveRDS(dd_permapprox_complete, file = path_permapprox_complete)
} else {
  dd_permapprox_complete <- readRDS(path_permapprox_complete)
}

summary(dd_permapprox_complete$constrained)
```

```{r combine_complete_clr_results}
dd_results_complete_all <- waddr_complete_results %>%
  mutate(
    taxon = base::unname(taxon_labels[taxon_id]),
    p_permapprox_unconstrained = dd_permapprox_complete$unconstrained$p_unadjusted,
    p_permapprox_constrained = dd_permapprox_complete$constrained$p_unadjusted,
    p_adj_permapprox_unconstrained = p.adjust(
      p_permapprox_unconstrained,
      method = "BH"
    ),
    p_adj_permapprox_constrained = p.adjust(
      p_permapprox_constrained,
      method = "BH"
    ),
    method_used_unconstrained = dd_permapprox_complete$unconstrained$method_used,
    method_used_constrained = dd_permapprox_complete$constrained$method_used,
    shape_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox_complete$unconstrained,
      "shape"
    ),
    shape_permapprox_constrained = extract_gpd_column(
      dd_permapprox_complete$constrained,
      "shape"
    ),
    scale_permapprox_unconstrained = extract_gpd_column(
      dd_permapprox_complete$unconstrained,
      "scale"
    ),
    scale_permapprox_constrained = extract_gpd_column(
      dd_permapprox_complete$constrained,
      "scale"
    )
  )

dd_permapprox_complete_summary <- dd_results_complete_all %>%
  summarise(
    n_taxa = n(),
    n_testable_waddr = sum(!is.na(pval)),
    n_testable_permapprox_unconstrained = sum(!is.na(p_permapprox_unconstrained)),
    n_testable_permapprox_constrained = sum(!is.na(p_permapprox_constrained)),
    n_zero_waddr = sum(pval == 0, na.rm = TRUE),
    n_zero_permapprox_unconstrained = sum(p_permapprox_unconstrained == 0, na.rm = TRUE),
    n_zero_permapprox_constrained = sum(p_permapprox_constrained == 0, na.rm = TRUE),
    n_sig_waddr_bh = sum(p.adj < alpha_diff_distribution, na.rm = TRUE),
    n_sig_permapprox_unconstrained_bh = sum(
      p_adj_permapprox_unconstrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    n_sig_permapprox_constrained_bh = sum(
      p_adj_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    )
  )

dd_permapprox_complete_summary
```

```{r save_complete_clr_permapprox_results}
saveRDS(
  dd_results_complete_all,
  file = path_target("diff_distribution_complete_clr_results_all.rds")
)

write_csv(
  dd_results_complete_all,
  file = path_target("diff_distribution_complete_clr_results_all.csv")
)

write_csv(
  dd_permapprox_complete_summary,
  file = path_target("diff_distribution_complete_clr_permapprox_summary.csv")
)
```

### P-values from complete CLR waddR and permApprox

This is the same zero-p-value diagnostic, but for the complete CLR approach
where originally zero counts remain in the distributions after multiplicative
replacement and CLR transformation.

```{r pvalue_plot_complete_clr_data}
pvalue_plot_complete_clr_df <- purrr::map_dfr(
  seq_len(nrow(plot_methods)),
  function(i) {
    p_col <- plot_methods$p_col[i]
    
    dd_results_complete_all %>%
      transmute(
        taxon_id,
        taxon,
        method_key = plot_methods$method_key[i],
        method_label = plot_methods$method_label[i],
        d_wass2 = `d.wass^2`,
        pvalue = .data[[p_col]],
        p_waddr = pval,
        p_permapprox_unconstrained,
        p_permapprox_constrained
      )
  }
) %>%
  mutate(
    zero_class = case_when(
      is.na(p_waddr) ~ "na_waddR",
      p_waddr > 0 ~ "nonzero_waddR",
      p_waddr == 0 & p_permapprox_unconstrained == 0 ~ "zero_both_unconstrained",
      p_waddr == 0 ~ "zero_waddR_only",
      TRUE ~ NA_character_
    ),
    zero_class = factor(
      zero_class,
      levels = c(
        "nonzero_waddR",
        "zero_waddR_only",
        "zero_both_unconstrained",
        "na_waddR"
      )
    ),
    method_key = factor(method_key, levels = plot_methods$method_key),
    method_label = factor(method_label, levels = plot_methods$method_label)
  )

facet_complete_clr_labels <- pvalue_plot_complete_clr_df %>%
  group_by(method_label) %>%
  summarise(
    n_zero = sum(pvalue == 0, na.rm = TRUE),
    n_non_na = sum(!is.na(pvalue)),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(method_label, "\n(zeros: ", n_zero, " / ", n_non_na, ")")
  )

facet_complete_clr_labeller <- setNames(
  facet_complete_clr_labels$label,
  facet_complete_clr_labels$method_label
)

pvalue_plot_complete_clr_df
```

```{r pvalues_complete_clr_waddr_permapprox, fig.width=10, fig.height=4.5, dpi=250}
pvalue_plot_complete_clr_df %>%
  mutate(
    pvalue_plot = case_when(
      is.na(pvalue) ~ NA_real_,
      pvalue == 0 ~ zero_repl,
      pvalue < zero_repl ~ NA_real_,
      TRUE ~ pvalue
    )
  ) %>%
  ggplot(aes(x = d_wass2, y = pvalue_plot)) +
  geom_point(
    data = function(x) filter(x, zero_class == "nonzero_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 3,
    alpha = 0.5,
    stroke = 0.5
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_waddR_only"),
    aes(color = zero_class, shape = zero_class),
    size = 3,
    alpha = 0.9,
    stroke = 0.5
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_both_unconstrained"),
    aes(color = zero_class, shape = zero_class),
    size = 3,
    alpha = 0.9,
    stroke = 0.5
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "na_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2.8,
    alpha = 0.9,
    stroke = 0.6
  ) +
  geom_hline(yintercept = 1 / (n_perm + 1), linetype = "dashed", color = "black") +
  facet_wrap(
    ~ method_label,
    nrow = 1,
    labeller = as_labeller(facet_complete_clr_labeller)
  ) +
  scale_y_log10(
    breaks = log_breaks,
    labels = function(x) {
      out <- scales::label_scientific()(x)
      out[x <= zero_repl * 10] <- "0"
      out
    }
  ) +
  coord_cartesian(ylim = c(zero_repl / 2, 1)) +
  scale_color_manual(
    values = c(
      "nonzero_waddR" = "grey40",
      "zero_waddR_only" = "#1f77b4",
      "zero_both_unconstrained" = "darkorange2",
      "na_waddR" = "grey20"
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  scale_shape_manual(
    values = c(
      "nonzero_waddR" = 16,
      "zero_waddR_only" = 17,
      "zero_both_unconstrained" = 16,
      "na_waddR" = 7
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR -> >0 in permApprox(U)",
      "zero_both_unconstrained" = "Zero in waddR -> 0 in permApprox(U)",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(d[W]^2),
    y = expression(p),
    color = NULL,
    shape = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.spacing = unit(1.6, "lines"),
    strip.text = element_text(size = 11),
    legend.position = c(0.23, 0.32),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )
```

### Constrained permApprox comparison for complete CLR waddR

This two-panel version is intended for the paper. It focuses on the one-stage
complete CLR analysis where waddR returns zero p-values and compares those
directly with the constrained permApprox refinement, labelled simply as
`permApprox`.

```{r pvalue_plot_complete_clr_constrained_data}
plot_methods_complete_constrained <- tibble(
  method_key = c("waddR", "permApprox_constrained"),
  method_label = c("waddR", "waddR + permApprox"),
  p_col = c("pval", "p_permapprox_constrained")
)

pvalue_plot_complete_clr_constrained_df <- purrr::map_dfr(
  seq_len(nrow(plot_methods_complete_constrained)),
  function(i) {
    p_col <- plot_methods_complete_constrained$p_col[i]
    
    dd_results_complete_all %>%
      transmute(
        taxon_id,
        taxon,
        method_key = plot_methods_complete_constrained$method_key[i],
        method_label = plot_methods_complete_constrained$method_label[i],
        d_wass2 = `d.wass^2`,
        pvalue = .data[[p_col]],
        p_waddr = pval,
        p_permapprox = p_permapprox_constrained
      )
  }
) %>%
  mutate(
    zero_class = case_when(
      is.na(p_waddr) ~ "na_waddR",
      p_waddr > 0 ~ "nonzero_waddR",
      p_waddr == 0 & p_permapprox == 0 ~ "zero_both",
      p_waddr == 0 ~ "zero_waddR_only",
      TRUE ~ NA_character_
    ),
    zero_class = factor(
      zero_class,
      levels = c(
        "nonzero_waddR",
        "zero_waddR_only",
        "zero_both",
        "na_waddR"
      )
    ),
    method_key = factor(
      method_key,
      levels = plot_methods_complete_constrained$method_key
    ),
    method_label = factor(
      method_label,
      levels = plot_methods_complete_constrained$method_label
    )
  )

facet_complete_constrained_labels <- pvalue_plot_complete_clr_constrained_df %>%
  group_by(method_label) %>%
  summarise(
    n_zero = sum(pvalue == 0, na.rm = TRUE),
    n_non_na = sum(!is.na(pvalue)),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(method_label, "\n(zeros: ", n_zero, " / ", n_non_na, ")")
  )

facet_complete_constrained_labeller <- setNames(
  facet_complete_constrained_labels$label,
  facet_complete_constrained_labels$method_label
)

pvalue_plot_complete_clr_constrained_df
```

```{r pvalues_complete_clr_waddr_permapprox_constrained_only, fig.width=10, fig.height=3.5, dpi=250}
zero_repl_constrained_plot <- 1e-60
log_breaks_constrained_plot <- c(
  1e-3,
  1e-10,
  1e-25,
  1e-50,
  zero_repl_constrained_plot
)

pvalue_plot_complete_clr_constrained_df %>%
  mutate(
    pvalue_plot = case_when(
      is.na(pvalue) ~ NA_real_,
      pvalue == 0 ~ zero_repl_constrained_plot,
      pvalue < zero_repl_constrained_plot ~ NA_real_,
      TRUE ~ pvalue
    )
  ) %>%
  ggplot(aes(x = d_wass2, y = pvalue_plot)) +
  geom_point(
    data = function(x) filter(x, zero_class == "nonzero_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 3,
    alpha = 0.5,
    stroke = 0.5
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "zero_waddR_only"),
    aes(color = zero_class, shape = zero_class),
    size = 3,
    alpha = 0.9,
    stroke = 0.5
  ) +
  geom_point(
    data = function(x) filter(x, zero_class == "na_waddR"),
    aes(color = zero_class, shape = zero_class),
    size = 2.8,
    alpha = 0.9,
    stroke = 0.6
  ) +
  geom_hline(yintercept = 1 / (n_perm + 1), linetype = "dashed", color = "black") +
  facet_wrap(
    ~ method_label,
    nrow = 1,
    labeller = as_labeller(facet_complete_constrained_labeller)
  ) +
  scale_y_log10(
    breaks = log_breaks_constrained_plot,
    labels = function(x) {
      out <- scales::label_scientific()(x)
      out[x <= zero_repl_constrained_plot * 10] <- "0"
      out
    }
  ) +
  coord_cartesian(ylim = c(zero_repl_constrained_plot / 2, 1)) +
  scale_color_manual(
    values = c(
      "nonzero_waddR" = "grey40",
      "zero_waddR_only" = "darkorange2",
      "na_waddR" = "grey20"
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  scale_shape_manual(
    values = c(
      "nonzero_waddR" = 16,
      "zero_waddR_only" = 16,
      "na_waddR" = 7
    ),
    labels = c(
      "nonzero_waddR" = "Non-zero in waddR",
      "zero_waddR_only" = "Zero in waddR",
      "na_waddR" = "NA in waddR"
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(d[W]^2),
    y = expression(p),
    color = NULL,
    shape = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 14),
    legend.text = element_text(size = 13),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.spacing = unit(1.6, "lines"),
    legend.position = c(0.5, 0.35),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )
```

### Volcano plot

The one-stage volcano plot uses the complete CLR Wasserstein statistic and the
constrained permApprox p-value refinement. Colours indicate BH significance at
the same alpha level used for the two-stage component plots.

```{r complete_clr_volcano_plot_data}
complete_clr_volcano_label_n <- 6
complete_clr_volcano_alpha_label <- paste0("BH < ", alpha_diff_distribution)

complete_clr_volcano_df <- dd_results_complete_all %>%
  filter(is.finite(`d.wass^2`), is.finite(p_permapprox_constrained)) %>%
  mutate(
    component = "Complete CLR distribution",
    neglog10_p = -log10(p_permapprox_constrained),
    significance_class = if_else(
      p_adj_permapprox_constrained < alpha_diff_distribution,
      complete_clr_volcano_alpha_label,
      "Not significant"
    ),
    significance_class = factor(
      significance_class,
      levels = c(complete_clr_volcano_alpha_label, "Not significant")
    ),
    label_taxon = stringr::str_trunc(
      if_else(
        is.na(taxon) | taxon == "",
        taxon_id,
        taxon
      ),
      width = volcano_label_width
    )
  ) %>%
  mutate(
    component = factor(component, levels = "Complete CLR distribution")
  )

complete_clr_volcano_label_df <- complete_clr_volcano_df %>%
  filter(significance_class == complete_clr_volcano_alpha_label) %>%
  arrange(p_adj_permapprox_constrained, desc(neglog10_p)) %>%
  slice_head(n = complete_clr_volcano_label_n)

complete_clr_volcano_labeller <- setNames(
  paste0(
    "Complete CLR distribution\n(",
    complete_clr_volcano_alpha_label,
    ": ",
    sum(
      complete_clr_volcano_df$p_adj_permapprox_constrained < alpha_diff_distribution,
      na.rm = TRUE
    ),
    " / ",
    sum(!is.na(complete_clr_volcano_df$p_adj_permapprox_constrained)),
    ")"
  ),
  "Complete CLR distribution"
)

complete_clr_volcano_df
```

```{r volcano_plot_one_stage, fig.width=4.5, fig.height=4.5, dpi=250}
volcano_one_stage <- ggplot(
  complete_clr_volcano_df,
  aes(x = `d.wass^2`, y = neglog10_p)
) +
  geom_point(
    aes(color = significance_class),
    size = 2.3,
    alpha = 0.85
  ) +
  geom_text_repel(
    data = complete_clr_volcano_label_df,
    aes(label = label_taxon),
    size = 4,
    min.segment.length = 0,
    box.padding = 0.35,
    max.overlaps = Inf,
    segment.color = "grey50",
    segment.alpha = 0.55,
    seed = permutation_seed
  ) +
  geom_hline(
    yintercept = -log10(alpha_diff_distribution),
    linetype = "dashed",
    color = "grey35"
  ) +
  facet_wrap(
    ~ component,
    nrow = 1,
    labeller = as_labeller(complete_clr_volcano_labeller)
  ) +
  scale_color_manual(
    values = setNames(
      c("#D55E00", "grey60"),
      c(complete_clr_volcano_alpha_label, "Not significant")
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(d[W]^2),
    y = expression(-log[10](p)),
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 14),
    legend.text = element_text(size = 16),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

```{r volcano_two_stage_one_stage_combined, fig.width=11, fig.height=5, dpi=250}
(
  volcano_combined_two_stage +
    labs(title = "Two-stage approach") +
    theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
) +
  (
    volcano_one_stage +
      labs(title = "One-stage approach") +
      theme(
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "bottom"
      )
  ) +
  plot_layout(widths = c(3, 1), guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 16)
  )
```


### Distribution diagnostics for selected genera

These plots use the complete CLR-transformed distributions from the one-stage
analysis. Originally zero counts are included through their multiplicative
replacement CLR values, matching the data passed to the one-stage Wasserstein
test.

```{r selected_taxa_complete_clr_distribution_data}
selected_complete_taxa_df <- dd_results_complete_all %>%
  filter(!is.na(`d.wass^2`), !is.na(p_permapprox_constrained)) %>%
  arrange(p_adj_permapprox_constrained, p_permapprox_constrained) %>%
  slice_head(n = n_distribution_taxa) %>%
  mutate(
    taxon_label = taxon,
    taxon_label = if_else(
      is.na(taxon_label) | taxon_label == "",
      taxon_id,
      taxon_label
    ),
    panel_label = paste0(
      stringr::str_trunc(taxon_label, width = 22),
      "\n(d = ", signif(d.wass, 3),
      "; p = ", sci(p_permapprox_constrained), ")"
    )
  )

selected_complete_taxa_ids <- selected_complete_taxa_df$taxon_id
selected_complete_panel_labels <- setNames(
  selected_complete_taxa_df$panel_label,
  selected_complete_taxa_ids
)

selected_complete_distribution_data <- make_distribution_diagnostic_data(
  results = waddr_complete_results,
  perm_stats = waddr_complete_perm_stats,
  clr_mat = X_clr,
  count_mat = X_counts,
  sample_data = dd_sample_data,
  selected_taxa = selected_complete_taxa_df,
  include_original_zeros = TRUE
)

selected_complete_taxa_df
```

```{r selected_taxa_complete_clr_distribution_diagnostics, fig.width=10, fig.height=7.5, dpi=250}
plot_distribution_diagnostics(
  diag_data = selected_complete_distribution_data,
  panel_labels = selected_complete_panel_labels,
  group_palette = dd_group_palette,
  ncol = n_distribution_taxa,
  axis_title_size = 14,
  axis_text_size = 14,
  strip_text_size = 13,
  legend_text_size = 16,
  legend_title_size = 16,
  row_spacing = 20
)
```

## Comparative p-value table

```{r table_pvalues_all_approaches, results="asis"}
pvalues_all_approaches_tbl <- dd_results_all %>%
  transmute(
    taxon_id,
    Taxon = taxon,
    `$p_{+}$` = p_permapprox_constrained,
    `$q_{+}$` = p_adj_permapprox_constrained,
    `$p_{0}$` = p.zero,
    `$q_{0}$` = p.adj.zero,
    `$p_{comb}$` = p_combined_permapprox_constrained,
    `$q_{comb}$` = p_adj_combined_permapprox_constrained
  ) %>%
  left_join(
    dd_results_complete_all %>%
      transmute(
        taxon_id,
        `$p_{1stage}$` = p_permapprox_constrained,
        `$q_{1stage}$` = p_adj_permapprox_constrained
      ),
    by = "taxon_id"
  ) %>%
  mutate(
    min_q = pmin(
      `$q_{+}$`,
      `$q_{0}$`,
      `$q_{comb}$`,
      `$q_{1stage}$`,
      na.rm = TRUE
    )
  ) %>%
  arrange(min_q, Taxon) %>%
  select(-taxon_id, -min_q) %>%
  mutate(
    across(
      where(is.numeric),
      sci
    )
  )

pvalues_all_approaches_table <- kbl(
  pvalues_all_approaches_tbl,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste0(
    "P-values and BH-adjusted p-values for the adapted two-stage ",
    "differential distribution analysis and the one-stage complete CLR ",
    "analysis. The non-zero, combined, and one-stage p-values use the ",
    "constrained \\texttt{permApprox} refinement."
  ),
  label = "application:diff-distribution-pvalues-all-approaches"
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 7)

pvalues_all_approaches_table

save_kable(
  pvalues_all_approaches_table,
  path_target("diff_distribution_pvalues_all_approaches_table.tex")
)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
