---
title: "Network analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "10_networks" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "147f8b0c-cb4b-427a-9549-8c8b8e3ea46b")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/net_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
# SpiecEasi from latentcor branch needed
if (as.character(utils::packageVersion("SpiecEasi")) != "2.0.1") {
  remotes::install_github("zdk123/SpiecEasi", ref = "latentcor")
}

library(conflicted)
library(phyloseq)
library(tidyverse)
library(NetCoMi)
library(permApprox)
library(igraph)
library(kableExtra)
library(patchwork)
library(ggplot2)

# For network comparison
library(foreach)
library(future)
library(doFuture)
library(doRNG)
library(progressr)
library(filematrix)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Aim

This script contains the network analyses for the application chapter. The first
part builds one complete network on the filtered genus-level data and evaluates
how sensitive the resulting network is to common preprocessing and association
choices. The complete-network analysis will compare the standard preprocessing
approach with the final method used in this thesis, and then vary one component
at a time: zero replacement, normalization, and association measure.

The second part will compare networks between the two main exclusive
breastfeeding groups used throughout this chapter, `EBF duration = 0` and
`EBF duration >=2`. The network-comparison analysis will contrast the classical
empirical permutation results with `permApprox`-refined p-values. For the
`permApprox` analysis we will fit both the unconstrained approximation and the
proposed constrained version, so that the unconstrained fit can be used to
inspect how many zero-valued permutation statistics would be expected without
the constraint.

## Set global parameters

```{r parameters}
zero_replacement_fraction <- 0.65
network_seed <- 42
permutation_seed <- 42
n_perm_network_comparison <- 999
cores <- 10

association_measure_final <- "spieceasi"
normalization_final <- "clr"
zero_replacement_final <- "multRepl"

refit_complete_network <- FALSE
refit_method_comparison <- FALSE
refit_network_comparison <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```
## Helper functions

```{r functions}
## Function to turn an association matrix into a graph object
# Arguments:
# - assoMat: association matrix
# - threshold: associations below the threshold are set to zero
# - dissTrans: dissimilarity transformation ("signed" or "unsigned")

transform_asso <- function(assoMat, thresh = NULL, dissTrans = "signed") {
    # Sparsification
    if (!is.null(thresh)) {
        assoMat[abs(assoMat) < thresh] <- 0
    }

    # Compute dissimilarity matrix
    if (dissTrans == "signed") {
        dissMat <- sqrt(0.5 * (1 - assoMat))
    } else {
        dissMat <- sqrt(1 - assoMat^2)
    }

    # Dissimilarity between nodes with zero correlation is set to 1
    # (these nodes are unconnected and thus should have maximum dissimilarity)
    dissMat[assoMat == 0] <- 1

    # Compute similarity matrix
    simMat <- 1 - dissMat

    # Turn into igraph object
    graphObj <- SpiecEasi::adj2igraph(simMat)

    return(list(graph = graphObj, adja = simMat, asso = assoMat, diss = dissMat))
}
```


## Prepare data

The network analyses use the same filtered genus-level object as the
differential abundance and differential distribution analyses. Counts are
represented as a samples-by-taxa matrix. Relative abundances are prepared here
because several network workflows use them directly. 

Since SpiecEasi does only provide pseudo count zero replacement, we need to 
apply zero replacement with multRepl beforehand. 

```{r prepare_matrices, warning=FALSE}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_net <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw)))

counts_raw <- counts_raw[sampdata_net$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_net$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_rel_multrepl <- zero_repl_object$rel_abund_repl

net_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction_orig = mean(counts_raw == 0),
  zero_fraction_repl = mean(counts_rel_multrepl == 0)
)

net_preprocessing_summary
```

Value imputed by multiplicative replacement: `r min(counts_rel[counts_rel > 0]) * 0.65`.

Rename taxonomic table and make rank Genus unique.

```{r}
phylo_genus <- renameTaxa(phylo_genus, 
                          pat = "<name>", 
                          substPat = "<name>_<subst_name>(<subst_R>)",
                          numDupli = "Genus")
```


## Complete network with standard preprocessing

This section will estimate the complete genus-level network using the final
preprocessing and association pipeline selected for the thesis. The resulting
object should be saved to the target directory and reused by downstream summary
tables and figures.

The preprocessing approach used throughout the thesis is:

1. Transform counts to relative abundances
2. Perform multiplicative zero replacement
3. Perform CLR transformation

### Network construction

```{r generate_net_single_main}
path_net_single_main <- path_target("net_single_main.rds")

if (file.exists(path_net_single_main) && !refit_complete_network) {
  net_single_main <- readRDS(path_net_single_main)
} else {
  net_single_main <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_main, path_net_single_main)
}
```

```{r}
# Extract association matrix
assomat_single_main <- net_single_main$assoMat1
adjamat_single_main <- net_single_main$adjaMat1

# Generate graph object
graph_single_main <- transform_asso(assomat_single_main)$graph
```

### Network analysis

#### Association heatmap

**For all taxa**

```{r single_association_heatmap, fig.width=20, fig.height=20}
plotHeat(assomat_single_main, textUpp = "none", textLow = "none")
```

**For non-singletons only**

```{r single_association_heatmap_nonzero, fig.width=15, fig.height=15}
adjamat_single_main_nonzero <- net_single_main$adjaMat1
diag(adjamat_single_main_nonzero) <- 0

taxa_all <- colnames(assomat_single_main)
taxa_with_associations <- taxa_all[rowSums(adjamat_single_main_nonzero) > 0]
assomat_single_main_no_singletons <- 
  assomat_single_main[taxa_with_associations, taxa_with_associations]

plotHeat(
  assomat_single_main_no_singletons,
  textUpp = "none",
  textLow = "none"
)
```

#### Association histogram

```{r single_association_histrogram, fig.width=8, fig.height=2}
n_zero_asso <- sum(assomat_single_main[lower.tri(assomat_single_main)] == 0)

data.frame(correlation = assomat_single_main[lower.tri(assomat_single_main)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.005) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso), size = 4) +
  coord_cartesian(ylim = c(0, 50)) +
  labs(x = "Estimated correlation",
       y = "Count",
       #title = "Estimated correlations after sparsification"
       ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )
```

```{r}
n_asso <- sum(lower.tri(assomat_single_main))
cat("Entries in the lower triangle:", n_asso)

cat("\nNon-zero entries in the lower triangle:", n_asso - n_zero_asso)
```

#### Degree distribution

```{r single_degree_distribution, fig.width=8, fig.height=2, dpi = 200}
# Compute degree distribution
ddist <- igraph::degree.distribution(graph_single_main)

# Data frame needed for ggplot2
df <- data.frame(
    Degree = as.factor((seq_along(ddist)) - 1), Fraction = ddist
)

ggplot(data = df, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))
```

#### Analysis with netAnalyze()

```{r}
props_single_main <- netAnalyze(net_single_main,
                                centrLCC = TRUE,
                                clustMethod = "cluster_fast_greedy",
                                hubPar = "eigenvector",
                                weightDeg = FALSE, 
                                normDeg = FALSE,
                                normBetw = FALSE,
                                normClose = FALSE,
                                normEigen = FALSE)
```

```{r}
print(summary(props_single_main))
```

### Graphlet correlation matrix

```{r single_gcm, fig.width=6, fig.height=6, dpi = 200}
plotGCM(props_single_main,
        legendCex = 1, 
        labCex = 1.1,
        sigTextCex = 1.2,
        titles = "",
        mar = c(1,1,0,1))
```


### Network plot

In the network plot, the following taxa are labelled: 

* Taxa that appeared in the differential association analysis as well as the differential distribution analysis.
* Most central taxa with respect to eigenvector centrality.
* The three taxa with a negative correlation to Bifidobacterium


```{r single_taxa_to_label}
## Most abundant taxa
top_genera_df <- readr::read_csv(
  path_source("02_filtering", "top_genera.csv"),
  show_col_types = FALSE
)

top_genera_plot <- top_genera_df$genus

## Taxa labelled in differential abundance plot
da_volcano_df <- readr::read_csv(
  path_source("08_diff_abundance", "diff_abundance_volcano_plot_data.csv"),
  show_col_types = FALSE
)

da_plot_taxa <- da_volcano_df %>%
  filter(p_plot_raw < 1e-5) %>%
  distinct(taxon_id, taxon) %>%
  { setNames(.$taxon, .$taxon_id) }


## Taxa labelled in differential distribution plot
alpha_diff_distribution <- 0.01
two_stage_component_label_n <- 6
complete_clr_volcano_label_n <- 6

dd_results_all <- readr::read_csv(
  path_source("09_diff_distribution", "diff_distribution_results_all.csv"),
  show_col_types = FALSE
)

dd_results_complete_all <- readr::read_csv(
  path_source("09_diff_distribution", "diff_distribution_complete_clr_results_all.csv"),
  show_col_types = FALSE
)

dd_two_stage_taxa <- dplyr::bind_rows(
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Non-zero distribution",
      native_statistic = `d.wass^2`,
      pvalue = p_permapprox_constrained,
      qvalue = p_adj_permapprox_constrained
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Zero prevalence",
      native_statistic = abs(zero_prevalence_z),
      pvalue = p.zero,
      qvalue = p.adj.zero
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Combined",
      native_statistic = fisher_statistic_permapprox_constrained,
      pvalue = p_combined_permapprox_constrained,
      qvalue = p_adj_combined_permapprox_constrained
    )
) %>%
  filter(is.finite(native_statistic), is.finite(pvalue), qvalue < alpha_diff_distribution) %>%
  group_by(component) %>%
  arrange(qvalue, desc(-log10(pvalue)), .by_group = TRUE) %>%
  slice_head(n = two_stage_component_label_n) %>%
  ungroup() %>%
  distinct(taxon_id, taxon)

dd_one_stage_taxa <- dd_results_complete_all %>%
  filter(
    is.finite(`d.wass^2`),
    is.finite(p_permapprox_constrained),
    p_adj_permapprox_constrained < alpha_diff_distribution
  ) %>%
  arrange(p_adj_permapprox_constrained, desc(-log10(p_permapprox_constrained))) %>%
  slice_head(n = complete_clr_volcano_label_n) %>%
  distinct(taxon_id, taxon)

dd_plot_taxa <- bind_rows(dd_two_stage_taxa, dd_one_stage_taxa) %>%
  distinct(taxon_id, taxon) %>%
  { setNames(.$taxon, .$taxon_id) }

# Define taxa with highest eigenvector centrality
taxa_high_eigen <- props_single_main$centralities$eigenv1 %>%
  sort(decreasing = TRUE) %>%
  head(10) %>%
  names() %>%
  { setNames(., .) }

## Taxa negatively correlated with Bifidobacterium
bif_neg_taxa <- assomat_single_main["Bifidobacterium", ] |>
  (\(x) x[x < 0])() |>
  names() |>
  (\(x) setNames(x, x))()

## Union across all sources
taxa_to_label <- c(top_genera_plot, da_plot_taxa, dd_plot_taxa, taxa_high_eigen, bif_neg_taxa)
taxa_to_label <- taxa_to_label[!duplicated(names(taxa_to_label))]

plot_labels <- setNames(
  rep("", ncol(net_single_main$adjaMat1)),
  colnames(net_single_main$adjaMat1)
)

plot_labels[taxa_to_label] <- taxa_to_label
```

```{r}
# Select nodes to plot
adja_single_main <- net_single_main$adjaMat1
diag(adja_single_main) <- 0

non_singleton_taxa <- names(which(
  rowSums(adja_single_main != 0) > 0 |
    colSums(adja_single_main != 0) > 0
))

taxa_to_label_in_network <- intersect(
  names(taxa_to_label),
  colnames(net_single_main$adjaMat1)
)

taxa_to_plot <- unique(c(
  non_singleton_taxa,
  taxa_to_label_in_network
))

network_labels <- setNames(
  rep("", ncol(net_single_main$adjaMat1)),
  colnames(net_single_main$adjaMat1)
)

network_labels[taxa_to_label_in_network] <- taxa_to_label[taxa_to_label_in_network]
```

#### Node colors: clusters; node size: eigenvector centrality

```{r single_net_plot_basic, fig.width=7, fig.height=6, dpi=250}
clustcols <- c("#E69F00", "#CC79A7", "#F0E442", "#56B4E9", "#0072B2", 
               "#009E73",  "#D55E00", "#33A02C", "#E69F00", "#CC79A7",
               "#F0E442", "#56B4E9", "#0072B2", "#009E73",  "#D55E00")

p <- plot(props_single_main,
          nodeFilter = "names",
          nodeFilterPar = taxa_to_plot,
          labelScale = FALSE,
          labels = network_labels,
          repulsion = 0.899999,
          nodeColor = "cluster", 
          colorVec = clustcols,
          nodeTransp = 40,
          nodeSize = "eigenvector",
          nodeSizeSpread = 3,
          hubBorderCol = "gray40",
          posCol = "gray50", 
          negCol = "red3",
          showTitle = FALSE,
          cexLabels = 0.8,
          mar = c(1, 1, 1, 1))
```

#### Nodes colored by phyla

```{r , fig.width=7, fig.height=6, dpi=250}
# Get phyla names
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]
table(phyla)

# Define phylum colors
phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)
```



```{r, fig.width=7, fig.height=6, dpi=250}
layout <- p$layout$layout1
layout["[Ruminococcus]_gnavugroup", ] <- c(-0.7,  -0.55)
layout["Pseudomonas", ] <- c(-0.1, -0.22)
layout["Blastococcus", ] <- c(-0.33, 0.13)
layout["8_Comamonadaceae(F)", ] <- c(-0.2, -0.1)
layout["Haemophilus", ] <- c(0.84, -0.5)
layout["Lachnospira", ] <- c(-0.34, -0.05)
```

```{r single_net_plot_col_phyla, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_main,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```


```{r single_net_plot_col_phyla_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_main,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Baseline network",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

## Alternative preprocessing steps

In this section we analyze, how the network changes when alternative 
preprocessing steps like filtering or zero replacement, or a different 
association measure is used. 

### Stronger filtering

In the network above, a prevalence filter of 1% was used. We now increase this 
to 5% and 10%. Everything else stays the same.

#### 5% prevalence filter

```{r, warning=FALSE}
phylo_genus_prev05 <- 
  readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev05.rds"))

counts_raw_prev05 <- get_sample_by_taxa(phylo_genus_prev05)

sampdata_net_prev05 <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw_prev05)) %>%
  arrange(match(SampleID, rownames(counts_raw_prev05)))

counts_raw_prev05 <- counts_raw_prev05[sampdata_net_prev05$SampleID, , 
                                       drop = FALSE]

stopifnot(identical(rownames(counts_raw_prev05), sampdata_net_prev05$SampleID))

counts_rel_prev05 <- counts_to_rel_abund(counts_raw_prev05)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel_prev05,
  frac = zero_replacement_fraction
)

counts_rel_multrepl_prev05 <- zero_repl_object$rel_abund_repl
```


```{r generate_net_single_prev05}
path_net_single_prev05 <- path_target("net_single_prev05.rds")

if (file.exists(path_net_single_prev05) && !refit_complete_network) {
  net_single_prev05 <- readRDS(path_net_single_prev05)
} else {
  net_single_prev05 <- NetCoMi::netConstruct(
    data = counts_rel_multrepl_prev05,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_prev05, path_net_single_prev05)
}
```

```{r}
props_single_prev05 <- netAnalyze(net_single_prev05,
                                  centrLCC = TRUE,
                                  clustMethod = "cluster_fast_greedy",
                                  hubPar = "eigenvector",
                                  weightDeg = FALSE, 
                                  normDeg = FALSE,
                                  normBetw = FALSE,
                                  normClose = FALSE,
                                  normEigen = FALSE)
```

```{r single_net_plot_prev05, fig.width=8.5, fig.height=6, dpi=250}
# Subset layout to prev05 taxa that exist in the main layout, preserving prev05 order.
# NetCoMi passes the layout matrix to qgraph positionally (row i = node i), so the
# row order must match colnames(adja1) after node filtering — which follows the
# prev05 adjacency matrix order.
taxa_prev05_in_layout <- colnames(net_single_prev05$adjaMat1)[
  colnames(net_single_prev05$adjaMat1) %in% rownames(layout)
]
layout_prev05 <- layout[taxa_prev05_in_layout, ]

plot(props_single_prev05,
     layout = layout_prev05,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 1.1,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1.1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 
```

```{r single_net_plot_prev05_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_prev05,
     layout = layout_prev05,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "5% prevalence filter",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### 10% prevalence filter

```{r, warning=FALSE}
phylo_genus_prev10 <- 
  readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev10.rds"))

counts_raw_prev10 <- get_sample_by_taxa(phylo_genus_prev10)

sampdata_net_prev10 <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw_prev10)) %>%
  arrange(match(SampleID, rownames(counts_raw_prev10)))

counts_raw_prev10 <- counts_raw_prev10[sampdata_net_prev10$SampleID, , 
                                       drop = FALSE]

stopifnot(identical(rownames(counts_raw_prev10), sampdata_net_prev10$SampleID))

counts_rel_prev10 <- counts_to_rel_abund(counts_raw_prev10)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel_prev10,
  frac = zero_replacement_fraction
)

counts_rel_multrepl_prev10 <- zero_repl_object$rel_abund_repl
```


```{r generate_net_single_prev10}
path_net_single_prev10 <- path_target("net_single_prev10.rds")

if (file.exists(path_net_single_prev10) && !refit_complete_network) {
  net_single_prev10 <- readRDS(path_net_single_prev10)
} else {
  net_single_prev10 <- NetCoMi::netConstruct(
    data = counts_rel_multrepl_prev10,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_prev10, path_net_single_prev10)
}
```

```{r}
props_single_prev10 <- netAnalyze(net_single_prev10,
                                  centrLCC = TRUE,
                                  clustMethod = "cluster_fast_greedy",
                                  hubPar = "eigenvector",
                                  weightDeg = FALSE, 
                                  normDeg = FALSE,
                                  normBetw = FALSE,
                                  normClose = FALSE,
                                  normEigen = FALSE)
```

```{r single_net_plot_prev10, fig.width=8.5, fig.height=6, dpi=250}
taxa_prev10_in_layout <- colnames(net_single_prev10$adjaMat1)[
  colnames(net_single_prev10$adjaMat1) %in% rownames(layout)
]
layout_prev10 <- layout[taxa_prev10_in_layout, ]

plot(props_single_prev10,
     layout = layout_prev10,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 1.1,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 1, 7))

legend(0.8, 1.1, cex = 1.2, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_prev10_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_prev10,
     layout = layout_prev10,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "10% prevalence filter",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

### Different zero replacement

In this section, we construct the network with the following zero replacement 
methods:

* pseudo count added to all counts (as example for what is not recommended)
* random pseudo counts below the detection limit
* alrEM
* Bayesian-multiplicative replacement

For all four methods, zero replacement is applied externally via
`compute_zero_repl_clr()` before passing the result to `netConstruct()` with
`zeroMethod = "none"`. This ensures the stored `countMat1` in each network
object reflects the actual zero-treated data, so that CLR-based node sizes are
computed from the correct matrix (see analysis of node size behaviour above).

**Minimum relative count and multRepl replacement value for reference**

```{r}
min(counts_rel[counts_rel > 0])
min(counts_rel[counts_rel > 0]) * 0.65
```

We use 2e-5 as pseudo count to have a round number close to the multRepl
replacement value.

#### Prepare zero-replaced matrices

```{r prepare_zero_repl_matrices, warning=FALSE}
zero_repl_pseudo <- compute_zero_repl_clr(
  counts_rel, method = "pseudo", pseudocount = 2e-5, zerosOnly = FALSE
)
counts_rel_pseudo <- zero_repl_pseudo$rel_abund_repl

set.seed(network_seed)
zero_repl_random <- compute_zero_repl_clr(
  counts_rel, method = "pseudo", pseudocount = 2e-5, zerosOnly = TRUE, 
  random = TRUE
)
counts_rel_random <- zero_repl_random$rel_abund_repl

if (file.exists(path_target("counts_rel_alrEM.rds"))) {
  counts_rel_alrEM <- readRDS(path_target("counts_rel_alrEM.rds"))
} else {
  zero_repl_alrEM <- compute_zero_repl_clr(
    counts_rel, method = "alrEM", z.warning = 0.99, z.delete = FALSE, 
    max.iter = 20
  )
  counts_rel_alrEM <- zero_repl_alrEM$rel_abund_repl
  saveRDS(counts_rel_alrEM, file = path_target("counts_rel_alrEM.rds"))
}

if (file.exists(path_target("counts_rel_bayesMult.rds"))) {
  zero_repl_bayesMult <- readRDS(path_target("counts_rel_bayesMult.rds"))
} else {
  zero_repl_bayesMult <- compute_zero_repl_clr(
    counts_rel, method = "bayesMult", z.warning = 0.99, z.delete = FALSE
  )
  counts_rel_bayesMult <- zero_repl_bayesMult$rel_abund_repl
  saveRDS(zero_repl_bayesMult, file = path_target("counts_rel_bayesMult.rds"))
}
```

#### Additive pseudo count

```{r generate_net_single_zero_pseudo}
path_net_single_zero_pseudo <- path_target("net_single_zero_pseudo.rds")

if (file.exists(path_net_single_zero_pseudo) && !refit_complete_network) {
  net_single_zero_pseudo <- readRDS(path_net_single_zero_pseudo)
} else {
  net_single_zero_pseudo <- NetCoMi::netConstruct(
    data = counts_rel_pseudo,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_pseudo, path_net_single_zero_pseudo)
}
```

```{r}
props_single_zero_pseudo <- netAnalyze(net_single_zero_pseudo,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_zero_pseudo, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_pseudo,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_pseudo_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_pseudo,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Additive pseudo count",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Random pseudo counts

```{r generate_net_single_zero_random}
path_net_single_zero_random <- path_target("net_single_zero_random.rds")

if (file.exists(path_net_single_zero_random) && !refit_complete_network) {
  net_single_zero_random <- readRDS(path_net_single_zero_random)
} else {
  net_single_zero_random <- NetCoMi::netConstruct(
    data = counts_rel_random,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_random, path_net_single_zero_random)
}
```

```{r}
props_single_zero_random <- netAnalyze(net_single_zero_random,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_zero_random, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_random,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_random_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_random,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Random pseudo counts (zeros only)",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Modified EM ALR algorithm

```{r generate_net_single_zero_alrEM}
path_net_single_zero_alrEM <- path_target("net_single_zero_alrEM.rds")

if (file.exists(path_net_single_zero_alrEM) && !refit_complete_network) {
  net_single_zero_alrEM <- readRDS(path_net_single_zero_alrEM)
} else {
  net_single_zero_alrEM <- NetCoMi::netConstruct(
    data = counts_rel_alrEM,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_alrEM, path_net_single_zero_alrEM)
}
```

```{r}
props_single_zero_alrEM <- netAnalyze(net_single_zero_alrEM,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_zero_alrEM, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_alrEM,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_alrEM_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_alrEM,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Modified EM alr-algorithm",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Bayesian-multiplicative replacement

```{r generate_net_single_zero_bayesMult}
path_net_single_zero_bayesMult <- path_target("net_single_zero_bayesMult.rds")

if (file.exists(path_net_single_zero_bayesMult) && !refit_complete_network) {
  net_single_zero_bayesMult <- readRDS(path_net_single_zero_bayesMult)
} else {
  net_single_zero_bayesMult <- NetCoMi::netConstruct(
    data = counts_rel_bayesMult,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_bayesMult, path_net_single_zero_bayesMult)
}
```

```{r}
props_single_zero_bayesMult <- netAnalyze(net_single_zero_bayesMult,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_zero_bayesMult, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_bayesMult,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_bayesMult_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_bayesMult,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Bayesian-multiplicative replacement",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

### Different association estimation method

Finally, we will also investigate how the network would look like for different
association estimation methods. Here, we consider the following alternatives:

* Sparse + low-rank decomposition (SPIEC-EASI slr)
* SPRING

#### Sparse + low-rank decomposition (SPIEC-EASI slr)

```{r generate_net_single_slr}
path_net_single_slr <- path_target("net_single_slr.rds")

if (file.exists(path_net_single_slr) && !refit_complete_network) {
  net_single_slr <- readRDS(path_net_single_slr)
} else {
  net_single_slr <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "spieceasi",
    measurePar = list(method = "slr",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_slr, path_net_single_slr)
}
```

```{r}
props_single_slr <- netAnalyze(net_single_slr,
                               centrLCC = TRUE,
                               clustMethod = "cluster_fast_greedy",
                               hubPar = "eigenvector",
                               weightDeg = FALSE, 
                               normDeg = FALSE,
                               normBetw = FALSE,
                               normClose = FALSE,
                               normEigen = FALSE)
```

```{r single_net_plot_slr, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_slr,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.73,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r}
layout_slr <- p$layout$layout1
layout_slr["Streptococcus", ] <- c(-0.1, 0.58)
layout_slr["Escherichia-Shigella", ] <- c(-0.1, -0.05)
layout_slr["Lachnospiraceae_ND3007_group", ] <- c(0.82, -0.28)
```


```{r single_net_plot_slr_ownlay, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_slr,
     layout = layout_slr,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.73,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_slr_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_slr,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPIEC-EASI slr",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))

```

#### SPRING

```{r generate_net_single_spring}
path_net_single_spring <- path_target("net_single_spring.rds")

if (file.exists(path_net_single_spring) && !refit_complete_network) {
  net_single_spring <- readRDS(path_net_single_spring)
} else {
  net_single_spring <- NetCoMi::netConstruct(
    data = counts_raw,
    measure = "spring",
    measurePar = list(nlambda = 10, 
                      rep.num = 10,
                      thresh  = 0.1,
                      Rmethod = "approx"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_spring, path_net_single_spring)
}
```

```{r}
props_single_spring <- netAnalyze(net_single_spring,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_spring, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_spring,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.899,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r}
layout_spring <- p$layout$layout1
layout_spring["Clostridium_sensu_strict1", ] <- c(-0.5, 0.58)
layout_spring["Terrisporobacter", ] <- c(0.43, -0.26)
layout_spring["Stenotrophomonas", ] <- c(-0.15, -0.45)
layout_spring["Streptococcus", ] <- c(0.51, -0.75)
```


```{r single_net_plot_spring_ownlay, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_spring,
     layout = layout_spring,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.899,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_spring_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_spring,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPRING",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Shrinkage proportionality

```{r generate_net_single_rhoshrink_nothresh}
net_single_rhoshrink_nothresh <- NetCoMi::netConstruct(
  data = counts_rel_multrepl,
  measure = "rhoshrink",
  filtTax = "none",
  filtSamp = "none",
  normMethod = "none",
  zeroMethod = "pseudo",
  zeroPar = list(pseudocount = 0),
  sparsMethod = "none",
  seed = 42
)

assomat_single_rhoshrink <- net_single_rhoshrink_nothresh$assoMat1
adjamat_single_rhoshrink <- net_single_rhoshrink_nothresh$adjaMat1
```

```{r single_rhoshrink_histrogram, fig.width=8, fig.height=2}
n_zero_asso <- 
  sum(assomat_single_rhoshrink[lower.tri(assomat_single_rhoshrink)] == 0)

data.frame(correlation = assomat_single_rhoshrink[lower.tri(assomat_single_rhoshrink)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.005) +
  labs(x = "Estimated correlation",
       y = "Count",
       #title = "Estimated correlations after sparsification"
       ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )
```

```{r generate_net_single_rhoshrink}
path_net_single_rhoshrink <- path_target("net_single_rhoshrink.rds")

if (file.exists(path_net_single_rhoshrink) && !refit_complete_network) {
  net_single_rhoshrink <- readRDS(path_net_single_rhoshrink)
} else {
  net_single_rhoshrink <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "rhoshrink",
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "pseudo",
    zeroPar = list(pseudocount = 0),
    sparsMethod = "threshold",
    thresh = 0.2,
    seed = 42
  )
  
  saveRDS(net_single_rhoshrink, path_net_single_rhoshrink)
}
```

```{r}
props_single_rhoshrink <- netAnalyze(net_single_rhoshrink,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```

```{r single_net_plot_rhoshrink, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_rhoshrink,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.899,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_rhoshrink_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_rhoshrink,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Shrinkage proportionality",
     cexTitle = 1.5,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

## Network comparison by EBF duration

### Analysis data set

The group comparison uses the two EBF groups that define the main contrast in
this chapter: no exclusive breastfeeding and exclusive breastfeeding for at
least two months. Samples with one month of exclusive breastfeeding are excluded
from this comparison.

```{r prepare_ebf_groups, warning=FALSE}
# Generate data set for EBF comparison
ebf_keep <- sampdata$`EBF duration` %in% c("0", ">=2")

phylo_genus_ebf <- prune_samples(ebf_keep, phylo_genus)
phylo_genus_ebf <- prune_taxa(taxa_sums(phylo_genus_ebf) > 0, phylo_genus_ebf)

metadata_ebf <- sampdata[ebf_keep, , drop = FALSE] %>%
  mutate(
    ebf_group = factor(
      as.character(`EBF duration`),
      levels = c("0", ">=2"),
      labels = c("0 months", ">=2 months")
    )
  )

counts_ebf_raw <- get_sample_by_taxa(phylo_genus_ebf)
counts_ebf_rel <- counts_to_rel_abund(counts_ebf_raw)

ebf_group <- metadata_ebf$ebf_group

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_ebf_rel,
  frac = zero_replacement_fraction
)

counts_ebf_rel_multrepl <- zero_repl_object$rel_abund_repl
```

```{r ebf_group_overview, eval=FALSE}
metadata_ebf %>%
  count(ebf_group, name = "n_samples")
```

### Network construction

```{r generate_net_ebf_main}
path_net_ebf_main <- path_target("net_ebf_main.rds")

if (file.exists(path_net_ebf_main) && !refit_complete_network) {
  net_ebf_main <- readRDS(path_net_ebf_main)
} else {
  net_ebf_main <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multrepl,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_ebf_main, path_net_ebf_main)
}
```

```{r}
# Extract association matrices
assomat_ebf_main1 <- net_ebf_main$assoMat1
assomat_ebf_main2 <- net_ebf_main$assoMat2
adjamat_ebf_main1 <- net_ebf_main$adjaMat1
adjamat_ebf_main2 <- net_ebf_main$adjaMat2

# Generate graph objectss
graph_ebf_main1 <- transform_asso(assomat_ebf_main1)$graph
graph_ebf_main2 <- transform_asso(assomat_ebf_main2)$graph
```

### Network analysis

#### Association histogram

```{r ebf_association_histrogram, fig.width=8, fig.height=4}


n_zero_asso1 <- sum(assomat_ebf_main1[lower.tri(assomat_ebf_main1)] == 0)
n_zero_asso2 <- sum(assomat_ebf_main2[lower.tri(assomat_ebf_main2)] == 0)

p1 <- data.frame(correlation = assomat_ebf_main1[lower.tri(assomat_ebf_main1)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso1), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "",
       y = "Count",
       ) +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- data.frame(correlation = assomat_ebf_main2[lower.tri(assomat_ebf_main2)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso2), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "Estimated correlation",
       y = "Count",
  ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )

p1 / p2
```

#### Degree distribution

```{r ebf_degree_distribution, fig.width=8, fig.height=4, dpi = 200}
# Compute degree distribution
ddist1 <- igraph::degree.distribution(graph_ebf_main1)
ddist2 <- igraph::degree.distribution(graph_ebf_main2)

# Data frame needed for ggplot2
df1 <- data.frame(
    Degree = (seq_along(ddist1) - 1), Fraction = ddist1
)
df2 <- data.frame(
    Degree = (seq_along(ddist2) - 1), Fraction = ddist2
)
df1[28:40, "Degree"] <- 28:40
df1[28:40, "Fraction"] <- 0

p1 <- ggplot(data = df1, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- ggplot(data = df2, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

p1 / p2
```

#### Analysis with netAnalyze()

```{r}
props_ebf_main <- netAnalyze(net_ebf_main,
                             centrLCC = TRUE,
                             clustMethod = "cluster_fast_greedy",
                             hubPar = "eigenvector",
                             weightDeg = FALSE, 
                             normDeg = FALSE,
                             normBetw = FALSE,
                             normClose = FALSE)
```

### Graphlet correlation matrix

```{r ebf_gcm, fig.width=10, fig.height=9, dpi = 200}
plotGCM(props_ebf_main,
        legendCex = 1.2, 
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

### Network plot

```{r}
plot(props_ebf_main,
     title1 = "EBF = 0", 
     title2 = "EBF >= 2")
```

### Generate permuted association matrices

```{r generate permuted association matrices, eval=FALSE}
permGroupMat <- createAssoPerm(props_ebf_main, 
                               nPerm = 1000, 
                               computeAsso = FALSE, 
                               seed = 42)

nPerm_all <- 1000
blocksize <- 20
repetitions <- nPerm_all / blocksize


cores <- 8

plan(multisession, workers = cores)
registerDoFuture()

with_progress({
  p <- progressr::progressor(along = 1:repetitions)
  
  tmp <- foreach(i = 1:repetitions,
                 .packages = c("NetCoMi")) %dorng% {
    p()
    file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
    bmat_file <- paste0(file_base, ".bmat")
    if (!file.exists(bmat_file)) {
      set.seed(permutation_seed + i)
      NetCoMi::createAssoPerm(props_ebf_main, nPerm = blocksize,
                              permGroupMat = permGroupMat[(i-1) * blocksize +
                                                            1:blocksize, ],
                              computeAsso = TRUE,
                              fileStoreAssoPerm = file_base,
                              storeCountsPerm = FALSE, append = FALSE)
    }
  }
})

# Reset to sequential execution
plan(sequential)
```


```{r, eval=FALSE}
# Combine the matrices and store them into a new file (because netCompare() 
# needs an external file)
assoPerm_all <- NULL

for (i in 1:repetitions) {
  file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
  
  assoPerm_tmp <- fm.open(filenamebase = file_base, readonly = TRUE)
  
  assoPerm_all <- rbind(assoPerm_all, as.matrix(assoPerm_tmp))
  
  close(assoPerm_tmp)
}

dim(assoPerm_all)
```
The combined matrix has 117000 rows and 234 columns, which is plausible because:

- n_rows = nPerm_all * n_taxa = 1000 * 117
- n_cols = n_groups * n_taxa * 2 * 117

We performed 1000 permutations, but want only 999 to get a round minimum possible
p-value. We therefore reduce the matrix to 999 * 117 = 116883 rows.

```{r, eval=FALSE}
# Keep the number of stored permutations consistent with netCompare().
n_perm_netcompare <- 999
n_taxa_netcompare <- nrow(props_ebf_main$input$adjaMat1)
n_rows_netcompare <- n_perm_netcompare * n_taxa_netcompare

stopifnot(nrow(assoPerm_all) >= n_rows_netcompare)
stopifnot(nrow(assoPerm_all) %% n_taxa_netcompare == 0)

assoPerm_netcompare <- assoPerm_all[seq_len(n_rows_netcompare), , drop = FALSE]
dim(assoPerm_netcompare)
```

```{r, eval=FALSE}
# Combine the permutation association matrices
fm.create.from.matrix(filenamebase = path_target("assoPerm_ebf"), 
                      mat = assoPerm_netcompare)
# >116883 x 234 filematrix with 8 byte "double" elements 
```

### Differential network analysis

We start by comparing local and global network characteristics between the two 
groups. This is done with NetCoMi' `netCompare()` function.

The statistical tests will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_netcompare_noBH}
path_comp_ebf_main_noBH <- path_target("comp_ebf_main_noBH.rds")

if (file.exists(path_comp_ebf_main_noBH) && !refit_complete_network) {
  comp_ebf_main_noBH <- readRDS(path_comp_ebf_main_noBH)
} else {
  comp_ebf_main_noBH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_noBH, path_comp_ebf_main_noBH)
}

print(summary(comp_ebf_main_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))

summary(pvals_comp_main_noBH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

```{r run_netcompare_BH}
path_comp_ebf_main_BH <- path_target("comp_ebf_main_BH.rds")

if (file.exists(path_comp_ebf_main_BH) && !refit_complete_network) {
  comp_ebf_main_BH <- readRDS(path_comp_ebf_main_BH)
} else {
  comp_ebf_main_BH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_BH, path_comp_ebf_main_BH)
}

print(summary(comp_ebf_main_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

#### permApprox refinement without constraint

```{r run_netcompare_GPD_noConstr_noBH}
path_comp_ebf_GPD_noConstr_noBH <- path_target("comp_ebf_GPD_noConstr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_noBH <- readRDS(path_comp_ebf_GPD_noConstr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )
  comp_ebf_GPD_noConstr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_noBH, path_comp_ebf_GPD_noConstr_noBH)
}

tmp <- data.frame(emp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrEmpirical$pvalDiffDeg,
                  pApp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrRefined$pvalDiffDeg)

print(summary(comp_ebf_GPD_noConstr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_noConstr_BH}
path_comp_ebf_GPD_noConstr_BH <- path_target("comp_ebf_GPD_noConstr_BH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_BH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_BH <- readRDS(path_comp_ebf_GPD_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  comp_ebf_GPD_noConstr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_BH, path_comp_ebf_GPD_noConstr_BH)
}

print(summary(comp_ebf_GPD_noConstr_BH, 
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"), 
              numbNodes = 5))

summary(comp_ebf_GPD_noConstr_BH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

#### permApprox refinement with constraint

```{r run_netcompare_GPD_constr_noBH}
path_comp_ebf_GPD_constr_noBH <- path_target("comp_ebf_GPD_constr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_constr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_constr_noBH <- readRDS(path_comp_ebf_GPD_constr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  comp_ebf_GPD_constr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_noBH, path_comp_ebf_GPD_constr_noBH)
}

print(summary(comp_ebf_GPD_constr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_constr_BH}
path_comp_ebf_GPD_constr_BH <- path_target("comp_ebf_GPD_constr_BH.rds")

if (file.exists(path_comp_ebf_GPD_constr_BH) && !refit_complete_network) {
  comp_ebf_GPD_constr_BH <- readRDS(path_comp_ebf_GPD_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  comp_ebf_GPD_constr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_BH, path_comp_ebf_GPD_constr_BH)
}

print(summary(comp_ebf_GPD_constr_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

### Differential association analysis

In this step we compare the associations itself by generating and plotting differential networks, which have an edge for differentially associated 
genera only.

The networks will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_diffnet_noBH}
path_diff_ebf_main_noBH <- path_target("diff_ebf_main_noBH.rds")

if (file.exists(path_diff_ebf_main_noBH) && !refit_complete_network) {
  diff_ebf_main_noBH <- readRDS(path_diff_ebf_main_noBH)
} else {
  diff_ebf_main_noBH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_noBH, path_diff_ebf_main_noBH)
}

summary(diff_ebf_main_noBH$pvalsVec)
```

```{r run_diffnet_BH}
path_diff_ebf_main_BH <- path_target("diff_ebf_main_BH.rds")

if (file.exists(path_diff_ebf_main_BH) && !refit_complete_network) {
  diff_ebf_main_BH <- readRDS(path_diff_ebf_main_BH)
} else {
  diff_ebf_main_BH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_BH, path_diff_ebf_main_BH)
}

summary(diff_ebf_main_BH$pAdjustVec)
```

#### permApprox refinement without constraint

```{r run_diffnet_noConstr_noBH}
path_diff_ebf_noConstr_noBH <- path_target("diff_ebf_noConstr_noBH.rds")

if (file.exists(path_diff_ebf_noConstr_noBH) && !refit_complete_network) {
  diff_ebf_noConstr_noBH <- readRDS(path_diff_ebf_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  diff_ebf_noConstr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_noConstr_noBH, path_diff_ebf_noConstr_noBH)
}

summary(diff_ebf_noConstr_noBH$pvalsVec)
```

```{r run_diffnet_noConstr_BH}
path_diff_ebf_noConstr_BH <- path_target("diff_ebf_noConstr_BH.rds")

if (file.exists(path_diff_ebf_noConstr_BH) && !refit_complete_network) {
  diff_ebf_noConstr_BH <- readRDS(path_diff_ebf_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )
  
  diff_ebf_noConstr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_noConstr_BH, path_diff_ebf_noConstr_BH)
}

summary(diff_ebf_noConstr_BH$pAdjustVec)
```

#### permApprox refinement with constraint

```{r run_diffnet_constr_noBH}
path_diff_ebf_constr_noBH <- path_target("diff_ebf_constr_noBH.rds")

if (file.exists(path_diff_ebf_constr_noBH) && !refit_complete_network) {
  diff_ebf_constr_noBH <- readRDS(path_diff_ebf_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )

  diff_ebf_constr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_constr_noBH, path_diff_ebf_constr_noBH)
}

summary(diff_ebf_constr_noBH$pvalsVec)
```

```{r run_diffnet_constr_BH}
path_diff_ebf_constr_BH <- path_target("diff_ebf_constr_BH.rds")

if (file.exists(path_diff_ebf_constr_BH) && !refit_complete_network) {
  diff_ebf_constr_BH <- readRDS(path_diff_ebf_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  diff_ebf_constr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_constr_BH, path_diff_ebf_constr_BH)
}

summary(diff_ebf_constr_BH$pAdjustVec)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
