---
title: "Network analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
always_allow_html: true
params:
  name: "10_network_analysis" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "147f8b0c-cb4b-427a-9549-8c8b8e3ea46b")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/net_"),
  echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
# SpiecEasi from latentcor branch needed
if (as.character(utils::packageVersion("SpiecEasi")) != "2.0.1") {
  remotes::install_github("zdk123/SpiecEasi", ref = "latentcor")
}

library(conflicted)
library(phyloseq)
library(tidyverse)
library(NetCoMi)
library(permApprox)
library(igraph)
library(kableExtra)
library(patchwork)
library(ggplot2)

# For network comparison
library(foreach)
library(future)
library(doFuture)
library(doRNG)
library(progressr)
library(filematrix)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

# Load global functions
application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = FALSE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Aim

This script contains the network analyses for the application chapter. The first
part builds one complete network on the filtered genus-level data and evaluates
how sensitive the resulting network is to common preprocessing and association
choices. The complete-network analysis will compare the standard preprocessing
approach with the final method used in this thesis, and then vary one component
at a time: zero replacement, normalization, and association measure.

The second part will compare networks between the two main exclusive
breastfeeding groups used throughout this chapter, `EBF duration = 0` and
`EBF duration >=2`. The network-comparison analysis will contrast the classical
empirical permutation results with `permApprox`-refined p-values. For the
`permApprox` analysis we will fit both the unconstrained approximation and the
proposed constrained version, so that the unconstrained fit can be used to
inspect how many zero-valued permutation statistics would be expected without
the constraint.

## Set global parameters

```{r parameters}
zero_replacement_fraction <- 0.65
network_seed <- 42
permutation_seed <- 42
n_perm_network_comparison <- 999
cores <- 10
n_abundant_network_labels <- 15
n_central_network_labels <- 10

association_measure_final <- "spieceasi"
normalization_final <- "clr"
zero_replacement_final <- "multRepl"

refit_complete_network <- FALSE
refit_method_comparison <- FALSE
refit_network_comparison <- FALSE
```

## Load data

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))
```

### Phyloseq object on genus level

```{r}
phylo_genus
```

## Helper functions

```{r functions}
## Function to turn an association matrix into a graph object
# Arguments:
# - assoMat: association matrix
# - threshold: associations below the threshold are set to zero
# - dissTrans: dissimilarity transformation ("signed" or "unsigned")

transform_asso <- function(assoMat, thresh = NULL, dissTrans = "signed") {
    # Sparsification
    if (!is.null(thresh)) {
        assoMat[abs(assoMat) < thresh] <- 0
    }

    # Compute dissimilarity matrix
    if (dissTrans == "signed") {
        dissMat <- sqrt(0.5 * (1 - assoMat))
    } else {
        dissMat <- sqrt(1 - assoMat^2)
    }

    # Dissimilarity between nodes with zero correlation is set to 1
    # (these nodes are unconnected and thus should have maximum dissimilarity)
    dissMat[assoMat == 0] <- 1

    # Compute similarity matrix
    simMat <- 1 - dissMat

    # Turn into igraph object
    graphObj <- SpiecEasi::adj2igraph(simMat)

    return(list(graph = graphObj, adja = simMat, asso = assoMat, diss = dissMat))
}

make_network_plot_labels <- function(
    props,
    top_abundant_taxa,
    da_taxa = NULL,
    dd_taxa = NULL,
    n_abundant = 10,
    n_central = 10,
    extra_taxa = NULL,
    label_overrides = NULL) {
  
  adja <- props$input$adjaMat1
  diag(adja) <- 0
  network_taxa <- colnames(adja)
  
  non_singleton_taxa <- names(which(
    rowSums(adja != 0) > 0 |
      colSums(adja != 0) > 0
  ))
  
  abundant_taxa <- top_abundant_taxa[seq_len(
    min(max(n_abundant, 0), length(top_abundant_taxa))
  )]
  
  centrality <- props$centralities$eigenv1
  if (is.null(centrality) || n_central <= 0) {
    central_taxa <- character(0)
  } else {
    centrality <- centrality[is.finite(centrality)]
    central_taxa <- centrality %>%
      sort(decreasing = TRUE) %>%
      head(n_central) %>%
      names() %>%
      (\(x) setNames(x, x))()
  }
  
  taxa_to_label <- c(abundant_taxa, da_taxa, dd_taxa, central_taxa, extra_taxa)
  taxa_to_label <- taxa_to_label[!duplicated(names(taxa_to_label))]
  
  taxa_to_label_in_network <- intersect(names(taxa_to_label), network_taxa)
  
  taxa_to_plot <- unique(c(non_singleton_taxa, taxa_to_label_in_network))
  
  network_labels <- setNames(rep("", length(network_taxa)), network_taxa)
  network_labels[taxa_to_label_in_network] <- taxa_to_label[taxa_to_label_in_network]
  
  if (!is.null(label_overrides)) {
    label_override_taxa <- intersect(names(label_overrides), names(network_labels))
    network_labels[label_override_taxa] <- label_overrides[label_override_taxa]
  }
  
  list(
    labels = network_labels,
    taxa_to_plot = taxa_to_plot,
    taxa_to_label = taxa_to_label,
    taxa_to_label_in_network = taxa_to_label_in_network,
    non_singleton_taxa = non_singleton_taxa
  )
}

extract_network_edges <- function(network, method) {
  adja <- network$adjaMat1
  diag(adja) <- 0
  
  edge_index <- which(lower.tri(adja), arr.ind = TRUE)
  taxon1 <- rownames(adja)[edge_index[, "row"]]
  taxon2 <- colnames(adja)[edge_index[, "col"]]
  edge_ids <- paste(
    pmin(taxon1, taxon2),
    pmax(taxon1, taxon2),
    sep = " -- "
  )
  
  tibble(
    edge_id = edge_ids,
    method = method,
    present = adja[edge_index] != 0
  ) %>%
    filter(present) %>%
    select(edge_id, method)
}

make_edge_presence <- function(networks) {
  edge_long <- purrr::imap_dfr(networks, extract_network_edges)
  
  edge_long %>%
    mutate(present = TRUE) %>%
    tidyr::pivot_wider(
      names_from = method,
      values_from = present,
      values_fill = FALSE
    )
}

format_taxon_label <- function(taxon, label_overrides = network_label_overrides) {
  dplyr::coalesce(label_overrides[taxon], taxon) %>%
    unname()
}

format_edge_label <- function(edge_id, label_overrides = network_label_overrides,
                              sep = " - ") {
  edge_taxa <- stringr::str_split_fixed(edge_id, " -- ", 2)
  paste(
    format_taxon_label(edge_taxa[, 1], label_overrides),
    format_taxon_label(edge_taxa[, 2], label_overrides),
    sep = sep
  )
}

extract_network_associations <- function(network, method) {
  adja <- network$adjaMat1
  diag(adja) <- 0
  
  asso <- network$assoMat1
  diag(asso) <- NA_real_
  
  edge_index <- which(lower.tri(adja), arr.ind = TRUE)
  taxon1 <- rownames(asso)[edge_index[, "row"]]
  taxon2 <- colnames(asso)[edge_index[, "col"]]
  edge_ids <- paste(
    pmin(taxon1, taxon2),
    pmax(taxon1, taxon2),
    sep = " -- "
  )
  
  tibble(
    edge_id = edge_ids,
    method = method,
    present = adja[edge_index] != 0,
    association = asso[edge_index]
  ) %>%
    filter(present) %>%
    select(edge_id, method, association)
}

plot_shared_edges <- function(edge_presence, networks, n_edges = Inf, 
                              text_size = 3.4) {
  method_cols <- setdiff(colnames(edge_presence), "edge_id")
  edge_membership <- edge_presence[, method_cols, drop = FALSE]
  edge_membership[] <- lapply(edge_membership, as.logical)
  
  shared_edges <- edge_presence$edge_id[
    rowSums(edge_membership, na.rm = TRUE) == length(method_cols)
  ]
  
  n_edges <- min(n_edges, length(shared_edges))
  
  if (length(shared_edges) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0.5, hjust = 0,
                 label = "No edges shared\nby all methods", size = text_size) +
        theme_void()
    )
  }
  
  shared_edge_summary <- purrr::imap_dfr(networks, 
                                         extract_network_associations) %>%
    filter(edge_id %in% shared_edges) %>%
    group_by(edge_id) %>%
    filter(n_distinct(method) == length(method_cols)) %>%
    summarise(
      mean_association = mean(association, na.rm = TRUE),
      min_association = min(association, na.rm = TRUE),
      max_association = max(association, na.rm = TRUE),
      mean_abs_association = mean(abs(association), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_abs_association)) %>%
    slice_head(n = n_edges) %>%
    mutate(
      edge_label = stringr::str_trunc(
        format_edge_label(edge_id),
        width = 30
      ),
      association_label = sprintf(
        "%+.2f [%+.2f, %+.2f]",
        mean_association,
        min_association,
        max_association
      ),
      y = rev(row_number()) * 0.68
    )
  
  title_text <- ifelse(length(shared_edges) == n_edges, 
                       paste0("Shared by all (n = ", length(shared_edges), ")"),
                       paste0("Shared by all (n = ", length(shared_edges),
                              "; shown = ", n_edges, ")")
                       )
  
  ggplot(shared_edge_summary) +
    annotate("text", x = 0, y = max(shared_edge_summary$y) + 0.85,
             hjust = 0, fontface = "bold", size = text_size + 0.4,
             label = title_text) +
    annotate("text", x = 0, y = max(shared_edge_summary$y) + 0.42,
             hjust = 0, size = text_size - 0.4,
             label = "mean association [range]") +
    geom_text(aes(x = 0, y = y, label = edge_label),
              hjust = 0, size = text_size) +
    geom_text(aes(x = 1.45, y = y, label = association_label),
              hjust = 1, size = text_size) +
    coord_cartesian(xlim = c(0, 1.45), clip = "off") +
    theme_void() +
    theme(plot.margin = margin(5.5, 5.5, 5.5, 0))
}

summarise_persistent_edges <- function(networks) {
  edge_presence <- make_edge_presence(networks)
  method_cols <- setdiff(colnames(edge_presence), "edge_id")
  edge_membership <- edge_presence[, method_cols, drop = FALSE]
  edge_membership[] <- lapply(edge_membership, as.logical)
  
  persistent_edge_ids <- edge_presence$edge_id[
    rowSums(edge_membership, na.rm = TRUE) == length(method_cols)
  ]
  
  purrr::imap_dfr(networks, extract_network_associations) %>%
    filter(edge_id %in% persistent_edge_ids) %>%
    tidyr::separate(edge_id, into = c("taxon1", "taxon2"), 
                    sep = " -- ", remove = FALSE) %>%
    group_by(edge_id, taxon1, taxon2) %>%
    filter(n_distinct(method) == length(method_cols)) %>%
    summarise(
      n_methods = n_distinct(method),
      mean_association = mean(association, na.rm = TRUE),
      min_association = min(association, na.rm = TRUE),
      max_association = max(association, na.rm = TRUE),
      mean_abs_association = mean(abs(association), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      taxon1_label = format_taxon_label(taxon1),
      taxon2_label = format_taxon_label(taxon2)
    ) %>%
    arrange(desc(mean_abs_association), taxon1, taxon2)
}

make_persistent_edge_network <- function(network, persistent_edges) {
  adja <- network$adjaMat1
  asso <- network$assoMat1
  diag(adja) <- 0
  diag(asso) <- 0
  
  edge_index <- which(lower.tri(adja), arr.ind = TRUE)
  taxon1 <- rownames(adja)[edge_index[, "row"]]
  taxon2 <- colnames(adja)[edge_index[, "col"]]
  edge_ids <- paste(
    pmin(taxon1, taxon2),
    pmax(taxon1, taxon2),
    sep = " -- "
  )
  
  keep_edge <- edge_ids %in% persistent_edges$edge_id
  persistent_adja <- adja * 0
  persistent_asso <- asso * 0
  
  persistent_adja[edge_index[keep_edge, , drop = FALSE]] <- 
    adja[edge_index[keep_edge, , drop = FALSE]]
  persistent_asso[edge_index[keep_edge, , drop = FALSE]] <- 
    asso[edge_index[keep_edge, , drop = FALSE]]
  
  persistent_adja <- persistent_adja + t(persistent_adja)
  persistent_asso <- persistent_asso + t(persistent_asso)
  
  network$adjaMat1 <- persistent_adja
  network$assoMat1 <- persistent_asso
  network
}

plot_edge_upset <- function(edge_presence, title, networks = NULL,
                            n_intersections = 20, n_shared_edges = Inf,
                            shared_text_size = 3.4) {
  method_cols <- setdiff(colnames(edge_presence), "edge_id")
  
  edge_patterns <- edge_presence %>%
    mutate(
      intersection = apply(
        across(all_of(method_cols)),
        1,
        \(x) paste(method_cols[as.logical(x)], collapse = " & ")
      )
    ) %>%
    filter(intersection != "") %>%
    count(intersection, name = "n_edges", sort = TRUE) %>%
    slice_head(n = n_intersections) %>%
    mutate(
      intersection_id = row_number(),
      intersection_id = factor(intersection_id, levels = intersection_id)
    )
  
  matrix_data <- edge_patterns %>%
    select(intersection, intersection_id) %>%
    tidyr::separate_longer_delim(intersection, delim = " & ") %>%
    rename(method = intersection) %>%
    mutate(active = TRUE) %>%
    tidyr::complete(
      intersection_id,
      method = method_cols,
      fill = list(active = FALSE)
    ) %>%
    mutate(method = factor(method, levels = rev(method_cols)))
  
  set_sizes <- edge_presence %>%
    summarise(across(all_of(method_cols), sum)) %>%
    tidyr::pivot_longer(
      cols = everything(),
      names_to = "method",
      values_to = "n_edges"
    )

  active_line_data <- matrix_data %>%
    filter(active) %>%
    add_count(intersection_id) %>%
    filter(n > 1)
  
  intersection_plot <- ggplot(edge_patterns, aes(x = intersection_id, 
                                                 y = n_edges)) +
    geom_col(width = 0.75, fill = "gray25") +
    geom_text(
      aes(label = n_edges),
      vjust = -0.35,
      size = 4
    ) +
    labs(title = title, x = NULL, y = "Edges") +
    expand_limits(y = max(edge_patterns$n_edges) * 1.08) +
    theme_bw() +
    theme(
      axis.title.y = element_text(margin = margin(r = -35), size = 14),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size = 12),
      axis.ticks.x = element_blank(),
      plot.title = element_text(face = "bold", size = 14)
    )
  
  matrix_plot <- ggplot(matrix_data, aes(x = intersection_id, y = method)) +
    geom_point(
      data = filter(matrix_data, !active),
      color = "gray85",
      size = 2.2
    ) +
    geom_line(
      data = active_line_data,
      aes(group = intersection_id),
      color = "gray25",
      linewidth = 0.5
    ) +
    geom_point(
      data = filter(matrix_data, active),
      color = "gray25",
      size = 2.8
    ) +
    scale_x_discrete(limits = levels(edge_patterns$intersection_id)) +
    labs(x = "Edge-set intersection", y = NULL) +
    theme_bw() +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 14),
      axis.ticks.x = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  set_size_plot <- ggplot(set_sizes,
                          aes(x = n_edges, 
                              y = factor(method, 
                                         levels = rev(method_cols)))) +
    geom_col(width = 0.7, fill = "gray25") +
    geom_text(
      aes(label = n_edges),
      hjust = -0.25,
      size = 4
    ) +
    labs(x = "Total edges", y = NULL) +
    expand_limits(x = max(set_sizes$n_edges) * 1.12) +
    theme_bw() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_blank(),
      axis.title.x = element_text(size = 14),
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
    
  shared_edge_plot <- if (is.null(networks)) {
    patchwork::plot_spacer()
  } else {
    plot_shared_edges(
      edge_presence = edge_presence,
      networks = networks,
      n_edges = n_shared_edges,
      text_size = shared_text_size
    )
  }
  
  (
    (intersection_plot | shared_edge_plot) /
      (matrix_plot | set_size_plot)
  ) +
    plot_layout(
      widths = c(4, 1),
      heights = c(3, 2)
    )
}
```


## Prepare data

The network analyses use the same filtered genus-level object as the
differential abundance and differential distribution analyses. Counts are
represented as a samples-by-taxa matrix. Relative abundances are prepared here
because several network workflows use them directly. 

Since SpiecEasi does only provide pseudo count zero replacement, we need to 
apply zero replacement with multRepl beforehand. 

```{r prepare_matrices, warning=FALSE}
counts_raw <- get_sample_by_taxa(phylo_genus)

sampdata_net <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw)) %>%
  arrange(match(SampleID, rownames(counts_raw)))

counts_raw <- counts_raw[sampdata_net$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_raw), sampdata_net$SampleID))

counts_rel <- counts_to_rel_abund(counts_raw)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel,
  frac = zero_replacement_fraction
)

counts_rel_multrepl <- zero_repl_object$rel_abund_repl

net_preprocessing_summary <- tibble(
  n_samples = nrow(counts_raw),
  n_taxa = ncol(counts_raw),
  min_library_size = min(rowSums(counts_raw)),
  median_library_size = median(rowSums(counts_raw)),
  max_library_size = max(rowSums(counts_raw)),
  zero_fraction_orig = mean(counts_raw == 0),
  zero_fraction_repl = mean(counts_rel_multrepl == 0)
)

net_preprocessing_summary
```

Value imputed by multiplicative replacement: `r min(counts_rel[counts_rel > 0]) * 0.65`.

Rename taxonomic table and make rank Genus unique.

```{r}
phylo_genus <- renameTaxa(phylo_genus, 
                          pat = "<name>", 
                          substPat = "<name>_<subst_name>(<subst_R>)",
                          numDupli = "Genus")
```

## Define taxa to label

In the respective network plots, the following taxa are labelled: 

* Highly abundant taxa
* Taxa that appeared in the differential association analysis as well as the differential distribution analysis
* Most central genera with respect to eigenvector centrality in the respective network
* The three genera with a negative correlation to Bifidobacterium in the baseline network

```{r}
# Most abundant taxa
genus_mat <- as(t(otu_table(phylo_genus)), "matrix")

genus_rel_mat <- genus_mat / rowSums(genus_mat)

genus_tax <- tax_table(phylo_genus) %>% 
  as("matrix") %>% 
  as.data.frame() %>% 
  rownames_to_column("taxon_id")

genus_labels <- genus_tax %>%
  transmute(
    taxon_id,
    genus = if_else(
      is.na(Genus) | Genus == "",
      taxon_id,
      Genus
    )
  )

top_genera_all <- tibble(
  taxon_id = colnames(genus_mat),
  mean_rel_abundance = colMeans(genus_rel_mat, na.rm = TRUE) * 100,
  prevalence = colMeans(genus_mat > 0) * 100
) %>%
  left_join(genus_labels, by = "taxon_id") %>%
  arrange(desc(mean_rel_abundance))

top_abundant_taxa <- top_genera_all$genus
names(top_abundant_taxa) <- top_abundant_taxa

## Taxa labelled in differential abundance plot
da_volcano_df <- readr::read_csv(
  path_source("08_diff_abundance", "diff_abundance_volcano_plot_data.csv"),
  show_col_types = FALSE
)

da_plot_taxa <- da_volcano_df %>%
  filter(p_plot_raw < 1e-5) %>%
  distinct(taxon_id, taxon) %>%
  { setNames(.$taxon, .$taxon_id) }


## Taxa labelled in differential distribution plot
alpha_diff_distribution <- 0.01
two_stage_component_label_n <- 6
complete_clr_volcano_label_n <- 6

dd_results_all <- readr::read_csv(
  path_source("09_diff_distribution", "diff_distribution_results_all.csv"),
  show_col_types = FALSE
)

dd_results_complete_all <- readr::read_csv(
  path_source("09_diff_distribution", "diff_distribution_complete_clr_results_all.csv"),
  show_col_types = FALSE
)

dd_two_stage_taxa <- dplyr::bind_rows(
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Non-zero distribution",
      native_statistic = `d.wass^2`,
      pvalue = p_permapprox_constrained,
      qvalue = p_adj_permapprox_constrained
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Zero prevalence",
      native_statistic = abs(zero_prevalence_z),
      pvalue = p.zero,
      qvalue = p.adj.zero
    ),
  dd_results_all %>%
    transmute(
      taxon_id,
      taxon,
      component = "Combined",
      native_statistic = fisher_statistic_permapprox_constrained,
      pvalue = p_combined_permapprox_constrained,
      qvalue = p_adj_combined_permapprox_constrained
    )
) %>%
  filter(is.finite(native_statistic), is.finite(pvalue), 
         qvalue < alpha_diff_distribution) %>%
  group_by(component) %>%
  arrange(qvalue, desc(-log10(pvalue)), .by_group = TRUE) %>%
  slice_head(n = two_stage_component_label_n) %>%
  ungroup() %>%
  distinct(taxon_id, taxon)

dd_one_stage_taxa <- dd_results_complete_all %>%
  filter(
    is.finite(`d.wass^2`),
    is.finite(p_permapprox_constrained),
    p_adj_permapprox_constrained < alpha_diff_distribution
  ) %>%
  arrange(p_adj_permapprox_constrained, 
          desc(-log10(p_permapprox_constrained))) %>%
  slice_head(n = complete_clr_volcano_label_n) %>%
  distinct(taxon_id, taxon)

dd_plot_taxa <- bind_rows(dd_two_stage_taxa, dd_one_stage_taxa) %>%
  distinct(taxon_id, taxon) %>%
  { setNames(.$taxon, .$taxon_id) }

network_label_overrides <- c(
  "8_Comamonadaceae(F)" = "Comamonadaceae(F)",
  "24_Enterobacteriaceae(F)" = "Enterobacteriaceae(F)"
)
```


## Baseline network with standard preprocessing

This section will estimate the complete genus-level network using the final
preprocessing and association pipeline selected for the thesis. The resulting
object should be saved to the target directory and reused by downstream summary
tables and figures.

The preprocessing approach used throughout the thesis is:

1. Transform counts to relative abundances
2. Perform multiplicative zero replacement
3. Perform CLR transformation

### Network construction

```{r generate_net_single_main}
path_net_single_main <- path_target("net_single_main.rds")

if (file.exists(path_net_single_main) && !refit_complete_network) {
  net_single_main <- readRDS(path_net_single_main)
} else {
  net_single_main <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_main, path_net_single_main)
}
```

```{r}
# Extract association matrix
assomat_single_main <- net_single_main$assoMat1
adjamat_single_main <- net_single_main$adjaMat1

# Generate graph object
graph_single_main <- transform_asso(assomat_single_main)$graph
```

### Network analysis

#### Association heatmap

**For all taxa**

```{r single_association_heatmap, fig.width=20, fig.height=20}
plotHeat(assomat_single_main, textUpp = "none", textLow = "none")
```

**For non-singletons only**

```{r single_association_heatmap_nonzero, fig.width=15, fig.height=15}
adjamat_single_main_nonzero <- net_single_main$adjaMat1
diag(adjamat_single_main_nonzero) <- 0

taxa_all <- colnames(assomat_single_main)
taxa_with_associations <- taxa_all[rowSums(adjamat_single_main_nonzero) > 0]
assomat_single_main_no_singletons <- 
  assomat_single_main[taxa_with_associations, taxa_with_associations]

plotHeat(
  assomat_single_main_no_singletons,
  textUpp = "none",
  textLow = "none"
)
```

#### Association histogram

```{r single_association_histrogram, fig.width=8, fig.height=2}
n_zero_asso <- sum(assomat_single_main[lower.tri(assomat_single_main)] == 0)

data.frame(correlation = assomat_single_main[lower.tri(assomat_single_main)]) %>%
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.005) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso), size = 4) +
  coord_cartesian(ylim = c(0, 50)) +
  labs(x = "Estimated correlation",
       y = "Count",
       #title = "Estimated correlations after sparsification"
       ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )
```

```{r}
n_asso <- sum(lower.tri(assomat_single_main))
cat("Entries in the lower triangle:", n_asso)

cat("\nNon-zero entries in the lower triangle:", n_asso - n_zero_asso)
```

#### Degree distribution

```{r single_degree_distribution, fig.width=8, fig.height=2, dpi = 200}
# Compute degree distribution
ddist <- igraph::degree.distribution(graph_single_main)

# Data frame needed for ggplot2
df <- data.frame(
    Degree = as.factor((seq_along(ddist)) - 1), Fraction = ddist
)

ggplot(data = df, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))
```

#### Analysis with netAnalyze()

```{r}
props_single_main <- netAnalyze(net_single_main,
                                centrLCC = TRUE,
                                clustMethod = "cluster_fast_greedy",
                                hubPar = "eigenvector",
                                weightDeg = FALSE, 
                                normDeg = FALSE,
                                normBetw = FALSE,
                                normClose = FALSE,
                                normEigen = FALSE)
```

```{r}
print(summary(props_single_main))
```

### Graphlet correlation matrix

```{r single_gcm, fig.width=6, fig.height=6, dpi = 200}
plotGCM(props_single_main,
        legendCex = 1, 
        labCex = 1.1,
        sigTextCex = 1.2,
        titles = "",
        mar = c(1,1,0,1))
```


### Network plot

In the network plot, the following taxa are labelled: 

* Highly abundant genera
* Genera that appeared in the differential association analysis as well as the differential distribution analysis.
* Most central genera with respect to eigenvector centrality.
* The three genera with a negative correlation to Bifidobacterium

```{r single_taxa_to_label}
single_main_plot_labels <- make_network_plot_labels(
  props = props_single_main,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 10,
  n_central = 10,
  #extra_taxa = bif_neg_taxa,
  label_overrides = network_label_overrides
)

taxa_to_label <- single_main_plot_labels$taxa_to_label
taxa_to_plot <- single_main_plot_labels$taxa_to_plot
network_labels <- single_main_plot_labels$labels
```

#### Node colors: clusters; node size: eigenvector centrality

```{r single_net_plot_basic, fig.width=7, fig.height=6, dpi=250}
clustcols <- c("#E69F00", "#CC79A7", "#F0E442", "#56B4E9", "#0072B2", 
               "#009E73",  "#D55E00", "#33A02C", "#E69F00", "#CC79A7",
               "#F0E442", "#56B4E9", "#0072B2", "#009E73",  "#D55E00")

set.seed(42)

p <- plot(props_single_main,
          nodeFilter = "names",
          nodeFilterPar = taxa_to_plot,
          labelScale = FALSE,
          labels = network_labels,
          repulsion = 0.899999,
          nodeColor = "cluster", 
          colorVec = clustcols,
          nodeTransp = 40,
          nodeSize = "eigenvector",
          nodeSizeSpread = 3,
          hubBorderCol = "gray40",
          posCol = "gray50", 
          negCol = "red3",
          showTitle = FALSE,
          cexLabels = 0.8,
          mar = c(1, 1, 1, 1))
```

We store this layout as main layout.

```{r}
layout <- p$layout$layout1
```


#### Nodes colored by phyla

```{r}
# Get phyla names
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]
table(phyla)

# Define phylum colors
phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)
```

```{r single_net_plot_col_phyla, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_main,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

**Add labels or further highly abundant taxa and improve layout**

```{r}
# Taxa negatively correlated with Bifidobacterium
bif_neg_taxa <- assomat_single_main["Bifidobacterium", ] %>%
  (\(x) x[x < 0])() %>%
  names() %>%
  (\(x) setNames(x, x))()

single_main_plot_labels <- make_network_plot_labels(
  props = props_single_main,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 15,
  n_central = 10,
  extra_taxa = bif_neg_taxa,
  label_overrides = network_label_overrides
)

taxa_to_label <- single_main_plot_labels$taxa_to_label
taxa_to_plot <- single_main_plot_labels$taxa_to_plot
network_labels <- single_main_plot_labels$labels
```

```{r}
# Improve layout
layout["[Ruminococcus]_gnavugroup", ] <- c(-0.7,  -0.55)
layout["Pseudomonas", ] <- c(-0.1, -0.22)
layout["Blastococcus", ] <- c(-0.33, 0.13)
layout["8_Comamonadaceae(F)", ] <- c(-0.2, -0.1)
layout["Haemophilus", ] <- c(0.84, -0.5)
layout["Lachnospira", ] <- c(-0.34, -0.05)
```


```{r single_net_plot_baseline, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_main,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```


```{r single_net_plot_baseline_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_main,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Baseline network",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

## Alternative preprocessing steps

In this section we analyze, how the network changes when alternative 
preprocessing steps like filtering or zero replacement, or a different 
association measure is used. 

### Stronger filtering

In the network above, a prevalence filter of 1% was used. We now increase this 
to 5% and 10%. Everything else stays the same.

#### 5% prevalence filter

```{r, warning=FALSE}
phylo_genus_prev05 <- 
  readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev05.rds"))

counts_raw_prev05 <- get_sample_by_taxa(phylo_genus_prev05)

sampdata_net_prev05 <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw_prev05)) %>%
  arrange(match(SampleID, rownames(counts_raw_prev05)))

counts_raw_prev05 <- counts_raw_prev05[sampdata_net_prev05$SampleID, , 
                                       drop = FALSE]

stopifnot(identical(rownames(counts_raw_prev05), sampdata_net_prev05$SampleID))

counts_rel_prev05 <- counts_to_rel_abund(counts_raw_prev05)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel_prev05,
  frac = zero_replacement_fraction
)

counts_rel_multrepl_prev05 <- zero_repl_object$rel_abund_repl
```


```{r generate_net_single_prev05}
path_net_single_prev05 <- path_target("net_single_prev05.rds")

if (file.exists(path_net_single_prev05) && !refit_complete_network) {
  net_single_prev05 <- readRDS(path_net_single_prev05)
} else {
  net_single_prev05 <- NetCoMi::netConstruct(
    data = counts_rel_multrepl_prev05,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_prev05, path_net_single_prev05)
}

dim(net_single_prev05$adjaMat1)
```

```{r}
props_single_prev05 <- netAnalyze(net_single_prev05,
                                  centrLCC = TRUE,
                                  clustMethod = "cluster_fast_greedy",
                                  hubPar = "eigenvector",
                                  weightDeg = FALSE, 
                                  normDeg = FALSE,
                                  normBetw = FALSE,
                                  normClose = FALSE,
                                  normEigen = FALSE)

prev05_plot_labels <- make_network_plot_labels(
  props = props_single_prev05,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- prev05_plot_labels$taxa_to_plot
network_labels <- prev05_plot_labels$labels
```

```{r single_net_plot_prev05, fig.width=8.5, fig.height=6, dpi=250}
# Subset layout to prev05 taxa that exist in the main layout, preserving prev05 order.
# NetCoMi passes the layout matrix to qgraph positionally (row i = node i), so the
# row order must match colnames(adja1) after node filtering — which follows the
# prev05 adjacency matrix order.
taxa_prev05_in_layout <- colnames(net_single_prev05$adjaMat1)[
  colnames(net_single_prev05$adjaMat1) %in% rownames(layout)
]
layout_prev05 <- layout[taxa_prev05_in_layout, ]

plot(props_single_prev05,
     layout = layout_prev05,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 1.1,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1.1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 
```

```{r single_net_plot_prev05_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_prev05,
     layout = layout_prev05,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "5% prevalence filter",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### 10% prevalence filter

```{r, warning=FALSE}
phylo_genus_prev10 <- 
  readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev10.rds"))

counts_raw_prev10 <- get_sample_by_taxa(phylo_genus_prev10)

sampdata_net_prev10 <- sampdata %>%
  filter(SampleID %in% rownames(counts_raw_prev10)) %>%
  arrange(match(SampleID, rownames(counts_raw_prev10)))

counts_raw_prev10 <- counts_raw_prev10[sampdata_net_prev10$SampleID, , 
                                       drop = FALSE]

stopifnot(identical(rownames(counts_raw_prev10), sampdata_net_prev10$SampleID))

counts_rel_prev10 <- counts_to_rel_abund(counts_raw_prev10)

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_rel_prev10,
  frac = zero_replacement_fraction
)

counts_rel_multrepl_prev10 <- zero_repl_object$rel_abund_repl
```


```{r generate_net_single_prev10}
path_net_single_prev10 <- path_target("net_single_prev10.rds")

if (file.exists(path_net_single_prev10) && !refit_complete_network) {
  net_single_prev10 <- readRDS(path_net_single_prev10)
} else {
  net_single_prev10 <- NetCoMi::netConstruct(
    data = counts_rel_multrepl_prev10,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_prev10, path_net_single_prev10)
}

dim(net_single_prev10$adjaMat1)
```

```{r}
props_single_prev10 <- netAnalyze(net_single_prev10,
                                  centrLCC = TRUE,
                                  clustMethod = "cluster_fast_greedy",
                                  hubPar = "eigenvector",
                                  weightDeg = FALSE, 
                                  normDeg = FALSE,
                                  normBetw = FALSE,
                                  normClose = FALSE,
                                  normEigen = FALSE)

prev10_plot_labels <- make_network_plot_labels(
  props = props_single_prev10,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- prev10_plot_labels$taxa_to_plot
network_labels <- prev10_plot_labels$labels
```

```{r single_net_plot_prev10, fig.width=8.5, fig.height=6, dpi=250}
taxa_prev10_in_layout <- colnames(net_single_prev10$adjaMat1)[
  colnames(net_single_prev10$adjaMat1) %in% rownames(layout)
]
layout_prev10 <- layout[taxa_prev10_in_layout, ]

plot(props_single_prev10,
     layout = layout_prev10,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 1.1,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 1, 7))

legend(0.8, 1.1, cex = 1.2, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_prev10_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_prev10,
     layout = layout_prev10,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "10% prevalence filter",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

### Different zero replacement

In this section, we construct the network with the following zero replacement 
methods:

* pseudo count added to all counts (as example for what is not recommended)
* multiplicative replacement with sample-wise detection limits
* random pseudo counts below the detection limit
* alrEM
* Bayesian-multiplicative replacement

For all methods, zero replacement is applied externally via
`compute_zero_repl_clr()` before passing the result to `netConstruct()` with
`zeroMethod = "none"`. This ensures the stored `countMat1` in each network
object reflects the actual zero-treated data, so that CLR-based node sizes are
computed from the correct matrix (see analysis of node size behaviour above).

Unlike the baseline multiplicative replacement, which uses one global detection
limit, the sample-wise variant defines the detection limit separately for each
sample. This reflects that the smallest observable relative abundance depends on
the sample's sequencing depth. It may therefore be more realistic for sampling
zeros, but it can also introduce library-size-dependent variation into the
CLR-transformed data. We therefore consider it as a sensitivity analysis rather
than replacing the baseline preprocessing.

**Minimum relative count and multRepl replacement value for reference**

```{r}
min(counts_rel[counts_rel > 0])
min(counts_rel[counts_rel > 0]) * 0.65
```

We use 2e-5 as pseudo count to have a round number close to the multRepl
replacement value.

#### Prepare zero-replaced matrices

```{r prepare_zero_repl_matrices, warning=FALSE}
zero_repl_pseudo <- compute_zero_repl_clr(
  counts_rel, method = "pseudo", pseudocount = 2e-5, zerosOnly = FALSE
)
counts_rel_pseudo <- zero_repl_pseudo$rel_abund_repl

zero_repl_multrepl_sample <- compute_zero_repl_clr(
  counts_rel, method = "multRepl", detection_limit = "sample",
  frac = zero_replacement_fraction
)
counts_rel_multrepl_sample <- zero_repl_multrepl_sample$rel_abund_repl

stopifnot(
  identical(dim(counts_rel_multrepl_sample), dim(counts_rel)),
  identical(dimnames(counts_rel_multrepl_sample), dimnames(counts_rel)),
  !any(counts_rel_multrepl_sample == 0),
  all(abs(rowSums(counts_rel_multrepl_sample) - 1) < 1e-8)
)

set.seed(network_seed)
zero_repl_random <- compute_zero_repl_clr(
  counts_rel, method = "pseudo", pseudocount = 2e-5, zerosOnly = TRUE, 
  random = TRUE
)
counts_rel_random <- zero_repl_random$rel_abund_repl

if (file.exists(path_target("counts_rel_alrEM.rds"))) {
  counts_rel_alrEM <- readRDS(path_target("counts_rel_alrEM.rds"))
} else {
  zero_repl_alrEM <- compute_zero_repl_clr(
    counts_rel, method = "alrEM", z.warning = 0.99, z.delete = FALSE, 
    max.iter = 20
  )
  counts_rel_alrEM <- zero_repl_alrEM$rel_abund_repl
  saveRDS(counts_rel_alrEM, file = path_target("counts_rel_alrEM.rds"))
}

if (file.exists(path_target("counts_rel_bayesMult.rds"))) {
  zero_repl_bayesMult <- readRDS(path_target("counts_rel_bayesMult.rds"))
} else {
  zero_repl_bayesMult <- compute_zero_repl_clr(
    counts_rel, method = "bayesMult", z.warning = 0.99, z.delete = FALSE
  )
  counts_rel_bayesMult <- zero_repl_bayesMult$rel_abund_repl
  saveRDS(zero_repl_bayesMult, file = path_target("counts_rel_bayesMult.rds"))
}
```

#### Additive pseudo count

```{r generate_net_single_zero_pseudo}
path_net_single_zero_pseudo <- path_target("net_single_zero_pseudo.rds")

if (file.exists(path_net_single_zero_pseudo) && !refit_complete_network) {
  net_single_zero_pseudo <- readRDS(path_net_single_zero_pseudo)
} else {
  net_single_zero_pseudo <- NetCoMi::netConstruct(
    data = counts_rel_pseudo,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_pseudo, path_net_single_zero_pseudo)
}
```

```{r}
props_single_zero_pseudo <- netAnalyze(net_single_zero_pseudo,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)

zero_pseudo_plot_labels <- make_network_plot_labels(
  props = props_single_zero_pseudo,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- zero_pseudo_plot_labels$taxa_to_plot
network_labels <- zero_pseudo_plot_labels$labels
```

```{r single_net_plot_zero_pseudo, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_pseudo,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_pseudo_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_pseudo,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Additive pseudo count",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Multiplicative replacement with sample-wise detection limit

```{r generate_net_single_zero_multrepl_sample}
path_net_single_zero_multrepl_sample <- 
  path_target("net_single_zero_multrepl_sample.rds")

if (file.exists(path_net_single_zero_multrepl_sample) && !refit_complete_network) {
  net_single_zero_multrepl_sample <- 
    readRDS(path_net_single_zero_multrepl_sample)
} else {
  net_single_zero_multrepl_sample <- NetCoMi::netConstruct(
    data = counts_rel_multrepl_sample,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_multrepl_sample, 
          path_net_single_zero_multrepl_sample)
}
```

```{r}
props_single_zero_multrepl_sample <- netAnalyze(
  net_single_zero_multrepl_sample,
  centrLCC = TRUE,
  clustMethod = "cluster_fast_greedy",
  hubPar = "eigenvector",
  weightDeg = FALSE, 
  normDeg = FALSE,
  normBetw = FALSE,
  normClose = FALSE,
  normEigen = FALSE
)

zero_multrepl_sample_plot_labels <- make_network_plot_labels(
  props = props_single_zero_multrepl_sample,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- zero_multrepl_sample_plot_labels$taxa_to_plot
network_labels <- zero_multrepl_sample_plot_labels$labels
```

```{r zero_multrepl_sample_edge_count}
tibble(
  method = c("Global detection limit", "Sample-wise detection limit"),
  n_edges = c(
    sum(net_single_main$adjaMat1[lower.tri(net_single_main$adjaMat1)] != 0),
    sum(net_single_zero_multrepl_sample$adjaMat1[
      lower.tri(net_single_zero_multrepl_sample$adjaMat1)
    ] != 0)
  )
)
```

```{r single_net_plot_zero_multrepl_sample, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_multrepl_sample,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_multrepl_sample_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_multrepl_sample,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "multRepl: sample-wise detection",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Random pseudo counts

```{r generate_net_single_zero_random}
path_net_single_zero_random <- path_target("net_single_zero_random.rds")

if (file.exists(path_net_single_zero_random) && !refit_complete_network) {
  net_single_zero_random <- readRDS(path_net_single_zero_random)
} else {
  net_single_zero_random <- NetCoMi::netConstruct(
    data = counts_rel_random,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_random, path_net_single_zero_random)
}
```

```{r}
props_single_zero_random <- netAnalyze(net_single_zero_random,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)

zero_random_plot_labels <- make_network_plot_labels(
  props = props_single_zero_random,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- zero_random_plot_labels$taxa_to_plot
network_labels <- zero_random_plot_labels$labels
```

```{r single_net_plot_zero_random, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_random,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_random_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_random,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Random pseudo counts (zeros only)",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Modified EM ALR algorithm

```{r generate_net_single_zero_alrEM}
path_net_single_zero_alrEM <- path_target("net_single_zero_alrEM.rds")

if (file.exists(path_net_single_zero_alrEM) && !refit_complete_network) {
  net_single_zero_alrEM <- readRDS(path_net_single_zero_alrEM)
} else {
  net_single_zero_alrEM <- NetCoMi::netConstruct(
    data = counts_rel_alrEM,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_alrEM, path_net_single_zero_alrEM)
}
```

```{r}
props_single_zero_alrEM <- netAnalyze(net_single_zero_alrEM,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)

zero_alrEM_plot_labels <- make_network_plot_labels(
  props = props_single_zero_alrEM,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- zero_alrEM_plot_labels$taxa_to_plot
network_labels <- zero_alrEM_plot_labels$labels
```

```{r single_net_plot_zero_alrEM, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_alrEM,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_alrEM_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_alrEM,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Modified EM alr-algorithm",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Bayesian-multiplicative replacement

```{r generate_net_single_zero_bayesMult}
path_net_single_zero_bayesMult <- path_target("net_single_zero_bayesMult.rds")

if (file.exists(path_net_single_zero_bayesMult) && !refit_complete_network) {
  net_single_zero_bayesMult <- readRDS(path_net_single_zero_bayesMult)
} else {
  net_single_zero_bayesMult <- NetCoMi::netConstruct(
    data = counts_rel_bayesMult,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_zero_bayesMult, path_net_single_zero_bayesMult)
}
```

```{r}
props_single_zero_bayesMult <- netAnalyze(net_single_zero_bayesMult,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)

zero_bayesMult_plot_labels <- make_network_plot_labels(
  props = props_single_zero_bayesMult,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- zero_bayesMult_plot_labels$taxa_to_plot
network_labels <- zero_bayesMult_plot_labels$labels
```

```{r single_net_plot_zero_bayesMult, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_zero_bayesMult,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_zero_bayesMult_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_zero_bayesMult,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Bayesian-multiplicative replacement",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

### Different association estimation method

Finally, we will also investigate how the network would look like for different
association estimation methods. Here, we consider the following alternatives:

* Sparse + low-rank decomposition (SPIEC-EASI slr)
* SPRING

#### Sparse + low-rank decomposition (SPIEC-EASI slr)

##### Find optimal rank

```{r generate_net_single_slr_opt_rank, eval=FALSE}

for (r in 2:5) {
  cat("\nr:", r)
  path_net_single_slr <- path_target(paste0("net_single_slr", "_r", r, ".rds"))
  
  if (!file.exists(path_net_single_slr) && !refit_complete_network) {
    net_single_slr <- NetCoMi::netConstruct(
      data = counts_rel_multrepl,
      measure = "spieceasi",
      measurePar = list(method = "slr",
                        nlambda = 30,
                        r = r,
                        pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                             thresh = 0.1), # StARS threshold
                        norm.params = list(method = "clr",
                                           types = "tru_counts"),
                        symBetaMode = "ave"),
      filtTax = "none",
      filtSamp = "none",
      normMethod = "none",
      zeroMethod = "none",
      sparsMethod = "none",
      seed = 42
    )
    
    saveRDS(net_single_slr, path_net_single_slr)
  }
}
```

```{r}
r_candidates <- 2:5

# Same CLR representation that is used inside SpiecEasi
X_clr <- t(SpiecEasi::clr(counts_rel_multrepl, mar = 1))

slr_bic <- function(S, L, X, r, tol = 1e-8) {
  
  Theta <- as.matrix(S - L)
  Theta <- (Theta + t(Theta)) / 2
  
  n <- nrow(X)
  p <- ncol(X)
  Sigma_hat <- stats::cov(X)
  
  chol_theta <- tryCatch(chol(Theta), error = function(e) NULL)
  
  if (is.null(chol_theta)) {
    return(Inf)
  }
  
  log_det_theta <- 2 * sum(log(diag(chol_theta)))
  
  # Gaussian log-likelihood up to an additive constant
  loglik <- n / 2 * (
    log_det_theta -
      sum(Sigma_hat * Theta)
  )
  
  n_edges <- sum(abs(S[upper.tri(S)]) > tol)
  
  # Degrees of freedom of a symmetric p x p matrix of rank r
  df_low_rank <- r * (2 * p - r + 1) / 2
  
  # Diagonal sparse precision entries + off-diagonal sparse edges + low-rank term
  df_total <- p + n_edges + df_low_rank
  
  -2 * loglik + log(n) * df_total
}

slr_bic_table <- do.call(
  rbind,
  lapply(r_candidates, function(r) {
    
    path_net_single_slr <- path_target(
      paste0("net_single_slr_r", r, ".rds")
    )
    
    net_single_slr <- readRDS(path_net_single_slr)
    
    spieceasi_fit <- net_single_slr$measureOut1
    
    opt_ind <- pulsar::opt.index(
      spieceasi_fit$select,
      "stars"
    )
    
    S_opt <- spieceasi_fit$est$icov[[opt_ind]]
    L_opt <- spieceasi_fit$est$resid[[opt_ind]]
    
    data.frame(
      r = r,
      opt_ind = opt_ind,
      lambda = spieceasi_fit$est$lambda[opt_ind],
      edges = sum(abs(S_opt[upper.tri(S_opt)]) > 1e-8),
      BIC = slr_bic(
        S = S_opt,
        L = L_opt,
        X = X_clr,
        r = r
      )
    )
  })
)

slr_bic_table <- slr_bic_table[order(slr_bic_table$BIC), ]

# SLR-BIC table, sorted by increasing BIC
slr_bic_table

r_opt <- slr_bic_table$r[1]

cat("Optimal r:", r_opt)
```

##### Plot low rank heatmaps

```{r}
library(pheatmap)
library(gridExtra)
library(grid)

r_candidates <- 2:5

# Load the fitted objects and extract the low-rank component
slr_low_rank <- lapply(r_candidates, function(r) {
  
  net_single_slr <- readRDS(
    path_target(paste0("net_single_slr_r", r, ".rds"))
  )
  
  spieceasi_fit <- net_single_slr$measureOut1
  
  opt_ind <- pulsar::opt.index(
    spieceasi_fit$select,
    criterion = "stars"
  )
  
  L_opt <- as.matrix(spieceasi_fit$est$resid[[opt_ind]])
  
  list(
    r = r,
    L = L_opt,
    lambda = spieceasi_fit$est$lambda[opt_ind]
  )
})

names(slr_low_rank) <- paste0("r_", r_candidates)

# Use the rank-2 component to define one common ordering of taxa
L_order <- slr_low_rank[["r_2"]]$L
diag(L_order) <- 0

taxon_order <- hclust(
  as.dist(1 - stats::cor(L_order))
)$order

taxon_names <- colnames(L_order)[taxon_order]
```

```{r}
# Determine a common symmetric colour range across all ranks
all_values <- unlist(lapply(slr_low_rank, function(x) {
  L <- x$L
  L[row(L) == col(L)] <- NA_real_
  L[upper.tri(L)]
}))

lim <- max(abs(all_values), na.rm = TRUE)

heatmap_grobs <- lapply(slr_low_rank, function(x) {
  
  L <- x$L[taxon_order, taxon_order]
  
  # Suppress the diagonal, which is not a pairwise association contribution
  diag(L) <- NA_real_
  
  pheatmap(
    mat = L,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    border_color = NA,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(101),
    breaks = seq(-lim, lim, length.out = 102),
    main = paste0(
      "Rank ", x$r,
      "  (StARS-selected \u03bb = ",
      formatC(x$lambda, format = "f", digits = 3),
      ")"
    ),
    silent = TRUE
  )$gtable
})
```

```{r slr_low_rank_heatmaps, fig.width=10, fig.height=10, dpi=250}
gridExtra::grid.arrange(
  grobs = heatmap_grobs,
  ncol = 2
)
```

##### Continue with optimal rank

```{r}
library(pheatmap)
library(gridExtra)
library(grid)

r_chosen <- 2

net_single_slr <- readRDS(
  path_target(paste0("net_single_slr_r", r_chosen, ".rds"))
)

spieceasi_fit <- net_single_slr$measureOut1

opt_ind <- pulsar::opt.index(
  spieceasi_fit$select,
  "stars"
)

S_opt <- as.matrix(spieceasi_fit$est$icov[[opt_ind]])
L_opt <- as.matrix(spieceasi_fit$est$resid[[opt_ind]])

# Use a common ordering for both panels
Theta_slr <- S_opt - L_opt
taxon_order <- hclust(dist(Theta_slr))$order

S_plot <- S_opt[taxon_order, taxon_order]
L_plot <- L_opt[taxon_order, taxon_order]

# Optional: suppress diagonal to focus on off-diagonal structure
diag(S_plot) <- NA_real_
diag(L_plot) <- NA_real_

# Common scale based only on sparse + low-rank parts
all_vals <- c(
  S_plot[upper.tri(S_plot)],
  L_plot[upper.tri(L_plot)]
)

# Robust symmetric range
lim <- as.numeric(
  quantile(abs(all_vals), probs = 0.99, na.rm = TRUE)
)

breaks <- seq(-lim, lim, length.out = 102)

make_heatmap <- function(mat, title) {
  pheatmap(
    mat = mat,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    border_color = NA,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(101),
    breaks = breaks,
    main = title,
    silent = TRUE
  )$gtable
}

g_sparse <- make_heatmap(S_plot, "Sparse part")
g_lowrank <- make_heatmap(L_plot, "Summed low-rank part")
```

```{r slr_decomposition_sparse_lowrank, fig.width=8.5, fig.height=4.5, dpi=250}
gridExtra::grid.arrange(g_sparse, g_lowrank, ncol = 2)
```

```{r load_net_single_slr}
net_single_slr <- readRDS(
  path_target(paste0("net_single_slr_r", r_chosen, ".rds"))
)
```

```{r}
props_single_slr <- netAnalyze(net_single_slr,
                               centrLCC = TRUE,
                               clustMethod = "cluster_fast_greedy",
                               hubPar = "eigenvector",
                               weightDeg = FALSE, 
                               normDeg = FALSE,
                               normBetw = FALSE,
                               normClose = FALSE,
                               normEigen = FALSE)
```

##### Network plot

```{r}
# Defines labels to plot
slr_plot_labels <- make_network_plot_labels(
  props = props_single_slr,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 10,
  n_central = 10,
  label_overrides = network_label_overrides
)

taxa_to_plot <- slr_plot_labels$taxa_to_plot
network_labels <- slr_plot_labels$labels
```

```{r single_net_plot_slr, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_slr,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.85,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r}
layout_slr <- p$layout$layout1
layout_slr["Bacteroides", ] <- c(-0.13, 0.12)
layout_slr["Staphylococcus", ] <- c(0.20, 0.04)
layout_slr["Haemophilus", ] <- c(0.18, -0.34)
```


```{r single_net_plot_slr_slrlay, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_slr,
     layout = layout_slr,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.73,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     edgeTranspLow = 30,
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_slr_slrlay_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_slr,
     layout = layout_slr,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.73,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPIEC-EASI slr",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

```{r single_net_plot_slr_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_slr,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPIEC-EASI slr",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))

```

#### Sparse + low rank (unsigned)

```{r}
path_net_single_slr_us <- path_target("net_single_slr_us.rds")

if (file.exists(path_net_single_slr_us) && !refit_complete_network) {
  net_single_slr_us <- readRDS(path_net_single_slr_us)
} else {
  net_single_slr_us <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "spieceasi",
    measurePar = list(method = "slr",
                      nlambda = 30,
                      r = 2,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    dissFunc = "unsigned",
    seed = 42
  )
  
  saveRDS(net_single_slr_us, path_net_single_slr_us)
}
```

```{r}
props_single_slr_us <- netAnalyze(net_single_slr_us,
                                  centrLCC = TRUE,
                                  clustMethod = "cluster_fast_greedy",
                                  hubPar = "eigenvector",
                                  weightDeg = FALSE, 
                                  normDeg = FALSE,
                                  normBetw = FALSE,
                                  normClose = FALSE,
                                  normEigen = FALSE)
```

```{r}
# Defines labels to plot
slr_us_plot_labels <- make_network_plot_labels(
  props = props_single_slr_us,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 15,
  n_central = 10,
  label_overrides = network_label_overrides
)

taxa_to_plot <- slr_us_plot_labels$taxa_to_plot
network_labels <- slr_us_plot_labels$labels
```

```{r single_net_plot_slr_us, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_slr_us,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.9,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```


#### SPRING

```{r generate_net_single_spring}
path_net_single_spring <- path_target("net_single_spring.rds")

if (file.exists(path_net_single_spring) && !refit_complete_network) {
  net_single_spring <- readRDS(path_net_single_spring)
} else {
  net_single_spring <- NetCoMi::netConstruct(
    data = counts_rel,
    measure = "spring",
    measurePar = list(nlambda = 30, 
                      rep.num = 30,
                      thresh  = 0.1,
                      Rmethod = "approx"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_single_spring, path_net_single_spring)
}
```

```{r}
props_single_spring <- netAnalyze(net_single_spring,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```


```{r}
spring_plot_labels <- make_network_plot_labels(
  props = props_single_spring,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 10,
  n_central = 10,
  label_overrides = network_label_overrides
)

taxa_to_plot <- spring_plot_labels$taxa_to_plot
network_labels <- spring_plot_labels$labels
```

```{r single_net_plot_spring, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_spring,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.6,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     edgeTranspLow = 30,
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r}
layout_spring <- p$layout$layout1
layout_spring["Faecalibacterium", ] <- c(-0.436, 0.75)
```


```{r single_net_plot_spring_ownlay, fig.width=8.5, fig.height=6, dpi=250}
plot(props_single_spring,
     layout = layout_spring,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.899,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     edgeTranspLow = 30,
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_spring_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_spring,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPRING",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

```{r single_net_plot_spring_slrlay_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_spring,
     layout = layout_slr,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPRING",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```


#### SPRING (unsigned)

```{r generate_net_single_spring_us}
path_net_single_spring_us <- path_target("net_single_spring_us.rds")

if (file.exists(path_net_single_spring_us) && !refit_complete_network) {
  net_single_spring_us <- readRDS(path_net_single_spring_us)
} else {
  net_single_spring_us <- NetCoMi::netConstruct(
    data = counts_rel,
    measure = "spring",
    measurePar = list(nlambda = 30, 
                      rep.num = 30,
                      thresh  = 0.1,
                      Rmethod = "approx"),
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    dissFunc = "unsigned",
    seed = 42
  )
  
  saveRDS(net_single_spring_us, path_net_single_spring_us)
}
```

```{r}
props_single_spring_us <- netAnalyze(net_single_spring_us,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)
```


```{r}
spring_plot_labels_us <- make_network_plot_labels(
  props = props_single_spring_us,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = 10,
  n_central = 10,
  label_overrides = network_label_overrides
)

taxa_to_plot <- spring_plot_labels_us$taxa_to_plot
network_labels <- spring_plot_labels_us$labels
```

```{r single_net_plot_spring_us, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_spring_us,
     #layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.99,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     edgeTranspLow = 30,
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_spring_us_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_spring_us,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "mclr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "SPRING",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

#### Shrinkage proportionality

```{r generate_net_single_rhoshrink_nothresh}
net_single_rhoshrink_nothresh <- NetCoMi::netConstruct(
  data = counts_rel_multrepl,
  measure = "rhoshrink",
  filtTax = "none",
  filtSamp = "none",
  normMethod = "none",
  zeroMethod = "pseudo",
  zeroPar = list(pseudocount = 0),
  sparsMethod = "none",
  seed = 42
)

assomat_single_rhoshrink <- net_single_rhoshrink_nothresh$assoMat1
adjamat_single_rhoshrink <- net_single_rhoshrink_nothresh$adjaMat1
```

```{r single_rhoshrink_histrogram, fig.width=8, fig.height=2}
n_zero_asso <- 
  sum(assomat_single_rhoshrink[lower.tri(assomat_single_rhoshrink)] == 0)

data.frame(correlation = assomat_single_rhoshrink[lower.tri(assomat_single_rhoshrink)]) %>%
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.005) +
  labs(x = "Estimated correlation",
       y = "Count",
       #title = "Estimated correlations after sparsification"
       ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )
```

```{r generate_net_single_rhoshrink}
path_net_single_rhoshrink <- path_target("net_single_rhoshrink.rds")

if (file.exists(path_net_single_rhoshrink) && !refit_complete_network) {
  net_single_rhoshrink <- readRDS(path_net_single_rhoshrink)
} else {
  net_single_rhoshrink <- NetCoMi::netConstruct(
    data = counts_rel_multrepl,
    measure = "rhoshrink",
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "pseudo",
    zeroPar = list(pseudocount = 0),
    sparsMethod = "threshold",
    thresh = 0.2,
    seed = 42
  )
  
  saveRDS(net_single_rhoshrink, path_net_single_rhoshrink)
}
```

```{r}
props_single_rhoshrink <- netAnalyze(net_single_rhoshrink,
                                       centrLCC = TRUE,
                                       clustMethod = "cluster_fast_greedy",
                                       hubPar = "eigenvector",
                                       weightDeg = FALSE, 
                                       normDeg = FALSE,
                                       normBetw = FALSE,
                                       normClose = FALSE,
                                       normEigen = FALSE)

rhoshrink_plot_labels <- make_network_plot_labels(
  props = props_single_rhoshrink,
  top_abundant_taxa = top_abundant_taxa,
  da_taxa = da_plot_taxa,
  dd_taxa = dd_plot_taxa,
  n_abundant = n_abundant_network_labels,
  n_central = n_central_network_labels,
  label_overrides = network_label_overrides
)

taxa_to_plot <- rhoshrink_plot_labels$taxa_to_plot
network_labels <- rhoshrink_plot_labels$labels
```

```{r single_net_plot_rhoshrink, fig.width=8.5, fig.height=6, dpi=250}
p <- plot(props_single_rhoshrink,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.899,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = FALSE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 7, 1, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

```{r single_net_plot_rhoshrink_nolabs, fig.width=5.8, fig.height=6, dpi=250}
plot(props_single_rhoshrink,
     layout = layout,
     nodeFilter = "names",
     nodeFilterPar = taxa_to_plot,
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.8999999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     title1 = "Shrinkage proportionality",
     cexTitle = 1.8,
     cexLabels = 0,
     cexHubLabels = 0,
     hubBorderCol = "gray40",
     mar = c(1, 1, 3, 1))
```

## Summary of edge overlap

The following UpSet-style plots compare the edge sets of the networks shown
above. Edges are encoded as present or absent based on the adjacency matrix. For
this summary, only the signed versions of the networks are included.

### Prevalence filtering

```{r edge_upset_prevalence, fig.width=10, fig.height=6, dpi=250}
networks_prevalence <- list(
  "1% prevalence" = net_single_main,
  "5% prevalence" = net_single_prev05,
  "10% prevalence" = net_single_prev10
)

edge_presence_prevalence <- make_edge_presence(networks_prevalence)

plot_edge_upset(
  edge_presence_prevalence,
  title = "Shared edges across prevalence filters",
  networks = networks_prevalence,
  shared_text_size = 4
)
```

```{r edge_upset_prevalence8, fig.width=10, fig.height=5, dpi=250}
networks_prevalence <- list(
  "1% prevalence" = net_single_main,
  "5% prevalence" = net_single_prev05,
  "10% prevalence" = net_single_prev10
)

edge_presence_prevalence <- make_edge_presence(networks_prevalence)

plot_edge_upset(
  edge_presence_prevalence,
  title = "Shared edges across prevalence filters",
  networks = networks_prevalence,
  shared_text_size = 4,
  n_shared_edges = 8
)
```

### Zero replacement

```{r edge_upset_zero_replacement, fig.width=10, fig.height=6, dpi=250}
networks_zero_replacement <- list(
  "multiplicative" = net_single_main,
  #"multiplicative sample-wise" = net_single_zero_multrepl_sample,
  "random pseudo count" = net_single_zero_random,
  "Modified EM alr" = net_single_zero_alrEM,
  "Bayesian mult." = net_single_zero_bayesMult
)

edge_presence_zero_replacement <- make_edge_presence(networks_zero_replacement)

plot_edge_upset(
  edge_presence_zero_replacement,
  title = "Shared edges across zero-replacement methods",
  networks = networks_zero_replacement,
  shared_text_size = 4
)
```

```{r edge_upset_zero_replacement8, fig.width=10, fig.height=5, dpi=250}
networks_zero_replacement <- list(
  "multiplicative" = net_single_main,
  #"multiplicative sample-wise" = net_single_zero_multrepl_sample,
  "random pseudo count" = net_single_zero_random,
  "Modified EM alr" = net_single_zero_alrEM,
  "Bayesian mult." = net_single_zero_bayesMult
)

edge_presence_zero_replacement <- make_edge_presence(networks_zero_replacement)

plot_edge_upset(
  edge_presence_zero_replacement,
  title = "Shared edges across zero-replacement methods",
  networks = networks_zero_replacement,
  shared_text_size = 4,
  n_shared_edges = 8
)
```

### Association measure

```{r edge_upset_association, fig.width=10, fig.height=6, dpi=250}
networks_association <- list(
  "SPIEC-EASI MB" = net_single_main,
  "SPIEC-EASI SLR" = net_single_slr,
  "SPRING" = net_single_spring
  #"rho shrinkage" = net_single_rhoshrink
)

edge_presence_association <- make_edge_presence(networks_association)

plot_edge_upset(
  edge_presence_association,
  title = "Shared edges across association measures",
  networks = networks_association,
  shared_text_size = 4
)
```

```{r edge_upset_association8, fig.width=10, fig.height=5, dpi=250}
networks_association <- list(
  "SPIEC-EASI MB" = net_single_main,
  "SPIEC-EASI SLR" = net_single_slr,
  "SPRING" = net_single_spring
  #"rho shrinkage" = net_single_rhoshrink
)

edge_presence_association <- make_edge_presence(networks_association)

plot_edge_upset(
  edge_presence_association,
  title = "Shared edges across association measures",
  networks = networks_association,
  shared_text_size = 4,
  n_shared_edges = 8
)
```

## Persistent edges across selected networks

The following summary uses a selected set of signed networks and identifies
taxa-pairs that are connected in all of them.

```{r select_persistent_edge_networks}
persistent_edge_networks <- list(
  "1% prevalence / multiplicative / SPIEC-EASI mb" = net_single_main,
  "5% prevalence" = net_single_prev05,
  "10% prevalence" = net_single_prev10,
  #"pseudo count" = net_single_zero_pseudo,
  #"multiplicative sample-wise" = net_single_zero_multrepl_sample,
  "random pseudo count" = net_single_zero_random,
  "Modified EM alr" = net_single_zero_alrEM,
  "Bayesian mult." = net_single_zero_bayesMult,
  "SPIEC-EASI slr" = net_single_slr,
  "SPRING" = net_single_spring
  #"rho shrinkage" = net_single_rhoshrink
)

persistent_edges <- summarise_persistent_edges(persistent_edge_networks)
```

```{r persistent_edge_table}
persistent_edges %>%
  mutate(
    mean_association = round(mean_association, 3),
    min_association = round(min_association, 3),
    max_association = round(max_association, 3),
    mean_abs_association = round(mean_abs_association, 3)
  ) %>%
  select(
    taxon1_label,
    taxon2_label,
    n_methods,
    mean_association,
    min_association,
    max_association,
    mean_abs_association
  ) %>%
  kableExtra::kbl(
    caption = "Taxa-pairs connected in all selected signed networks.",
    col.names = c(
      "Taxon 1",
      "Taxon 2",
      "Networks",
      "Mean association",
      "Minimum",
      "Maximum",
      "Mean absolute association"
    )
  ) %>%
  kableExtra::kable_styling(full_width = FALSE)
```

```{r persistent_edge_network_prepare}
persistent_taxa <- persistent_edges %>%
  select(taxon1, taxon2) %>%
  unlist(use.names = FALSE) %>%
  unique()

net_single_persistent <- make_persistent_edge_network(
  network = net_single_main,
  persistent_edges = persistent_edges
)

props_single_persistent <- netAnalyze(
  net_single_persistent,
  centrLCC = TRUE,
  clustMethod = "cluster_fast_greedy",
  hubPar = "eigenvector",
  weightDeg = FALSE, 
  normDeg = FALSE,
  normBetw = FALSE,
  normClose = FALSE,
  normEigen = FALSE
)

persistent_network_labels <- setNames(
  rep("", ncol(net_single_persistent$adjaMat1)),
  colnames(net_single_persistent$adjaMat1)
)

persistent_label_taxa <- intersect(persistent_taxa, names(persistent_network_labels))
persistent_network_labels[persistent_label_taxa] <- persistent_label_taxa

label_override_taxa <- intersect(
  names(network_label_overrides),
  persistent_label_taxa
)
persistent_network_labels[label_override_taxa] <- 
  network_label_overrides[label_override_taxa]

persistent_layout <- layout[
  intersect(rownames(layout), colnames(net_single_persistent$adjaMat1)),
  ,
  drop = FALSE
]
```

```{r persistent_edge_network_plot, fig.width=5.9, fig.height=6, dpi=250}
plot(
  props_single_persistent,
  layout = persistent_layout,
  nodeFilter = "names",
  nodeFilterPar = persistent_taxa,
  labelScale = FALSE,
  labels = persistent_network_labels,
  repulsion = 0.899,
  nodeColor = "feature", 
  featVecCol = phyla, 
  colorVec =  phylcol,
  nodeSize = "clr",
  nodeSizeSpread = 4,
  cexNodes = 2,
  nodeTransp = 40,
  hubTransp = 40,
  posCol = "gray50", 
  negCol = "red3",
  showTitle = TRUE,
  title1 = "Persistent edges",
  cexTitle = 1.8,
  cexLabels = 1,
  hubLabelFont = 0.9,
  cexHubLabels = 0.9,
  hubBorderCol = "gray40",
  mar = c(1, 1, 3, 2))
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
