---
title: "Network comparison between EBF groups"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "11_network_comparison"
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "2e531120-aad0-4546-bf2a-05d6465091a5")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 8, 
  fig.height = 5, 
  dpi = 250,
  fig.path = paste0('figures/', params$name, "/net_"),
  echo = FALSE)
```

This document compares genus-level microbial association networks between
infants without exclusive breastfeeding and infants with at least two months of
exclusive breastfeeding. The first part contrasts several network construction
approaches descriptively. The second part conducts permutation-based network
comparison for one selected descriptive network.

```{r packages, message=FALSE, warning=FALSE}
# SpiecEasi from latentcor branch needed
if (as.character(utils::packageVersion("SpiecEasi")) != "2.0.1") {
  remotes::install_github("zdk123/SpiecEasi", ref = "latentcor")
}

library(conflicted)
library(phyloseq)
library(tidyverse)
library(NetCoMi)
library(permApprox)
library(igraph)
library(patchwork)
library(ggplot2)
library(pheatmap)
library(gridExtra)
library(grid)

# For permutation association matrices
library(foreach)
library(future)
library(doFuture)
library(doRNG)
library(progressr)
library(filematrix)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)
conflicts_prefer(base::union)

application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# Keep cached networks, permutation matrices, and permutation-test results.
projthis::proj_create_dir_target(params$name, clean = FALSE)

path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
```

## Parameters

```{r parameters}
zero_replacement_fraction <- 0.65
network_seed <- 42
permutation_seed <- 42
n_perm_network_comparison <- 999
cores <- 10

slr_rank <- 2
selected_permutation_method <- "corshrink_bayesMult"

refit_network_comparison <- FALSE
refit_permutation_matrices <- FALSE
refit_permutation_tests <- FALSE
```

## Load And Prepare Data

The comparison uses the two EBF groups that define the main contrast in this
chapter: no exclusive breastfeeding and exclusive breastfeeding for at least two
months. Samples with one month of exclusive breastfeeding are excluded.

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))

phylo_genus <- renameTaxa(
  phylo_genus,
  pat = "<name>",
  substPat = "<name>_<subst_name>(<subst_R>)",
  numDupli = "Genus"
)
```

```{r prepare_ebf_groups, warning=FALSE}
ebf_keep <- sampdata$`EBF duration` %in% c("0", ">=2")

phylo_genus_ebf <- prune_samples(ebf_keep, phylo_genus)
phylo_genus_ebf <- prune_taxa(taxa_sums(phylo_genus_ebf) > 0, phylo_genus_ebf)

metadata_ebf <- sampdata[ebf_keep, , drop = FALSE] %>%
  mutate(
    ebf_group = factor(
      as.character(`EBF duration`),
      levels = c("0", ">=2"),
      labels = c("0 months", ">=2 months")
    )
  )

counts_ebf_raw <- get_sample_by_taxa(phylo_genus_ebf)
metadata_ebf <- metadata_ebf %>%
  filter(SampleID %in% rownames(counts_ebf_raw)) %>%
  arrange(match(SampleID, rownames(counts_ebf_raw)))
counts_ebf_raw <- counts_ebf_raw[metadata_ebf$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_ebf_raw), metadata_ebf$SampleID))

counts_ebf_rel <- counts_to_rel_abund(counts_ebf_raw)
ebf_group <- metadata_ebf$ebf_group

zero_repl_multRepl <- compute_multrepl_clr(
  rel_abund_mat = counts_ebf_rel,
  frac = zero_replacement_fraction
)
counts_ebf_rel_multRepl <- zero_repl_multRepl$rel_abund_repl

path_counts_ebf_rel_bayesMult <- path_target("counts_ebf_rel_bayesMult.rds")

if (file.exists(path_counts_ebf_rel_bayesMult) && !refit_network_comparison) {
  zero_repl_bayesMult <- readRDS(path_counts_ebf_rel_bayesMult)
} else {
  zero_repl_bayesMult <- compute_zero_repl_clr(
    rel_abund_mat = counts_ebf_rel,
    method = "bayesMult",
    z.warning = 0.99,
    z.delete = FALSE
  )
  saveRDS(zero_repl_bayesMult, path_counts_ebf_rel_bayesMult)
}

counts_ebf_rel_bayesMult <- zero_repl_bayesMult$rel_abund_repl

metadata_ebf %>%
  count(ebf_group, name = "n_samples")
```

```{r plot_metadata}
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]

phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)

genus_mat <- get_sample_by_taxa(phylo_genus)
genus_rel_mat <- counts_to_rel_abund(genus_mat)

top_abundant_taxa <- colMeans(genus_rel_mat, na.rm = TRUE) %>%
  sort(decreasing = TRUE) %>%
  names()
names(top_abundant_taxa) <- top_abundant_taxa

network_label_overrides <- c(
  "8_Comamonadaceae(F)" = "Comamonadaceae(F)",
  "24_Enterobacteriaceae(F)" = "Enterobacteriaceae(F)"
)
```

## Helper Functions

```{r comparison_helpers}
transform_asso <- function(assoMat, thresh = NULL, dissTrans = "signed") {
  if (!is.null(thresh)) {
    assoMat[abs(assoMat) < thresh] <- 0
  }

  if (dissTrans == "signed") {
    dissMat <- sqrt(0.5 * (1 - assoMat))
  } else {
    dissMat <- sqrt(1 - assoMat^2)
  }

  dissMat[assoMat == 0] <- 1
  simMat <- 1 - dissMat
  graphObj <- SpiecEasi::adj2igraph(simMat)

  list(graph = graphObj, adja = simMat, asso = assoMat, diss = dissMat)
}

make_network_plot_labels <- function(
    props,
    top_abundant_taxa,
    n_abundant = 10,
    n_central = 10,
    extra_taxa = NULL,
    label_overrides = network_label_overrides) {

  adja1 <- props$input$adjaMat1
  adja2 <- props$input$adjaMat2
  diag(adja1) <- 0
  diag(adja2) <- 0
  network_taxa <- union(colnames(adja1), colnames(adja2))

  non_singleton_taxa1 <- names(which(
    rowSums(adja1 != 0) > 0 |
      colSums(adja1 != 0) > 0
  ))
  non_singleton_taxa2 <- names(which(
    rowSums(adja2 != 0) > 0 |
      colSums(adja2 != 0) > 0
  ))
  non_singleton_taxa <- union(non_singleton_taxa1, non_singleton_taxa2)

  abundant_taxa <- top_abundant_taxa[seq_len(
    min(max(n_abundant, 0), length(top_abundant_taxa))
  )]

  get_central_taxa <- function(centrality) {
    if (is.null(centrality) || n_central <= 0) {
      return(character(0))
    }
    centrality <- centrality[is.finite(centrality)]
    central_taxa <- centrality %>%
      sort(decreasing = TRUE) %>%
      head(n_central) %>%
      names()
    names(central_taxa) <- central_taxa
    central_taxa
  }

  central_taxa <- c(
    get_central_taxa(props$centralities$eigenv1),
    get_central_taxa(props$centralities$eigenv2)
  )
  central_taxa <- central_taxa[!duplicated(names(central_taxa))]

  taxa_to_label <- c(abundant_taxa, central_taxa, extra_taxa)
  taxa_to_label <- taxa_to_label[!duplicated(names(taxa_to_label))]
  taxa_to_label_in_network <- intersect(names(taxa_to_label), network_taxa)

  taxa_to_plot <- unique(c(non_singleton_taxa, taxa_to_label_in_network))
  network_labels <- setNames(rep("", length(network_taxa)), network_taxa)
  network_labels[taxa_to_label_in_network] <- taxa_to_label[taxa_to_label_in_network]

  if (!is.null(label_overrides)) {
    label_override_taxa <- intersect(names(label_overrides), names(network_labels))
    network_labels[label_override_taxa] <- label_overrides[label_override_taxa]
  }

  network_labels <- stringr::str_replace(
    network_labels,
    "^\\d+_([^_]+\\(F\\))$",
    "\\1"
  )

  list(
    labels = network_labels,
    taxa_to_plot = taxa_to_plot,
    taxa_to_label = taxa_to_label,
    taxa_to_label_in_network = taxa_to_label_in_network,
    non_singleton_taxa = non_singleton_taxa
  )
}

extract_group_network_objects <- function(net) {
  list(
    asso1 = net$assoMat1,
    asso2 = net$assoMat2,
    adja1 = net$adjaMat1,
    adja2 = net$adjaMat2,
    graph1 = transform_asso(net$assoMat1)$graph,
    graph2 = transform_asso(net$assoMat2)$graph
  )
}

analyze_group_network <- function(net) {
  netAnalyze(
    net,
    centrLCC = TRUE,
    clustMethod = "cluster_fast_greedy",
    hubPar = "eigenvector",
    weightDeg = FALSE,
    normDeg = FALSE,
    normBetw = FALSE,
    normClose = FALSE,
    normEigen = FALSE
  )
}

plot_group_association_histograms <- function(net_objects, binwidth = 0.01,
                                              xlim = c(-0.2, 0.8),
                                              ylim = c(0, 50)) {
  make_plot <- function(asso, x_title = FALSE) {
    n_zero <- sum(asso[lower.tri(asso)] == 0)
    data.frame(correlation = asso[lower.tri(asso)]) %>%
      ggplot(aes(x = correlation)) +
      geom_histogram(binwidth = binwidth) +
      annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
               label = paste0("n = ", n_zero), size = 4) +
      coord_cartesian(ylim = ylim, xlim = xlim) +
      labs(
        x = if (x_title) "Estimated association" else "",
        y = "Count"
      ) +
      theme_bw() +
      theme(
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title.x = if (x_title) element_text(size = 14) else element_blank(),
        axis.text.x = if (x_title) element_text(size = 12) else element_blank(),
        axis.ticks.x = if (x_title) element_line() else element_blank()
      )
  }

  make_plot(net_objects$asso1, x_title = FALSE) /
    make_plot(net_objects$asso2, x_title = TRUE)
}

plot_group_degree_distribution <- function(net_objects) {
  ddist1 <- igraph::degree.distribution(net_objects$graph1)
  ddist2 <- igraph::degree.distribution(net_objects$graph2)

  max_len <- max(length(ddist1), length(ddist2))
  ddist1 <- c(ddist1, rep(0, max_len - length(ddist1)))
  ddist2 <- c(ddist2, rep(0, max_len - length(ddist2)))
  degree_seq <- 0:(max_len - 1)

  make_plot <- function(ddist, x_title = FALSE) {
    data.frame(
      Degree = as.factor(degree_seq),
      Fraction = ddist
    ) %>%
      ggplot(aes(x = Degree, y = Fraction, group = 1)) +
      geom_line(color = "gray30") +
      geom_point(size = 2, color = "gray30") +
      theme_bw() +
      theme(
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title.x = if (x_title) element_text(size = 14) else element_blank(),
        axis.text.x = if (x_title) element_text(size = 12) else element_blank(),
        axis.ticks.x = if (x_title) element_line() else element_blank()
      )
  }

  make_plot(ddist1, x_title = FALSE) /
    make_plot(ddist2, x_title = TRUE)
}

plot_group_network <- function(props, same_layout = FALSE, labels = NULL,
                               repulsion = 0.9, title1 = "EBF duration = 0",
                               title2 = "EBF duration >= 2",
                               labelScale = FALSE, cexLabels = 0.9,
                               phylum_legend_x = -0.15,
                               phylum_legend_y = -0.25,
                               asso_legend_x   = -0.15,
                               asso_legend_y   = 1.2,
                               doPlot = TRUE,
                               layout = NULL, 
                               highlightHubs = TRUE,
                               legends = TRUE,
                               mar = c(1,3,3,3)) {
  
  rmSinglesArg <- ifelse(same_layout, "inboth", TRUE)
  
  plot_args <- list(
    x = props,
    layout = layout,
    sameLayout = same_layout,
    rmSingles = rmSinglesArg,
    title1 = title1,
    title2 = title2,
    labelScale = labelScale,
    labels = labels,
    repulsion = repulsion,
    nodeColor = "feature",
    featVecCol = phyla,
    colorVec = phylcol,
    nodeSize = "clr",
    nodeSizeSpread = 4,
    cexNodes = 2,
    nodeTransp = 40,
    hubTransp = 40,
    posCol = "gray50",
    negCol = "red3",
    showTitle = TRUE,
    highlightHubs = highlightHubs,
    cexLabels = cexLabels,
    cexHubLabels = cexLabels,
    cexTitle = 1.4,
    hubBorderCol = "gray40",
    mar = mar,
    doPlot = doPlot
  )
  if (same_layout) {
    plot_args$layoutGroup <- "union"
  }
  
  p <- do.call(plot, plot_args)
  
  if (doPlot && legends) {
    phylcol_transp <- colToTransp(phylcol, 40)
    
    legend(phylum_legend_x, phylum_legend_y, cex = 1, pt.cex = 2.5, title = NULL,
           legend = levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
           xpd = NA)
    
    legend(asso_legend_x, asso_legend_y, cex = 1.1, title = "Estimated association:",
           legend = c("+", "-"), lty = 1, lwd = 3, col = c("gray50", "red3"),
           bty = "n", horiz = TRUE, xpd = NA)
  }
  
  invisible(p)
}

build_plot_labels <- function(props, n_abundant = 15, n_central = 10) {
  make_network_plot_labels(
    props = props,
    top_abundant_taxa = top_abundant_taxa,
    n_abundant = n_abundant,
    n_central = n_central,
    label_overrides = network_label_overrides
  )$labels
}

blank_plot_labels <- function(labels) {
  out <- rep("", length(labels))
  names(out) <- names(labels)
  out
}

extract_global_network_properties <- function(props, method) {
  tibble(
    method = method,
    group = c("EBF = 0", "EBF >= 2"),
    relative_lcc_size = c(
      props$globalPropsLCC$lccSizeRel1,
      props$globalPropsLCC$lccSizeRel2
    ),
    density = c(
      props$globalProps$density1,
      props$globalProps$density2
    ),
    natural_connectivity = c(
      props$globalProps$natConnect1,
      props$globalProps$natConnect2
    )
  )
}

make_global_property_table <- function(global_properties, property) {
  global_properties %>%
    select(group, method, value = all_of(property)) %>%
    mutate(value = round(value, 3)) %>%
    pivot_wider(names_from = method, values_from = value)
}
```

## Descriptive Network Comparisons

The following sections construct and describe EBF-group-specific networks
without permutation testing.


### SPIEC-EASI MB with multRepl

```{r generate_net_ebf_spieceasi_multRepl}
path_net_ebf_spieceasi_multRepl <- path_target("net_ebf_spieceasi_multRepl.rds")

if (file.exists(path_net_ebf_spieceasi_multRepl) && !refit_network_comparison) {
  net_ebf_spieceasi_multRepl <- readRDS(path_net_ebf_spieceasi_multRepl)
} else {
  net_ebf_spieceasi_multRepl <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multRepl,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_spieceasi_multRepl, path_net_ebf_spieceasi_multRepl)
}

objects_ebf_spieceasi_multRepl <- extract_group_network_objects(
  net_ebf_spieceasi_multRepl
)
props_ebf_spieceasi_multRepl <- analyze_group_network(
  net_ebf_spieceasi_multRepl
)
labels_ebf_spieceasi_multRepl <- build_plot_labels(
  props_ebf_spieceasi_multRepl
)
```

#### netAnalyze summary

```{r ebf_spieceasi_multRepl_netAnalyze_summary}
print(summary(props_ebf_spieceasi_multRepl,
              groupNames = c("EBF = 0", "EBF >= 2")))
```


#### Association Histogram

```{r ebf_spieceasi_multRepl_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_spieceasi_multRepl)
```

#### Degree Distribution

```{r ebf_spieceasi_multRepl_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_spieceasi_multRepl)
```

#### Graphlet Correlation Matrix

```{r ebf_spieceasi_multRepl_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_spieceasi_multRepl,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_spieceasi_multRepl_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_multRepl,
  same_layout = FALSE,
  labels = labels_ebf_spieceasi_multRepl,
  repulsion = 0.8
)
```

```{r ebf_spieceasi_multRepl_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_multRepl,
  same_layout = TRUE,
  labels = labels_ebf_spieceasi_multRepl,
  repulsion = 0.79999
)
```

```{r ebf_spieceasi_multRepl_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_multRepl,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_spieceasi_multRepl),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```

### SPIEC-EASI MB with BayesMult replacement

```{r generate_net_ebf_spieceasi_bayesMult}
path_net_ebf_spieceasi_bayesMult <- path_target("net_ebf_spieceasi_bayesMult.rds")

if (file.exists(path_net_ebf_spieceasi_bayesMult) && !refit_network_comparison) {
  net_ebf_spieceasi_bayesMult <- readRDS(path_net_ebf_spieceasi_bayesMult)
} else {
  net_ebf_spieceasi_bayesMult <- NetCoMi::netConstruct(
    data = counts_ebf_rel_bayesMult,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_spieceasi_bayesMult, path_net_ebf_spieceasi_bayesMult)
}

objects_ebf_spieceasi_bayesMult <- extract_group_network_objects(
  net_ebf_spieceasi_bayesMult
)
props_ebf_spieceasi_bayesMult <- analyze_group_network(
  net_ebf_spieceasi_bayesMult
)
labels_ebf_spieceasi_bayesMult <- build_plot_labels(
  props_ebf_spieceasi_bayesMult
)
```

#### netAnalyze summary

```{r ebf_spieceasi_bayesMult_netAnalyze_summary}
print(summary(props_ebf_spieceasi_bayesMult,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

```{r ebf_spieceasi_bayesMult_global_property_extract}
# Relative LCC size
props_ebf_spieceasi_bayesMult$globalPropsLCC$lccSizeRel1
props_ebf_spieceasi_bayesMult$globalPropsLCC$lccSizeRel2

# Natural connectivity
props_ebf_spieceasi_bayesMult$globalProps$natConnect1
props_ebf_spieceasi_bayesMult$globalProps$natConnect2

# Density
props_ebf_spieceasi_bayesMult$globalProps$density1
props_ebf_spieceasi_bayesMult$globalProps$density2
```


#### Association Histogram

```{r ebf_spieceasi_bayesMult_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_spieceasi_bayesMult)
```

#### Degree Distribution

```{r ebf_spieceasi_bayesMult_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_spieceasi_bayesMult)
```

#### Graphlet Correlation Matrix

```{r ebf_spieceasi_bayesMult_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_spieceasi_bayesMult,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_spieceasi_bayesMult_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_bayesMult,
  same_layout = FALSE,
  labels = labels_ebf_spieceasi_bayesMult,
  repulsion = 0.699
)
```

```{r ebf_spieceasi_bayesMult_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_bayesMult,
  same_layout = TRUE,
  labels = labels_ebf_spieceasi_bayesMult,
  repulsion = 0.79999
)
```

```{r ebf_spieceasi_bayesMult_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_bayesMult,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_spieceasi_bayesMult),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```



### SPIEC-EASI SLR

```{r generate_net_ebf_slr}
path_net_ebf_slr <- path_target(paste0("net_ebf_slr_r", slr_rank, ".rds"))

if (file.exists(path_net_ebf_slr) && !refit_network_comparison) {
  net_ebf_slr <- readRDS(path_net_ebf_slr)
} else {
  net_ebf_slr <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multRepl,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "slr",
                      nlambda = 30,
                      r = slr_rank,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_slr, path_net_ebf_slr)
}

objects_ebf_slr <- extract_group_network_objects(net_ebf_slr)
props_ebf_slr <- analyze_group_network(net_ebf_slr)
labels_ebf_slr <- build_plot_labels(props_ebf_slr)
```

#### netAnalyze summary

```{r ebf_slr_netAnalyze_summary}
print(summary(props_ebf_slr,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

#### Low-Rank Heatmaps

```{r ebf_slr_low_rank_prepare}
extract_slr_low_rank <- function(spieceasi_fit, taxa_names) {
  opt_ind <- pulsar::opt.index(
    spieceasi_fit$select,
    criterion = "stars"
  )

  L_opt <- as.matrix(spieceasi_fit$est$resid[[opt_ind]])

  if (length(taxa_names) != nrow(L_opt) || length(taxa_names) != ncol(L_opt)) {
    stop(
      "Taxon names from the SLR association matrix do not match the low-rank ",
      "matrix dimensions."
    )
  }

  dimnames(L_opt) <- list(taxa_names, taxa_names)

  list(
    L = L_opt,
    lambda = spieceasi_fit$est$lambda[opt_ind]
  )
}

slr_low_rank_ebf <- list(
  "EBF = 0" = extract_slr_low_rank(
    net_ebf_slr$measureOut1,
    taxa_names = colnames(net_ebf_slr$assoMat1)
  ),
  "EBF >= 2" = extract_slr_low_rank(
    net_ebf_slr$measureOut2,
    taxa_names = colnames(net_ebf_slr$assoMat2)
  )
)

common_taxa_slr <- Reduce(
  intersect,
  lapply(slr_low_rank_ebf, function(x) colnames(x$L))
)

if (length(common_taxa_slr) == 0) {
  stop("No common taxa found between the two SLR low-rank matrices.")
}

slr_low_rank_summary <- tibble(
  group = names(slr_low_rank_ebf),
  n_taxa = vapply(slr_low_rank_ebf, function(x) ncol(x$L), integer(1)),
  n_common_taxa = length(common_taxa_slr),
  lambda = vapply(slr_low_rank_ebf, function(x) x$lambda, numeric(1))
)

slr_low_rank_summary

slr_low_rank_ebf <- lapply(slr_low_rank_ebf, function(x) {
  x$L <- x$L[common_taxa_slr, common_taxa_slr, drop = FALSE]
  x
})

L_order <- Reduce(
  "+",
  lapply(slr_low_rank_ebf, function(x) {
    L <- x$L
    diag(L) <- 0
    L
  })
) / length(slr_low_rank_ebf)

slr_low_rank_cor <- stats::cor(L_order, use = "pairwise.complete.obs")
slr_low_rank_diss <- 1 - slr_low_rank_cor
slr_low_rank_diss[!is.finite(slr_low_rank_diss)] <- 1
diag(slr_low_rank_diss) <- 0

taxon_order_slr <- hclust(
  as.dist(slr_low_rank_diss)
)$order

all_slr_low_rank_values <- unlist(lapply(slr_low_rank_ebf, function(x) {
  L <- x$L
  L[row(L) == col(L)] <- NA_real_
  L[upper.tri(L)]
}))

lim_slr_low_rank <- max(abs(all_slr_low_rank_values), na.rm = TRUE)
if (!is.finite(lim_slr_low_rank) || lim_slr_low_rank == 0) {
  lim_slr_low_rank <- 1
}

slr_low_rank_heatmap_grobs <- lapply(names(slr_low_rank_ebf), function(group_name) {
  x <- slr_low_rank_ebf[[group_name]]
  L <- x$L[taxon_order_slr, taxon_order_slr, drop = FALSE]
  diag(L) <- NA_real_

  pheatmap(
    mat = L,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    border_color = NA,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(101),
    breaks = seq(-lim_slr_low_rank, lim_slr_low_rank, length.out = 102),
    main = paste0(
      group_name,
      ": rank ",
      slr_rank,
      " (StARS-selected lambda = ",
      formatC(x$lambda, format = "f", digits = 3),
      ")"
    ),
    silent = TRUE
  )$gtable
})
```

```{r ebf_slr_low_rank_heatmaps, fig.width=10, fig.height=5, dpi=250}
gridExtra::grid.arrange(
  grobs = slr_low_rank_heatmap_grobs,
  ncol = 2
)
```

#### Association Histogram

```{r ebf_slr_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_slr)
```

#### Degree Distribution

```{r ebf_slr_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_slr)
```

#### Graphlet Correlation Matrix

```{r ebf_slr_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_slr,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_slr_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_slr, 
                   same_layout = FALSE, 
                   labels = labels_ebf_slr,
                   repulsion = 0.7,
                   labelScale = TRUE,
                   cexLabels = 4)
```

```{r ebf_slr_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_slr, 
                   same_layout = TRUE, 
                   labels = labels_ebf_slr,
                   repulsion = 0.7,
                   labelScale = TRUE,
                   cexLabels = 4)
```

```{r ebf_slr_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_slr,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_slr),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```

### SPRING

```{r generate_net_ebf_spring}
path_net_ebf_spring <- path_target("net_ebf_spring.rds")

if (file.exists(path_net_ebf_spring) && !refit_network_comparison) {
  net_ebf_spring <- readRDS(path_net_ebf_spring)
} else {
  net_ebf_spring <- NetCoMi::netConstruct(
    data = counts_ebf_raw,
    group = ebf_group,
    measure = "spring",
    measurePar = list(nlambda = 30,
                      rep.num = 30,
                      thresh = 0.1,
                      Rmethod = "approx"),
    jointPrepro = FALSE,
    filtTax = c("totalReads", "numbSamp"),
    filtTaxPar = list(totalReads = 100, numbSamp = 5),
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_spring, path_net_ebf_spring)
}

objects_ebf_spring <- extract_group_network_objects(net_ebf_spring)
props_ebf_spring <- analyze_group_network(net_ebf_spring)
labels_ebf_spring <- build_plot_labels(props_ebf_spring)
```

#### netAnalyze summary

```{r ebf_spring_netAnalyze_summary}
print(summary(props_ebf_spring,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

```{r}
dim(net_ebf_spring$countMat1)
dim(net_ebf_spring$countMat2)
```

The filtering reduced the data set to 65 taxa in each group.

#### Association Histogram

```{r ebf_spring_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_spring)
```

#### Degree Distribution

```{r ebf_spring_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_spring)
```

#### Graphlet Correlation Matrix

```{r ebf_spring_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_spring,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_spring_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_spring, same_layout = FALSE, 
                        labels = labels_ebf_spring, repulsion = 0.8599, 
                        doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

layout1["Anaerostipes", ] <- c(0.31, 0.33)
layout1["Enterococcus", ] <- c(1, -0.23)
layout1["24_Enterobacteriaceae(F)", ] <- c(0.97, 0.19)

layout2["Bacteroides", ] <- c(0.14, 0.94)

plot_group_network(props_ebf_spring, same_layout = FALSE, 
                   labels = labels_ebf_spring, layout = list(layout1, layout2))
```

```{r ebf_spring_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_spring, same_layout = TRUE, 
                        labels = labels_ebf_spring, repulsion = 0.89999999,
                        doPlot = FALSE)
layout <- p$layout$layout1

layout["Blautia",  2] <- 0.53
layout["Bifidobacterium", 2] <- 0.24
layout["Fusicatenibacter", 2] <- 0.16
layout["Bacteroides", ] <- c(-0.75, -0.85)
layout["Clostridium_sensu_strict1", 1] <- 0.65

plot_group_network(props_ebf_spring, same_layout = TRUE, 
                   labels = labels_ebf_spring, layout = layout)
```

```{r ebf_spring_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spring, 
  same_layout = TRUE, 
  labels = blank_plot_labels(labels_ebf_spring),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```


### rhoshrink with multRepl

```{r generate_net_ebf_rhoshrink_multRepl}
path_net_ebf_rhoshrink_multRepl <- path_target("net_ebf_rhoshrink_multRepl.rds")

if (file.exists(path_net_ebf_rhoshrink_multRepl) && !refit_network_comparison) {
  net_ebf_rhoshrink_multRepl <- readRDS(path_net_ebf_rhoshrink_multRepl)
} else {
  net_ebf_rhoshrink_multRepl <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multRepl,
    group = ebf_group,
    measure = "rhoshrink",
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "threshold",
    thresh = 0.3,
    seed = network_seed
  )

  saveRDS(net_ebf_rhoshrink_multRepl, path_net_ebf_rhoshrink_multRepl)
}

objects_ebf_rhoshrink_multRepl <- extract_group_network_objects(net_ebf_rhoshrink_multRepl)
props_ebf_rhoshrink_multRepl <- analyze_group_network(net_ebf_rhoshrink_multRepl)
labels_ebf_rhoshrink_multRepl <- build_plot_labels(props_ebf_rhoshrink_multRepl)
```

#### netAnalyze summary

```{r ebf_rhoshrink_multRepl_netAnalyze_summary}
print(summary(props_ebf_rhoshrink_multRepl,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

#### Association Histogram

```{r ebf_rhoshrink_multRepl_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_rhoshrink_multRepl)
```

#### Degree Distribution

```{r ebf_rhoshrink_multRepl_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_rhoshrink_multRepl)
```

#### Graphlet Correlation Matrix

```{r ebf_rhoshrink_multRepl_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_rhoshrink_multRepl,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_rhoshrink_multRepl_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_rhoshrink_multRepl, same_layout = FALSE, 
                        labels = labels_ebf_rhoshrink_multRepl, repulsion = 0.89999, 
                        doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

plot_group_network(props_ebf_rhoshrink_multRepl, same_layout = FALSE, 
                   labels = labels_ebf_rhoshrink_multRepl, layout = list(layout1, layout2))
```

```{r ebf_rhoshrink_multRepl_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_rhoshrink_multRepl, same_layout = TRUE, 
                        labels = labels_ebf_rhoshrink_multRepl,
                        doPlot = FALSE)
layout <- p$layout$layout1

plot_group_network(props_ebf_rhoshrink_multRepl, same_layout = TRUE, 
                   labels = labels_ebf_rhoshrink_multRepl, layout = layout,
                   phylum_legend_x = -0.18,
                   phylum_legend_y = 1.35,
                   asso_legend_x   = -0.15,
                   asso_legend_y   = -0.8)
```

```{r ebf_rhoshrink_multRepl_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_rhoshrink_multRepl,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_rhoshrink_multRepl),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```

### rhoshrink with bayesMult replacement

```{r generate_net_ebf_rhoshrink_bayesMult}
path_net_ebf_rhoshrink_bayesMult <- path_target("net_ebf_rhoshrink_bayesMult.rds")

if (file.exists(path_net_ebf_rhoshrink_bayesMult) && !refit_network_comparison) {
  net_ebf_rhoshrink_bayesMult <- readRDS(path_net_ebf_rhoshrink_bayesMult)
} else {
  net_ebf_rhoshrink_bayesMult <- NetCoMi::netConstruct(
    data = counts_ebf_rel_bayesMult,
    group = ebf_group,
    measure = "rhoshrink",
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "threshold",
    thresh = 0.15,
    seed = network_seed
  )

  saveRDS(net_ebf_rhoshrink_bayesMult, path_net_ebf_rhoshrink_bayesMult)
}

objects_ebf_rhoshrink_bayesMult <- extract_group_network_objects(net_ebf_rhoshrink_bayesMult)
props_ebf_rhoshrink_bayesMult <- analyze_group_network(net_ebf_rhoshrink_bayesMult)
labels_ebf_rhoshrink_bayesMult <- build_plot_labels(props_ebf_rhoshrink_bayesMult)
```

#### netAnalyze summary

```{r ebf_rhoshrink_bayesMult_netAnalyze_summary}
print(summary(props_ebf_rhoshrink_bayesMult,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

#### Association Histogram

```{r ebf_rhoshrink_bayesMult_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_rhoshrink_bayesMult)
```

#### Degree Distribution

```{r ebf_rhoshrink_bayesMult_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_rhoshrink_bayesMult)
```

#### Graphlet Correlation Matrix

```{r ebf_rhoshrink_bayesMult_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_rhoshrink_bayesMult,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_rhoshrink_bayesMult_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_rhoshrink_bayesMult, same_layout = FALSE, 
                        labels = labels_ebf_rhoshrink_bayesMult, repulsion = 0.899999, 
                        doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

layout2["Bifidobacterium", 2] <- 0.96
layout1["Ruminococcus", 2] <- 0.82

plot_group_network(props_ebf_rhoshrink_bayesMult, same_layout = FALSE, 
                   labels = labels_ebf_rhoshrink_bayesMult, layout = list(layout1, layout2))
```

```{r ebf_rhoshrink_bayesMult_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_rhoshrink_bayesMult, same_layout = TRUE, 
                        labels = labels_ebf_rhoshrink_bayesMult,
                        doPlot = FALSE)
layout <- p$layout$layout1

layout["Escherichia-Shigella",  2] <- -0.54
layout["[Ruminococcus]_gauvreauii_group", 2] <- -0.34

plot_group_network(props_ebf_rhoshrink_bayesMult, same_layout = TRUE, 
                   labels = labels_ebf_rhoshrink_bayesMult, layout = layout,
                   phylum_legend_x = -0.18,
                   phylum_legend_y = 1.35,
                   asso_legend_x   = -0.15,
                   asso_legend_y   = -0.8)
```

```{r ebf_rhoshrink_bayesMult_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_rhoshrink_bayesMult,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_rhoshrink_bayesMult),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```



### corshrink with multiplicative replacement

```{r generate_net_ebf_corshrink_multRepl}
path_net_ebf_corshrink_multRepl <- path_target("net_ebf_corshrink_multRepl.rds")

if (file.exists(path_net_ebf_corshrink_multRepl) && !refit_network_comparison) {
  net_ebf_corshrink_multRepl <- readRDS(path_net_ebf_corshrink_multRepl)
} else {
  net_ebf_corshrink_multRepl <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multRepl,
    group = ebf_group,
    measure = "corshrink",
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "clr",
    zeroMethod = "none",
    sparsMethod = "threshold",
    thresh = 0.3,
    seed = network_seed
  )

  saveRDS(net_ebf_corshrink_multRepl, path_net_ebf_corshrink_multRepl)
}

objects_ebf_corshrink_multRepl <- extract_group_network_objects(net_ebf_corshrink_multRepl)
props_ebf_corshrink_multRepl <- analyze_group_network(net_ebf_corshrink_multRepl)
labels_ebf_corshrink_multRepl <- build_plot_labels(props_ebf_corshrink_multRepl)
```

#### netAnalyze summary

```{r ebf_corshrink_multRepl_netAnalyze_summary}
print(summary(props_ebf_corshrink_multRepl,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

#### Association Histogram

```{r ebf_corshrink_multRepl_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_corshrink_multRepl)
```

#### Degree Distribution

```{r ebf_corshrink_multRepl_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_corshrink_multRepl)
```

#### Graphlet Correlation Matrix

```{r ebf_corshrink_multRepl_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_corshrink_multRepl,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_corshrink_multRepl_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_corshrink_multRepl, same_layout = FALSE, 
                        labels = labels_ebf_corshrink_multRepl, repulsion = 0.8, 
                        doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

plot_group_network(props_ebf_corshrink_multRepl, same_layout = FALSE, 
                   labels = labels_ebf_corshrink_multRepl, layout = list(layout1, layout2))
```


```{r ebf_corshrink_multRepl_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_corshrink_multRepl, same_layout = TRUE, 
                        labels = labels_ebf_corshrink_multRepl,
                        doPlot = FALSE)

layout <- p$layout$layout1

plot_group_network(props_ebf_corshrink_multRepl, same_layout = TRUE, 
                   labels = labels_ebf_corshrink_multRepl, layout = layout,
                   phylum_legend_x = -0.18,
                   phylum_legend_y = 1.35,
                   asso_legend_x   = -0.15,
                   asso_legend_y   = -0.8)
```

```{r ebf_corshrink_multRepl_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_corshrink_multRepl,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_corshrink_multRepl),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```


### corshrink with bayesMult replacement

```{r generate_net_ebf_corshrink_bayesMult}
path_net_ebf_corshrink_bayesMult <- path_target("net_ebf_corshrink_bayesMult.rds")

if (file.exists(path_net_ebf_corshrink_bayesMult) && !refit_network_comparison) {
  net_ebf_corshrink_bayesMult <- readRDS(path_net_ebf_corshrink_bayesMult)
} else {
  net_ebf_corshrink_bayesMult <- NetCoMi::netConstruct(
    data = counts_ebf_rel_bayesMult,
    group = ebf_group,
    measure = "corshrink",
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "clr",
    zeroMethod = "none",
    sparsMethod = "threshold",
    thresh = 0.15,
    seed = network_seed
  )

  saveRDS(net_ebf_corshrink_bayesMult, path_net_ebf_corshrink_bayesMult)
}

objects_ebf_corshrink_bayesMult <- extract_group_network_objects(net_ebf_corshrink_bayesMult)
props_ebf_corshrink_bayesMult <- analyze_group_network(net_ebf_corshrink_bayesMult)
labels_ebf_corshrink_bayesMult <- build_plot_labels(props_ebf_corshrink_bayesMult)
```

#### netAnalyze summary

```{r ebf_corshrink_bayesMult_netAnalyze_summary}
print(summary(props_ebf_corshrink_bayesMult,
              groupNames = c("EBF = 0", "EBF >= 2")))
```

#### Association Histogram

```{r ebf_corshrink_bayesMult_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_corshrink_bayesMult)
```

#### Degree Distribution

```{r ebf_corshrink_bayesMult_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_corshrink_bayesMult)
```

#### Graphlet Correlation Matrix

```{r ebf_corshrink_bayesMult_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_corshrink_bayesMult,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_corshrink_bayesMult_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_corshrink_bayesMult, same_layout = FALSE, 
                        labels = labels_ebf_corshrink_bayesMult, repulsion = 0.899, 
                        doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

layout2["Bacteroides", 2] <- 0.43
layout2["Bifidobacterium", 2] <- -0.46

plot_group_network(props_ebf_corshrink_bayesMult, same_layout = FALSE, 
                   labels = labels_ebf_corshrink_bayesMult, layout = list(layout1, layout2))
```


```{r ebf_corshrink_bayesMult_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_corshrink_bayesMult, same_layout = TRUE, 
                        labels = labels_ebf_corshrink_bayesMult,
                        doPlot = FALSE)
layout <- p$layout$layout1
layout["Phascolarctobacterium", 2] <- -0.58
layout["Clostridium_sensu_strict1", 2] <- -0.75
layout["Escherichia-Shigella", 2] <- -0.68

plot_group_network(props_ebf_corshrink_bayesMult, same_layout = TRUE, 
                   labels = labels_ebf_corshrink_bayesMult, layout = layout,
                   phylum_legend_x = -0.18,
                   phylum_legend_y = 1.35,
                   asso_legend_x   = -0.15,
                   asso_legend_y   = -0.8)
```

```{r ebf_corshrink_bayesMult_network_plot_nolabs, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_corshrink_bayesMult,
  same_layout = TRUE,
  labels = blank_plot_labels(labels_ebf_corshrink_bayesMult),
  legends = FALSE,
  mar = c(1, 1.5, 3, 1.5)
)
```

## Comparison Of Global Network Properties

The following tables summarize three global network properties for the
construction methods shown in the thesis.

```{r descriptive_global_network_properties}
descriptive_global_properties <- bind_rows(
  extract_global_network_properties(
    props_ebf_spieceasi_bayesMult,
    method = "SPIEC-EASI MB bayesMult"
  ),
  extract_global_network_properties(
    props_ebf_slr,
    method = "SPIEC-EASI SLR"
  ),
  extract_global_network_properties(
    props_ebf_spring,
    method = "SPRING"
  ),
  extract_global_network_properties(
    props_ebf_corshrink_multRepl,
    method = "corshrink multRepl"
  ),
  extract_global_network_properties(
    props_ebf_corshrink_bayesMult,
    method = "corshrink bayesMult"
  ),
  extract_global_network_properties(
    props_ebf_rhoshrink_multRepl,
    method = "rhoshrink multRepl"
  ),
  extract_global_network_properties(
    props_ebf_rhoshrink_bayesMult,
    method = "rhoshrink bayesMult"
  )
)
```

### Relative LCC Size

```{r descriptive_relative_lcc_size_table}
make_global_property_table(
  descriptive_global_properties,
  property = "relative_lcc_size"
)
```

### Density

```{r descriptive_density_table}
make_global_property_table(
  descriptive_global_properties,
  property = "density"
)
```

### Natural Connectivity

```{r descriptive_natural_connectivity_table}
make_global_property_table(
  descriptive_global_properties,
  property = "natural_connectivity"
)
```


## Select Network For Permutation Testing

The permutation-testing part can be run for any network constructed in the
descriptive comparison above. Change `selected_permutation_method` in the
parameter chunk to switch the network and the folder used for all
permutation-related files.

```{r select_permutation_network}
permutation_networks <- list(
  spieceasi_multRepl = list(
    label = "SPIEC-EASI MB with multRepl",
    net = net_ebf_spieceasi_multRepl,
    props = props_ebf_spieceasi_multRepl
  ),
  spieceasi_bayesMult = list(
    label = "SPIEC-EASI MB with bayesMult",
    net = net_ebf_spieceasi_bayesMult,
    props = props_ebf_spieceasi_bayesMult
  ),
  spieceasi_slr = list(
    label = paste0("SPIEC-EASI SLR, rank ", slr_rank),
    net = net_ebf_slr,
    props = props_ebf_slr
  ),
  spring = list(
    label = "SPRING",
    net = net_ebf_spring,
    props = props_ebf_spring
  ),
  rhoshrink_multRepl = list(
    label = "rhoshrink with multRepl",
    net = net_ebf_rhoshrink_multRepl,
    props = props_ebf_rhoshrink_multRepl
  ),
  rhoshrink_bayesMult = list(
    label = "rhoshrink with bayesMult",
    net = net_ebf_rhoshrink_bayesMult,
    props = props_ebf_rhoshrink_bayesMult
  ),
  corshrink_multRepl = list(
    label = "corshrink with multRepl",
    net = net_ebf_corshrink_multRepl,
    props = props_ebf_corshrink_multRepl
  ),
  corshrink_bayesMult = list(
    label = "corshrink with bayesMult",
    net = net_ebf_corshrink_bayesMult,
    props = props_ebf_corshrink_bayesMult
  )
)

if (!selected_permutation_method %in% names(permutation_networks)) {
  stop(
    "Unknown selected_permutation_method. Choose one of: ",
    paste(names(permutation_networks), collapse = ", ")
  )
}

selected_permutation_network <- permutation_networks[[selected_permutation_method]]
net_ebf_perm <- selected_permutation_network$net
props_ebf_perm <- selected_permutation_network$props

permutation_output_dir <- path_target(
  file.path("permutation_tests", selected_permutation_method)
)
dir.create(permutation_output_dir, recursive = TRUE, showWarnings = FALSE)

path_perm <- function(...) {
  file.path(permutation_output_dir, ...)
}

tibble(
  selected_method = selected_permutation_method,
  selected_label = selected_permutation_network$label,
  n_taxa_group1 = ncol(props_ebf_perm$input$adjaMat1),
  n_taxa_group2 = ncol(props_ebf_perm$input$adjaMat2),
  output_dir = permutation_output_dir
)
```

## Generate Or Reuse Permuted Association Matrices

The following chunks are kept for reproducibility and are not evaluated during
normal rendering. The final filematrix at `path_perm("assoPerm_ebf")` is used
by both `netCompare()` and `diffnet()`.

```{r generate_permuted_association_matrices, eval=FALSE}
permGroupMat <- createAssoPerm(
  props_ebf_perm,
  nPerm = 1000,
  computeAsso = FALSE,
  seed = permutation_seed
)

nPerm_all <- 1000
blocksize <- 20
repetitions <- nPerm_all / blocksize
perm_asso_dir <- path_perm("perm_asso_matrices")
dir.create(perm_asso_dir, recursive = TRUE, showWarnings = FALSE)

cores_perm <- 8

plan(multisession, workers = cores_perm)
registerDoFuture()

with_progress({
  p <- progressr::progressor(along = 1:repetitions)

  tmp <- foreach(i = 1:repetitions,
                 .packages = c("NetCoMi")) %dorng% {
    p()
    file_base <- file.path(perm_asso_dir, paste0("assoPerm_ebf", i))
    bmat_file <- paste0(file_base, ".bmat")
    if (!file.exists(bmat_file) || refit_permutation_matrices) {
      set.seed(permutation_seed + i)
      NetCoMi::createAssoPerm(
        props_ebf_perm,
        nPerm = blocksize,
        permGroupMat = permGroupMat[(i - 1) * blocksize + 1:blocksize, ],
        computeAsso = TRUE,
        fileStoreAssoPerm = file_base,
        storeCountsPerm = FALSE,
        append = FALSE
      )
    }
  }
})

plan(sequential)
```

```{r combine_permuted_association_matrices, eval=FALSE}
assoPerm_all <- NULL
perm_asso_dir <- path_perm("perm_asso_matrices")

for (i in 1:repetitions) {
  file_base <- file.path(perm_asso_dir, paste0("assoPerm_ebf", i))
  assoPerm_tmp <- fm.open(filenamebase = file_base, readonly = TRUE)
  assoPerm_all <- rbind(assoPerm_all, as.matrix(assoPerm_tmp))
  close(assoPerm_tmp)
}

dim(assoPerm_all)

n_perm_netcompare <- n_perm_network_comparison
n_taxa_netcompare <- nrow(props_ebf_perm$input$adjaMat1)
n_rows_netcompare <- n_perm_netcompare * n_taxa_netcompare

stopifnot(nrow(assoPerm_all) >= n_rows_netcompare)
stopifnot(nrow(assoPerm_all) %% n_taxa_netcompare == 0)

assoPerm_netcompare <- assoPerm_all[seq_len(n_rows_netcompare), , drop = FALSE]
dim(assoPerm_netcompare)

fm.create.from.matrix(
  filenamebase = path_perm("assoPerm_ebf"),
  mat = assoPerm_netcompare
)
```

```{r check_permuted_association_matrices}
asso_perm_base <- path_perm("assoPerm_ebf")

if (!file.exists(paste0(asso_perm_base, ".bmat"))) {
  warning(
    "Permutation association filematrix not found at path_perm('assoPerm_ebf'). ",
    "Run the unevaluated permutation-generation chunks before refitting ",
    "permutation tests."
  )
} else {
  assoPerm_check <- tryCatch(
    fm.open(filenamebase = asso_perm_base, readonly = TRUE),
    error = function(e) e
  )

  if (inherits(assoPerm_check, "error")) {
    warning("Could not open permutation association filematrix: ",
            conditionMessage(assoPerm_check))
  } else {
    expected_rows <- n_perm_network_comparison *
      nrow(props_ebf_perm$input$adjaMat1)
    expected_cols <- 2 * ncol(props_ebf_perm$input$adjaMat1)

    stopifnot(nrow(assoPerm_check) == expected_rows)
    stopifnot(ncol(assoPerm_check) == expected_cols)
    close(assoPerm_check)
  }
}
```

## Permutation-Based Differential Network Analysis

We compare local and global network characteristics with NetCoMi's
`netCompare()` using empirical p-values, unconstrained `permApprox` refinement,
and constrained `permApprox` refinement.

### Empirical P-Values

```{r run_netcompare_emp_noBH}
path_comp_ebf_emp_noBH <- path_perm("comp_ebf_emp_noBH.rds")

if (file.exists(path_comp_ebf_emp_noBH) && !refit_permutation_tests) {
  comp_ebf_emp_noBH <- readRDS(path_comp_ebf_emp_noBH)
} else {
  comp_ebf_emp_noBH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_emp_noBH, path_comp_ebf_emp_noBH)
}

print(summary(comp_ebf_emp_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_emp_BH}
path_comp_ebf_emp_BH <- path_perm("comp_ebf_emp_BH.rds")

if (file.exists(path_comp_ebf_emp_BH) && !refit_permutation_tests) {
  comp_ebf_emp_BH <- readRDS(path_comp_ebf_emp_BH)
} else {
  comp_ebf_emp_BH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_emp_BH, path_comp_ebf_emp_BH)
}

print(summary(comp_ebf_emp_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

### permApprox Refinement Without Constraint

```{r run_netcompare_GPD_noConstr_noBH}
path_comp_ebf_GPD_noConstr_noBH <- path_perm("comp_ebf_GPD_noConstr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_noBH) && !refit_permutation_tests) {
  comp_ebf_GPD_noConstr_noBH <- readRDS(path_comp_ebf_GPD_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  comp_ebf_GPD_noConstr_noBH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_noConstr_noBH, path_comp_ebf_GPD_noConstr_noBH)
}

print(summary(comp_ebf_GPD_noConstr_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_GPD_noConstr_BH}
path_comp_ebf_GPD_noConstr_BH <- path_perm("comp_ebf_GPD_noConstr_BH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_BH) && !refit_permutation_tests) {
  comp_ebf_GPD_noConstr_BH <- readRDS(path_comp_ebf_GPD_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  comp_ebf_GPD_noConstr_BH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_noConstr_BH, path_comp_ebf_GPD_noConstr_BH)
}

print(summary(comp_ebf_GPD_noConstr_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

### permApprox Refinement With Constraint

```{r run_netcompare_GPD_constr_noBH}
path_comp_ebf_GPD_constr_noBH <- path_perm("comp_ebf_GPD_constr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_constr_noBH) && !refit_permutation_tests) {
  comp_ebf_GPD_constr_noBH <- readRDS(path_comp_ebf_GPD_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  comp_ebf_GPD_constr_noBH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_constr_noBH, path_comp_ebf_GPD_constr_noBH)
}

print(summary(comp_ebf_GPD_constr_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_GPD_constr_BH}
path_comp_ebf_GPD_constr_BH <- path_perm("comp_ebf_GPD_constr_BH.rds")

if (file.exists(path_comp_ebf_GPD_constr_BH) && !refit_permutation_tests) {
  comp_ebf_GPD_constr_BH <- readRDS(path_comp_ebf_GPD_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  comp_ebf_GPD_constr_BH <- netCompare(
    props_ebf_perm,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_constr_BH, path_comp_ebf_GPD_constr_BH)
}

print(summary(comp_ebf_GPD_constr_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

### Combined netCompare Summary

```{r netcompare_combined_summary_helpers}
normalise_result_name <- function(x) {
  stringr::str_replace_all(tolower(x), "[^a-z0-9]", "")
}

find_numeric_field <- function(x, candidates) {
  candidate_keys <- normalise_result_name(candidates)

  search_one <- function(obj) {
    if (is.null(obj)) {
      return(NULL)
    }

    obj_names <- names(obj)
    if (!is.null(obj_names)) {
      obj_keys <- normalise_result_name(obj_names)
      hit <- which(obj_keys %in% candidate_keys)
      if (length(hit) > 0) {
        value <- obj[[hit[1]]]
        if (is.numeric(value) && length(value) >= 1) {
          return(as.numeric(value[1]))
        }
      }
    }

    if (is.list(obj)) {
      for (i in seq_along(obj)) {
        value <- search_one(obj[[i]])
        if (!is.null(value)) {
          return(value)
        }
      }
    }

    NULL
  }

  value <- search_one(x)
  if (is.null(value)) {
    NA_real_
  } else {
    value
  }
}

get_global_pvalue <- function(comp, candidates, refined = FALSE, lcc = FALSE) {
  containers <- if (lcc) {
    if (refined) {
      list(comp$pvalDiffGlobalLCCRefined, comp$pvalDiffGlobalLCCEmpirical)
    } else {
      list(comp$pvalDiffGlobalLCCEmpirical)
    }
  } else {
    if (refined) {
      list(comp$pvalDiffGlobalRefined, comp$pvalDiffGlobalEmpirical)
    } else {
      list(comp$pvalDiffGlobalEmpirical)
    }
  }

  for (container in containers) {
    p_value <- find_numeric_field(container, candidates)
    if (is.finite(p_value)) {
      return(p_value)
    }
  }

  NA_real_
}

get_prop_value <- function(props, containers, candidates, group) {
  containers <- unlist(containers, recursive = FALSE)
  candidates <- unlist(candidates)
  suffix <- if (identical(group, 1L)) "1" else "2"
  candidates <- unique(c(
    paste0(candidates, suffix),
    candidates
  ))

  for (container_name in containers) {
    value <- find_numeric_field(props[[container_name]], candidates)
    if (is.finite(value)) {
      return(value)
    }
  }

  NA_real_
}

make_global_summary_rows <- function(rows, lcc = FALSE) {
  purrr::pmap_dfr(rows, function(label, containers, fields, p_fields, ...) {
    p_fields <- unlist(p_fields)
    value1 <- get_prop_value(props_ebf_perm, containers, fields, group = 1L)
    value2 <- get_prop_value(props_ebf_perm, containers, fields, group = 2L)

    tibble(
      measure = label,
      group_ebf_0 = value1,
      group_ebf_ge2 = value2,
      abs_difference = abs(value1 - value2),
      p_empirical = get_global_pvalue(
        comp_ebf_emp_noBH,
        p_fields,
        refined = FALSE,
        lcc = lcc
      ),
      p_unconstrained = get_global_pvalue(
        comp_ebf_GPD_noConstr_noBH,
        p_fields,
        refined = TRUE,
        lcc = lcc
      ),
      p_constrained = get_global_pvalue(
        comp_ebf_GPD_constr_noBH,
        p_fields,
        refined = TRUE,
        lcc = lcc
      )
    )
  })
}

format_netcompare_table <- function(df, title, name_width = 30) {
  cat(title, "\n")
  cat(strrep("~", nchar(title)), "\n\n")
  cat(sprintf(
    paste0("%-", name_width, "s %10s %10s %10s %10s %10s %10s\n"),
    "",
    "EBF = 0",
    "EBF >= 2",
    "abs.diff.",
    "p.emp",
    "p.uncon.",
    "p.constr."
  ))

  purrr::pwalk(df, function(measure, group_ebf_0, group_ebf_ge2,
                            abs_difference, p_empirical, p_unconstrained,
                            p_constrained, ...) {
    cat(sprintf(
      paste0("%-", name_width, "s %10.3f %10.3f %10.3f %10.6f %10.6f %10.6f\n"),
      stringr::str_trunc(measure, name_width, ellipsis = ""),
      group_ebf_0,
      group_ebf_ge2,
      abs_difference,
      p_empirical,
      p_unconstrained,
      p_constrained
    ))
  })

  cat("\n")
}

get_centrality_pvalues_wide <- function(comp_emp, comp_unconstr, comp_constr) {
  centrality_fields <- list(
    degree = c("pvalDiffDeg", "pvalDiffDegree"),
    betweenness = c("pvalDiffBetw", "pvalDiffBetween", "pvalDiffBetweenness"),
    closeness = c("pvalDiffClose", "pvalDiffCloseness"),
    eigenvector = c("pvalDiffEigen", "pvalDiffEigenv", "pvalDiffEigenvector")
  )

  get_pvector <- function(x, fields) {
    for (field in fields) {
      if (!is.null(x[[field]])) {
        return(x[[field]])
      }
    }
    NULL
  }

  pvalues <- purrr::imap_dfr(centrality_fields, function(fields, centrality_measure) {
    p_emp <- get_pvector(comp_emp$pvalDiffCentrEmpirical, fields)
    p_unconstr <- get_pvector(comp_unconstr$pvalDiffCentrRefined, fields)
    p_constr <- get_pvector(comp_constr$pvalDiffCentrRefined, fields)

    if (is.null(p_emp) || is.null(p_unconstr) || is.null(p_constr)) {
      return(tibble())
    }

    genus <- names(p_emp)
    if (is.null(genus)) {
      genus <- as.character(seq_along(p_emp))
    }

    tibble(
      centrality_measure = centrality_measure,
      genus = genus,
      p_empirical = as.numeric(p_emp),
      p_unconstrained = as.numeric(p_unconstr),
      p_constrained = as.numeric(p_constr)
    )
  })

  if (nrow(pvalues) == 0) {
    return(tibble(
      centrality_measure = character(),
      genus = character(),
      p_empirical = numeric(),
      p_unconstrained = numeric(),
      p_constrained = numeric()
    ))
  }

  pvalues
}

get_centrality_values_long <- function(props) {
  centrality_fields <- list(
    degree = list(group1 = c("degree1", "deg1"),
                  group2 = c("degree2", "deg2")),
    betweenness = list(group1 = c("between1", "betweenness1", "betw1"),
                       group2 = c("between2", "betweenness2", "betw2")),
    closeness = list(group1 = c("closeness1", "close1"),
                     group2 = c("closeness2", "close2")),
    eigenvector = list(group1 = c("eigenv1", "eigenvector1"),
                       group2 = c("eigenv2", "eigenvector2"))
  )

  get_vector <- function(fields) {
    for (field in fields) {
      if (!is.null(props$centralities[[field]])) {
        return(props$centralities[[field]])
      }
    }
    NULL
  }

  centrality_values <- purrr::imap_dfr(centrality_fields, function(fields, centrality_measure) {
    value1 <- get_vector(fields$group1)
    value2 <- get_vector(fields$group2)

    if (is.null(value1) || is.null(value2)) {
      return(tibble())
    }

    genera <- union(names(value1), names(value2))

    tibble(
      centrality_measure = centrality_measure,
      genus = genera,
      group_ebf_0 = as.numeric(value1[genera]),
      group_ebf_ge2 = as.numeric(value2[genera])
    )
  })

  if (nrow(centrality_values) == 0) {
    return(tibble(
      centrality_measure = character(),
      genus = character(),
      group_ebf_0 = numeric(),
      group_ebf_ge2 = numeric()
    ))
  }

  centrality_values
}

format_centrality_summary <- function(df, top_n = 5, name_width = 30) {
  centrality_labels <- c(
    degree = "Degree centrality",
    betweenness = "Betweenness centrality",
    closeness = "Closeness centrality",
    eigenvector = "Eigenvector centrality"
  )

  for (measure_id in unique(df$centrality_measure)) {
    title <- unname(centrality_labels[measure_id])
    if (is.na(title)) {
      title <- measure_id
    }
    cat(title, "\n")
    cat(strrep("~", nchar(title)), "\n\n")
    cat(sprintf(
      paste0("%-", name_width, "s %10s %10s %10s %10s %10s %10s\n"),
      "Genus",
      "EBF = 0",
      "EBF >= 2",
      "abs.diff.",
      "p.emp",
      "p.uncon.",
      "p.constr."
    ))

    df %>%
      filter(centrality_measure == measure_id) %>%
      arrange(min_raw_p) %>%
      slice_head(n = top_n) %>%
      purrr::pwalk(function(centrality_measure, genus, group_ebf_0,
                            group_ebf_ge2, abs_difference, p_empirical,
                            p_unconstrained, p_constrained, min_raw_p, ...) {
        cat(sprintf(
          paste0("%-", name_width, "s %10.3f %10.3f %10.3f %10.6f %10.6f %10.6f\n"),
          stringr::str_trunc(genus, name_width, ellipsis = ""),
          group_ebf_0,
          group_ebf_ge2,
          abs_difference,
          p_empirical,
          p_unconstrained,
          p_constrained
        ))
      })

    cat("\n")
  }
}
```

```{r netcompare_combined_summary_print}
lcc_global_rows <- tibble::tribble(
  ~label, ~containers, ~fields, ~p_fields,
  "Relative LCC size", list(c("globalPropsLCC")), list(c("lccSizeRel", "relLCCSize", "LCCSizeRel")),
  list(c("pvallccSizeRel", "pvallccsize")),
  "Clustering coefficient", list(c("globalPropsLCC")), list(c("clustCoef", "clusteringCoefficient")),
  list(c("pvalClustCoef")),
  "Modularity", list(c("globalPropsLCC")), list(c("modularity")),
  list(c("pvalModul")),
  "Positive edge percentage", list(c("globalPropsLCC")), list(c("pep", "posEdgePerc", "posEdgePercentage")),
  list(c("pvalPEP")),
  "Edge density", list(c("globalPropsLCC")), list(c("density", "edgeDensity")),
  list(c("pvalDensity")),
  "Natural connectivity", list(c("globalPropsLCC", "globalProps")), list(c("natConnect", "naturalConnectivity")),
  list(c("pvalNatConnect")),
  "Vertex connectivity", list(c("globalPropsLCC")), list(c("vertConnect", "vertexConnect", "vertexConnectivity")),
  list(c("pvalVertConnect")),
  "Edge connectivity", list(c("globalPropsLCC")), list(c("edgeConnect", "edgeConnectivity")),
  list(c("pvalEdgeConnect")),
  "Average dissimilarity*", list(c("globalPropsLCC")), list(c("avDiss", "avgDiss", "averageDissimilarity")),
  list(c("pvalavDiss")),
  "Average path length**", list(c("globalPropsLCC")), list(c("avPath", "avPathLength", "avgPathLength", "averagePathLength")),
  list(c("pvalavPath"))
)

whole_network_rows <- tibble::tribble(
  ~label, ~containers, ~fields, ~p_fields,
  "Number of components", list(c("globalProps")), list(c("nComp", "nComponents", "numberOfComponents")),
  list(c("pvalnComp")),
  "Clustering coefficient", list(c("globalProps")), list(c("clustCoef", "clusteringCoefficient")),
  list(c("pvalClustCoef")),
  "Modularity", list(c("globalProps")), list(c("modularity")),
  list(c("pvalModul")),
  "Positive edge percentage", list(c("globalProps")), list(c("pep", "posEdgePerc", "posEdgePercentage")),
  list(c("pvalPEP")),
  "Edge density", list(c("globalProps")), list(c("density", "edgeDensity")),
  list(c("pvalDensity")),
  "Natural connectivity", list(c("globalProps")), list(c("natConnect", "naturalConnectivity")),
  list(c("pvalNatConnect"))
)

netcompare_lcc_global_summary <- make_global_summary_rows(lcc_global_rows, lcc = TRUE)
netcompare_whole_global_summary <- make_global_summary_rows(whole_network_rows, lcc = FALSE)

netcompare_centrality_summary <- get_centrality_pvalues_wide(
  comp_emp = comp_ebf_emp_noBH,
  comp_unconstr = comp_ebf_GPD_noConstr_noBH,
  comp_constr = comp_ebf_GPD_constr_noBH
) %>%
  left_join(
    get_centrality_values_long(props_ebf_perm),
    by = c("centrality_measure", "genus")
  ) %>%
  mutate(
    abs_difference = abs(group_ebf_0 - group_ebf_ge2),
    min_raw_p = pmin(
      p_empirical,
      p_unconstrained,
      p_constrained,
      na.rm = TRUE
    ),
    min_raw_p = if_else(is.infinite(min_raw_p), NA_real_, min_raw_p)
  )

cat("Global network properties\n")
cat("=========================\n\n")
format_netcompare_table(
  netcompare_lcc_global_summary,
  title = "Largest connected component (LCC):"
)
format_netcompare_table(
  netcompare_whole_global_summary,
  title = "Whole network:"
)

cat("Centrality measures\n")
cat("===================\n\n")
format_centrality_summary(netcompare_centrality_summary, top_n = 5)

cat("p-values: one-tailed test with null hypothesis diff=0\n")
cat("  *: Dissimilarity = 1 - edge weight\n")
cat(" **: Path length = Units with average dissimilarity\n")
```

## Permutation-Based Differential Association Analysis

Differential association analysis is performed with `diffnet()` on the selected
network and the same stored permutation association matrices.

### Empirical P-Values

```{r run_diffnet_noBH}
path_diff_ebf_emp_noBH <- path_perm("diff_ebf_emp_noBH.rds")

if (file.exists(path_diff_ebf_emp_noBH) && !refit_permutation_tests) {
  diff_ebf_emp_noBH <- readRDS(path_diff_ebf_emp_noBH)
} else {
  diff_ebf_emp_noBH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_emp_noBH, path_diff_ebf_emp_noBH)
}

summary(diff_ebf_emp_noBH$pvalsVec)
```

```{r run_diffnet_BH}
path_diff_ebf_emp_BH <- path_perm("diff_ebf_emp_BH.rds")

if (file.exists(path_diff_ebf_emp_BH) && !refit_permutation_tests) {
  diff_ebf_emp_BH <- readRDS(path_diff_ebf_emp_BH)
} else {
  diff_ebf_emp_BH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    adjust = "BH",
    adjustNonzeroOnly = TRUE,
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_emp_BH, path_diff_ebf_emp_BH)
}

summary(diff_ebf_emp_BH$pAdjustVec)
```

```{r}
plot(diff_ebf_emp_BH)
```

### permApprox Refinement Without Constraint

```{r run_diffnet_noConstr_noBH}
path_diff_ebf_noConstr_noBH <- path_perm("diff_ebf_noConstr_noBH.rds")

if (file.exists(path_diff_ebf_noConstr_noBH) && !refit_permutation_tests) {
  diff_ebf_noConstr_noBH <- readRDS(path_diff_ebf_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  diff_ebf_noConstr_noBH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_noConstr_noBH, path_diff_ebf_noConstr_noBH)
}

summary(diff_ebf_noConstr_noBH$pvalsVec)
```

```{r run_diffnet_noConstr_BH}
path_diff_ebf_noConstr_BH <- path_perm("diff_ebf_noConstr_BH.rds")

if (file.exists(path_diff_ebf_noConstr_BH) && !refit_permutation_tests) {
  diff_ebf_noConstr_BH <- readRDS(path_diff_ebf_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  diff_ebf_noConstr_BH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    adjustNonzeroOnly = TRUE,
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_noConstr_BH, path_diff_ebf_noConstr_BH)
}

summary(diff_ebf_noConstr_BH$pAdjustVec)
```


```{r plot_diffnet_main_BH, fig.width=10, fig.height=7}
plot(diff_ebf_noConstr_BH, 
     edgeTransp = 0)
```


### permApprox Refinement With Constraint

```{r run_diffnet_constr_noBH}
path_diff_ebf_constr_noBH <- path_perm("diff_ebf_constr_noBH.rds")

if (file.exists(path_diff_ebf_constr_noBH) && !refit_permutation_tests) {
  diff_ebf_constr_noBH <- readRDS(path_diff_ebf_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  diff_ebf_constr_noBH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_constr_noBH, path_diff_ebf_constr_noBH)
}

summary(diff_ebf_constr_noBH$pvalsVec)
```

```{r run_diffnet_constr_BH}
path_diff_ebf_constr_BH <- path_perm("diff_ebf_constr_BH.rds")

if (file.exists(path_diff_ebf_constr_BH) && !refit_permutation_tests) {
  diff_ebf_constr_BH <- readRDS(path_diff_ebf_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  diff_ebf_constr_BH <- diffnet(
    net_ebf_perm,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_perm("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_constr_BH, path_diff_ebf_constr_BH)
}

summary(diff_ebf_constr_BH$pAdjustVec)
```

### Fisher's Z-Test

```{r run_diffnet_fisher_BH}
path_diff_ebf_fisher_BH <- path_perm("diff_ebf_fisher_BH.rds")

if (file.exists(path_diff_ebf_fisher_BH) && !refit_permutation_tests) {
  diff_ebf_fisher_BH <- readRDS(path_diff_ebf_fisher_BH)
} else {
  diff_ebf_fisher_BH <- diffnet(
    net_ebf_perm,
    diffMethod = "fisherTest",
    adjust = "BH",
    adjustNonzeroOnly = TRUE,
    verbose = TRUE
  )

  saveRDS(diff_ebf_fisher_BH, path_diff_ebf_fisher_BH)
}

summary(diff_ebf_fisher_BH$pvalsVec)
summary(diff_ebf_fisher_BH$pAdjustVec)
```

```{r diffnet_plot_helper}
plot_diffnet_with_adjust <- function(diff_obj, adjust = c("none", "BH"), ...) {
  adjust <- match.arg(adjust)
  plot_try <- tryCatch(
    plot(diff_obj, adjust = adjust, ...),
    error = function(e) e
  )

  if (!inherits(plot_try, "error")) {
    return(invisible(plot_try))
  }

  p_adjust_try <- tryCatch(
    plot(diff_obj, pAdjust = adjust, ...),
    error = function(e) e
  )

  if (!inherits(p_adjust_try, "error")) {
    return(invisible(p_adjust_try))
  }

  warning(
    "Could not pass an adjustment selector to plot.diffnet(); ",
    "falling back to the default plot method. First error: ",
    conditionMessage(plot_try)
  )
  plot(diff_obj, ...)
}
```

```{r plot_diffnet_fisher_raw, fig.width=10, fig.height=7}
plot_diffnet_with_adjust(
  diff_ebf_fisher_BH,
  adjust = "none",
  edgeTransp = 0
)
```

```{r plot_diffnet_fisher_BH, fig.width=10, fig.height=7}
plot_diffnet_with_adjust(
  diff_ebf_fisher_BH,
  adjust = "BH",
  edgeTransp = 0
)
```

## Raw P-Value Comparison

This section compares unadjusted p-values from empirical permutation testing,
unconstrained `permApprox` refinement, and constrained `permApprox` refinement.

```{r pvalue_comparison_helpers}
safe_pvec <- function(x) {
  if (is.null(x)) {
    return(numeric(0))
  }
  as.numeric(x)
}

make_ranked_palette <- function(n, palette = "Spectral") {
  if (is.function(palette)) {
    return(palette(n))
  }

  if (length(palette) > 1) {
    return(grDevices::colorRampPalette(palette)(n))
  }

  if (requireNamespace("RColorBrewer", quietly = TRUE) &&
      palette %in% rownames(RColorBrewer::brewer.pal.info)) {
    max_colors <- RColorBrewer::brewer.pal.info[palette, "maxcolors"]
    base_colors <- RColorBrewer::brewer.pal(max(3, max_colors), palette)
    return(grDevices::colorRampPalette(base_colors)(n))
  }

  grDevices::colorRampPalette(grDevices::hcl.colors(11, palette = palette))(n)
}

extract_association_differences <- function(net, absolute = FALSE) {
  asso1 <- as.matrix(net$assoEst1)
  asso2 <- as.matrix(net$assoEst2)

  if (!identical(dim(asso1), dim(asso2))) {
    stop("Association matrices must have identical dimensions.")
  }

  taxa <- colnames(asso1)
  if (is.null(taxa)) {
    taxa <- as.character(seq_len(ncol(asso1)))
  }

  upper_tri <- upper.tri(asso1)
  association_difference <- asso2[upper_tri] - asso1[upper_tri]

  if (absolute) {
    association_difference <- abs(association_difference)
  }

  taxa_pairs <- which(upper_tri, arr.ind = TRUE)

  tibble(
    test_id = seq_along(association_difference),
    taxon1 = taxa[taxa_pairs[, "row"]],
    taxon2 = taxa[taxa_pairs[, "col"]],
    association_difference = as.numeric(association_difference)
  )
}

extract_diffnet_method_pvalues <- function(method, diff_obj) {
  p_value <- safe_pvec(diff_obj$pvalsVec)
  p_adjust <- safe_pvec(diff_obj$pAdjustVec)

  if (length(p_adjust) == 0) {
    p_adjust <- rep(NA_real_, length(p_value))
  }

  if (length(p_adjust) != length(p_value)) {
    stop(
      "Raw and BH-adjusted p-value vectors have different lengths for method ",
      method,
      "."
    )
  }

  tibble(
    test_id = seq_along(p_value),
    method = method,
    p_value = p_value,
    p_adjust = p_adjust
  )
}

extract_diffnet_pvalues <- function(absolute = FALSE, palette = "Spectral") {
  pvalues_long <- bind_rows(
    extract_diffnet_method_pvalues(
      method = "empirical",
      diff_obj = diff_ebf_emp_BH
    ),
    extract_diffnet_method_pvalues(
      method = "unconstrained",
      diff_obj = diff_ebf_noConstr_BH
    ),
    extract_diffnet_method_pvalues(
      method = "constrained",
      diff_obj = diff_ebf_constr_BH
    ),
    extract_diffnet_method_pvalues(
      method = "fisher",
      diff_obj = diff_ebf_fisher_BH
    )
  )

  association_differences <- extract_association_differences(net_ebf_perm,
                                                             absolute = absolute)

  if (any(table(pvalues_long$method) != nrow(association_differences))) {
    stop(
      "The number of association differences does not match the number of ",
      "diffnet p-values."
    )
  }

  association_colors <- association_differences %>%
    arrange(abs(association_difference), test_id) %>%
    mutate(
      association_color = make_ranked_palette(n(), palette = palette)
    ) %>%
    select(test_id, association_color)

  pvalues_long %>%
    left_join(association_differences, by = "test_id") %>%
    left_join(association_colors, by = "test_id")
}

get_named_pvalues <- function(x, field_candidates) {
  for (field in field_candidates) {
    if (!is.null(x[[field]])) {
      out <- x[[field]]
      return(out)
    }
  }
  NULL
}

extract_centrality_pvalues <- function(comp_emp, comp_unconstr, comp_constr) {
  centrality_fields <- list(
    degree = c("pvalDiffDeg", "pvalDiffDegree"),
    betweenness = c("pvalDiffBetw", "pvalDiffBetween", "pvalDiffBetweenness"),
    closeness = c("pvalDiffClose", "pvalDiffCloseness"),
    eigenvector = c("pvalDiffEigen", "pvalDiffEigenv", "pvalDiffEigenvector")
  )

  empirical_list <- comp_emp$pvalDiffCentrEmpirical
  unconstr_list <- comp_unconstr$pvalDiffCentrRefined
  constr_list <- comp_constr$pvalDiffCentrRefined

  centrality_pvalues_wide <- purrr::imap_dfr(centrality_fields, function(fields, measure) {
    p_emp <- get_named_pvalues(empirical_list, fields)
    p_unconstr <- get_named_pvalues(unconstr_list, fields)
    p_constr <- get_named_pvalues(constr_list, fields)

    if (is.null(p_emp) || is.null(p_unconstr) || is.null(p_constr)) {
      return(tibble())
    }

    taxon <- names(p_emp)
    if (is.null(taxon)) {
      taxon <- seq_along(p_emp)
    }

    tibble(
      measure = measure,
      taxon = taxon,
      empirical = as.numeric(p_emp),
      unconstrained = as.numeric(p_unconstr),
      constrained = as.numeric(p_constr)
    )
  })

  if (nrow(centrality_pvalues_wide) == 0) {
    return(tibble(
      measure = character(),
      taxon = character(),
      method = character(),
      p_value = numeric()
    ))
  }

  centrality_pvalues_wide %>%
    pivot_longer(
      cols = c(empirical, unconstrained, constrained),
      names_to = "method",
      values_to = "p_value"
    )
}

plot_pvalue_pairs <- function(pval_long, id_cols, method_order = NULL) {
  if (is.null(method_order)) {
    method_order <- unique(pval_long$method)
  }

  pval_wide <- pval_long %>%
    mutate(method = factor(method, levels = method_order)) %>%
    pivot_wider(names_from = method, values_from = p_value)

  pair_df <- combn(method_order, 2, simplify = FALSE) %>%
    purrr::map_dfr(function(method_pair) {
      pval_wide %>%
        transmute(
          across(all_of(id_cols)),
          comparison = paste(method_pair, collapse = " vs "),
          x = .data[[method_pair[1]]],
          y = .data[[method_pair[2]]]
        )
    }) %>%
    filter(is.finite(x), is.finite(y), x > 0, y > 0)

  if (nrow(pair_df) == 0) {
    stop("No positive finite p-value pairs available for log-scaled plotting.")
  }

  pvalue_limits <- range(c(pair_df$x, pair_df$y), na.rm = TRUE)

  ggplot(pair_df, aes(x = x, y = y)) +
    geom_abline(slope = 1, intercept = 0, color = "gray60") +
    geom_point(alpha = 0.7, size = 1.5) +
    facet_wrap(~ comparison) +
    scale_x_log10(limits = pvalue_limits) +
    scale_y_log10(limits = pvalue_limits) +
    coord_equal() +
    labs(x = "p-value, method 1",
         y = "p-value, method 2") +
    theme_bw()
}

plot_diffnet_pvalues_by_association_difference <- function(
    pval_long,
    method_order = c("empirical", "unconstrained", "constrained", "fisher")) {
  method_labels <- c(
    empirical = "Empirical",
    unconstrained = "GPD unconstrained",
    constrained = "GPD constrained",
    fisher = "Fisher test"
  )

  zero_counts <- pval_long %>%
    group_by(method) %>%
    summarise(
      n_zero = sum(p_value == 0, na.rm = TRUE),
      n_total = sum(!is.na(p_value)),
      .groups = "drop"
    ) %>%
    mutate(
      method = factor(method, levels = method_order),
      method_label = paste0(
        method_labels[as.character(method)],
        "\n(zeros: ", n_zero, " / ", n_total, ")"
      )
    ) %>%
    arrange(method) %>%
    mutate(method = as.character(method))

  plot_df <- pval_long %>%
    left_join(zero_counts %>% select(method, method_label), by = "method") %>%
    mutate(
      method = factor(method, levels = method_order),
      method_label = factor(method_label, levels = zero_counts$method_label),
      significant_BH = !is.na(p_adjust) & p_adjust < 0.05,
      point_color = if_else(
        significant_BH,
        association_color,
        "gray65"
      )
    ) %>%
    filter(
      is.finite(association_difference),
      is.finite(p_value),
      p_value >= 0
    )

  if (nrow(plot_df) == 0) {
    stop("No finite non-negative p-values available for log-scaled plotting.")
  }

  positive_pvalues <- plot_df$p_value[plot_df$p_value > 0]
  if (length(positive_pvalues) == 0) {
    stop("At least one positive p-value is needed to define the log scale.")
  }
  
  min_positive_pvalue <- min(positive_pvalues, na.rm = TRUE)
  max_positive_pvalue <- max(positive_pvalues, na.rm = TRUE)
  zero_pvalue_position <- min_positive_pvalue / 10
  y_lower_limit <- zero_pvalue_position / 1.5
  y_breaks <- c(
    zero_pvalue_position,
    10^seq(
      floor(log10(min_positive_pvalue)),
      ceiling(log10(max_positive_pvalue))
    )
  )
  y_breaks <- y_breaks[
    y_breaks >= zero_pvalue_position &
      y_breaks <= max_positive_pvalue
  ]

  plot_df <- plot_df %>%
    mutate(
      p_value_plot = if_else(
        p_value == 0,
        zero_pvalue_position,
        p_value
      )
    )

  ggplot(plot_df, aes(x = association_difference, y = p_value_plot)) +
    geom_point(
      aes(color = point_color),
      alpha = 0.65,
      size = 1.4
    ) +
    facet_wrap(~ method_label, nrow = 1) +
    scale_color_identity() +
    scale_y_log10(
      limits = c(y_lower_limit, max_positive_pvalue),
      expand = c(0, 0),
      breaks = y_breaks,
      minor_breaks = NULL,
      labels = if_else(
        y_breaks == zero_pvalue_position,
        "0",
        scales::label_scientific(digits = 1)(y_breaks)
      )
    ) +
    labs(
      x = "Difference between associations",
      y = "p-value"
    ) +
    theme_bw()
}
```

### diffnet P-Values

```{r diffnet_raw_pvalue_comparison, fig.width=10, fig.height=5, dpi=250}
diffnet_pvalues_long <- extract_diffnet_pvalues(
  palette = "Spectral"
)

diffnet_pvalue_summary <- diffnet_pvalues_long %>%
  group_by(method) %>%
  summarise(
    n = sum(is.finite(p_value)),
    n_zero = sum(p_value == 0, na.rm = TRUE),
    min = min(p_value, na.rm = TRUE),
    min_positive = {
      positive_pvalues <- p_value[is.finite(p_value) & p_value > 0]
      if (length(positive_pvalues) == 0) {
        NA_real_
      } else {
        min(positive_pvalues)
      }
    },
    q25 = quantile(p_value, 0.25, na.rm = TRUE),
    median = median(p_value, na.rm = TRUE),
    q75 = quantile(p_value, 0.75, na.rm = TRUE),
    max = max(p_value, na.rm = TRUE),
    .groups = "drop"
  )

diffnet_pvalue_summary

plot_diffnet_pvalues_by_association_difference(
  diffnet_pvalues_long,
  method_order = c("empirical", "unconstrained", "constrained", "fisher")
)
```

### netCompare Centrality P-Values

```{r netcompare_centrality_raw_pvalue_comparison, fig.width=10, fig.height=6, dpi=250}
netcompare_centrality_pvalues_long <- extract_centrality_pvalues(
  comp_emp = comp_ebf_emp_noBH,
  comp_unconstr = comp_ebf_GPD_noConstr_noBH,
  comp_constr = comp_ebf_GPD_constr_noBH
)

netcompare_centrality_pvalue_summary <- netcompare_centrality_pvalues_long %>%
  group_by(measure, method) %>%
  summarise(
    n = sum(is.finite(p_value)),
    min = min(p_value, na.rm = TRUE),
    q25 = quantile(p_value, 0.25, na.rm = TRUE),
    median = median(p_value, na.rm = TRUE),
    q75 = quantile(p_value, 0.75, na.rm = TRUE),
    max = max(p_value, na.rm = TRUE),
    .groups = "drop"
  )

netcompare_centrality_pvalue_summary

plot_pvalue_pairs(
  netcompare_centrality_pvalues_long,
  id_cols = c("measure", "taxon"),
  method_order = c("empirical", "unconstrained", "constrained")
) +
  facet_grid(measure ~ comparison)
```

## Files Written

```{r list-files-target}
projthis::proj_dir_info(path_target(), recursive = TRUE)
```
