---
title: "Network comparison between EBF groups"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "11_network_comparison"
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "2e531120-aad0-4546-bf2a-05d6465091a5")
```

This document compares genus-level microbial association networks between
infants without exclusive breastfeeding and infants with at least two months of
exclusive breastfeeding. The first part contrasts several network construction
approaches descriptively. The second part conducts permutation-based network
comparison for the baseline network only.

```{r packages, message=FALSE, warning=FALSE}
# SpiecEasi from latentcor branch needed
if (as.character(utils::packageVersion("SpiecEasi")) != "2.0.1") {
  remotes::install_github("zdk123/SpiecEasi", ref = "latentcor")
}

library(conflicted)
library(phyloseq)
library(tidyverse)
library(NetCoMi)
library(permApprox)
library(igraph)
library(patchwork)
library(ggplot2)
library(pheatmap)
library(gridExtra)
library(grid)

# For permutation association matrices
library(foreach)
library(future)
library(doFuture)
library(doRNG)
library(progressr)
library(filematrix)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::count)

application_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
source(file.path(application_path, "functions.R"))
```

```{r directories}
# Keep cached networks, permutation matrices, and permutation-test results.
projthis::proj_create_dir_target(params$name, clean = FALSE)

path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
```

## Parameters

```{r parameters}
zero_replacement_fraction <- 0.65
network_seed <- 42
permutation_seed <- 42
n_perm_network_comparison <- 999
cores <- 10

n_abundant_network_labels <- 15
n_central_network_labels <- 10
slr_rank <- 2

refit_network_comparison <- FALSE
refit_permutation_matrices <- FALSE
refit_permutation_tests <- FALSE
```

## Load And Prepare Data

The comparison uses the two EBF groups that define the main contrast in this
chapter: no exclusive breastfeeding and exclusive breastfeeding for at least two
months. Samples with one month of exclusive breastfeeding are excluded.

```{r load_data}
phylo_genus <- readRDS(path_source("02_filtering", "bact_phylo_2m_genus_prev01.rds"))
sampdata <- readRDS(path_source("00_build_dataset", "sampdata.rds"))

phylo_genus <- renameTaxa(
  phylo_genus,
  pat = "<name>",
  substPat = "<name>_<subst_name>(<subst_R>)",
  numDupli = "Genus"
)
```

```{r prepare_ebf_groups, warning=FALSE}
ebf_keep <- sampdata$`EBF duration` %in% c("0", ">=2")

phylo_genus_ebf <- prune_samples(ebf_keep, phylo_genus)
phylo_genus_ebf <- prune_taxa(taxa_sums(phylo_genus_ebf) > 0, phylo_genus_ebf)

metadata_ebf <- sampdata[ebf_keep, , drop = FALSE] %>%
  mutate(
    ebf_group = factor(
      as.character(`EBF duration`),
      levels = c("0", ">=2"),
      labels = c("0 months", ">=2 months")
    )
  )

counts_ebf_raw <- get_sample_by_taxa(phylo_genus_ebf)
metadata_ebf <- metadata_ebf %>%
  filter(SampleID %in% rownames(counts_ebf_raw)) %>%
  arrange(match(SampleID, rownames(counts_ebf_raw)))
counts_ebf_raw <- counts_ebf_raw[metadata_ebf$SampleID, , drop = FALSE]

stopifnot(identical(rownames(counts_ebf_raw), metadata_ebf$SampleID))

counts_ebf_rel <- counts_to_rel_abund(counts_ebf_raw)
ebf_group <- metadata_ebf$ebf_group

zero_repl_multrepl <- compute_multrepl_clr(
  rel_abund_mat = counts_ebf_rel,
  frac = zero_replacement_fraction
)
counts_ebf_rel_multrepl <- zero_repl_multrepl$rel_abund_repl

path_counts_ebf_rel_bayesMult <- path_target("counts_ebf_rel_bayesMult.rds")

if (file.exists(path_counts_ebf_rel_bayesMult) && !refit_network_comparison) {
  zero_repl_bayesMult <- readRDS(path_counts_ebf_rel_bayesMult)
} else {
  zero_repl_bayesMult <- compute_zero_repl_clr(
    rel_abund_mat = counts_ebf_rel,
    method = "bayesMult",
    z.warning = 0.99,
    z.delete = FALSE
  )
  saveRDS(zero_repl_bayesMult, path_counts_ebf_rel_bayesMult)
}

counts_ebf_rel_bayesMult <- zero_repl_bayesMult$rel_abund_repl

metadata_ebf %>%
  count(ebf_group, name = "n_samples")
```

```{r plot_metadata}
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]

phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)

genus_mat <- get_sample_by_taxa(phylo_genus)
genus_rel_mat <- counts_to_rel_abund(genus_mat)

top_abundant_taxa <- colMeans(genus_rel_mat, na.rm = TRUE) %>%
  sort(decreasing = TRUE) %>%
  names()
names(top_abundant_taxa) <- top_abundant_taxa

network_label_overrides <- c(
  "Bifidobacterium" = "Bifidobacterium",
  "Bacteroides" = "Bacteroides",
  "Staphylococcus" = "Staphylococcus",
  "Haemophilus" = "Haemophilus"
)
```

## Helper Functions

```{r comparison_helpers}
transform_asso <- function(assoMat, thresh = NULL, dissTrans = "signed") {
  if (!is.null(thresh)) {
    assoMat[abs(assoMat) < thresh] <- 0
  }

  if (dissTrans == "signed") {
    dissMat <- sqrt(0.5 * (1 - assoMat))
  } else {
    dissMat <- sqrt(1 - assoMat^2)
  }

  dissMat[assoMat == 0] <- 1
  simMat <- 1 - dissMat
  graphObj <- SpiecEasi::adj2igraph(simMat)

  list(graph = graphObj, adja = simMat, asso = assoMat, diss = dissMat)
}

make_network_plot_labels <- function(
    props,
    top_abundant_taxa,
    n_abundant = 10,
    n_central = 10,
    extra_taxa = NULL,
    label_overrides = network_label_overrides) {

  adja <- props$input$adjaMat1
  diag(adja) <- 0
  network_taxa <- colnames(adja)

  non_singleton_taxa <- names(which(
    rowSums(adja != 0) > 0 |
      colSums(adja != 0) > 0
  ))

  abundant_taxa <- top_abundant_taxa[seq_len(
    min(max(n_abundant, 0), length(top_abundant_taxa))
  )]

  centrality <- props$centralities$eigenv1
  if (is.null(centrality) || n_central <= 0) {
    central_taxa <- character(0)
  } else {
    centrality <- centrality[is.finite(centrality)]
    central_taxa <- centrality %>%
      sort(decreasing = TRUE) %>%
      head(n_central) %>%
      names()
    names(central_taxa) <- central_taxa
  }

  taxa_to_label <- c(abundant_taxa, central_taxa, extra_taxa)
  taxa_to_label <- taxa_to_label[!duplicated(names(taxa_to_label))]
  taxa_to_label_in_network <- intersect(names(taxa_to_label), network_taxa)

  taxa_to_plot <- unique(c(non_singleton_taxa, taxa_to_label_in_network))
  network_labels <- setNames(rep("", length(network_taxa)), network_taxa)
  network_labels[taxa_to_label_in_network] <- taxa_to_label[taxa_to_label_in_network]

  if (!is.null(label_overrides)) {
    label_override_taxa <- intersect(names(label_overrides), names(network_labels))
    network_labels[label_override_taxa] <- label_overrides[label_override_taxa]
  }

  list(
    labels = network_labels,
    taxa_to_plot = taxa_to_plot,
    taxa_to_label = taxa_to_label,
    taxa_to_label_in_network = taxa_to_label_in_network,
    non_singleton_taxa = non_singleton_taxa
  )
}

extract_group_network_objects <- function(net) {
  list(
    asso1 = net$assoMat1,
    asso2 = net$assoMat2,
    adja1 = net$adjaMat1,
    adja2 = net$adjaMat2,
    graph1 = transform_asso(net$assoMat1)$graph,
    graph2 = transform_asso(net$assoMat2)$graph
  )
}

analyze_group_network <- function(net) {
  netAnalyze(
    net,
    centrLCC = TRUE,
    clustMethod = "cluster_fast_greedy",
    hubPar = "eigenvector",
    weightDeg = FALSE,
    normDeg = FALSE,
    normBetw = FALSE,
    normClose = FALSE
  )
}

plot_group_association_histograms <- function(net_objects, binwidth = 0.01,
                                              xlim = c(-0.2, 0.8),
                                              ylim = c(0, 50)) {
  make_plot <- function(asso, x_title = FALSE) {
    n_zero <- sum(asso[lower.tri(asso)] == 0)
    data.frame(correlation = asso[lower.tri(asso)]) %>%
      ggplot(aes(x = correlation)) +
      geom_histogram(binwidth = binwidth) +
      annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
               label = paste0("n = ", n_zero), size = 4) +
      coord_cartesian(ylim = ylim, xlim = xlim) +
      labs(
        x = if (x_title) "Estimated association" else "",
        y = "Count"
      ) +
      theme_bw() +
      theme(
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title.x = if (x_title) element_text(size = 14) else element_blank(),
        axis.text.x = if (x_title) element_text(size = 12) else element_blank(),
        axis.ticks.x = if (x_title) element_line() else element_blank()
      )
  }

  make_plot(net_objects$asso1, x_title = FALSE) /
    make_plot(net_objects$asso2, x_title = TRUE)
}

plot_group_degree_distribution <- function(net_objects) {
  ddist1 <- igraph::degree.distribution(net_objects$graph1)
  ddist2 <- igraph::degree.distribution(net_objects$graph2)

  max_len <- max(length(ddist1), length(ddist2))
  ddist1 <- c(ddist1, rep(0, max_len - length(ddist1)))
  ddist2 <- c(ddist2, rep(0, max_len - length(ddist2)))
  degree_seq <- 0:(max_len - 1)

  make_plot <- function(ddist, x_title = FALSE) {
    data.frame(
      Degree = as.factor(degree_seq),
      Fraction = ddist
    ) %>%
      ggplot(aes(x = Degree, y = Fraction, group = 1)) +
      geom_line(color = "gray30") +
      geom_point(size = 2, color = "gray30") +
      theme_bw() +
      theme(
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title.x = if (x_title) element_text(size = 14) else element_blank(),
        axis.text.x = if (x_title) element_text(size = 12) else element_blank(),
        axis.ticks.x = if (x_title) element_line() else element_blank()
      )
  }

  make_plot(ddist1, x_title = FALSE) /
    make_plot(ddist2, x_title = TRUE)
}

plot_group_network <- function(props, same_layout = FALSE, labels = NULL,
                               repulsion = 0.9, title1 = "EBF duration = 0",
                               title2 = "EBF duration >= 2",
                               labelScale = FALSE, cexLabels = 0.9,
                               phylum_legend_x = -0.15,
                               phylum_legend_y = -0.1,
                               asso_legend_x   = -0.15,
                               asso_legend_y   = 1.2,
                               doPlot = TRUE,
                               layout = NULL) {
  plot_args <- list(
    x = props,
    layout = layout,
    sameLayout = same_layout,
    rmSingles = TRUE,
    title1 = title1,
    title2 = title2,
    labelScale = labelScale,
    labels = labels,
    repulsion = repulsion,
    nodeColor = "feature",
    featVecCol = phyla,
    colorVec = phylcol,
    nodeSize = "clr",
    nodeSizeSpread = 4,
    cexNodes = 2,
    nodeTransp = 40,
    hubTransp = 40,
    posCol = "gray50",
    negCol = "red3",
    showTitle = TRUE,
    cexLabels = cexLabels,
    cexHubLabels = cexLabels,
    hubBorderCol = "gray40",
    mar = c(1, 3, 3, 3),
    doPlot = doPlot
  )
  if (same_layout) {
    plot_args$layoutGroup <- "union"
  }
  
  p <- do.call(plot, plot_args)
  
  if (doPlot) {
    phylcol_transp <- colToTransp(phylcol, 40)
    
    legend(phylum_legend_x, phylum_legend_y, cex = 1, pt.cex = 2.5, title = "Phylum:",
           legend = levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
           xpd = NA)
    
    legend(asso_legend_x, asso_legend_y, cex = 1, title = "Estimated association:",
           legend = c("+", "-"), lty = 1, lwd = 3, col = c("gray50", "red3"),
           bty = "n", horiz = TRUE, xpd = NA)
  }
  
  invisible(p)
}

build_plot_labels <- function(props) {
  make_network_plot_labels(
    props = props,
    top_abundant_taxa = top_abundant_taxa,
    n_abundant = n_abundant_network_labels,
    n_central = n_central_network_labels
  )$labels
}
```

## Descriptive Network Comparisons

The following sections construct and describe EBF-group-specific networks
without permutation testing.

### SPIEC-EASI SLR

```{r generate_net_ebf_slr}
path_net_ebf_slr <- path_target(paste0("net_ebf_slr_r", slr_rank, ".rds"))

if (file.exists(path_net_ebf_slr) && !refit_network_comparison) {
  net_ebf_slr <- readRDS(path_net_ebf_slr)
} else {
  net_ebf_slr <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multrepl,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "slr",
                      nlambda = 30,
                      r = slr_rank,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_slr, path_net_ebf_slr)
}

objects_ebf_slr <- extract_group_network_objects(net_ebf_slr)
props_ebf_slr <- analyze_group_network(net_ebf_slr)
labels_ebf_slr <- build_plot_labels(props_ebf_slr)
```

#### Low-Rank Heatmaps

```{r ebf_slr_low_rank_prepare}
extract_slr_low_rank <- function(spieceasi_fit, taxa_names) {
  opt_ind <- pulsar::opt.index(
    spieceasi_fit$select,
    criterion = "stars"
  )

  L_opt <- as.matrix(spieceasi_fit$est$resid[[opt_ind]])

  if (length(taxa_names) != nrow(L_opt) || length(taxa_names) != ncol(L_opt)) {
    stop(
      "Taxon names from the SLR association matrix do not match the low-rank ",
      "matrix dimensions."
    )
  }

  dimnames(L_opt) <- list(taxa_names, taxa_names)

  list(
    L = L_opt,
    lambda = spieceasi_fit$est$lambda[opt_ind]
  )
}

slr_low_rank_ebf <- list(
  "EBF = 0" = extract_slr_low_rank(
    net_ebf_slr$measureOut1,
    taxa_names = colnames(net_ebf_slr$assoMat1)
  ),
  "EBF >= 2" = extract_slr_low_rank(
    net_ebf_slr$measureOut2,
    taxa_names = colnames(net_ebf_slr$assoMat2)
  )
)

common_taxa_slr <- Reduce(
  intersect,
  lapply(slr_low_rank_ebf, function(x) colnames(x$L))
)

if (length(common_taxa_slr) == 0) {
  stop("No common taxa found between the two SLR low-rank matrices.")
}

slr_low_rank_summary <- tibble(
  group = names(slr_low_rank_ebf),
  n_taxa = vapply(slr_low_rank_ebf, function(x) ncol(x$L), integer(1)),
  n_common_taxa = length(common_taxa_slr),
  lambda = vapply(slr_low_rank_ebf, function(x) x$lambda, numeric(1))
)

slr_low_rank_summary

slr_low_rank_ebf <- lapply(slr_low_rank_ebf, function(x) {
  x$L <- x$L[common_taxa_slr, common_taxa_slr, drop = FALSE]
  x
})

L_order <- Reduce(
  "+",
  lapply(slr_low_rank_ebf, function(x) {
    L <- x$L
    diag(L) <- 0
    L
  })
) / length(slr_low_rank_ebf)

slr_low_rank_cor <- stats::cor(L_order, use = "pairwise.complete.obs")
slr_low_rank_diss <- 1 - slr_low_rank_cor
slr_low_rank_diss[!is.finite(slr_low_rank_diss)] <- 1
diag(slr_low_rank_diss) <- 0

taxon_order_slr <- hclust(
  as.dist(slr_low_rank_diss)
)$order

all_slr_low_rank_values <- unlist(lapply(slr_low_rank_ebf, function(x) {
  L <- x$L
  L[row(L) == col(L)] <- NA_real_
  L[upper.tri(L)]
}))

lim_slr_low_rank <- max(abs(all_slr_low_rank_values), na.rm = TRUE)
if (!is.finite(lim_slr_low_rank) || lim_slr_low_rank == 0) {
  lim_slr_low_rank <- 1
}

slr_low_rank_heatmap_grobs <- lapply(names(slr_low_rank_ebf), function(group_name) {
  x <- slr_low_rank_ebf[[group_name]]
  L <- x$L[taxon_order_slr, taxon_order_slr, drop = FALSE]
  diag(L) <- NA_real_

  pheatmap(
    mat = L,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    border_color = NA,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(101),
    breaks = seq(-lim_slr_low_rank, lim_slr_low_rank, length.out = 102),
    main = paste0(
      group_name,
      ": rank ",
      slr_rank,
      " (StARS-selected lambda = ",
      formatC(x$lambda, format = "f", digits = 3),
      ")"
    ),
    silent = TRUE
  )$gtable
})
```

```{r ebf_slr_low_rank_heatmaps, fig.width=10, fig.height=5, dpi=250}
gridExtra::grid.arrange(
  grobs = slr_low_rank_heatmap_grobs,
  ncol = 2
)
```

#### Association Histogram

```{r ebf_slr_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_slr)
```

#### Degree Distribution

```{r ebf_slr_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_slr)
```

#### Graphlet Correlation Matrix

```{r ebf_slr_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_slr,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_slr_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_slr, 
                   same_layout = FALSE, 
                   labels = labels_ebf_slr,
                   repulsion = 0.7,
                   labelScale = TRUE,
                   cexLabels = 4)
```

```{r ebf_slr_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_slr, 
                   same_layout = TRUE, 
                   labels = labels_ebf_slr,
                   repulsion = 0.7,
                   labelScale = TRUE,
                   cexLabels = 4)
```

### SPRING

```{r generate_net_ebf_spring}
path_net_ebf_spring <- path_target("net_ebf_spring.rds")

if (file.exists(path_net_ebf_spring) && !refit_network_comparison) {
  net_ebf_spring <- readRDS(path_net_ebf_spring)
} else {
  net_ebf_spring <- NetCoMi::netConstruct(
    data = counts_ebf_raw,
    group = ebf_group,
    measure = "spring",
    measurePar = list(nlambda = 30,
                      rep.num = 30,
                      thresh = 0.1,
                      Rmethod = "approx"),
    jointPrepro = FALSE,
    filtTax = c("totalReads", "numbSamp"),
    filtTaxPar = list(totalReads = 100, numbSamp = 5),
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_spring, path_net_ebf_spring)
}

objects_ebf_spring <- extract_group_network_objects(net_ebf_spring)
props_ebf_spring <- analyze_group_network(net_ebf_spring)
labels_ebf_spring <- build_plot_labels(props_ebf_spring)
```

```{r}
dim(net_ebf_spring$countMat1)
dim(net_ebf_spring$countMat2)
```

The filtering reduced the data set to 65 taxa in each group.

#### Association Histogram

```{r ebf_spring_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_spring)
```

#### Degree Distribution

```{r ebf_spring_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_spring)
```

#### Graphlet Correlation Matrix

```{r ebf_spring_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_spring,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_spring_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
p <- plot_group_network(props_ebf_spring, same_layout = FALSE, 
                   labels = labels_ebf_spring, repulsion = 0.8599, doPlot = FALSE)
layout1 <- p$layout$layout1
layout2 <- p$layout$layout2

layout1["Anaerostipes", ] <- c(0.31, 0.33)
layout1["Enterococcus", ] <- c(1, -0.23)

layout2["Bacteroides", ] <- c(0.14, 0.94)

plot_group_network(props_ebf_spring, same_layout = FALSE, 
                   labels = labels_ebf_spring, layout = list(layout1, layout2))
```

```{r ebf_spring_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_spring, same_layout = TRUE, labels = labels_ebf_spring)
```

### SPIEC-EASI MB With BayesMult Zero Replacement

```{r generate_net_ebf_spieceasi_bayesMult}
path_net_ebf_spieceasi_bayesMult <- path_target("net_ebf_spieceasi_bayesMult.rds")

if (file.exists(path_net_ebf_spieceasi_bayesMult) && !refit_network_comparison) {
  net_ebf_spieceasi_bayesMult <- readRDS(path_net_ebf_spieceasi_bayesMult)
} else {
  net_ebf_spieceasi_bayesMult <- NetCoMi::netConstruct(
    data = counts_ebf_rel_bayesMult,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_spieceasi_bayesMult, path_net_ebf_spieceasi_bayesMult)
}

objects_ebf_spieceasi_bayesMult <- extract_group_network_objects(
  net_ebf_spieceasi_bayesMult
)
props_ebf_spieceasi_bayesMult <- analyze_group_network(
  net_ebf_spieceasi_bayesMult
)
labels_ebf_spieceasi_bayesMult <- build_plot_labels(
  props_ebf_spieceasi_bayesMult
)
```

#### Association Histogram

```{r ebf_spieceasi_bayesMult_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_spieceasi_bayesMult)
```

#### Degree Distribution

```{r ebf_spieceasi_bayesMult_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_spieceasi_bayesMult)
```

#### Graphlet Correlation Matrix

```{r ebf_spieceasi_bayesMult_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_spieceasi_bayesMult,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_spieceasi_bayesMult_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_bayesMult,
  same_layout = FALSE,
  labels = labels_ebf_spieceasi_bayesMult
)
```

```{r ebf_spieceasi_bayesMult_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_spieceasi_bayesMult,
  same_layout = TRUE,
  labels = labels_ebf_spieceasi_bayesMult
)
```

### rhoshrink

```{r generate_net_ebf_rhoshrink}
path_net_ebf_rhoshrink <- path_target("net_ebf_rhoshrink.rds")

if (file.exists(path_net_ebf_rhoshrink) && !refit_network_comparison) {
  net_ebf_rhoshrink <- readRDS(path_net_ebf_rhoshrink)
} else {
  net_ebf_rhoshrink <- NetCoMi::netConstruct(
    data = counts_ebf_rel_bayesMult,
    group = ebf_group,
    measure = "rhoshrink",
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "threshold",
    thresh = 0.25,
    seed = network_seed
  )

  saveRDS(net_ebf_rhoshrink, path_net_ebf_rhoshrink)
}

objects_ebf_rhoshrink <- extract_group_network_objects(net_ebf_rhoshrink)
props_ebf_rhoshrink <- analyze_group_network(net_ebf_rhoshrink)
labels_ebf_rhoshrink <- build_plot_labels(props_ebf_rhoshrink)
```

#### Association Histogram

```{r ebf_rhoshrink_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_rhoshrink)
```

#### Degree Distribution

```{r ebf_rhoshrink_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_rhoshrink)
```

#### Graphlet Correlation Matrix

```{r ebf_rhoshrink_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_rhoshrink,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

#### Network Plot

```{r ebf_rhoshrink_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_rhoshrink,
  same_layout = FALSE,
  labels = labels_ebf_rhoshrink
)
```

```{r ebf_rhoshrink_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(
  props_ebf_rhoshrink,
  same_layout = TRUE,
  labels = labels_ebf_rhoshrink
)
```

## Baseline Network For Permutation Testing

Permutation testing is restricted to the baseline network from
`10_network_analysis`: SPIEC-EASI with the MB method, using multiplicative zero
replacement before `netConstruct()`. The descriptive methods above are not used
for permutation testing because preliminary runs failed under the extreme
sparsity of the data.

```{r generate_net_ebf_main}
path_net_ebf_main <- path_target("net_ebf_main.rds")

if (file.exists(path_net_ebf_main) && !refit_network_comparison) {
  net_ebf_main <- readRDS(path_net_ebf_main)
} else {
  net_ebf_main <- NetCoMi::netConstruct(
    data = counts_ebf_rel_multrepl,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30,
                                           thresh = 0.1),
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "none",
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = network_seed
  )

  saveRDS(net_ebf_main, path_net_ebf_main)
}

objects_ebf_main <- extract_group_network_objects(net_ebf_main)
props_ebf_main <- analyze_group_network(net_ebf_main)
labels_ebf_main <- build_plot_labels(props_ebf_main)
```

### Baseline Network Diagnostics

```{r ebf_main_association_histogram, fig.width=8, fig.height=4}
plot_group_association_histograms(objects_ebf_main)
```

```{r ebf_main_degree_distribution, fig.width=8, fig.height=4, dpi=200}
plot_group_degree_distribution(objects_ebf_main)
```

```{r ebf_main_gcm, fig.width=10, fig.height=9, dpi=200}
plotGCM(props_ebf_main,
        legendCex = 1.2,
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

```{r ebf_main_network_plot_different_layouts, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_main, same_layout = FALSE, labels = labels_ebf_main)
```

```{r ebf_main_network_plot_same_layout, fig.width=12, fig.height=6, dpi=250}
plot_group_network(props_ebf_main, same_layout = TRUE, labels = labels_ebf_main)
```

## Generate Or Reuse Permuted Association Matrices

The following chunks are kept for reproducibility and are not evaluated during
normal rendering. The final filematrix at `path_target("assoPerm_ebf")` is used
by both `netCompare()` and `diffnet()`.

```{r generate_permuted_association_matrices, eval=FALSE}
permGroupMat <- createAssoPerm(
  props_ebf_main,
  nPerm = 1000,
  computeAsso = FALSE,
  seed = permutation_seed
)

nPerm_all <- 1000
blocksize <- 20
repetitions <- nPerm_all / blocksize
perm_asso_dir <- file.path("data", "10_networks", "perm_asso_matrices")

cores_perm <- 8

plan(multisession, workers = cores_perm)
registerDoFuture()

with_progress({
  p <- progressr::progressor(along = 1:repetitions)

  tmp <- foreach(i = 1:repetitions,
                 .packages = c("NetCoMi")) %dorng% {
    p()
    file_base <- file.path(perm_asso_dir, paste0("assoPerm_ebf", i))
    bmat_file <- paste0(file_base, ".bmat")
    if (!file.exists(bmat_file) || refit_permutation_matrices) {
      set.seed(permutation_seed + i)
      NetCoMi::createAssoPerm(
        props_ebf_main,
        nPerm = blocksize,
        permGroupMat = permGroupMat[(i - 1) * blocksize + 1:blocksize, ],
        computeAsso = TRUE,
        fileStoreAssoPerm = file_base,
        storeCountsPerm = FALSE,
        append = FALSE
      )
    }
  }
})

plan(sequential)
```

```{r combine_permuted_association_matrices, eval=FALSE}
assoPerm_all <- NULL
perm_asso_dir <- file.path("data", "10_networks", "perm_asso_matrices")

for (i in 1:repetitions) {
  file_base <- file.path(perm_asso_dir, paste0("assoPerm_ebf", i))
  assoPerm_tmp <- fm.open(filenamebase = file_base, readonly = TRUE)
  assoPerm_all <- rbind(assoPerm_all, as.matrix(assoPerm_tmp))
  close(assoPerm_tmp)
}

dim(assoPerm_all)

n_perm_netcompare <- n_perm_network_comparison
n_taxa_netcompare <- nrow(props_ebf_main$input$adjaMat1)
n_rows_netcompare <- n_perm_netcompare * n_taxa_netcompare

stopifnot(nrow(assoPerm_all) >= n_rows_netcompare)
stopifnot(nrow(assoPerm_all) %% n_taxa_netcompare == 0)

assoPerm_netcompare <- assoPerm_all[seq_len(n_rows_netcompare), , drop = FALSE]
dim(assoPerm_netcompare)

fm.create.from.matrix(
  filenamebase = path_target("assoPerm_ebf"),
  mat = assoPerm_netcompare
)
```

```{r check_permuted_association_matrices}
asso_perm_base <- path_target("assoPerm_ebf")

if (!file.exists(paste0(asso_perm_base, ".bmat"))) {
  warning(
    "Permutation association filematrix not found at path_target('assoPerm_ebf'). ",
    "Run the unevaluated permutation-generation chunks before refitting ",
    "permutation tests."
  )
} else {
  assoPerm_check <- tryCatch(
    fm.open(filenamebase = asso_perm_base, readonly = TRUE),
    error = function(e) e
  )

  if (inherits(assoPerm_check, "error")) {
    warning("Could not open permutation association filematrix: ",
            conditionMessage(assoPerm_check))
  } else {
    expected_rows <- n_perm_network_comparison *
      nrow(props_ebf_main$input$adjaMat1)
    expected_cols <- 2 * ncol(props_ebf_main$input$adjaMat1)

    stopifnot(nrow(assoPerm_check) == expected_rows)
    stopifnot(ncol(assoPerm_check) == expected_cols)
    close(assoPerm_check)
  }
}
```

## Permutation-Based Differential Network Analysis

We compare local and global network characteristics with NetCoMi's
`netCompare()` using empirical p-values, unconstrained `permApprox` refinement,
and constrained `permApprox` refinement.

### Empirical P-Values

```{r run_netcompare_noBH}
path_comp_ebf_main_noBH <- path_target("comp_ebf_main_noBH.rds")

if (file.exists(path_comp_ebf_main_noBH) && !refit_permutation_tests) {
  comp_ebf_main_noBH <- readRDS(path_comp_ebf_main_noBH)
} else {
  comp_ebf_main_noBH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_main_noBH, path_comp_ebf_main_noBH)
}

print(summary(comp_ebf_main_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_BH}
path_comp_ebf_main_BH <- path_target("comp_ebf_main_BH.rds")

if (file.exists(path_comp_ebf_main_BH) && !refit_permutation_tests) {
  comp_ebf_main_BH <- readRDS(path_comp_ebf_main_BH)
} else {
  comp_ebf_main_BH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_main_BH, path_comp_ebf_main_BH)
}

print(summary(comp_ebf_main_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

### permApprox Refinement Without Constraint

```{r run_netcompare_GPD_noConstr_noBH}
path_comp_ebf_GPD_noConstr_noBH <- path_target("comp_ebf_GPD_noConstr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_noBH) && !refit_permutation_tests) {
  comp_ebf_GPD_noConstr_noBH <- readRDS(path_comp_ebf_GPD_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  comp_ebf_GPD_noConstr_noBH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_noConstr_noBH, path_comp_ebf_GPD_noConstr_noBH)
}

print(summary(comp_ebf_GPD_noConstr_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_GPD_noConstr_BH}
path_comp_ebf_GPD_noConstr_BH <- path_target("comp_ebf_GPD_noConstr_BH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_BH) && !refit_permutation_tests) {
  comp_ebf_GPD_noConstr_BH <- readRDS(path_comp_ebf_GPD_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  comp_ebf_GPD_noConstr_BH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_noConstr_BH, path_comp_ebf_GPD_noConstr_BH)
}

print(summary(comp_ebf_GPD_noConstr_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

### permApprox Refinement With Constraint

```{r run_netcompare_GPD_constr_noBH}
path_comp_ebf_GPD_constr_noBH <- path_target("comp_ebf_GPD_constr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_constr_noBH) && !refit_permutation_tests) {
  comp_ebf_GPD_constr_noBH <- readRDS(path_comp_ebf_GPD_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  comp_ebf_GPD_constr_noBH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_constr_noBH, path_comp_ebf_GPD_constr_noBH)
}

print(summary(comp_ebf_GPD_constr_noBH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

```{r run_netcompare_GPD_constr_BH}
path_comp_ebf_GPD_constr_BH <- path_target("comp_ebf_GPD_constr_BH.rds")

if (file.exists(path_comp_ebf_GPD_constr_BH) && !refit_permutation_tests) {
  comp_ebf_GPD_constr_BH <- readRDS(path_comp_ebf_GPD_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  comp_ebf_GPD_constr_BH <- netCompare(
    props_ebf_main,
    permTest = TRUE,
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(comp_ebf_GPD_constr_BH, path_comp_ebf_GPD_constr_BH)
}

print(summary(comp_ebf_GPD_constr_BH,
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"),
              numbNodes = 5))
```

## Permutation-Based Differential Association Analysis

Differential association analysis is performed with `diffnet()` on the same
baseline network and the same stored permutation association matrices.

### Empirical P-Values

```{r run_diffnet_noBH}
path_diff_ebf_main_noBH <- path_target("diff_ebf_main_noBH.rds")

if (file.exists(path_diff_ebf_main_noBH) && !refit_permutation_tests) {
  diff_ebf_main_noBH <- readRDS(path_diff_ebf_main_noBH)
} else {
  diff_ebf_main_noBH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_main_noBH, path_diff_ebf_main_noBH)
}

summary(diff_ebf_main_noBH$pvalsVec)
```

```{r run_diffnet_BH}
path_diff_ebf_main_BH <- path_target("diff_ebf_main_BH.rds")

if (file.exists(path_diff_ebf_main_BH) && !refit_permutation_tests) {
  diff_ebf_main_BH <- readRDS(path_diff_ebf_main_BH)
} else {
  diff_ebf_main_BH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_main_BH, path_diff_ebf_main_BH)
}

summary(diff_ebf_main_BH$pAdjustVec)
```

### permApprox Refinement Without Constraint

```{r run_diffnet_noConstr_noBH}
path_diff_ebf_noConstr_noBH <- path_target("diff_ebf_noConstr_noBH.rds")

if (file.exists(path_diff_ebf_noConstr_noBH) && !refit_permutation_tests) {
  diff_ebf_noConstr_noBH <- readRDS(path_diff_ebf_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  diff_ebf_noConstr_noBH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_noConstr_noBH, path_diff_ebf_noConstr_noBH)
}

summary(diff_ebf_noConstr_noBH$pvalsVec)
```

```{r run_diffnet_noConstr_BH}
path_diff_ebf_noConstr_BH <- path_target("diff_ebf_noConstr_BH.rds")

if (file.exists(path_diff_ebf_noConstr_BH) && !refit_permutation_tests) {
  diff_ebf_noConstr_BH <- readRDS(path_diff_ebf_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )

  diff_ebf_noConstr_BH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_noConstr_BH, path_diff_ebf_noConstr_BH)
}

summary(diff_ebf_noConstr_BH$pAdjustVec)
```

### permApprox Refinement With Constraint

```{r run_diffnet_constr_noBH}
path_diff_ebf_constr_noBH <- path_target("diff_ebf_constr_noBH.rds")

if (file.exists(path_diff_ebf_constr_noBH) && !refit_permutation_tests) {
  diff_ebf_constr_noBH <- readRDS(path_diff_ebf_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  diff_ebf_constr_noBH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "none",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_constr_noBH, path_diff_ebf_constr_noBH)
}

summary(diff_ebf_constr_noBH$pvalsVec)
```

```{r run_diffnet_constr_BH}
path_diff_ebf_constr_BH <- path_target("diff_ebf_constr_BH.rds")

if (file.exists(path_diff_ebf_constr_BH) && !refit_permutation_tests) {
  diff_ebf_constr_BH <- readRDS(path_diff_ebf_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group))),
    thresh_method = "ftr"
  )

  diff_ebf_constr_BH <- diffnet(
    net_ebf_main,
    diffMethod = "permute",
    nPerm = n_perm_network_comparison,
    fileLoadAssoPerm = path_target("assoPerm_ebf"),
    refinePermPvals = TRUE,
    permApproxParams = list(
      method = "gpd",
      gpd_ctrl = gpd_ctrl
    ),
    adjust = "BH",
    cores = cores,
    verbose = TRUE,
    seed = permutation_seed
  )

  saveRDS(diff_ebf_constr_BH, path_diff_ebf_constr_BH)
}

summary(diff_ebf_constr_BH$pAdjustVec)
```

## Raw P-Value Comparison

This section compares unadjusted p-values from empirical permutation testing,
unconstrained `permApprox` refinement, and constrained `permApprox` refinement.

```{r pvalue_comparison_helpers}
safe_pvec <- function(x) {
  if (is.null(x)) {
    return(numeric(0))
  }
  as.numeric(x)
}

extract_diffnet_pvalues <- function() {
  tibble(
    test_id = seq_along(diff_ebf_main_noBH$pvalsVec),
    empirical = safe_pvec(diff_ebf_main_noBH$pvalsVec),
    unconstrained = safe_pvec(diff_ebf_noConstr_noBH$pvalsVec),
    constrained = safe_pvec(diff_ebf_constr_noBH$pvalsVec)
  ) %>%
    pivot_longer(
      cols = c(empirical, unconstrained, constrained),
      names_to = "method",
      values_to = "p_value"
    )
}

get_named_pvalues <- function(x, field_candidates) {
  for (field in field_candidates) {
    if (!is.null(x[[field]])) {
      out <- x[[field]]
      return(out)
    }
  }
  NULL
}

extract_centrality_pvalues <- function(comp_emp, comp_unconstr, comp_constr) {
  centrality_fields <- list(
    degree = c("pvalDiffDeg", "pvalDiffDegree"),
    betweenness = c("pvalDiffBetw", "pvalDiffBetween", "pvalDiffBetweenness"),
    closeness = c("pvalDiffClose", "pvalDiffCloseness"),
    eigenvector = c("pvalDiffEigen", "pvalDiffEigenv", "pvalDiffEigenvector")
  )

  empirical_list <- comp_emp$pvalDiffCentrEmpirical
  unconstr_list <- comp_unconstr$pvalDiffCentrRefined
  constr_list <- comp_constr$pvalDiffCentrRefined

  centrality_pvalues_wide <- purrr::imap_dfr(centrality_fields, function(fields, measure) {
    p_emp <- get_named_pvalues(empirical_list, fields)
    p_unconstr <- get_named_pvalues(unconstr_list, fields)
    p_constr <- get_named_pvalues(constr_list, fields)

    if (is.null(p_emp) || is.null(p_unconstr) || is.null(p_constr)) {
      return(tibble())
    }

    taxon <- names(p_emp)
    if (is.null(taxon)) {
      taxon <- seq_along(p_emp)
    }

    tibble(
      measure = measure,
      taxon = taxon,
      empirical = as.numeric(p_emp),
      unconstrained = as.numeric(p_unconstr),
      constrained = as.numeric(p_constr)
    )
  })

  if (nrow(centrality_pvalues_wide) == 0) {
    return(tibble(
      measure = character(),
      taxon = character(),
      method = character(),
      p_value = numeric()
    ))
  }

  centrality_pvalues_wide %>%
    pivot_longer(
      cols = c(empirical, unconstrained, constrained),
      names_to = "method",
      values_to = "p_value"
    )
}

plot_pvalue_pairs <- function(pval_long, id_cols) {
  pval_wide <- pval_long %>%
    pivot_wider(names_from = method, values_from = p_value)

  pair_df <- bind_rows(
    pval_wide %>%
      transmute(across(all_of(id_cols)),
                comparison = "Empirical vs unconstrained",
                x = empirical,
                y = unconstrained),
    pval_wide %>%
      transmute(across(all_of(id_cols)),
                comparison = "Empirical vs constrained",
                x = empirical,
                y = constrained),
    pval_wide %>%
      transmute(across(all_of(id_cols)),
                comparison = "Unconstrained vs constrained",
                x = unconstrained,
                y = constrained)
  )

  ggplot(pair_df, aes(x = -log10(x), y = -log10(y))) +
    geom_abline(slope = 1, intercept = 0, color = "gray60") +
    geom_point(alpha = 0.7, size = 1.5) +
    facet_wrap(~ comparison, scales = "free") +
    labs(x = "-log10 p-value, method 1",
         y = "-log10 p-value, method 2") +
    theme_bw()
}
```

### diffnet P-Values

```{r diffnet_raw_pvalue_comparison, fig.width=10, fig.height=3.5, dpi=250}
diffnet_pvalues_long <- extract_diffnet_pvalues()

diffnet_pvalue_summary <- diffnet_pvalues_long %>%
  group_by(method) %>%
  summarise(
    n = sum(is.finite(p_value)),
    min = min(p_value, na.rm = TRUE),
    q25 = quantile(p_value, 0.25, na.rm = TRUE),
    median = median(p_value, na.rm = TRUE),
    q75 = quantile(p_value, 0.75, na.rm = TRUE),
    max = max(p_value, na.rm = TRUE),
    .groups = "drop"
  )

diffnet_pvalue_summary

plot_pvalue_pairs(diffnet_pvalues_long, id_cols = "test_id")
```

### netCompare Centrality P-Values

```{r netcompare_centrality_raw_pvalue_comparison, fig.width=10, fig.height=6, dpi=250}
netcompare_centrality_pvalues_long <- extract_centrality_pvalues(
  comp_emp = comp_ebf_main_noBH,
  comp_unconstr = comp_ebf_GPD_noConstr_noBH,
  comp_constr = comp_ebf_GPD_constr_noBH
)

netcompare_centrality_pvalue_summary <- netcompare_centrality_pvalues_long %>%
  group_by(measure, method) %>%
  summarise(
    n = sum(is.finite(p_value)),
    min = min(p_value, na.rm = TRUE),
    q25 = quantile(p_value, 0.25, na.rm = TRUE),
    median = median(p_value, na.rm = TRUE),
    q75 = quantile(p_value, 0.75, na.rm = TRUE),
    max = max(p_value, na.rm = TRUE),
    .groups = "drop"
  )

netcompare_centrality_pvalue_summary

plot_pvalue_pairs(
  netcompare_centrality_pvalues_long,
  id_cols = c("measure", "taxon")
) +
  facet_grid(measure ~ comparison, scales = "free")
```

## Files Written

```{r list-files-target}
projthis::proj_dir_info(path_target(), recursive = TRUE)
```
