---
title: "Network comparison between EBF groups"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: 
  github_document: default
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
params:
  name: "11_network_comparison" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "2e531120-aad0-4546-bf2a-05d6465091a5")
```

The purpose of this document is ...

```{r packages}
library("conflicted")
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Network comparison with SPRING associations

### Analysis data set

The group comparison uses the two EBF groups that define the main contrast in
this chapter: no exclusive breastfeeding and exclusive breastfeeding for at
least two months. Samples with one month of exclusive breastfeeding are excluded
from this comparison.

We start by using SPRING associations for the comparison because it avoids the 
additional uncertainty introduced by zero replacement.

```{r prepare_ebf_groups, warning=FALSE}
# Generate data set for EBF comparison
ebf_keep <- sampdata$`EBF duration` %in% c("0", ">=2")

phylo_genus_ebf <- prune_samples(ebf_keep, phylo_genus)
phylo_genus_ebf <- prune_taxa(taxa_sums(phylo_genus_ebf) > 0, phylo_genus_ebf)

metadata_ebf <- sampdata[ebf_keep, , drop = FALSE] %>%
  mutate(
    ebf_group = factor(
      as.character(`EBF duration`),
      levels = c("0", ">=2"),
      labels = c("0 months", ">=2 months")
    )
  )

counts_ebf_raw <- get_sample_by_taxa(phylo_genus_ebf)
counts_ebf_rel <- counts_to_rel_abund(counts_ebf_raw)

ebf_group <- metadata_ebf$ebf_group

zero_repl_object <- compute_multrepl_clr(
  rel_abund_mat = counts_ebf_rel,
  frac = zero_replacement_fraction
)

counts_ebf_rel_multrepl <- zero_repl_object$rel_abund_repl
```

```{r ebf_group_overview, eval=FALSE}
metadata_ebf %>%
  count(ebf_group, name = "n_samples")
```

### Network construction

```{r generate_net_ebf_spring}
path_net_ebf_spring <- path_target("net_ebf_spring.rds")

if (file.exists(path_net_ebf_spring) && !refit_complete_network) {
  net_ebf_spring <- readRDS(path_net_ebf_spring)
} else {
  net_ebf_spring <- NetCoMi::netConstruct(
    group = ebf_group,
    data = counts_ebf_raw,
    measure = "spring",
    measurePar = list(nlambda = 30, 
                      rep.num = 30,
                      thresh  = 0.1,
                      Rmethod = "approx"),
    jointPrepro = FALSE,
    filtTax = c("totalReads", "numbSamp"),
    filtTaxPar = list(totalReads = 100, numbSamp = 5),
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "none",
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_ebf_spring, path_net_ebf_spring)
}
```

```{r}
# Extract association matrices
assomat_ebf_spring1 <- net_ebf_spring$assoMat1
assomat_ebf_spring2 <- net_ebf_spring$assoMat2
adjamat_ebf_spring1 <- net_ebf_spring$adjaMat1
adjamat_ebf_spring2 <- net_ebf_spring$adjaMat2

# Generate graph objectss
graph_ebf_spring1 <- transform_asso(assomat_ebf_spring1)$graph
graph_ebf_spring2 <- transform_asso(assomat_ebf_spring2)$graph
```

### Network analysis

#### Association histogram

```{r ebf_spring_association_histrogram, fig.width=8, fig.height=4}
n_zero_asso1 <- sum(assomat_ebf_spring1[lower.tri(assomat_ebf_spring1)] == 0)
n_zero_asso2 <- sum(assomat_ebf_spring2[lower.tri(assomat_ebf_spring2)] == 0)

p1 <- data.frame(correlation = assomat_ebf_spring1[lower.tri(assomat_ebf_spring1)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso1), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "",
       y = "Count",
       ) +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- data.frame(correlation = assomat_ebf_spring2[lower.tri(assomat_ebf_spring2)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso2), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "Estimated correlation",
       y = "Count",
  ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )

p1 / p2
```

#### Degree distribution

```{r ebf_spring_degree_distribution, fig.width=8, fig.height=4, dpi = 200}
# Compute degree distributions
ddist1 <- igraph::degree.distribution(graph_ebf_spring1)
ddist2 <- igraph::degree.distribution(graph_ebf_spring2)

max_len <- max(length(ddist1), length(ddist2))

ddist1 <- c(ddist1, rep(0, max_len - length(ddist1)))
ddist2 <- c(ddist2, rep(0, max_len - length(ddist2)))

degree_seq <- 0:(max_len - 1)

df1 <- data.frame(
  Degree = as.factor(degree_seq),
  Fraction = ddist1
)

df2 <- data.frame(
  Degree = as.factor(degree_seq),
  Fraction = ddist2
)

p1 <- ggplot(data = df1, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- ggplot(data = df2, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

p1 / p2
```

#### Analysis with netAnalyze()

```{r}
props_ebf_spring <- netAnalyze(net_ebf_spring,
                             centrLCC = TRUE,
                             clustMethod = "cluster_fast_greedy",
                             hubPar = "eigenvector",
                             weightDeg = FALSE, 
                             normDeg = FALSE,
                             normBetw = FALSE,
                             normClose = FALSE)
```

### Graphlet correlation matrix

```{r ebf_spring_gcm, fig.width=10, fig.height=9, dpi = 200}
plotGCM(props_ebf_spring,
        legendCex = 1.2, 
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

### Network plot

```{r}
# Get phyla names
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]
table(phyla)

# Define phylum colors
phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)
```

```{r, fig.width=12, fig.height=6, dpi=250}
plot(props_ebf_spring,
     sameLayout = FALSE,
     title1 = "EBF = 0", 
     title2 = "EBF >= 2",
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.939999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 3, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```


```{r, fig.width=12, fig.height=6, dpi=250}
plot(props_ebf_spring,
     sameLayout = TRUE,
     layoutGroup = "union",
     title1 = "EBF = 0", 
     title2 = "EBF >= 2",
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.94,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 3, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

### Generate permuted association matrices

```{r generate permuted association matrices for spring, eval=FALSE}
permGroupMat <- createAssoPerm(props_ebf_spring, 
                               nPerm = 1000, 
                               computeAsso = FALSE, 
                               seed = 42)

nPerm_all <- 1000
blocksize <- 20
repetitions <- nPerm_all / blocksize


cores <- 8

plan(multisession, workers = cores)
registerDoFuture()

with_progress({
  p <- progressr::progressor(along = 1:repetitions)
  
  tmp <- foreach(i = 1:repetitions,
                 .packages = c("NetCoMi")) %dorng% {
    p()
    file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
    bmat_file <- paste0(file_base, ".bmat")
    if (!file.exists(bmat_file)) {
      set.seed(permutation_seed + i)
      NetCoMi::createAssoPerm(props_ebf_spring, nPerm = blocksize,
                              permGroupMat = permGroupMat[(i-1) * blocksize +
                                                            1:blocksize, ],
                              computeAsso = TRUE,
                              fileStoreAssoPerm = file_base,
                              storeCountsPerm = FALSE, append = FALSE)
    }
  }
})

# Reset to sequential execution
plan(sequential)
```


```{r, eval=FALSE}
# Combine the matrices and store them into a new file (because netCompare() 
# needs an external file)
assoPerm_all <- NULL

for (i in 1:repetitions) {
  file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
  
  assoPerm_tmp <- fm.open(filenamebase = file_base, readonly = TRUE)
  
  assoPerm_all <- rbind(assoPerm_all, as.matrix(assoPerm_tmp))
  
  close(assoPerm_tmp)
}

dim(assoPerm_all)
```
The combined matrix has 117000 rows and 234 columns, which is plausible because:

- n_rows = nPerm_all * n_taxa = 1000 * 117
- n_cols = n_groups * n_taxa * 2 * 117

We performed 1000 permutations, but want only 999 to get a round minimum possible
p-value. We therefore reduce the matrix to 999 * 117 = 116883 rows.

```{r, eval=FALSE}
# Keep the number of stored permutations consistent with netCompare().
n_perm_netcompare <- 999
n_taxa_netcompare <- nrow(props_ebf_main$input$adjaMat1)
n_rows_netcompare <- n_perm_netcompare * n_taxa_netcompare

stopifnot(nrow(assoPerm_all) >= n_rows_netcompare)
stopifnot(nrow(assoPerm_all) %% n_taxa_netcompare == 0)

assoPerm_netcompare <- assoPerm_all[seq_len(n_rows_netcompare), , drop = FALSE]
dim(assoPerm_netcompare)
```

```{r, eval=FALSE}
# Combine the permutation association matrices
fm.create.from.matrix(filenamebase = path_target("assoPerm_ebf"), 
                      mat = assoPerm_netcompare)
# >116883 x 234 filematrix with 8 byte "double" elements 
```

### Differential network analysis

We start by comparing local and global network characteristics between the two 
groups. This is done with NetCoMi' `netCompare()` function.

The statistical tests will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_netcompare_spring_noBH}
path_comp_ebf_main_noBH <- path_target("comp_ebf_main_noBH.rds")

if (file.exists(path_comp_ebf_main_noBH) && !refit_complete_network) {
  comp_ebf_main_noBH <- readRDS(path_comp_ebf_main_noBH)
} else {
  comp_ebf_main_noBH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_noBH, path_comp_ebf_main_noBH)
}

print(summary(comp_ebf_main_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))

summary(pvals_comp_main_noBH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

```{r run_netcompare_BH}
path_comp_ebf_main_BH <- path_target("comp_ebf_main_BH.rds")

if (file.exists(path_comp_ebf_main_BH) && !refit_complete_network) {
  comp_ebf_main_BH <- readRDS(path_comp_ebf_main_BH)
} else {
  comp_ebf_main_BH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_BH, path_comp_ebf_main_BH)
}

print(summary(comp_ebf_main_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

#### permApprox refinement without constraint

```{r run_netcompare_GPD_noConstr_noBH}
path_comp_ebf_GPD_noConstr_noBH <- path_target("comp_ebf_GPD_noConstr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_noBH <- readRDS(path_comp_ebf_GPD_noConstr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )
  comp_ebf_GPD_noConstr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_noBH, path_comp_ebf_GPD_noConstr_noBH)
}

tmp <- data.frame(emp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrEmpirical$pvalDiffDeg,
                  pApp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrRefined$pvalDiffDeg)

print(summary(comp_ebf_GPD_noConstr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_noConstr_BH}
path_comp_ebf_GPD_noConstr_BH <- path_target("comp_ebf_GPD_noConstr_BH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_BH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_BH <- readRDS(path_comp_ebf_GPD_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  comp_ebf_GPD_noConstr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_BH, path_comp_ebf_GPD_noConstr_BH)
}

print(summary(comp_ebf_GPD_noConstr_BH, 
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"), 
              numbNodes = 5))

summary(comp_ebf_GPD_noConstr_BH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

#### permApprox refinement with constraint

```{r run_netcompare_GPD_constr_noBH}
path_comp_ebf_GPD_constr_noBH <- path_target("comp_ebf_GPD_constr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_constr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_constr_noBH <- readRDS(path_comp_ebf_GPD_constr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  comp_ebf_GPD_constr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_noBH, path_comp_ebf_GPD_constr_noBH)
}

print(summary(comp_ebf_GPD_constr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_constr_BH}
path_comp_ebf_GPD_constr_BH <- path_target("comp_ebf_GPD_constr_BH.rds")

if (file.exists(path_comp_ebf_GPD_constr_BH) && !refit_complete_network) {
  comp_ebf_GPD_constr_BH <- readRDS(path_comp_ebf_GPD_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  comp_ebf_GPD_constr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_BH, path_comp_ebf_GPD_constr_BH)
}

print(summary(comp_ebf_GPD_constr_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

### Differential association analysis

In this step we compare the associations itself by generating and plotting differential networks, which have an edge for differentially associated 
genera only.

The networks will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_diffnet_noBH}
path_diff_ebf_main_noBH <- path_target("diff_ebf_main_noBH.rds")

if (file.exists(path_diff_ebf_main_noBH) && !refit_complete_network) {
  diff_ebf_main_noBH <- readRDS(path_diff_ebf_main_noBH)
} else {
  diff_ebf_main_noBH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_noBH, path_diff_ebf_main_noBH)
}

summary(diff_ebf_main_noBH$pvalsVec)
```

```{r run_diffnet_BH}
path_diff_ebf_main_BH <- path_target("diff_ebf_main_BH.rds")

if (file.exists(path_diff_ebf_main_BH) && !refit_complete_network) {
  diff_ebf_main_BH <- readRDS(path_diff_ebf_main_BH)
} else {
  diff_ebf_main_BH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_BH, path_diff_ebf_main_BH)
}

summary(diff_ebf_main_BH$pAdjustVec)
```

#### permApprox refinement without constraint

```{r run_diffnet_noConstr_noBH}
path_diff_ebf_noConstr_noBH <- path_target("diff_ebf_noConstr_noBH.rds")

if (file.exists(path_diff_ebf_noConstr_noBH) && !refit_complete_network) {
  diff_ebf_noConstr_noBH <- readRDS(path_diff_ebf_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  diff_ebf_noConstr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_noConstr_noBH, path_diff_ebf_noConstr_noBH)
}

summary(diff_ebf_noConstr_noBH$pvalsVec)
```

```{r run_diffnet_noConstr_BH}
path_diff_ebf_noConstr_BH <- path_target("diff_ebf_noConstr_BH.rds")

if (file.exists(path_diff_ebf_noConstr_BH) && !refit_complete_network) {
  diff_ebf_noConstr_BH <- readRDS(path_diff_ebf_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )
  
  diff_ebf_noConstr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_noConstr_BH, path_diff_ebf_noConstr_BH)
}

summary(diff_ebf_noConstr_BH$pAdjustVec)
```

#### permApprox refinement with constraint

```{r run_diffnet_constr_noBH}
path_diff_ebf_constr_noBH <- path_target("diff_ebf_constr_noBH.rds")

if (file.exists(path_diff_ebf_constr_noBH) && !refit_complete_network) {
  diff_ebf_constr_noBH <- readRDS(path_diff_ebf_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )

  diff_ebf_constr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_constr_noBH, path_diff_ebf_constr_noBH)
}

summary(diff_ebf_constr_noBH$pvalsVec)
```

```{r run_diffnet_constr_BH}
path_diff_ebf_constr_BH <- path_target("diff_ebf_constr_BH.rds")

if (file.exists(path_diff_ebf_constr_BH) && !refit_complete_network) {
  diff_ebf_constr_BH <- readRDS(path_diff_ebf_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  diff_ebf_constr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_constr_BH, path_diff_ebf_constr_BH)
}

summary(diff_ebf_constr_BH$pAdjustVec)
```


## Network comparison with SPIEC-EASI associations


### Network construction

```{r generate_net_ebfspieceasi}
path_net_ebfspieceasi <- path_target("net_ebf_spieceasi.rds")

if (file.exists(path_net_ebfspieceasi) && !refit_complete_network) {
  net_ebfspieceasi <- readRDS(path_net_ebfspieceasi)
} else {
  net_ebf_spieceasi <- NetCoMi::netConstruct(
    data = counts_ebf_rel,
    group = ebf_group,
    measure = "spieceasi",
    measurePar = list(method = "mb",
                      nlambda = 30,
                      pulsar.params = list(rep.num = 30, # Subsamples in StARS
                                           thresh = 0.1), # StARS threshold
                      norm.params = list(method = "clr",
                                         types = "tru_counts"),
                      symBetaMode = "ave"),
    jointPrepro = FALSE,
    filtTax = "totalReads",
    filtTaxPar = list(totalReads = 0.001),
    filtSamp = "none",
    normMethod = "none",
    zeroMethod = "bayesMult",
    zeroPar = list(z.warning = 0.99, z.delete = FALSE),
    sparsMethod = "none",
    seed = 42
  )
  
  saveRDS(net_ebfspieceasi, path_net_ebfspieceasi)
}
```


```{r}
# Extract association matrices
assomat_ebf_spieceasi1 <- net_ebf_spieceasi$assoMat1
assomat_ebf_spieceasi2 <- net_ebf_spieceasi$assoMat2
adjamat_ebf_spieceasi1 <- net_ebf_spieceasi$adjaMat1
adjamat_ebf_spieceasi2 <- net_ebf_spieceasi$adjaMat2

# Generate graph objectss
graph_ebf_spieceasi1 <- transform_asso(assomat_ebf_spieceasi1)$graph
graph_ebf_spieceasi2 <- transform_asso(assomat_ebf_spieceasi2)$graph
```

### Network analysis

#### Association histogram

```{r ebf_spieceasi_association_histrogram, fig.width=8, fig.height=4}
n_zero_asso1 <- sum(assomat_ebf_spieceasi1[lower.tri(assomat_ebf_spieceasi1)] == 0)
n_zero_asso2 <- sum(assomat_ebf_spieceasi2[lower.tri(assomat_ebf_spieceasi2)] == 0)

p1 <- data.frame(correlation = assomat_ebf_spieceasi1[lower.tri(assomat_ebf_spieceasi1)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso1), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "",
       y = "Count",
       ) +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- data.frame(correlation = assomat_ebf_spieceasi2[lower.tri(assomat_ebf_spieceasi2)]) |>
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.01) +
  annotate("text", x = 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("n = ", n_zero_asso2), size = 4) +
  coord_cartesian(ylim = c(0, 50), xlim = c(-0.2, 0.8)) +
  labs(x = "Estimated correlation",
       y = "Count",
  ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
  )

p1 / p2
```

#### Degree distribution

```{r ebf_degree_distribution, fig.width=8, fig.height=4, dpi = 200}
# Compute degree distributions
ddist1 <- igraph::degree.distribution(graph_ebf_spieceasi1)
ddist2 <- igraph::degree.distribution(graph_ebf_spieceasi2)

max_len <- max(length(ddist1), length(ddist2))

ddist1 <- c(ddist1, rep(0, max_len - length(ddist1)))
ddist2 <- c(ddist2, rep(0, max_len - length(ddist2)))

degree_seq <- 0:(max_len - 1)

df1 <- data.frame(
  Degree = as.factor(degree_seq),
  Fraction = ddist1
)

df2 <- data.frame(
  Degree = as.factor(degree_seq),
  Fraction = ddist2
)

p1 <- ggplot(data = df1, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

p2 <- ggplot(data = df2, aes(x = Degree, y = Fraction, group = 1)) +
  geom_line(color = "gray30") +
  geom_point(size = 2, color = "gray30") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

p1 / p2
```

#### Analysis with netAnalyze()

```{r}
props_ebf_spieceasi <- netAnalyze(net_ebf_spieceasi,
                             centrLCC = TRUE,
                             clustMethod = "cluster_fast_greedy",
                             hubPar = "eigenvector",
                             weightDeg = FALSE, 
                             normDeg = FALSE,
                             normBetw = FALSE,
                             normClose = FALSE)
```

### Graphlet correlation matrix

```{r ebf_gcm, fig.width=10, fig.height=9, dpi = 200}
plotGCM(props_ebf_spieceasi,
        legendCex = 1.2, 
        titleCex = 1.5,
        labCex = 1.2,
        sigTextCex = 1.3)
```

### Network plot

```{r}
# Get phyla names
taxtab <- as(tax_table(phylo_genus), "matrix")
phyla <- as.factor(taxtab[, "Phylum"])
names(phyla) <- taxtab[, "Genus"]
table(phyla)

# Define phylum colors
phylcol <- c(
  Actinobacteriota = "#E69F00",
  Bacteroidota = "#CC79A7",
  Desulfobacterota = "#00408A",
  Firmicutes = "#56B4E9",
  Fusobacteriota = "#F0E442",
  Proteobacteria = "#009E73",
  Verrucomicrobiota = "#D55E00"
)
```

```{r, fig.width=12, fig.height=6, dpi=250}
plot(props_ebf_spieceasi,
     sameLayout = FALSE,
     title1 = "EBF = 0", 
     title2 = "EBF >= 2",
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.939999,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 3, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```


```{r, fig.width=12, fig.height=6, dpi=250}
plot(props_ebf_spieceasi,
     sameLayout = TRUE,
     layoutGroup = "union",
     title1 = "EBF = 0", 
     title2 = "EBF >= 2",
     labelScale = FALSE,
     labels = network_labels,
     repulsion = 0.94,
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  phylcol,
     nodeSize = "clr",
     nodeSizeSpread = 4,
     cexNodes = 2,
     nodeTransp = 40,
     hubTransp = 40,
     posCol = "gray50", 
     negCol = "red3",
     showTitle = TRUE,
     cexLabels = 0.9,
     cexHubLabels = 0.9,
     hubBorderCol = "gray40",
     mar = c(1, 3, 3, 3))

# Colors used in the legend should be equally transparent as in the plot
phylcol_transp <- colToTransp(phylcol, 40)

# xpd = NA allows drawing outside the plot region into the margins
legend(-1.7, 1.1, cex = 1, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = phylcol_transp, bty = "n", pch = 16,
       xpd = NA) 

legend(0.65, 1.1, cex = 1, title = "Estimated association:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("gray50","red3"), 
       bty = "n", horiz = TRUE, xpd = NA)
```

### Generate permuted association matrices

```{r generate permuted association matrices, eval=FALSE}
permGroupMat <- createAssoPerm(props_ebf_main, 
                               nPerm = 1000, 
                               computeAsso = FALSE, 
                               seed = 42)

nPerm_all <- 1000
blocksize <- 20
repetitions <- nPerm_all / blocksize


cores <- 8

plan(multisession, workers = cores)
registerDoFuture()

with_progress({
  p <- progressr::progressor(along = 1:repetitions)
  
  tmp <- foreach(i = 1:repetitions,
                 .packages = c("NetCoMi")) %dorng% {
    p()
    file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
    bmat_file <- paste0(file_base, ".bmat")
    if (!file.exists(bmat_file)) {
      set.seed(permutation_seed + i)
      NetCoMi::createAssoPerm(props_ebf_main, nPerm = blocksize,
                              permGroupMat = permGroupMat[(i-1) * blocksize +
                                                            1:blocksize, ],
                              computeAsso = TRUE,
                              fileStoreAssoPerm = file_base,
                              storeCountsPerm = FALSE, append = FALSE)
    }
  }
})

# Reset to sequential execution
plan(sequential)
```


```{r, eval=FALSE}
# Combine the matrices and store them into a new file (because netCompare() 
# needs an external file)
assoPerm_all <- NULL

for (i in 1:repetitions) {
  file_base <- paste0("./data/10_networks/perm_asso_matrices/assoPerm_ebf", i)
  
  assoPerm_tmp <- fm.open(filenamebase = file_base, readonly = TRUE)
  
  assoPerm_all <- rbind(assoPerm_all, as.matrix(assoPerm_tmp))
  
  close(assoPerm_tmp)
}

dim(assoPerm_all)
```
The combined matrix has 117000 rows and 234 columns, which is plausible because:

- n_rows = nPerm_all * n_taxa = 1000 * 117
- n_cols = n_groups * n_taxa * 2 * 117

We performed 1000 permutations, but want only 999 to get a round minimum possible
p-value. We therefore reduce the matrix to 999 * 117 = 116883 rows.

```{r, eval=FALSE}
# Keep the number of stored permutations consistent with netCompare().
n_perm_netcompare <- 999
n_taxa_netcompare <- nrow(props_ebf_main$input$adjaMat1)
n_rows_netcompare <- n_perm_netcompare * n_taxa_netcompare

stopifnot(nrow(assoPerm_all) >= n_rows_netcompare)
stopifnot(nrow(assoPerm_all) %% n_taxa_netcompare == 0)

assoPerm_netcompare <- assoPerm_all[seq_len(n_rows_netcompare), , drop = FALSE]
dim(assoPerm_netcompare)
```

```{r, eval=FALSE}
# Combine the permutation association matrices
fm.create.from.matrix(filenamebase = path_target("assoPerm_ebf"), 
                      mat = assoPerm_netcompare)
# >116883 x 234 filematrix with 8 byte "double" elements 
```

### Differential network analysis

We start by comparing local and global network characteristics between the two 
groups. This is done with NetCoMi' `netCompare()` function.

The statistical tests will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_netcompare_noBH}
path_comp_ebf_main_noBH <- path_target("comp_ebf_main_noBH.rds")

if (file.exists(path_comp_ebf_main_noBH) && !refit_complete_network) {
  comp_ebf_main_noBH <- readRDS(path_comp_ebf_main_noBH)
} else {
  comp_ebf_main_noBH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_noBH, path_comp_ebf_main_noBH)
}

print(summary(comp_ebf_main_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))

summary(pvals_comp_main_noBH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

```{r run_netcompare_BH}
path_comp_ebf_main_BH <- path_target("comp_ebf_main_BH.rds")

if (file.exists(path_comp_ebf_main_BH) && !refit_complete_network) {
  comp_ebf_main_BH <- readRDS(path_comp_ebf_main_BH)
} else {
  comp_ebf_main_BH <- netCompare(props_ebf_main, 
                              permTest = TRUE, 
                              nPerm = 999,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(comp_ebf_main_BH, path_comp_ebf_main_BH)
}

print(summary(comp_ebf_main_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

#### permApprox refinement without constraint

```{r run_netcompare_GPD_noConstr_noBH}
path_comp_ebf_GPD_noConstr_noBH <- path_target("comp_ebf_GPD_noConstr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_noBH <- readRDS(path_comp_ebf_GPD_noConstr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none",
    thresh_method = "ftr"
  )
  comp_ebf_GPD_noConstr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_noBH, path_comp_ebf_GPD_noConstr_noBH)
}

tmp <- data.frame(emp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrEmpirical$pvalDiffDeg,
                  pApp = comp_ebf_GPD_noConstr_noBH$pvalDiffCentrRefined$pvalDiffDeg)

print(summary(comp_ebf_GPD_noConstr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_noConstr_BH}
path_comp_ebf_GPD_noConstr_BH <- path_target("comp_ebf_GPD_noConstr_BH.rds")

if (file.exists(path_comp_ebf_GPD_noConstr_BH) && !refit_complete_network) {
  comp_ebf_GPD_noConstr_BH <- readRDS(path_comp_ebf_GPD_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  comp_ebf_GPD_noConstr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_noConstr_BH, path_comp_ebf_GPD_noConstr_BH)
}

print(summary(comp_ebf_GPD_noConstr_BH, 
              groupNames = c("EBF = 0", "EBF >= 2"),
              showCentr = c("degree", "between", "closeness"), 
              numbNodes = 5))

summary(comp_ebf_GPD_noConstr_BH <- comp_ebf_main_noBH$pvalDiffCentrEmpirical$pvalDiffEigen)
```

#### permApprox refinement with constraint

```{r run_netcompare_GPD_constr_noBH}
path_comp_ebf_GPD_constr_noBH <- path_target("comp_ebf_GPD_constr_noBH.rds")

if (file.exists(path_comp_ebf_GPD_constr_noBH) && !refit_complete_network) {
  comp_ebf_GPD_constr_noBH <- readRDS(path_comp_ebf_GPD_constr_noBH)
} else {
  
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  comp_ebf_GPD_constr_noBH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "none",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_noBH, path_comp_ebf_GPD_constr_noBH)
}

print(summary(comp_ebf_GPD_constr_noBH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

```{r run_netcompare_GPD_constr_BH}
path_comp_ebf_GPD_constr_BH <- path_target("comp_ebf_GPD_constr_BH.rds")

if (file.exists(path_comp_ebf_GPD_constr_BH) && !refit_complete_network) {
  comp_ebf_GPD_constr_BH <- readRDS(path_comp_ebf_GPD_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  comp_ebf_GPD_constr_BH <- 
    netCompare(props_ebf_main, 
               permTest = TRUE, 
               nPerm = 999,
               fileLoadAssoPerm = path_target("assoPerm_ebf"),
               refinePermPvals = TRUE,
               permApproxParams = list(
                 method = "gpd",
                 gpd_ctrl = gpd_ctrl
               ),
               adjust = "BH",
               cores = 10,
               verbose = TRUE,
               seed = 42)
  
  saveRDS(comp_ebf_GPD_constr_BH, path_comp_ebf_GPD_constr_BH)
}

print(summary(comp_ebf_GPD_constr_BH, 
        groupNames = c("EBF = 0", "EBF >= 2"),
        showCentr = c("degree", "between", "closeness"), 
        numbNodes = 5))
```

### Differential association analysis

In this step we compare the associations itself by generating and plotting differential networks, which have an edge for differentially associated 
genera only.

The networks will be based on three different types of p-values:

* Empirical p-values
* Empirical p-values with permApprox refinement (without support constraints)
* Empirical p-values with permApprox refinement (with support constraints)

#### Empirical p-values

```{r run_diffnet_noBH}
path_diff_ebf_main_noBH <- path_target("diff_ebf_main_noBH.rds")

if (file.exists(path_diff_ebf_main_noBH) && !refit_complete_network) {
  diff_ebf_main_noBH <- readRDS(path_diff_ebf_main_noBH)
} else {
  diff_ebf_main_noBH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "none",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_noBH, path_diff_ebf_main_noBH)
}

summary(diff_ebf_main_noBH$pvalsVec)
```

```{r run_diffnet_BH}
path_diff_ebf_main_BH <- path_target("diff_ebf_main_BH.rds")

if (file.exists(path_diff_ebf_main_BH) && !refit_complete_network) {
  diff_ebf_main_BH <- readRDS(path_diff_ebf_main_BH)
} else {
  diff_ebf_main_BH <- diffnet(net_ebf_main,
                              diffMethod = "permute",
                              nPerm = 999L,
                              fileLoadAssoPerm = path_target("assoPerm_ebf"),
                              adjust = "BH",
                              cores = 10,
                              verbose = TRUE,
                              seed = 42)
  
  saveRDS(diff_ebf_main_BH, path_diff_ebf_main_BH)
}

summary(diff_ebf_main_BH$pAdjustVec)
```

#### permApprox refinement without constraint

```{r run_diffnet_noConstr_noBH}
path_diff_ebf_noConstr_noBH <- path_target("diff_ebf_noConstr_noBH.rds")

if (file.exists(path_diff_ebf_noConstr_noBH) && !refit_complete_network) {
  diff_ebf_noConstr_noBH <- readRDS(path_diff_ebf_noConstr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )

  diff_ebf_noConstr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_noConstr_noBH, path_diff_ebf_noConstr_noBH)
}

summary(diff_ebf_noConstr_noBH$pvalsVec)
```

```{r run_diffnet_noConstr_BH}
path_diff_ebf_noConstr_BH <- path_target("diff_ebf_noConstr_BH.rds")

if (file.exists(path_diff_ebf_noConstr_BH) && !refit_complete_network) {
  diff_ebf_noConstr_BH <- readRDS(path_diff_ebf_noConstr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "none"
  )
  
  diff_ebf_noConstr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_noConstr_BH, path_diff_ebf_noConstr_BH)
}

summary(diff_ebf_noConstr_BH$pAdjustVec)
```

#### permApprox refinement with constraint

```{r run_diffnet_constr_noBH}
path_diff_ebf_constr_noBH <- path_target("diff_ebf_constr_noBH.rds")

if (file.exists(path_diff_ebf_constr_noBH) && !refit_complete_network) {
  diff_ebf_constr_noBH <- readRDS(path_diff_ebf_constr_noBH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )

  diff_ebf_constr_noBH <- diffnet(net_ebf_main,
                                    diffMethod = "permute",
                                    nPerm = 999L,
                                    fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                    refinePermPvals = TRUE,
                                    permApproxParams = list(
                                      method = "gpd",
                                      gpd_ctrl = gpd_ctrl
                                    ),
                                    adjust = "none",
                                    cores = 10,
                                    verbose = TRUE,
                                    seed = 42)
  
  saveRDS(diff_ebf_constr_noBH, path_diff_ebf_constr_noBH)
}

summary(diff_ebf_constr_noBH$pvalsVec)
```

```{r run_diffnet_constr_BH}
path_diff_ebf_constr_BH <- path_target("diff_ebf_constr_BH.rds")

if (file.exists(path_diff_ebf_constr_BH) && !refit_complete_network) {
  diff_ebf_constr_BH <- readRDS(path_diff_ebf_constr_BH)
} else {
  gpd_ctrl <- permApprox::make_gpd_ctrl(
    constraint = "support_at_obs",
    sample_size = min(as.vector(table(ebf_group)))
  )
  
  diff_ebf_constr_BH <- diffnet(net_ebf_main,
                                  diffMethod = "permute",
                                  nPerm = 999L,
                                  fileLoadAssoPerm = path_target("assoPerm_ebf"),
                                  refinePermPvals = TRUE,
                                  permApproxParams = list(
                                    method = "gpd",
                                    gpd_ctrl = gpd_ctrl
                                  ),
                                  adjust = "BH",
                                  cores = 10,
                                  verbose = TRUE,
                                  seed = 42)
  
  saveRDS(diff_ebf_constr_BH, path_diff_ebf_constr_BH)
}

summary(diff_ebf_constr_BH$pAdjustVec)
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
