% !TeX root = ../thesis.tex

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

My deepest gratitude goes to my supervisor, Christian L. Müller. Although he wasn't my direct advisor at first, he was involved in my projects from a very early stage of my doctoral work. I can still hear my supervisor at the time, Anne-Laure Boulesteix, saying, ``There's a new professor at Helmholtz Munich who's doing exactly the same thing as you. Why don't you meet with him?'' Who would have guessed back then that this would lead to such a long and wonderful collaboration. I thank him for integrating me into his group from the very beginning, for supporting me in my projects at all times, for critically evaluating my work where necessary, for teaching me so much valuable knowledge, for facilitating my attendance at many great conferences, and above all, for always having a motivating word ready when I felt like I was at a loss.

Next, I would like to thank my second referee and former supervisor, Anne-Laure Boulesteix, for her wonderful support during the first years of my doctoral studies, and beyond. Throughout my entire doctoral program, I was able to turn to her with any questions, whether statistical or organizational, and always received a prompt response. Although the focus of my work shifted somewhat away from her direct area of expertise, she was always there for me, asked the right questions at the right times, and thus contributed significantly to improving my projects. 

One person who made my doctoral studies possible in the first place is Erika von Mutius. Although it wasn't originally planned that way, she agreed to expand my master's thesis into a doctoral dissertation and offered me a position at her institute at Helmholtz Munich. I thank her heartily for this, as well as for the many fruitful discussions about my projects, her input from a medical perspective, and our consistently good collaboration. 

I would also like to express my sincere gratitude to my third referee, Christophe \text{Ambroise}, for agreeing to review my work and for his patience, as the completion of my dissertation has been delayed beyond the original schedule. Thank you also to Martin Depner for the great collaboration during my time at Helmholtz Munich and for his valuable ideas, particularly regarding network analysis, which helped improve my projects. Moreover, I want to thank the the Munich Center for Machine Learning for the financial support. Thanks also to the administration of the Department of Statistics at LMU, especially to Brigitte Maxa, who was always there for me with my contract and organizational issues.

A big thank you also goes to the entire Biomedical Statistics and Data Science group at LMU, especially Mara Stadler, Medina Feldl, Viet Tran, Oleg Vlasovets, Roberto Olayo Alarcon, Daniele Pugno, and Johannes Ostner. I thank them for the many great conversations, for the productive discussions about my projects during group meetings, for the wonderful company on outings and trips (especially our U.S. trip), for all the support they provided in mentoring my seminar participants, and for being such a wonderful group that you won't find another like it anytime soon.

Last but not least, I would like to thank my family. A huge thank you goes to my mom, Cornelia Krügel, for always believing in me, for stepping back and giving me the space to focus on my dissertation, even though I know she would have liked to see me or speak to me on the phone much more often, and simply for always being there for me. And as for my husband, Mario Peschel, I hardly know where to begin. Maybe with the fact that I wouldn't have started my dissertation at all if he hadn't encouraged me to do so. Above all, I thank him for always finding a way to make me smile, even in my lowest moments, for always encouraging me to keep going, for never getting tired of showing me how proud he is of me, for never complaining when I spent long days at the computer rather than spending our free time together or going on vacation, and simply for making my life happier in so many ways.

\cleardoublepage